### 2012-11-22 articles/Article-RFI-distribution/article-rfi-distribution.tex
Importing RFI distribution paper

\documentclass[useAMS,usenatbib]{mn2e} 
%\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[font={it},labelfont={bf}]{caption}
%\usepackage[round]{natbib}
\usepackage{units}
%\usepackage[utopia]{mathdesign}
\usepackage{times} %A
%\usepackage{fourier} %B
%\usepackage{lmodern} % C
%\usepackage{mathpazo} % D
%\usepackage[T1]{fontenc}
%\usepackage{euler} % D
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

\usepackage{fixltx2e} % fixes placing of image positions

\definecolor{customhdrcolor}{rgb}{0.0,0.0,0.0}
%\definecolor{customcitecolor}{rgb}{0.0,0.25,0.25}
\definecolor{customcitecolor}{rgb}{0.0,0.5,0.75}
%\definecolor{customlinkcolor}{rgb}{0.0,0.0,1.0}
\definecolor{customlinkcolor}{rgb}{0.0,0.5,0.75}

\usepackage[colorlinks=true,linkcolor=customlinkcolor,urlcolor=customlinkcolor,citecolor=customcitecolor,pdftex]{hyperref}

\ifpdf\pdfinfo{/Title      (The brightness and spatial distributions of terrestrial radio sources)
               /Author     (A. R. Offringa et al.)
               /Keywords   (dark ages, reionization, first stars;atmospheric effects;instrumentation: interferometers;methods: observational;techniques: interferometric;radio continuum: general;rfi)
        }
\else\usepackage{graphics}\fi

\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\title[Brightness and spatial distributions of terrestrial radio sources]{The brightness and spatial distributions of terrestrial radio sources}

\author[A.~R.~Offringa et al.]{
  A.~R.~Offringa$^{1,2}$\thanks{E-mail:
\url{offringa@mso.anu.edu.au}},
  A.~G.~de~Bruyn$^{2,3}$,
   and S.~Zaroubi$^{2}$\\
$^{1}$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
$^{2}$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
$^{3}$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands
}

\begin{document}

\date{Accepted [TODO date]. Received [TODO date]; in original form [TODO date]}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\pubyear{2012}

\maketitle
\label{firstpage}

\begin{abstract}
Feeble undetected sources of radio-frequency interference (RFI) might become visible after integration of long radio observations if they are coherently present over time. Thereby, they might occlude the astronomical signal of interest. This issue is especially important for epoch of reionization (EoR) projects that try to detect the very faint redshifted HI signals of the earliest objects in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR. The histograms shows significant features, of which a power law with estimated exponent around -1.6 to -1.5 is most striking. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows the Hata propagation model. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. Since these observations have achieved the thermal (sky) noise limit, a logical implication is that most of the RFI sources do not add coherently.
\end{abstract}

\begin{keywords}
atmospheric effects -- instrumentation: interferometers -- methods: observational -- techniques: interferometric -- radio continuum: general -- dark ages, reionization, first stars
\end{keywords}

\section{Introduction}
Radio astronomy concerns itself with the observation of radiation from celestial sources at radio wavelengths. However, astronomical radio observations can be affected by radio-frequency interference (RFI), which makes it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}. Many techniques have been designed to mitigate its effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, LOFAR-RFI-pipeline}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}. In most cases, these methods excise enough of the interference to calibrate and image the data.

Regular radio observations record one or a few sessions of typical a few hours and results are integrated. In these cases, it often suffices to only excise interference that is apparent and thus above the noise to reach the thermal noise limit of the instrument. A new challenge arises when one desires much deeper observations, and hunderds of hours of observations need to be integrated. In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but is \index{coherency (RFI)} coherently \index{RFI!coherency} present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for 21-cm Epoch of Reionisation (EoR) experiments, because these involve long integration times. Several such experiments are underway, both to measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or global signal \citep{edges}. For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array \citep{de-bruyn-eor-ursi-2011}.

This work explores the information that is present in interference distributions, in order to analyse possible low-level interference that is not detectable by standard detection methods. Our approach is similar to the radio source count analysis that is used in cosmology \citep{vonhoerner-radio-source-counts}, also named \index{log N--log S plot} $\log N$ -- $\log S$ analysis, where $N$ and $S$ refer to the celestial source count and brightness respectively. The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe. We analyse such a double-logarithmic plot for the case of terrestrial sources, with the ultimate goal of predicting their full spatial and brightness distributions. This results in a better insight in the effects of low-level interference and allows one to simulate the effects of interference more accurately. A distribution can be expressed either as a differential or cumulative function. This work uses differential distributions, because they are better suited for visualizing the parameters we are interested in, although the two basically convey the same information.

%TODO maybe also discuss atmospheric (lightning), e.g. see 'Harry M. Hall, A New Model for ``Impulsive'' phenomena: application to atmospheric-noise communication channels', as contains also discussion on energy distributions.

In Sect.~\ref{sec:distribution-prediction}, we predict the terrestrial interfering source distribution based on various assumptions. Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse the histograms. This is followed in Sect.~\ref{sec:dist-data} by a short description of the two LOFAR data sets that have been used to perform the experiment. The results of analysing the sets are presented in Sect.~\ref{sec:dist-results}. Finally, in Sect.~\ref{sec:dist-discussion} the results are discussed and conclusions are drawn.

\section{Predicting the brightness distribution} \label{sec:distribution-prediction}
\index{brightness distribution}\index{spatial distribution}\index{power law}\index{transmitters!distribution}
Interference is generated by many different kind of transmitters, and these will have different spatial and brightness distributions (``spatial'' refers here to the distribution on the Earth). For example, \index{aeroplanes} aeroplanes and \index{satellite} satellites have widely different heights, while other sources are ground-based. Even ground-based sources might be spread differently. For example, it can be expected that \index{CB devices} citizens' band (CB) devices, that are often used in cars, are differently distributed from broadcasting transmitters. The frequencies of deliberately transmitting sources of a particular class of devices are constrained by the bands that have been allocated for the given class, which might allow one to distinguish different sources to some extent.

\index{morphology}
In a time-frequency plot, interfering sources can have complex structures. They can also be intermittent and different sources might overlap in time-frequency space. An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a time-frequency diagram (dynamic spectrum). Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in the time-frequency diagram. It is often not clear what constitutes a single interfering source, hence it is hard to count individual sources. Instead, we will count the number of times a given brightness occurs in time-frequency space. This --- as well as many other effects --- will of course influence the distribution. First we will derive the expected intrinsic source distribution. After that, we will analyse the issues that arise when measuring the distribution by counting samples.

\begin{figure*}
\begin{center}
\includegraphics[height=8cm]{img/TF-L43786_L220-rfi-example.pdf}
\caption{A small part of an observation displayed in a time-frequency plot. The features with significantly higher values are caused by interference. Some of these have a constant frequency, while others are more erratic.}
\label{fig:dist-rfi-example}
\end{center}
\end{figure*}

In every dynamic spectrum we can measure the number of times that the flux density is within a particular range. Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity. We will refer to this quantity with the statistical term \index{frequency density} ``frequency density'' (not to be confused with the physical radiation frequency). We will now start by deriving a prediction of the frequency density function of ground-based interfering sources. Consider an interfering point source of strength $I$. This source is observed by an interferometer that consists of two antennae or stations with gains $g_1,g_2$, which include all instrumental effects. The antennae are located at distances $r_1,r_2$ from the source. The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
\begin{equation}
 \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{r_1 r_2},
\end{equation}
with $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$. Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth. We will limit our analysis to cross-correlated antennae; the auto-correlations will be ignored.

We assume that the source is observed fully coherent, but a possible decoherence factor can be absorbed in the gains. Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the \index{narrow-band condition} narrow-band condition. For example, if the bandwidth of the signal $\Delta\nu=1$ kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c / (2 \pi \Delta \nu) \approx$ 50 km.

Now, we can treat the interferometer geometrically as a single point, as both antennae will see the same distribution. Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and gain $g$ as
\begin{equation} \label{eq:amplitude-fall-off-in-free-space}
 S(r) = \frac{Ig}{r^2}.
\end{equation}

Next, we assume that all sources have equal constant strength $I$. Moreover, we assume that the sources have a uniform spatial distribution in a two-dimensional plane. This is obviously a simplification, as the sources are actually distributed on the surface of a sphere. Beyond some distance, the Earth will also partly block the path between transmitter and receiver. Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid and only for a limited distance. Using these assumptions, we can express the expected cumulative frequency density of sources $f_I(r)$ at distance $r$ as
\begin{equation} \label{eq:cumulative-distance-function}
F_{\textrm{distance} \le r}(r)=c\pi r^2,
\end{equation}
for some constant $c$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance of $r$ away.

We need the reverse of Eq.~\eqref{eq:cumulative-distance-function} to predict the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance. Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - c\pi r^2$, with $N$ the total number of sources.

The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can now be calculated with
\begin{align} \label{eq:cumulative-rfi-distribution}
 F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{S}^{-1}(S))\\
\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{S})\\
\notag = & N-\frac{c\pi Ig}{S}
\end{align}
where $\mathcal{S}^{-1}$ refers to the inverse of $\mathcal{S}(r)$, i.e., the function that returns the distance $r$ for a given amplitude $S$. Finally, the differential frequency density can be calculated by taking the derivative,
\begin{align} \label{eq:two-dimensional-distribution}
 f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{c\pi Ig}{S^2}.
\end{align}

Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we predict to see a \index{slope (log-log plot)} slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane.

\subsection{Spherical case} \index{spherical distribution}
If we do take into account the fact that the sources lie on a sphere, Eq.~\eqref{eq:cumulative-distance-function} changes. In this case, the number of sources corresponds with the surface area of a hemisphere. By using the formula for the surface area of a hemisphere and some basic geometry, one finds that
\begin{equation}\label{eq:cumulative-spherical}
\tilde{F}_{\textrm{distance} \le r} = c 2\pi R^2\left(1-\cos(2\arcsin \frac{r}{2R})\right) \hspace{0.4cm}\textrm{with } r \le 2R,
\end{equation}
with $R$ the radius of the Earth. Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated. Therefore, the spherical and 2-dimensional cases lead to a distribution with equal power law. Evaluating Eq.~\eqref{eq:cumulative-spherical} is limited to $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth. This causes distant sources to settle below the horizon. Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.

\subsection{Propagation effects} \index{propagation (EM)}
So far, we have assumed that the electromagnetic radiation propagates through \index{free space} free space, resulting in a $r^{-2}$ fall-off. In reality, the radiation will be affected by complicated propagational effects. Because of the irregular surface of the Earth (including urban areas) and the absence of a direct \index{line of sight} line of sight between transmitter and receiver, the propagation mode might be indirect. Multiple indirect paths might be formed by reflection, refraction or diffraction of the electromagnetic wave.

A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. \citeauthor{hata-propagation-loss} gives the following analytical estimate for $L_p$, the electromagnetic propagation loss over land:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in m; $h_m$ the height of the receiving antenna in m; $r$ the distance between the antennae in m; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the \index{urban density|see{population density}} urban \index{population density} density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with \index{transmitters!height} transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

Collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, and converting from a subtrahend in decibels to a flux density factor $L_S$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
The model is based on urban areas, but corrections are given by \citeauthor{hata-propagation-loss} for sub-urban areas with a lower population density and for open areas. These corrections are independent of $r$, thus would only change $\zeta$. Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{\zeta Ig}{r^\eta},
\end{equation}
with $\eta = 0.655 \log_{10} h_b - 4.49$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(c*%pi*r^2,S),solve(S(r)=S,r);
\begin{align}
f_p(S)
= & \frac{d}{dS} c \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta}
= \frac{c 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free space propagation effects, the observed log-log histogram is predicted to have a $\frac{2}{\eta}-1$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for \index{transmitters!height} transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

\subsection{Noise contribution} \label{sec:histogram-noise}\index{noise}
The full measured distribution will consist of the power-law distribution combined with that of the noise and the celestial signal. For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong sources. Noise, however, will have a contribution. The real and imaginary components of receiver noise are independent and identically Gaussian distributed. Consequently, an amplitude $x$ will be Rayleigh distributed: \index{Rayleigh distribution}
\begin{equation}\label{eq:rayleigh-formula}
f_\textrm{noise}(x) =
\begin{cases}
\frac{x}{\sigma^2} e^{\frac{-x^2}{2\sigma^2}} & x > 0, \\
0 & \textrm{otherwise.}
\end{cases}
\end{equation}
Because most of the samples will be unaffected by RFI, this will be the dominating distribution. The Rayleigh distribution is plotted together with the -2 power-law distribution of Eq.~\eqref{eq:two-dimensional-distribution} in Fig.~\ref{fig:rayleigh-and-rfi-distributions}.

\begin{figure}
\begin{center}\hspace{-5mm}\includegraphics[width=8cm]{img/plot-rayleigh-and-rfi-trimmed}
\caption{The Rayleigh and power-law distributions in a log-log plot. The power-law distribution (Eq.~\eqref{eq:two-dimensional-distribution}) has a constant slope of -2. The slope of the Rayleigh distribution in the limit of the origin is 1. Its maximum occurs where the amplitude value equals its mode $\sigma$, which is 1 in this example. For higher amplitudes, its slope decreases exponentially.}
\label{fig:rayleigh-and-rfi-distributions}
\end{center}
\end{figure}

So far, these are the expected histograms for pure noise and pure RFI that propagates through free space. However, the measured distribution is a mixture of the two. A perfect RFI detector would separate the samples in two distributions; one that is not affected by RFI, and therefore contains noise only, and one that is the sum of RFI and noise. Real RFI detectors can separate these distributions to some extent, but due to false positives and false negatives, the histograms will get mixed nevertheless. Even more, we still need to take into account that the RFI classified samples are also affected by noise. Although this effect is moderate, because 'most' RFI samples are of much higher amplitude than what is added because of the noise, it is the low-level RFI samples which we are interested in, and the relative effect of noise on those samples is large.

\begin{figure}
\begin{center}\hspace{-2mm}\includegraphics[width=8.5cm]{img/plot-rayleigh-and-rfi-combined-trimmed}
\caption{Histograms of simulated samples that all have a contribution of noise and RFI. Various settings of the parameters were used, and samples were drawn as described in Eq.~\eqref{eq:drawing-amplitudes}. Solid lines: the combined distributions, dashed lines: the original distributions before mixing. }
\label{fig:rayleigh-and-rfi-combined}
\end{center}
\end{figure}

To understand the RFI-with-noise distribution, one needs to consider the individual contributions of the real and imaginary components. The real and imaginary components of samples that are affected by RFI, will follow a \index{bivariate distribution} bivariate distribution: each complex component is the sum of a complex component of RFI and a noise sample. As said, the noise comes from a Gaussian distribution with zero mean. The RFI distribution is assumed to be a power law with yet unknown exponent.

To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as algebraic calculation is not trivial. The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes. A sample can be drawn from the RFI distribution by normalizing and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable. Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity. Therefore, to \index{drawing samples} draw a sample by using the \index{inversion method} inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$. By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-c \pi r_L^2$ for $S < Ig/r_L^2$. Consequently, by sampling a uniform value $x_u$ with $0 \le x_u \le N-c \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ the power-law distribution.

A single complex RFI contaminated sample $S_{\textrm{RFI}}$ can then be sampled with
\begin{equation} \label{eq:drawing-rfi-distribution}
\textrm{real}(S_{\textrm{RFI}})\leftarrow \frac{Ig}{x_u r_L^2}\textrm{,\hspace{1mm} imag}(S_{\textrm{RFI}}) \leftarrow \frac{Ig}{x_u r_L^2},
\end{equation}
with $x_u \sim U(0, 1)$, and the corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
\begin{equation} \label{eq:drawing-combined-distribution}
\textrm{real}(S) \leftarrow x_n + \frac{Ig}{x_u r_L^2}
\textrm{,\hspace{1mm} imag}(S) \leftarrow x_n + \frac{Ig}{x_u r_L^2},
\end{equation}
with $x_n \sim N(\mu=0; \sigma=1)$. Of course, the final amplitude sample $S_A$ can be calculated with 
\begin{equation} \label{eq:drawing-amplitudes}
S_A \leftarrow |S| = \sqrt{\textrm{real}({S})^2+\textrm{imag}({S})^2}.
\end{equation}
An example of distribution curves for various settings of $Ig/r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.

If we assume that $I$, the average source strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-combined-distribution} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram. Although calibration could in theory solve $g$, almost all sources will come in through the edges of the beam, and finding the expected values for the gains is therefore hard. The effects of these parameters on the histogram will be further discussed in \S\ref{sec:parameter-variability}.

Fig.~\ref{fig:rayleigh-and-rfi-combined} shows that the left side of the graph corresponds with the original Rayleigh distribution. From the right side of the histogram, we are able to estimate $Ig/r_L^2$, as the placing of the power-law curve is independent of the Rayleigh distribution for large $S$. Methods to constrain the parameters of the RFI distribution will be discussed in \S\ref{sec:rfi-distribution-constraints}. The curves in Fig.~\ref{fig:rayleigh-and-rfi-combined} are from histograms of samples that are all contaminated by both noise and RFI. In practice, we can not make this distinction 100\% accurately, as RFI detectors have a limited accuracy. Consequently, the observed histogram will be the sum of two types of histograms: the first histogram being contaminated by both noise and RFI, the second one only by noise.

% Exponential:
% fs(x):=c*%pi*I*g/x^2;
%
% Gaussian:
% fn(x):=1/(sqrt(2*%pi)*sigma)*exp(-x^2/(2*sigma^2));
%
% Convolution:
% fc(x):=romberg(fs(x)*fn(x+tau), tau, -inf, inf);
%
% plot2d([log(fc(x)),log(fs(x)),log(fn(x))], [x,0.01, 100], [y,-10,10], [logx]), sigma=1, I=1, g=1, c=1;

\subsection{Parameter variability} \label{sec:parameter-variability}
In reality, the parameters $c$, $I$ and $g$, which are the source density per unit area, source strength and instrumental gain respectively, will not be constant over time and frequency but have a stochastic nature. However, since each specific value for these parameters produces a power law, the combined distribution will still show a power law, as long as the parameters follow a distribution that is steep at high amplitudes (in $\log$--$\log$ space), such as a Gaussian or uniform distribution.

%\begin{figure}[tb]
%\begin{center}
%\includegraphics[width=12cm]{img/plot-passband.pdf}
%\caption{Effect of the low and high-band frequency responses on the amplitude histogram for different distributions.}
%\label{fig:plot-passband-histogram}
%\end{center}
%\end{figure}

One instrumental effect that is absorbed in $g$ is the \index{frequency response|see{band-pass}} frequency response of the instrument, i.e., the antenna response in combination with the passband of the analogue and digital filters. Because the data that are analysed in Sect.~\ref{sec:dist-results} have initially not been band-pass calibrated, the instrumental response is not uniform over frequency. We determined that the variation due to the \index{band-pass} band-pass is about one order of magnitude in the LBA and about a factor of two in the HBA.

The effect of the band-pass on the data distribution is consequently limited to one order of magnitude or less. If the RFI sources have an apparent power-law distribution with a certain exponent, the distribution will have a feature at low amplitudes due to the frequency response, but a power law with equal exponent will still appear when the distribution is observed over a wide enough range.

%In Fig.~\ref{fig:plot-passband-histogram}, the histograms of the frequency response are given. To also demonstrate the effect of mixing distributions, histograms are added for samples that are drawn from a uniform, Gaussian and -2 power-law distribution as in Eq.~\eqref{eq:drawing-rfi-distribution} with $Ig/r_L^2=1$, that have been multiplied by the passband. The frequency responses were determined by measuring the standard deviation of the data over frequency after flagging the data for interference. As expected, it can still be determined that the intrinsic distribution is a -2 power law.

\index{beam}\index{LOFAR!beam}
Another effect that is absorbed in $g$, is the beam of the instrument. At the point of writing, LOFAR beam models are still being developed and are not yet well parametrized. However, since it is likely that most RFI sources are observed at the edges of the beam, it can be expected that the beam will have a benign effect on the histogram of an observation, comparable with the effect of the frequency response.

\section{Methods} \label{sec:distribution-methods}
In this section we will briefly discuss how the histograms are created, how the slope of the underlying RFI distribution is estimated and show how to constrain some of the intrinsic RFI parameters.

\subsection{Creating a histogram} \label{sec:histogram}\index{histogram}
While creating a histogram is trivial, it is important to note that it is necessary to have a variable bin size. This is mandated by the large dynamic range of the histogram that we are interested in. Therefore, we chose to have a bin size that increases linearly with the amplitude $S$, and the frequency distribution is normalized by the bin size after counting. Consequently, in parts of the histogram that have a sparse number of samples, the outlying samples will follow a $1/S$ curve, or a -1 slope in a log-log plot. This can be seen in the tails of the distributions of Fig.~\ref{fig:rayleigh-and-rfi-combined}. This, however, is not an intrinsic feature of the data but a consequence of the \index{binning} binning method.

\subsection{Estimating the slope}\index{slope}
An automated procedure is used to calculate the slope of the observed log-log histogram. First, the \index{mode (Rayleigh)} mode $\hat\sigma$ of the Rayleigh distribution is estimated by finding the amplitude with the maximum occurrences, i.e., the amplitude corresponding to the peak of the histogram. Then, the maximum amplitude $S_\textrm{max}$ is calculated, i.e., the sample with highest amplitude. Finally, the slope is estimated using \index{linear regression} linear regression over the interval $R$,
\begin{equation}
R=\left[20\hat\sigma; 10^{\frac{1}{2}\left(\log_{10} 20\hat\sigma + \log_{10} S_\textrm{max}\right) } \right].
\end{equation}
In terms of the log-log histogram, the start of the interval corresponds with the point that is $1.3$ units to the right of the peak in the histogram, and the interval end corresponds with the point halfway between the start point and the maximum amplitude. This interval starts near the start of the RFI slope but past where the Rayleigh curve dominates. It also does not include the noisy data at the end of the curve.% TODO The curves are not always completely straight in this interval, but since we use the same stable way to determine the interval over which to calculate the slope, we can at least compare them.

\index{power law}
One should note that fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}. However, because of our enormous sample size, which allows fitting the line over a large interval, the estimator will be sufficiently accurate for our purpose. Nevertheless, we will additionally calculate a maximum likelihood estimator for comparison. The maximum-likelihood estimator for the exponent in a power-law distribution is given by the \index{Hill estimator} Hill estimator \citep{hill-estimator, power-law-distribution}, defined as:
\begin{equation} \label{eq:hill}
 \alpha_H = 1 + N \left(\sum\limits_{i=1}^{N} \ln \frac{x_i}{x_\textrm{max}} \right)^{-1}.
\end{equation}
However, this estimator assumes the distribution is not cut-off at a high point. In our case, the distribution is cut-off, for example because of the limit of the analogue-digital converter (see \S\ref{sec:rfi-distribution-constraints}). Therefore, using this estimator will result in an estimate that is steeper (i.e., more negative) than the actual distribution. Moreover, the estimator requires an iteration over all data points, and because of the size of the data sets this is somewhat impractical. However,  we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ the bin frequency. Because the bin size is rather small, the resulting estimate will be close to the normal Hill estimator and one does not have to iterate over all data.

\subsection{Determining RFI distribution limits} \label{sec:rfi-distribution-constraints}
In this section we will show methods to find $S_U$ and $S_L$, the upper and lower flux limits of the power-law distribution at which the power law breaks down. Once the exponent of the power-law part of the distribution is estimated with the previously discussed techniques, the distribution can be extrapolated to find a lower flux limit. Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the following frequency density function $g$:
\begin{equation} \label{eq:amplitude-power-law}
g(S) = \beta S^\alpha.
\end{equation}
\begin{figure}
\begin{center}
\includegraphics[width=7.5cm]{img/explanation-lower-constraint/FindingCounts}
\includegraphics[width=7.5cm]{img/explanation-lower-constraint/FindingConstraints}
\caption{Cartoon of how a constraint on the lower fall-off point of the power-law distribution can be determined. Note that the labelled areas are areas as occupied in a linear plot, i.e., the integration of the density function. Areas in a log-log plot are not linearly related to the integral. There are two ways to estimate the lower constraint $S_L$: (i) the areas $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal if $Ig/r^2$ is constant during the observation, and (ii) if one assumes $Ig/r^2\sim$ uniform, then $N_a+N_b = N_\textrm{total}-N_\textrm{part}$.}
\label{fig:explanation-constraints}
\end{center}
\end{figure}

Assume also that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}. The hypothetical upper limit $S_U$ of the distribution can now be found, i.e., the highest amplitude that would be observed when the distribution follows the power law up to that point, by solving
\begin{equation} \label{eq:upper-limit-constrained}
 \int\limits_{S_1}^{S_U} g(S) dS = N_\textrm{part}.
\end{equation}
In practice, the observed histogram will break down beyond some amplitude because of several reasons. First and most importantly, the samples itself are digitized with an analogue-to-digital converter (ADC) with limited range. Second, we observe for a limited time and the frequency count is discrete. Because the chance of finding a sample with a very high intensity is low, it is unlikely to observe samples beyond some amplitude within the finite observing time. Finally, under the assumption of a uniform spatial distribution of RFI transmitters, samples with very high amplitude would have to be produced by transmitters that are very close to the telescope. However, it is likely that the uniform spatial distribution of transmitters will break down at closer distances.

Solving Eq.~\eqref{eq:upper-limit-constrained} results in
\begin{equation} \label{eq:upper-limit}
S_U = \sqrt[\alpha+1]{\frac{\alpha+1}{\beta} N_\textrm{part} + S_1^{\alpha+1}}.
\end{equation}
In some cases there will be no solution for $S_U$. The reason for this is that the integral of the distribution function converges, and the total number of samples is finite even when $S \rightarrow \infty$. If $N_\textrm{part}$ exceeds the integrated value of the fitted distribution function, there is no solution. This can happen when the empirical distribution is not limited at the high end, or when it contains features that are not in the model.

Similar to the calculation of $S_U$ and by using $N_\textrm{total}$ as an upper limit to the total number of samples that are affected by the power-law distribution, one can estimate the lower limit $S_L$. For Fig.~\ref{fig:explanation-constraints} this means that the areas labelled $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal. Solving this equation, one finds that
\begin{equation} \label{eq:lower-limit-1}
S_L = \sqrt[\alpha+1]{\frac{\alpha+1}{\beta}\left(N_\textrm{part} - N_\textrm{total} \right) + S_1^{\alpha+1}}.
\end{equation}

With the assumption that $Ig/r_L^2\sim$ a uniform distribution, the area labelled in Fig.~\ref{fig:explanation-constraints} as $N_b$ is also part of the RFI distribution, and a stronger constraint $\tilde S_L$ can be found by solving
\begin{equation}
 \int\limits_{\tilde S_L}^{S_U} g(S) dS + \tilde S_{L} g(\tilde S_{L}) = N_\textrm{total} - N_\textrm{part},
\end{equation}
which yields
\begin{equation} \label{eq:lower-limit-2}
\tilde S_L = \sqrt[\alpha+1]{ - \frac{1}{\alpha} \left( \frac{\alpha+1}{\beta} \left(N_\textrm{part} - N_\textrm{total}\right) + S_1^{\alpha+1} \right) }.
\end{equation}

With estimates of $\alpha$, $\beta$, $S_L$ and $S_U$, one has obtained a parametrization of the RFI distribution. As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is $Ig/r^2$ -- assuming free space propagation for the moment -- and therefore $S_L = Ig/r^2$. For Hata's propagation model, the more generic solution $S_L = \zeta Ig / r^\alpha$ is found. This value represents the apparent brightness of the sources that are the furthest away from the telescope. With a fully parametrized distribution of the effect of RFI sources, an empirical model for RFI effects can be made. Moreover, one can calculate $E(S_R)$, the expected apparent strength of RFI:
\begin{align} \label{eq:expected-value-rfi-def}
E(S_R) & = \frac{1}{N_{LU}}\int\limits_{S_L}^{S_U} \beta S^\alpha S dS
 = \frac{\beta}{N_{LU}} \left[ \frac{1}{\alpha+2} S^{\alpha+2} \right]_{S_L}^{S_U} \\
 & = \frac{\beta \left( S_U^{\alpha+2} - S_L^{\alpha+2} \right) }{N_{LU} \left( \alpha+2\right) }
\end{align}
Here, $N_{LU}$ is the number of samples between $S_L$ and $S_U$ after normalizing for the bin size:
\begin{align} \label{eq:n-between-sl-su}
N_{LU} = \int\limits_{S_L}^{S_U} \beta S^\alpha dS.
\end{align}
Substitution and simplification of these two results in
\begin{align} \label{eq:expected-value-rfi}
 E(S_R) & = \frac
{\left( S_U^{\alpha+2} - S_L^{\alpha+2}\right)\left(\alpha+1\right)} 
{\left(S_U^{\alpha+1} - S_L^{\alpha+1}\right) \left(\alpha+2 \right)}.
\end{align}
This is essentially the flux density that is caused by RFI without using RFI detection or excision algorithms. $E(S_R)$ has the same units as $S_L$ and $S_U$, thus after calibration (see \S\ref{sec:calibration}) could be given in Jy. In practice, the increase of system noise after correlation is much less severe because of RFI flagging, which excises a part of the RFI. One can assume that all RFI above a certain power level is found by the detector. Since modern RFI detection algorithms can find all RFI that is detectable ``by eye'' \citep{post-correlation-rfi-classification}, this power level will be near the level of the noise mode. In
Offringa et al. (in preparation), % TODO fill in citation if available
the false-positives rate for the AOFlagger is estimated to be 0.5\%. An estimate of $S_d$, the average lower limit of detected RFI, can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the ``real number'' (true positives) of RFI samples. Therefore, the limit is calculated similar to Eq.~\eqref{eq:lower-limit-1}, where the term $N_\textrm{part} - N_\textrm{total}$ needs to be replaced with $N_\textrm{RFI}$, which equals the total number of samples detected as RFI minus the 0.5\% false positives.

Finally, $E(S_\textrm{leak})$, which is the expected value of leaked RFI not detected by the flagger, can be calculated by replacing $S_U$ with $S_d$ in the numerator of Eq.~\eqref{eq:expected-value-rfi} and subtracting the removed number of samples from the total of number of samples. Assume that a fraction of $1-\eta$ samples have been detected as RFI, then
\begin{align} \label{eq:expected-value-leaked-rfi}
E(S_\textrm{leak}) & = \frac{1}{\eta N_{LU}}\int\limits_{S_L}^{S_d} \beta S^\alpha S dS
= \frac{\left( S_d^{\alpha+2} - S_L^{\alpha+2} \right) \left( \alpha+1 \right) } {\eta \left( S_U^{\alpha+1} - S_L^{\alpha+1}\right) \left( \alpha+2 \right)}.
\end{align}
This is the average contribution that leaked RFI will have on a single sample. It has the same units as the parameters $S_L$, $S_U$ and $S_d$. Typical values for $\eta$ are $0.95$--$0.99$.

\subsection{Calibration} \label{sec:calibration}
We can assign flux densities to the horizontal axis of the histogram by using the system equivalent flux density (SEFD) of a single station. The current \index{SEFD (LOFAR)} LOFAR SEFD is found to be approximately 3000~Jy for the HBA core stations and 1500~Jy for the remote stations in the frequency range from $125$--$175$ MHz. For all Dutch LBA stations, in the frequency range $40$--$70$ MHz the SEFD is approximately 30,000~Jy. The standard deviation $\sigma$ in the real and imaginary values is related to the SEFD with % TODO citation
\begin{equation}
 \sigma = \frac{\textrm{SEFD}}{\sqrt{2 \Delta \nu \Delta t}},
\end{equation}
where $\Delta \nu$ is the bandwidth and $\Delta t$ is the correlator integration time. The standard deviation will appear as the mode of the Rayleigh distribution. By fitting a Rayleigh function with fitting parameter $\sigma$ to the distribution, one finds the corresponding flux density scale.

RFI sources will enter through the distant sidelobes of the station beams from many unknown directions. Moreover, models for the full beam are often hard to construct. Therefore, we will not try to calibrate the beam, and the flux densities in the histogram are apparent quantities. Consequently, we will not be able to say something about the true intrinsic power levels of RFI sources.

\subsection{Error analysis}
An estimate for the standard deviation of the slope estimator $\hat \alpha$ can be found by calculating $\textrm{SE}(\hat \alpha)$, the \emph{standard error} of $\hat \alpha$. The standard error of the slope of a straight line \citep[pp. 32--35]{acton-analaysis-of-straight-lines} is given by
\begin{equation} \label{eq:stderr-slope}
 \textrm{SE}(\hat \alpha) = \sqrt{\frac{SS_{yy}-\hat\alpha SS_{xy}}{\left(n - 2\right) SS_{xx}}},
\end{equation}
where $SS_{xx}$, $SS_{xy}$ and $SS_{yy}$ are the sums of squares, e.g., $SS_{xy}=\sum_{i=1}^n (x_i - \bar x) (y_i - \bar y)$ and $n$ is the number of samples. However, we found that this is not a representable error in our case, because the errors in the slope are not normally distributed. The noise in the part of the histogram over which the slope is calculated, is very small due to the large amount of samples. Consequently, the estimated standard deviation of $\hat \alpha$ is very low. Nevertheless, when the slope is calculated over subsections of the original range over which the slope is calculated, one finds the line is not completely straight and the slope is changing more than is predicted by the standard error. Therefore, we introduce an error estimate $\epsilon_\alpha$ that quantifies a normalized standard deviation of the slope over the range. This error is formed by calculating the slope over $n_\alpha$ smaller subranges in the histogram, creating $n_\alpha$ estimates $\alpha_i$. If the errors in $\alpha_i$ are normally distributed with mean zero, the standard deviation over the full range will be $\sqrt{n_\alpha}$ times smaller. Therefore, an estimate of the variance of $\hat \alpha$ can be calculated with
\begin{equation}
 \epsilon_{\hat\alpha} = \sqrt{\frac{\sum \left( \alpha_i - \hat\alpha \right)^2 }{n^2_\alpha - n_\alpha}}.
\end{equation}
This estimate is slightly depending on the number of subranges that is used, $n_\alpha$, but we found that $\epsilon_{\hat\alpha}$ is more representative than the standard error of $\hat\alpha$.

The Hill estimator of Eq.~\eqref{eq:hill} is a different estimation method for the exponent in the power-law distribution, and yields therefore also a different standard error. The standard error of the Hill estimator is \citep{power-law-distribution}
\begin{equation} \label{eq:stderr-hill}
 \textrm{SE}(\hat \alpha_H) = \frac{-\alpha - 1}{\sqrt{n}} + \mathcal{O}(\frac{1}{n}).
\end{equation}
Because the number of samples is very large ($>10^{11}$), the $\mathcal{O}$-term will be very small. Therefore, we will calculate the quantity without the term. As with the standard error for the slope of a straight line in Eq.~\eqref{eq:stderr-slope}, the standard error for the Hill estimator yields very small quantities. Again, this is because it assumes the underlying power law has a fixed exponential, while in our case the power law seems to vary. Therefore, this value is given only for completeness.

Because our distributions are huge, we decided not to do goodness-of-fit tests, because these require many similar distributions to be simulated in order to reach accurate decision. Instead, we will try to evaluate the distributions visually.

\section{Data description} \label{sec:dist-data}
We have analysed the distributions of two data sets. Both data sets are 24 h LOFAR RFI surveys and are extensively analysed in
Offringa et al. (submitted to A\&A, 2012). % TODO
In one set, the low-band antennae (LBA) were used and the frequency range 30.1--77.5 MHz was recorded, while in the other the high-band antennae (HBA) were used to record the frequency range 115.0--163.3 MHz. More stations were used in the LBA set. The specifications of the two sets are listed in Table~\ref{table:dist-data-specs}. The stations that have been used are geometrically spread over an area of about 80~km and 30~km in diameter at maximum for the LBA and HBA sets respectively.

Although we have used Hata's model to estimate the RFI log-log histogram slope, our frequency range falls partly outside the frequency range over which Hata's model has been verified. However, according to Hata's model the observing frequency does not influence the power-law exponent in the frequency range 150--1500 MHz, thus it can be assumed the exponent will at least not significantly differ over the HBA range.

To detect RFI, the AOFlagger \citep{post-correlation-rfi-classification, LOFAR-RFI-pipeline} was used. However, since this flagger is partly amplitude-based, it is likely that low-level RFI will leak through the detector. Since it is also low-level RFI we are interested in, we will analyse unflagged data and the RFI classified data.

\begin{table}
\caption{Data set specifications}\label{table:dist-data-specs}
\begin{center}
\begin{tabular}{lrr}
                    & \textbf{LBA set}& \textbf{HBA set} \\
\hline
\hline
Observation date    & 2011-10-09      & 2010-12-27 \\
Start time          & 06:50 UTC       & 0:00 UTC \\
Length              & 24 h           & 24 h \\
Time resolution     & 1 s             & 1 s \\
\hline
Frequency range     &  30.1--77.5 MHz & 115.0--163.3 MHz\\
Frequency resolution & 0.76 kHz    & 0.76 kHz \\
Number of stations  &  33           & 13 \\
Total size          & 96.3 TiB        & 18.6 TiB \\
\hline
Field               & NCP             & NCP \\
Amount RFI detected & 1.77\%       & 3.18\% \\
\hline
\hline
\end{tabular}
\end{center}
\end{table}

\begin{figure*}
\begin{center}
\includegraphics[width=12cm]{img/histograms-raw-trimmed}
\caption{The histograms of the two data sets before pass-band correction and flux calibration.}
\label{fig:Histograms-raw}
\end{center}
\end{figure*}

\section{Results} \label{sec:dist-results}
In this section we present the histograms of the LBA and HBA sets and the results that were obtained by applying the methodology discussed in Sect.~\ref{sec:distribution-methods}.

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/plot-lba-dist-per-frequency-trimmed}
\caption{Histograms for $5$ different $0.2$ MHz LBA sub-bands without pass-band correction and flux calibration. The continuous lines represent the data before RFI flagging. The dashes lines are the histograms of the samples that have been classified as RFI.}
\label{fig:plot-dist-per-frequency-LBA}
\end{center}
\end{figure}

\subsection{Histogram analysis} \index{histogram}
Fig.~\ref{fig:Histograms-raw} shows the histograms with logarithmic axes for the LBA and HBA sets. In both sets, it is clear that at least one variate with a Rayleigh and one with a power-law distribution have been observed. The left part of the histogram matches the Rayleigh distribution well up to the mode of the distribution. The bulge around the mode of the LBA histogram is wider due to the larger effect of the antenna response as discussed in \S\ref{sec:parameter-variability}. As can be seen in Fig.~\ref{fig:plot-dist-per-frequency-LBA}, the Rayleigh-bulges of individual sub-bands are not that wide, but they are not aligned because of the differing noise levels. This effect is not so strong in histograms of the HBA sub-bands in Fig.~\ref{fig:plot-dist-per-frequency-HBA}, because the HBA antenna response changes less over frequency.

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/plot-hba-dist-per-frequency-trimmed}
\caption{Histograms for $5$ different $0.2$ MHz HBA sub-bands without pass-band correction and flux calibration. The continuous lines represent the data before RFI flagging. The dashes lines are the histograms of the samples that have been classified as RFI.}
\label{fig:plot-dist-per-frequency-HBA}
\end{center}
\end{figure}

It is to be expected that the RFI-dominated part of the distributions at different frequencies will reflect the underlying source populations. Both Figs.~\ref{fig:plot-dist-per-frequency-LBA} and \ref{fig:plot-dist-per-frequency-HBA} show that the power-law part of the distributions are very different for different sub-bands. Nevertheless, combining the data of all the sub-bands results in reasonably-stable power-law distributions. This indicates that the individual sub-bands have an approximately similar underlying power-law distribution, that is not yet apparent because not enough samples are combined in the histograms of individual sub-bands.

\begin{figure*}
\begin{center}
\includegraphics[width=13cm]{img/histograms-corrected-trimmed}
\caption{Observed LBA distribution after pass-band correction and flux calibration. $S_{l1}$ and $S_{l2}$ denote the limits of the distribution with a sharp lower cut-off (Eq.~\eqref{eq:lower-limit-1}) and uniform lower limit (Eq.~\eqref{eq:lower-limit-2}), $S_d$ is the average lower limit of RFI that is detected.}
\label{fig:histogram-passband-corrected}
\end{center}
\end{figure*}

\index{band-pass}
To make sure that the antenna response does not influence the result of the slope, we have also analysed the curves after a simple band-pass calibration. This was performed by dividing each sub-band by its standard deviation after RFI excision. Because the standard deviation of the distribution might be affected by the RFI tail of the distribution, we compare the two histograms to make sure the power-law distribution is not significantly changed. The resulting histograms are shown in Fig.~\ref{fig:histogram-passband-corrected}. This procedure makes the bulge of the LBA histogram similar to the bulge of a Rayleigh curve and extends the power-law part. Nevertheless, it does not change the log-log slope of the power law in either histograms. This validates that the variable gain that is caused by the antenna response does not change the observed power law. Consequently, it can be expected that other stochastic effects, such as the intrinsic source strength and the beam gain due to a differing direction of arrival, will similarly not effect the power law. Because the pass-band corrected histograms provide a more representative analysis, we will use the corrected histograms for further analysis.

\begin{figure*}
\begin{center}
\includegraphics[width=13cm]{img/histograms-with-fits-trimmed}
\caption{Least-squares fits of Rayleigh distributions to the observed LBA and HBA histograms, after pass-band correction but without flux calibration.}
\label{fig:histograms-with-fits}
\end{center}
\end{figure*}

The Rayleigh parts of the distributions are plotted in Fig.~\ref{fig:histograms-with-fits}, along with a least-squares fit and its residuals. Both histograms follow the Rayleigh distribution for about five orders of magnitude, which is validated by the residuals that show only noise. It breaks down about one order of magnitude before the mode of the distributions. This is because of the multi-variate nature of the distributions, as was described in \S\ref{sec:histogram-noise}.

\begin{figure*}
\begin{center}
\includegraphics[width=13cm]{img/slopes-trimmed}
\caption{The slope of the band-pass corrected log-log histogram as a function of the brightness. The horizontal lines indicate the fitted slope over the full (semi-) stable region. The horizontal axis is not calibrated.}
\label{fig:plot-slopes}
\end{center}
\end{figure*}

If we go back to Fig.~\ref{fig:histogram-passband-corrected}, we see that in the LBA the power law is stable for about three orders of magnitude, and one order more in the HBA. Fig.~\ref{fig:plot-slopes} shows the slope of the log-log plot as a function of amplitude, which was constructed by performing linear regression in a sliding window, with a window size of 1 decade. The HBA shows very little structure in the slope, but the HBA is less stable and shows some features in its power-law part. Linear regression on the power-law part of the log-log plot results in a slope of $-1.62$ for the LBA and $-1.53$ for the HBA. These and the other derived quantities have been summarized in Table~\ref{table:dist-data-quantities}. Although the HBA slope does not show any other significant features besides the Rayleigh and power-law curves, the LBA power law ends with a bulge around an amplitude of $10^6$. This bulge is caused by a very strong source affecting lots of samples, and is an apparent single outlier in the spatial distribution. We found this is caused by RFI observed for about an hour in the late afternoon in the lower LBA frequency regime, around $30$--$40$~MHz. Leaving this frequency range out flattens the bulge significantly, but does not completely eliminate it, because the source turned the receivers in non-linear state, causing leakage at lower intensity levels in the other sub-bands. Unlike linear regression, the fitting region of the Hill estimator is not limited at the high end. Consequently, because of the bulge, the Hill estimator evaluates for the LBA into a slope that is less steep, with a value of $-1.53$. For the HBA set, the Hill estimator is equal to the $-1.53$ value found by linear regression.

%\begin{figure}[tbp]
%\begin{center}
%\includegraphics[width=12cm]{img/plot-slope-over-frequency-trimmed}
%\caption{Slope of the log-log histogram slope over frequency in the LBA set.}
%\label{fig:plot-slope-over-frequency-LBA}
%\end{center}
%\end{figure}

%\begin{figure}[tbp]
%\begin{center}
%\includegraphics[width=12cm]{img/slope-over-frequency-HBA/plot-slope-over-frequency}
%\caption{Slope of the log-log histogram over frequency in the HBA set.}
%\label{fig:plot-slope-over-frequency-HBA}
%\end{center}
%\end{figure}

%\begin{figure}[tbp]
%\begin{center}
%\includegraphics[width=12cm]{img/slope-over-baseline-HBA/plot-slope-over-baseline-wstddev}
%\caption{Slope of the log-log histogram over the baseline length for the HBA set. A single sub-band at 150 MHz was used. Error bars indicate $5\sigma$ level.}
%\label{fig:plot-slope-over-baseline-HBA}
%\end{center}
%\end{figure}

\begin{table*}
\centering
\begin{minipage}{12cm}
\caption{Estimated distribution quantities per data set. }\label{table:dist-data-quantities}
\begin{tabular}{@{}llrr@{}}
\textbf{Symbol} & \textbf{Name} & \textbf{LBA set}& \textbf{HBA set} \\
\hline
\hline
% LBA fitting: 10-2000
$N_\textrm{total}$ & Total number of samples in histogram & $8.0\times10^{11}$ & $5.4\times10^{11}$ \\
$\sigma$ & Rayleigh mode (assumed to be SEFD/$\sqrt{2\Delta t \Delta \nu}$) & $770$ Jy & $77$ Jy \\
\hline
\multicolumn{4}{l}{\textit{Estimators for power-law distribution parameters}} \\
\hline
$\alpha$ & Exponent of power law in RFI distribution      & $-1.62$ & $-1.53$ \\ %-1.616 & -1.523
$SE(\alpha)$ & Standard error of $\alpha$ & $2.8 \times 10^{-3}$ & $6.9 \times 10^{-4}$ \\
$\alpha_H$ & Hill estimator for power-law exponent & $-1.53$ & $-1.53$ \\
$SE(\alpha_H)$ & Standard error of $\alpha_H$ & $8.9 \times 10^{-6}$ & $1.0\times10^{-5}$ \\
$\epsilon_{\alpha}$ & Sampled estimate of standard deviation of $\alpha$ & $6.1 \times 10^{-2}$ & $1.2 \times 10^{-2}$ \\
$\beta$ & Scaling factor of power law with exponent $\alpha$ & $4.0 \times 10^{17}$ & $3.4\times 10^{15}$ \\
\hline
\multicolumn{4}{l}{\textit{Limits}} \\
\hline
$S_L$ & Constraint on lower fall-off point of power law & $21$ mJy & $6.2$ mJy \\ % $10^{-6.42}$ & $10^{-4.31}$
$\tilde{S}_L$ & As $S_L$, but assuming $Ig/r^2\sim$ uniform & $47$ mJy & $14$ mJy \\ % $10^{-6.08}$ & $10^{-3.96}$
$S_d$ & Expected lowest apparent level of RFI detected & $26$ Jy & $5.7$ Jy\\ % $2.3 \times 10^{-2}$ & $4.5 \times 10^{-2}$
% $S_U$ & Upper fall-off point & n.a. & $10^{2.62}$ \\
%$\epsilon_R$ & Error in Rayleigh part & TODO & \\
$E(S_R)$ & Apparent RFI flux density & $2700$ Jy & $140$ Jy \\
$E(S_\textrm{leak})$& Residual apparent RFI flux density after excision & $484$--$496$ mJy & $167$--$171$ mJy \\
 & Same as above, but by assuming 10\% occupancy & $384$ mJy & $120$ mJy \\ % 97--123 mJy
REFD& RFI equivalent flux density & $18.9$--$19.3$ Jy & $6.5$--$6.7$ Jy \\
\hline
\hline
\end{tabular}
\end{minipage}
\end{table*}

On the assumption that the histogram is zero below amplitude $S_L$, we find that $S_L=21$ mJy for the LBA and $S_L=6.2$ mJy for the HBA (see Table~\ref{table:dist-data-quantities}). If instead it is assumed that the histogram has a uniform distribution below some amplitude $\tilde{S_L}$, we find that the amplitude at which the power-law distribution breaks down is approximately a factor two higher. The two different assumptions on how the power-law distribution breaks down have a small effect on $E(S_\textrm{leak})$, the expected value of the leaked RFI. By using $\tilde{S}_L$ instead of $S_L$, it is a few percent lower. By assuming a 100\% RFI occupancy, we find that the expected value of \index{RFI!leakage} leaked \index{leakage (RFI)} RFI is $484$--$496$ mJy for the LBA and $167$--$171$ mJy. With $10$\% occupancy, the value for $E(S_\textrm{leak})$ is about $25$\% reduced. The RFI occupancy only starts to have a significant effect on $E(S_\textrm{leak})$ if it is well below $10$\%.

\section{Conclusions and discussion} \label{sec:dist-discussion}
We have analysed the histogram of visibility amplitudes of LOFAR observations and found that, within a significant range of the histogram, the contribution of RFI sources follows a power-law distribution. The found power-law exponents of $-1.62$ and $-1.53$ for the 30--78~MHz LBA and 115--163~MHz HBA observations respectively, can be explained by a uniform spatial distribution of RFI sources, affected by propagation described surprisingly well by Hata's electromagnetic propagation model. Taken at face value these exponents imply in Hata's model that the average \index{transmitters!height} transmitting heights for sources affecting the LBA and HBA are $79$ and $13$~m respectively. Hata's model only goes down to 150 MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies, e.g. because the effect of the ionosphere becomes stronger. Intervals for the exponents with representative $3\sigma$ confidence boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.53;-1.49]$ for the HBA. The estimate of the HBA is thus more accurate, because its histogram deviates less from the power law.

On the assumption that the power-law distribution for RFI sources will continue down into the noise, we have constructed a full parametrization of the RFI apparent flux distribution. By assuming that all samples contain some contribution of RFI, we find that the average flux density of RFI that leaks through the detector is $484$--$496$~mJy for the LBA and $167$--$171$~mJy for the HBA (depending on the used method). These values should be compared to the noise in individual samples of $770$~Jy (LBA) and $77$~Jy (HBA) (see Table~\ref{table:dist-data-quantities}), and are upper limits for what can be expected. If in fact not all samples are affected by RFI, the leaked RFI flux will be smaller, and will of course be zero in the extreme case that the detector has found and removed all RFI.

In experiments such as the LOFAR EoR project, a simulation pipeline is used to create a realistic estimate of the signal that can be expected. Currently, these simulations do not include the effects of RFI. With the construction of empirical models for the RFI source distributions, we are one step closer to including these effects in the simulation. Using Eq.~\eqref{eq:drawing-rfi-distribution}, one can sample the strength of a single RFI source, add the feature to the data and run the AOFlagger. What is still needed for accurate simulation, is to obtain a likely distribution for the duration that one such source affects the data. For example, it is neither realistic that all RFI sources are continuously transmitting nor that they affect only one sample. The RFI detector is highly depending on the morphology of the feature in the time-frequency domain. Finally, the \index{coherency (RFI)} coherency \index{RFI!coherency} properties of the RFI might be even more important to simulate correctly, but these have been not been explored. However, these have large implications for observations with high sensitivity. This will be discussed in the next section.

The derived values for the average lower level of detected RFI, $S_d$, show that the AOFlagger has detected a large part of the RFI that is well below the sample noise. In both sets, $S_d$ is more than one order of magnitude below the Rayleigh mode. This can be explained with two of the algorithms it implements. The first one is the \index{SumThreshold} {\tt SumThreshold} method \citep{post-correlation-rfi-classification}, that thresholds on combinations of samples, and is thus able to detect RFI that is weaker than the sample noise. The second one is the \index{SIR operator} scale-invariant rank (SIR) operator \citep{scale-invariant-rank-operator}. This operator is not dependent on the sample amplitude, but flags based on morphology.

\subsection{Implications for very long integrations}
In theory, faint RFI could impose a fundamental limit on the attainable \index{thermal sky noise} noise limit of long integrations. As an example, we will analyse the situation for the \index{LOFAR-EoR} LOFAR EoR project. This project will use the LOFAR high-band antennae to collect on the order of 50--100 night-time observations of 6 h for a few target fields. The final resolution required for signal extraction will be about $1$~MHz. The project will use about $60$~stations, each of which provides two polarized feeds. This will bring the noise level in a single 6~h observation in 1~MHz bandwidth to
\begin{equation}
 \sigma_\textrm{eor-night} = \textrm{SEFD} \left(2\Delta t \Delta \nu N_\textrm{feed} N_\textrm{interferometers} \right)^{-\frac{1}{2}} \approx 250\hspace{1mm} \mu\textrm{Jy}.
\end{equation}
Therefore, after 100 nights the thermal noise level will be $25\hspace{1mm}\mu\textrm{Jy}$.

Because some RFI sources might be \index{stationary RFI} stationary, the signals from these sources will add coherently over time. Therefore, the amount that time integration can lower the flux density of RFI might be limited. Additionally, some RFI sources will be received by multiple stations of the array, and by multiple feeds of the individual antennae. Therefore, integrating data from different interferometers and data from the two polarized feeds might also not bring the noise level that is caused by RFI down. In summary, unlike normal noise, the RFI might be coherent over time, interferometer and feed.

On the other hand, many RFI signals observed in the LOFAR bands have a limited bandwidth. Indeed, the majority of the detected RFI sources affect only one or a few LOFAR channels of 0.76 kHz. Therefore, frequency averaging will lower the flux density of the RFI signal. If the frequency range contains only one stationary RFI source, the strength of this RFI source will go down linearly with the total bandwidth. If we assume that all channels are affected by RFI sources and all transmit in approximately one channel, then the noise addition that is produced by RFI will go down with the square root of the number of averaged channels. This is a consequence of the random phase that different RFI sources have.

In summary, some class of stationary RFI sources are expected to be coherent over time, polarization and interferometer, but not over frequency. Therefore, in this case the noise level at which RFI leakage approximately becomes relevant is given by
\begin{equation} \label{eq:rfi-noise-level}
 \sigma_\textrm{RFI} = \frac{\textrm{REFD}}{\sqrt{2 \Delta \nu}},
\end{equation}
where \index{REFD} $\textrm{REFD}$ is the RFI equivalent flux density at $1$ Hz and $1$ s resolution for a single station, in analogue to how the SEFD is defined. This only holds when the observational integrated bandwidth $\Delta \nu$ is substantially higher than the average bandwidth of a single RFI source. Otherwise, if the $\Delta \nu$ is small relative to the average bandwidth of RFI sources, some RFI might show up earlier. The empirically found upper limits in this work are $\textrm{REFD}_\textrm{LBA}=18.9$--$19.3$~Jy and $\textrm{REFD}_\textrm{HBA}=6.5$--$6.7$~Jy (see Table~\ref{table:dist-data-quantities}).

For the EoR project with $1$~MHz resolution, Eq.~\ref{eq:rfi-noise-level} results in $\sigma_\textrm{RFI}\approx 4.7$ mJy. However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per subband (Yatawatta et al., \textit{in preparation}), and the resulting images show no signs of RFI. Clearly, Eq.~\ref{eq:rfi-noise-level} is therefore not applicable to most of the RFI that is observed with LOFAR. In the following section we will discuss effects that could cause a reduced contribution of RFI.

\subsection{Interference-reducing effects} \label{sec:coherence-reduction} \index{RFI!coherency} \index{coherency (RFI)}
When integrating data, it is likely that the actual noise limit will be significantly lower than the given upper limit, which was determined on high resolution. There are several reasons for this, which we will summarize one by one.
\begin{itemize}
 \item Many RFI sources have a variable \index{geometric phase} geometric phase, because they move or because their path of propagation changes. This would cause them to sum (partly) incoherently over time, and thus go approximately down with the noise.
 \item Many RFI sources will be seen by only a few stations. This would make the histogram go down more quickly at lower amplitudes.
 \item For the shortest baselines at 150~MHz, the far field starts around 1~km. Some RFI sources will therefore be in the near field, especially in the longer baselines. In that case, a source will not add up coherently over the interferometers that see the particular source, as the interferometers see them with different phases.
 \item Many stationary sources are not constant over time, thus will be attenuated somewhat by \index{averaging} averaging.
 \item We have assumed 100\% of the spectrum is occupied by RFI. If, say, only 10\% of the spectrum is occupied, the expected value of the \index{leakage (RFI)} leaked \index{RFI!leakage} RFI level decreases with about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all. With current data, one can only speculate how much the electromagnetic spectrum is truly occupied.
 \item Stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other. Individually, the sources will have fixed geometric phases, but because the baselines are much longer than the wavelength, their phases will become uniformly spread. Therefore, they will add incoherently.
\end{itemize}

Two other effects can improve the RFI situation as well:
\begin{itemize}
 \item \index{fringe stopping} Fringe stopping interferometers can partly average out RFI sources. Nevertheless, stationary RFI that is averaged out by fringe stopping will leave artefacts behind in the field centre \citep{post-correlation-filtering}. This is not relevant when observing the \index{NCP} North Celestial Pole --- which is one of the LOFAR EoR fields --- because no fringe stopping is applied when observing the NCP.
 \item The Fourier transform that is involved in data imaging will localize the contribution from \index{stationary RFI} stationary RFI near the NCP. If RFI artefacts would show in the image of the NCP field, they can easily be detected and possibly be removed, or processing could ignore data near the pole.
\end{itemize}

Because of the above two arguments, when considering RFI it is a risk to use the NCP as one of the EoR target fields. At the same time, this field is useful for analysing the RFI coherency properties. It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright \index{foreground sources} foreground sources. Preliminary analysis of EoR NCP observations of a single night have reached the \index{thermal sky noise} thermal noise (Yatawatta et al., \textit{in preparation}), but do not show leaked RFI at the pole.

Finally, future RFI excision strategies will further enhance detection accuracy. Currently, RFI excision is applied only on the raw data from single interferometers at high resolution. Once data from a large number of nights are collected, it will be possible to detect and excise RFI more accurately, by looking at the averaged data from multiple nights and/or multiple interferometers. We have shown that the current detection algorithms can detect RFI well below the noise. Therefore, if data from different nights or interferometers are summed, and there is stationary RFI in the set, it will become detectable. All RFI that is below the noise but is not stationary, will act like normal noise and will therefore be harmless.

With the current strategy, it is likely that the LOFAR EoR project will encounter some RFI on some frequencies when averaging lots of observing nights, although this still has to be seen. To mitigate this leaked RFI, the detection can be executed at higher signal-to-noise levels. The current results indicate that a lot of RFI is not coherent, and the situation is promising. Considering the current RFI results, and the availability of further mitigation steps, it is very likely to assume that RFI will not be problematic for the detection Epoch of Reionisation.

\label{lastpage}
\section*{Acknowledgments}
TODO

\bibliographystyle{mn2e}
\bibliography{references}

\end{document}

### 2012-11-22 articles/Article-RFI-distribution/article-rfi-distribution.tex
Finished article

@@
 \documentclass[useAMS,usenatbib]{mn2e} 
-%\documentclass{article}
 \usepackage{graphicx}
@@
 \usepackage[font={it},labelfont={bf}]{caption}
-%\usepackage[round]{natbib}
 \usepackage{units}
-%\usepackage[utopia]{mathdesign}
-\usepackage{times} %A
-%\usepackage{fourier} %B
-%\usepackage{lmodern} % C
-%\usepackage{mathpazo} % D
-%\usepackage[T1]{fontenc}
-%\usepackage{euler} % D
+\usepackage{times}
 \usepackage{color}
@@
 Thereby, they might occlude the astronomical signal of interest.
-This issue is especially important for epoch of reionization (EoR) projects that try to detect the very faint redshifted HI signals of the earliest objects in the Universe.
+This issue is especially important for epoch of reionization (EoR) projects that try to detect the faint redshifted HI signals of the earliest objects in the Universe.
 We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR.
-The histograms shows significant features, of which a power law with estimated exponent around -1.6 to -1.5 is most striking.
-With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows the Hata propagation model.
+The histograms shows a RFI distribution that follows a power-law distribution with an estimated exponent around -1.6 to -1.5.
+With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models.
 Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present.
@@
 In these cases, it often suffices to only excise interference that is apparent and thus above the noise to reach the thermal noise limit of the instrument.
-A new challenge arises when one desires much deeper observations, and hunderds of hours of observations need to be integrated.
-In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but is \index{coherency (RFI)} coherently \index{RFI!coherency} present in the data.
+A new challenge arises when one desires much deeper observations, and hundreds of hours of observations need to be integrated.
+In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but is coherently present in the data.
 Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest.
@@
 This work explores the information that is present in interference distributions, in order to analyse possible low-level interference that is not detectable by standard detection methods.
-Our approach is similar to the radio source count analysis that is used in cosmology \citep{vonhoerner-radio-source-counts}, also named \index{log N--log S plot} $\log N$ -- $\log S$ analysis, where $N$ and $S$ refer to the celestial source count and brightness respectively.
+Our approach is similar to the radio source count analysis that is used in cosmology \citep{vonhoerner-radio-source-counts}, also named $\log N$ -- $\log S$ analysis, where $N$ and $S$ refer to the celestial source count and brightness respectively.
 The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe.
@@
 \section{Predicting the brightness distribution} \label{sec:distribution-prediction}
-\index{brightness distribution}\index{spatial distribution}\index{power law}\index{transmitters!distribution}
 Interference is generated by many different kind of transmitters, and these will have different spatial and brightness distributions (``spatial'' refers here to the distribution on the Earth).
-For example, \index{aeroplanes} aeroplanes and \index{satellite} satellites have widely different heights, while other sources are ground-based.
+For example, aeroplanes and satellites have widely different heights, while other sources are ground-based.
 Even ground-based sources might be spread differently.
-For example, it can be expected that \index{CB devices} citizens' band (CB) devices, that are often used in cars, are differently distributed from broadcasting transmitters.
+For example, it can be expected that citizens' band (CB) devices, that are often used in cars, are differently distributed from broadcasting transmitters.
 The frequencies of deliberately transmitting sources of a particular class of devices are constrained by the bands that have been allocated for the given class, which might allow one to distinguish different sources to some extent.
 
-\index{morphology}
 In a time-frequency plot, interfering sources can have complex structures.
@@
 Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity.
-We will refer to this quantity with the statistical term \index{frequency density} ``frequency density'' (not to be confused with the physical radiation frequency).
+We will refer to this quantity with the statistical term ``frequency density'' (not to be confused with the physical radiation frequency).
 We will now start by deriving a prediction of the frequency density function of ground-based interfering sources.
@@
 We assume that the source is observed fully coherent, but a possible decoherence factor can be absorbed in the gains.
-Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the \index{narrow-band condition} narrow-band condition.
+Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the narrow-band condition.
 For example, if the bandwidth of the signal $\Delta\nu=1$ kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c / (2 \pi \Delta \nu) \approx$ 50 km.
@@
 
-Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we predict to see a \index{slope (log-log plot)} slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane.
+Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we predict to see a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane.
+Therefore, RFI follows a power-law distribution.
 
-\subsection{Spherical case} \index{spherical distribution}
-If we do take into account the fact that the sources lie on a sphere, Eq.~\eqref{eq:cumulative-distance-function} changes.
+\subsection{Spherical case}
+If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes.
 In this case, the number of sources corresponds with the surface area of a hemisphere.
@@
 
-\subsection{Propagation effects} \index{propagation (EM)}
-So far, we have assumed that the electromagnetic radiation propagates through \index{free space} free space, resulting in a $r^{-2}$ fall-off.
+\subsection{Propagation effects}
+So far, we have assumed that the electromagnetic radiation propagates through free space, resulting in a $r^{-2}$ fall-off.
 In reality, the radiation will be affected by complicated propagational effects.
-Because of the irregular surface of the Earth (including urban areas) and the absence of a direct \index{line of sight} line of sight between transmitter and receiver, the propagation mode might be indirect.
+Because of the irregular surface of the Earth (including urban areas) and the absence of a direct line of sight between transmitter and receiver, the propagation mode might be indirect.
 Multiple indirect paths might be formed by reflection, refraction or diffraction of the electromagnetic wave.
@@
 \end{align}
-where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in m; $h_m$ the height of the receiving antenna in m; $r$ the distance between the antennae in m; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the \index{urban density|see{population density}} urban \index{population density} density.
-\citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with \index{transmitters!height} transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.
+where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in m; $h_m$ the height of the receiving antenna in m; $r$ the distance between the antennae in m; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density.
+\citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.
 
-Collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, and converting from a subtrahend in decibels to a flux density factor $L_S$, results in
+Converting from a subtrahend in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
 \begin{equation} \label{eq:propagation-loss-factor}
@@
 \end{figure}
-This is valid for \index{transmitters!height} transmitters that have a height of 30--200 m, the range over which Hata's model was defined.
+This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined.
 This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively.
@@
 
-\subsection{Noise contribution} \label{sec:histogram-noise}\index{noise}
+\subsection{Noise contribution} \label{sec:histogram-noise}
 The full measured distribution will consist of the power-law distribution combined with that of the noise and the celestial signal.
@@
 The real and imaginary components of receiver noise are independent and identically Gaussian distributed.
-Consequently, an amplitude $x$ will be Rayleigh distributed: \index{Rayleigh distribution}
+Consequently, an amplitude $x$ will be Rayleigh distributed:
 \begin{equation}\label{eq:rayleigh-formula}
@@
 
-To understand the RFI-with-noise distribution, one needs to consider the individual contributions of the real and imaginary components.
-The real and imaginary components of samples that are affected by RFI, will follow a \index{bivariate distribution} bivariate distribution: each complex component is the sum of a complex component of RFI and a noise sample.
-As said, the noise comes from a Gaussian distribution with zero mean.
-The RFI distribution is assumed to be a power law with yet unknown exponent.
+To understand the distribution of samples consisting of RFI + noise, one needs to consider the individual contributions of the real and imaginary components.
+The real and imaginary components of samples that are affected by RFI, will follow a bivariate distribution: each complex component is the sum of a complex component of RFI and a noise sample.
+The noise comes from a Gaussian distribution with zero mean.
+The RFI distribution is assumed to be a power-law distribution with yet unknown exponent.
 
@@
 Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity.
-Therefore, to \index{drawing samples} draw a sample by using the \index{inversion method} inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$.
+Therefore, to draw a sample by using the inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$.
 By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-c \pi r_L^2$ for $S < Ig/r_L^2$.
-Consequently, by sampling a uniform value $x_u$ with $0 \le x_u \le N-c \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ the power-law distribution.
+Consequently, by sampling $x_u$ from a uniform distribution with $0 \le x_u \le N-c \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ the power-law distribution.
 
-A single complex RFI contaminated sample $S_{\textrm{RFI}}$ can then be sampled with
+By normalizing $x_u$, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
 \begin{equation} \label{eq:drawing-rfi-distribution}
-\textrm{real}(S_{\textrm{RFI}})\leftarrow \frac{Ig}{x_u r_L^2}\textrm{,\hspace{1mm} imag}(S_{\textrm{RFI}}) \leftarrow \frac{Ig}{x_u r_L^2},
+\textrm{real}(S_{\textrm{RFI}})\leftarrow \frac{Ig}{x_u r_L^2}\textrm{,\hspace{1mm} imag}(S_{\textrm{RFI}}) \leftarrow \frac{Ig}{y_u r_L^2},
 \end{equation}
-with $x_u \sim U(0, 1)$, and the corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
+with $x_u, y_u \sim U(0, 1)$, and the corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
 \begin{equation} \label{eq:drawing-combined-distribution}
 \textrm{real}(S) \leftarrow x_n + \frac{Ig}{x_u r_L^2}
-\textrm{,\hspace{1mm} imag}(S) \leftarrow x_n + \frac{Ig}{x_u r_L^2},
+\textrm{,\hspace{1mm} imag}(S) \leftarrow y_n + \frac{Ig}{y_u r_L^2},
 \end{equation}
@@
 
-One instrumental effect that is absorbed in $g$ is the \index{frequency response|see{band-pass}} frequency response of the instrument, i.e., the antenna response in combination with the passband of the analogue and digital filters.
+One instrumental effect that is absorbed in $g$ is the frequency response of the instrument, i.e., the antenna response in combination with the passband of the analogue and digital filters.
 Because the data that are analysed in Sect.~\ref{sec:dist-results} have initially not been band-pass calibrated, the instrumental response is not uniform over frequency.
-We determined that the variation due to the \index{band-pass} band-pass is about one order of magnitude in the LBA and about a factor of two in the HBA.
+We determined that the variation due to the band-pass is about one order of magnitude in the LBA and about a factor of two in the HBA.
 
@@
 
-\index{beam}\index{LOFAR!beam}
 Another effect that is absorbed in $g$, is the beam of the instrument.
@@
 
-\subsection{Creating a histogram} \label{sec:histogram}\index{histogram}
+\subsection{Creating a histogram} \label{sec:histogram}
 While creating a histogram is trivial, it is important to note that it is necessary to have a variable bin size.
@@
 This can be seen in the tails of the distributions of Fig.~\ref{fig:rayleigh-and-rfi-combined}.
-This, however, is not an intrinsic feature of the data but a consequence of the \index{binning} binning method.
+This, however, is not an intrinsic feature of the data but a consequence of the binning method.
 
-\subsection{Estimating the slope}\index{slope}
-An automated procedure is used to calculate the slope of the observed log-log histogram.
-First, the \index{mode (Rayleigh)} mode $\hat\sigma$ of the Rayleigh distribution is estimated by finding the amplitude with the maximum occurrences, i.e., the amplitude corresponding to the peak of the histogram.
-Then, the maximum amplitude $S_\textrm{max}$ is calculated, i.e., the sample with highest amplitude.
-Finally, the slope is estimated using \index{linear regression} linear regression over the interval $R$,
-\begin{equation}
-R=\left[20\hat\sigma; 10^{\frac{1}{2}\left(\log_{10} 20\hat\sigma + \log_{10} S_\textrm{max}\right) } \right].
-\end{equation}
-In terms of the log-log histogram, the start of the interval corresponds with the point that is $1.3$ units to the right of the peak in the histogram, and the interval end corresponds with the point halfway between the start point and the maximum amplitude.
+\subsection{Estimating $\sigma$ and slope parameters}
+The mode $\hat\sigma$ of the Rayleigh distribution is estimated by finding the amplitude with the maximum occurrences, i.e., the amplitude corresponding to the peak of the histogram.
+The slope is estimated using linear regression over a visually selected interval.
+%\begin{equation}
+%R=\left[20\hat\sigma; 10^{\frac{1}{2}\left(\log_{10} 20\hat\sigma + \log_{10} S_\textrm{max}\right) } \right].
+%\end{equation}
+%In terms of the log-log histogram, the start of the interval corresponds with the point that is $1.3$ units to the right of the peak in the histogram, and the interval end corresponds with the point halfway between the start point and the maximum amplitude.
 This interval starts near the start of the RFI slope but past where the Rayleigh curve dominates.
-It also does not include the noisy data at the end of the curve.% TODO The curves are not always completely straight in this interval, but since we use the same stable way to determine the interval over which to calculate the slope, we can at least compare them.
+It also does not include the noisy data at the end of the curve.
+The curves are not always completely straight in this interval, but since we use the same stable way to determine the interval over which to calculate the slope, we can at least compare them.
 
-\index{power law}
 One should note that fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}.
@@
 Nevertheless, we will additionally calculate a maximum likelihood estimator for comparison.
-The maximum-likelihood estimator for the exponent in a power-law distribution is given by the \index{Hill estimator} Hill estimator \citep{hill-estimator, power-law-distribution}, defined as:
+The maximum-likelihood estimator for the exponent in a power-law distribution is given by the Hill estimator \citep{hill-estimator, power-law-distribution}, defined as:
 \begin{equation} \label{eq:hill}
@@
 However, this estimator assumes the distribution is not cut-off at a high point.
-In our case, the distribution is cut-off, for example because of the limit of the analogue-digital converter (see \S\ref{sec:rfi-distribution-constraints}).
+In our case, the distribution is cut-off at both sides, for example because of the limit of the analogue-digital converter (see \S\ref{sec:rfi-distribution-constraints}).
 Therefore, using this estimator will result in an estimate that is steeper (i.e., more negative) than the actual distribution.
@@
 However, we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ the bin frequency.
-Because the bin size is rather small, the resulting estimate will be close to the normal Hill estimator and one does not have to iterate over all data.
+Because our bin size is small and the distribution is well sampled, the resulting estimate will be close to the normal Hill estimator.
+This overcomes the need to iterate over all data.
 
@@
 For Fig.~\ref{fig:explanation-constraints} this means that the areas labelled $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal.
-Solving this equation, one finds that
+Solving this equality, one finds that
 \begin{equation} \label{eq:lower-limit-1}
@@
 In
-Offringa et al. (in preparation), % TODO fill in citation if available
+\citet{lofar-radio-environment} % TODO correct citation if full citation available
 the false-positives rate for the AOFlagger is estimated to be 0.5\%.
-An estimate of $S_d$, the average lower limit of detected RFI, can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the ``real number'' (true positives) of RFI samples.
+An estimate of $S_d$, the average lower limit of detected RFI, can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the real number (true positives) of RFI samples.
 Therefore, the limit is calculated similar to Eq.~\eqref{eq:lower-limit-1}, where the term $N_\textrm{part} - N_\textrm{total}$ needs to be replaced with $N_\textrm{RFI}$, which equals the total number of samples detected as RFI minus the 0.5\% false positives.
@@
 Finally, $E(S_\textrm{leak})$, which is the expected value of leaked RFI not detected by the flagger, can be calculated by replacing $S_U$ with $S_d$ in the numerator of Eq.~\eqref{eq:expected-value-rfi} and subtracting the removed number of samples from the total of number of samples.
-Assume that a fraction of $1-\eta$ samples have been detected as RFI, then
+Assume that a fraction of $\eta$ samples are not detected as RFI and $1-\eta$ have been detected as RFI, then
 \begin{align} \label{eq:expected-value-leaked-rfi}
@@
 We can assign flux densities to the horizontal axis of the histogram by using the system equivalent flux density (SEFD) of a single station.
-The current \index{SEFD (LOFAR)} LOFAR SEFD is found to be approximately 3000~Jy for the HBA core stations and 1500~Jy for the remote stations in the frequency range from $125$--$175$ MHz.
+The current LOFAR SEFD is found to be approximately 3000~Jy for the HBA core stations and 1500~Jy for the remote stations in the frequency range from $125$--$175$ MHz.
 For all Dutch LBA stations, in the frequency range $40$--$70$ MHz the SEFD is approximately 30,000~Jy.
@@
 \end{equation}
-This estimate is slightly depending on the number of subranges that is used, $n_\alpha$, but we found that $\epsilon_{\hat\alpha}$ is more representative than the standard error of $\hat\alpha$.
+This estimate is slightly depending on the number of subranges that is used, $n_\alpha$, because the errors are not Gaussian distributed, but we found that $\epsilon_{\hat\alpha}$ is more representative than the standard error of $\hat\alpha$.
 
@@
 We have analysed the distributions of two data sets. Both data sets are 24 h LOFAR RFI surveys and are extensively analysed in
-Offringa et al. (submitted to A\&A, 2012). % TODO
+\citet{lofar-radio-environment}. % TODO
 In one set, the low-band antennae (LBA) were used and the frequency range 30.1--77.5 MHz was recorded, while in the other the high-band antennae (HBA) were used to record the frequency range 115.0--163.3 MHz.
@@
 
-To detect RFI, the AOFlagger \citep{post-correlation-rfi-classification, LOFAR-RFI-pipeline} was used.
+To detect RFI, the AOFlagger \citep{post-correlation-rfi-classification, LOFAR-RFI-pipeline, scale-invariant-rank-operator} was used.
 However, since this flagger is partly amplitude-based, it is likely that low-level RFI will leak through the detector.
@@
 
-\subsection{Histogram analysis} \index{histogram}
+\subsection{Histogram analysis}
 Fig.~\ref{fig:Histograms-raw} shows the histograms with logarithmic axes for the LBA and HBA sets.
@@
 
-\index{band-pass}
 To make sure that the antenna response does not influence the result of the slope, we have also analysed the curves after a simple band-pass calibration.
@@
 Although the HBA slope does not show any other significant features besides the Rayleigh and power-law curves, the LBA power law ends with a bulge around an amplitude of $10^6$.
-This bulge is caused by a very strong source affecting lots of samples, and is an apparent single outlier in the spatial distribution.
+This bulge is caused by a very strong source affecting lots of samples, and is a single outlier in the spatial distribution.
 We found this is caused by RFI observed for about an hour in the late afternoon in the lower LBA frequency regime, around $30$--$40$~MHz.
@@
 By using $\tilde{S}_L$ instead of $S_L$, it is a few percent lower.
-By assuming a 100\% RFI occupancy, we find that the expected value of \index{RFI!leakage} leaked \index{leakage (RFI)} RFI is $484$--$496$ mJy for the LBA and $167$--$171$ mJy.
-With $10$\% occupancy, the value for $E(S_\textrm{leak})$ is about $25$\% reduced.
+By assuming a 100\% RFI occupancy, we find that the expected value of leaked RFI is $484$--$496$ mJy for the LBA and $167$--$171$ mJy.
+By assuming $10$\% occupancy, the value for $E(S_\textrm{leak})$ is about $25$\% reduced.
 The RFI occupancy only starts to have a significant effect on $E(S_\textrm{leak})$ if it is well below $10$\%.
@@
 The found power-law exponents of $-1.62$ and $-1.53$ for the 30--78~MHz LBA and 115--163~MHz HBA observations respectively, can be explained by a uniform spatial distribution of RFI sources, affected by propagation described surprisingly well by Hata's electromagnetic propagation model.
-Taken at face value these exponents imply in Hata's model that the average \index{transmitters!height} transmitting heights for sources affecting the LBA and HBA are $79$ and $13$~m respectively.
+Taken at face value these exponents imply in Hata's model that the average transmitting heights for sources affecting the LBA and HBA are $79$ and $13$~m respectively.
 Hata's model only goes down to 150 MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies, e.g.
@@
 With the construction of empirical models for the RFI source distributions, we are one step closer to including these effects in the simulation.
-Using Eq.~\eqref{eq:drawing-rfi-distribution}, one can sample the strength of a single RFI source, add the feature to the data and run the AOFlagger.
+Using Eq.~\eqref{eq:drawing-rfi-distribution}, one can sample a realistic strength of a single RFI source, add the feature to the data and run the AOFlagger.
 What is still needed for accurate simulation, is to obtain a likely distribution for the duration that one such source affects the data.
@@
 The RFI detector is highly depending on the morphology of the feature in the time-frequency domain.
-Finally, the \index{coherency (RFI)} coherency \index{RFI!coherency} properties of the RFI might be even more important to simulate correctly, but these have been not been explored.
+Finally, the coherency properties of the RFI might be even more important to simulate correctly, but these have been not been explored.
 However, these have large implications for observations with high sensitivity.
@@
 This can be explained with two of the algorithms it implements.
-The first one is the \index{SumThreshold} {\tt SumThreshold} method \citep{post-correlation-rfi-classification}, that thresholds on combinations of samples, and is thus able to detect RFI that is weaker than the sample noise.
-The second one is the \index{SIR operator} scale-invariant rank (SIR) operator \citep{scale-invariant-rank-operator}.
+The first one is the {\tt SumThreshold} method \citep{post-correlation-rfi-classification}, that thresholds on combinations of samples, and is thus able to detect RFI that is weaker than the sample noise.
+The second one is the scale-invariant rank (SIR) operator \citep{scale-invariant-rank-operator}.
 This operator is not dependent on the sample amplitude, but flags based on morphology.
@@
 \subsection{Implications for very long integrations}
-In theory, faint RFI could impose a fundamental limit on the attainable \index{thermal sky noise} noise limit of long integrations.
-As an example, we will analyse the situation for the \index{LOFAR-EoR} LOFAR EoR project.
+In theory, faint RFI could impose a fundamental limit on the attainable noise limit of long integrations.
+As an example, we will analyse the situation for the LOFAR EoR project.
 This project will use the LOFAR high-band antennae to collect on the order of 50--100 night-time observations of 6 h for a few target fields.
@@
 
-Because some RFI sources might be \index{stationary RFI} stationary, the signals from these sources will add coherently over time.
+Because some RFI sources might be stationary, the signals from these sources will add coherently over time.
 Therefore, the amount that time integration can lower the flux density of RFI might be limited.
-Additionally, some RFI sources will be received by multiple stations of the array, and by multiple feeds of the individual antennae.
-Therefore, integrating data from different interferometers and data from the two polarized feeds might also not bring the noise level that is caused by RFI down.
-In summary, unlike normal noise, the RFI might be coherent over time, interferometer and feed.
+Additionally, some RFI sources are received by multiple stations of the array, and by multiple feeds of the individual antennae.
+Therefore, integrating data from different interferometers and data from the two polarized feeds might not lower the noise level that is caused by RFI at some point.
+In summary, RFI is unlike normal noise and might be coherent over time, interferometer and feed.
 
@@
 If the frequency range contains only one stationary RFI source, the strength of this RFI source will go down linearly with the total bandwidth.
-If we assume that all channels are affected by RFI sources and all transmit in approximately one channel, then the noise addition that is produced by RFI will go down with the square root of the number of averaged channels.
+If we assume that all channels are affected by RFI sources and all these sources transmit in approximately one channel, then the noise addition that is produced by RFI will go down with the square root of the number of averaged channels.
 This is a consequence of the random phase that different RFI sources have.
@@
 \end{equation}
-where \index{REFD} $\textrm{REFD}$ is the RFI equivalent flux density at $1$ Hz and $1$ s resolution for a single station, in analogue to how the SEFD is defined.
+where $\textrm{REFD}$ is the RFI equivalent flux density at $1$ Hz and $1$ s resolution for a single station, in analogue to how the SEFD is defined.
 This only holds when the observational integrated bandwidth $\Delta \nu$ is substantially higher than the average bandwidth of a single RFI source.
@@
 For the EoR project with $1$~MHz resolution, Eq.~\ref{eq:rfi-noise-level} results in $\sigma_\textrm{RFI}\approx 4.7$ mJy.
-However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per subband (Yatawatta et al., \textit{in preparation}), and the resulting images show no signs of RFI.
-Clearly, Eq.~\ref{eq:rfi-noise-level} is therefore not applicable to most of the RFI that is observed with LOFAR.
+However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz subband (Yatawatta et al., \textit{in preparation}), and the resulting images show no signs of RFI.
+This implies that either the upper limit is far from the actual RFI situation, or Eq.~\ref{eq:rfi-noise-level} is not applicable to most of the RFI that is observed with LOFAR.
 In the following section we will discuss effects that could cause a reduced contribution of RFI.
 
-\subsection{Interference-reducing effects} \label{sec:coherence-reduction} \index{RFI!coherency} \index{coherency (RFI)}
+\subsection{Interference-reducing effects} \label{sec:coherence-reduction}
 When integrating data, it is likely that the actual noise limit will be significantly lower than the given upper limit, which was determined on high resolution.
@@
 \begin{itemize}
- \item Many RFI sources have a variable \index{geometric phase} geometric phase, because they move or because their path of propagation changes.
+ \item Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes.
 This would cause them to sum (partly) incoherently over time, and thus go approximately down with the noise.
@@
 In that case, a source will not add up coherently over the interferometers that see the particular source, as the interferometers see them with different phases.
- \item Many stationary sources are not constant over time, thus will be attenuated somewhat by \index{averaging} averaging.
+ \item Many stationary sources are not constant over time, thus will be attenuated somewhat by averaging.
  \item We have assumed 100\% of the spectrum is occupied by RFI.
-If, say, only 10\% of the spectrum is occupied, the expected value of the \index{leakage (RFI)} leaked \index{RFI!leakage} RFI level decreases with about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all.
+If, say, only 10\% of the spectrum is occupied, the expected value of the leaked RFI level decreases with about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all.
 With current data, one can only speculate how much the electromagnetic spectrum is truly occupied.
@@
 \begin{itemize}
- \item \index{fringe stopping} Fringe stopping interferometers can partly average out RFI sources.
+ \item Fringe stopping interferometers can partly average out RFI sources.
 Nevertheless, stationary RFI that is averaged out by fringe stopping will leave artefacts behind in the field centre \citep{post-correlation-filtering}.
-This is not relevant when observing the \index{NCP} North Celestial Pole --- which is one of the LOFAR EoR fields --- because no fringe stopping is applied when observing the NCP.
- \item The Fourier transform that is involved in data imaging will localize the contribution from \index{stationary RFI} stationary RFI near the NCP.
+This is not relevant when observing the North Celestial Pole --- which is one of the LOFAR EoR fields --- because no fringe stopping is applied when observing the NCP.
+ \item The Fourier transform that is involved in data imaging will localize the contribution from stationary RFI near the NCP.
 If RFI artefacts would show in the image of the NCP field, they can easily be detected and possibly be removed, or processing could ignore data near the pole.
@@
 At the same time, this field is useful for analysing the RFI coherency properties.
-It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright \index{foreground sources} foreground sources.
-Preliminary analysis of EoR NCP observations of a single night have reached the \index{thermal sky noise} thermal noise (Yatawatta et al., \textit{in preparation}), but do not show leaked RFI at the pole.
+It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright foreground sources.
+Preliminary analysis of EoR NCP observations of a single night have reached the thermal noise (Yatawatta et al., \textit{in preparation}), but do not show leaked RFI at the pole.
 
@@
 The current results indicate that a lot of RFI is not coherent, and the situation is promising.
-Considering the current RFI results, and the availability of further mitigation steps, it is very likely to assume that RFI will not be problematic for the detection Epoch of Reionisation.
+Considering the current RFI results, and the availability of further mitigation steps, it is very likely to assume that RFI will not be problematic for the detection Epoch of Reionisation with LOFAR.
 

### 2012-12-17 articles/Article-RFI-distribution/article-rfi-distribution.tex
Implemented Saleems comments

@@
 Feeble undetected sources of radio-frequency interference (RFI) might become visible after integration of long radio observations if they are coherently present over time.
-Thereby, they might occlude the astronomical signal of interest.
-This issue is especially important for epoch of reionization (EoR) projects that try to detect the faint redshifted HI signals of the earliest objects in the Universe.
+Thereby, they might obstruct the detection of the astronomical signal of interest.
+This issue is especially important for Epoch of Reionization (EoR) projects that try to detect the faint redshifted HI signals of the earliest objects in the Universe.
 We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR.
-The histograms shows a RFI distribution that follows a power-law distribution with an estimated exponent around -1.6 to -1.5.
+The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.6 to -1.5.
 With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models.
@@
 However, astronomical radio observations can be affected by radio-frequency interference (RFI), which makes it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}.
-Many techniques have been designed to mitigate its effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, LOFAR-RFI-pipeline}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.
+Many techniques have been designed to mitigate this effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, LOFAR-RFI-pipeline}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.
 In most cases, these methods excise enough of the interference to calibrate and image the data.
 
-Regular radio observations record one or a few sessions of typical a few hours and results are integrated.
+Typical radio observations record a few hours of data, and the results are integrated.
 In these cases, it often suffices to only excise interference that is apparent and thus above the noise to reach the thermal noise limit of the instrument.
-A new challenge arises when one desires much deeper observations, and hundreds of hours of observations need to be integrated.
+A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated.
 In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but is coherently present in the data.
 Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest.
-This is very relevant for 21-cm Epoch of Reionisation (EoR) experiments, because these involve long integration times.
-Several such experiments are underway, both to measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or global signal \citep{edges}.
+This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times.
+Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}.
 For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest.
@@
 
-In Sect.~\ref{sec:distribution-prediction}, we predict the terrestrial interfering source distribution based on various assumptions.
+This paper is organised as follows: in Sect.~\ref{sec:distribution-prediction}, we predict the terrestrial interfering source distribution based on various assumptions.
 Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse the histograms.
@@
 
-We assume that the source is observed fully coherent, but a possible decoherence factor can be absorbed in the gains.
+We assume that the source observed is fully coherent, but a possible decoherence factor can be absorbed in the gains.
 Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the narrow-band condition.
@@
 
-Next, we assume that all sources have equal constant strength $I$.
-Moreover, we assume that the sources have a uniform spatial distribution in a two-dimensional plane.
+Next, we assume that all sources have equal constant strength $I$ and follow a uniform spatial distribution in a two-dimensional plane.
 This is obviously a simplification, as the sources are actually distributed on the surface of a sphere.
-Beyond some distance, the Earth will also partly block the path between transmitter and receiver.
-Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid and only for a limited distance.
+Hence, beyond some distance, the Earth will partly block the path between transmitter and receiver.
+Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid.
 Using these assumptions, we can express the expected cumulative frequency density of sources $f_I(r)$ at distance $r$ as
@@
 \begin{align} \label{eq:cumulative-rfi-distribution}
- F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{S}^{-1}(S))\\
+ F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{S}^{-1}(r))\\
 \notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{S})\\
@@
 \end{align}
-where $\mathcal{S}^{-1}$ refers to the inverse of $\mathcal{S}(r)$, i.e., the function that returns the distance $r$ for a given amplitude $S$.
+where $\mathcal{S}^{-1}$ is the inverse of $\mathcal{S}(r)$, i.e., the function that returns the distance $r$ for a given amplitude $S$.
 Finally, the differential frequency density can be calculated by taking the derivative,
@@
 A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}.
+The original model is based on urban areas, but corrections are given by \citeauthor{hata-propagation-loss} for sub-urban areas with a lower population density and for open areas.
 \citeauthor{hata-propagation-loss} gives the following analytical estimate for $L_p$, the electromagnetic propagation loss over land:
@@
 \end{align}
-where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in m; $h_m$ the height of the receiving antenna in m; $r$ the distance between the antennae in m; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density.
+where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in meters; $h_m$ the height of the receiving antenna in meters; $r$ the distance between the antennae in meters; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density.
 \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.
@@
 \end{equation}
-The model is based on urban areas, but corrections are given by \citeauthor{hata-propagation-loss} for sub-urban areas with a lower population density and for open areas.
-These corrections are independent of $r$, thus would only change $\zeta$.
 Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density.
@@
 The curves in Fig.~\ref{fig:rayleigh-and-rfi-combined} are from histograms of samples that are all contaminated by both noise and RFI.
-In practice, we can not make this distinction 100\% accurately, as RFI detectors have a limited accuracy.
+In practice, we can not make this distinction accurately, as RFI detectors have a limited accuracy.
 Consequently, the observed histogram will be the sum of two types of histograms: the first histogram being contaminated by both noise and RFI, the second one only by noise.
@@
 At the point of writing, LOFAR beam models are still being developed and are not yet well parametrized.
-However, since it is likely that most RFI sources are observed at the edges of the beam, it can be expected that the beam will have a benign effect on the histogram of an observation, comparable with the effect of the frequency response.
+However, since it is likely that most RFI sources are observed at the edges of the beam, it can be expected that the beam will have a limited effect on the histogram of an observation, comparable with the effect of the frequency response.
 
@@
 In practice, the observed histogram will break down beyond some amplitude because of several reasons.
-First and most importantly, the samples itself are digitized with an analogue-to-digital converter (ADC) with limited range.
+First and most importantly, the samples themselves are digitized with an analogue-to-digital converter (ADC) with limited range.
 Second, we observe for a limited time and the frequency count is discrete.
@@
 In some cases there will be no solution for $S_U$.
-The reason for this is that the integral of the distribution function converges, and the total number of samples is finite even when $S \rightarrow \infty$.
+This situation arises because the integral of the distribution function is finite even when $S_U \rightarrow \infty$.
 If $N_\textrm{part}$ exceeds the integrated value of the fitted distribution function, there is no solution.
@@
 where $SS_{xx}$, $SS_{xy}$ and $SS_{yy}$ are the sums of squares, e.g., $SS_{xy}=\sum_{i=1}^n (x_i - \bar x) (y_i - \bar y)$ and $n$ is the number of samples.
-However, we found that this is not a representable error in our case, because the errors in the slope are not normally distributed.
+However, we found that this is not a representative error in our case, because the errors in the slope are not normally distributed.
 The noise in the part of the histogram over which the slope is calculated, is very small due to the large amount of samples.
@@
 Because the number of samples is very large ($>10^{11}$), the $\mathcal{O}$-term will be very small.
-Therefore, we will calculate the quantity without the term.
+Therefore, we will calculate the quantity without this term.
 As with the standard error for the slope of a straight line in Eq.~\eqref{eq:stderr-slope}, the standard error for the Hill estimator yields very small quantities.
@@
 Fig.~\ref{fig:plot-slopes} shows the slope of the log-log plot as a function of amplitude, which was constructed by performing linear regression in a sliding window, with a window size of 1 decade.
-The HBA shows very little structure in the slope, but the HBA is less stable and shows some features in its power-law part.
+The HBA shows very little structure in the slope, but the LBA is less stable and shows some features in its power-law part.
 Linear regression on the power-law part of the log-log plot results in a slope of $-1.62$ for the LBA and $-1.53$ for the HBA.
@@
 By using $\tilde{S}_L$ instead of $S_L$, it is a few percent lower.
-By assuming a 100\% RFI occupancy, we find that the expected value of leaked RFI is $484$--$496$ mJy for the LBA and $167$--$171$ mJy.
+By assuming a 100\% RFI occupancy, we find that the expected value of leaked RFI is $484$--$496$ mJy for the LBA and $167$--$171$ mJy for the HBA.
 By assuming $10$\% occupancy, the value for $E(S_\textrm{leak})$ is about $25$\% reduced.
@@
 For the EoR project with $1$~MHz resolution, Eq.~\ref{eq:rfi-noise-level} results in $\sigma_\textrm{RFI}\approx 4.7$ mJy.
-However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz subband (Yatawatta et al., \textit{in preparation}), and the resulting images show no signs of RFI.
+However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz subband (Yatawatta et al., \textit{submitted}), and the resulting images show no signs of RFI.
 This implies that either the upper limit is far from the actual RFI situation, or Eq.~\ref{eq:rfi-noise-level} is not applicable to most of the RFI that is observed with LOFAR.
@@
  \item Fringe stopping interferometers can partly average out RFI sources.
-Nevertheless, stationary RFI that is averaged out by fringe stopping will leave artefacts behind in the field centre \citep{post-correlation-filtering}.
+Nevertheless, stationary RFI that is averaged out by fringe stopping will leave artefacts in the field centre \citep{post-correlation-filtering}.
 This is not relevant when observing the North Celestial Pole --- which is one of the LOFAR EoR fields --- because no fringe stopping is applied when observing the NCP.
@@
 It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright foreground sources.
-Preliminary analysis of EoR NCP observations of a single night have reached the thermal noise (Yatawatta et al., \textit{in preparation}), but do not show leaked RFI at the pole.
+Preliminary analysis of EoR NCP observations of a single night have reached the thermal noise (Yatawatta et al., \textit{submitted}), but do not show leaked RFI at the pole.
 
@@
 \section*{Acknowledgments}
-TODO
+LOFAR, the Low-Frequency Array designed and constructed by ASTRON, has facilities in several countries, that are owned by various parties (each with their own funding sources), and that are collectively operated by the International LOFAR Telescope (ILT) foundation under a joint scientific policy.
 

### 2012-12-19 articles/Article-RFI-distribution/article-rfi-distribution.tex
Updating article with Wim Brouw's comments

@@
                /Author     (A. R. Offringa et al.)
-               /Keywords   (dark ages, reionization, first stars;atmospheric effects;instrumentation: interferometers;methods: observational;techniques: interferometric;radio continuum: general;rfi)
+               /Keywords   (dark ages, reionisation, first stars;atmospheric effects;instrumentation: interferometers;methods: observational;techniques: interferometric;radio continuum: general;rfi)
         }
@@
 Thereby, they might obstruct the detection of the astronomical signal of interest.
-This issue is especially important for Epoch of Reionization (EoR) projects that try to detect the faint redshifted HI signals of the earliest objects in the Universe.
+This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals of the earliest objects in the Universe.
 We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR.
@@
 \begin{keywords}
-atmospheric effects -- instrumentation: interferometers -- methods: observational -- techniques: interferometric -- radio continuum: general -- dark ages, reionization, first stars
+atmospheric effects -- instrumentation: interferometers -- methods: observational -- techniques: interferometric -- radio continuum: general -- dark ages, reionisation, first stars
 \end{keywords}
@@
 A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated.
-In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but is coherently present in the data.
+In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be coherently present in the data.
 Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest.
@@
 Even ground-based sources might be spread differently.
-For example, it can be expected that citizens' band (CB) devices, that are often used in cars, are differently distributed from broadcasting transmitters.
+For example, it can be expected that citizens' band (CB) devices, that are often used in cars, are distributed differently from broadcasting transmitters.
 The frequencies of deliberately transmitting sources of a particular class of devices are constrained by the bands that have been allocated for the given class, which might allow one to distinguish different sources to some extent.
@@
 Consider an interfering point source of strength $I$.
-This source is observed by an interferometer that consists of two antennae or stations with gains $g_1,g_2$, which include all instrumental effects.
-The antennae are located at distances $r_1,r_2$ from the source.
+This source is observed by an interferometer that consists of two antennas or stations with gains $g_1,g_2$, which include all instrumental effects.
+The antennas are located at distances $r_1,r_2$ from the source.
 The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
@@
 Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth.
-We will limit our analysis to cross-correlated antennae; the auto-correlations will be ignored.
+We will limit our analysis to cross-correlated antennas; the auto-correlations will be ignored.
 
@@
 
-Now, we can treat the interferometer geometrically as a single point, as both antennae will see the same distribution.
+Now, we can treat the interferometer geometrically as a single point, as both antennas will see the same distribution.
 Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and gain $g$ as
@@
 
-Next, we assume that all sources have equal constant strength $I$ and follow a uniform spatial distribution in a two-dimensional plane.
-This is obviously a simplification, as the sources are actually distributed on the surface of a sphere.
-Hence, beyond some distance, the Earth will partly block the path between transmitter and receiver.
+Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in a two-dimensional plane.
+This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights.
+Beyond some distance, the Earth will partly block the path between transmitter and receiver.
 Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid.
@@
 \end{equation}
-for some constant $c$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance of $r$ away.
+for some constant $c$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance $r$ away.
 
@@
 If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes.
-In this case, the number of sources corresponds with the surface area of a hemisphere.
-By using the formula for the surface area of a hemisphere and some basic geometry, one finds that
+In this case, the RFI sources are distributed over the surface of a hemisphere.
+If we assume their difference in height is negligible, and by using the formula for the surface area of a hemisphere and some basic geometry, one finds that
 \begin{equation}\label{eq:cumulative-spherical}
@@
 \subsection{Propagation effects}
-So far, we have assumed that the electromagnetic radiation propagates through free space, resulting in a $r^{-2}$ fall-off.
+So far, we have assumed that the electromagnetic radiation propagates through free space, resulting in an $r^{-2}$ fall-off.
 In reality, the radiation will be affected by complicated propagational effects.
@@
 \end{align}
-where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in meters; $h_m$ the height of the receiving antenna in meters; $r$ the distance between the antennae in meters; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density.
+where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in meters; $h_m$ the height of the receiving antenna in meters; $r$ the distance between the antennas in meters; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density.
 \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.
@@
 Real RFI detectors can separate these distributions to some extent, but due to false positives and false negatives, the histograms will get mixed nevertheless.
-Even more, we still need to take into account that the RFI classified samples are also affected by noise.
-Although this effect is moderate, because 'most' RFI samples are of much higher amplitude than what is added because of the noise, it is the low-level RFI samples which we are interested in, and the relative effect of noise on those samples is large.
+Even more, we still need to take into account that the samples classified as RFI are also affected by noise.
+Although this effect is moderate, because 'most' RFI samples are of much higher amplitude than what is added because of the noise, we are interested in the low-level RFI samples, and the relative effect of noise on those samples is large.
 
@@
 
-To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as algebraic calculation is not trivial.
+To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial.
 The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes.
@@
 \end{equation}
-with $x_n \sim N(\mu=0; \sigma=1)$. Of course, the final amplitude sample $S_A$ can be calculated with 
+with $x_n \sim N(\mu=0; \sigma=1)$. The final amplitude sample $S_A$ can be calculated with 
 \begin{equation} \label{eq:drawing-amplitudes}
@@
 
-One instrumental effect that is absorbed in $g$ is the frequency response of the instrument, i.e., the antenna response in combination with the passband of the analogue and digital filters.
+One instrumental effect that is absorbed in $g$ is the frequency response of the instrument, i.e., the antenna response in combination with the band-pass of the analogue and digital filters.
 Because the data that are analysed in Sect.~\ref{sec:dist-results} have initially not been band-pass calibrated, the instrumental response is not uniform over frequency.
-We determined that the variation due to the band-pass is about one order of magnitude in the LBA and about a factor of two in the HBA.
+We determined that the variation due to the band-pass is about one order of magnitude for the low-band antennas (LBA, 30.1--77.5 MHz) and about a factor of two for the high-band antennas (HBA, 115.0--163.3 MHz).
 
@@
 %In Fig.~\ref{fig:plot-passband-histogram}, the histograms of the frequency response are given.
-To also demonstrate the effect of mixing distributions, histograms are added for samples that are drawn from a uniform, Gaussian and -2 power-law distribution as in Eq.~\eqref{eq:drawing-rfi-distribution} with $Ig/r_L^2=1$, that have been multiplied by the passband.
+To also demonstrate the effect of mixing distributions, histograms are added for samples that are drawn from a uniform, Gaussian and -2 power-law distribution as in Eq.~\eqref{eq:drawing-rfi-distribution} with $Ig/r_L^2=1$, that have been multiplied by the band-pass.
 The frequency responses were determined by measuring the standard deviation of the data over frequency after flagging the data for interference.
@@
 At the point of writing, LOFAR beam models are still being developed and are not yet well parametrized.
-However, since it is likely that most RFI sources are observed at the edges of the beam, it can be expected that the beam will have a limited effect on the histogram of an observation, comparable with the effect of the frequency response.
+It is likely that most RFI sources are observed at the edges of the beam.
+Because most sources will be observed with similar gains (within one order of magnitude), it can be expected that the beam will have a limited effect on the histogram properties of an observation, and is comparable with the effect of the frequency response.
 
@@
 \subsection{Estimating $\sigma$ and slope parameters}
-The mode $\hat\sigma$ of the Rayleigh distribution is estimated by finding the amplitude with the maximum occurrences, i.e., the amplitude corresponding to the peak of the histogram.
+The mode $\sigma$ of the Rayleigh distribution is estimated by finding the amplitude with the maximum occurrences, i.e., the amplitude corresponding to the peak of the histogram.
 The slope is estimated using linear regression over a visually selected interval.
@@
 Because our bin size is small and the distribution is well sampled, the resulting estimate will be close to the normal Hill estimator.
-This overcomes the need to iterate over all data.
+This avoids the need to iterate over all data.
 
@@
 For Fig.~\ref{fig:explanation-constraints} this means that the areas labelled $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal.
+The area labelled $N_b$ is assumed to be zero for now, which assumes the power law has a sharp cut-off on the left side (e.g., because of the curvature of the Earth).
 Solving this equality, one finds that
@@
 \citet{lofar-radio-environment}. % TODO
-In one set, the low-band antennae (LBA) were used and the frequency range 30.1--77.5 MHz was recorded, while in the other the high-band antennae (HBA) were used to record the frequency range 115.0--163.3 MHz.
+In one set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz.
 More stations were used in the LBA set.
@@
 \includegraphics[width=12cm]{img/histograms-raw-trimmed}
-\caption{The histograms of the two data sets before pass-band correction and flux calibration.}
+\caption{The histograms of the two data sets before band-pass correction and flux calibration.}
 \label{fig:Histograms-raw}
@@
 \includegraphics[width=8cm]{img/plot-lba-dist-per-frequency-trimmed}
-\caption{Histograms for $5$ different $0.2$ MHz LBA sub-bands without pass-band correction and flux calibration.
+\caption{Histograms for $5$ different $0.2$ MHz LBA sub-bands without band-pass correction and flux calibration.
 The continuous lines represent the data before RFI flagging.
-The dashes lines are the histograms of the samples that have been classified as RFI.}
+The dashed lines are the histograms of the samples that have been classified as RFI.}
 \label{fig:plot-dist-per-frequency-LBA}
@@
 \includegraphics[width=8cm]{img/plot-hba-dist-per-frequency-trimmed}
-\caption{Histograms for $5$ different $0.2$ MHz HBA sub-bands without pass-band correction and flux calibration.
+\caption{Histograms for $5$ different $0.2$ MHz HBA sub-bands without band-pass correction and flux calibration.
 The continuous lines represent the data before RFI flagging.
-The dashes lines are the histograms of the samples that have been classified as RFI.}
+The dashed lines are the histograms of the samples that have been classified as RFI.}
 \label{fig:plot-dist-per-frequency-HBA}
@@
 \includegraphics[width=13cm]{img/histograms-corrected-trimmed}
-\caption{Observed LBA distribution after pass-band correction and flux calibration.
+\caption{Observed LBA distribution after band-pass correction and flux calibration.
 $S_{l1}$ and $S_{l2}$ denote the limits of the distribution with a sharp lower cut-off (Eq.~\eqref{eq:lower-limit-1}) and uniform lower limit (Eq.~\eqref{eq:lower-limit-2}), $S_d$ is the average lower limit of RFI that is detected.}
@@
 Consequently, it can be expected that other stochastic effects, such as the intrinsic source strength and the beam gain due to a differing direction of arrival, will similarly not effect the power law.
-Because the pass-band corrected histograms provide a more representative analysis, we will use the corrected histograms for further analysis.
+Because the band-pass corrected histograms provide a more representative analysis, we will use the corrected histograms for further analysis.
 
@@
 \includegraphics[width=13cm]{img/histograms-with-fits-trimmed}
-\caption{Least-squares fits of Rayleigh distributions to the observed LBA and HBA histograms, after pass-band correction but without flux calibration.}
+\caption{Least-squares fits of Rayleigh distributions to the observed LBA and HBA histograms, after band-pass correction but without flux calibration.}
 \label{fig:histograms-with-fits}
@@
 We found this is caused by RFI observed for about an hour in the late afternoon in the lower LBA frequency regime, around $30$--$40$~MHz.
-Leaving this frequency range out flattens the bulge significantly, but does not completely eliminate it, because the source turned the receivers in non-linear state, causing leakage at lower intensity levels in the other sub-bands.
+Leaving this frequency range out flattens the bulge significantly, but does not completely eliminate it, because the source turned the receivers to a non-linear state, causing leakage at lower intensity levels in the other sub-bands.
 Unlike linear regression, the fitting region of the Hill estimator is not limited at the high end.
@@
 $N_\textrm{total}$ & Total number of samples in histogram & $8.0\times10^{11}$ & $5.4\times10^{11}$ \\
-$\sigma$ & Rayleigh mode (assumed to be SEFD/$\sqrt{2\Delta t \Delta \nu}$) & $770$ Jy & $77$ Jy \\
+$\sigma$ & Rayleigh mode (assumed to be SEFD/$\sqrt{2\Delta t \Delta \nu}$, & $770$ Jy & $77$ Jy \\
+ & where SEFD is the System Equivalent Flux Density) & & \\
 \hline
@@
 As an example, we will analyse the situation for the LOFAR EoR project.
-This project will use the LOFAR high-band antennae to collect on the order of 50--100 night-time observations of 6 h for a few target fields.
+This project will use the LOFAR high-band antennas to collect on the order of 50--100 night-time observations of 6 h for a few target fields.
 The final resolution required for signal extraction will be about $1$~MHz.
@@
 Therefore, the amount that time integration can lower the flux density of RFI might be limited.
-Additionally, some RFI sources are received by multiple stations of the array, and by multiple feeds of the individual antennae.
+Additionally, some RFI sources are received by multiple stations of the array, and by multiple feeds of the individual antennas.
 Therefore, integrating data from different interferometers and data from the two polarized feeds might not lower the noise level that is caused by RFI at some point.
@@
  \item For the shortest baselines at 150~MHz, the far field starts around 1~km.
-Some RFI sources will therefore be in the near field, especially in the longer baselines.
+Some RFI sources will therefore be in the near field, especially for the longer baselines.
 In that case, a source will not add up coherently over the interferometers that see the particular source, as the interferometers see them with different phases.

### 2012-12-19 articles/Article-RFI-distribution/article-rfi-distribution.tex
Add authors Wim and Benedetta

@@
   A.~G.~de~Bruyn$^{2,3}$,
-   and S.~Zaroubi$^{2}$\\
+   S.~Zaroubi$^{2}$,
+   W.~Brouw$^{2,3}$,
+   B.~Ciardi$^{4}$\\
 $^{1}$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
 $^{2}$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
-$^{3}$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands
+$^{3}$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
+$^{4}$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany.
 }

### 2013-01-04 articles/Article-RFI-distribution/article-rfi-distribution.tex
Slight modification of abstract
Adding new authors

--- 2012-12-19 articles/Article-RFI-distribution/article-rfi-distribution.tex
@@
 A.~G.~de~Bruyn$^{2,3}$,
-   and S.~Zaroubi$^{2}$\\
+S.~Zaroubi$^2$,
+W.~Brouw$^{2,3}$,
+B.~Ciardi$^4$,
+I.~T.~Iliav$^5$,
+\newauthor
+G.~J.~A.~Harker$^6$,
+L.~V.~E.~Koopmans$^2$,
+G.~Mellema$^7$,
+S.~J.~Wijnholds$^3$
+\\
 $^{1}$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
 $^{2}$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
-$^{3}$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands
+$^{3}$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
+$^{4}$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
+$^{5}$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
+$^{6}$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
+$^7$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
 }
@@
 Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present.
-Since these observations have achieved the thermal (sky) noise limit, a logical implication is that most of the RFI sources do not add coherently.
+Since these observations have achieved the thermal (sky) noise limit, a possible implication is that most of the RFI sources do not add coherently.
 \end{abstract}

### 2013-01-07 articles/Article-RFI-distribution/article-rfi-distribution.tex
First part of Garrelts comments

@@
 \begin{abstract}
-Feeble undetected sources of radio-frequency interference (RFI) might become visible after integration of long radio observations if they are coherently present over time.
-Thereby, they might obstruct the detection of the astronomical signal of interest.
-This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals of the earliest objects in the Universe.
+Feeble undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are coherently present over time.
+Thereby, they might obstruct the detection of the weak astronomical signals of interest.
+This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest objects in the Universe.
 We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR.
-The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.6 to -1.5.
+The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
 With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models.
 Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present.
-Since these observations have achieved the thermal (sky) noise limit, a possible implication is that most of the RFI sources do not add coherently.
+Since these observations have achieved the thermal (sky) noise limit, we conclude that it is probable that most of the RFI sources do not add coherently.
 \end{abstract}
@@
 Many techniques have been designed to mitigate this effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, LOFAR-RFI-pipeline}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.
-In most cases, these methods excise enough of the interference to calibrate and image the data.
+In most cases, these methods excise enough of the interference to allow calibration and imaging of the data.
 
 Typical radio observations record a few hours of data, and the results are integrated.
-In these cases, it often suffices to only excise interference that is apparent and thus above the noise to reach the thermal noise limit of the instrument.
+In these cases, it often suffices to only excise interference that is apparent and thus above the noise, to reach the thermal noise limit of the instrument.
 A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated.
@@
 For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest.
-In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array \citep{de-bruyn-eor-ursi-2011}.
+In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) \citep{de-bruyn-eor-ursi-2011}.
 
@@
 A distribution can be expressed either as a differential or cumulative function.
-This work uses differential distributions, because they are better suited for visualizing the parameters we are interested in, although the two basically convey the same information.
+This work uses differential distributions, because they are better suited for visualizing the parameters we are interested in, although the two distributions basically convey the same information.
 
@@
 Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid.
-Using these assumptions, we can express the expected cumulative frequency density of sources $f_I(r)$ at distance $r$ as
+Using these assumptions, we can express the expected cumulative frequency density of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
 \begin{equation} \label{eq:cumulative-distance-function}
@@
 
-Converting from a subtrahend in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
+Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
 \begin{equation} \label{eq:propagation-loss-factor}
@@
 By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-c \pi r_L^2$ for $S < Ig/r_L^2$.
-Consequently, by sampling $x_u$ from a uniform distribution with $0 \le x_u \le N-c \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ the power-law distribution.
+Consequently, by sampling $x_u$ from a uniform distribution with $0 \le x_u \le N-c \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ a power-law distribution.
 
@@
 Once the exponent of the power-law part of the distribution is estimated with the previously discussed techniques, the distribution can be extrapolated to find a lower flux limit.
-Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the following frequency density function $g$:
+Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the following frequency density function $h$:
 \begin{equation} \label{eq:amplitude-power-law}
-g(S) = \beta S^\alpha.
+h(S) = \beta S^\alpha.
 \end{equation}
@@
 \begin{equation} \label{eq:upper-limit-constrained}
- \int\limits_{S_1}^{S_U} g(S) dS = N_\textrm{part}.
+ \int\limits_{S_1}^{S_U} h(S) dS = N_\textrm{part}.
 \end{equation}
@@
 \begin{equation}
- \int\limits_{\tilde S_L}^{S_U} g(S) dS + \tilde S_{L} g(\tilde S_{L}) = N_\textrm{total} - N_\textrm{part},
+ \int\limits_{\tilde S_L}^{S_U} h(S) dS + \tilde S_{L} h(\tilde S_{L}) = N_\textrm{total} - N_\textrm{part},
 \end{equation}
@@
 The stations that have been used are geometrically spread over an area of about 80~km and 30~km in diameter at maximum for the LBA and HBA sets respectively.
+For EoR detection experiments, the HBA are more important than the LBA, because they cover the frequency range of the redshifted EoR signal.
 
@@
 Number of stations  &  33           & 13 \\
-Total size          & 96.3 TiB        & 18.6 TiB \\
+Total size          & 96.3 TB        & 18.6 TB \\
 \hline
@@
 We found this is caused by RFI observed for about an hour in the late afternoon in the lower LBA frequency regime, around $30$--$40$~MHz.
-Leaving this frequency range out flattens the bulge significantly, but does not completely eliminate it, because the source turned the receivers to a non-linear state, causing leakage at lower intensity levels in the other sub-bands.
+Leaving this frequency range out flattens the bulge significantly, but does not completely eliminate it, because the source put the receivers in a non-linear state, causing leakage at lower intensity levels in the other sub-bands.
 Unlike linear regression, the fitting region of the Hill estimator is not limited at the high end.
@@
 Because some RFI sources might be stationary, the signals from these sources will add coherently over time.
-Therefore, the amount that time integration can lower the flux density of RFI might be limited.
+Therefore, the amount that time integration can decrease the flux density of RFI might be limited.
 Additionally, some RFI sources are received by multiple stations of the array, and by multiple feeds of the individual antennas.
@@
 \subsection{Interference-reducing effects} \label{sec:coherence-reduction}
-When integrating data, it is likely that the actual noise limit will be significantly lower than the given upper limit, which was determined on high resolution.
+When integrating data, it is likely that the actual noise limit from low-level RFI will be significantly lower than the given upper limit, which was determined at highest LOFAR resolution, before further averaging.
 There are several reasons for this, which we will summarize one by one.
@@
 
-With the current strategy, it is likely that the LOFAR EoR project will encounter some RFI on some frequencies when averaging lots of observing nights, although this still has to be seen.
+With the current strategy, it is likely that the LOFAR EoR project will encounter some RFI on some frequencies when averaging lots of observing nights, although this still remains to be seen.
 To mitigate this leaked RFI, the detection can be executed at higher signal-to-noise levels.
 The current results indicate that a lot of RFI is not coherent, and the situation is promising.
-Considering the current RFI results, and the availability of further mitigation steps, it is very likely to assume that RFI will not be problematic for the detection Epoch of Reionisation with LOFAR.
+Considering the current RFI results, and the availability of further mitigation steps, we conclude that RFI will likely not be problematic for the detection Epoch of Reionisation with LOFAR.
 

### 2013-01-07 articles/Article-RFI-distribution/article-rfi-distribution.tex
Small change

--- 2013-01-04 articles/Article-RFI-distribution/article-rfi-distribution.tex
@@
 Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present.
-Since these observations have achieved the thermal (sky) noise limit, a possible implication is that most of the RFI sources do not add coherently.
+These observations almost reach the thermal noise and are thought not to be limited by RFI.
+A possible implication is that most RFI sources do not add coherently.
 \end{abstract}

### 2013-01-07 articles/Article-RFI-distribution/article-rfi-distribution.tex

--- 2013-01-07 articles/Article-RFI-distribution/article-rfi-distribution.tex
@@
 \begin{abstract}
-Feeble undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are coherently present over time.
-Thereby, they might obstruct the detection of the weak astronomical signals of interest.
-This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest objects in the Universe.
+Feeble undetected sources of radio-frequency interference (RFI) might become visible after integration of long radio observations if they are coherently present over time.
+Thereby, they might obstruct the detection of the astronomical signal of interest.
+This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals of the earliest objects in the Universe.
 We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR.
-The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
+The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.6 to -1.5.
 With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models.
 Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present.
-Since these observations have achieved the thermal (sky) noise limit, we conclude that it is probable that most of the RFI sources do not add coherently.
+These observations almost reach the thermal noise and are thought not to be limited by RFI.
+We conclude that it is likely that most RFI sources do not add coherently.
 \end{abstract}

### 2013-01-07 articles/Article-RFI-distribution/article-rfi-distribution.tex
Stefan's comments part I

--- 2013-01-07 articles/Article-RFI-distribution/article-rfi-distribution.tex
@@
 \begin{abstract}
-Feeble undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are coherently present over time.
+Faint undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are stationary and coherently present over time.
 Thereby, they might obstruct the detection of the weak astronomical signals of interest.
@@
 We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR.
-The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
+The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
 With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models.
@@
 A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated.
-In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be coherently present in the data.
+In such a case, weak interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be coherently present in the data.
 Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest.
@@
 We analyse such a double-logarithmic plot for the case of terrestrial sources, with the ultimate goal of predicting their full spatial and brightness distributions.
-This results in a better insight in the effects of low-level interference and allows one to simulate the effects of interference more accurately.
+This results in a better insight into the effects of low-level interference and allows one to simulate the effects of interference more accurately.
 A distribution can be expressed either as a differential or cumulative function.
@@
 For example, it can be expected that citizens' band (CB) devices, that are often used in cars, are distributed differently from broadcasting transmitters.
-The frequencies of deliberately transmitting sources of a particular class of devices are constrained by the bands that have been allocated for the given class, which might allow one to distinguish different sources to some extent.
+For deliberate transmitters, the frequency at which interference occurs can identify the involved class of devices, because devices are constrained by the bands that have been allocated for the given class.
 
@@
 
-Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we predict to see a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane.
-Therefore, RFI follows a power-law distribution.
+Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we predict to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane.
 
@@
 This is mandated by the large dynamic range of the histogram that we are interested in.
-Therefore, we chose to have a bin size that increases linearly with the amplitude $S$, and the frequency distribution is normalized by the bin size after counting.
+Therefore, we chose to have a bin size that increases linearly with the amplitude $S$, and the frequency counts are divided by the bin size after counting.
 Consequently, in parts of the histogram that have a sparse number of samples, the outlying samples will follow a $1/S$ curve, or a -1 slope in a log-log plot.
@@
 Therefore, using this estimator will result in an estimate that is steeper (i.e., more negative) than the actual distribution.
-Moreover, the estimator requires an iteration over all data points, and because of the size of the data sets this is somewhat impractical.
+Moreover, the estimator requires reading and processing all data points, and because of the size of the data sets this is somewhat impractical.
 However, we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ the bin frequency.
@@
 Both Figs.~\ref{fig:plot-dist-per-frequency-LBA} and \ref{fig:plot-dist-per-frequency-HBA} show that the power-law part of the distributions are very different for different sub-bands.
-Nevertheless, combining the data of all the sub-bands results in reasonably-stable power-law distributions.
+Nevertheless, combining the data of all the sub-bands results in reasonably stable power-law distributions.
 This indicates that the individual sub-bands have an approximately similar underlying power-law distribution, that is not yet apparent because not enough samples are combined in the histograms of individual sub-bands.
@@
 This validates that the variable gain that is caused by the antenna response does not change the observed power law.
-Consequently, it can be expected that other stochastic effects, such as the intrinsic source strength and the beam gain due to a differing direction of arrival, will similarly not effect the power law.
-Because the band-pass corrected histograms provide a more representative analysis, we will use the corrected histograms for further analysis.
+Consequently, it can be expected that other stochastic effects, such as the intrinsic source strength and the beam gain due to a differing direction of arrival, will similarly not affect the power law.
+Because the band-pass corrected histograms should provide a more accurate analysis, we will use the corrected histograms for further analysis.
 
@@
 It breaks down about one order of magnitude before the mode of the distributions.
-This is because of the multi-variate nature of the distributions, as was described in \S\ref{sec:histogram-noise}.
+This is because of the multi-component nature of the distributions, as was described in \S\ref{sec:histogram-noise}.
 
@@
  \item We have assumed 100\% of the spectrum is occupied by RFI.
-If, say, only 10\% of the spectrum is occupied, the expected value of the leaked RFI level decreases with about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all.
-With current data, one can only speculate how much the electromagnetic spectrum is truly occupied.
+If, say, only 10\% of the spectrum is occupied, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all.
+With current data, one can only speculate how much of the electromagnetic spectrum is truly occupied.
  \item Stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other.
@@
 The current results indicate that a lot of RFI is not coherent, and the situation is promising.
-Considering the current RFI results, and the availability of further mitigation steps, we conclude that RFI will likely not be problematic for the detection Epoch of Reionisation with LOFAR.
+Considering the current RFI results, and the availability of further mitigation steps, we conclude that RFI will likely not be problematic for the detection of the Epoch of Reionisation with LOFAR.
 

### 2013-01-08 articles/Article-RFI-distribution/article-rfi-distribution.tex

@@
 We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR.
-The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
+The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
 With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models.
 Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present.
-Since these observations have achieved the thermal (sky) noise limit, we conclude that it is probable that most of the RFI sources do not add coherently.
+These observations almost reach the thermal noise and are thought not to be limited by RFI.
+Therefore, we conclude that it is likely that most RFI sources do not add coherently, in which case RFI will not prevent LOFAR from detecting the EoR.
 \end{abstract}
@@
 \section{Predicting the brightness distribution} \label{sec:distribution-prediction}
-Interference is generated by many different kind of transmitters, and these will have different spatial and brightness distributions (``spatial'' refers here to the distribution on the Earth).
+Interference is generated by many different kinds of transmitters, and these will have different spatial and brightness distributions (``spatial'' refers here to the distribution on the Earth).
 For example, aeroplanes and satellites have widely different heights, while other sources are ground-based.
@@
 Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth.
+The formula represents a spherically propagating wave in free space.
 We will limit our analysis to cross-correlated antennas; the auto-correlations will be ignored.
@@
 Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the narrow-band condition.
-For example, if the bandwidth of the signal $\Delta\nu=1$ kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c / (2 \pi \Delta \nu) \approx$ 50 km.
+For example, if the bandwidth of the signal $\Delta\nu=1$ kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c (2 \pi \Delta \nu)^{-1} \approx$ 50 km.
 
 Now, we can treat the interferometer geometrically as a single point, as both antennas will see the same distribution.
-Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and gain $g$ as
+Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and interferometric gain $g$ as
 \begin{equation} \label{eq:amplitude-fall-off-in-free-space}
@@
 \begin{equation} \label{eq:cumulative-distance-function}
-F_{\textrm{distance} \le r}(r)=c\pi r^2,
+F_{\textrm{distance} \le r}(r)=\rho \pi r^2,
 \end{equation}
-for some constant $c$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance $r$ away.
+for some constant $\rho$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance $r$ away.
 
 We need the reverse of Eq.~\eqref{eq:cumulative-distance-function} to predict the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance.
-Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - c\pi r^2$, with $N$ the total number of sources.
+Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2$, with $N$ the total number of sources.
 
@@
 \notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{S})\\
-\notag = & N-\frac{c\pi Ig}{S}
+\notag = & N-\frac{\rho \pi Ig}{S}
 \end{align}
-where $\mathcal{S}^{-1}$ is the inverse of $\mathcal{S}(r)$, i.e., the function that returns the distance $r$ for a given amplitude $S$.
+where $\mathcal{S}^{-1}(r)$ is the inverse of $\mathcal{S}(r)$, i.e., the function that returns the distance $r$ for a given amplitude $S$.
 Finally, the differential frequency density can be calculated by taking the derivative,
 \begin{align} \label{eq:two-dimensional-distribution}
- f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{c\pi Ig}{S^2}.
+ f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{\rho \pi Ig}{S^2}.
 \end{align}
@@
 If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes.
-In this case, the RFI sources are distributed over the surface of a hemisphere.
+In this case, the RFI sources are distributed over the surface of a sphere.
 If we assume their difference in height is negligible, and by using the formula for the surface area of a hemisphere and some basic geometry, one finds that
 \begin{equation}\label{eq:cumulative-spherical}
-\tilde{F}_{\textrm{distance} \le r} = c 2\pi R^2\left(1-\cos(2\arcsin \frac{r}{2R})\right) \hspace{0.4cm}\textrm{with } r \le 2R,
+\tilde{F}_{\textrm{distance} \le r} = \rho 2\pi R^2\left(1-\cos(2\arcsin \frac{r}{2R})\right) \hspace{0.4cm}\textrm{with } r \le 2R,
 \end{equation}
@@
 % diff(c*%pi*r^2,S),solve(S(r)=S,r);
+% (c=rho)
 \begin{align}
 f_p(S)
-= & \frac{d}{dS} c \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta}
-= \frac{c 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
+= & \frac{d}{dS} \rho \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta}
+= \frac{\rho 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
 \end{align}
-Consequently, due to non-free space propagation effects, the observed log-log histogram is predicted to have a $\frac{2}{\eta}-1$ slope. By substituting $\eta$, one finds
+Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $\frac{2}{\eta}-1$ slope. By substituting $\eta$, one finds
 \begin{equation}
@@
 Therefore, to draw a sample by using the inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$.
-By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-c \pi r_L^2$ for $S < Ig/r_L^2$.
-Consequently, by sampling $x_u$ from a uniform distribution with $0 \le x_u \le N-c \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ a power-law distribution.
+By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-\rho \pi r_L^2$ for $S < Ig/r_L^2$.
+Consequently, by sampling $x_u$ from a uniform distribution with $0 \le x_u \le N-\rho \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ a power-law distribution.
 
@@
 \subsection{Parameter variability} \label{sec:parameter-variability}
-In reality, the parameters $c$, $I$ and $g$, which are the source density per unit area, source strength and instrumental gain respectively, will not be constant over time and frequency but have a stochastic nature.
+In reality, the parameters $\rho$, $I$ and $g$, which are the source density per unit area, source strength and instrumental gain respectively, will not be constant over time and frequency but have a stochastic nature.
 However, since each specific value for these parameters produces a power law, the combined distribution will still show a power law, as long as the parameters follow a distribution that is steep at high amplitudes (in $\log$--$\log$ space), such as a Gaussian or uniform distribution.
-
-%\begin{figure}[tb]
-%\begin{center}
-%\includegraphics[width=12cm]{img/plot-passband.pdf}
-%\caption{Effect of the low and high-band frequency responses on the amplitude histogram for different distributions.}
-%\label{fig:plot-passband-histogram}
-%\end{center}
-%\end{figure}
 
@@
 If the RFI sources have an apparent power-law distribution with a certain exponent, the distribution will have a feature at low amplitudes due to the frequency response, but a power law with equal exponent will still appear when the distribution is observed over a wide enough range.
-
-%In Fig.~\ref{fig:plot-passband-histogram}, the histograms of the frequency response are given.
-To also demonstrate the effect of mixing distributions, histograms are added for samples that are drawn from a uniform, Gaussian and -2 power-law distribution as in Eq.~\eqref{eq:drawing-rfi-distribution} with $Ig/r_L^2=1$, that have been multiplied by the band-pass.
-The frequency responses were determined by measuring the standard deviation of the data over frequency after flagging the data for interference.
-As expected, it can still be determined that the intrinsic distribution is a -2 power law.
 
@@
 In
-\citet{lofar-radio-environment} % TODO correct citation if full citation available
+\citet{lofar-radio-environment}
 the false-positives rate for the AOFlagger is estimated to be 0.5\%.
@@
 We have analysed the distributions of two data sets. Both data sets are 24 h LOFAR RFI surveys and are extensively analysed in
-\citet{lofar-radio-environment}. % TODO
+\citet{lofar-radio-environment}.
 In one set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz.
@@
 
-To detect RFI, the AOFlagger \citep{post-correlation-rfi-classification, LOFAR-RFI-pipeline, scale-invariant-rank-operator} was used.
+To detect RFI, the AOFlagger \citep{LOFAR-RFI-pipeline, scale-invariant-rank-operator} was used.
 However, since this flagger is partly amplitude-based, it is likely that low-level RFI will leak through the detector.
@@
 Nevertheless, combining the data of all the sub-bands results in reasonably stable power-law distributions.
-This indicates that the individual sub-bands have an approximately similar underlying power-law distribution, that is not yet apparent because not enough samples are combined in the histograms of individual sub-bands.
+The variation could be caused by the different power law exponents that source populations at different frequencies might have.
+It could also be caused by approximately similar underlying power-law distribution that only differ in the number of transmitters.
+In that case, the underlying power law might not yet be apparent, because not enough samples are combined.
 

### 2013-01-09 articles/Article-RFI-distribution/article-rfi-distribution.tex
Some comments of Leon

@@
 Thereby, they might obstruct the detection of the weak astronomical signals of interest.
-This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest objects in the Universe.
-We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR.
+This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe.
+We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of visibility data observed with LOFAR, similar to radio source count analyses that are used in cosmology.
 The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
@@
 These observations almost reach the thermal noise and are thought not to be limited by RFI.
-Therefore, we conclude that it is likely that most RFI sources do not add coherently, in which case RFI will not prevent LOFAR from detecting the EoR.
+Therefore, we conclude that it is likely that most RFI sources do not add coherently, in which case RFI will not prevent an EoR detection with LOFAR.
 \end{abstract}
@@
 For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest.
-In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) \citep{de-bruyn-eor-ursi-2011}.
+In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, van Haarlem et al.
+2013, \textit{submitted}).
+% TODO correct reference if van Haarlem et al is published
 
@@
 We will now start by deriving a prediction of the frequency density function of ground-based interfering sources.
-Consider an interfering point source of strength $I$.
+Consider an interfering point source of strength $I$ measured in flux density.
 This source is observed by an interferometer that consists of two antennas or stations with gains $g_1,g_2$, which include all instrumental effects.
@@
 \end{equation}
-with $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$.
+with (amplitude) gains $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$.
 Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth.
@@
 
-Now, we can treat the interferometer geometrically as a single point, as both antennas will see the same distribution.
-Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and interferometric gain $g$ as
+Although two antennas do not necessarily observe the same RFI sources, for source count analysis we can treat the interferometer geometrically as a single point, as both antennas will see the same distribution.
+Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and interferometric gain $g = g_1 g_2$ as
 \begin{equation} \label{eq:amplitude-fall-off-in-free-space}
@@
 \begin{align} \label{eq:cumulative-rfi-distribution}
- F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{S}^{-1}(r))\\
+ F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{R}(S))\\
 \notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{S})\\
@@
 \end{align}
-where $\mathcal{S}^{-1}(r)$ is the inverse of $\mathcal{S}(r)$, i.e., the function that returns the distance $r$ for a given amplitude $S$.
+where $\mathcal{R}(S)=\mathcal{S}^{-1}$, the inverse of $\mathcal{S}$, i.e., the function that returns the distance $r$ for a given amplitude $S$.
 Finally, the differential frequency density can be calculated by taking the derivative,
@@
 Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we predict to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane.
+For this, we have assumed that the variables $I$, $g$ and $\rho$ are constants.
+In reality, these variables have a stochastic nature.
+The effect of this will be discussed in \S\ref{sec:parameter-variability}.
 
@@
 \end{equation}
-with $R$ the radius of the Earth.
+with $R$ the radius of the Earth and $r$ the Euclidean distance between the source and the antenna.
 Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated.
@@
 Because of the irregular surface of the Earth (including urban areas) and the absence of a direct line of sight between transmitter and receiver, the propagation mode might be indirect.
-Multiple indirect paths might be formed by reflection, refraction or diffraction of the electromagnetic wave.
+Multiple indirect paths might be formed that include reflection, refraction, diffraction and absorption of the electromagnetic wave.
 
@@
 For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong sources.
-Noise, however, will have a contribution.
-The real and imaginary components of receiver noise are independent and identically Gaussian distributed.
-Consequently, an amplitude $x$ will be Rayleigh distributed:
+For example, the strongest apparent source in the NCP EoR field is around 5 Jy (Yattawata et al.
+2013, \textit{submitted}).
+% TODO fill in details.
+The standard deviation of the noise, however, is around 100~Jy on highest LOFAR resolutions, and will have a larger contribution.
+
+The real and imaginary components of receiver noise are independent and identically Gaussian distributed with zero mean and equal variance.
+Consequently, an amplitude $x$ will be Rayleigh distributed \citep[\S6-2]{papoulis-stochastic-processes}:
 \begin{equation}\label{eq:rayleigh-formula}
@@
 By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-\rho \pi r_L^2$ for $S < Ig/r_L^2$.
-Consequently, by sampling $x_u$ from a uniform distribution with $0 \le x_u \le N-\rho \pi r_L^2$ and applying the inverse of Eq.~\eqref{eq:cumulative-rfi-distribution}, we will sample amplitudes $S$ with $S \ge Ig/r_L^2$, and $S\sim$ a power-law distribution.
+Consequently, if $x_u$ is sampled from a uniform distribution with $0 \le x_u \le N-\rho \pi r_L^2$, and $x_u$ is the input to the inverse of the cumulative source count function in Eq.~\eqref{eq:cumulative-rfi-distribution}, the result is an amplitude sample $S$ with $S \ge Ig/r_L^2$, and $S\sim$ a power-law distribution.
 
@@
 \section{Data description} \label{sec:dist-data}
-We have analysed the distributions of two data sets. Both data sets are 24 h LOFAR RFI surveys and are extensively analysed in
-\citet{lofar-radio-environment}.
+We have analysed the distributions of two data sets. Both data sets are 24-h LOFAR RFI surveys and are extensively analysed in
+\citet{lofar-radio-environment}. We refer to van Haarlem et al. (2013, \textit{submitted}) for a full description of the capabilities of LOFAR.
+% TODO correct reference if van Haarlem et al is published
 In one set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz.
@@
 
+% TODO appendix with calculation of 2D case == 3D case?
+
 \bibliographystyle{mn2e}

### 2013-01-15 articles/Article-RFI-distribution/article-rfi-distribution.tex
Correcting Iliev''s name
Some fixes based on comments by internal ref

@@
 B.~Ciardi$^4$,
-I.~T.~Iliav$^5$,
+I.~T.~Iliev$^5$,
 \newauthor
@@
 This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe.
-We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of visibility data observed with LOFAR, similar to radio source count analyses that are used in cosmology.
+We explore the low-frequency RFI situation by studying brightness histograms of visibility data observed with LOFAR, similar to radio source count analyses that are used in cosmology.
 The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
@@
 Typical radio observations record a few hours of data, and the results are integrated.
-In these cases, it often suffices to only excise interference that is apparent and thus above the noise, to reach the thermal noise limit of the instrument.
+In these cases, excising only the interference that is apparent and thus above the noise often suffices, i.e., the observation can still reach the thermal noise limit of the instrument.
 A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated.
@@
 Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}.
+Ground-based Cosmic Microwave Background (CMB) experiments are another class of experiments involving long integration times (e.g., \citealt{ska-cmb-subrahmanyan}).
 For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest.
@@
 This work explores the information that is present in interference distributions, in order to analyse possible low-level interference that is not detectable by standard detection methods.
-Our approach is similar to the radio source count analysis that is used in cosmology \citep{vonhoerner-radio-source-counts}, also named $\log N$ -- $\log S$ analysis, where $N$ and $S$ refer to the celestial source count and brightness respectively.
+Our approach is similar to the radio source count analysis that is used in cosmology \citep{condon-cosm-evol-of-radio-sources}, also named $\log N$ -- $\log S$ analysis, where $N$ and $S$ refer to the celestial source count and brightness respectively.
 The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe.
@@
 This --- as well as many other effects --- will of course influence the distribution.
-First we will derive the expected intrinsic source distribution.
-After that, we will analyse the issues that arise when measuring the distribution by counting samples.
+If sources overlap in the time-frequency space, the situation is somewhat similar to the case where multiple unresolved celestial radio sources in the reception pattern of a telescope only allow observation of a sum of sources.
+However, in that case it is still possible to validate radio source models by comparing $\log N$ -- $\log S$ histograms \citep{scheuer-analysing-faint-stars}.
+
+First we will derive the expected intrinsic source distribution for interfering radio sources. After that, we will analyse the issues that arise when measuring the distribution by counting samples.
 
@@
 
-\subsection{Spherical case}
+\subsection{Spherical case} \label{sec:spherical-geometry}
 If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes.
@@
 with $R$ the radius of the Earth and $r$ the Euclidean distance between the source and the antenna.
-Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated.
+As shown in Appendix~\ref{sec:proof-spherical-case}, Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated.
 Therefore, the spherical and 2-dimensional cases lead to a distribution with equal power law.
@@
 
+We assume that the stochastic nature of $I$, that is caused by the spread of transmitters with different intrinsic strengths, has little effect on the $\log N$ -- $\log S$ histograms.
+Nevertheless, it is unlikely that $I$ follows a uniform or Gaussian distribution, because the distribution will contain few strong transmitters (such as radio stations) and many weak transmitters (such as remote controls).
+Therefore, $I$ might follow a power law by itself.
+Strong transmitters are probably on average transmitting longer, thus will also contaminate more samples.
+This weakens the effect of $I$ on the histogram somewhat.
+%TODO
+
 One instrumental effect that is absorbed in $g$ is the frequency response of the instrument, i.e., the antenna response in combination with the band-pass of the analogue and digital filters.
@@
 For Fig.~\ref{fig:explanation-constraints} this means that the areas labelled $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal.
-The area labelled $N_b$ is assumed to be zero for now, which assumes the power law has a sharp cut-off on the left side (e.g., because of the curvature of the Earth).
-Solving this equality, one finds that
+The area labelled $N_b$ is assumed to be zero for now, which assumes the power law has a sharp cut-off on the left side, e.g., because of the curvature of the Earth.
+Solving $N_a = N_\textrm{total}-N_\textrm{part}$ results in
 \begin{equation} \label{eq:lower-limit-1}
@@
 
-% TODO appendix with calculation of 2D case == 3D case?
+\appendix
+\section{Calculations with spherical geometry} \label{sec:proof-spherical-case}
+As stated in \S\ref{sec:spherical-geometry}, the distance distribution of sources for a flat geometry and a spherical surface are equal for distances smaller than the diameter of the Earth.
+We proof this by showing that the spherical distance function $\tilde{F}_{\textrm{distance} \le r}(r)$ of Eq.~\ref{eq:cumulative-spherical} equals the flat distance function $F_{\textrm{distance} \le r}(r)$ of Eq.~\ref{eq:cumulative-distance-function}.
+\begin{proof}
+\begin{eqnarray} \notag
+\tilde{F}_{\textrm{distance} \le r}(r) & = & \rho 2 \pi R^2 \left[1  - \cos\left(2 \arcsin \frac{r}{2R}\right)\right] \\\notag
+& & \textit{\{\hspace{1mm}Using $\cos 2\phi = 1 - 2\sin^2 \phi$\hspace{1mm}\}} \\\notag
+& = & \rho 2 \pi R^2 \left[1 - \left(1 - 2 \sin^2(\arcsin \frac{r}{2R})\right)\right] \\\notag
+& = & \rho 2 \pi R^2 \left[2 \sin^2(\arcsin \frac{r}{2R})\right] \\\notag
+& & \textit{\{\hspace{1mm}Assuming $r \le 2R$\hspace{1mm}\}} \\\notag
+& = & \rho 2 \pi R^2 \left[2 \left(\frac{r}{2R}\right)^2\right] \\\notag
+& = & \rho \pi r^2 \\
+& = & F_{\textrm{distance} \le r}(r)
+\end{eqnarray}
+\end{proof}
 

### 2013-01-16 articles/Article-RFI-distribution/article-rfi-distribution.tex
A days work...

@@
 % (c=rho)
-\begin{align}
+\begin{align} \label{eq:frequency-density-with-propagation}
 f_p(S)
-= & \frac{d}{dS} \rho \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta}
+= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta} \right]
 = \frac{\rho 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
@@
 Even more, we still need to take into account that the samples classified as RFI are also affected by noise.
-Although this effect is moderate, because 'most' RFI samples are of much higher amplitude than what is added because of the noise, we are interested in the low-level RFI samples, and the relative effect of noise on those samples is large.
+Although this effect is moderate, because `most' RFI samples are of much higher amplitude than what is added because of the noise, we are interested in the low-level RFI samples, and the relative effect of noise on those samples is large.
 
@@
 The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes.
-A sample can be drawn from the RFI distribution by normalizing and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable.
+A sample can be drawn from the RFI distribution by scaling and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable.
 Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity.
-Therefore, to draw a sample by using the inversion method, this would require a uniformly distributed variable $\sim U(0,\infty)$.
-By assuming that there are no samples beyond some limiting distance $r_L$, then $F_{\textrm{amplitude}\le S}(S) = N-\rho \pi r_L^2$ for $S < Ig/r_L^2$.
-Consequently, if $x_u$ is sampled from a uniform distribution with $0 \le x_u \le N-\rho \pi r_L^2$, and $x_u$ is the input to the inverse of the cumulative source count function in Eq.~\eqref{eq:cumulative-rfi-distribution}, the result is an amplitude sample $S$ with $S \ge Ig/r_L^2$, and $S\sim$ a power-law distribution.
-
-By normalizing $x_u$, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
+Therefore, one needs to assume that there are no samples beyond some limiting distance $r_L$ to invert the cumulative function.
+With this assumption, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
 \begin{equation} \label{eq:drawing-rfi-distribution}
-\textrm{real}(S_{\textrm{RFI}})\leftarrow \frac{Ig}{x_u r_L^2}\textrm{,\hspace{1mm} imag}(S_{\textrm{RFI}}) \leftarrow \frac{Ig}{y_u r_L^2},
+S_{\textrm{RFI}} \leftarrow \frac{Ig}{x_u r_L^2} e^{i 2 \pi y_u},
 \end{equation}
-with $x_u, y_u \sim U(0, 1)$, and the corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
+with $x_u, y_u \sim U(0, 1)$.
+The result is an amplitude sample $S$, with $S \ge Ig/r_L^2$, that follows a power-law distribution.
+The corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
 \begin{equation} \label{eq:drawing-combined-distribution}
-\textrm{real}(S) \leftarrow x_n + \frac{Ig}{x_u r_L^2}
-\textrm{,\hspace{1mm} imag}(S) \leftarrow y_n + \frac{Ig}{y_u r_L^2},
+\textrm{real}(S) \leftarrow x_n + \frac{Ig \cos 2 \pi y_u}{x_u r_L^2}
+\textrm{,\hspace{1mm} imag}(S) \leftarrow y_n + \frac{Ig \sin 2 \pi y_u}{x_u r_L^2},
 \end{equation}
-with $x_n \sim N(\mu=0; \sigma=1)$. The final amplitude sample $S_A$ can be calculated with 
+with $x_n \sim N(\mu=0; \sigma)$. The final amplitude sample $S_A$ can be calculated with 
 \begin{equation} \label{eq:drawing-amplitudes}
@@
 
-% Exponential:
-% fs(x):=c*%pi*I*g/x^2;
-%
-% Gaussian:
-% fn(x):=1/(sqrt(2*%pi)*sigma)*exp(-x^2/(2*sigma^2));
-%
-% Convolution:
-% fc(x):=romberg(fs(x)*fn(x+tau), tau, -inf, inf);
-%
-% plot2d([log(fc(x)),log(fs(x)),log(fn(x))], [x,0.01, 100], [y,-10,10], [logx]), sigma=1, I=1, g=1, c=1;
+For completeness, we also show the formula for sampling apparent source strengths that include surface propagation effects, which can be used for sampling realistic source amplitudes in simulations.
+This is calculated by inversion and scaling of the frequency density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
+\begin{equation}
+\textrm{real}(S_{\textrm{RFI}}) \leftarrow \frac{\zeta I g}{x_u^{\eta/2} r_L^\eta},
+\textrm{imag}(S_{\textrm{RFI}}) \leftarrow \frac{\zeta I g}{x_u^{\eta/2} r_L^\eta}.
+\end{equation}
 
@@
 However, since each specific value for these parameters produces a power law, the combined distribution will still show a power law, as long as the parameters follow a distribution that is steep at high amplitudes (in $\log$--$\log$ space), such as a Gaussian or uniform distribution.
-
-We assume that the stochastic nature of $I$, that is caused by the spread of transmitters with different intrinsic strengths, has little effect on the $\log N$ -- $\log S$ histograms.
-Nevertheless, it is unlikely that $I$ follows a uniform or Gaussian distribution, because the distribution will contain few strong transmitters (such as radio stations) and many weak transmitters (such as remote controls).
-Therefore, $I$ might follow a power law by itself.
-Strong transmitters are probably on average transmitting longer, thus will also contaminate more samples.
-This weakens the effect of $I$ on the histogram somewhat.
-%TODO
 
@@
 Because most sources will be observed with similar gains (within one order of magnitude), it can be expected that the beam will have a limited effect on the histogram properties of an observation, and is comparable with the effect of the frequency response.
+
+We assume that the stochastic nature of $I$, that is caused by the spread of transmitters with different intrinsic strengths, also has little effect on the $\log N$ -- $\log S$ histograms.
+Nevertheless, it is unlikely that $I$ follows a uniform or Gaussian distribution, because the distribution will contain few strong transmitters (such as radio stations) and many weak transmitters (such as remote controls).
+Therefore, $I$ might follow a power law by itself.
+On average, strong transmitters transmit longer, thus will also contaminate more samples.
+High-power transmitters, such as radio stations, have a typical transmitting power in the order of 100 kW.
+Low-power transmitters, such as remote controls, transmit with an order of 100 mW.
+Therefore, these devices have a spread of around 6 orders of magnitudes in power.
+As long as the exponent of this distribution is less steep (i.e., less negative) compared to the power law caused by the spatial distribution, the spatial distribution will dominate the histogram at high amplitudes.
+With a spatial $-1.5$ power law and the given transmitting powers, the low-power transmitters should contaminate a factor of $10^9$ more samples compared to the high-power transmitters to dominate the high-amplitude distribution, which is unlikely.
 
@@
 Nevertheless, we will additionally calculate a maximum likelihood estimator for comparison.
-The maximum-likelihood estimator for the exponent in a power-law distribution is given by the Hill estimator \citep{hill-estimator, power-law-distribution}, defined as:
+The maximum-likelihood estimator for the exponent in a power-law distribution is given by the Hill estimator $\hat \alpha_H$ \citep{hill-estimator, power-law-distribution}, defined as:
 \begin{equation} \label{eq:hill}
- \alpha_H = 1 + N \left(\sum\limits_{i=1}^{N} \ln \frac{x_i}{x_\textrm{max}} \right)^{-1}.
+ \hat \alpha_H = 1 + N \left(\sum\limits_{i=1}^{N} \ln \frac{x_i}{x_\textrm{min}} \right)^{-1},
 \end{equation}
+with $N$ the number of samples and $x_i$ for $0 < i \le N$ the samples drawn from a power law with lower bound $x_\textrm{min}$.
 However, this estimator assumes the distribution is not cut-off at a high point.

### 2013-01-17 articles/Article-RFI-distribution/article-rfi-distribution.tex
Fixed several formulae and fixed all ref I comments

@@
 with
-\begin{equation}
+\begin{equation} \label{eq:zeta-definition}
  \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
@@
 \begin{equation} \label{eq:drawing-combined-distribution}
-\textrm{real}(S) \leftarrow x_n + \frac{Ig \cos 2 \pi y_u}{x_u r_L^2}
-\textrm{,\hspace{1mm} imag}(S) \leftarrow y_n + \frac{Ig \sin 2 \pi y_u}{x_u r_L^2},
+S \leftarrow v_n + w_n i + \frac{Ig}{x_u r_L^2} e^{i 2 \pi y_u},
 \end{equation}
-with $x_n \sim N(\mu=0; \sigma)$. The final amplitude sample $S_A$ can be calculated with 
+with $v_n, w_n \sim N(\mu=0; \sigma)$. The final amplitude sample $S_A$ can be calculated with 
 \begin{equation} \label{eq:drawing-amplitudes}
-S_A \leftarrow |S| = \sqrt{\textrm{real}({S})^2+\textrm{imag}({S})^2}.
+S_A \leftarrow |S| = \sqrt{\textrm{real}^2({S})+\textrm{imag}^2({S})}.
 \end{equation}
@@
 
-For completeness, we also show the formula for sampling apparent source strengths that include surface propagation effects, which can be used for sampling realistic source amplitudes in simulations.
-This is calculated by inversion and scaling of the frequency density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
-\begin{equation}
-\textrm{real}(S_{\textrm{RFI}}) \leftarrow \frac{\zeta I g}{x_u^{\eta/2} r_L^\eta},
-\textrm{imag}(S_{\textrm{RFI}}) \leftarrow \frac{\zeta I g}{x_u^{\eta/2} r_L^\eta}.
+For completeness we show the formula for sampling apparent source strengths that include surface propagation effects, which can be used for sampling realistic source amplitudes in simulations.
+This can be derived by integration, scaling and inversion of the frequency density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
+\begin{equation} \label{eq:sample-with-propagation}
+\tilde{S}_{\textrm{RFI}} \leftarrow \frac{\zeta I g}{x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
 \end{equation}
+Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 \le x_u,y_u < 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.
 
@@
 Nevertheless, it is unlikely that $I$ follows a uniform or Gaussian distribution, because the distribution will contain few strong transmitters (such as radio stations) and many weak transmitters (such as remote controls).
-Therefore, $I$ might follow a power law by itself.
-On average, strong transmitters transmit longer, thus will also contaminate more samples.
+Therefore, variable $I$ might follow a power law by itself.
+It is likely that strong transmitters transmit more on average, and therefore contaminate more samples as well.
 High-power transmitters, such as radio stations, have a typical transmitting power in the order of 100 kW.
@@
 Therefore, these devices have a spread of around 6 orders of magnitudes in power.
-As long as the exponent of this distribution is less steep (i.e., less negative) compared to the power law caused by the spatial distribution, the spatial distribution will dominate the histogram at high amplitudes.
+As long as the exponent of the power law of $I$ is less steep (i.e., less negative) compared to the power law caused by the spatial distribution, the spatial distribution will dominate the histogram at high amplitudes.
 With a spatial $-1.5$ power law and the given transmitting powers, the low-power transmitters should contaminate a factor of $10^9$ more samples compared to the high-power transmitters to dominate the high-amplitude distribution, which is unlikely.
+Therefore, it is likely that the spatial distribution will dominate the power law in the histogram.
 
@@
 \end{equation}
-with $N$ the number of samples and $x_i$ for $0 < i \le N$ the samples drawn from a power law with lower bound $x_\textrm{min}$.
+with $N$ the number of samples and $x_i$ for $0 < i \le N$ the samples that follow a power law with lower bound $x_\textrm{min}$.
 However, this estimator assumes the distribution is not cut-off at a high point.
@@
 Moreover, the estimator requires reading and processing all data points, and because of the size of the data sets this is somewhat impractical.
-However, we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ the bin frequency.
+However, we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ is the bin frequency.
 Because our bin size is small and the distribution is well sampled, the resulting estimate will be close to the normal Hill estimator.
@@
 With the construction of empirical models for the RFI source distributions, we are one step closer to including these effects in the simulation.
-Using Eq.~\eqref{eq:drawing-rfi-distribution}, one can sample a realistic strength of a single RFI source, add the feature to the data and run the AOFlagger.
+Using Eq.~\eqref{eq:sample-with-propagation}, one can sample a realistic strength of a single RFI source, add the feature to the data and run the AOFlagger.
 What is still needed for accurate simulation, is to obtain a likely distribution for the duration that one such source affects the data.
@@
 There are several reasons for this, which we will summarize one by one.
-\begin{itemize}
- \item Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes.
+
+There are several effects related to the origin of the interference, which can reduce their effect.
+One such effect, is when the RFI sources have a variable geometric phase.
+Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes.
 This would cause them to sum (partly) incoherently over time, and thus go approximately down with the noise.
- \item Many RFI sources will be seen by only a few stations. This would make the histogram go down more quickly at lower amplitudes.
- \item For the shortest baselines at 150~MHz, the far field starts around 1~km.
+Moreover, many RFI sources will be seen by only a few stations or are not constant over time, which further attenuates their effect.
+Finally, for the shortest baselines at 150~MHz, the far field starts around 1~km.
 Some RFI sources will therefore be in the near field, especially for the longer baselines.
-In that case, a source will not add up coherently over the interferometers that see the particular source, as the interferometers see them with different phases.
- \item Many stationary sources are not constant over time, thus will be attenuated somewhat by averaging.
- \item We have assumed 100\% of the spectrum is occupied by RFI.
-If, say, only 10\% of the spectrum is occupied, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all.
-With current data, one can only speculate how much of the electromagnetic spectrum is truly occupied.
- \item Stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other.
+In that case, a source will not add up coherently over the interferometers that see the particular source, as the interferometers see it with different phases.
+
+A large number of RFI sources together can also behave like normal incoherent noise.
+Stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other.
 Individually, the sources will have fixed geometric phases, but because the baselines are much longer than the wavelength, their phases will become uniformly spread.
 Therefore, they will add incoherently.
-\end{itemize}
 
-Two other effects can improve the RFI situation as well:
-\begin{itemize}
- \item Fringe stopping interferometers can partly average out RFI sources.
+Sometimes, standard data processing also attenuates RFI, mainly because of fringe stopping and the imaging of data.
+Fringe stopping interferometers can partly average out RFI sources.
 Nevertheless, stationary RFI that is averaged out by fringe stopping will leave artefacts in the field centre \citep{post-correlation-filtering}.
 This is not relevant when observing the North Celestial Pole --- which is one of the LOFAR EoR fields --- because no fringe stopping is applied when observing the NCP.
- \item The Fourier transform that is involved in data imaging will localize the contribution from stationary RFI near the NCP.
+Imaging sometimes attenuates RFI, because the Fourier transform that is involved in data imaging will localize the contribution from stationary RFI near the NCP.
 If RFI artefacts would show in the image of the NCP field, they can easily be detected and possibly be removed, or processing could ignore data near the pole.
-\end{itemize}
 
@@
 Preliminary analysis of EoR NCP observations of a single night have reached the thermal noise (Yatawatta et al., \textit{submitted}), but do not show leaked RFI at the pole.
+
+On top of all these possible RFI reducing effects, our given RFI constraints could be too pessimistic because we have assumed 100\% of the spectrum is occupied by RFI.
+If, say, only 10\% of the spectrum is occupied, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all.
+With current data, one can only speculate how much of the electromagnetic spectrum is truly occupied.
 

### 2013-01-17 articles/Article-RFI-distribution/article-rfi-distribution.tex
Slight rewording

@@
 Nevertheless, it is unlikely that $I$ follows a uniform or Gaussian distribution, because the distribution will contain few strong transmitters (such as radio stations) and many weak transmitters (such as remote controls).
-Therefore, variable $I$ might follow a power law by itself.
+Therefore, variable $I$ might follow a power-law distribution by itself.
 It is likely that strong transmitters transmit more on average, and therefore contaminate more samples as well.
 High-power transmitters, such as radio stations, have a typical transmitting power in the order of 100 kW.
-Low-power transmitters, such as remote controls, transmit with an order of 100 mW.
+Low-power transmitters, such as remote controls, transmit with an order of 100 mW or even less.
 Therefore, these devices have a spread of around 6 orders of magnitudes in power.

### 2013-01-18 articles/Article-RFI-distribution/article-rfi-distribution.tex
Textuals

@@
 For example, the strongest apparent source in the NCP EoR field is around 5 Jy (Yattawata et al.
-2013, \textit{submitted}).
+2013, \textit{in press}).
 % TODO fill in details.
-The standard deviation of the noise, however, is around 100~Jy on highest LOFAR resolutions, and will have a larger contribution.
+The standard deviation of the noise, however, is around 100~Jy on highest LOFAR resolutions, and will have a larger contribution on the histogram.
 
@@
 To understand the distribution of samples consisting of RFI + noise, one needs to consider the individual contributions of the real and imaginary components.
-The real and imaginary components of samples that are affected by RFI, will follow a bivariate distribution: each complex component is the sum of a complex component of RFI and a noise sample.
+The real and imaginary components of samples that are affected by RFI, will follow a multi-variate distribution: each complex component is the sum of a complex component of RFI and a noise sample.
 The noise comes from a Gaussian distribution with zero mean.
@@
 To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial.
-The histogram can be numerically estimated by drawing complex real and imaginary samples from the two distributions and calculating and counting the amplitudes.
+The histogram can be numerically estimated by drawing complex samples from the two distributions and calculating and counting the amplitudes.
 A sample can be drawn from the RFI distribution by scaling and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable.
@@
 with $x_u, y_u \sim U(0, 1)$.
-The result is an amplitude sample $S$, with $S \ge Ig/r_L^2$, that follows a power-law distribution.
+The result is an amplitude sample $S_{\textrm{RFI}}$, with $S_{\textrm{RFI}} \ge Ig/r_L^2$, that follows a power-law distribution.
 The corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
@@
 
-For completeness we show the formula for sampling apparent source strengths that include surface propagation effects, which can be used for sampling realistic source amplitudes in simulations.
+For completeness we show the formula for sampling apparent source strengths that include propagation effects according to Hata's model, which can be used for sampling realistic source amplitudes in simulations.
 This can be derived by integration, scaling and inversion of the frequency density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
@@
 Finally, $E(S_\textrm{leak})$, which is the expected value of leaked RFI not detected by the flagger, can be calculated by replacing $S_U$ with $S_d$ in the numerator of Eq.~\eqref{eq:expected-value-rfi} and subtracting the removed number of samples from the total of number of samples.
-Assume that a fraction of $\eta$ samples are not detected as RFI and $1-\eta$ have been detected as RFI, then
+Assume that a fraction of $\kappa$ samples are not detected as RFI and $1-\kappa$ have been detected as RFI, then
 \begin{align} \label{eq:expected-value-leaked-rfi}
-E(S_\textrm{leak}) & = \frac{1}{\eta N_{LU}}\int\limits_{S_L}^{S_d} \beta S^\alpha S dS
-= \frac{\left( S_d^{\alpha+2} - S_L^{\alpha+2} \right) \left( \alpha+1 \right) } {\eta \left( S_U^{\alpha+1} - S_L^{\alpha+1}\right) \left( \alpha+2 \right)}.
+E(S_\textrm{leak}) & = \frac{1}{\kappa N_{LU}}\int\limits_{S_L}^{S_d} \beta S^\alpha S dS
+= \frac{\left( S_d^{\alpha+2} - S_L^{\alpha+2} \right) \left( \alpha+1 \right) } {\kappa \left( S_U^{\alpha+1} - S_L^{\alpha+1}\right) \left( \alpha+2 \right)}.
 \end{align}
-This is the average contribution that leaked RFI will have on a single sample. It has the same units as the parameters $S_L$, $S_U$ and $S_d$. Typical values for $\eta$ are $0.95$--$0.99$.
+This is the average contribution that leaked RFI will have on a single sample. It has the same units as the parameters $S_L$, $S_U$ and $S_d$. Typical values for $\kappa$ are $0.95$--$0.99$.
 
@@
 Fig.~\ref{fig:Histograms-raw} shows the histograms with logarithmic axes for the LBA and HBA sets.
-In both sets, it is clear that at least one variate with a Rayleigh and one with a power-law distribution have been observed.
+In both sets, it is clear that at least one component with a Rayleigh and one with a power-law distribution have been observed.
 The left part of the histogram matches the Rayleigh distribution well up to the mode of the distribution.
@@
 Nevertheless, combining the data of all the sub-bands results in reasonably stable power-law distributions.
-The variation could be caused by the different power law exponents that source populations at different frequencies might have.
-It could also be caused by approximately similar underlying power-law distribution that only differ in the number of transmitters.
-In that case, the underlying power law might not yet be apparent, because not enough samples are combined.
+The variation could be caused by the different power-law exponents that source populations at different frequencies might have.
+It could also be caused by a differing number of transmitters.
+In that case, the underlying power law might not always be apparent, because not enough samples are combined.
 
@@
 For the EoR project with $1$~MHz resolution, Eq.~\ref{eq:rfi-noise-level} results in $\sigma_\textrm{RFI}\approx 4.7$ mJy.
-However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz subband (Yatawatta et al., \textit{submitted}), and the resulting images show no signs of RFI.
+However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz subband (Yatawatta et al.
+2013, \textit{in press}), and the resulting images show no signs of RFI.
 This implies that either the upper limit is far from the actual RFI situation, or Eq.~\ref{eq:rfi-noise-level} is not applicable to most of the RFI that is observed with LOFAR.
@@
 
-Sometimes, standard data processing also attenuates RFI, mainly because of fringe stopping and the imaging of data.
+Some standard data processing actions can also attenuate RFI, especially fringe stopping and the imaging of data.
 Fringe stopping interferometers can partly average out RFI sources.
@@
 It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright foreground sources.
-Preliminary analysis of EoR NCP observations of a single night have reached the thermal noise (Yatawatta et al., \textit{submitted}), but do not show leaked RFI at the pole.
+Preliminary analysis of EoR NCP observations of a single night have reached the thermal noise (Yatawatta et al.
+2013, \textit{in press}), but do not show leaked RFI at the pole.
 

### 2013-01-18 articles/Article-RFI-distribution/article-rfi-distribution.tex
Removed appendix, rewrote spherical derivation

@@
 In this case, the RFI sources are distributed over the surface of a sphere.
-If we assume their difference in height is negligible, and by using the formula for the surface area of a hemisphere and some basic geometry, one finds that
-\begin{equation}\label{eq:cumulative-spherical}
-\tilde{F}_{\textrm{distance} \le r} = \rho 2\pi R^2\left(1-\cos(2\arcsin \frac{r}{2R})\right) \hspace{0.4cm}\textrm{with } r \le 2R,
+If we assume their difference in height is negligible, one can use the formula for the surface area of a spherical cap,
+\begin{equation}\label{eq:spherical-cap-area}
+ A_\textrm{cap} = \pi \left( a^2 + h^2 \right) \hspace{0.4cm}\textrm{with } a \le R \wedge h \le 2R,
 \end{equation}
-with $R$ the radius of the Earth and $r$ the Euclidean distance between the source and the antenna.
-As shown in Appendix~\ref{sec:proof-spherical-case}, Eqs.~\eqref{eq:cumulative-distance-function} and \eqref{eq:cumulative-spherical} have the same value over the range they can be evaluated.
-Therefore, the spherical and 2-dimensional cases lead to a distribution with equal power law.
+with $R$ the radius of the Earth, $a$ the radius of the base of the cap and $h$ the height of the cap.
+Because the Euclidean distance $r$ between the source and the antenna is $r^2 = a^2 + h^2$, the resulting number of sources within Euclidean distance $r$ becomes
+\begin{equation} \label{eq:cumulative-spherical}
+\tilde{F}_{\textrm{distance} \le r}(r) = \rho A_\textrm{cap} = \rho \pi r^2 \hspace{0.4cm}\textrm{with } r \le 2R,
+\end{equation}
+which equals the formule for a planar source distribution given in Eq.~\eqref{eq:cumulative-distance-function}.
+Therefore, the spherical and two-dimensional cases lead to a distribution with equal power law.
 Evaluating Eq.~\eqref{eq:cumulative-spherical} is limited to $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth.
@@
 
-\appendix
-\section{Calculations with spherical geometry} \label{sec:proof-spherical-case}
-As stated in \S\ref{sec:spherical-geometry}, the distance distribution of sources for a flat geometry and a spherical surface are equal for distances smaller than the diameter of the Earth.
-We proof this by showing that the spherical distance function $\tilde{F}_{\textrm{distance} \le r}(r)$ of Eq.~\ref{eq:cumulative-spherical} equals the flat distance function $F_{\textrm{distance} \le r}(r)$ of Eq.~\ref{eq:cumulative-distance-function}.
-\begin{proof}
-\begin{eqnarray} \notag
-\tilde{F}_{\textrm{distance} \le r}(r) & = & \rho 2 \pi R^2 \left[1  - \cos\left(2 \arcsin \frac{r}{2R}\right)\right] \\\notag
-& & \textit{\{\hspace{1mm}Using $\cos 2\phi = 1 - 2\sin^2 \phi$\hspace{1mm}\}} \\\notag
-& = & \rho 2 \pi R^2 \left[1 - \left(1 - 2 \sin^2(\arcsin \frac{r}{2R})\right)\right] \\\notag
-& = & \rho 2 \pi R^2 \left[2 \sin^2(\arcsin \frac{r}{2R})\right] \\\notag
-& & \textit{\{\hspace{1mm}Assuming $r \le 2R$\hspace{1mm}\}} \\\notag
-& = & \rho 2 \pi R^2 \left[2 \left(\frac{r}{2R}\right)^2\right] \\\notag
-& = & \rho \pi r^2 \\
-& = & F_{\textrm{distance} \le r}(r)
-\end{eqnarray}
-\end{proof}
-
 \bibliographystyle{mn2e}

### 2013-01-18 articles/Article-RFI-distribution/article-rfi-distribution.tex
Corrected big sign mistake\!

@@
 \begin{abstract}
-Faint undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are stationary and coherently present over time.
+Faint undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are stationary and consistently present over time.
 Thereby, they might obstruct the detection of the weak astronomical signals of interest.
@@
 With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models.
-Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present.
+Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources add up consistently.
 These observations almost reach the thermal noise and are thought not to be limited by RFI.
-Therefore, we conclude that it is likely that most RFI sources do not add coherently, in which case RFI will not prevent an EoR detection with LOFAR.
+Therefore, we conclude that it is likely that most RFI sources do not linearly increase the noise, in which case RFI will not prevent an EoR detection with LOFAR.
 \end{abstract}
@@
 Many techniques have been designed to mitigate this effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, LOFAR-RFI-pipeline}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.
-In most cases, these methods excise enough of the interference to allow calibration and imaging of the data.
+In most cases, and with the help of carefull spectrum allocation management and radio-quiet zones, these methods excise enough of the interference to allow calibration and imaging of the data.
 
@@
 The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe.
-We analyse such a double-logarithmic plot for the case of terrestrial sources, with the ultimate goal of predicting their full spatial and brightness distributions.
+We analyse such a double-logarithmic plot for the case of terrestrial sources, with the ultimate goal of estimating their full spatial and brightness distributions.
 This results in a better insight into the effects of low-level interference and allows one to simulate the effects of interference more accurately.
@@
 
-This paper is organised as follows: in Sect.~\ref{sec:distribution-prediction}, we predict the terrestrial interfering source distribution based on various assumptions.
+This paper is organised as follows: in Sect.~\ref{sec:distribution-prediction}, we calculate a model for terrestrial interfering source distribution based on various assumptions.
 Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse the histograms.
@@
 
-\section{Predicting the brightness distribution} \label{sec:distribution-prediction}
+\section{Modelling the brightness distribution} \label{sec:distribution-prediction}
 Interference is generated by many different kinds of transmitters, and these will have different spatial and brightness distributions (``spatial'' refers here to the distribution on the Earth).
@@
 We will refer to this quantity with the statistical term ``frequency density'' (not to be confused with the physical radiation frequency).
-We will now start by deriving a prediction of the frequency density function of ground-based interfering sources.
+We will now start by estimating the frequency density function of ground-based interfering sources.
 Consider an interfering point source of strength $I$ measured in flux density.
@@
 
-We need the reverse of Eq.~\eqref{eq:cumulative-distance-function} to predict the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance.
+We need the reverse of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance.
 Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2$, with $N$ the total number of sources.
+Because $r$ is restricted, $N$ is finite.
 
@@
 
-Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we predict to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane.
+Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we expect to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane.
 For this, we have assumed that the variables $I$, $g$ and $\rho$ are constants.
@@
 In this case, the RFI sources are distributed over the surface of a sphere.
-If we assume their difference in height is negligible, one can use the formula for the surface area of a spherical cap,
+If we assume their difference in height is negligible, we can use the formula for the surface area of a spherical cap,
 \begin{equation}\label{eq:spherical-cap-area}
@@
 \end{equation}
-with $\eta = 0.655 \log_{10} h_b - 4.49$, one finds the frequency density function $f_p$ that considers propagation effects,
+with $\eta = 4.49 - 0.655 \log_{10} h_b$, one finds the frequency density function $f_p$ that considers propagation effects,
 % S(r):=I*g/(r^alpha);
 % assume(g>0,I>0,S>0);
-% diff(c*%pi*r^2,S),solve(S(r)=S,r);
-% (c=rho)
+% diff(rho*%pi*r^2,S),solve(S(r)=S,r);
 \begin{align} \label{eq:frequency-density-with-propagation}
@@
 \end{align}
-Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $\frac{2}{\eta}-1$ slope. By substituting $\eta$, one finds
+Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
 \begin{equation}

### 2013-01-18 articles/Article-RFI-distribution/article-rfi-distribution.tex
Dealt with comments from ref II

@@
 \begin{equation}
- S(r) = \frac{\zeta Ig}{r^\eta},
+ S(r) = \frac{Ig}{\zeta r^\eta},
 \end{equation}
@@
 f_p(S)
-= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta} \right]
-= \frac{\rho 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
+= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{Ig }{\zeta S} \right)^{2/\eta} \right]
+= \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{\zeta S} \right)^{2/\eta}.
 \end{align}
@@
 
-The real and imaginary components of receiver noise are independent and identically Gaussian distributed with zero mean and equal variance.
+The real and imaginary components of the noise in the cross-correlations are independent and identically Gaussian distributed with zero mean and equal variance.
 Consequently, an amplitude $x$ will be Rayleigh distributed \citep[\S6-2]{papoulis-stochastic-processes}:
@@
 If we assume that $I$, the average source strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-combined-distribution} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram.
-Although calibration could in theory solve $g$, almost all sources will come in through the edges of the beam, and finding the expected values for the gains is therefore hard.
+Although calibration could in theory solve $g$, almost all sources will come in through the far-sidelobes of the phased-array beam, and finding the expected values for the gains is therefore hard.
 The effects of these parameters on the histogram will be further discussed in \S\ref{sec:parameter-variability}.
@@
 \begin{equation} \label{eq:sample-with-propagation}
-\tilde{S}_{\textrm{RFI}} \leftarrow \frac{\zeta I g}{x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
+\tilde{S}_{\textrm{RFI}} \leftarrow \frac{I g}{\zeta x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
 \end{equation}
@@
 \subsection{Parameter variability} \label{sec:parameter-variability}
-In reality, the parameters $\rho$, $I$ and $g$, which are the source density per unit area, source strength and instrumental gain respectively, will not be constant over time and frequency but have a stochastic nature.
+In reality, the parameters $\rho$, $I$ and $g$, which are the source density per unit area, source strength and instrumental gain respectively, will not be constant, but can change over time and frequency.
+Therefore, they act like stochastic variables.
 However, since each specific value for these parameters produces a power law, the combined distribution will still show a power law, as long as the parameters follow a distribution that is steep at high amplitudes (in $\log$--$\log$ space), such as a Gaussian or uniform distribution.
@@
 As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is $Ig/r^2$ -- assuming free space propagation for the moment -- and therefore $S_L = Ig/r^2$.
-For Hata's propagation model, the more generic solution $S_L = \zeta Ig / r^\alpha$ is found.
+For Hata's propagation model, the more generic solution $S_L = Ig / (\zeta r^\alpha)$ is found.
 This value represents the apparent brightness of the sources that are the furthest away from the telescope.
@@
 Taken at face value these exponents imply in Hata's model that the average transmitting heights for sources affecting the LBA and HBA are $79$ and $13$~m respectively.
+There are no 79 m high transmitters nearby LOFAR stations in the LBA frequency range.
 Hata's model only goes down to 150 MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies, e.g.
@@
 Therefore, if data from different nights or interferometers are summed, and there is stationary RFI in the set, it will become detectable.
-All RFI that is below the noise but is not stationary, will act like normal noise and will therefore be harmless.
+All RFI that is at that point still below the noise, will act like normal noise and will therefore be harmless.
 

### 2013-01-22 articles/Article-RFI-distribution/article-rfi-distribution.tex
Slight work on ref II comments
Finalizing referee comments

@@
 A.~G.~de~Bruyn$^{2,3}$,
-S.~Zaroubi$^2$,
-W.~Brouw$^{2,3}$,
-B.~Ciardi$^4$,
-I.~T.~Iliev$^5$,
-\newauthor
-G.~J.~A.~Harker$^6$,
-L.~V.~E.~Koopmans$^2$,
-G.~Mellema$^7$,
-S.~J.~Wijnholds$^3$
+S.~Zaroubi$^2$%,
+%W.~Brouw$^{2,3}$,
+%B.~Ciardi$^4$,
+%I.~T.~Iliev$^5$,
+%\newauthor
+%G.~J.~A.~Harker$^6$,
+%L.~V.~E.~Koopmans$^2$,
+%G.~Mellema$^7$,
+%S.~J.~Wijnholds$^3$
 \\
@@
 $^{3}$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
-$^{4}$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
-$^{5}$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
-$^{6}$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
-$^7$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
+%$^{4}$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
+%$^{5}$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
+%$^{6}$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
+%$^7$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
 }
@@
 \begin{equation}\label{eq:spherical-cap-area}
- A_\textrm{cap} = \pi \left( a^2 + h^2 \right) \hspace{0.4cm}\textrm{with } a \le R \wedge h \le 2R,
+ A_\textrm{cap} = \pi \left( a^2 + h^2 \right) \hspace{0.4cm}\textrm{with } (a \le R) \wedge (h \le 2R),
 \end{equation}
@@
 Therefore, the spherical and two-dimensional cases lead to a distribution with equal power law.
-Evaluating Eq.~\eqref{eq:cumulative-spherical} is limited to $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth.
+Eq.~\eqref{eq:cumulative-spherical} is valid for $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth.
 This causes distant sources to settle below the horizon.
@@
 Various settings of the parameters were used, and samples were drawn as described in Eq.~\eqref{eq:drawing-amplitudes}.
-Solid lines: the combined distributions, dashed lines: the original distributions before mixing.
+Solid lines: the combined distributions, dashed lines: the power-law distributions before mixing.
 }
@@
 
-We assume that the stochastic nature of $I$, that is caused by the spread of transmitters with different intrinsic strengths, also has little effect on the $\log N$ -- $\log S$ histograms.
-Nevertheless, it is unlikely that $I$ follows a uniform or Gaussian distribution, because the distribution will contain few strong transmitters (such as radio stations) and many weak transmitters (such as remote controls).
+The stochastic nature of $I$, that is caused by the spread of transmitters with different intrinsic strengths, might also have an effect on the $\log N$ -- $\log S$ histograms.
+It is unlikely that $I$ follows a uniform or Gaussian distribution, because the distribution will contain few strong transmitters (such as radio stations) and many weak transmitters (such as remote controls).
 Therefore, variable $I$ might follow a power-law distribution by itself.
@@
 Therefore, it is likely that the spatial distribution will dominate the power law in the histogram.
+Otherwise, a turn-over point should be visible in the histogram.
 
@@
 \begin{tabular}{@{}llrr@{}}
-\textbf{Symbol} & \textbf{Name} & \textbf{LBA set}& \textbf{HBA set} \\
+\textbf{Symbol} & \textbf{Description} & \textbf{LBA set}& \textbf{HBA set} \\
 \hline
@@
 $\beta$ & Scaling factor of power law with exponent $\alpha$ & $4.0 \times 10^{17}$ & $3.4\times 10^{15}$ \\
+$\eta$ & Radiation fall-off speed for $\alpha$ ($\eta=2$ is free space) & 3.23 & 3.77 \\
 \hline
@@
 We have shown that the current detection algorithms can detect RFI well below the noise.
-Therefore, if data from different nights or interferometers are summed, and there is stationary RFI in the set, it will become detectable.
-All RFI that is at that point still below the noise, will act like normal noise and will therefore be harmless.
+Therefore, if data from different nights or interferometers are summed, and there is stationary RFI in the set, it will become detectable if it is consistently present.
+If it is still below the noise and not detectable, but behaves similar to the RFI we are already seeing, it will act like normal noise and will therefore be harmless.
 

### 2013-01-22 articles/Article-RFI-distribution/article-rfi-distribution.tex
Affiliations, etc

@@
 \author[A.~R.~Offringa et al.]{
-  A.~R.~Offringa$^{1,2}$\thanks{E-mail:
+  A.~R.~Offringa$^{1,2,3}$\thanks{E-mail:
 \url{offringa@mso.anu.edu.au}},
-A.~G.~de~Bruyn$^{2,3}$,
-S.~Zaroubi$^2$%,
-%W.~Brouw$^{2,3}$,
-%B.~Ciardi$^4$,
-%I.~T.~Iliev$^5$,
+A.~G.~de~Bruyn$^{3,4}$ and
+S.~Zaroubi$^3$%,
+%W.~Brouw$^{3,4}$,
+%B.~Ciardi$^5$,
+%I.~T.~Iliev$^6$,
 %\newauthor
-%G.~J.~A.~Harker$^6$,
-%L.~V.~E.~Koopmans$^2$,
-%G.~Mellema$^7$,
-%S.~J.~Wijnholds$^3$
+%G.~J.~A.~Harker$^7$,
+%L.~V.~E.~Koopmans$^3$,
+%G.~Mellema$^8$,
+%S.~J.~Wijnholds$^4$
 \\
 $^{1}$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
-$^{2}$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
-$^{3}$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
-%$^{4}$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
-%$^{5}$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
-%$^{6}$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
-%$^7$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
+$^{2}$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO)\\
+$^{3}$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
+$^{4}$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
+%$^{5}$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
+%$^{6}$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
+%$^{7}$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
+%$^8$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
 }
@@
 We explore the low-frequency RFI situation by studying brightness histograms of visibility data observed with LOFAR, similar to radio source count analyses that are used in cosmology.
-The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
+The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
 With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models.

### 2013-01-22 cotter/cotter.cpp
Cotter will now use the times from the raw date files -- which are 3 second more accurate compared to the ones in the meta files

@@ -142,7 +142,37 @@ void Cotter::Run(const char *outputFilename, size_t timeAvgFactor, size_t freqAv
 		do {
 			initializeReader();
 			
+			bool firstRead = (bufferPos == 0 && chunkIndex == 0);
+			
 			bool moreAvailableInCurrentFile = _reader->Read(bufferPos, _curChunkEnd-_curChunkStart);
+			
+			if(firstRead)
+			{
+				std::time_t startTime = _reader->StartTime();
+				std::tm startTimeTm;
+				gmtime_r(&startTime, &startTimeTm);
+				if(startTimeTm.tm_year+1900 != _mwaConfig.Header().year ||
+					startTimeTm.tm_mon+1 != _mwaConfig.Header().month ||
+					startTimeTm.tm_mday != _mwaConfig.Header().day ||
+					startTimeTm.tm_hour != _mwaConfig.Header().refHour ||
+					startTimeTm.tm_min != _mwaConfig.Header().refMinute ||
+					startTimeTm.tm_sec != _mwaConfig.Header().refSecond)
+				{
+					std::cout << "WARNING: start time according to raw files is "
+						<< startTimeTm.tm_year+1900  << '-' << startTimeTm.tm_mon+1 << '-' << startTimeTm.tm_mday << ' '
+						<< startTimeTm.tm_hour << ':' << startTimeTm.tm_min << ':' << startTimeTm.tm_sec << ",\nbut meta files say "
+						<< _mwaConfig.Header().year << '-' << _mwaConfig.Header().month << '-' << _mwaConfig.Header().day << ' '
+						<< _mwaConfig.Header().refHour << ':' << _mwaConfig.Header().refMinute << ':' << _mwaConfig.Header().refSecond <<
+					" !\nWill use start time from raw file, which should be most accurate.\n";
+					_mwaConfig.HeaderRW().year = startTimeTm.tm_year+1900;
+					_mwaConfig.HeaderRW().month = startTimeTm.tm_mon+1;
+					_mwaConfig.HeaderRW().day = startTimeTm.tm_mday;
+					_mwaConfig.HeaderRW().refHour = startTimeTm.tm_hour;
+					_mwaConfig.HeaderRW().refMinute = startTimeTm.tm_min;
+					_mwaConfig.HeaderRW().refSecond = startTimeTm.tm_sec;
[…]

### 2013-01-22 cotter/geometry.h

@@ -219,6 +219,7 @@ public:
 		Z = (EARTH_RAD_WGS84*(1.0-E_SQUARED)/chi + height_meters)*s_lat;
 	}
 	
+	
 };
 
 #endif

### 2013-01-22 cotter/gpufilereader.cpp

@@ -8,7 +8,6 @@
 void GPUFileReader::openFiles()
 {
 	int status = 0;
-
 	for(size_t i=0; i!=_filenames.size(); ++i)
 	{
 		fitsfile *fptr;
@@ -22,6 +21,17 @@ void GPUFileReader::openFiles()
 			
 			_fitsHDUCounts.push_back(hduCount);
 			std::cout << "There are " << hduCount << " HDUs in file " << _filenames[i] << '\n';
+			
+			long thisFileTime;
+			fits_read_key(fptr, TLONG, "TIME", &thisFileTime, 0, &status);
+			checkStatus(status);
+			
+			if(i == 0) _startTime = thisFileTime;
+			if(_startTime != thisFileTime)
+			{
+				std::cout << "WARNING: file number " << (i+1) << " of current time range has different start time!\n"
+					"Current file start time: " << thisFileTime << " previous file had: " << _startTime << ".\n";
+			}
 		}
 		else {
 			throwError(status, std::string("Cannot open file ") + _filenames[i]);

### 2013-01-22 cotter/gpufilereader.h

@@ -1,5 +1,6 @@
 #include <string>
 #include <vector>
+#include <ctime>
 
 #include <fitsio.h>
 #include <stdexcept>
@@ -66,6 +67,7 @@ class GPUFileReader : private FitsUser
 		{
 			return _isConjugated[(ant1 * 2 + pol1) * _nAntenna * 2 + (ant2 * 2 + pol2)];
 		}
+		std::time_t StartTime() const { return _startTime; }
 	private:
 		const static int pfb_output_to_input[64];
 		
@@ -94,4 +96,5 @@ class GPUFileReader : private FitsUser
 		std::vector<BaselineBuffer> _mappedBuffers;
 		std::vector<size_t> _corrInputToOutput;
 		std::vector<bool> _isConjugated;
+		std::time_t _startTime;
 };

### 2013-01-22 cotter/metafitsfile.cpp

@@ -162,7 +162,10 @@ void MetaFitsFile::parseKeyword(MWAHeader &header, MWAHeaderExt &headerExt, cons
 		header.fieldName = stripBand(headerExt.filename);
 	}
 	else if(name == "DATE-OBS")
+	{
+		parseFitsDate(keyValue, header.year, header.month, header.day, header.refHour, header.refMinute, header.refSecond);
 		headerExt.dateRequestedMJD = parseFitsDateToMJD(keyValue);
+	}
 	else if(name == "RAPHASE")
 		header.raHrs = atof(keyValue) * (24.0 / 360.0);
 	else if(name == "DECPHASE")
@@ -202,7 +205,7 @@ void MetaFitsFile::parseKeyword(MWAHeader &header, MWAHeaderExt &headerExt, cons
 	else if(name == "FREQCENT")
 		header.centralFrequencyMHz = atof(keyValue);
 	else if(name == "DATESTRT")
-		parseFitsDate(keyValue, header.year, header.month, header.day, header.refHour, header.refMinute, header.refSecond);
+		; //parseFitsDate(keyValue, header.year, header.month, header.day, header.refHour, header.refMinute, header.refSecond);
 	else if(name == "DATE")
 		; // Date that metafits was created; ignored.
 	else if(name == "EXPOSURE" || name == "MJD" || name == "LST" || name == "HA" || name == "AZIMUTH" || name == "ALTITUDE" || name == "SUN-DIST" || name == "MOONDIST" || name == "JUP-DIST" || name == "GRIDNUM" || name == "RECVRS" || name == "CHANNELS" || name == "SUN-ALT" || name == "TILEFLAG" || name == "NAV_FREQ" || name == "FINECHAN" || name == "TIMEOFF")

### 2013-01-22 cotter/mwaconfig.cpp

@@ -186,7 +186,12 @@ void MWAHeader::Validate(bool lockPointing)
 	if(haHrsStart == -99.0 && lockPointing)
 		throw std::runtime_error("HA must be specified in header when locking pointing.");
 	
-	dateFirstScanMJD = 0.5*(integrationTime/86400.0) + Geometry::GetMJD(
+	dateFirstScanMJD = GetDateFirstScanFromFields();
+}
+
+double MWAHeader::GetDateFirstScanFromFields() const
+{
+	return 0.5*(integrationTime/86400.0) + Geometry::GetMJD(
 		year, month, day, refHour, refMinute, refSecond);
 }
 

### 2013-01-22 cotter/mwaconfig.h

@@ -51,6 +51,7 @@ struct MWAHeader
 			return dateFirstScanMJD + (integrationTime/86400.0)*nScans;
 		}
 		double GetStartDateMJD() const;
+		double GetDateFirstScanFromFields() const;
 		
 		void Validate(bool lockPointing);
 	private:
@@ -137,6 +138,8 @@ class MWAConfig
 		static double ArrayLongitudeRad();
 		static double ArrayHeightMeters();
 		static double VelocityFactor();
+		
+		MWAHeader &HeaderRW() { return _header; }
 	private:
 		std::vector<MWAInput> _inputs;
 		std::vector<MWAAntenna> _antennae;

### 2013-01-24 articles/Article-RFI-distribution/article-rfi-distribution.tex
Leons and others comments

@@
 \url{offringa@mso.anu.edu.au}},
-A.~G.~de~Bruyn$^{3,4}$ and
-S.~Zaroubi$^3$%,
-%W.~Brouw$^{3,4}$,
-%B.~Ciardi$^5$,
-%I.~T.~Iliev$^6$,
-%\newauthor
-%G.~J.~A.~Harker$^7$,
-%L.~V.~E.~Koopmans$^3$,
-%G.~Mellema$^8$,
-%S.~J.~Wijnholds$^4$
-\\
-$^{1}$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
-$^{2}$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO)\\
-$^{3}$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
-$^{4}$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
-%$^{5}$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
-%$^{6}$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
-%$^{7}$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
-%$^8$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
+A.~G.~de~Bruyn$^{3,4}$,
+S.~Zaroubi$^3$,
+L.~V.~E.~Koopmans$^3$,
+\newauthor
+S.~J.~Wijnholds$^4$,
+F.~B.~Abdalla$^5$,
+W.~Brouw$^{3,4}$,
+B.~Ciardi$^6$,
+I.~T.~Iliev$^7$,
+G.~J.~A.~Harker$^8$,
+\newauthor
+G.~Mellema$^8$\\
+$^1$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
+$^2$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO)\\
+$^3$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
+$^4$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
+$^5$ UCL Department of Physics and Astronomy, London WC1E 6BT, United Kingdom\\
+$^6$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
+$^7$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
+$^8$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
+$^9$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
 }
@@
 \end{equation}
-with (amplitude) gains $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$.
+with (real-valued amplitude) gains $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$.
 Here, $r_\textrm{L}$ is a limiting distance, which will be well below the diameter of the Earth.
@@
 \begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
-\caption{Effect of transmitter height on the slope of a log-log histogram.}
+\caption{Effect of transmitter height on the slope of a log-log histogram.
+According to Hata's model, this is valid for the range 30--200 m.
+The trend of the slope will not continue indefinitely when increasing the height further.
+Instead it will converge to a $-2$ slope, which corresponds with free-space propagation.}
 \label{fig:plot-transmitter-height}
@@
 However, the measured distribution is a mixture of the two.
-A perfect RFI detector would separate the samples in two distributions; one that is not affected by RFI, and therefore contains noise only, and one that is the sum of RFI and noise.
-Real RFI detectors can separate these distributions to some extent, but due to false positives and false negatives, the histograms will get mixed nevertheless.
+A perfect RFI detector would classify the samples in two groups; one group of samples that are not affected by RFI, and therefore contain noise only, and one one group of samples that are contaminated, and are the sum of RFI and noise.
+Real RFI detectors can classify samples into these groups to some extent, but due to false positives and false negatives, the groups will get mixed nevertheless.
 Even more, we still need to take into account that the samples classified as RFI are also affected by noise.
@@
 Another effect that is absorbed in $g$, is the beam of the instrument.
-At the point of writing, LOFAR beam models are still being developed and are not yet well parametrized.
+At the point of writing, LOFAR beam models are still being developed and are not yet well parametrized near the horizon.
 It is likely that most RFI sources are observed at the edges of the beam.
-Because most sources will be observed with similar gains (within one order of magnitude), it can be expected that the beam will have a limited effect on the histogram properties of an observation, and is comparable with the effect of the frequency response.
+Nevertheless, most sources will be observed with similar gains (within one order of magnitude), and it can be expected that the beam will have a limited effect on the histogram properties of an observation.
+It is therefore comparable with the effect of the frequency response.
 
@@
 
-Assume also that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}.
+$S_1$ and $S_2$ are selected by visually inspection of the histogram.
+Assume that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}, and that the effect of the Rayleigh component on the histogram $> S_1$ is negligible.
 The hypothetical upper limit $S_U$ of the distribution can now be found, i.e., the highest amplitude that would be observed when the distribution follows the power law up to that point, by solving
@@
 \end{align}
-This is essentially the flux density that is caused by RFI without using RFI detection or excision algorithms.
+This is essentially the average flux density that is caused by RFI without using RFI detection or excision algorithms.
 $E(S_R)$ has the same units as $S_L$ and $S_U$, thus after calibration (see \S\ref{sec:calibration}) could be given in Jy.
-In practice, the increase of system noise after correlation is much less severe because of RFI flagging, which excises a part of the RFI.
+In practice, the increase of data noise after correlation is much less severe because of RFI flagging, which excises a part of the RFI.
 One can assume that all RFI above a certain power level is found by the detector.
@@
 However, we found that this is not a representative error in our case, because the errors in the slope are not normally distributed.
-The noise in the part of the histogram over which the slope is calculated, is very small due to the large amount of samples.
+The noise in the part of the histogram over which the slope is calculated, is very small due to the large number of samples.
 Consequently, the estimated standard deviation of $\hat \alpha$ is very low.
@@
 The left part of the histogram matches the Rayleigh distribution well up to the mode of the distribution.
-The bulge around the mode of the LBA histogram is wider due to the larger effect of the antenna response as discussed in \S\ref{sec:parameter-variability}.
+The bulge around the mode of the LBA histogram is wider due to the larger effect of the antenna response, i.e., variability of $g$ as discussed in \S\ref{sec:parameter-variability}.
 As can be seen in Fig.~\ref{fig:plot-dist-per-frequency-LBA}, the Rayleigh-bulges of individual sub-bands are not that wide, but they are not aligned because of the differing noise levels.
@@
 because the effect of the ionosphere becomes stronger.
-Intervals for the exponents with representative $3\sigma$ confidence boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.53;-1.49]$ for the HBA.
+Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.53;-1.49]$ for the HBA.
 The estimate of the HBA is thus more accurate, because its histogram deviates less from the power law.
@@
 On the assumption that the power-law distribution for RFI sources will continue down into the noise, we have constructed a full parametrization of the RFI apparent flux distribution.
-By assuming that all samples contain some contribution of RFI, we find that the average flux density of RFI that leaks through the detector is $484$--$496$~mJy for the LBA and $167$--$171$~mJy for the HBA (depending on the used method).
+By assuming that all samples contain some contribution of RFI, we find that the average flux density of RFI after excision by automated flagging is $484$--$496$~mJy for the LBA and $167$--$171$~mJy for the HBA.
 These values should be compared to the noise in individual samples of $770$~Jy (LBA) and $77$~Jy (HBA) (see Table~\ref{table:dist-data-quantities}), and are upper limits for what can be expected.
@@
 
-In experiments such as the LOFAR EoR project, a simulation pipeline is used to create a realistic estimate of the signal that can be expected.
+In projects such as the EoR detection experiment with LOFAR, a simulation pipeline is used to create a realistic estimate of the signal that can be expected.
 Currently, these simulations do not include the effects of RFI.
@@
 
-On top of all these possible RFI reducing effects, our given RFI constraints could be too pessimistic because we have assumed 100\% of the spectrum is occupied by RFI.
-If, say, only 10\% of the spectrum is occupied, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all.
+On top of all these possible RFI reducing effects, our given RFI constraints could be too pessimistic because we have assumed 100\% of the spectrum is occupied by RFI over the entire duration of the observation.
+If, say, only 10\% of the samples are affected by RFI, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all.
 With current data, one can only speculate how much of the electromagnetic spectrum is truly occupied.

### 2013-01-30 articles/Article-RFI-distribution/article-rfi-distribution.tex

@@
 \url{offringa@mso.anu.edu.au}},
-A.~G.~de~Bruyn$^{3,4}$,
-S.~Zaroubi$^3$,
-L.~V.~E.~Koopmans$^3$,
-\newauthor
-S.~J.~Wijnholds$^4$,
-F.~B.~Abdalla$^5$,
-W.~Brouw$^{3,4}$,
-B.~Ciardi$^6$,
-I.~T.~Iliev$^7$,
-G.~J.~A.~Harker$^8$,
-\newauthor
-G.~Mellema$^8$\\
+A.~G.~de~Bruyn$^{3,4}$
+and S.~Zaroubi$^3$%,
+%L.~V.~E.~Koopmans$^3$,
+%\newauthor
+%S.~J.~Wijnholds$^4$,
+%F.~B.~Abdalla$^5$,
+%W.~Brouw$^{3,4}$,
+%B.~Ciardi$^6$,
+%I.~T.~Iliev$^7$,
+%G.~J.~A.~Harker$^8$,
+%\newauthor
+%G.~Mellema$^8$
+\\
 $^1$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
@@
 $^4$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
-$^5$ UCL Department of Physics and Astronomy, London WC1E 6BT, United Kingdom\\
-$^6$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
-$^7$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
-$^8$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
-$^9$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
+%$^5$ UCL Department of Physics and Astronomy, London WC1E 6BT, United Kingdom\\
+%$^6$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
+%$^7$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
+%$^8$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
+%$^9$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
 }
@@
 This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe.
-We explore the low-frequency RFI situation by studying brightness histograms of visibility data observed with LOFAR, similar to radio source count analyses that are used in cosmology.
+We explore the low-frequency RFI situation by studying brightness histograms of visibility data observed with LOFAR, similar to radio-source-count analyses that are used in cosmology.
 The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
@@
 Many techniques have been designed to mitigate this effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, LOFAR-RFI-pipeline}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.
-In most cases, and with the help of carefull spectrum allocation management and radio-quiet zones, these methods excise enough of the interference to allow calibration and imaging of the data.
+In most cases, and with the help of careful spectrum allocation management and radio-quiet zones, these methods excise enough of the interference to allow calibration and imaging of the data.
 
@@
 This work explores the information that is present in interference distributions, in order to analyse possible low-level interference that is not detectable by standard detection methods.
-Our approach is similar to the radio source count analysis that is used in cosmology \citep{condon-cosm-evol-of-radio-sources}, also named $\log N$ -- $\log S$ analysis, where $N$ and $S$ refer to the celestial source count and brightness respectively.
+Our approach is similar to the radio-source-count analyses that are used in cosmology \citep{condon-cosm-evol-of-radio-sources}, also named $\log N$ -- $\log S$ analyses, where $N$ and $S$ refer to the celestial source count and brightness respectively.
 The slope in such a plot contains information about source populations, their luminosity functions and the geometry of the Universe.
@@
 
-We assume that the source observed is fully coherent, but a possible decoherence factor can be absorbed in the gains.
+We assume that the source observed is fully coherent, but a possible de-coherence factor can be absorbed in the gains.
 Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the narrow-band condition.
@@
 \end{equation}
-which equals the formule for a planar source distribution given in Eq.~\eqref{eq:cumulative-distance-function}.
+which equals the formula for a planar source distribution given in Eq.~\eqref{eq:cumulative-distance-function}.
 Therefore, the spherical and two-dimensional cases lead to a distribution with equal power law.
@@
 Therefore, we introduce an error estimate $\epsilon_\alpha$ that quantifies a normalized standard deviation of the slope over the range.
-This error is formed by calculating the slope over $n_\alpha$ smaller subranges in the histogram, creating $n_\alpha$ estimates $\alpha_i$.
+This error is formed by calculating the slope over $n_\alpha$ smaller sub-ranges in the histogram, creating $n_\alpha$ estimates $\alpha_i$.
 If the errors in $\alpha_i$ are normally distributed with mean zero, the standard deviation over the full range will be $\sqrt{n_\alpha}$ times smaller.
@@
 \end{equation}
-This estimate is slightly depending on the number of subranges that is used, $n_\alpha$, because the errors are not Gaussian distributed, but we found that $\epsilon_{\hat\alpha}$ is more representative than the standard error of $\hat\alpha$.
+This estimate is slightly depending on the number of sub-ranges that is used, $n_\alpha$, because the errors are not Gaussian distributed, but we found that $\epsilon_{\hat\alpha}$ is more representative than the standard error of $\hat\alpha$.
 
@@
 For the EoR project with $1$~MHz resolution, Eq.~\ref{eq:rfi-noise-level} results in $\sigma_\textrm{RFI}\approx 4.7$ mJy.
-However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz subband (Yatawatta et al.
+However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz sub-band (Yatawatta et al.
 2013, \textit{in press}), and the resulting images show no signs of RFI.

### 2013-01-31 articles/Article-RFI-distribution/article-rfi-distribution.tex
Textual

@@
 However, the measured distribution is a mixture of the two.
-A perfect RFI detector would classify the samples in two groups; one group of samples that are not affected by RFI, and therefore contain noise only, and one one group of samples that are contaminated, and are the sum of RFI and noise.
+A perfect RFI detector would classify the samples in two groups; one group of samples that are not affected by RFI, and therefore contain noise only, and one group of samples that are contaminated, and are the sum of RFI and noise.
 Real RFI detectors can classify samples into these groups to some extent, but due to false positives and false negatives, the groups will get mixed nevertheless.

### 2013-02-26 articles/Article-RFI-distribution/article-rfi-distribution.tex
Author list full and suggestions by Zarka

@@
 
-\author[A.~R.~Offringa et al.]{
-  A.~R.~Offringa$^{1,2,3}$\thanks{E-mail:
+\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2,3}$\thanks{E-mail:
 \url{offringa@mso.anu.edu.au}},
-A.~G.~de~Bruyn$^{3,4}$
-and S.~Zaroubi$^3$%,
-%L.~V.~E.~Koopmans$^3$,
-%\newauthor
-%S.~J.~Wijnholds$^4$,
-%F.~B.~Abdalla$^5$,
-%W.~Brouw$^{3,4}$,
-%B.~Ciardi$^6$,
-%I.~T.~Iliev$^7$,
-%G.~J.~A.~Harker$^8$,
-%\newauthor
-%G.~Mellema$^8$
+A.~G.~de~Bruyn$^{4,3}$, S.~Zaroubi$^{3}$, L.~V.~E.~Koopmans$^{3}$, \newauthor
+S.~J.~Wijnholds$^{4}$, F.~B.~Abdalla$^{5}$, W.~N.~Brouw$^{3,4}$, B.~Ciardi$^{6}$, I.~T.~Iliev$^{7}$, \newauthor
+G.~J.~A.~Harker$^{8}$, G.~Mellema$^{9}$, G.~Bernardi$^{10}$, P.~Zarka$^{11}$, A.~Ghosh$^{3}$,\newauthor A.~Alexov$^{12}$, J.~Anderson$^{13}$, A.~Asgekar$^{4}$, I.~M.~Avruch$^{14,3}$, R.~Beck$^{13}$,\newauthor
+M.~E.~Bell$^{2}$, M.~R.~Bell$^{6}$, M.~J.~Bentum$^{4}$, P.~Best$^{15}$, L.~B\^{i}rzan$^{16}$,
+F.~Breitling$^{17}$, \newauthor
+J.~Broderick$^{18}$, M.~Br\"uggen$^{19}$, H.~R.~Butcher$^{4,1}$, F.~de~Gasperin$^{19}$, E.~de~Geus$^{4}$,\newauthor
+M.~de~Vos$^{4}$, S.~Duscha$^{4}$, J.~Eisl\"offel$^{20}$, R.~A.~Fallows$^{4}$, C.~Ferrari$^{21}$,\newauthor
+W.~Frieswijk$^{4}$, M.~A.~Garrett$^{4,16}$, J.~Grie\ss{}meier$^{22}$, T.~E.~Hassall$^{18}$, A.~Horneffer$^{13}$,\newauthor
+M.~Iacobelli$^{16}$, E.~Juette$^{23}$, A. ~Karastergiou$^{24}$, W.~Klijn$^{4}$, V.~I.~Kondratiev$^{4,25}$,\newauthor
+M.~Kuniyoshi$^{13}$, G.~Kuper$^{4}$, J.~van~Leeuwen$^{4,26}$, M.~Loose$^{4}$, P.~Maat$^{4}$,\newauthor
+G.~Macario$^{21}$, G.~Mann$^{17}$, J.~P.~McKean$^{4}$, H.~Meulman$^{4}$, M.~J.~Norden$^{4}$, 
+\newauthor
+E.~Orru$^{4}$, H.~Paas$^{27}$, M.~Pandey-Pommier$^{28}$, R.~Pizzo$^{4}$, A.~G.~Polatidis$^{4}$,\newauthor
+D.~Rafferty$^{16}$, W.~Reich$^{13}$, R.~van~Nieuwpoort$^{4}$, H.~R\"ottgering$^{16}$, \newauthor
+A.~M.~M.~Scaife$^{18}$, J.~Sluman$^{4}$, O.~Smirnov$^{29,30}$, C.~Sobey$^{13}$, M.~Tagger$^{22}$,\newauthor
+Y.~Tang$^{4}$, C.~Tasse$^{11}$, S.~ter~Veen$^{31}$, C.~Toribio$^{4}$, R.~Vermeulen$^{4}$, C.~Vocks$^{17}$,\newauthor
+R.~J.~van~Weeren$^{10}$, M.~W.~Wise$^{4,26}$, O.~Wucknitz$^{32,13}$
 \\
-$^1$Australian National University, RSAA Mt. Stromlo Observatory, Cotter Road, Weston Creek ACT 2611, Australia\\
-$^2$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO)\\
-$^3$University of Groningen, Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands\\
-$^4$ASTRON, PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
-%$^5$ UCL Department of Physics and Astronomy, London WC1E 6BT, United Kingdom\\
-%$^6$Max-Planck Institute for Astrophysics, Karl-Schwarzschild-Strasse 1, 85748 Garching bei M\"unchen, Germany\\
-%$^7$University of Sussex, Falmer, Brighton BN1 9QH, UK\\
-%$^8$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA\\
-%$^9$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden 
+$^{1}$RSAA, Australian National University, Mt Stromlo Observatory, via Cotter Road, Weston, ACT 2611, Australia \\
+$^{2}$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO) \\
+$^{3}$Kapteyn Astronomical Institute, PO Box 800, 9700 AV Groningen, The Netherlands \\
+$^{4}$Netherlands Institute for Radio Astronomy (ASTRON), Postbus 2, 7990 AA Dwingeloo, The Netherlands \\
+$^{5}$UCL Department of Physics and Astronomy, London WC1E 6BT, United Kingdom \\
+$^{6}$Max Planck Institute for Astrophysics, Karl Schwarzschild Str. 1, 85741 Garching, Germany \\
+$^{7}$University of Sussex, Falmer, Brighton BN1 9QH, UK \\
+$^{8}$Center for Astrophysics and Space Astronomy, University of Colorado Boulder, CO 80309, USA \\
+$^{9}$Stockholm University, AlbaNova University Center, Stockholm Observatory, SE-106 91 Stockholm, Sweden \\
+$^{10}$Harvard-Smithsonian Center for Astrophysics, 60 Garden Street, Cambridge, MA 02138, USA \\
+$^{11}$LESIA, UMR CNRS 8109, Observatoire de Paris, 92195   Meudon, France \\
+$^{12}$Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD 21218, USA \\
+$^{13}$Max-Planck-Institut f\"{u}r Radioastronomie, Auf dem H\"ugel 69, 53121 Bonn, Germany \\
+$^{14}$SRON Netherlands Insitute for Space Research, Sorbonnelaan 2, 3584 CA, Utrecht, The Netherlands \\
+$^{15}$Institute for Astronomy, University of Edinburgh, Royal Observatory of Edinburgh, Blackford Hill, Edinburgh EH9 3HJ, UK \\
+$^{16}$Leiden Observatory, Leiden University, PO Box 9513, 2300 RA Leiden, The Netherlands \\
+$^{17}$Leibniz-Institut f\"{u}r Astrophysik Potsdam (AIP), An der Sternwarte 16, 14482 Potsdam, Germany \\
+$^{18}$School of Physics and Astronomy, University of Southampton, Southampton, SO17 1BJ, UK \\
+$^{19}$University of Hamburg, Gojenbergsweg 112, 21029 Hamburg, Germany \\
+$^{20}$Th\"{u}ringer Landessternwarte, Sternwarte 5, D-07778 Tautenburg, Germany \\
+$^{21}$Laboratoire Lagrange, UMR7293, Universit\`{e} de Nice Sophia-Antipolis, CNRS, Observatoire de la C\'{o}te d'Azur, 06300 Nice, France \\
+$^{22}$Laboratoire de Physique et Chimie de l' Environnement et de l' Espace, LPC2E UMR 7328 CNRS, 45071 Orl\'{e}ans Cedex 02, France \\
+$^{23}$Astronomisches Institut der Ruhr-Universit\"{a}t Bochum, Universitaetsstrasse 150, 44780 Bochum, Germany \\
+$^{24}$Astrophysics, University of Oxford, Denys Wilkinson Building, Keble Road, Oxford OX1 3RH \\
+$^{25}$Astro Space Center of the Lebedev Physical Institute, Profsoyuznaya str. 84/32, Moscow 117997, Russia \\
+$^{26}$Astronomical Institute 'Anton Pannekoek', University of Amsterdam, Postbus 94249, 1090 GE Amsterdam, The Netherlands \\
+$^{27}$Center for Information Technology (CIT), University of Groningen, The Netherlands \\
+$^{28}$Centre de Recherche Astrophysique de Lyon, Observatoire de Lyon, 9 av Charles Andr\'{e}, 69561 Saint Genis Laval Cedex, France \\
+$^{29}$Centre for Radio Astronomy Techniques \& Technologies (RATT), Department of Physics and Elelctronics, Rhodes University, PO Box 94, Grahamstown 6140, South Africa \\
+$^{30}$SKA South Africa, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa \\
+$^{31}$Department of Astrophysics/IMAPP, Radboud University Nijmegen, P.O. Box 9010, 6500 GL Nijmegen, The Netherlands \\
+$^{32}$Argelander-Institut f\"{u}r Astronomie, University of Bonn, Auf dem H\"{u}gel 71, 53121, Bonn, Germany
 }
@@
 
+\label{firstpage}
 \maketitle
-\label{firstpage}
 
@@
 
-Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in a two-dimensional plane.
+Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in the local two-dimensional horizontal plane.
 This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights.
@@
 The trend of the slope will not continue indefinitely when increasing the height further.
-Instead it will converge to a $-2$ slope, which corresponds with free-space propagation.}
+Instead it will converge to a $-2$ slope, which corresponds to free-space propagation.}
 \label{fig:plot-transmitter-height}
@@
 
-\subsection{Noise contribution} \label{sec:histogram-noise}
-The full measured distribution will consist of the power-law distribution combined with that of the noise and the celestial signal.
+\subsection{Thermal noise contribution} \label{sec:histogram-noise}
+The full measured distribution will consist of the power-law distribution combined with that of the thermal noise and the celestial signal.
 For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong sources.
@@
 
-\label{lastpage}
 \section*{Acknowledgments}
@@
 
+\label{lastpage}
+
 \end{document}

### 2013-02-28 articles/Article-RFI-distribution/article-rfi-distribution.tex
Acknowledgments and final touch

@@
 M.~Kuniyoshi$^{13}$, G.~Kuper$^{4}$, J.~van~Leeuwen$^{4,26}$, M.~Loose$^{4}$, P.~Maat$^{4}$,\newauthor
-G.~Macario$^{21}$, G.~Mann$^{17}$, J.~P.~McKean$^{4}$, H.~Meulman$^{4}$, M.~J.~Norden$^{4}$, 
+G.~Maccario$^{21}$, G.~Mann$^{17}$, J.~P.~McKean$^{4}$, H.~Meulman$^{4}$, M.~J.~Norden$^{4}$, 
 \newauthor
@@
 Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources add up consistently.
-These observations almost reach the thermal noise and are thought not to be limited by RFI.
+This is in contrast with the actual observations that almost reach the thermal noise and are thought not to be limited by RFI.
 Therefore, we conclude that it is likely that most RFI sources do not linearly increase the noise, in which case RFI will not prevent an EoR detection with LOFAR.
@@
 However, astronomical radio observations can be affected by radio-frequency interference (RFI), which makes it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}.
-Many techniques have been designed to mitigate this effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, LOFAR-RFI-pipeline}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.
+Many techniques have been designed to mitigate this effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, post-correlation-rfi-classification}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.
 In most cases, and with the help of careful spectrum allocation management and radio-quiet zones, these methods excise enough of the interference to allow calibration and imaging of the data.
@@
 For the EoR project with $1$~MHz resolution, Eq.~\ref{eq:rfi-noise-level} results in $\sigma_\textrm{RFI}\approx 4.7$ mJy.
-However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz sub-band (Yatawatta et al.
-2013, \textit{in press}), and the resulting images show no signs of RFI.
+However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz sub-band \citep{ncp-eor-yatawatta}, and the resulting images show no signs of RFI.
 This implies that either the upper limit is far from the actual RFI situation, or Eq.~\ref{eq:rfi-noise-level} is not applicable to most of the RFI that is observed with LOFAR.
@@
 It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright foreground sources.
-Preliminary analysis of EoR NCP observations of a single night have reached the thermal noise (Yatawatta et al.
-2013, \textit{in press}), but do not show leaked RFI at the pole.
+Preliminary analysis of EoR NCP observations of a single night have reached the thermal noise \citep{ncp-eor-yatawatta}, but do not show leaked RFI at the pole.
 
@@
 LOFAR, the Low-Frequency Array designed and constructed by ASTRON, has facilities in several countries, that are owned by various parties (each with their own funding sources), and that are collectively operated by the International LOFAR Telescope (ILT) foundation under a joint scientific policy.
+Parts of this research were conducted by the Australian Research Council Centre of Excellence for All-sky Astrophysics (CAASTRO), through project number CE110001020.
+C.
+Ferrari and G.
+Maccario acknowledge financial support by the ``Agence Nationale de la Recherche'' through grant ANR-09-JCJC-0001-01.
 

### 2013-02-28 articles/Article-RFI-distribution/article-rfi-distribution.tex
Minor fix to abstract

@@
 Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources add up consistently.
-This is in contrast with the actual observations that almost reach the thermal noise and are thought not to be limited by RFI.
-Therefore, we conclude that it is likely that most RFI sources do not linearly increase the noise, in which case RFI will not prevent an EoR detection with LOFAR.
+This is in contrast with actual observations, which almost reach the thermal noise and are thought not to be limited by RFI.
+Therefore, we conclude that it is likely that most RFI sources do not linearly increase the noise.
+We believe that RFI will not prevent an EoR detection with LOFAR.
 \end{abstract}

### 2013-02-28 articles/Article-RFI-distribution/article-rfi-distribution.tex
Fixing Matthews comments

@@
 Radio astronomy concerns itself with the observation of radiation from celestial sources at radio wavelengths.
-However, astronomical radio observations can be affected by radio-frequency interference (RFI), which makes it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}.
+However, astronomical radio observations can be affected by radio-frequency interference (RFI), which might make it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}.
 Many techniques have been designed to mitigate this effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, post-correlation-rfi-classification}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.
@@
 Ground-based Cosmic Microwave Background (CMB) experiments are another class of experiments involving long integration times (e.g., \citealt{ska-cmb-subrahmanyan}).
-For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest.
+For these experiments, it is important to know the possible effect of low-level interference on the data, as these might overshadow or alter the signal of interest.
 In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, van Haarlem et al.

### 2013-03-01 articles/Article-RFI-distribution/article-rfi-distribution.tex
Fixed suggestions by Matthew and Bryans comment on affiliation of Martin

@@
 G.~J.~A.~Harker$^{8}$, G.~Mellema$^{9}$, G.~Bernardi$^{10}$, P.~Zarka$^{11}$, A.~Ghosh$^{3}$,\newauthor A.~Alexov$^{12}$, J.~Anderson$^{13}$, A.~Asgekar$^{4}$, I.~M.~Avruch$^{14,3}$, R.~Beck$^{13}$,\newauthor
-M.~E.~Bell$^{2}$, M.~R.~Bell$^{6}$, M.~J.~Bentum$^{4}$, P.~Best$^{15}$, L.~B\^{i}rzan$^{16}$,
-F.~Breitling$^{17}$, \newauthor
-J.~Broderick$^{18}$, M.~Br\"uggen$^{19}$, H.~R.~Butcher$^{4,1}$, F.~de~Gasperin$^{19}$, E.~de~Geus$^{4}$,\newauthor
-M.~de~Vos$^{4}$, S.~Duscha$^{4}$, J.~Eisl\"offel$^{20}$, R.~A.~Fallows$^{4}$, C.~Ferrari$^{21}$,\newauthor
-W.~Frieswijk$^{4}$, M.~A.~Garrett$^{4,16}$, J.~Grie\ss{}meier$^{22}$, T.~E.~Hassall$^{18}$, A.~Horneffer$^{13}$,\newauthor
-M.~Iacobelli$^{16}$, E.~Juette$^{23}$, A. ~Karastergiou$^{24}$, W.~Klijn$^{4}$, V.~I.~Kondratiev$^{4,25}$,\newauthor
-M.~Kuniyoshi$^{13}$, G.~Kuper$^{4}$, J.~van~Leeuwen$^{4,26}$, M.~Loose$^{4}$, P.~Maat$^{4}$,\newauthor
-G.~Maccario$^{21}$, G.~Mann$^{17}$, J.~P.~McKean$^{4}$, H.~Meulman$^{4}$, M.~J.~Norden$^{4}$, 
-\newauthor
-E.~Orru$^{4}$, H.~Paas$^{27}$, M.~Pandey-Pommier$^{28}$, R.~Pizzo$^{4}$, A.~G.~Polatidis$^{4}$,\newauthor
-D.~Rafferty$^{16}$, W.~Reich$^{13}$, R.~van~Nieuwpoort$^{4}$, H.~R\"ottgering$^{16}$, \newauthor
-A.~M.~M.~Scaife$^{18}$, J.~Sluman$^{4}$, O.~Smirnov$^{29,30}$, C.~Sobey$^{13}$, M.~Tagger$^{22}$,\newauthor
-Y.~Tang$^{4}$, C.~Tasse$^{11}$, S.~ter~Veen$^{31}$, C.~Toribio$^{4}$, R.~Vermeulen$^{4}$, C.~Vocks$^{17}$,\newauthor
-R.~J.~van~Weeren$^{10}$, M.~W.~Wise$^{4,26}$, O.~Wucknitz$^{32,13}$
+M.~E.~Bell$^{2,15}$, M.~R.~Bell$^{6}$, M.~J.~Bentum$^{4}$, P.~Best$^{16}$, L.~B\^{i}rzan$^{17}$, F.~Breitling$^{18}$,\newauthor
+J.~Broderick$^{19}$, M.~Br\"uggen$^{20}$, H.~R.~Butcher$^{4,1}$, F.~de~Gasperin$^{20}$, E.~de~Geus$^{4}$,\newauthor
+M.~de~Vos$^{4}$, S.~Duscha$^{4}$, J.~Eisl\"offel$^{21}$, R.~A.~Fallows$^{4}$, C.~Ferrari$^{22}$,\newauthor
+W.~Frieswijk$^{4}$, M.~A.~Garrett$^{4,17}$, J.~Grie\ss{}meier$^{23}$, T.~E.~Hassall$^{19}$, A.~Horneffer$^{13}$,\newauthor
+M.~Iacobelli$^{17}$, E.~Juette$^{24}$, A. ~Karastergiou$^{25}$, W.~Klijn$^{4}$, V.~I.~Kondratiev$^{4,26}$,\newauthor
+M.~Kuniyoshi$^{13}$, G.~Kuper$^{4}$, J.~van~Leeuwen$^{4,27}$, M.~Loose$^{4}$, P.~Maat$^{4}$,\newauthor
+G.~Macario$^{22}$, G.~Mann$^{18}$, J.~P.~McKean$^{4}$, H.~Meulman$^{4}$, M.~J.~Norden$^{4}$,\newauthor
+E.~Orru$^{4}$, H.~Paas$^{28}$, M.~Pandey-Pommier$^{29}$, R.~Pizzo$^{4}$, A.~G.~Polatidis$^{4}$,\newauthor
+D.~Rafferty$^{17}$, W.~Reich$^{13}$, R.~van~Nieuwpoort$^{4}$, H.~R\"ottgering$^{17}$,\newauthor
+A.~M.~M.~Scaife$^{19}$, J.~Sluman$^{4}$, O.~Smirnov$^{30,31}$, C.~Sobey$^{13}$, M.~Tagger$^{23}$,\newauthor
+Y.~Tang$^{4}$, C.~Tasse$^{11}$, S.~ter~Veen$^{32}$, C.~Toribio$^{4}$, R.~Vermeulen$^{4}$, C.~Vocks$^{18}$,\newauthor
+R.~J.~van~Weeren$^{10}$, M.~W.~Wise$^{4,27}$, O.~Wucknitz$^{33,13}$
 \\
@@
 $^{14}$SRON Netherlands Insitute for Space Research, Sorbonnelaan 2, 3584 CA, Utrecht, The Netherlands \\
-$^{15}$Institute for Astronomy, University of Edinburgh, Royal Observatory of Edinburgh, Blackford Hill, Edinburgh EH9 3HJ, UK \\
-$^{16}$Leiden Observatory, Leiden University, PO Box 9513, 2300 RA Leiden, The Netherlands \\
-$^{17}$Leibniz-Institut f\"{u}r Astrophysik Potsdam (AIP), An der Sternwarte 16, 14482 Potsdam, Germany \\
-$^{18}$School of Physics and Astronomy, University of Southampton, Southampton, SO17 1BJ, UK \\
-$^{19}$University of Hamburg, Gojenbergsweg 112, 21029 Hamburg, Germany \\
-$^{20}$Th\"{u}ringer Landessternwarte, Sternwarte 5, D-07778 Tautenburg, Germany \\
-$^{21}$Laboratoire Lagrange, UMR7293, Universit\`{e} de Nice Sophia-Antipolis, CNRS, Observatoire de la C\'{o}te d'Azur, 06300 Nice, France \\
-$^{22}$Laboratoire de Physique et Chimie de l' Environnement et de l' Espace, LPC2E UMR 7328 CNRS, 45071 Orl\'{e}ans Cedex 02, France \\
-$^{23}$Astronomisches Institut der Ruhr-Universit\"{a}t Bochum, Universitaetsstrasse 150, 44780 Bochum, Germany \\
-$^{24}$Astrophysics, University of Oxford, Denys Wilkinson Building, Keble Road, Oxford OX1 3RH \\
-$^{25}$Astro Space Center of the Lebedev Physical Institute, Profsoyuznaya str. 84/32, Moscow 117997, Russia \\
-$^{26}$Astronomical Institute 'Anton Pannekoek', University of Amsterdam, Postbus 94249, 1090 GE Amsterdam, The Netherlands \\
-$^{27}$Center for Information Technology (CIT), University of Groningen, The Netherlands \\
-$^{28}$Centre de Recherche Astrophysique de Lyon, Observatoire de Lyon, 9 av Charles Andr\'{e}, 69561 Saint Genis Laval Cedex, France \\
-$^{29}$Centre for Radio Astronomy Techniques \& Technologies (RATT), Department of Physics and Elelctronics, Rhodes University, PO Box 94, Grahamstown 6140, South Africa \\
-$^{30}$SKA South Africa, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa \\
-$^{31}$Department of Astrophysics/IMAPP, Radboud University Nijmegen, P.O. Box 9010, 6500 GL Nijmegen, The Netherlands \\
-$^{32}$Argelander-Institut f\"{u}r Astronomie, University of Bonn, Auf dem H\"{u}gel 71, 53121, Bonn, Germany
+$^{15}$Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia \\
+$^{16}$Institute for Astronomy, University of Edinburgh, Royal Observatory of Edinburgh, Blackford Hill, Edinburgh EH9 3HJ, UK \\
+$^{17}$Leiden Observatory, Leiden University, PO Box 9513, 2300 RA Leiden, The Netherlands \\
+$^{18}$Leibniz-Institut f\"{u}r Astrophysik Potsdam (AIP), An der Sternwarte 16, 14482 Potsdam, Germany \\
+$^{19}$School of Physics and Astronomy, University of Southampton, Southampton, SO17 1BJ, UK \\
+$^{20}$University of Hamburg, Gojenbergsweg 112, 21029 Hamburg, Germany \\
+$^{21}$Th\"{u}ringer Landessternwarte, Sternwarte 5, D-07778 Tautenburg, Germany \\
+$^{22}$Laboratoire Lagrange, UMR7293, Universit\`{e} de Nice Sophia-Antipolis, CNRS, Observatoire de la C\'{o}te d'Azur, 06300 Nice, France \\
+$^{23}$Laboratoire de Physique et Chimie de l' Environnement et de l' Espace, LPC2E UMR 7328 CNRS, 45071 Orl\'{e}ans Cedex 02, France \\
+$^{24}$Astronomisches Institut der Ruhr-Universit\"{a}t Bochum, Universitaetsstrasse 150, 44780 Bochum, Germany \\
+$^{25}$Astrophysics, University of Oxford, Denys Wilkinson Building, Keble Road, Oxford OX1 3RH \\
+$^{26}$Astro Space Center of the Lebedev Physical Institute, Profsoyuznaya str. 84/32, Moscow 117997, Russia \\
+$^{27}$Astronomical Institute 'Anton Pannekoek', University of Amsterdam, Postbus 94249, 1090 GE Amsterdam, The Netherlands \\
+$^{28}$Center for Information Technology (CIT), University of Groningen, The Netherlands \\
+$^{29}$Centre de Recherche Astrophysique de Lyon, Observatoire de Lyon, 9 av Charles Andr\'{e}, 69561 Saint Genis Laval Cedex, France \\
+$^{30}$Centre for Radio Astronomy Techniques \& Technologies (RATT), Department of Physics and Elelctronics, Rhodes University, PO Box 94, Grahamstown 6140, South Africa \\
+$^{31}$SKA South Africa, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa \\
+$^{32}$Department of Astrophysics/IMAPP, Radboud University Nijmegen, P.O. Box 9010, 6500 GL Nijmegen, The Netherlands \\
+$^{33}$Argelander-Institut f\"{u}r Astronomie, University of Bonn, Auf dem H\"{u}gel 71, 53121, Bonn, Germany
 }

### 2013-03-01 articles/Article-RFI-distribution/pack.sh
Adding packing script

@@ -1,0 +1,1 @@
+tar -czvf rfi-distributions.tar.gz article-rfi-distribution.tex references.bib mn2e.bst mn2e.cls img/TF-L43786_L220-rfi-example.pdf img/transmitter-height/plot-transmitter-height-trimmed.pdf img/plot-rayleigh-and-rfi-trimmed.pdf img/plot-rayleigh-and-rfi-combined-trimmed.pdf img/explanation-lower-constraint/FindingCounts.pdf img/explanation-lower-constraint/FindingConstraints.pdf img/histograms-raw-trimmed.pdf img/plot-lba-dist-per-frequency-trimmed.pdf img/plot-hba-dist-per-frequency-trimmed.pdf img/histograms-corrected-trimmed.pdf img/histograms-with-fits-trimmed.pdf img/slopes-trimmed.pdf

### 2013-03-21 articles/Article-compression/compression.tex
First work on compression art

\documentclass{article} 
\usepackage[style=nature,backend=biber]{biblatex}
\usepackage{graphicx}
\usepackage[font={it},labelfont={bf}]{caption}
\usepackage{times}
\usepackage{color}
%\usepackage{fixltx2e} % fixes placing of image positions
\usepackage{authblk}

\definecolor{customhdrcolor}{rgb}{0.0,0.0,0.0}
%\definecolor{customcitecolor}{rgb}{0.0,0.25,0.25}
\definecolor{customcitecolor}{rgb}{0.0,0.5,0.75}
%\definecolor{customlinkcolor}{rgb}{0.0,0.0,1.0}
\definecolor{customlinkcolor}{rgb}{0.0,0.5,0.75}

\usepackage[colorlinks=true,linkcolor=customlinkcolor,urlcolor=customlinkcolor,citecolor=customcitecolor,pdftex]{hyperref}

\ifpdf\pdfinfo{/Title      (Compressing raw radio-astronomical data)
               /Author     (A. R. Offringa et al.)
               /Keywords   (compression)
        }
\else\usepackage{graphics}\fi

\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\addbibresource{references.bib}
\addbibresource{compressrefs.bib}

\title{Compression of interferometric radio-astronomical data}

\author[1,2]{A.~R.~Offringa}
\affil[1]{RSAA, Australian National University, Mt Stromlo Observatory, via Cotter Road, Weston, ACT 2611, Australia}
\affil[2]{ARC Centre of Excellence for All-Sky Astrophysics (CAASTRO)}
\begin{document}

\label{firstpage}
\maketitle

\begin{abstract}
% Articles have a summary, separate from the main text, of up to 150 words, which does not have references, and does not contain numbers, abbreviations, acronyms or measurements unless essential. It is aimed at readers outside the discipline. This summary contains a paragraph (2-3 sentences) of basic-level introduction to the field; a brief account of the background and rationale of the work; a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent); and finally, 2-3 sentences putting the main findings into general context so it is clear how the results described in the paper have moved the field forwards.
... TODO ...
\end{abstract}

\section{Background}
% up to 500 words of referenced text expanding on the background to the work (some overlap with the summary is acceptable), before proceeding to a concise, focused account of the findings, ending with one or two short paragraphs of discussion.
Unprocessed data from interferometric radio observatories consists of the correlated signals from all pairs of antennas that are part of the interferometric array. These correlated signals are measured in many spectral channels, and integrated over short timesteps before they are recorded to disk (TODO cite Thompson).

Modern interferometric radio observatories have very high spectral and temporal resolutions (TODO cite JVLA, LOFAR, ALMA, ATCA, MWA). It is often not desirable to decrease the resolution before archiving, because resolution in both dimensions provides scientific value. For example, high temporal resolutions are desirable for pulsar observations(TODO cite), variability studies(TODO cite) and advanced calibration methods that solve for the ionospheric disturbances and instrumental variabilities (TODO cite Sagecal). On the other hand, a high frequency resolution is necessary for widefield imaging of the sky (TODO cite), spectral-line studies (TODO cite) and the detection of man-made interference~\cite{lofar-radio-environment}. Therefore, the samples outputted by the correlator need to be archived at its original high resolution, which can lead to data volumes of several petabytes and complex network architectures to allow acceptable transfer rates (TODO cite). While current interferometers consist of up to a few hundred of antenna elements (TODO cite VLA, LOFAR), future instruments will have thousands of antenna elements that will produce data at even higher resolutions, thereby increasing the data volume and associated storage costs by several orders of magnitudes (TODO cite SKA).

Various multi-media and scientific applications have used and enhanced a variety of different compression techniques to decrease the storage costs. For audio compression, entropy-based encoding schemes (Huffman-encoding \cite{huffman}) and prediction schemes (TODO cite Chebinov polynomials) are important building blocks. Wavelet decomposition is another important building block for image and video compression algorithms (TODO cite). However, data from radio interferometric radio observatories have a low signal-to-noise ratio (SNR), especially at high resolutions (TODO cite Thompson), and noise-like signals are inherently hard to compress without loss of information (TODO cite).

% TODO describe compression in the context of space missions, as it is interesting

% TODO compression in the context of radio astronomy? E.g., PAPER's compression technique

\section{Methods}
We consider compressing correlator output samples that consist of complex values. For radio data, each complex component is normally stored as a 32-bit IEEE 754 single precision floating point value\footnote{See e.g. ``FITS Interferometry Data Interchange Format'', AIPS memo \#114, \S4.1;\\and ``MeasurementSet definition version 2.0'', CASA Memo \#229.}. We test two compression algorithms on recorded data from several telescopes. The first method is a loss-less encoding technique using a Huffman encoding \cite{huffman}, while the second method is a lossy encoding technique in the form of a non-linear minimized-squared-error quantization scheme (TODO cite?). Both encodings are optimized for normally distributed complex samples with zero mean. The distribution of correlated samples will approximate such a distribution more closely when the SNR is low (TODO cite) and does not contain interfering sources (TODO cite my own).

After we determined that loss-less compression is only effective for the 8-bit exponents of the floating point values, we avoid very large dictionary sizes by applying the Huffman encoding on the 8-bit exponents only. To overcome loss of compression rate due to integer bit counts required to represent symbols (TODO cite), the exponents of two consecutive samples are bitwise concatinated and encoded as such.

In the lossy case, the standard deviation is determined dynamically, and individually stored for each pair of antennas and at each correlator time-step. The lossy quantization scheme is created by uniform sampling of the inverse of the corresponding cummulative distribution function.

\section{Findings}

\begin{table}
 \caption{Compression performances. The compressed sizes are the resulting file size relative to the original size. The speed is measured relative to writing the data without compression.}
 \begin{tabular}{|l|r|r|r|}
  \hline
  Observation & Size & \multicolumn{4}{ |c| }{Huffman} & Lossy\\
  \hline
  MWA HydA & ..MB & 86\% & .. \\
  \hline
 \end{tabular}

\end{table}

\section{Discussion}

% TODO mention 16 quantization of VLA data

\printbibliography

\label{lastpage}

\end{document}

### 2013-03-22 articles/Article-compression/compression.tex
Work on paper

@@
 % up to 500 words of referenced text expanding on the background to the work (some overlap with the summary is acceptable), before proceeding to a concise, focused account of the findings, ending with one or two short paragraphs of discussion.
-Unprocessed data from interferometric radio observatories consists of the correlated signals from all pairs of antennas that are part of the interferometric array.
-These correlated signals are measured in many spectral channels, and integrated over short timesteps before they are recorded to disk (TODO cite Thompson).
+Unprocessed data from interferometric radio observatories consists of correlated signals from all the pairs of antennas that are part of the interferometric array.
+These correlated signals are measured in many spectral channels, and integrated over short timesteps before they are recorded to disk \cite{thompson-radio-interferometry}.
 
-Modern interferometric radio observatories have very high spectral and temporal resolutions (TODO cite JVLA, LOFAR, ALMA, ATCA, MWA).
+Correlators in modern interferometric radio observatories output data at high spectral and temporal resolutions \cite{evla-perley-2011,mwa-2013-tingay,atca-broadband-backend-2011} (TODO cite LOFAR).
 It is often not desirable to decrease the resolution before archiving, because resolution in both dimensions provides scientific value.
-For example, high temporal resolutions are desirable for pulsar observations(TODO cite), variability studies(TODO cite) and advanced calibration methods that solve for the ionospheric disturbances and instrumental variabilities (TODO cite Sagecal).
-On the other hand, a high frequency resolution is necessary for widefield imaging of the sky (TODO cite), spectral-line studies (TODO cite) and the detection of man-made interference~\cite{lofar-radio-environment}.
-Therefore, the samples outputted by the correlator need to be archived at its original high resolution, which can lead to data volumes of several petabytes and complex network architectures to allow acceptable transfer rates (TODO cite).
-While current interferometers consist of up to a few hundred of antenna elements (TODO cite VLA, LOFAR), future instruments will have thousands of antenna elements that will produce data at even higher resolutions, thereby increasing the data volume and associated storage costs by several orders of magnitudes (TODO cite SKA).
+For example, high temporal resolution is desirable for pulsar observations~\cite{ska-pulsars-2009}, variability studies~\cite{lofar-pulsars-and-transients-2011} and advanced calibration methods that solve for the ionospheric disturbances and instrumental variabilities \cite{kazemi-clustered-cal-2013,revisiting-me-ii}.
+On the other hand, high frequency resolution is necessary for widefield imaging of the sky~\cite{big-book-on-synthesis-imaging}, spectral-line studies~\cite{evla-high-resolution-lines}(TODO cite more?) and accurate removal of man-made interference~\cite{lofar-radio-environment}.
+This can lead to data volumes of several petabytes, which have to be supported by complex network architectures to allow acceptable transfer rates \cite{lofar-processing-arch-2005} (TODO cite more).
+While current interferometers consist of up to a few hundred of antenna elements \cite{evla-perley-2011,mwa-2013-tingay,atca-broadband-backend-2011} (TODO cite LOFAR), future instruments will have thousands of antenna elements that will produce data at even higher resolutions or with multiple beams, thereby increasing the data volume and associated storage costs by several orders of magnitudes \cite{apertif-2010} (TODO cite SKA, ESKAP(?)).
+For these reasons, this paper considers the compression of correlated interferometric data.
 
+% This might go...
 Various multi-media and scientific applications have used and enhanced a variety of different compression techniques to decrease the storage costs.
@@
 
+% TODO think about what this compression will do on HI observations(!)
+
 \section{Methods}
@@
 The first method is a loss-less encoding technique using a Huffman encoding \cite{huffman}, while the second method is a lossy encoding technique in the form of a non-linear minimized-squared-error quantization scheme (TODO cite?).
-Both encodings are optimized for normally distributed complex samples with zero mean.
+Both encodings are a-priory optimized for normally distributed complex samples with zero mean and a certain standard deviation.
 The distribution of correlated samples will approximate such a distribution more closely when the SNR is low (TODO cite) and does not contain interfering sources (TODO cite my own).
 
-After we determined that loss-less compression is only effective for the 8-bit exponents of the floating point values, we avoid very large dictionary sizes by applying the Huffman encoding on the 8-bit exponents only.
-To overcome loss of compression rate due to integer bit counts required to represent symbols (TODO cite), the exponents of two consecutive samples are bitwise concatinated and encoded as such.
+Loss-less compression is only effective for the 8-bit exponents of the floating point values, because the mantissa and sign are uniformly distributed and therefore incompressible without loss.
+We avoid very large dictionary sizes by applying the Huffman encoding on the 8-bit exponents only.
+To decrease loss of compression rate due to the integer number of bits required to represent symbols (TODO cite), the exponents of two consecutive samples are bitwise concatinated and encoded as such.
 
+% TODO Maybe we should do the same for the lossless encoding?
 In the lossy case, the standard deviation is determined dynamically, and individually stored for each pair of antennas and at each correlator time-step.
@@
 \begin{table}
- \caption{Compression performances. The compressed sizes are the resulting file size relative to the original size. The speed is measured relative to writing the data without compression.}
- \begin{tabular}{|l|r|r|r|}
+ \caption{Compression performances. The compressed sizes are the resulting file size relative to the original size. The speed is measured relative to reading/writing the data without compression.}
+ \begin{tabular}{|l|c|c|c|c|c|}
   \hline
-  Observation & Size & \multicolumn{4}{ |c| }{Huffman} & Lossy\\
+   & \multicolumn{2}{|c|}{Huffman} & \multicolumn{2}{|c|}{Lossy} \\
+  Observation &  \multicolumn{1}{|c}{Size} & \multicolumn{1}{c|}{Enc speed} & \multicolumn{1}{|c}{Size}&\multicolumn{1}{c|}{Enc speed} \\
   \hline
-  MWA HydA & ..MB & 86\% & .. \\
+  MWA HydA & 86\% & ..MB/s & 25\% & ..MB/s \\
   \hline

### 2013-03-28 articles/Article-RFI-distribution/article-rfi-distribution.tex
Fixing acknowledgement for Giulia Maca..
First work on compression art
Initial processing of referee comments

@@
 \pagerange{\pageref{firstpage}--\pageref{lastpage}}
-\pubyear{2012}
+\pubyear{2013}
 
@@
 \begin{abstract}
-Faint undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are stationary and consistently present over time.
+Faint undetected sources of radio-frequency interference (RFI) might become visible after long-time integration of radio observations if they are consistently present over time.
 Thereby, they might obstruct the detection of the weak astronomical signals of interest.
 This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe.
-We explore the low-frequency RFI situation by studying brightness histograms of visibility data observed with LOFAR, similar to radio-source-count analyses that are used in cosmology.
+We explore the RFI situation at 30--163~MHz by studying brightness histograms of visibility data observed with LOFAR, similar to radio-source-count analyses that are used in cosmology.
 The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
 With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models.
-Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources add up consistently.
+Extrapolation of the power law implies that the current LOFAR EoR observations should be severely RFI limited if the integrated strength of RFI sources are consistent, i.e., they are observed as coherent sources.
 This is in contrast with actual observations, which almost reach the thermal noise and are thought not to be limited by RFI.
-Therefore, we conclude that it is likely that most RFI sources do not linearly increase the noise.
-We believe that RFI will not prevent an EoR detection with LOFAR.
+Therefore, we conclude that it is unlikely that there are undetected RFI sources that will become visible after long-time integration.
+Consequently, it is likely that RFI will not prevent an EoR detection with LOFAR.
 \end{abstract}
@@
 However, astronomical radio observations can be affected by radio-frequency interference (RFI), which might make it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}.
-Many techniques have been designed to mitigate this effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, post-correlation-rfi-classification}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.
-In most cases, and with the help of careful spectrum allocation management and radio-quiet zones, these methods excise enough of the interference to allow calibration and imaging of the data.
+The careful management of spectrum allocation and the construction of radio-quiet zones help to limit the number of harmful transmitters.
+If harmful RFI is observed nevertheless, the use of RFI mitigation methods can sometimes clean the data sufficiently to allow succesful calibration and imaging.
+Many techniques have been designed to mitigate the effects of RFI, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, post-correlation-rfi-classification}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.
 
@@
 A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated.
-In such a case, weak interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be coherently present in the data.
+In such a case, weak interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be persistently present in the data.
 Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest.
@@
 This results in a better insight into the effects of low-level interference and allows one to simulate the effects of interference more accurately.
-A distribution can be expressed either as a differential or cumulative function.
-This work uses differential distributions, because they are better suited for visualizing the parameters we are interested in, although the two distributions basically convey the same information.
+% A distribution can be expressed either as the probability density function (PDF) or as cumulative density function (CDF).
+This work uses PDFs, because they are better suited for visualizing the parameters we are interested in, although the two distributions basically convey the same information.
 
@@
 This paper is organised as follows: in Sect.~\ref{sec:distribution-prediction}, we calculate a model for terrestrial interfering source distribution based on various assumptions.
-Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse the histograms.
+Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse brightness histograms of LOFAR data.
 This is followed in Sect.~\ref{sec:dist-data} by a short description of the two LOFAR data sets that have been used to perform the experiment.
@@
 \begin{center}
-\includegraphics[height=8cm]{img/TF-L43786_L220-rfi-example.pdf}
+\includegraphics[height=8cm]{img/TF-L43786_L220-rfi-example-xaxistext.pdf}
 \caption{A small part of an observation displayed in a time-frequency plot.
@@
 Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity.
-We will refer to this quantity with the statistical term ``frequency density'' (not to be confused with the physical radiation frequency).
-We will now start by estimating the frequency density function of ground-based interfering sources.
-Consider an interfering point source of strength $I$ measured in flux density.
+We will refer to this quantity with the term ``rate density''.
+We will now start by estimating the rate density function of ground-based interfering sources.
+Consider an interfering point source of strength $I$ that denotes the transmitting power normalized by the observational channel resolution (e.g., measured in W/Hz).
 This source is observed by an interferometer that consists of two antennas or stations with gains $g_1,g_2$, which include all instrumental effects.
@@
 \begin{equation}
- \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{r_1 r_2},
+ \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{4\pi r_1 r_2},
 \end{equation}
@@
 Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the narrow-band condition.
-For example, if the bandwidth of the signal $\Delta\nu=1$ kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c (2 \pi \Delta \nu)^{-1} \approx$ 50 km.
+With a frequency resolution $\Delta\nu=$ 0.76~kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c (2 \pi \Delta \nu)^{-1} \approx$ 50~km.
+Because the velocity resolution of LOFAR is 1.5~km/s at 150~MHz, and larger at lower frequencies, a Doppler frequency shift due to movement of the source will only be significant if its velocity is at least 1.5~km/s relative to the antennas.
+Since the relative velocities to different antennas in the array will be similar for such high-velocity transmitters (i.e., satellites), there will be hardly any decorrelation because of Doppler shifting.
 
@@
 \begin{equation} \label{eq:amplitude-fall-off-in-free-space}
- S(r) = \frac{Ig}{r^2}.
+ S(r) = \frac{Ig}{4\pi r^2}.
 \end{equation}
@@
 Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid.
-Using these assumptions, we can express the expected cumulative frequency density of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
+Using these assumptions, we can express the expected cumulative rate density of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
 \begin{equation} \label{eq:cumulative-distance-function}
@@
 
-We need the reverse of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance.
+We need the complement of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance.
 Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2$, with $N$ the total number of sources.
@@
  F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{R}(S))\\
-\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{S})\\
-\notag = & N-\frac{\rho \pi Ig}{S}
+\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{4\pi S})\\
+\notag = & N-\frac{\rho Ig}{4S}
 \end{align}
 where $\mathcal{R}(S)=\mathcal{S}^{-1}$, the inverse of $\mathcal{S}$, i.e., the function that returns the distance $r$ for a given amplitude $S$.
-Finally, the differential frequency density can be calculated by taking the derivative,
+Finally, the rate density can be calculated by taking the derivative,
 \begin{align} \label{eq:two-dimensional-distribution}
- f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{\rho \pi Ig}{S^2}.
+ f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{\rho Ig}{4S^2}.
 \end{align}
@@
 
-\subsection{Spherical case} \label{sec:spherical-geometry}
-If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes.
-In this case, the RFI sources are distributed over the surface of a sphere.
+%\subsection{Spherical case} \label{sec:spherical-geometry}
+%If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes.
 If we assume their difference in height is negligible, we can use the formula for the surface area of a spherical cap,
-\begin{equation}\label{eq:spherical-cap-area}
- A_\textrm{cap} = \pi \left( a^2 + h^2 \right) \hspace{0.4cm}\textrm{with } (a \le R) \wedge (h \le 2R),
-\end{equation}
-with $R$ the radius of the Earth, $a$ the radius of the base of the cap and $h$ the height of the cap.
+%\begin{equation}\label{eq:spherical-cap-area}
+% A_\textrm{cap} = \pi \left( a^2 + h^2 \right) \hspace{0.4cm}\textrm{with } (a \le R) \wedge (h \le 2R),
+%\end{equation}
+%with $R$ the radius of the Earth, $a$ the radius of the base of the cap and $h$ the height of the cap.
 Because the Euclidean distance $r$ between the source and the antenna is $r^2 = a^2 + h^2$, the resulting number of sources within Euclidean distance $r$ becomes
-\begin{equation} \label{eq:cumulative-spherical}
-\tilde{F}_{\textrm{distance} \le r}(r) = \rho A_\textrm{cap} = \rho \pi r^2 \hspace{0.4cm}\textrm{with } r \le 2R,
-\end{equation}
-which equals the formula for a planar source distribution given in Eq.~\eqref{eq:cumulative-distance-function}.
+%\begin{equation} \label{eq:cumulative-spherical}
+%\tilde{F}_{\textrm{distance} \le r}(r) = \rho A_\textrm{cap} = \rho \pi r^2 \hspace{0.4cm}\textrm{with } r \le 2R,
+%\end{equation}
+%which equals the formula for a planar source distribution given in Eq.~\eqref{eq:cumulative-distance-function}.
 Therefore, the spherical and two-dimensional cases lead to a distribution with equal power law.
@@
 A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}.
-The original model is based on urban areas, but corrections are given by \citeauthor{hata-propagation-loss} for sub-urban areas with a lower population density and for open areas.
-\citeauthor{hata-propagation-loss} gives the following analytical estimate for $L_p$, the electromagnetic propagation loss over land:
+The original model is based on urban areas, but corrections are given by Hata for sub-urban areas with a lower population density and for open areas.
+Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss over land:
 \begin{align} 
@@
 \begin{equation}
- S(r) = \frac{Ig}{\zeta r^\eta},
+ S(r) = \frac{Ig}{4\pi \zeta r^\eta},
 \end{equation}
-with $\eta = 4.49 - 0.655 \log_{10} h_b$, one finds the frequency density function $f_p$ that considers propagation effects,
+with $\eta = 4.49 - 0.655 \log_{10} h_b$, one finds the rate density function $f_p$ that considers propagation effects,
 % S(r):=I*g/(r^alpha);
@@
 f_p(S)
-= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{Ig }{\zeta S} \right)^{2/\eta} \right]
-= \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{\zeta S} \right)^{2/\eta}.
+= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{Ig }{4\pi\zeta S} \right)^{2/\eta} \right]
+= \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{4\pi\zeta S} \right)^{2/\eta}.
 \end{align}
@@
 
-To understand the distribution of samples consisting of RFI + noise, one needs to consider the individual contributions of the real and imaginary components.
-The real and imaginary components of samples that are affected by RFI, will follow a multi-variate distribution: each complex component is the sum of a complex component of RFI and a noise sample.
-The noise comes from a Gaussian distribution with zero mean.
-The RFI distribution is assumed to be a power-law distribution with yet unknown exponent.
-
 To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial.
 The histogram can be numerically estimated by drawing complex samples from the two distributions and calculating and counting the amplitudes.
-A sample can be drawn from the RFI distribution by scaling and inverting the cumulative frequency function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable.
+A sample can be drawn from the RFI distribution by scaling and inverting the cumulative rate density function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable.
 Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity.
@@
 \begin{equation} \label{eq:drawing-rfi-distribution}
-S_{\textrm{RFI}} \leftarrow \frac{Ig}{x_u r_L^2} e^{i 2 \pi y_u},
+S_{\textrm{RFI}} \leftarrow \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
 \end{equation}
-with $x_u, y_u \sim U(0, 1)$.
+where $x_u$ and $y_u$ are independently drawn from a uniform distribution between 0 and 1.
 The result is an amplitude sample $S_{\textrm{RFI}}$, with $S_{\textrm{RFI}} \ge Ig/r_L^2$, that follows a power-law distribution.
@@
 \begin{equation} \label{eq:drawing-combined-distribution}
-S \leftarrow v_n + w_n i + \frac{Ig}{x_u r_L^2} e^{i 2 \pi y_u},
+S \leftarrow v_n + w_n i + \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
 \end{equation}
@@
 For completeness we show the formula for sampling apparent source strengths that include propagation effects according to Hata's model, which can be used for sampling realistic source amplitudes in simulations.
-This can be derived by integration, scaling and inversion of the frequency density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
+This can be derived by integration, scaling and inversion of the rate density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
 \begin{equation} \label{eq:sample-with-propagation}
@@
 This is mandated by the large dynamic range of the histogram that we are interested in.
-Therefore, we chose to have a bin size that increases linearly with the amplitude $S$, and the frequency counts are divided by the bin size after counting.
+Therefore, we chose to have a bin size that increases linearly with the amplitude $S$, and the rate counts are divided by the bin size after counting.
 Consequently, in parts of the histogram that have a sparse number of samples, the outlying samples will follow a $1/S$ curve, or a -1 slope in a log-log plot.
@@
 Moreover, the estimator requires reading and processing all data points, and because of the size of the data sets this is somewhat impractical.
-However, we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ is the bin frequency.
+However, we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ is the bin rate.
 Because our bin size is small and the distribution is well sampled, the resulting estimate will be close to the normal Hill estimator.
@@
 Once the exponent of the power-law part of the distribution is estimated with the previously discussed techniques, the distribution can be extrapolated to find a lower flux limit.
-Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the following frequency density function $h$:
+Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the following rate density function $h$:
 \begin{equation} \label{eq:amplitude-power-law}
@@
 First and most importantly, the samples themselves are digitized with an analogue-to-digital converter (ADC) with limited range.
-Second, we observe for a limited time and the frequency count is discrete.
+Second, we observe for a limited time and the rate count is discrete.
 Because the chance of finding a sample with a very high intensity is low, it is unlikely to observe samples beyond some amplitude within the finite observing time.
@@
 
-%\begin{figure}[tbp]
-%\begin{center}
-%\includegraphics[width=12cm]{img/plot-slope-over-frequency-trimmed}
-%\caption{Slope of the log-log histogram slope over frequency in the LBA set.}
-%\label{fig:plot-slope-over-frequency-LBA}
-%\end{center}
-%\end{figure}
-
-%\begin{figure}[tbp]
-%\begin{center}
-%\includegraphics[width=12cm]{img/slope-over-frequency-HBA/plot-slope-over-frequency}
-%\caption{Slope of the log-log histogram over frequency in the HBA set.}
-%\label{fig:plot-slope-over-frequency-HBA}
-%\end{center}
-%\end{figure}
-
-%\begin{figure}[tbp]
-%\begin{center}
-%\includegraphics[width=12cm]{img/slope-over-baseline-HBA/plot-slope-over-baseline-wstddev}
-%\caption{Slope of the log-log histogram over the baseline length for the HBA set. A single sub-band at 150 MHz was used. Error bars indicate $5\sigma$ level.}
-%\label{fig:plot-slope-over-baseline-HBA}
-%\end{center}
-%\end{figure}
-
 \begin{table*}
@@
 Ferrari and G.
-Maccario acknowledge financial support by the ``Agence Nationale de la Recherche'' through grant ANR-09-JCJC-0001-01.
+Macario acknowledge financial support by the ``Agence Nationale de la Recherche'' through grant ANR-09-JCJC-0001-01.
 

### 2013-03-28 articles/Article-compression/compression.tex
Some work on the compression article

@@
 This can lead to data volumes of several petabytes, which have to be supported by complex network architectures to allow acceptable transfer rates \cite{lofar-processing-arch-2005} (TODO cite more).
-While current interferometers consist of up to a few hundred of antenna elements \cite{evla-perley-2011,mwa-2013-tingay,atca-broadband-backend-2011} (TODO cite LOFAR), future instruments will have thousands of antenna elements that will produce data at even higher resolutions or with multiple beams, thereby increasing the data volume and associated storage costs by several orders of magnitudes \cite{apertif-2010} (TODO cite SKA, ESKAP(?)).
+While current interferometers consist of up to a few hundred of antenna elements \cite{evla-perley-2011,mwa-2013-tingay,atca-broadband-backend-2011} (TODO cite LOFAR), future instruments will have thousands of antenna elements that will produce data at even higher resolutions or with multiple beams, thereby increasing the data volume and associated storage costs by several orders of magnitudes \cite{apertif-2010,ska-station-config-2013} (TODO cite ESKAP(?)).
 For these reasons, this paper considers the compression of correlated interferometric data.
 
-% This might go...
-Various multi-media and scientific applications have used and enhanced a variety of different compression techniques to decrease the storage costs.
-For audio compression, entropy-based encoding schemes (Huffman-encoding \cite{huffman}) and prediction schemes (TODO cite Chebinov polynomials) are important building blocks.
-Wavelet decomposition is another important building block for image and video compression algorithms (TODO cite).
-However, data from radio interferometric radio observatories have a low signal-to-noise ratio (SNR), especially at high resolutions (TODO cite Thompson), and noise-like signals are inherently hard to compress without loss of information (TODO cite).
+Various multi-media and scientific applications have applied a variety of different compression techniques to decrease storage or transmission costs.
+Linear quantization can be effective for compression of noisy astronomical image \cite{astro-image-compression-2010}.
+Entropy-based encoding schemes such as Huffman encoding \cite{huffman} and prediction schemes (TODO cite Chebyshev polynomials) can optimize the representation of data.
+Decomposition into basis functions can provide a way to remove the less important bits in the data, such as the wavelet decomposition of audio data to remove tones that are nearly inaudible \cite{wavelet-audio-compression-1993}.
+In radio astronomy, linear quantization of the captured antenna voltages prior to correlation is common and well understood \cite{quantization-efficiency-2007}.
 
@@
 
+% TODO raw data compression is different than image compression in the way that we don't care about the error in individual pixels; the total error is important!
+
 \section{Methods}
 We consider compressing correlator output samples that consist of complex values.
-For radio data, each complex component is normally stored as a 32-bit IEEE 754 single precision floating point value\footnote{See e.g.
+These data commonly have a low signal-to-noise ratio (SNR), especially at high resolutions \cite{thompson-radio-interferometry}, and noise-like signals are inherently hard to compress without loss of information \cite{shannon-entropy-definition-1949}.
+In radio data, each complex component is normally stored as a 32-bit IEEE 754 single precision floating point value\footnote{See e.g.
 ``FITS Interferometry Data Interchange Format'', AIPS memo \#114, \S4.1;\\and ``MeasurementSet definition version 2.0'', CASA Memo \#229.}.
 We test two compression algorithms on recorded data from several telescopes.
-The first method is a loss-less encoding technique using a Huffman encoding \cite{huffman}, while the second method is a lossy encoding technique in the form of a non-linear minimized-squared-error quantization scheme (TODO cite?).
+The first method is a loss-less technique using Huffman encoding \cite{huffman}, while the second method is a lossy encoding in the form of a non-linear minimized-squared-error quantization scheme \cite{least-squares-quantization-2006}.
 Both encodings are a-priory optimized for normally distributed complex samples with zero mean and a certain standard deviation.
@@
 
-Loss-less compression is only effective for the 8-bit exponents of the floating point values, because the mantissa and sign are uniformly distributed and therefore incompressible without loss.
+Loss-less compression is only effective for the 8-bit exponents of the floating point values, because the mantissa and sign follow closely a uniform distribution and are therefore incompressible without loss \cite{shannon-entropy-definition-1949}.
 We avoid very large dictionary sizes by applying the Huffman encoding on the 8-bit exponents only.

### 2013-04-01 articles/Article-RFI-distribution/article-rfi-distribution.tex
Work on ref report

@@
 \begin{abstract}
-Faint undetected sources of radio-frequency interference (RFI) might become visible after long-time integration of radio observations if they are consistently present over time.
+Faint undetected sources of radio-frequency interference (RFI) might become visible after long time-integration of radio observations if they are consistently present over time.
 Thereby, they might obstruct the detection of the weak astronomical signals of interest.

### 2013-04-02 articles/Article-RFI-distribution/article-rfi-distribution.tex
Work on referee comments

@@
 
-We need the complement of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance.
+We need $F_{\textrm{distance} \ge r}(r)$, the complement of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance.
 Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2$, with $N$ the total number of sources.
@@
 A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}.
-The original model is based on urban areas, but corrections are given by Hata for sub-urban areas with a lower population density and for open areas.
-Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss over land:
+The original model is based on urban areas, but corrections are given by Hata for suburban areas with a lower population density and for open areas.
+Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss between two ground-based antennas:
 \begin{align} 
@@
 \begin{equation} \label{eq:propagation-loss-factor}
- L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
+ L_S = \frac{1}{10} 10^{L_p} = \zeta r^{4.49 - 0.655 \log_{10} h_b},
 \end{equation}
@@
 \end{equation}
-with $\eta = 4.49 - 0.655 \log_{10} h_b$, one finds the rate density function $f_p$ that considers propagation effects,
+with $\eta = 4.49 - 0.655 \log_{10} h_b$ as seen in Eq.~\eqref{eq:propagation-loss-factor}, one finds the rate density function $f_p$ that considers propagation effects,
 % S(r):=I*g/(r^alpha);
@@
 The power-law distribution (Eq.~\eqref{eq:two-dimensional-distribution}) has a constant slope of -2.
+Its integral is not equal to 1 in this example, because some form of flattening would be required to make this possible.
 The slope of the Rayleigh distribution in the limit of the origin is 1.
@@
 \end{equation}
-where $x_u$ and $y_u$ are independently drawn from a uniform distribution between 0 and 1.
+where $0 < x_u, y_u \le 1$ are independently drawn from a uniform distribution.
 The result is an amplitude sample $S_{\textrm{RFI}}$, with $S_{\textrm{RFI}} \ge Ig/r_L^2$, that follows a power-law distribution.
@@
 If we assume that $I$, the average source strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-combined-distribution} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram.
-Although calibration could in theory solve $g$, almost all sources will come in through the far-sidelobes of the phased-array beam, and finding the expected values for the gains is therefore hard.
+Although calibration could in theory solve $g$, almost all RFI sources will come in through the far-sidelobes of the beam, and finding the expected values for the gains is therefore hard.
 The effects of these parameters on the histogram will be further discussed in \S\ref{sec:parameter-variability}.
 
-Fig.~\ref{fig:rayleigh-and-rfi-combined} shows that the left side of the graph corresponds with the original Rayleigh distribution.
+%Fig.~\ref{fig:rayleigh-and-rfi-combined} shows that the left side of the graph corresponds with the original Rayleigh distribution.
 From the right side of the histogram, we are able to estimate $Ig/r_L^2$, as the placing of the power-law curve is independent of the Rayleigh distribution for large $S$.
@@
 \end{equation}
-Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 \le x_u,y_u < 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.
+Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 < x_u,y_u \le 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.
 
@@
 In reality, the parameters $\rho$, $I$ and $g$, which are the source density per unit area, source strength and instrumental gain respectively, will not be constant, but can change over time and frequency.
-Therefore, they act like stochastic variables.
+Therefore, they are stochastic variables.
 However, since each specific value for these parameters produces a power law, the combined distribution will still show a power law, as long as the parameters follow a distribution that is steep at high amplitudes (in $\log$--$\log$ space), such as a Gaussian or uniform distribution.
@@
 Because the data that are analysed in Sect.~\ref{sec:dist-results} have initially not been band-pass calibrated, the instrumental response is not uniform over frequency.
-We determined that the variation due to the band-pass is about one order of magnitude for the low-band antennas (LBA, 30.1--77.5 MHz) and about a factor of two for the high-band antennas (HBA, 115.0--163.3 MHz).
-
-The effect of the band-pass on the data distribution is consequently limited to one order of magnitude or less.
-If the RFI sources have an apparent power-law distribution with a certain exponent, the distribution will have a feature at low amplitudes due to the frequency response, but a power law with equal exponent will still appear when the distribution is observed over a wide enough range.
+We determined that the gain variation due to the band-pass is about one order of magnitude for the low-band antennas (LBA, 30.1--77.5 MHz) and about a factor of two for the high-band antennas (HBA, 115.0--163.3 MHz).
+The frequency dependency of the gains due to the band-pass will consequently smooth the data in the brightness histogram in horizontal direction by one order of magnitude or less.
 
@@
 
-One should note that fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}.
+Fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}.
 However, because of our enormous sample size, which allows fitting the line over a large interval, the estimator will be sufficiently accurate for our purpose.
-Nevertheless, we will additionally calculate a maximum likelihood estimator for comparison.
+Nevertheless, we will additionally calculate a maximum-likelihood estimator for comparison.
 The maximum-likelihood estimator for the exponent in a power-law distribution is given by the Hill estimator $\hat \alpha_H$ \citep{hill-estimator, power-law-distribution}, defined as:
@@
 Therefore, using this estimator will result in an estimate that is steeper (i.e., more negative) than the actual distribution.
-Moreover, the estimator requires reading and processing all data points, and because of the size of the data sets this is somewhat impractical.
-However, we can calculate Eq.~\eqref{eq:hill} by adding the bin centre values $N$ times to the calculation, where $N$ is the bin rate.
-Because our bin size is small and the distribution is well sampled, the resulting estimate will be close to the normal Hill estimator.
-This avoids the need to iterate over all data.
 
@@
 
-$S_1$ and $S_2$ are selected by visually inspection of the histogram.
+$S_1$ and $S_2$ are selected by visual inspection of the histogram.
 Assume that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}, and that the effect of the Rayleigh component on the histogram $> S_1$ is negligible.
@@
 \end{equation}
-In some cases there will be no solution for $S_U$.
-This situation arises because the integral of the distribution function is finite even when $S_U \rightarrow \infty$.
-If $N_\textrm{part}$ exceeds the integrated value of the fitted distribution function, there is no solution.
-This can happen when the empirical distribution is not limited at the high end, or when it contains features that are not in the model.
+In some cases there will be no solution for $S_U$. This can happen when the empirical distribution is not limited at the high end, or when it contains features that are not in the model.
 
@@
 With estimates of $\alpha$, $\beta$, $S_L$ and $S_U$, one has obtained a parametrization of the RFI distribution.
-As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is $Ig/r^2$ -- assuming free space propagation for the moment -- and therefore $S_L = Ig/r^2$.
-For Hata's propagation model, the more generic solution $S_L = Ig / (\zeta r^\alpha)$ is found.
+As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is $Ig/(4\pi r^2)$ -- assuming free space propagation for the moment -- and therefore $S_L = Ig/(4\pi r^2)$.
+For Hata's propagation model, the more generic solution $S_L = Ig / (4\pi \zeta r^\alpha)$ is found.
 This value represents the apparent brightness of the sources that are the furthest away from the telescope.
@@
 Since modern RFI detection algorithms can find all RFI that is detectable ``by eye'' \citep{post-correlation-rfi-classification}, this power level will be near the level of the noise mode.
+We use the AOFlagger for RFI detection, which will be described in Sect.~\ref{sec:dist-data}.
 In
@@
 the false-positives rate for the AOFlagger is estimated to be 0.5\%.
+
 An estimate of $S_d$, the average lower limit of detected RFI, can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the real number (true positives) of RFI samples.
-Therefore, the limit is calculated similar to Eq.~\eqref{eq:lower-limit-1}, where the term $N_\textrm{part} - N_\textrm{total}$ needs to be replaced with $N_\textrm{RFI}$, which equals the total number of samples detected as RFI minus the 0.5\% false positives.
+Therefore, the limit is calculated similar to Eq.~\eqref{eq:lower-limit-1}, where the term $N_\textrm{part} - N_\textrm{total}$ needs to be replaced with $N_\textrm{RFI}$, which equals the total number of samples detected as RFI minus the false positives.
 
@@
 
-To detect RFI, the AOFlagger \citep{LOFAR-RFI-pipeline, scale-invariant-rank-operator} was used.
-However, since this flagger is partly amplitude-based, it is likely that low-level RFI will leak through the detector.
+To detect RFI, the AOFlagger \citep{LOFAR-RFI-pipeline} is used.
+This RFI detector estimates the sky contribution by iteratively applying a high-pass filter to the visibility amplitudes of a single baseline in the time-frequency plane.
+Subsequently, it flags line-shaped features with the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}.
+Finally, the scale-invariant rank operator, a morphological technique to search for contaminated samples, is applied on the two-dimensional flag mask \citep{scale-invariant-rank-operator}.
+
+Because the AOFlagger detector is partly amplitude-based, it is likely that low-level RFI will leak through the detector.
 Since it is also low-level RFI we are interested in, we will analyse unflagged data and the RFI classified data.

### 2013-04-05 cotter/fitsuser.h
Some work on fitswriter that I did during 3GC3

@@ -6,8 +6,8 @@
 class FitsUser
 {
 	protected:
-		void checkStatus(int status);
-		void throwError(int status, const std::string &msg = std::string());
+		static void checkStatus(int status);
+		static void throwError(int status, const std::string &msg = std::string());
 };
 
 #endif

### 2013-04-05 cotter/fitswriter.cpp

@@ -1,21 +1,68 @@
 #include "fitswriter.h"
 
+#include <cstdio>
+
 FitsWriter::FitsWriter(const char* filename)
 {
+	/** If the file already exists, remove it */
+	FILE *fp = std::fopen(filename,"r");
+  if (fp!=NULL) {
+    std::fclose(fp);
+    std::remove(filename);
+  }
+  
 	int status = 0;
-	if(fits_open_file(&_fptr, filename, READONLY, &status))
+  if(fits_create_file(&_fptr, filename, &status))
 		throwError(status, std::string("Cannot open file ") + filename);
 }
 
 FitsWriter::~FitsWriter()
 {
-	int status;
+	int status = 0;
 	if(fits_close_file(_fptr, &status))
 		throwError(status, std::string("Cannot close file "));
 }
 
+void FitsWriter::initGroupHeader()
+{
+	int status = 0;
+  /* Write the group header and initialize random groups.
+	 * The data are grouped in visibilities that belong to the same time step and
+	 * baseline. Within a group, the visibilities are sorted as 
+	 * (real, imag, weight) x nrpolarizations (XX,XY,..) x frequency channel
+	 * The preamble consists of:
+	 * - U, V, W in nanoseconds
+	 * - baseline
+	 * - time offset in days
+	 * This call will set the keywords
+	 * - PCOUNT=5
+	 * - GROUPS=T
+	 * - GCOUNT=nr baselines x time step count
+	 */
+	const unsigned NAXIS = 6;
+	long naxes[NAXIS];
+  naxes[0] = 0;
+  naxes[1] = 3;  // real, imaginary, weight
+  naxes[2] = 4;  // polarization count
+  naxes[3] = _bandInfo.channels.size();
+  naxes[4] = 1;
+  naxes[5] = 1;
+	const unsigned nGroupParams = 5; // u,v,w,baseline,time
+	const unsigned nRows = 0;
+  fits_write_grphdr(_fptr, TRUE, FLOAT_IMG, NAXIS, naxes, nGroupParams,
+										nRows, TRUE, &status);
+	checkStatus(status);
+	
+  setKeywordToFloat("BSCALE", 1.0);
+}
+
 void FitsWriter::WriteBandInfo(const std::string& name, const std::vector<ChannelInfo>& channels, double refFreq, double totalBandwidth, bool flagRow)
 {
+	_bandInfo.name = name;
+	_bandInfo.channels = channels;
+	_bandInfo.refFreq = refFreq;
[…]

### 2013-04-05 cotter/fitswriter.h

@@ -28,14 +28,39 @@ class FitsWriter : public Writer, private FitsUser
 		virtual void WriteRow(double time, double timeCentroid, size_t antenna1, size_t antenna2, double u, double v, double w, double interval, const std::complex<float>* data, const bool* flags, const float *weights);
 		
 	private:
-		void writeDataDescEntry(size_t spectralWindowId, size_t polarizationId, bool flagRow);
-		void writeFeedEntries(const std::vector<Writer::AntennaInfo>& antennae, double time);
+		void initGroupHeader();
+		void setKeywordToDouble(const char *keywordName, double value) const
+		{
+			int status = 0;
+			if(fits_update_key(_fptr, TDOUBLE, keywordName, &value, NULL, &status))
+				throwError(status, "Could not write keyword");
+		}
+		void setKeywordToFloat(const char *keywordName, float value) const
+		{
+			int status = 0;
+			if(fits_update_key(_fptr, TFLOAT, keywordName, &value, NULL, &status))
+				throwError(status, "Could not write keyword");
+		}
+		void setKeywordToString(const char *keywordName, const char *value) const
+		{
+			int status = 0;
+			if(fits_update_key(_fptr, TFLOAT, keywordName, const_cast<char*>(value), NULL, &status))
+				throwError(status, "Could not write keyword");
+		}
+		void setKeywordToString(const char *keywordName, const std::string &value) const
+		{
+			setKeywordToString(keywordName, value.c_str());
+		}
 		
-		class MSWriterData *_data;
-		size_t _rowIndex;
-		
-		size_t _nChannels;
 		fitsfile *_fptr;
+		
+		struct {
+			std::string name;
+			std::vector<ChannelInfo> channels;
+			double refFreq;
+			double totalBandwidth;
+			bool flagRow;
+		} _bandInfo;
 };
 
 #endif

### 2013-04-05 articles/Article-compression/compression.tex
Work on compr art

--- 2013-03-22 articles/Article-compression/compression.tex
@@
 Unprocessed data from interferometric radio observatories consists of correlated signals from all the pairs of antennas that are part of the interferometric array.
-These correlated signals are measured in many spectral channels, and integrated over short timesteps before they are recorded to disk \cite{thompson-radio-interferometry}.
+These correlated signals are measured in spectral channels, and integrated over short timesteps before they are recorded to disk \cite{thompson-radio-interferometry}.
 

### 2013-04-05 articles/Article-RFI-distribution/article-rfi-distribution.tex

@@
 \begin{abstract}
-Faint undetected sources of radio-frequency interference (RFI) might become visible after long time-integration of radio observations if they are consistently present over time.
+Faint undetected sources of radio-frequency interference (RFI) might become visible after long integrations of radio observations if they are consistently present over time.
 Thereby, they might obstruct the detection of the weak astronomical signals of interest.
@@
 With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models.
-Extrapolation of the power law implies that the current LOFAR EoR observations should be severely RFI limited if the integrated strength of RFI sources are consistent, i.e., they are observed as coherent sources.
+Extrapolation of the power law implies that the current LOFAR EoR observations should be severely RFI limited if the strength of RFI sources remain strong after integration, i.e., they are observed as coherent sources.
 This is in contrast with actual observations, which almost reach the thermal noise and are thought not to be limited by RFI.
@@
 
-In a time-frequency plot, interfering sources can have complex structures.
+In time-frequency space, interfering sources can have complex structures.
 They can also be intermittent and different sources might overlap in time-frequency space.
-An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a time-frequency diagram (dynamic spectrum).
-Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in the time-frequency diagram.
+An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a dynamic spectrum.
+Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in time-frequency space.
 It is often not clear what constitutes a single interfering source, hence it is hard to count individual sources.
@@
 \includegraphics[height=8cm]{img/TF-L43786_L220-rfi-example-xaxistext.pdf}
-\caption{A small part of an observation displayed in a time-frequency plot.
+\caption{A small part of an observation displayed in a dynamic spectrum.
 The features with significantly higher values are caused by interference.
@@
 We can assign flux densities to the horizontal axis of the histogram by using the system equivalent flux density (SEFD) of a single station.
-The current LOFAR SEFD is found to be approximately 3000~Jy for the HBA core stations and 1500~Jy for the remote stations in the frequency range from $125$--$175$ MHz.
-For all Dutch LBA stations, in the frequency range $40$--$70$ MHz the SEFD is approximately 30,000~Jy.
+The current LOFAR SEFD is found to be approximately 3400~Jy for the HBA core stations and 1500~Jy for the remote stations in the frequency range from $125$--$175$ MHz.
+For all Dutch LBA stations, in the frequency range $40$--$70$ MHz the SEFD is approximately 34,000~Jy.
 The standard deviation $\sigma$ in the real and imaginary values is related to the SEFD with % TODO citation
@@
 \section{Data description} \label{sec:dist-data}
+% TODO correct reference if van Haarlem et al is published
 We have analysed the distributions of two data sets. Both data sets are 24-h LOFAR RFI surveys and are extensively analysed in
-\citet{lofar-radio-environment}. We refer to van Haarlem et al. (2013, \textit{submitted}) for a full description of the capabilities of LOFAR.
-% TODO correct reference if van Haarlem et al is published
-In one set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz.
+\citet{lofar-radio-environment}.
+We refer to van Haarlem et al.
+(2013, \textit{submitted}) for a full description of the capabilities of LOFAR.
+The analyses will cover only Dutch stations.
+Each Dutch station consists of 96 dipole low-band antennas (LBA) and one or two fields totalling 48~tiles of 4x4 bow-tie high-band antennas (HBA).
+The core area of LOFAR is located near the village of Exloo in the Netherlands, where the station density is at its highest.
+The six most densely packed stations are on the Superterp, an elevated area surrounded by water situated 3~km North of Exloo.
+A radio-quiet zone of 2~km around the Superterp has been established, but is relatively small and households exist within 1~km of the Superterp.
+
+In the two data sets, we have used the correlation coefficients of cross-correlated stations, i.e., the raw visibilities.
+In one data set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz.
 More stations were used in the LBA set.
@@
 Field               & NCP             & NCP \\
-Amount RFI detected & 1.77\%       & 3.18\% \\
+Amount of RFI detected & & \\
+by the AOFlagger    & 1.77\%       & 3.18\% \\
 \hline
@@
 %$\epsilon_R$ & Error in Rayleigh part & TODO & \\
-$E(S_R)$ & Apparent RFI flux density & $2700$ Jy & $140$ Jy \\
+$E(S_R)$ & Apparent RFI flux density & $2,700$ Jy & $140$ Jy \\
 $E(S_\textrm{leak})$& Residual apparent RFI flux density after excision & $484$--$496$ mJy & $167$--$171$ mJy \\
@@
 REFD& RFI equivalent flux density & $18.9$--$19.3$ Jy & $6.5$--$6.7$ Jy \\
+\hline
+\multicolumn{4}{l}{\textit{Average station temperatures}} \\
+\hline
+$T_\textrm{sys}$ & System temperature (in clean bands)  & 5,000 K & 640 K \\
+$T_\textrm{R}$ & Temperature of RFI & 17,000 K & 1,200 K \\
+$T_\textrm{leak}$ & Temperature of undetected RFI & 3.2 K & 1.4 K \\
 \hline
@@
 Taken at face value these exponents imply in Hata's model that the average transmitting heights for sources affecting the LBA and HBA are $79$ and $13$~m respectively.
-There are no 79 m high transmitters nearby LOFAR stations in the LBA frequency range.
-Hata's model only goes down to 150 MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies, e.g.
+There are no 79~m high transmitters nearby LOFAR stations in the LBA frequency range.
+Additionally, Hata's model only goes down to 150~MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies, e.g.
 because the effect of the ionosphere becomes stronger.
-Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.53;-1.49]$ for the HBA.
+Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.57;-1.49]$ for the HBA.
 The estimate of the HBA is thus more accurate, because its histogram deviates less from the power law.
+This translates to boundaries on the average transmitter height of $[0.6; 800]$ and $[3.1; 23]$~m for the LBA and HBA respectively.
+Therefore, the LBA measurements are clearly not accurate enough to be conclusive.
+Moreover, because the power-law distribution analyses involve many assumptions, it is uncertain whether the analyses are sufficiently accurate for making these detailed conclusions.
 
@@
 If in fact not all samples are affected by RFI, the leaked RFI flux will be smaller, and will of course be zero in the extreme case that the detector has found and removed all RFI.
+
+\citet{lf-interference-temperature-rogers} TODO
 
@@
 \begin{equation}
- \sigma_\textrm{eor-night} = \textrm{SEFD} \left(2\Delta t \Delta \nu N_\textrm{feed} N_\textrm{interferometers} \right)^{-\frac{1}{2}} \approx 250\hspace{1mm} \mu\textrm{Jy}.
+ \sigma_\textrm{eor-night} = \textrm{SEFD} \left(2\Delta t \Delta \nu N_\textrm{feed} N_\textrm{interferometers} \right)^{-\frac{1}{2}} \approx 250\hspace{1mm} \mu\textrm{Jy},
 \end{equation}
+where $N_\textrm{feed}=2$ is the number of feeds per antenna and $N_\textrm{interferometers}=\frac{1}{2}60\times59$ is the number of interferometers.
 Therefore, after 100 nights the thermal noise level will be $25\hspace{1mm}\mu\textrm{Jy}$.
@@
 It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright foreground sources.
-Preliminary analysis of EoR NCP observations of a single night have reached the thermal noise \citep{ncp-eor-yatawatta}, but do not show leaked RFI at the pole.
+Preliminary analysis of EoR NCP observations of a single night have almost reached the thermal noise \citep{ncp-eor-yatawatta}, but do not show leaked RFI at the pole.
 

### 2013-04-05 articles/Article-RFI-distribution/article-rfi-distribution.tex
Todays work on the paper

@@
 % TODO correct reference if van Haarlem et al is published
+\nocite{lofar-overview-2013}
 
@@
 \caption{The Rayleigh and power-law distributions in a log-log plot.
-The power-law distribution (Eq.~\eqref{eq:two-dimensional-distribution}) has a constant slope of -2.
-Its integral is not equal to 1 in this example, because some form of flattening would be required to make this possible.
+The power-law distribution (Eq.~\eqref{eq:two-dimensional-distribution}) has a constant slope of -2 over the range it is defined.
 The slope of the Rayleigh distribution in the limit of the origin is 1.
@@
 It is likely that strong transmitters transmit more on average, and therefore contaminate more samples as well.
-High-power transmitters, such as radio stations, have a typical transmitting power in the order of 100 kW.
-Low-power transmitters, such as remote controls, transmit with an order of 100 mW or even less.
+High-power transmitters, such as radio stations, have a typical transmitting power in the order of 100~kW.
+Low-power transmitters, such as remote controls, transmit with an order of 100~mW or even less.
 Therefore, these devices have a spread of around 6 orders of magnitudes in power.
@@
 $T_\textrm{sys}$ & System temperature (in clean bands)  & 5,000 K & 640 K \\
-$T_\textrm{R}$ & Temperature of RFI & 17,000 K & 1,200 K \\
+$T_\textrm{R}$ & RFI Temperature & 17,000 K & 1,200 K \\
 $T_\textrm{leak}$ & Temperature of undetected RFI & 3.2 K & 1.4 K \\
@@
 
-\citet{lf-interference-temperature-rogers} TODO
+The apparent RFI flux densities can be converted to a RFI station temperature that excludes the system noise and sky noise components.
+If we use a station efficiency factor $\eta_\textrm{st}=1$ and effective areas LBA $A_\textrm{eff}=398$ and HBA $A_\textrm{eff}=512$ with again 100\% RFI occupancy, our models lead to RFI temperatures of 17,000~K and 1,200~K for respectively the LBA and the HBA.
+These are relatively high compared to for example the survey by \citet{lf-interference-temperature-rogers}, who report that on two different sites, 20\% and 27\% of the spectrum has a temperature above 450~K in the range of 50--1500~MHz.
+However, our post-detection RFI station temperatures, which arise from the residual apparent RFI flux density estimates, are 3.2~K and 1.4~K for the LBA and HBA respectively.
+Due to LOFAR's high resolution and accurate flagging strategy, this is achieved by flagging a relatively small data percentage of 1.8 (LBA) and 3.2\% (HBA).
 

### 2013-04-09 cotter/cotter.cpp
Fixing cotter for 128T

@@ -314,17 +314,12 @@ void Cotter::Run(const char *outputFilename, size_t timeAvgFactor, size_t freqAv
 void Cotter::createReader(const std::vector< std::string >& curFileset)
 {
 	delete _reader; // might be 0, but that's ok.
-	_reader = new GPUFileReader();
+	_reader = new GPUFileReader(_mwaConfig.NAntennae(), _mwaConfig.Header().nChannels);
 
 	for(std::vector<std::string>::const_iterator i=curFileset.begin(); i!=curFileset.end(); ++i)
 		_reader->AddFile(i->c_str());
 	
 	_reader->Initialize();
-	
-	if(_reader->AntennaCount() != _mwaConfig.NAntennae())
-		throw std::runtime_error("nAntennae in GPU files do not match nAntennae in config");
-	if(_reader->ChannelCount() != _mwaConfig.Header().nChannels)
-		throw std::runtime_error("nChannels in GPU files do not match nChannels in config");
 }
 
 void Cotter::initializeReader()

### 2013-04-09 cotter/gpufilereader.cpp

@@ -10,8 +10,15 @@ void GPUFileReader::openFiles()
 	int status = 0;
 	for(size_t i=0; i!=_filenames.size(); ++i)
 	{
-		fitsfile *fptr;
-		if(!fits_open_file(&fptr, _filenames[i].c_str(), READONLY, &status))
+		const std::string &curFilename = _filenames[i];
+		fitsfile *fptr = 0;
+		bool hasStartTime = false;
+		if(curFilename.empty())
+		{
+			std::cout << "(Skipping unavailable file)\n";
+			_fitsFiles.push_back(0);
+		}
+		else if(!fits_open_file(&fptr, curFilename.c_str(), READONLY, &status))
 		{
 			_fitsFiles.push_back(fptr);
 			
@@ -26,7 +33,11 @@ void GPUFileReader::openFiles()
 			fits_read_key(fptr, TLONG, "TIME", &thisFileTime, 0, &status);
 			checkStatus(status);
 			
-			if(i == 0) _startTime = thisFileTime;
+			if(!hasStartTime) 
+			{
+				_startTime = thisFileTime;
+				hasStartTime = true;
+			}
 			if(_startTime != thisFileTime)
 			{
 				std::cout << "WARNING: file number " << (i+1) << " of current time range has different start time!\n"
@@ -47,17 +58,17 @@ void GPUFileReader::closeFiles()
 	{
 		int status = 0;
 		fitsfile *fptr = _fitsFiles[i];
-		fits_close_file(fptr, &status);
-		checkStatus(status);
+		if(fptr != 0)
+		{
+			fits_close_file(fptr, &status);
+			checkStatus(status);
+		}
 	}
 	_fitsFiles.clear();
 	_isOpen = false;
 }
 
 bool GPUFileReader::Read(size_t &bufferPos, size_t count) {
-	_nAntenna = 32;
-	_nChannelsInTotal = 1536*2;
-	
 	const size_t nPol = 4;
 	const size_t nBaselines = (_nAntenna + 1) * _nAntenna / 2;
 	const size_t gpuMatrixSizePerFile = _nChannelsInTotal * nBaselines * nPol / _filenames.size(); // cuda matrix length per file
@@ -83,85 +94,90 @@ bool GPUFileReader::Read(size_t &bufferPos, size_t count) {
 
 			fitsfile *fptr = _fitsFiles[iFile];
 
-			int status = 0, hduType = 0;
-			fits_movabs_hdu(fptr, _currentHDU, &hduType, &status);
[…]

### 2013-04-09 cotter/gpufilereader.h

@@ -33,7 +33,7 @@
 class GPUFileReader : private FitsUser
 {
 	public:
-		GPUFileReader() : _isOpen(false), _nAntenna(32), _nChannelsInTotal(3072), _bufferSize(0) { }
+		GPUFileReader(size_t nAntenna, size_t nChannelsInTotal) : _isOpen(false), _nAntenna(nAntenna), _nChannelsInTotal(nChannelsInTotal), _bufferSize(0) { }
 		~GPUFileReader() { closeFiles(); }
 		
 		void AddFile(const char *filename) { _filenames.push_back(std::string(filename)); }
@@ -69,7 +69,8 @@ class GPUFileReader : private FitsUser
 		}
 		std::time_t StartTime() const { return _startTime; }
 	private:
-		const static int pfb_output_to_input[64];
+		const static int single_pfb_output_to_input[64];
+		std::vector<int> pfb_output_to_input;
 		
 		GPUFileReader(const GPUFileReader &) { }
 		void operator=(const GPUFileReader &) { }
@@ -77,6 +78,7 @@ class GPUFileReader : private FitsUser
 		void closeFiles();
 		void findStopHDU();
 		void initMapping();
+		void initializePFBMapping();
 		BaselineBuffer &getBuffer(size_t antenna1, size_t antenna2)
 		{
 			return _buffers[_nAntenna*antenna1 + antenna2];

### 2013-04-09 cotter/main.cpp

@@ -151,8 +151,7 @@ int main(int argc, char **argv)
 		{
 			if(i >= fileSets[j].size() || fileSets[j][i].empty()) {
 				std::ostringstream errstr;
-				errstr << "Unable to correctly order the raw GPU input files: missing information from GPU box " << (j+1) << ", timerange " << i << ". Maybe you are missing an input file?";
-				throw std::runtime_error(errstr.str());
+				std::cout << "Missing information from GPU box " << i << ", timerange " << j << ". Maybe you are missing an input file?\n";
 			}
 		}
 	}

### 2013-04-09 articles/Article-RFI-distribution/article-rfi-distribution.tex
Article fixes

@@
 \setlength{\pdfpagewidth}{\paperwidth}
+
+\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
 
@@
 \begin{abstract}
-Faint undetected sources of radio-frequency interference (RFI) might become visible after long integrations of radio observations if they are consistently present over time.
+Faint undetected sources of radio-frequency interference (RFI) might become visible after \editmark{long integrations} of radio observations if they \editmark{are consistently} present over time.
 Thereby, they might obstruct the detection of the weak astronomical signals of interest.
 This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe.
-We explore the RFI situation at 30--163~MHz by studying brightness histograms of visibility data observed with LOFAR, similar to radio-source-count analyses that are used in cosmology.
+We explore the RFI situation \editmark{at 30--163~MHz} by studying brightness histograms of visibility data observed with LOFAR, similar to radio-source-count analyses that are used in cosmology.
 The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
 With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models.
-Extrapolation of the power law implies that the current LOFAR EoR observations should be severely RFI limited if the strength of RFI sources remain strong after integration, i.e., they are observed as coherent sources.
+Extrapolation of the power law \editmark{implies} that the current LOFAR EoR observations should be severely RFI limited if the \editmark{strength of RFI sources remain strong after integration}.
 This is in contrast with actual observations, which almost reach the thermal noise and are thought not to be limited by RFI.
-Therefore, we conclude that it is unlikely that there are undetected RFI sources that will become visible after long-time integration.
-Consequently, it is likely that RFI will not prevent an EoR detection with LOFAR.
+Therefore, we conclude that it is \editmark{unlikely that there are undetected RFI sources that will become visible after long integration.
+Consequently, it is likely} that RFI will not prevent an EoR detection with LOFAR.
 \end{abstract}
@@
 However, astronomical radio observations can be affected by radio-frequency interference (RFI), which might make it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}.
-The careful management of spectrum allocation and the construction of radio-quiet zones help to limit the number of harmful transmitters.
-If harmful RFI is observed nevertheless, the use of RFI mitigation methods can sometimes clean the data sufficiently to allow succesful calibration and imaging.
-Many techniques have been designed to mitigate the effects of RFI, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, post-correlation-rfi-classification}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.
+\editmark{The careful management of spectrum allocation and the construction of radio-quiet zones help to limit the number of harmful transmitters.
+If harmful RFI is observed nevertheless, the use of RFI mitigation methods can sometimes clean the data sufficiently to allow succesful calibration and imaging.} Many techniques have been designed to mitigate the effects of RFI, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, post-correlation-rfi-classification}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.
 
@@
 A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated.
-In such a case, weak interference caused by stationary sources might not manifest itself above the noise in individual observations, but might be persistently present in the data.
+In such a case, weak interference caused by stationary \editmark{RFI sources} might not manifest itself above the noise in individual observations, but might be \editmark{persistently} present in the data.
 Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest.
@@
 For these experiments, it is important to know the possible effect of low-level interference on the data, as these might overshadow or alter the signal of interest.
-In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, van Haarlem et al.
-2013, \textit{submitted}).
+
+In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, \citealt{lofar-overview-2013}, \textit{submitted}).
+\editmark{The LOFAR EoR project aims to detect the redshifted 21-cm signals from the EoR using the LOFAR HBA antennas (115--190~MHz, $z_\textrm{HI}$=11.4--6.5).
+Several fields will be observed over 100~nights, to achieve sufficient sensitivity to allow the signal extraction.
+An EoR calibration pipeline has been designed that solves for ionospheric and instrumental effects in approximately hundred directions using the SAGE algorithm \citep{sage-calibration-ii}.
+Initial results from commissioning observations show that in a single night the thermal noise level can almost be reached \citep{ncp-eor-yatawatta}.}
 % TODO correct reference if van Haarlem et al is published
-\nocite{lofar-overview-2013}
 
@@
 This results in a better insight into the effects of low-level interference and allows one to simulate the effects of interference more accurately.
-% A distribution can be expressed either as the probability density function (PDF) or as cumulative density function (CDF).
-This work uses PDFs, because they are better suited for visualizing the parameters we are interested in, although the two distributions basically convey the same information.
 
@@
 This paper is organised as follows: in Sect.~\ref{sec:distribution-prediction}, we calculate a model for terrestrial interfering source distribution based on various assumptions.
-Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse brightness histograms of LOFAR data.
+Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse \editmark{brightness histograms of LOFAR data}.
 This is followed in Sect.~\ref{sec:dist-data} by a short description of the two LOFAR data sets that have been used to perform the experiment.
@@
 
-In time-frequency space, interfering sources can have complex structures.
-They can also be intermittent and different sources might overlap in time-frequency space.
-An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a dynamic spectrum.
-Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in time-frequency space.
+In \editmark{time-frequency space}, interfering sources can have complex structures.
+They can also be intermittent and different sources might overlap in \editmark{time-frequency space}.
+An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a \editmark{dynamic spectrum}.
+Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in \editmark{time-frequency space}.
 It is often not clear what constitutes a single interfering source, hence it is hard to count individual sources.
@@
 \includegraphics[height=8cm]{img/TF-L43786_L220-rfi-example-xaxistext.pdf}
-\caption{A small part of an observation displayed in a dynamic spectrum.
+\caption{A \editmark{dynamic spectrum} of a small part of an observation.
 The features with significantly higher values are caused by interference.
@@
 Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity.
-We will refer to this quantity with the term ``rate density''.
+We will refer to this quantity with the term \editmark{``rate density''}.
 We will now start by estimating the rate density function of ground-based interfering sources.
-Consider an interfering point source of strength $I$ that denotes the transmitting power normalized by the observational channel resolution (e.g., measured in W/Hz).
+Consider an interfering point source of strength $I$ that denotes \editmark{the transmitting power normalized by the observational channel resolution (e.g., measured in W/Hz)}.
 This source is observed by an interferometer that consists of two antennas or stations with gains $g_1,g_2$, which include all instrumental effects.
@@
 The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
-\begin{equation}
+\editmark{\begin{equation}
  \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{4\pi r_1 r_2},
-\end{equation}
+\end{equation}}
 with (real-valued amplitude) gains $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$.
@@
 Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the narrow-band condition.
-With a frequency resolution $\Delta\nu=$ 0.76~kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c (2 \pi \Delta \nu)^{-1} \approx$ 50~km.
-Because the velocity resolution of LOFAR is 1.5~km/s at 150~MHz, and larger at lower frequencies, a Doppler frequency shift due to movement of the source will only be significant if its velocity is at least 1.5~km/s relative to the antennas.
-Since the relative velocities to different antennas in the array will be similar for such high-velocity transmitters (i.e., satellites), there will be hardly any decorrelation because of Doppler shifting.
+\editmark{With a frequency resolution $\Delta\nu=$ 0.76~kHz}, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c (2 \pi \Delta \nu)^{-1} \approx$ 50~\editmark{km}.
+\editmark{Because the velocity resolution of LOFAR is 1.5~km/s at 150~MHz, and larger at lower frequencies, a Doppler frequency shift due to movement of the source will only be significant if its velocity is at least 1.5~km/s relative to the antennas.
+Since the relative velocities towards different antennas in the array will be similar for such high-velocity transmitters (i.e., satellites), there will be hardly any decorrelation because of Doppler shifting.}
 
@@
 Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and interferometric gain $g = g_1 g_2$ as
-\begin{equation} \label{eq:amplitude-fall-off-in-free-space}
+\editmark{\begin{equation} \label{eq:amplitude-fall-off-in-free-space}
  S(r) = \frac{Ig}{4\pi r^2}.
-\end{equation}
+\end{equation}}
 
@@
 Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid.
-Using these assumptions, we can express the expected cumulative rate density of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
+Using these assumptions, we can express the expected cumulative \editmark{rate density} of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
 \begin{equation} \label{eq:cumulative-distance-function}
@@
 
-We need $F_{\textrm{distance} \ge r}(r)$, the complement of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance.
+We need \editmark{$F_{\textrm{distance} \ge r}(r)$, the complement} of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance.
 Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2$, with $N$ the total number of sources.
@@
 The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can now be calculated with
-\begin{align} \label{eq:cumulative-rfi-distribution}
+\editmark{\begin{align} \label{eq:cumulative-rfi-distribution}
  F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{R}(S))\\
@@
 \notag = & N-\frac{\rho Ig}{4S}
-\end{align}
+\end{align}}
 where $\mathcal{R}(S)=\mathcal{S}^{-1}$, the inverse of $\mathcal{S}$, i.e., the function that returns the distance $r$ for a given amplitude $S$.
-Finally, the rate density can be calculated by taking the derivative,
-\begin{align} \label{eq:two-dimensional-distribution}
+Finally, the \editmark{rate density} can be calculated by taking the derivative,
+\editmark{\begin{align} \label{eq:two-dimensional-distribution}
  f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{\rho Ig}{4S^2}.
-\end{align}
+\end{align}}
 
@@
 The original model is based on urban areas, but corrections are given by Hata for suburban areas with a lower population density and for open areas.
-Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss between two ground-based antennas:
+Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss \editmark{between two ground-based antennas}:
 \begin{align} 
@@
 Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
-\begin{equation} \label{eq:propagation-loss-factor}
+\editmark{\begin{equation} \label{eq:propagation-loss-factor}
  L_S = \frac{1}{10} 10^{L_p} = \zeta r^{4.49 - 0.655 \log_{10} h_b},
-\end{equation}
+\end{equation}}
 with
@@
 Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
-\begin{equation}
+\editmark{\begin{equation}
  S(r) = \frac{Ig}{4\pi \zeta r^\eta},
-\end{equation}
+\end{equation}}
 with $\eta = 4.49 - 0.655 \log_{10} h_b$ as seen in Eq.~\eqref{eq:propagation-loss-factor}, one finds the rate density function $f_p$ that considers propagation effects,
@@
 % diff(rho*%pi*r^2,S),solve(S(r)=S,r);
-\begin{align} \label{eq:frequency-density-with-propagation}
+\editmark{\begin{align} \label{eq:frequency-density-with-propagation}
 f_p(S)
@@
 = \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{4\pi\zeta S} \right)^{2/\eta}.
-\end{align}
+\end{align}}
 Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
@@
 The full measured distribution will consist of the power-law distribution combined with that of the thermal noise and the celestial signal.
-For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong sources.
-For example, the strongest apparent source in the NCP EoR field is around 5 Jy (Yattawata et al.
-2013, \textit{in press}).
-% TODO fill in details.
+For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong \editmark{celestial} sources.
+For example, the strongest apparent \editmark{celestial} source in the NCP EoR field is around 5 Jy \citep{ncp-eor-yatawatta}.
 The standard deviation of the noise, however, is around 100~Jy on highest LOFAR resolutions, and will have a larger contribution on the histogram.
@@
 \caption{The Rayleigh and power-law distributions in a log-log plot.
-The power-law distribution (Eq.~\eqref{eq:two-dimensional-distribution}) has a constant slope of -2 over the range it is defined.
+The power-law distribution (Eq.~\eqref{eq:two-dimensional-distribution}) has a constant slope of -2 \editmark{over the range it is defined}.
 The slope of the Rayleigh distribution in the limit of the origin is 1.
@@
 The histogram can be numerically estimated by drawing complex samples from the two distributions and calculating and counting the amplitudes.
-A sample can be drawn from the RFI distribution by scaling and inverting the cumulative rate density function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable.
-Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of sources will go to infinity.
-Therefore, one needs to assume that there are no samples beyond some limiting distance $r_L$ to invert the cumulative function.
+A sample can be drawn from the RFI distribution by scaling and inverting the cumulative \editmark{rate density} function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable.
+Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of \editmark{RFI sources} will go to infinity.
+Therefore, one needs to assume that there are no \editmark{sources} beyond some limiting distance $r_L$ to invert the cumulative function.
 With this assumption, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
-\begin{equation} \label{eq:drawing-rfi-distribution}
+\editmark{\begin{equation} \label{eq:drawing-rfi-distribution}
 S_{\textrm{RFI}} \leftarrow \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
-\end{equation}
-where $0 < x_u, y_u \le 1$ are independently drawn from a uniform distribution.
-The result is an amplitude sample $S_{\textrm{RFI}}$, with $S_{\textrm{RFI}} \ge Ig/r_L^2$, that follows a power-law distribution.
+\end{equation}}
+\editmark{where $0 < x_u, y_u \le 1$ are independently drawn from a uniform distribution}.
+The result is an amplitude sample $S_{\textrm{RFI}}$, with \editmark{$S_{\textrm{RFI}} \ge Ig/4\pi r_L^2$}, that follows a power-law distribution.
 The corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
-\begin{equation} \label{eq:drawing-combined-distribution}
+\editmark{\begin{equation} \label{eq:drawing-combined-distribution}
 S \leftarrow v_n + w_n i + \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
-\end{equation}
+\end{equation}}
 with $v_n, w_n \sim N(\mu=0; \sigma)$. The final amplitude sample $S_A$ can be calculated with 
@@
 \end{equation}
-An example of distribution curves for various settings of $Ig/r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.
+An example of distribution curves for various settings of \editmark{$Ig/4\pi r_L^2$} is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.
 
-If we assume that $I$, the average source strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-combined-distribution} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram.
-Although calibration could in theory solve $g$, almost all RFI sources will come in through the far-sidelobes of the beam, and finding the expected values for the gains is therefore hard.
+If we assume that $I$, the average \editmark{RFI source} strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-combined-distribution} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram.
+Although calibration could in theory solve $g$, almost all \editmark{RFI} sources will come in through the far-sidelobes of \editmark{the beam}, and finding the expected values for the gains is therefore hard.
 The effects of these parameters on the histogram will be further discussed in \S\ref{sec:parameter-variability}.
@@
 For completeness we show the formula for sampling apparent source strengths that include propagation effects according to Hata's model, which can be used for sampling realistic source amplitudes in simulations.
-This can be derived by integration, scaling and inversion of the rate density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
+This can be derived by integration, scaling and inversion of the \editmark{rate density} function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
 \begin{equation} \label{eq:sample-with-propagation}
@@
 \subsection{Parameter variability} \label{sec:parameter-variability}
-In reality, the parameters $\rho$, $I$ and $g$, which are the source density per unit area, source strength and instrumental gain respectively, will not be constant, but can change over time and frequency.
-Therefore, they are stochastic variables.
+In reality, the parameters $\rho$, $I$ and $g$, which are the RFI source density per unit area, \editmark{RFI source} strength and instrumental gain respectively, will not be constant, but can change over time and frequency.
+Therefore, \editmark{they are} stochastic variables.
 However, since each specific value for these parameters produces a power law, the combined distribution will still show a power law, as long as the parameters follow a distribution that is steep at high amplitudes (in $\log$--$\log$ space), such as a Gaussian or uniform distribution.
@@
 We determined that the gain variation due to the band-pass is about one order of magnitude for the low-band antennas (LBA, 30.1--77.5 MHz) and about a factor of two for the high-band antennas (HBA, 115.0--163.3 MHz).
-The frequency dependency of the gains due to the band-pass will consequently smooth the data in the brightness histogram in horizontal direction by one order of magnitude or less.
+\editmark{The frequency dependency of the gains due to the band-pass will consequently smooth the data in the brightness histogram in horizontal direction by one order of magnitude or less.}
 
@@
 This is mandated by the large dynamic range of the histogram that we are interested in.
-Therefore, we chose to have a bin size that increases linearly with the amplitude $S$, and the rate counts are divided by the bin size after counting.
+Therefore, we chose to have a bin size that increases linearly with the amplitude $S$, and the \editmark{rate counts} are divided by the bin size after counting.
 Consequently, in parts of the histogram that have a sparse number of samples, the outlying samples will follow a $1/S$ curve, or a -1 slope in a log-log plot.
@@
 
-Fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}.
+\editmark{Fitting} straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}.
 However, because of our enormous sample size, which allows fitting the line over a large interval, the estimator will be sufficiently accurate for our purpose.
@@
 Once the exponent of the power-law part of the distribution is estimated with the previously discussed techniques, the distribution can be extrapolated to find a lower flux limit.
-Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the following rate density function $h$:
+Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the following \editmark{rate density} function $h$:
 \begin{equation} \label{eq:amplitude-power-law}
@@
 Areas in a log-log plot are not linearly related to the integral.
-There are two ways to estimate the lower constraint $S_L$: (i) the areas $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal if $Ig/r^2$ is constant during the observation, and (ii) if one assumes $Ig/r^2\sim$ uniform, then $N_a+N_b = N_\textrm{total}-N_\textrm{part}$.}
+There are two ways to estimate the lower constraint $S_L$: (i) the areas $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal if \editmark{$Ig/r_L^\eta$} is constant during the observation, and (ii) if one assumes \editmark{$Ig/r_L^\eta\sim$} uniform, then $N_a+N_b = N_\textrm{total}-N_\textrm{part}$.}
 \label{fig:explanation-constraints}
@@
 
-$S_1$ and $S_2$ are selected by visual inspection of the histogram.
+$S_1$ and $S_2$ are selected by \editmark{visual} inspection of the histogram.
 Assume that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}, and that the effect of the Rayleigh component on the histogram $> S_1$ is negligible.
@@
 First and most importantly, the samples themselves are digitized with an analogue-to-digital converter (ADC) with limited range.
-Second, we observe for a limited time and the rate count is discrete.
+Second, we observe for a limited time and the \editmark{rate count} is discrete.
 Because the chance of finding a sample with a very high intensity is low, it is unlikely to observe samples beyond some amplitude within the finite observing time.
@@
 \end{equation}
-In some cases there will be no solution for $S_U$. This can happen when the empirical distribution is not limited at the high end, or when it contains features that are not in the model.
+In some cases there will be no solution for $S_U$. \editmark{This can happen when the empirical distribution is not limited at the high end, or when it contains features that are not in the model.}
 
@@
 
-With the assumption that $Ig/r_L^2\sim$ a uniform distribution, the area labelled in Fig.~\ref{fig:explanation-constraints} as $N_b$ is also part of the RFI distribution, and a stronger constraint $\tilde S_L$ can be found by solving
+With the assumption that \editmark{$Ig/r_L^\eta\sim$} a uniform distribution, the area labelled in Fig.~\ref{fig:explanation-constraints} as $N_b$ is also part of the RFI distribution, and a stronger constraint $\tilde S_L$ can be found by solving
 \begin{equation}
@@
 With estimates of $\alpha$, $\beta$, $S_L$ and $S_U$, one has obtained a parametrization of the RFI distribution.
-As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is $Ig/(4\pi r^2)$ -- assuming free space propagation for the moment -- and therefore $S_L = Ig/(4\pi r^2)$.
-For Hata's propagation model, the more generic solution $S_L = Ig / (4\pi \zeta r^\alpha)$ is found.
-This value represents the apparent brightness of the sources that are the furthest away from the telescope.
+As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is \editmark{$Ig/4\pi r^2$} -- assuming free space propagation for the moment -- and therefore \editmark{$S_L = Ig/4\pi r^2$}.
+For Hata's propagation model, the more generic solution \editmark{$S_L = Ig / 4\pi \zeta r^\eta$} is found.
+This value represents the apparent brightness of the \editmark{RFI sources} that are furthest away from the telescope.
 With a fully parametrized distribution of the effect of RFI sources, an empirical model for RFI effects can be made.
@@
 Since modern RFI detection algorithms can find all RFI that is detectable ``by eye'' \citep{post-correlation-rfi-classification}, this power level will be near the level of the noise mode.
-We use the AOFlagger for RFI detection, which will be described in Sect.~\ref{sec:dist-data}.
-In
+\editmark{We use the AOFlagger for RFI detection, which will be described in Sect.~\ref{sec:dist-data}.} In
 \citet{lofar-radio-environment}
@@
 We can assign flux densities to the horizontal axis of the histogram by using the system equivalent flux density (SEFD) of a single station.
-The current LOFAR SEFD is found to be approximately 3400~Jy for the HBA core stations and 1500~Jy for the remote stations in the frequency range from $125$--$175$ MHz.
-For all Dutch LBA stations, in the frequency range $40$--$70$ MHz the SEFD is approximately 34,000~Jy.
+The current LOFAR SEFD is found to be approximately \editmark{3400~Jy} for the HBA core stations and \editmark{1700~Jy} for the remote stations in the frequency range from $125$--$175$ MHz.
+For all Dutch LBA stations, in the frequency range $40$--$70$ MHz the SEFD is approximately \editmark{34,000~Jy}.
 The standard deviation $\sigma$ in the real and imaginary values is related to the SEFD with % TODO citation
@@
 (2013, \textit{submitted}) for a full description of the capabilities of LOFAR.
-The analyses will cover only Dutch stations.
+\editmark{The analyses will cover only Dutch stations.
 Each Dutch station consists of 96 dipole low-band antennas (LBA) and one or two fields totalling 48~tiles of 4x4 bow-tie high-band antennas (HBA).
@@
 The six most densely packed stations are on the Superterp, an elevated area surrounded by water situated 3~km North of Exloo.
-A radio-quiet zone of 2~km around the Superterp has been established, but is relatively small and households exist within 1~km of the Superterp.
+A radio-quiet zone of 2~km around the Superterp has been established, but is relatively small and households exist within 1~km of the Superterp.}
 
-In the two data sets, we have used the correlation coefficients of cross-correlated stations, i.e., the raw visibilities.
-In one data set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz.
+\editmark{In the two data sets, we have used the correlation coefficients of cross-correlated stations, i.e., the raw visibilities.} In one data set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz.
 More stations were used in the LBA set.
@@
 To detect RFI, the AOFlagger \citep{LOFAR-RFI-pipeline} is used.
-This RFI detector estimates the sky contribution by iteratively applying a high-pass filter to the visibility amplitudes of a single baseline in the time-frequency plane.
+\editmark{This RFI detector estimates the sky contribution by iteratively applying a high-pass filter to the visibility amplitudes of a single baseline in the time-frequency plane.
 Subsequently, it flags line-shaped features with the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}.
-Finally, the scale-invariant rank operator, a morphological technique to search for contaminated samples, is applied on the two-dimensional flag mask \citep{scale-invariant-rank-operator}.
+Finally, the scale-invariant rank operator, a morphological technique to search for contaminated samples, is applied on the two-dimensional flag mask \citep{scale-invariant-rank-operator}.}
 
@@
 Field               & NCP             & NCP \\
-Amount of RFI detected & & \\
-by the AOFlagger    & 1.77\%       & 3.18\% \\
+\editmark{Amount of RFI detected} & & \\
+\editmark{by the AOFlagger}    & 1.77\%       & 3.18\% \\
 \hline
@@
 
-It is to be expected that the RFI-dominated part of the distributions at different frequencies will reflect the underlying source populations.
+It is to be expected that the RFI-dominated part of the distributions at different frequencies will reflect the underlying \editmark{RFI source} populations.
 Both Figs.~\ref{fig:plot-dist-per-frequency-LBA} and \ref{fig:plot-dist-per-frequency-HBA} show that the power-law part of the distributions are very different for different sub-bands.
@@
 This validates that the variable gain that is caused by the antenna response does not change the observed power law.
-Consequently, it can be expected that other stochastic effects, such as the intrinsic source strength and the beam gain due to a differing direction of arrival, will similarly not affect the power law.
+Consequently, it can be expected that other stochastic effects, such as the intrinsic \editmark{RFI source} strength and the beam gain due to a differing direction of arrival, will similarly not affect the power law.
 Because the band-pass corrected histograms should provide a more accurate analysis, we will use the corrected histograms for further analysis.
@@
 Although the HBA slope does not show any other significant features besides the Rayleigh and power-law curves, the LBA power law ends with a bulge around an amplitude of $10^6$.
-This bulge is caused by a very strong source affecting lots of samples, and is a single outlier in the spatial distribution.
+This bulge is caused by a very strong \editmark{RFI} source affecting lots of samples, and is a single outlier in the spatial distribution.
 We found this is caused by RFI observed for about an hour in the late afternoon in the lower LBA frequency regime, around $30$--$40$~MHz.
@@
 $S_L$ & Constraint on lower fall-off point of power law & $21$ mJy & $6.2$ mJy \\ % $10^{-6.42}$ & $10^{-4.31}$
-$\tilde{S}_L$ & As $S_L$, but assuming $Ig/r^2\sim$ uniform & $47$ mJy & $14$ mJy \\ % $10^{-6.08}$ & $10^{-3.96}$
+$\tilde{S}_L$ & As $S_L$, but assuming \editmark{$Ig/r^\eta\sim$} uniform & $47$ mJy & $14$ mJy \\ % $10^{-6.08}$ & $10^{-3.96}$
 $S_d$ & Expected lowest apparent level of RFI detected & $26$ Jy & $5.7$ Jy\\ % $2.3 \times 10^{-2}$ & $4.5 \times 10^{-2}$
@@
 \hline
-\multicolumn{4}{l}{\textit{Average station temperatures}} \\
+\multicolumn{4}{l}{\editmark{\textit{Average station temperatures}}} \\
 \hline
@@
 because the effect of the ionosphere becomes stronger.
-Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.57;-1.49]$ for the HBA.
+Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and \editmark{$[-1.57;-1.49]$} for the HBA.
 The estimate of the HBA is thus more accurate, because its histogram deviates less from the power law.
-This translates to boundaries on the average transmitter height of $[0.6; 800]$ and $[3.1; 23]$~m for the LBA and HBA respectively.
+\editmark{This translates to boundaries on the average transmitter height of $[0.6; 800]$ and $[3.1; 23]$~m for the LBA and HBA respectively.
 Therefore, the LBA measurements are clearly not accurate enough to be conclusive.
 Moreover, because the power-law distribution analyses involve many assumptions, it is uncertain whether the analyses are sufficiently accurate for making these detailed conclusions.
+}
 
@@
 
-The apparent RFI flux densities can be converted to a RFI station temperature that excludes the system noise and sky noise components.
+\editmark{The apparent RFI flux densities can be converted to a RFI station temperature that excludes the system noise and sky noise components.
 If we use a station efficiency factor $\eta_\textrm{st}=1$ and effective areas LBA $A_\textrm{eff}=398$ and HBA $A_\textrm{eff}=512$ with again 100\% RFI occupancy, our models lead to RFI temperatures of 17,000~K and 1,200~K for respectively the LBA and the HBA.
@@
 However, our post-detection RFI station temperatures, which arise from the residual apparent RFI flux density estimates, are 3.2~K and 1.4~K for the LBA and HBA respectively.
-Due to LOFAR's high resolution and accurate flagging strategy, this is achieved by flagging a relatively small data percentage of 1.8 (LBA) and 3.2\% (HBA).
+Due to LOFAR's high resolution and accurate flagging strategy, this is achieved by flagging a relatively small data percentage of 1.8 (LBA) and 3.2\% (HBA).}
 
@@
 \end{equation}
-where $N_\textrm{feed}=2$ is the number of feeds per antenna and $N_\textrm{interferometers}=\frac{1}{2}60\times59$ is the number of interferometers.
-Therefore, after 100 nights the thermal noise level will be $25\hspace{1mm}\mu\textrm{Jy}$.
+\editmark{where $N_\textrm{feed}=2$ is the number of feeds per antenna and $N_\textrm{interferometers}=\frac{1}{2}60\times59$ is the number of interferometers.} Therefore, after 100 nights the thermal noise level will be $25\hspace{1mm}\mu\textrm{Jy}$.
 
-Because some RFI sources might be stationary, the signals from these sources will add coherently over time.
+Because some RFI sources might be stationary, the signals from these sources will add \editmark{consistently} over time.
 Therefore, the amount that time integration can decrease the flux density of RFI might be limited.
@@
 Therefore, integrating data from different interferometers and data from the two polarized feeds might not lower the noise level that is caused by RFI at some point.
-In summary, RFI is unlike normal noise and might be coherent over time, interferometer and feed.
+In summary, RFI is unlike normal noise and might \editmark{add consistent} over time, interferometer and feed.
 
@@
 Therefore, frequency averaging will lower the flux density of the RFI signal.
-If the frequency range contains only one stationary RFI source, the strength of this RFI source will go down linearly with the total bandwidth.
+If the frequency range contains only one stationary RFI source, the strength of this source will go down linearly with the total bandwidth.
 If we assume that all channels are affected by RFI sources and all these sources transmit in approximately one channel, then the noise addition that is produced by RFI will go down with the square root of the number of averaged channels.
@@
 
-In summary, some class of stationary RFI sources are expected to be coherent over time, polarization and interferometer, but not over frequency.
+In summary, some class of stationary RFI sources are expected to add \editmark{consistently} over time, polarization and interferometer, but not over frequency.
 Therefore, in this case the noise level at which RFI leakage approximately becomes relevant is given by
@@
 Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes.
-This would cause them to sum (partly) incoherently over time, and thus go approximately down with the noise.
+This would cause them to sum (partly) \editmark{inconsistently} over time, and thus go approximately down with the noise.
 Moreover, many RFI sources will be seen by only a few stations or are not constant over time, which further attenuates their effect.
@@
 Individually, the sources will have fixed geometric phases, but because the baselines are much longer than the wavelength, their phases will become uniformly spread.
-Therefore, they will add incoherently.
+Therefore, they will add inconsistently.
 
@@
 At the same time, this field is useful for analysing the RFI coherency properties.
-It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright foreground sources.
-Preliminary analysis of EoR NCP observations of a single night have almost reached the thermal noise \citep{ncp-eor-yatawatta}, but do not show leaked RFI at the pole.
+It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright \editmark{celestial} foreground sources.
+Preliminary analysis of EoR NCP observations of a single night have \editmark{almost} reached the thermal noise \citep{ncp-eor-yatawatta}, but do not show leaked RFI at the pole.
 
@@
 To mitigate this leaked RFI, the detection can be executed at higher signal-to-noise levels.
-The current results indicate that a lot of RFI is not coherent, and the situation is promising.
+The current results indicate that a lot of RFI \editmark{does not add up consistently}, and the situation is promising.
 Considering the current RFI results, and the availability of further mitigation steps, we conclude that RFI will likely not be problematic for the detection of the Epoch of Reionisation with LOFAR.

### 2013-05-21 articles/Article-RFI-distribution/article-rfi-distribution.tex
Article is ready for resubmission

@@
 With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models.
-Extrapolation of the power law \editmark{implies} that the current LOFAR EoR observations should be severely RFI limited if the \editmark{strength of RFI sources remain strong after integration}.
+Extrapolation of the power law \editmark{implies} that the current LOFAR EoR observations should be severely RFI limited if the \editmark{strength of RFI sources remains strong after time integration}.
 This is in contrast with actual observations, which almost reach the thermal noise and are thought not to be limited by RFI.
 Therefore, we conclude that it is \editmark{unlikely that there are undetected RFI sources that will become visible after long integration.
-Consequently, it is likely} that RFI will not prevent an EoR detection with LOFAR.
+Consequently, there is no indication that RFI will prevent an EoR detection with LOFAR.}
 \end{abstract}
@@
 
-Although two antennas do not necessarily observe the same RFI sources, for source count analysis we can treat the interferometer geometrically as a single point, as both antennas will see the same distribution.
+Although two antennas do not necessarily observe the same RFI sources, for source-count analysis we can treat the interferometer geometrically as a single point, as both antennas will see the same distribution.
 Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and interferometric gain $g = g_1 g_2$ as
@@
 This can be derived by integration, scaling and inversion of the \editmark{rate density} function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
-\begin{equation} \label{eq:sample-with-propagation}
-\tilde{S}_{\textrm{RFI}} \leftarrow \frac{I g}{\zeta x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
-\end{equation}
+\editmark{\begin{equation} \label{eq:sample-with-propagation}
+% === Proof of formula below ===
+% From definition above:
+%   F(S) = N-\rho \pi ( Ig / 4 \pi \zeta S ) ^ (2/\eta)
+% We want S to be in the interval [ S_l ; \inf ], where: (from equation of S(r) above)
+%   S_l = I g / 4 \pi \zeta r_l^\eta,
+% the flux density of the most distance sources at distance r_l that are still visible.
+% This means F(S) will be in the interval [ F(\inf) ; F(S_l) ], and:
+%   F(\inf) = N
+%    F(S_l) = N - \rho \pi r_l^2
+% Because F is the CDF, we can sample a strength by calculating the S
+% from a uniformly distributed F within this interval.
+% If x_u is a uniform variable between 0 and 1, then N - x_u \rho \pi r_l^2 is a uniform
+% variable in F's interval. From this we calculate S with the definition of F(S) :
+%   N - x_u \rho \pi r_l^2 = N-\rho \pi ( Ig / 4 \pi \zeta S ) ^ (2/\eta)
+%       x_u \rho \pi r_l^2 = \rho \pi ( Ig / 4 \pi \zeta S ) ^ (2/\eta)
+%                x_u r_l^2 = ( Ig / 4 \pi \zeta S ) ^ (2/eta)
+%   (x_u r_l^2 ) ^ (eta/2) = Ig / 4 \pi \zeta S
+ %   x_u^(\eta/2) r_l^\eta = Ig / 4 \pi \zeta S
+%                        S = Ig / ( 4 \pi \zeta x_u^(\eta/2) r_l^\eta )
+\tilde{S}_{\textrm{RFI}} \leftarrow \frac{I g}{4\pi \zeta x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
+\end{equation}}
 Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 < x_u,y_u \le 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.
@@
 It is likely that strong transmitters transmit more on average, and therefore contaminate more samples as well.
-High-power transmitters, such as radio stations, have a typical transmitting power in the order of 100~kW.
+High-power transmitters, such as radio stations, have a \editmark{typical equivalent isotropically radiated power (EIRP) in the order of 10--100~kW}.
 Low-power transmitters, such as remote controls, transmit with an order of 100~mW or even less.
@@
 The slope is estimated using linear regression over a visually selected interval.
+\editmark{We have validated that the slope does not significantly change by using a slightly different interval.}
 %\begin{equation}
@@
 With estimates of $\alpha$, $\beta$, $S_L$ and $S_U$, one has obtained a parametrization of the RFI distribution.
-As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is \editmark{$Ig/4\pi r^2$} -- assuming free space propagation for the moment -- and therefore \editmark{$S_L = Ig/4\pi r^2$}.
-For Hata's propagation model, the more generic solution \editmark{$S_L = Ig / 4\pi \zeta r^\eta$} is found.
+As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is \editmark{$S_L = Ig / 4\pi \zeta r_L^\eta$}.
 This value represents the apparent brightness of the \editmark{RFI sources} that are furthest away from the telescope.
@@
 The six most densely packed stations are on the Superterp, an elevated area surrounded by water situated 3~km North of Exloo.
-A radio-quiet zone of 2~km around the Superterp has been established, but is relatively small and households exist within 1~km of the Superterp.}
+A radio-quiet zone of 2~km around the Superterp has been established, but is relatively small and households exist within 1~km of the Superterp.
+With the help of the spectrum allocation registry, the most-obvious transmitters can easily be identified and ignored in LOFAR data \citep{lofar-radio-environment}.
+However, many interfering sources have an unknown origin.}
 
@@
 In that case, the underlying power law might not always be apparent, because not enough samples are combined.
+\editmark{By making distributions over different frequency ranges, we have verified that the power law is not dominated by a few obvious and known sources.}
 
@@
 
-Because some RFI sources might be stationary, the signals from these sources will add \editmark{consistently} over time.
+Because some RFI sources might be stationary, the signals from these sources will add \editmark{consistently} over time\editmark{, meaning that the geometrical phase will be the same every day}.
 Therefore, the amount that time integration can decrease the flux density of RFI might be limited.
@@
 It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright \editmark{celestial} foreground sources.
-Preliminary analysis of EoR NCP observations of a single night have \editmark{almost} reached the thermal noise \citep{ncp-eor-yatawatta}, but do not show leaked RFI at the pole.
+Preliminary analysis of EoR NCP observations of a single night have \editmark{almost} reached the thermal noise, but do not show leaked RFI at the pole \citep[\S4.3]{ncp-eor-yatawatta}.
 

### 2013-05-21 articles/Article-RFI-distribution/pack.sh
SUBMITTED

@@ -1,1 +1,1 @@
-tar -czvf rfi-distributions.tar.gz article-rfi-distribution.tex references.bib mn2e.bst mn2e.cls img/TF-L43786_L220-rfi-example.pdf img/transmitter-height/plot-transmitter-height-trimmed.pdf img/plot-rayleigh-and-rfi-trimmed.pdf img/plot-rayleigh-and-rfi-combined-trimmed.pdf img/explanation-lower-constraint/FindingCounts.pdf img/explanation-lower-constraint/FindingConstraints.pdf img/histograms-raw-trimmed.pdf img/plot-lba-dist-per-frequency-trimmed.pdf img/plot-hba-dist-per-frequency-trimmed.pdf img/histograms-corrected-trimmed.pdf img/histograms-with-fits-trimmed.pdf img/slopes-trimmed.pdf
+tar -czvf rfi-distributions.tar.gz article-rfi-distribution.tex references.bib mn2e.bst mn2e.cls img/TF-L43786_L220-rfi-example-xaxistext.pdf img/transmitter-height/plot-transmitter-height-trimmed.pdf img/plot-rayleigh-and-rfi-trimmed.pdf img/plot-rayleigh-and-rfi-combined-trimmed.pdf img/explanation-lower-constraint/FindingCounts.pdf img/explanation-lower-constraint/FindingConstraints.pdf img/histograms-raw-trimmed.pdf img/plot-lba-dist-per-frequency-trimmed.pdf img/plot-hba-dist-per-frequency-trimmed.pdf img/histograms-corrected-trimmed.pdf img/histograms-with-fits-trimmed.pdf img/slopes-trimmed.pdf

### 2013-05-21 articles/Article-RFI-distribution/article-rfi-distribution.tex
SUBMITTED

@@
 \begin{abstract}
-Faint undetected sources of radio-frequency interference (RFI) might become visible after \editmark{long integrations} of radio observations if they \editmark{are consistently} present over time.
+Faint undetected sources of radio-frequency interference (RFI) might become visible \editmark{in long radio observations when} they \editmark{are consistently} present over time.
 Thereby, they might obstruct the detection of the weak astronomical signals of interest.
@@
 We explore the RFI situation \editmark{at 30--163~MHz} by studying brightness histograms of visibility data observed with LOFAR, similar to radio-source-count analyses that are used in cosmology.
-The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
+\editmark{An empirical RFI distribution model is derived that allows the simulation of RFI in radio observations.} The brightness histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
 With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models.
@@
 This is in contrast with actual observations, which almost reach the thermal noise and are thought not to be limited by RFI.
-Therefore, we conclude that it is \editmark{unlikely that there are undetected RFI sources that will become visible after long integration.
+Therefore, we conclude that it is \editmark{unlikely that there are undetected RFI sources that will become visible in long observations.
 Consequently, there is no indication that RFI will prevent an EoR detection with LOFAR.}

### 2013-06-07 articles/Article-RFI-distribution/article-rfi-distribution.tex
Removing edit marks for next revision

--- 2013-05-21 articles/Article-RFI-distribution/article-rfi-distribution.tex
@@
 \begin{abstract}
-Faint undetected sources of radio-frequency interference (RFI) might become visible after \editmark{long integrations} of radio observations if they \editmark{are consistently} present over time.
+Faint undetected sources of radio-frequency interference (RFI) might become visible in long radio observations when they are consistently present over time.
 Thereby, they might obstruct the detection of the weak astronomical signals of interest.
 This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe.
-We explore the RFI situation \editmark{at 30--163~MHz} by studying brightness histograms of visibility data observed with LOFAR, similar to radio-source-count analyses that are used in cosmology.
-The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
+We explore the RFI situation at 30--163~MHz by studying brightness histograms of visibility data observed with LOFAR, similar to radio-source-count analyses that are used in cosmology.
+An empirical RFI distribution model is derived that allows the simulation of RFI in radio observations.
+The brightness histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5.
 With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models.
-Extrapolation of the power law \editmark{implies} that the current LOFAR EoR observations should be severely RFI limited if the \editmark{strength of RFI sources remains strong after time integration}.
+Extrapolation of the power law implies that the current LOFAR EoR observations should be severely RFI limited if the strength of RFI sources remains strong after time integration.
 This is in contrast with actual observations, which almost reach the thermal noise and are thought not to be limited by RFI.
-Therefore, we conclude that it is \editmark{unlikely that there are undetected RFI sources that will become visible after long integration.
-Consequently, there is no indication that RFI will prevent an EoR detection with LOFAR.}
+Therefore, we conclude that it is unlikely that there are undetected RFI sources that will become visible in long observations.
+Consequently, there is no indication that RFI will prevent an EoR detection with LOFAR.
 \end{abstract}
@@
 However, astronomical radio observations can be affected by radio-frequency interference (RFI), which might make it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}.
-\editmark{The careful management of spectrum allocation and the construction of radio-quiet zones help to limit the number of harmful transmitters.
-If harmful RFI is observed nevertheless, the use of RFI mitigation methods can sometimes clean the data sufficiently to allow succesful calibration and imaging.} Many techniques have been designed to mitigate the effects of RFI, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, post-correlation-rfi-classification}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.
+The careful management of spectrum allocation and the construction of radio-quiet zones help to limit the number of harmful transmitters.
+If harmful RFI is observed nevertheless, the use of RFI mitigation methods can sometimes clean the data sufficiently to allow succesful calibration and imaging.
+Many techniques have been designed to mitigate the effects of RFI, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, post-correlation-rfi-classification}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}.
 
@@
 A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated.
-In such a case, weak interference caused by stationary \editmark{RFI sources} might not manifest itself above the noise in individual observations, but might be \editmark{persistently} present in the data.
+In such a case, weak interference caused by stationary RFI sources might not manifest itself above the noise in individual observations, but might be persistently present in the data.
 Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest.
@@
 In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, \citealt{lofar-overview-2013}, \textit{submitted}).
-\editmark{The LOFAR EoR project aims to detect the redshifted 21-cm signals from the EoR using the LOFAR HBA antennas (115--190~MHz, $z_\textrm{HI}$=11.4--6.5).
+The LOFAR EoR project aims to detect the redshifted 21-cm signals from the EoR using the LOFAR HBA antennas (115--190~MHz, $z_\textrm{HI}$=11.4--6.5).
 Several fields will be observed over 100~nights, to achieve sufficient sensitivity to allow the signal extraction.
 An EoR calibration pipeline has been designed that solves for ionospheric and instrumental effects in approximately hundred directions using the SAGE algorithm \citep{sage-calibration-ii}.
-Initial results from commissioning observations show that in a single night the thermal noise level can almost be reached \citep{ncp-eor-yatawatta}.}
+Initial results from commissioning observations show that in a single night the thermal noise level can almost be reached \citep{ncp-eor-yatawatta}.
 % TODO correct reference if van Haarlem et al is published
@@
 
-%TODO maybe also discuss atmospheric (lightning), e.g.
+% Somewhat related: atmospheric (lightning), e.g.
 see 'Harry M.
@@
 This paper is organised as follows: in Sect.~\ref{sec:distribution-prediction}, we calculate a model for terrestrial interfering source distribution based on various assumptions.
-Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse \editmark{brightness histograms of LOFAR data}.
+Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse brightness histograms of LOFAR data.
 This is followed in Sect.~\ref{sec:dist-data} by a short description of the two LOFAR data sets that have been used to perform the experiment.
@@
 
-In \editmark{time-frequency space}, interfering sources can have complex structures.
-They can also be intermittent and different sources might overlap in \editmark{time-frequency space}.
-An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a \editmark{dynamic spectrum}.
-Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in \editmark{time-frequency space}.
+In time-frequency space, interfering sources can have complex structures.
+They can also be intermittent and different sources might overlap in time-frequency space.
+An example of interfering sources can be seen in Fig.~\ref{fig:dist-rfi-example}, which shows raw visibility data of one baseline of a LOFAR observation in a dynamic spectrum.
+Because many sources change over time, are repetitive or affect multiple channels, many sources produce multiple unconnected features in time-frequency space.
 It is often not clear what constitutes a single interfering source, hence it is hard to count individual sources.
@@
 \includegraphics[height=8cm]{img/TF-L43786_L220-rfi-example-xaxistext.pdf}
-\caption{A \editmark{dynamic spectrum} of a small part of an observation.
+\caption{A dynamic spectrum of a small part of an observation.
 The features with significantly higher values are caused by interference.
@@
 Dividing this quantity by the total number of samples yields the relative number of events as a function of intensity.
-We will refer to this quantity with the term \editmark{``rate density''}.
+We will refer to this quantity with the term ``rate density''.
 We will now start by estimating the rate density function of ground-based interfering sources.
-Consider an interfering point source of strength $I$ that denotes \editmark{the transmitting power normalized by the observational channel resolution (e.g., measured in W/Hz)}.
+Consider an interfering point source of strength $I$ that denotes the transmitting power normalized by the observational channel resolution (e.g., measured in W/Hz).
 This source is observed by an interferometer that consists of two antennas or stations with gains $g_1,g_2$, which include all instrumental effects.
@@
 The interferometer will record an apparent instantaneous strength $\mathcal{S}$ of
-\editmark{\begin{equation}
+\begin{equation}
  \mathcal{S}(r_1, r_2) = I \frac{g_1 g_2}{4\pi r_1 r_2},
-\end{equation}}
+\end{equation}
 with (real-valued amplitude) gains $g_1,g_2>0$ and $r_\textrm{L} > r_1,r_2 > 0$.
@@
 Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the narrow-band condition.
-\editmark{With a frequency resolution $\Delta\nu=$ 0.76~kHz}, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c (2 \pi \Delta \nu)^{-1} \approx$ 50~\editmark{km}.
-\editmark{Because the velocity resolution of LOFAR is 1.5~km/s at 150~MHz, and larger at lower frequencies, a Doppler frequency shift due to movement of the source will only be significant if its velocity is at least 1.5~km/s relative to the antennas.
-Since the relative velocities towards different antennas in the array will be similar for such high-velocity transmitters (i.e., satellites), there will be hardly any decorrelation because of Doppler shifting.}
+With a frequency resolution $\Delta\nu=$ 0.76~kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c (2 \pi \Delta \nu)^{-1} \approx$ 50~km.
+Because the velocity resolution of LOFAR is 1.5~km/s at 150~MHz, and larger at lower frequencies, a Doppler frequency shift due to movement of the source will only be significant if its velocity is at least 1.5~km/s relative to the antennas.
+Since the relative velocities towards different antennas in the array will be similar for such high-velocity transmitters (i.e., satellites), there will be hardly any decorrelation because of Doppler shifting.
 
@@
 Then, we can express the received amplitude $\mathcal{S}$ for a given distance $r$ and interferometric gain $g = g_1 g_2$ as
-\editmark{\begin{equation} \label{eq:amplitude-fall-off-in-free-space}
+\begin{equation} \label{eq:amplitude-fall-off-in-free-space}
  S(r) = \frac{Ig}{4\pi r^2}.
-\end{equation}}
+\end{equation}
 
@@
 Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid.
-Using these assumptions, we can express the expected cumulative \editmark{rate density} of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
+Using these assumptions, we can express the expected cumulative rate density of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
 \begin{equation} \label{eq:cumulative-distance-function}
@@
 
-We need \editmark{$F_{\textrm{distance} \ge r}(r)$, the complement} of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance.
+We need $F_{\textrm{distance} \ge r}(r)$, the complement of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance.
 Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2$, with $N$ the total number of sources.
@@
 The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can now be calculated with
-\editmark{\begin{align} \label{eq:cumulative-rfi-distribution}
+\begin{align} \label{eq:cumulative-rfi-distribution}
  F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{R}(S))\\
@@
 \notag = & N-\frac{\rho Ig}{4S}
-\end{align}}
+\end{align}
 where $\mathcal{R}(S)=\mathcal{S}^{-1}$, the inverse of $\mathcal{S}$, i.e., the function that returns the distance $r$ for a given amplitude $S$.
-Finally, the \editmark{rate density} can be calculated by taking the derivative,
-\editmark{\begin{align} \label{eq:two-dimensional-distribution}
+Finally, the rate density can be calculated by taking the derivative,
+\begin{align} \label{eq:two-dimensional-distribution}
  f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{\rho Ig}{4S^2}.
-\end{align}}
+\end{align}
 
@@
 The original model is based on urban areas, but corrections are given by Hata for suburban areas with a lower population density and for open areas.
-Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss \editmark{between two ground-based antennas}:
+Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss between two ground-based antennas:
 \begin{align} 
@@
 Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
-\editmark{\begin{equation} \label{eq:propagation-loss-factor}
+\begin{equation} \label{eq:propagation-loss-factor}
  L_S = \frac{1}{10} 10^{L_p} = \zeta r^{4.49 - 0.655 \log_{10} h_b},
-\end{equation}}
+\end{equation}
 with
@@
 Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
-\editmark{\begin{equation}
+\begin{equation}
  S(r) = \frac{Ig}{4\pi \zeta r^\eta},
-\end{equation}}
+\end{equation}
 with $\eta = 4.49 - 0.655 \log_{10} h_b$ as seen in Eq.~\eqref{eq:propagation-loss-factor}, one finds the rate density function $f_p$ that considers propagation effects,
@@
 % diff(rho*%pi*r^2,S),solve(S(r)=S,r);
-\editmark{\begin{align} \label{eq:frequency-density-with-propagation}
+\begin{align} \label{eq:frequency-density-with-propagation}
 f_p(S)
@@
 = \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{4\pi\zeta S} \right)^{2/\eta}.
-\end{align}}
+\end{align}
 Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
@@
 The full measured distribution will consist of the power-law distribution combined with that of the thermal noise and the celestial signal.
-For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong \editmark{celestial} sources.
-For example, the strongest apparent \editmark{celestial} source in the NCP EoR field is around 5 Jy \citep{ncp-eor-yatawatta}.
+For now, we will ignore the contribution of the celestial signal, as its contribution to the amplitude distribution will be minimal when observing fields without strong celestial sources.
+For example, the strongest apparent celestial source in the NCP EoR field is around 5 Jy \citep{ncp-eor-yatawatta}.
 The standard deviation of the noise, however, is around 100~Jy on highest LOFAR resolutions, and will have a larger contribution on the histogram.
@@
 \caption{The Rayleigh and power-law distributions in a log-log plot.
-The power-law distribution (Eq.~\eqref{eq:two-dimensional-distribution}) has a constant slope of -2 \editmark{over the range it is defined}.
+The power-law distribution (Eq.~\eqref{eq:two-dimensional-distribution}) has a constant slope of -2 over the range it is defined.
 The slope of the Rayleigh distribution in the limit of the origin is 1.
@@
 The histogram can be numerically estimated by drawing complex samples from the two distributions and calculating and counting the amplitudes.
-A sample can be drawn from the RFI distribution by scaling and inverting the cumulative \editmark{rate density} function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable.
-Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of \editmark{RFI sources} will go to infinity.
-Therefore, one needs to assume that there are no \editmark{sources} beyond some limiting distance $r_L$ to invert the cumulative function.
+A sample can be drawn from the RFI distribution by scaling and inverting the cumulative rate density function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable.
+Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of RFI sources will go to infinity.
+Therefore, one needs to assume that there are no sources beyond some limiting distance $r_L$ to invert the cumulative function.
 With this assumption, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
-\editmark{\begin{equation} \label{eq:drawing-rfi-distribution}
+\begin{equation} \label{eq:drawing-rfi-distribution}
 S_{\textrm{RFI}} \leftarrow \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
-\end{equation}}
-\editmark{where $0 < x_u, y_u \le 1$ are independently drawn from a uniform distribution}.
-The result is an amplitude sample $S_{\textrm{RFI}}$, with \editmark{$S_{\textrm{RFI}} \ge Ig/4\pi r_L^2$}, that follows a power-law distribution.
+\end{equation}
+where $0 < x_u, y_u \le 1$ are independently drawn from a uniform distribution.
+The result is an amplitude sample $S_{\textrm{RFI}}$, with $S_{\textrm{RFI}} \ge Ig/4\pi r_L^2$, that follows a power-law distribution.
 The corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
-\editmark{\begin{equation} \label{eq:drawing-combined-distribution}
+\begin{equation} \label{eq:drawing-combined-distribution}
 S \leftarrow v_n + w_n i + \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
-\end{equation}}
+\end{equation}
 with $v_n, w_n \sim N(\mu=0; \sigma)$. The final amplitude sample $S_A$ can be calculated with 
@@
 \end{equation}
-An example of distribution curves for various settings of \editmark{$Ig/4\pi r_L^2$} is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.
+An example of distribution curves for various settings of $Ig/4\pi r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.
 
-If we assume that $I$, the average \editmark{RFI source} strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-combined-distribution} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram.
-Although calibration could in theory solve $g$, almost all \editmark{RFI} sources will come in through the far-sidelobes of \editmark{the beam}, and finding the expected values for the gains is therefore hard.
+If we assume that $I$, the average RFI source strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-combined-distribution} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram.
+Although calibration could in theory solve $g$, almost all RFI sources will come in through the far-sidelobes of the beam, and finding the expected values for the gains is therefore hard.
 The effects of these parameters on the histogram will be further discussed in \S\ref{sec:parameter-variability}.
@@
 For completeness we show the formula for sampling apparent source strengths that include propagation effects according to Hata's model, which can be used for sampling realistic source amplitudes in simulations.
-This can be derived by integration, scaling and inversion of the \editmark{rate density} function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
-\editmark{\begin{equation} \label{eq:sample-with-propagation}
+This can be derived by integration, scaling and inversion of the rate density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
+\begin{equation} \label{eq:sample-with-propagation}
 % === Proof of formula below ===
@@
 \tilde{S}_{\textrm{RFI}} \leftarrow \frac{I g}{4\pi \zeta x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
-\end{equation}}
+\end{equation}
 Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 < x_u,y_u \le 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.
@@
 \subsection{Parameter variability} \label{sec:parameter-variability}
-In reality, the parameters $\rho$, $I$ and $g$, which are the RFI source density per unit area, \editmark{RFI source} strength and instrumental gain respectively, will not be constant, but can change over time and frequency.
-Therefore, \editmark{they are} stochastic variables.
+In reality, the parameters $\rho$, $I$ and $g$, which are the RFI source density per unit area, RFI source strength and instrumental gain respectively, will not be constant, but can change over time and frequency.
+Therefore, they are stochastic variables.
 However, since each specific value for these parameters produces a power law, the combined distribution will still show a power law, as long as the parameters follow a distribution that is steep at high amplitudes (in $\log$--$\log$ space), such as a Gaussian or uniform distribution.
@@
 We determined that the gain variation due to the band-pass is about one order of magnitude for the low-band antennas (LBA, 30.1--77.5 MHz) and about a factor of two for the high-band antennas (HBA, 115.0--163.3 MHz).
-\editmark{The frequency dependency of the gains due to the band-pass will consequently smooth the data in the brightness histogram in horizontal direction by one order of magnitude or less.}
+The frequency dependency of the gains due to the band-pass will consequently smooth the data in the brightness histogram in horizontal direction by one order of magnitude or less.
 
@@
 It is likely that strong transmitters transmit more on average, and therefore contaminate more samples as well.
-High-power transmitters, such as radio stations, have a \editmark{typical equivalent isotropically radiated power (EIRP) in the order of 10--100~kW}.
+High-power transmitters, such as radio stations, have a typical equivalent isotropically radiated power (EIRP) in the order of 10--100~kW.
 Low-power transmitters, such as remote controls, transmit with an order of 100~mW or even less.
@@
 This is mandated by the large dynamic range of the histogram that we are interested in.
-Therefore, we chose to have a bin size that increases linearly with the amplitude $S$, and the \editmark{rate counts} are divided by the bin size after counting.
+Therefore, we chose to have a bin size that increases linearly with the amplitude $S$, and the rate counts are divided by the bin size after counting.
 Consequently, in parts of the histogram that have a sparse number of samples, the outlying samples will follow a $1/S$ curve, or a -1 slope in a log-log plot.
@@
 The slope is estimated using linear regression over a visually selected interval.
-\editmark{We have validated that the slope does not significantly change by using a slightly different interval.}
+We have validated that the slope does not significantly change by using a slightly different interval.
 %\begin{equation}
@@
 
-\editmark{Fitting} straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}.
+Fitting straight lines to the distribution curve in a log-log plot is not the most accurate way of estimating the exponent of a power-law distribution \citep{power-law-distribution}.
 However, because of our enormous sample size, which allows fitting the line over a large interval, the estimator will be sufficiently accurate for our purpose.
@@
 Once the exponent of the power-law part of the distribution is estimated with the previously discussed techniques, the distribution can be extrapolated to find a lower flux limit.
-Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the following \editmark{rate density} function $h$:
+Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the following rate density function $h$:
 \begin{equation} \label{eq:amplitude-power-law}
@@
 Areas in a log-log plot are not linearly related to the integral.
-There are two ways to estimate the lower constraint $S_L$: (i) the areas $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal if \editmark{$Ig/r_L^\eta$} is constant during the observation, and (ii) if one assumes \editmark{$Ig/r_L^\eta\sim$} uniform, then $N_a+N_b = N_\textrm{total}-N_\textrm{part}$.}
+There are two ways to estimate the lower constraint $S_L$: (i) the areas $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal if $Ig/r_L^\eta$ is constant during the observation, and (ii) if one assumes $Ig/r_L^\eta\sim$ uniform, then $N_a+N_b = N_\textrm{total}-N_\textrm{part}$.}
 \label{fig:explanation-constraints}
@@
 
-$S_1$ and $S_2$ are selected by \editmark{visual} inspection of the histogram.
+$S_1$ and $S_2$ are selected by visual inspection of the histogram.
 Assume that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}, and that the effect of the Rayleigh component on the histogram $> S_1$ is negligible.
@@
 First and most importantly, the samples themselves are digitized with an analogue-to-digital converter (ADC) with limited range.
-Second, we observe for a limited time and the \editmark{rate count} is discrete.
+Second, we observe for a limited time and the rate count is discrete.
 Because the chance of finding a sample with a very high intensity is low, it is unlikely to observe samples beyond some amplitude within the finite observing time.
@@
 \end{equation}
-In some cases there will be no solution for $S_U$. \editmark{This can happen when the empirical distribution is not limited at the high end, or when it contains features that are not in the model.}
+In some cases there will be no solution for $S_U$. This can happen when the empirical distribution is not limited at the high end, or when it contains features that are not in the model.
 
@@
 
-With the assumption that \editmark{$Ig/r_L^\eta\sim$} a uniform distribution, the area labelled in Fig.~\ref{fig:explanation-constraints} as $N_b$ is also part of the RFI distribution, and a stronger constraint $\tilde S_L$ can be found by solving
+With the assumption that $Ig/r_L^\eta\sim$ a uniform distribution, the area labelled in Fig.~\ref{fig:explanation-constraints} as $N_b$ is also part of the RFI distribution, and a stronger constraint $\tilde S_L$ can be found by solving
 \begin{equation}
@@
 With estimates of $\alpha$, $\beta$, $S_L$ and $S_U$, one has obtained a parametrization of the RFI distribution.
-As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is \editmark{$S_L = Ig / 4\pi \zeta r_L^\eta$}.
-This value represents the apparent brightness of the \editmark{RFI sources} that are furthest away from the telescope.
+As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is $S_L = Ig / 4\pi \zeta r_L^\eta$.
+This value represents the apparent brightness of the RFI sources that are furthest away from the telescope.
 With a fully parametrized distribution of the effect of RFI sources, an empirical model for RFI effects can be made.
@@
 Since modern RFI detection algorithms can find all RFI that is detectable ``by eye'' \citep{post-correlation-rfi-classification}, this power level will be near the level of the noise mode.
-\editmark{We use the AOFlagger for RFI detection, which will be described in Sect.~\ref{sec:dist-data}.} In
-\citet{lofar-radio-environment}
-the false-positives rate for the AOFlagger is estimated to be 0.5\%.
+We use the AOFlagger for RFI detection, which will be described in Sect.~\ref{sec:dist-data}.
+In
+\citet{lofar-radio-environment} the false-positives rate for the AOFlagger is estimated to be 0.5\%.
 
@@
 We can assign flux densities to the horizontal axis of the histogram by using the system equivalent flux density (SEFD) of a single station.
-The current LOFAR SEFD is found to be approximately \editmark{3400~Jy} for the HBA core stations and \editmark{1700~Jy} for the remote stations in the frequency range from $125$--$175$ MHz.
-For all Dutch LBA stations, in the frequency range $40$--$70$ MHz the SEFD is approximately \editmark{34,000~Jy}.
+The current LOFAR SEFD is found to be approximately 3400~Jy for the HBA core stations and 1700~Jy for the remote stations in the frequency range from $125$--$175$ MHz.
+For all Dutch LBA stations, in the frequency range $40$--$70$ MHz the SEFD is approximately 34,000~Jy.
 The standard deviation $\sigma$ in the real and imaginary values is related to the SEFD with % TODO citation
@@
 (2013, \textit{submitted}) for a full description of the capabilities of LOFAR.
-\editmark{The analyses will cover only Dutch stations.
+The analyses will cover only Dutch stations.
 Each Dutch station consists of 96 dipole low-band antennas (LBA) and one or two fields totalling 48~tiles of 4x4 bow-tie high-band antennas (HBA).
@@
 With the help of the spectrum allocation registry, the most-obvious transmitters can easily be identified and ignored in LOFAR data \citep{lofar-radio-environment}.
-However, many interfering sources have an unknown origin.}
+However, many interfering sources have an unknown origin.
 
-\editmark{In the two data sets, we have used the correlation coefficients of cross-correlated stations, i.e., the raw visibilities.} In one data set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz.
+In the two data sets, we have used the correlation coefficients of cross-correlated stations, i.e., the raw visibilities.
+In one data set, the low-band antennas (LBA) were used and the frequency range 30.1--77.5~MHz was recorded, while in the other the high-band antennas (HBA) were used to record the frequency range 115.0--163.3~MHz.
 More stations were used in the LBA set.
@@
 To detect RFI, the AOFlagger \citep{LOFAR-RFI-pipeline} is used.
-\editmark{This RFI detector estimates the sky contribution by iteratively applying a high-pass filter to the visibility amplitudes of a single baseline in the time-frequency plane.
+This RFI detector estimates the sky contribution by iteratively applying a high-pass filter to the visibility amplitudes of a single baseline in the time-frequency plane.
 Subsequently, it flags line-shaped features with the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}.
-Finally, the scale-invariant rank operator, a morphological technique to search for contaminated samples, is applied on the two-dimensional flag mask \citep{scale-invariant-rank-operator}.}
+Finally, the scale-invariant rank operator, a morphological technique to search for contaminated samples, is applied on the two-dimensional flag mask \citep{scale-invariant-rank-operator}.
 
@@
 Field               & NCP             & NCP \\
-\editmark{Amount of RFI detected} & & \\
-\editmark{by the AOFlagger}    & 1.77\%       & 3.18\% \\
+Amount of RFI detected & & \\
+by the AOFlagger    & 1.77\%       & 3.18\% \\
 \hline
@@
 
-It is to be expected that the RFI-dominated part of the distributions at different frequencies will reflect the underlying \editmark{RFI source} populations.
+It is to be expected that the RFI-dominated part of the distributions at different frequencies will reflect the underlying RFI source populations.
 Both Figs.~\ref{fig:plot-dist-per-frequency-LBA} and \ref{fig:plot-dist-per-frequency-HBA} show that the power-law part of the distributions are very different for different sub-bands.
@@
 In that case, the underlying power law might not always be apparent, because not enough samples are combined.
-\editmark{By making distributions over different frequency ranges, we have verified that the power law is not dominated by a few obvious and known sources.}
+By making distributions over different frequency ranges, we have verified that the power law is not dominated by a few obvious and known sources.
 
@@
 This validates that the variable gain that is caused by the antenna response does not change the observed power law.
-Consequently, it can be expected that other stochastic effects, such as the intrinsic \editmark{RFI source} strength and the beam gain due to a differing direction of arrival, will similarly not affect the power law.
+Consequently, it can be expected that other stochastic effects, such as the intrinsic RFI source strength and the beam gain due to a differing direction of arrival, will similarly not affect the power law.
 Because the band-pass corrected histograms should provide a more accurate analysis, we will use the corrected histograms for further analysis.
@@
 Although the HBA slope does not show any other significant features besides the Rayleigh and power-law curves, the LBA power law ends with a bulge around an amplitude of $10^6$.
-This bulge is caused by a very strong \editmark{RFI} source affecting lots of samples, and is a single outlier in the spatial distribution.
+This bulge is caused by a very strong RFI source affecting lots of samples, and is a single outlier in the spatial distribution.
 We found this is caused by RFI observed for about an hour in the late afternoon in the lower LBA frequency regime, around $30$--$40$~MHz.
@@
 $S_L$ & Constraint on lower fall-off point of power law & $21$ mJy & $6.2$ mJy \\ % $10^{-6.42}$ & $10^{-4.31}$
-$\tilde{S}_L$ & As $S_L$, but assuming \editmark{$Ig/r^\eta\sim$} uniform & $47$ mJy & $14$ mJy \\ % $10^{-6.08}$ & $10^{-3.96}$
+$\tilde{S}_L$ & As $S_L$, but assuming $Ig/r^\eta\sim$ uniform & $47$ mJy & $14$ mJy \\ % $10^{-6.08}$ & $10^{-3.96}$
 $S_d$ & Expected lowest apparent level of RFI detected & $26$ Jy & $5.7$ Jy\\ % $2.3 \times 10^{-2}$ & $4.5 \times 10^{-2}$
@@
 \hline
-\multicolumn{4}{l}{\editmark{\textit{Average station temperatures}}} \\
+\multicolumn{4}{l}{\textit{Average station temperatures}} \\
 \hline
@@
 because the effect of the ionosphere becomes stronger.
-Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and \editmark{$[-1.57;-1.49]$} for the HBA.
+Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.57;-1.49]$ for the HBA.
 The estimate of the HBA is thus more accurate, because its histogram deviates less from the power law.
-\editmark{This translates to boundaries on the average transmitter height of $[0.6; 800]$ and $[3.1; 23]$~m for the LBA and HBA respectively.
+This translates to boundaries on the average transmitter height of $[0.6; 800]$ and $[3.1; 23]$~m for the LBA and HBA respectively.
 Therefore, the LBA measurements are clearly not accurate enough to be conclusive.
 Moreover, because the power-law distribution analyses involve many assumptions, it is uncertain whether the analyses are sufficiently accurate for making these detailed conclusions.
-}
 
@@
 
-\editmark{The apparent RFI flux densities can be converted to a RFI station temperature that excludes the system noise and sky noise components.
+The apparent RFI flux densities can be converted to a RFI station temperature that excludes the system noise and sky noise components.
 If we use a station efficiency factor $\eta_\textrm{st}=1$ and effective areas LBA $A_\textrm{eff}=398$ and HBA $A_\textrm{eff}=512$ with again 100\% RFI occupancy, our models lead to RFI temperatures of 17,000~K and 1,200~K for respectively the LBA and the HBA.
@@
 However, our post-detection RFI station temperatures, which arise from the residual apparent RFI flux density estimates, are 3.2~K and 1.4~K for the LBA and HBA respectively.
-Due to LOFAR's high resolution and accurate flagging strategy, this is achieved by flagging a relatively small data percentage of 1.8 (LBA) and 3.2\% (HBA).}
+Due to LOFAR's high resolution and accurate flagging strategy, this is achieved by flagging a relatively small data percentage of 1.8 (LBA) and 3.2\% (HBA).
 
@@
 \end{equation}
-\editmark{where $N_\textrm{feed}=2$ is the number of feeds per antenna and $N_\textrm{interferometers}=\frac{1}{2}60\times59$ is the number of interferometers.} Therefore, after 100 nights the thermal noise level will be $25\hspace{1mm}\mu\textrm{Jy}$.
+where $N_\textrm{feed}=2$ is the number of feeds per antenna and $N_\textrm{interferometers}=\frac{1}{2}60\times59$ is the number of interferometers.
+Therefore, after 100 nights the thermal noise level will be $25\hspace{1mm}\mu\textrm{Jy}$.
 
-Because some RFI sources might be stationary, the signals from these sources will add \editmark{consistently} over time\editmark{, meaning that the geometrical phase will be the same every day}.
+Because some RFI sources might be stationary, the signals from these sources will add consistently over time, meaning that the geometrical phase will be the same every day.
 Therefore, the amount that time integration can decrease the flux density of RFI might be limited.
@@
 Therefore, integrating data from different interferometers and data from the two polarized feeds might not lower the noise level that is caused by RFI at some point.
-In summary, RFI is unlike normal noise and might \editmark{add consistent} over time, interferometer and feed.
+In summary, RFI is unlike normal noise and might add consistent over time, interferometer and feed.
 
@@
 
-In summary, some class of stationary RFI sources are expected to add \editmark{consistently} over time, polarization and interferometer, but not over frequency.
+In summary, some class of stationary RFI sources are expected to add consistently over time, polarization and interferometer, but not over frequency.
 Therefore, in this case the noise level at which RFI leakage approximately becomes relevant is given by
@@
 Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes.
-This would cause them to sum (partly) \editmark{inconsistently} over time, and thus go approximately down with the noise.
+This would cause them to sum (partly) inconsistently over time, and thus go approximately down with the noise.
 Moreover, many RFI sources will be seen by only a few stations or are not constant over time, which further attenuates their effect.
@@
 At the same time, this field is useful for analysing the RFI coherency properties.
-It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright \editmark{celestial} foreground sources.
-Preliminary analysis of EoR NCP observations of a single night have \editmark{almost} reached the thermal noise, but do not show leaked RFI at the pole \citep[\S4.3]{ncp-eor-yatawatta}.
+It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright celestial foreground sources.
+Preliminary analysis of EoR NCP observations of a single night have almost reached the thermal noise, but do not show leaked RFI at the pole \citep[\S4.3]{ncp-eor-yatawatta}.
 
@@
 To mitigate this leaked RFI, the detection can be executed at higher signal-to-noise levels.
-The current results indicate that a lot of RFI \editmark{does not add up consistently}, and the situation is promising.
+The current results indicate that a lot of RFI does not add up consistently, and the situation is promising.
 Considering the current RFI results, and the availability of further mitigation steps, we conclude that RFI will likely not be problematic for the detection of the Epoch of Reionisation with LOFAR.

### 2013-07-04 articles/Article-RFI-distribution/article-rfi-distribution.tex
Trimming less relevant parts to accomodate referee complaint of article too long

@@
 
-In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, \citealt{lofar-overview-2013}, \textit{submitted}).
+In this article, we will connect new insights about RFI to the angular \editmark{EoR} experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, \citealt{lofar-overview-2013}).
 The LOFAR EoR project aims to detect the redshifted 21-cm signals from the EoR using the LOFAR HBA antennas (115--190~MHz, $z_\textrm{HI}$=11.4--6.5).
@@
 Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse brightness histograms of LOFAR data.
-This is followed in Sect.~\ref{sec:dist-data} by a short description of the two LOFAR data sets that have been used to perform the experiment.
+\editmark{Sect.~\ref{sec:dist-data} describes} the two LOFAR data sets that have been used to perform the experiment.
 The results of analysing the sets are presented in Sect.~\ref{sec:dist-results}.
@@
 
-First we will derive the expected intrinsic source distribution for interfering radio sources. After that, we will analyse the issues that arise when measuring the distribution by counting samples.
+\editmark{It is common knowledge that in a uniform Euclidian Universe source counts behave like power-law distributions.
+For sources on a flat surface, the power law will have a -2 exponent}.
+We will derive this expected intrinsic source distribution for interfering radio sources.
+After that, we will analyse the issues that arise when measuring the distribution by counting samples.
 
@@
 Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in the local two-dimensional horizontal plane.
-This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights.
-Beyond some distance, the Earth will partly block the path between transmitter and receiver.
-Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid.
-Using these assumptions, we can express the expected cumulative rate density of sources $F_{\textrm{distance} \le r}(r)$ at distance $r$ as
-\begin{equation} \label{eq:cumulative-distance-function}
-F_{\textrm{distance} \le r}(r)=\rho \pi r^2,
+\editmark{These assumptions are obviously simplications, but we will address these later.
+%This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights.
+Beyond some distance, the Earth will partly block the path between transmitter and \editmark{receiver.
+Using} these assumptions, we can express the expected \editmark{inverse cumulative rate density of sources at distance $r$ as
+\begin{equation} \label{eq:inverse-cumulative-distance-function}
+%F_{\textrm{distance} \le r}(r)=\rho \pi r^2, (<-- non inverted)
+F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2,
 \end{equation}
-for some constant $\rho$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance $r$ away.
-
+with $N$ the total number of sources and for some constant $\rho$ that represents the number of sources per unit area.}
+%In other words, we will observe $F$ sources that are further than a distance $r$ away.
 We need $F_{\textrm{distance} \ge r}(r)$, the complement of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance.
@@
 Because $r$ is restricted, $N$ is finite.
-
-The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can now be calculated with
+The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can \editmark{be calculated from this with}
 \begin{align} \label{eq:cumulative-rfi-distribution}
- F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{R}(S))\\
-\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{4\pi S})\\
-\notag = & N-\frac{\rho Ig}{4S}
+ F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{R}(S))
+%\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{4\pi S})\\
+ = N-\frac{\rho Ig}{4S}
 \end{align}
@@
 So far, we have assumed that the electromagnetic radiation propagates through free space, resulting in an $r^{-2}$ fall-off.
-In reality, the radiation will be affected by complicated propagational effects.
-Because of the irregular surface of the Earth (including urban areas) and the absence of a direct line of sight between transmitter and receiver, the propagation mode might be indirect.
+In reality, the radiation will be affected by complicated propagational effects \editmark{due to the surface of the Earth}.
+%Because of the irregular surface of the Earth (including urban areas) and the absence of a direct line of sight between transmitter and receiver, the propagation mode might be indirect.
 Multiple indirect paths might be formed that include reflection, refraction, diffraction and absorption of the electromagnetic wave.
-
 A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}.
@@
 
-Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
+Converting from a subtracted term in decibels to a flux density factor $L_S$ \editmark{results in}
 \begin{equation} \label{eq:propagation-loss-factor}
- L_S = \frac{1}{10} 10^{L_p} = \zeta r^{4.49 - 0.655 \log_{10} h_b},
+ L_S = \frac{1}{10} 10^{L_p} = \zeta r^\eta,
 \end{equation}
-with
-\begin{equation} \label{eq:zeta-definition}
- \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
+\editmark{with $\eta$ and $\zeta$ given by}
+\begin{align} 
+ \label{eq:zeta-definition} \eta & = 4.49 - 0.655 \log_{10} h_b,\\
+ \label{eq:eta-definition}  \zeta & = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
+\end{align}
+Note that according to Hata's model, the exponent of the power law $\eta$ depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density.
+To find the rate density function $f_p$ that considers propagation effects, one can replace $S(r)$ in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} with one that includes the propagation effects,
+\begin{equation}
+ S(r) = \frac{Ig}{4\pi \zeta r^\eta}.
 \end{equation}
-Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density.
-Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
-\begin{equation}
- S(r) = \frac{Ig}{4\pi \zeta r^\eta},
-\end{equation}
-with $\eta = 4.49 - 0.655 \log_{10} h_b$ as seen in Eq.~\eqref{eq:propagation-loss-factor}, one finds the rate density function $f_p$ that considers propagation effects,
+\editmark{The} resulting rate density function $f_p$ is
 % S(r):=I*g/(r^alpha);
@@
 \end{figure}
-This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined.
-This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively.
+\editmark{This} yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively.
 In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.
@@
 
-So far, these are the expected histograms for pure noise and pure RFI that propagates through free space.
-However, the measured distribution is a mixture of the two.
-A perfect RFI detector would classify the samples in two groups; one group of samples that are not affected by RFI, and therefore contain noise only, and one group of samples that are contaminated, and are the sum of RFI and noise.
-Real RFI detectors can classify samples into these groups to some extent, but due to false positives and false negatives, the groups will get mixed nevertheless.
-Even more, we still need to take into account that the samples classified as RFI are also affected by noise.
-Although this effect is moderate, because `most' RFI samples are of much higher amplitude than what is added because of the noise, we are interested in the low-level RFI samples, and the relative effect of noise on those samples is large.
-
 \begin{figure}
@@
 \caption{Histograms of simulated samples that all have a contribution of noise and RFI.
-Various settings of the parameters were used, and samples were drawn as described in Eq.~\eqref{eq:drawing-amplitudes}.
+Various settings of the parameters were used, and samples were drawn as described in Eq.~\eqref{eq:drawing-rfi-with-propagation}.
 Solid lines: the combined distributions, dashed lines: the power-law distributions before mixing.
@@
 
-To calculate the corresponding amplitude distribution that is formed by the combination of the real and imaginary component distributions, it is easier to perform numerical simulations, as analytic derivation is not trivial.
-The histogram can be numerically estimated by drawing complex samples from the two distributions and calculating and counting the amplitudes.
-A sample can be drawn from the RFI distribution by scaling and inverting the cumulative rate density function in Eq.~\eqref{eq:cumulative-rfi-distribution} and evaluating it for a uniformly distributed variable.
-Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of RFI sources will go to infinity.
-Therefore, one needs to assume that there are no sources beyond some limiting distance $r_L$ to invert the cumulative function.
-With this assumption, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with
-\begin{equation} \label{eq:drawing-rfi-distribution}
-S_{\textrm{RFI}} \leftarrow \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
-\end{equation}
-where $0 < x_u, y_u \le 1$ are independently drawn from a uniform distribution.
-The result is an amplitude sample $S_{\textrm{RFI}}$, with $S_{\textrm{RFI}} \ge Ig/4\pi r_L^2$, that follows a power-law distribution.
-The corresponding formula for drawing a sample $S$ that is contaminated by both RFI and noise is
-\begin{equation} \label{eq:drawing-combined-distribution}
-S \leftarrow v_n + w_n i + \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
-\end{equation}
-with $v_n, w_n \sim N(\mu=0; \sigma)$. The final amplitude sample $S_A$ can be calculated with 
-\begin{equation} \label{eq:drawing-amplitudes}
-S_A \leftarrow |S| = \sqrt{\textrm{real}^2({S})+\textrm{imag}^2({S})}.
-\end{equation}
-An example of distribution curves for various settings of $Ig/4\pi r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.
-
-If we assume that $I$, the average RFI source strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-combined-distribution} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram.
-Although calibration could in theory solve $g$, almost all RFI sources will come in through the far-sidelobes of the beam, and finding the expected values for the gains is therefore hard.
-The effects of these parameters on the histogram will be further discussed in \S\ref{sec:parameter-variability}.
-
-%Fig.~\ref{fig:rayleigh-and-rfi-combined} shows that the left side of the graph corresponds with the original Rayleigh distribution.
-From the right side of the histogram, we are able to estimate $Ig/r_L^2$, as the placing of the power-law curve is independent of the Rayleigh distribution for large $S$.
-Methods to constrain the parameters of the RFI distribution will be discussed in \S\ref{sec:rfi-distribution-constraints}.
-The curves in Fig.~\ref{fig:rayleigh-and-rfi-combined} are from histograms of samples that are all contaminated by both noise and RFI.
-In practice, we can not make this distinction accurately, as RFI detectors have a limited accuracy.
-Consequently, the observed histogram will be the sum of two types of histograms: the first histogram being contaminated by both noise and RFI, the second one only by noise.
-
-For completeness we show the formula for sampling apparent source strengths that include propagation effects according to Hata's model, which can be used for sampling realistic source amplitudes in simulations.
-This can be derived by integration, scaling and inversion of the rate density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
-\begin{equation} \label{eq:sample-with-propagation}
+So far, these are the expected histograms for pure noise and pure RFI that propagates through free space.
+However, the measured distribution is a mixture of the \editmark{two.
+%A perfect RFI detector would classify the samples in two groups; one group of samples that are not affected by RFI, and therefore contain noise only, and one group of samples that are contaminated, and are the sum of RFI and noise.
+Real RFI detectors can classify samples into these groups to some extent, but due to false positives and false negatives, the groups will get mixed nevertheless.
+Even more, we still need to take into account that the samples classified as RFI are also affected by noise.
+Although this effect is moderate, because `most' RFI samples are of much higher amplitude than what is added because of the noise, we are interested in the low-level RFI samples, and the relative effect of noise on those samples is large.
+Analytic derivation of the corresponding mixed amplitude distribution is not trivial, but the distributions can easily be} estimated by drawing complex samples from the two distributions and calculating and counting the amplitudes.
+A sample can be drawn from the RFI distribution by integration, scaling and inversion of the rate density function in Eq.~\eqref{eq:frequency-density-with-propagation}.
+%Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of RFI sources will go to infinity.
+\editmark{To invert the cumulative function, one needs to assume that there are no sources beyond some limiting distance $r_L$}.
+With this assumption, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with:
+\begin{equation} \label{eq:drawing-rfi-with-propagation}
 % === Proof of formula below ===
@@
 %                        S = Ig / ( 4 \pi \zeta x_u^(\eta/2) r_l^\eta )
-\tilde{S}_{\textrm{RFI}} \leftarrow \frac{I g}{4\pi \zeta x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
+S_{\textrm{RFI}} \leftarrow \frac{I g}{4\pi \zeta x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
 \end{equation}
-Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 < x_u,y_u \le 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.
+Here, $S_{\textrm{RFI}}$ is a new \editmark{complex RFI sample that follows a power-law distribution; $\eta$ and $\zeta$ are defined in Eqs.~\eqref{eq:zeta-definition} and \eqref{eq:eta-definition};} $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 < x_u,y_u \le 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.
+% Free-space propagating drawing formula:
+%\begin{equation} \label{eq:drawing-rfi-distribution}
+%S_{\textrm{RFI}} \leftarrow \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
+%\end{equation}
+%where $0 < x_u, y_u \le 1$ are independently drawn from a uniform distribution.
+The result is an amplitude sample $S_{\textrm{RFI}}$, with $S_{\textrm{RFI}} \ge Ig/4\pi r_L^2$, that follows a power-law distribution.
+A sample $S$ that is contaminated by both RFI and noise \editmark{can be drawn with
+$S \leftarrow v_n + w_n i + S_{\textrm{RFI}}$, with $v_n, w_n \sim N(\mu=0; \sigma)$.
+An} example of distribution curves of $S$ for $\eta=2$ and various settings of $Ig/4\pi \zeta r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.
+
+%If we assume that $I$, the average RFI source strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-rfi-with-propagation} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram.
+Although calibration could in theory solve $g$, almost all RFI sources will come in through the far-sidelobes of the beam, and finding the expected values for the gains is therefore hard.
+The effects of these parameters on the histogram will be further discussed in \S\ref{sec:parameter-variability}.
+
+%Fig.~\ref{fig:rayleigh-and-rfi-combined} shows that the left side of the graph corresponds with the original Rayleigh distribution.
+From the right side of the histogram, we are able to estimate $Ig/r_L^2$, as the placing of the power-law curve is independent of the Rayleigh distribution for large $S$.
+Methods to constrain the parameters of the RFI distribution will be discussed in \S\ref{sec:rfi-distribution-constraints}.
+The curves in Fig.~\ref{fig:rayleigh-and-rfi-combined} are from histograms of samples that are all contaminated by both noise and RFI.
+In practice, we can not make this distinction accurately, as RFI detectors have a limited accuracy.
+Consequently, the observed histogram will be the sum of two types of histograms: the first histogram being contaminated by both noise and RFI, the second one only by noise.
 
@@
 Therefore, we chose to have a bin size that increases linearly with the amplitude $S$, and the rate counts are divided by the bin size after counting.
-Consequently, in parts of the histogram that have a sparse number of samples, the outlying samples will follow a $1/S$ curve, or a -1 slope in a log-log plot.
+%Consequently, in parts of the histogram that have a sparse number of samples, the outlying samples will follow a $1/S$ curve, or a -1 slope in a log-log plot.
 This can be seen in the tails of the distributions of Fig.~\ref{fig:rayleigh-and-rfi-combined}.
@@
 \end{equation}
-with $N$ the number of samples and $x_i$ for $0 < i \le N$ the samples that follow a power law with lower bound $x_\textrm{min}$.
-However, this estimator assumes the distribution is not cut-off at a high point.
+with $N$ the number of samples and $x_i$ for $0 < i \le N$ the samples that follow a power law with \editmark{lower bound $x_\textrm{min}$.}
+%However, this estimator assumes the distribution is not cut-off at a high point.
 In our case, the distribution is cut-off at both sides, for example because of the limit of the analogue-digital converter (see \S\ref{sec:rfi-distribution-constraints}).
@@
 
-\subsection{Determining RFI distribution limits} \label{sec:rfi-distribution-constraints}
-In this section we will show methods to find $S_U$ and $S_L$, the upper and lower flux limits of the power-law distribution at which the power law breaks down.
-Once the exponent of the power-law part of the distribution is estimated with the previously discussed techniques, the distribution can be extrapolated to find a lower flux limit.
-Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the following rate density function $h$:
-\begin{equation} \label{eq:amplitude-power-law}
-h(S) = \beta S^\alpha.
-\end{equation}
 \begin{figure}
 \begin{center}
-\includegraphics[width=7.5cm]{img/explanation-lower-constraint/FindingCounts}
+%\includegraphics[width=7.5cm]{img/explanation-lower-constraint/FindingCounts}
 \includegraphics[width=7.5cm]{img/explanation-lower-constraint/FindingConstraints}
@@
 
+\subsection{Determining RFI distribution limits} \label{sec:rfi-distribution-constraints}
+\editmark{In this section we will show how to put upper and lower constraints on the power-law distribution.}
+%show methods to find $S_U$ and $S_L$, the upper and lower flux limits of the power-law distribution at which the power law breaks down.
+Once the exponent of the power-law part of the distribution is estimated with the previously discussed techniques, the distribution can be extrapolated to find a lower flux limit.
+Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the \editmark{rate density function $h(S) = \beta S^\alpha$}.
 $S_1$ and $S_2$ are selected by visual inspection of the histogram.
 Assume that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}, and that the effect of the Rayleigh component on the histogram $> S_1$ is negligible.
-The hypothetical upper limit $S_U$ of the distribution can now be found, i.e., the highest amplitude that would be observed when the distribution follows the power law up to that point, by solving
+The hypothetical upper limit $S_U$ of the distribution can be \editmark{found by} solving
 \begin{equation} \label{eq:upper-limit-constrained}
@@
 \end{equation}
-In practice, the observed histogram will break down beyond some amplitude because of several reasons.
-First and most importantly, the samples themselves are digitized with an analogue-to-digital converter (ADC) with limited range.
-Second, we observe for a limited time and the rate count is discrete.
-Because the chance of finding a sample with a very high intensity is low, it is unlikely to observe samples beyond some amplitude within the finite observing time.
-Finally, under the assumption of a uniform spatial distribution of RFI transmitters, samples with very high amplitude would have to be produced by transmitters that are very close to the telescope.
+The observed histogram will break down beyond some amplitude because \editmark{of several reasons: the samples are digitized with an analogue-to-digital converter (ADC) with limited range; we observe for a limited time and the rate count is discrete; and, under the assumption of a uniform spatial distribution of RFI transmitters, samples with} very high amplitude would have to be produced by transmitters that are very close to the telescope.
 However, it is likely that the uniform spatial distribution of transmitters will break down at closer distances.
@@
 \end{equation}
-In some cases there will be no solution for $S_U$. This can happen when the empirical distribution is not limited at the high end, or when it contains features that are not in the model.
+%In some cases there will be no solution for $S_U$. This can happen when the empirical distribution is not limited at the high end, or when it contains features that are not in the model.
 
-Similar to the calculation of $S_U$ and by using $N_\textrm{total}$ as an upper limit to the total number of samples that are affected by the power-law distribution, one can estimate the lower limit $S_L$.
-For Fig.~\ref{fig:explanation-constraints} this means that the areas labelled $N_a$ and $N_\textrm{total}-N_\textrm{part}$ are equal.
+\editmark{One can estimate the lower limit $S_L$ in a similar way.
+\editmark{This can be solved by assuming the area labelled $N_a$ in Fig.~\ref{fig:explanation-constraints}} equals the number of samples to the left of $S_1$}.
 The area labelled $N_b$ is assumed to be zero for now, which assumes the power law has a sharp cut-off on the left side, e.g., because of the curvature of the Earth.
@@
 
-With the assumption that $Ig/r_L^\eta\sim$ a uniform distribution, the area labelled in Fig.~\ref{fig:explanation-constraints} as $N_b$ is also part of the RFI distribution, and a stronger constraint $\tilde S_L$ can be found by solving
-\begin{equation}
- \int\limits_{\tilde S_L}^{S_U} h(S) dS + \tilde S_{L} h(\tilde S_{L}) = N_\textrm{total} - N_\textrm{part},
-\end{equation}
-which yields
+With the assumption that $Ig/r_L^\eta\sim$ a uniform distribution, the area labelled in Fig.~\ref{fig:explanation-constraints} as $N_b$ is also part of the RFI distribution, and a stronger constraint $\tilde S_L$ can be found\editmark{, yielding}
+% By solving:
+%\begin{equation}
+% \int\limits_{\tilde S_L}^{S_U} h(S) dS + \tilde S_{L} h(\tilde S_{L}) = N_\textrm{total} - N_\textrm{part},
+%\end{equation}
 \begin{equation} \label{eq:lower-limit-2}
@@
 E(S_R) & = \frac{1}{N_{LU}}\int\limits_{S_L}^{S_U} \beta S^\alpha S dS
- = \frac{\beta}{N_{LU}} \left[ \frac{1}{\alpha+2} S^{\alpha+2} \right]_{S_L}^{S_U} \\
- & = \frac{\beta \left( S_U^{\alpha+2} - S_L^{\alpha+2} \right) }{N_{LU} \left( \alpha+2\right) }
+ = \frac{\beta}{N_{LU}} \left[ \frac{1}{\alpha+2} S^{\alpha+2} \right]_{S_L}^{S_U}
+% & = \frac{\beta \left( S_U^{\alpha+2} - S_L^{\alpha+2} \right) }{N_{LU} \left( \alpha+2\right) }
 \end{align}
-Here, $N_{LU}$ is the number of samples between $S_L$ and $S_U$ after normalizing for the bin size:
-\begin{align} \label{eq:n-between-sl-su}
-N_{LU} = \int\limits_{S_L}^{S_U} \beta S^\alpha dS.
-\end{align}
-Substitution and simplification of these two results in
+Here, $N_{LU}$ is the number of samples between $S_L$ and $S_U$ after normalizing for the \editmark{bin size. Evaluating this results in}
+%\begin{align} \label{eq:n-between-sl-su}
+%N_{LU} = \int\limits_{S_L}^{S_U} \beta S^\alpha dS.
+%\end{align}
+%Substitution and simplification of these two results in
 \begin{align} \label{eq:expected-value-rfi}
@@
 We use the AOFlagger for RFI detection, which will be described in Sect.~\ref{sec:dist-data}.
-In
-\citet{lofar-radio-environment} the false-positives rate for the AOFlagger is estimated to be 0.5\%.
 
-An estimate of $S_d$, the average lower limit of detected RFI, can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the real number (true positives) of RFI samples.
+\editmark{Another interesting parameter is $S_d$, the} average lower limit of detected RFI.
+It can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the real number (true positives) of RFI samples.
 Therefore, the limit is calculated similar to Eq.~\eqref{eq:lower-limit-1}, where the term $N_\textrm{part} - N_\textrm{total}$ needs to be replaced with $N_\textrm{RFI}$, which equals the total number of samples detected as RFI minus the false positives.
+\editmark{In \citet{lofar-radio-environment} the false-positives rate for the AOFlagger is estimated to be 0.5\%.}
 
@@
 where $SS_{xx}$, $SS_{xy}$ and $SS_{yy}$ are the sums of squares, e.g., $SS_{xy}=\sum_{i=1}^n (x_i - \bar x) (y_i - \bar y)$ and $n$ is the number of samples.
-However, we found that this is not a representative error in our case, because the errors in the slope are not normally distributed.
-The noise in the part of the histogram over which the slope is calculated, is very small due to the large number of samples.
+However, we found that this is not a representative error in our case, because the errors in the slope are not normally \editmark{distributed.
+%The noise in the part of the histogram over which the slope is calculated, is very small due to the large number of samples.
 Consequently, the estimated standard deviation of $\hat \alpha$ is very low.
 Nevertheless, when the slope is calculated over subsections of the original range over which the slope is calculated, one finds the line is not completely straight and the slope is changing more than is predicted by the standard error.
-Therefore, we introduce an error estimate $\epsilon_\alpha$ that quantifies a normalized standard deviation of the slope over the range.
+Therefore}, we also introduce an error estimate $\epsilon_\alpha$ that quantifies a normalized standard deviation of the slope over the range.
 This error is formed by calculating the slope over $n_\alpha$ smaller sub-ranges in the histogram, creating $n_\alpha$ estimates $\alpha_i$.
-If the errors in $\alpha_i$ are normally distributed with mean zero, the standard deviation over the full range will be $\sqrt{n_\alpha}$ times smaller.
-Therefore, an estimate of the variance of $\hat \alpha$ can be calculated with
+If the errors in $\alpha_i$ are normally distributed with \editmark{zero mean, an} estimate of the variance of $\hat \alpha$ can be calculated with
 \begin{equation}
@@
 
-The Hill estimator of Eq.~\eqref{eq:hill} is a different estimation method for the exponent in the power-law distribution, and yields therefore also a different standard error.
-The standard error of the Hill estimator is \citep{power-law-distribution}
+\editmark{The standard error of the Hill estimator of Eq.~\eqref{eq:hill} is \citep{power-law-distribution}}
 \begin{equation} \label{eq:stderr-hill}
@@
 Because the number of samples is very large ($>10^{11}$), the $\mathcal{O}$-term will be very small.
-Therefore, we will calculate the quantity without this term.
-As with the standard error for the slope of a straight line in Eq.~\eqref{eq:stderr-slope}, the standard error for the Hill estimator yields very small quantities.
+Therefore, we will calculate the quantity without this \editmark{term.} %As with the standard error for the slope of a straight line in Eq.~\eqref{eq:stderr-slope}, the standard error for the Hill estimator yields very small quantities.
 Again, this is because it assumes the underlying power law has a fixed exponential, while in our case the power law seems to vary.
@@
 
-Because our distributions are huge, we decided not to do goodness-of-fit tests, because these require many similar distributions to be simulated in order to reach accurate decision.
+%\editmark{Because} our distributions are huge, we decided not to do goodness-of-fit tests, because these require many similar distributions to be simulated in order to reach accurate decision.
 Instead, we will try to evaluate the distributions visually.
@@
 \citet{lofar-radio-environment}.
-We refer to van Haarlem et al.
-(2013, \textit{submitted}) for a full description of the capabilities of LOFAR.
+We refer to \citet{lofar-overview-2013} for a full description of the capabilities of LOFAR.
 The analyses will cover only Dutch stations.
@@
 As can be seen in Fig.~\ref{fig:plot-dist-per-frequency-LBA}, the Rayleigh-bulges of individual sub-bands are not that wide, but they are not aligned because of the differing noise levels.
-This effect is not so strong in histograms of the HBA sub-bands in Fig.~\ref{fig:plot-dist-per-frequency-HBA}, because the HBA antenna response changes less over frequency.
+%This effect is not so strong in histograms of the HBA sub-bands in Fig.~\ref{fig:plot-dist-per-frequency-HBA}, because the HBA antenna response changes less over frequency.
 
@@
 Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.57;-1.49]$ for the HBA.
-The estimate of the HBA is thus more accurate, because its histogram deviates less from the power law.
 This translates to boundaries on the average transmitter height of $[0.6; 800]$ and $[3.1; 23]$~m for the LBA and HBA respectively.
@@
 With the construction of empirical models for the RFI source distributions, we are one step closer to including these effects in the simulation.
-Using Eq.~\eqref{eq:sample-with-propagation}, one can sample a realistic strength of a single RFI source, add the feature to the data and run the AOFlagger.
+Using Eq.~\eqref{eq:drawing-rfi-with-propagation}, one can sample a realistic strength of a single RFI source, add the feature to the data and run the AOFlagger.
 What is still needed for accurate simulation, is to obtain a likely distribution for the duration that one such source affects the data.
@@
 In theory, faint RFI could impose a fundamental limit on the attainable noise limit of long integrations.
-As an example, we will analyse the situation for the LOFAR EoR project.
+\editmark{We} will analyse the situation for the LOFAR EoR project.
 This project will use the LOFAR high-band antennas to collect on the order of 50--100 night-time observations of 6 h for a few target fields.
@@
 Therefore, the amount that time integration can decrease the flux density of RFI might be limited.
-Additionally, some RFI sources are received by multiple stations of the array, and by multiple feeds of the individual antennas.
+%Additionally, some RFI sources are received by multiple stations of the array, and by multiple feeds of the individual antennas.
 Therefore, integrating data from different interferometers and data from the two polarized feeds might not lower the noise level that is caused by RFI at some point.
 In summary, RFI is unlike normal noise and might add consistent over time, interferometer and feed.
-
-On the other hand, many RFI signals observed in the LOFAR bands have a limited bandwidth.
+\editmark{On the other hand}, many RFI signals observed in the LOFAR bands have a limited bandwidth.
 Indeed, the majority of the detected RFI sources affect only one or a few LOFAR channels of 0.76 kHz.
@@
 This only holds when the observational integrated bandwidth $\Delta \nu$ is substantially higher than the average bandwidth of a single RFI source.
-Otherwise, if the $\Delta \nu$ is small relative to the average bandwidth of RFI sources, some RFI might show up earlier.
 The empirically found upper limits in this work are $\textrm{REFD}_\textrm{LBA}=18.9$--$19.3$~Jy and $\textrm{REFD}_\textrm{HBA}=6.5$--$6.7$~Jy (see Table~\ref{table:dist-data-quantities}).
@@
 \subsection{Interference-reducing effects} \label{sec:coherence-reduction}
-When integrating data, it is likely that the actual noise limit from low-level RFI will be significantly lower than the given upper limit, which was determined at highest LOFAR resolution, before further averaging.
-There are several reasons for this, which we will summarize one by one.
+When integrating data, it is likely that the actual noise limit from low-level RFI will be significantly lower than the given upper limit, which was determined at highest LOFAR resolution.
+\editmark{There are several reasons for this: Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes; many RFI sources will be seen by only a few stations or are not constant over time; for the shortest baselines at 150~MHz, the far field starts around 1~km, and some RFI sources will therefore be in the near field; and finally, a large number of stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other.}
 
-There are several effects related to the origin of the interference, which can reduce their effect.
-One such effect, is when the RFI sources have a variable geometric phase.
-Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes.
-This would cause them to sum (partly) inconsistently over time, and thus go approximately down with the noise.
-Moreover, many RFI sources will be seen by only a few stations or are not constant over time, which further attenuates their effect.
-Finally, for the shortest baselines at 150~MHz, the far field starts around 1~km.
-Some RFI sources will therefore be in the near field, especially for the longer baselines.
-In that case, a source will not add up coherently over the interferometers that see the particular source, as the interferometers see it with different phases.
-
-A large number of RFI sources together can also behave like normal incoherent noise.
-Stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other.
-Individually, the sources will have fixed geometric phases, but because the baselines are much longer than the wavelength, their phases will become uniformly spread.
-Therefore, they will add inconsistently.
-
-Some standard data processing actions can also attenuate RFI, especially fringe stopping and the imaging of data.
-Fringe stopping interferometers can partly average out RFI sources.
+\editmark{Fringe} stopping interferometers can partly average out RFI sources.
 Nevertheless, stationary RFI that is averaged out by fringe stopping will leave artefacts in the field centre \citep{post-correlation-filtering}.
 This is not relevant when observing the North Celestial Pole --- which is one of the LOFAR EoR fields --- because no fringe stopping is applied when observing the NCP.
-Imaging sometimes attenuates RFI, because the Fourier transform that is involved in data imaging will localize the contribution from stationary RFI near the NCP.
+\editmark{Imaging of the interferometric data} will localize the contribution from stationary RFI near the NCP.
 If RFI artefacts would show in the image of the NCP field, they can easily be detected and possibly be removed, or processing could ignore data near the pole.
-
-Because of the above two arguments, when considering RFI it is a risk to use the NCP as one of the EoR target fields.
+Because of these arguments, when considering RFI it is a risk to use the NCP as one of the EoR target fields.
 At the same time, this field is useful for analysing the RFI coherency properties.
-It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright celestial foreground sources.
 Preliminary analysis of EoR NCP observations of a single night have almost reached the thermal noise, but do not show leaked RFI at the pole \citep[\S4.3]{ncp-eor-yatawatta}.
 
-On top of all these possible RFI reducing effects, our given RFI constraints could be too pessimistic because we have assumed 100\% of the spectrum is occupied by RFI over the entire duration of the observation.
+\editmark{Our given RFI constraints could be too pessimistic, because} we have assumed 100\% of the spectrum is occupied by RFI over the entire duration of the observation.
 If, say, only 10\% of the samples are affected by RFI, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all.
@@
 
-Finally, future RFI excision strategies will further enhance detection accuracy.
-Currently, RFI excision is applied only on the raw data from single interferometers at high resolution.
-Once data from a large number of nights are collected, it will be possible to detect and excise RFI more accurately, by looking at the averaged data from multiple nights and/or multiple interferometers.
-We have shown that the current detection algorithms can detect RFI well below the noise.
-Therefore, if data from different nights or interferometers are summed, and there is stationary RFI in the set, it will become detectable if it is consistently present.
-If it is still below the noise and not detectable, but behaves similar to the RFI we are already seeing, it will act like normal noise and will therefore be harmless.
-
-With the current strategy, it is likely that the LOFAR EoR project will encounter some RFI on some frequencies when averaging lots of observing nights, although this still remains to be seen.
+\editmark{Finally, future RFI excision strategies can further enhance detection accuracy.
+Once data from a large number of nights are collected, it will be possible to detect and excise RFI more accurately.
+With} the current strategy, it is likely that the LOFAR EoR project will encounter some RFI on some frequencies when averaging lots of observing nights, although this still remains to be seen.
 To mitigate this leaked RFI, the detection can be executed at higher signal-to-noise levels.

### 2013-07-08 articles/Article-RFI-distribution/article-rfi-distribution.tex
Ready for resubmit\!

@@
 
+%\newcommand{\editmark}[1]{#1}
 \newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
@@
 
+\begin{figure*}
+\begin{center}
+\includegraphics[height=8cm]{img/TF-L43786_L220-rfi-example-xaxistext.pdf}
+\caption{A dynamic spectrum of a small part of an observation.
+The features with significantly higher values are caused by interference.
+Some of these have a constant frequency, while others are more erratic.}
+\label{fig:dist-rfi-example}
+\end{center}
+\end{figure*}
+
 In time-frequency space, interfering sources can have complex structures.
@@
 After that, we will analyse the issues that arise when measuring the distribution by counting samples.
-
-\begin{figure*}
-\begin{center}
-\includegraphics[height=8cm]{img/TF-L43786_L220-rfi-example-xaxistext.pdf}
-\caption{A dynamic spectrum of a small part of an observation.
-The features with significantly higher values are caused by interference.
-Some of these have a constant frequency, while others are more erratic.}
-\label{fig:dist-rfi-example}
-\end{center}
-\end{figure*}
 
@@
 
-Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we expect to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane.
-For this, we have assumed that the variables $I$, $g$ and $\rho$ are constants.
+Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we expect to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional \editmark{plane.} %For this, we have assumed that the variables $I$, $g$ and $\rho$ are constants.
 In reality, these variables have a stochastic nature.
@@
 A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}.
-The original model is based on urban areas, but corrections are given by Hata for suburban areas with a lower population density and for open areas.
-Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss between two ground-based antennas:
+%The original model is based on urban areas, but corrections are given by Hata for suburban areas with a lower population density and for open areas.
+\editmark{Hata} gives the following analytical estimate for $L_p$, the electromagnetic propagation loss between two ground-based antennas:
 \begin{align} 
@@
 In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.
+
+\editmark{We note that a uniform distribution of meteors or aircrafts which reflect free-space propagating RFI can create a power-law distribution with a similar slope: a uniform two-dimensional distribution of reflecting sources will create a $-1.5$ slope, while a uniform three-dimensional distribution will create a $-1.75$ slope.
+With brightness-distribution analyses one can therefore not distinguish between transmitters affected by Hata's propagation model and reflectors affected by free-space propagation.
+Reflected RFI might become relevant at lower amplitude levels.}
 
@@
 Otherwise, a turn-over point should be visible in the histogram.
+
+\editmark{From Eq.~\eqref{eq:frequency-density-with-propagation} it can be seen that the $\rho$, $I$ and $g$ parameters have the same effect of scaling the power-law distribution, and do not change its shape or slope.
+Therefore, with distribution analyses one can e.g.
+not determine whether the distribution is dominated by low-power sources within the horizon or by scattered signals from over the horizon.
+The horizon of an antenna is estimated 	with $\sqrt{2hr}$ \citep{bullington-propagation}, with $r$ the radius of the Earth and $h$ the height of the antenna.
+For LOFAR, the horizon is at about 5~km.}
 
@@
 There are no 79~m high transmitters nearby LOFAR stations in the LBA frequency range.
-Additionally, Hata's model only goes down to 150~MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies, e.g.
-because the effect of the ionosphere becomes stronger.
-Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.57;-1.49]$ for the HBA.
-This translates to boundaries on the average transmitter height of $[0.6; 800]$ and $[3.1; 23]$~m for the LBA and HBA respectively.
+Additionally, Hata's model only goes down to 150~MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies.
+Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.57;-1.49]$ for the HBA\editmark{, giving} average transmitter heights of $[0.6; 800]$ and $[3.1; 23]$~m for the LBA and HBA respectively.
 Therefore, the LBA measurements are clearly not accurate enough to be conclusive.
@@
 \subsection{Implications for very long integrations}
-In theory, faint RFI could impose a fundamental limit on the attainable noise limit of long integrations.
+Faint RFI could impose a fundamental limit on the attainable noise limit of long integrations.
 \editmark{We} will analyse the situation for the LOFAR EoR project.
@@
 In summary, RFI is unlike normal noise and might add consistent over time, interferometer and feed.
-\editmark{On the other hand}, many RFI signals observed in the LOFAR bands have a limited bandwidth.
-Indeed, the majority of the detected RFI sources affect only one or a few LOFAR channels of 0.76 kHz.
+\editmark{On the other hand}, many RFI signals observed in the LOFAR bands have a limited bandwidth, and the majority of the detected RFI sources affect only one or a few LOFAR channels of 0.76 kHz.
 Therefore, frequency averaging will lower the flux density of the RFI signal.
@@
 When integrating data, it is likely that the actual noise limit from low-level RFI will be significantly lower than the given upper limit, which was determined at highest LOFAR resolution.
-\editmark{There are several reasons for this: Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes; many RFI sources will be seen by only a few stations or are not constant over time; for the shortest baselines at 150~MHz, the far field starts around 1~km, and some RFI sources will therefore be in the near field; and finally, a large number of stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other.}
+\editmark{There are several reasons for this: Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes; many RFI sources will be seen by only a few stations or are not constant over time; for the shortest baselines at 150~MHz, the far field starts around 1~km, and some RFI sources will therefore be in the near field; and finally, a large number of stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other.
+These arguments are valid only for interferometric arrays.
+Global EoR experiments that use a single antenna will not benefit from these effects, and will still be limited by low-level RFI.}
 
@@
 This is not relevant when observing the North Celestial Pole --- which is one of the LOFAR EoR fields --- because no fringe stopping is applied when observing the NCP.
-\editmark{Imaging of the interferometric data} will localize the contribution from stationary RFI near the NCP.
+\editmark{Imaging of the data} will localize the contribution from stationary RFI near the NCP.
 If RFI artefacts would show in the image of the NCP field, they can easily be detected and possibly be removed, or processing could ignore data near the pole.
-Because of these arguments, when considering RFI it is a risk to use the NCP as one of the EoR target fields.
+Because of these arguments, it is a risk to use the NCP as one of the EoR target fields.
 At the same time, this field is useful for analysing the RFI coherency properties.
@@
 
-\editmark{Our given RFI constraints could be too pessimistic, because} we have assumed 100\% of the spectrum is occupied by RFI over the entire duration of the observation.
-If, say, only 10\% of the samples are affected by RFI, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all.
+\editmark{Because we have assumed 100\% of the spectrum is occupied by RFI, our given RFI constraints could be too pessimistic.
+If} only 10\% of the samples are affected by RFI, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all.
 With current data, one can only speculate how much of the electromagnetic spectrum is truly occupied.

### 2013-07-16 articles/Article-RFI-distribution/article-rfi-distribution.tex
Fixing final comment by referee

@@
 \newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
+
+% To make Dutch ``tussenvoegsels'' work correctly in Latex such as ``de Bruyn'', we use this command
+% It fixes ordering and uppercases.
+% In text, it should be written with uppercase, as in ``De Bruyn''
+\DeclareRobustCommand{\TUSSEN}[3]{#2}
 
@@
 
-\editmark{It is common knowledge that in a uniform Euclidian Universe source counts behave like power-law distributions.
-For sources on a flat surface, the power law will have a -2 exponent}.
+It is common knowledge that in a uniform Euclidian Universe source counts behave like power-law distributions.
+\editmark{The differential distribution for sources on a flat surface is a power law with -2 exponent}.
 We will derive this expected intrinsic source distribution for interfering radio sources.
@@
 
+% To make Dutch ``tussenvoegsels'' work correctly in Latex such as ``de Bruyn'', we use this command
+% In bibliography, it should be written with lowercase, as in ``de Bruyn''
+\DeclareRobustCommand{\TUSSEN}[3]{#3}
+
 \bibliographystyle{mn2e}

### 2013-07-16 articles/Article-RFI-distribution/article-rfi-distribution.tex
Removed all editmarks and extra comments

@@
 
-\usepackage[colorlinks=true,linkcolor=customlinkcolor,urlcolor=customlinkcolor,citecolor=customcitecolor,pdftex]{hyperref}
+%\usepackage[colorlinks=true,linkcolor=customlinkcolor,urlcolor=customlinkcolor,citecolor=customcitecolor,pdftex]{hyperref}
+\usepackage[pdftex]{hyperref}
 
@@
 %\newcommand{\editmark}[1]{#1}
-\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
+%\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
 
@@
 
-In this article, we will connect new insights about RFI to the angular \editmark{EoR} experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, \citealt{lofar-overview-2013}).
+In this article, we will connect new insights about RFI to the angular EoR experiment that is using the Low-Frequency Array (LOFAR) (\citealt{de-bruyn-eor-ursi-2011}, \citealt{lofar-overview-2013}).
 The LOFAR EoR project aims to detect the redshifted 21-cm signals from the EoR using the LOFAR HBA antennas (115--190~MHz, $z_\textrm{HI}$=11.4--6.5).
@@
 Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse brightness histograms of LOFAR data.
-\editmark{Sect.~\ref{sec:dist-data} describes} the two LOFAR data sets that have been used to perform the experiment.
+Sect.~\ref{sec:dist-data} describes the two LOFAR data sets that have been used to perform the experiment.
 The results of analysing the sets are presented in Sect.~\ref{sec:dist-results}.
@@
 It is common knowledge that in a uniform Euclidian Universe source counts behave like power-law distributions.
-\editmark{The differential distribution for sources on a flat surface is a power law with -2 exponent}.
+The differential distribution for sources on a flat surface is a power law with -2 exponent.
 We will derive this expected intrinsic source distribution for interfering radio sources.
@@
 Next, we assume that all RFI sources have equal constant strength $I$ and follow a uniform spatial distribution in the local two-dimensional horizontal plane.
-\editmark{These assumptions are obviously simplications, but we will address these later.
-%This is obviously a simplification, as the RFI sources are actually distributed on the surface of a sphere with different heights.
-Beyond some distance, the Earth will partly block the path between transmitter and \editmark{receiver.
-Using} these assumptions, we can express the expected \editmark{inverse cumulative rate density of sources at distance $r$ as
+These assumptions are obviously simplications, but we will address these later.
+Using these assumptions, we can express the expected inverse cumulative rate density of sources at distance $r$ as
 \begin{equation} \label{eq:inverse-cumulative-distance-function}
@@
 \end{equation}
-with $N$ the total number of sources and for some constant $\rho$ that represents the number of sources per unit area.}
-%In other words, we will observe $F$ sources that are further than a distance $r$ away.
-We need $F_{\textrm{distance} \ge r}(r)$, the complement of Eq.~\eqref{eq:cumulative-distance-function} to calculate the amplitude distribution, because sources with an amplitude of \emph{at most} a given strength will have a distance of \emph{at least} some distance.
-Sources with at least a distance of $r$ are given by $F_{\textrm{distance} \ge r}(r) = N - \rho \pi r^2$, with $N$ the total number of sources.
-Because $r$ is restricted, $N$ is finite.
-The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can \editmark{be calculated from this with}
+with $N$ the total number of sources and for some constant $\rho$ that represents the number of sources per unit area.
+The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can be calculated from this with
 \begin{align} \label{eq:cumulative-rfi-distribution}
@@
 
-Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we expect to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional \editmark{plane.} %For this, we have assumed that the variables $I$, $g$ and $\rho$ are constants.
-In reality, these variables have a stochastic nature.
-The effect of this will be discussed in \S\ref{sec:parameter-variability}.
-
-%\subsection{Spherical case} \label{sec:spherical-geometry}
-%If we do take into account the fact that the sources lie on a sphere instead of a plane, Eq.~\eqref{eq:cumulative-distance-function} changes.
-If we assume their difference in height is negligible, we can use the formula for the surface area of a spherical cap,
-%\begin{equation}\label{eq:spherical-cap-area}
-% A_\textrm{cap} = \pi \left( a^2 + h^2 \right) \hspace{0.4cm}\textrm{with } (a \le R) \wedge (h \le 2R),
-%\end{equation}
-%with $R$ the radius of the Earth, $a$ the radius of the base of the cap and $h$ the height of the cap.
-Because the Euclidean distance $r$ between the source and the antenna is $r^2 = a^2 + h^2$, the resulting number of sources within Euclidean distance $r$ becomes
-%\begin{equation} \label{eq:cumulative-spherical}
-%\tilde{F}_{\textrm{distance} \le r}(r) = \rho A_\textrm{cap} = \rho \pi r^2 \hspace{0.4cm}\textrm{with } r \le 2R,
-%\end{equation}
-%which equals the formula for a planar source distribution given in Eq.~\eqref{eq:cumulative-distance-function}.
-Therefore, the spherical and two-dimensional cases lead to a distribution with equal power law.
-Eq.~\eqref{eq:cumulative-spherical} is valid for $r \le 2R$, but in practice the lowest amplitude of observed sources is further constrained because of the curvature of the Earth.
-This causes distant sources to settle below the horizon.
-Nevertheless, some of those sources might still be visible due to reflection of the ionosphere and other (semi-)spherical propagation effects.
+Therefore, if we plot the histogram of the RFI amplitudes in a $\log$-$\log$ plot, we expect to see a power law with a slope of $-2$ over the interval in which the RFI sources are spread like uniform sources on a two-dimensional plane.
 
@@
 So far, we have assumed that the electromagnetic radiation propagates through free space, resulting in an $r^{-2}$ fall-off.
-In reality, the radiation will be affected by complicated propagational effects \editmark{due to the surface of the Earth}.
-%Because of the irregular surface of the Earth (including urban areas) and the absence of a direct line of sight between transmitter and receiver, the propagation mode might be indirect.
-Multiple indirect paths might be formed that include reflection, refraction, diffraction and absorption of the electromagnetic wave.
+In reality, the radiation will be affected by complicated propagational effects due to the surface of the Earth.
 A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}.
-%The original model is based on urban areas, but corrections are given by Hata for suburban areas with a lower population density and for open areas.
-\editmark{Hata} gives the following analytical estimate for $L_p$, the electromagnetic propagation loss between two ground-based antennas:
+Hata gives the following analytical estimate for $L_p$, the electromagnetic propagation loss between two ground-based antennas:
 \begin{align} 
@@
 
-Converting from a subtracted term in decibels to a flux density factor $L_S$ \editmark{results in}
+Converting from a subtracted term in decibels to a flux density factor $L_S$ results in
 \begin{equation} \label{eq:propagation-loss-factor}
@@
 \end{equation}
-\editmark{with $\eta$ and $\zeta$ given by}
+with $\eta$ and $\zeta$ given by
 \begin{align} 
@@
 \end{equation}
-\editmark{The} resulting rate density function $f_p$ is
+The resulting rate density function $f_p$ is
 % S(r):=I*g/(r^alpha);
@@
 \end{figure}
-\editmark{This} yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively.
+This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively.
 In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.
 
-\editmark{We note that a uniform distribution of meteors or aircrafts which reflect free-space propagating RFI can create a power-law distribution with a similar slope: a uniform two-dimensional distribution of reflecting sources will create a $-1.5$ slope, while a uniform three-dimensional distribution will create a $-1.75$ slope.
+We note that a uniform distribution of meteors or aircrafts which reflect free-space propagating RFI can create a power-law distribution with a similar slope: a uniform two-dimensional distribution of reflecting sources will create a $-1.5$ slope, while a uniform three-dimensional distribution will create a $-1.75$ slope.
 With brightness-distribution analyses one can therefore not distinguish between transmitters affected by Hata's propagation model and reflectors affected by free-space propagation.
-Reflected RFI might become relevant at lower amplitude levels.}
+Reflected RFI might become relevant at lower amplitude levels.
 
@@
 So far, these are the expected histograms for pure noise and pure RFI that propagates through free space.
-However, the measured distribution is a mixture of the \editmark{two.
-%A perfect RFI detector would classify the samples in two groups; one group of samples that are not affected by RFI, and therefore contain noise only, and one group of samples that are contaminated, and are the sum of RFI and noise.
-Real RFI detectors can classify samples into these groups to some extent, but due to false positives and false negatives, the groups will get mixed nevertheless.
-Even more, we still need to take into account that the samples classified as RFI are also affected by noise.
-Although this effect is moderate, because `most' RFI samples are of much higher amplitude than what is added because of the noise, we are interested in the low-level RFI samples, and the relative effect of noise on those samples is large.
-Analytic derivation of the corresponding mixed amplitude distribution is not trivial, but the distributions can easily be} estimated by drawing complex samples from the two distributions and calculating and counting the amplitudes.
+However, the measured distribution is a mixture of the two.
+Analytic derivation of the corresponding mixed amplitude distribution is not trivial, but the distributions can easily be estimated by drawing complex samples from the two distributions and calculating and counting the amplitudes.
 A sample can be drawn from the RFI distribution by integration, scaling and inversion of the rate density function in Eq.~\eqref{eq:frequency-density-with-propagation}.
-%Note however that $F_{\textrm{amplitude}\le S}(S)$ is not limited; when decreasing the amplitude $S$ towards zero, the number of RFI sources will go to infinity.
-\editmark{To invert the cumulative function, one needs to assume that there are no sources beyond some limiting distance $r_L$}.
+To invert the cumulative function, one needs to assume that there are no sources beyond some limiting distance $r_L$.
 With this assumption, a single complex RFI contaminated sample $S_{\textrm{RFI}}$ can be sampled with:
@@
 \end{equation}
-Here, $S_{\textrm{RFI}}$ is a new \editmark{complex RFI sample that follows a power-law distribution; $\eta$ and $\zeta$ are defined in Eqs.~\eqref{eq:zeta-definition} and \eqref{eq:eta-definition};} $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 < x_u,y_u \le 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.
-% Free-space propagating drawing formula:
-%\begin{equation} \label{eq:drawing-rfi-distribution}
-%S_{\textrm{RFI}} \leftarrow \frac{Ig}{4\pi x_u r_L^2} e^{i 2 \pi y_u},
-%\end{equation}
-%where $0 < x_u, y_u \le 1$ are independently drawn from a uniform distribution.
-The result is an amplitude sample $S_{\textrm{RFI}}$, with $S_{\textrm{RFI}} \ge Ig/4\pi r_L^2$, that follows a power-law distribution.
-A sample $S$ that is contaminated by both RFI and noise \editmark{can be drawn with
+Here, $S_{\textrm{RFI}}$ is a new complex RFI sample that follows a power-law distribution; $\eta$ and $\zeta$ are defined in Eqs.~\eqref{eq:zeta-definition} and \eqref{eq:eta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 < x_u,y_u \le 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.
+A sample $S$ that is contaminated by both RFI and noise can be drawn with
 $S \leftarrow v_n + w_n i + S_{\textrm{RFI}}$, with $v_n, w_n \sim N(\mu=0; \sigma)$.
-An} example of distribution curves of $S$ for $\eta=2$ and various settings of $Ig/4\pi \zeta r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.
-
-%If we assume that $I$, the average RFI source strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-rfi-with-propagation} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram.
-Although calibration could in theory solve $g$, almost all RFI sources will come in through the far-sidelobes of the beam, and finding the expected values for the gains is therefore hard.
-The effects of these parameters on the histogram will be further discussed in \S\ref{sec:parameter-variability}.
-
-%Fig.~\ref{fig:rayleigh-and-rfi-combined} shows that the left side of the graph corresponds with the original Rayleigh distribution.
-From the right side of the histogram, we are able to estimate $Ig/r_L^2$, as the placing of the power-law curve is independent of the Rayleigh distribution for large $S$.
-Methods to constrain the parameters of the RFI distribution will be discussed in \S\ref{sec:rfi-distribution-constraints}.
-The curves in Fig.~\ref{fig:rayleigh-and-rfi-combined} are from histograms of samples that are all contaminated by both noise and RFI.
-In practice, we can not make this distinction accurately, as RFI detectors have a limited accuracy.
-Consequently, the observed histogram will be the sum of two types of histograms: the first histogram being contaminated by both noise and RFI, the second one only by noise.
+An example of distribution curves of $S$ for $\eta=2$ and various settings of $Ig/4\pi \zeta r_L^2$ is given in Fig.~\ref{fig:rayleigh-and-rfi-combined}.
 
@@
 
-\editmark{From Eq.~\eqref{eq:frequency-density-with-propagation} it can be seen that the $\rho$, $I$ and $g$ parameters have the same effect of scaling the power-law distribution, and do not change its shape or slope.
+From Eq.~\eqref{eq:frequency-density-with-propagation} it can be seen that the $\rho$, $I$ and $g$ parameters have the same effect of scaling the power-law distribution, and do not change its shape or slope.
 Therefore, with distribution analyses one can e.g.
@@
 The horizon of an antenna is estimated with $\sqrt{2hr}$ \citep{bullington-propagation}, with $r$ the radius of the Earth and $h$ the height of the antenna.
-For LOFAR, the horizon is at about 5~km.}
+For LOFAR, the horizon is at about 5~km.
 
@@
 \end{equation}
-with $N$ the number of samples and $x_i$ for $0 < i \le N$ the samples that follow a power law with \editmark{lower bound $x_\textrm{min}$.}
-%However, this estimator assumes the distribution is not cut-off at a high point.
-In our case, the distribution is cut-off at both sides, for example because of the limit of the analogue-digital converter (see \S\ref{sec:rfi-distribution-constraints}).
-Therefore, using this estimator will result in an estimate that is steeper (i.e., more negative) than the actual distribution.
+with $N$ the number of samples and $x_i$ for $0 < i \le N$ the samples that follow a power law with lower bound $x_\textrm{min}$.
 
@@
 \begin{center}
-%\includegraphics[width=7.5cm]{img/explanation-lower-constraint/FindingCounts}
 \includegraphics[width=7.5cm]{img/explanation-lower-constraint/FindingConstraints}
@@
 \subsection{Determining RFI distribution limits} \label{sec:rfi-distribution-constraints}
-\editmark{In this section we will show how to put upper and lower constraints on the power-law distribution.}
-%show methods to find $S_U$ and $S_L$, the upper and lower flux limits of the power-law distribution at which the power law breaks down.
-Once the exponent of the power-law part of the distribution is estimated with the previously discussed techniques, the distribution can be extrapolated to find a lower flux limit.
-Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the \editmark{rate density function $h(S) = \beta S^\alpha$}.
+In this section we will show how to put upper and lower constraints on the power-law distribution.
+Assume that we have found a power law with exponent $\alpha$ and factor $\beta$ over an amplitude region $[S_1 ; S_2]$, resulting in the rate density function $h(S) = \beta S^\alpha$.
 $S_1$ and $S_2$ are selected by visual inspection of the histogram.
 Assume that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}, and that the effect of the Rayleigh component on the histogram $> S_1$ is negligible.
-The hypothetical upper limit $S_U$ of the distribution can be \editmark{found by} solving
+The hypothetical upper limit $S_U$ of the distribution can be found by solving
 \begin{equation} \label{eq:upper-limit-constrained}
@@
 \end{equation}
-The observed histogram will break down beyond some amplitude because \editmark{of several reasons: the samples are digitized with an analogue-to-digital converter (ADC) with limited range; we observe for a limited time and the rate count is discrete; and, under the assumption of a uniform spatial distribution of RFI transmitters, samples with} very high amplitude would have to be produced by transmitters that are very close to the telescope.
+The observed histogram will break down beyond some amplitude because of several reasons: the samples are digitized with an analogue-to-digital converter (ADC) with limited range; we observe for a limited time and the rate count is discrete; and, under the assumption of a uniform spatial distribution of RFI transmitters, samples with very high amplitude would have to be produced by transmitters that are very close to the telescope.
 However, it is likely that the uniform spatial distribution of transmitters will break down at closer distances.
@@
 
-\editmark{One can estimate the lower limit $S_L$ in a similar way.
-\editmark{This can be solved by assuming the area labelled $N_a$ in Fig.~\ref{fig:explanation-constraints}} equals the number of samples to the left of $S_1$}.
+One can estimate the lower limit $S_L$ in a similar way.
+This can be solved by assuming the area labelled $N_a$ in Fig.~\ref{fig:explanation-constraints} equals the number of samples to the left of $S_1$.
 The area labelled $N_b$ is assumed to be zero for now, which assumes the power law has a sharp cut-off on the left side, e.g., because of the curvature of the Earth.
@@
 
-With the assumption that $Ig/r_L^\eta\sim$ a uniform distribution, the area labelled in Fig.~\ref{fig:explanation-constraints} as $N_b$ is also part of the RFI distribution, and a stronger constraint $\tilde S_L$ can be found\editmark{, yielding}
+With the assumption that $Ig/r_L^\eta\sim$ a uniform distribution, the area labelled in Fig.~\ref{fig:explanation-constraints} as $N_b$ is also part of the RFI distribution, and a stronger constraint $\tilde S_L$ can be found, yielding
 % By solving:
@@
 \end{equation}
-
 With estimates of $\alpha$, $\beta$, $S_L$ and $S_U$, one has obtained a parametrization of the RFI distribution.
@@
 \end{align}
-Here, $N_{LU}$ is the number of samples between $S_L$ and $S_U$ after normalizing for the \editmark{bin size. Evaluating this results in}
+Here, $N_{LU}$ is the number of samples between $S_L$ and $S_U$ after normalizing for the bin size. Evaluating this results in
 %\begin{align} \label{eq:n-between-sl-su}
@@
 
-\editmark{Another interesting parameter is $S_d$, the} average lower limit of detected RFI.
+Another interesting parameter is $S_d$, the average lower limit of detected RFI.
 It can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the real number (true positives) of RFI samples.
 Therefore, the limit is calculated similar to Eq.~\eqref{eq:lower-limit-1}, where the term $N_\textrm{part} - N_\textrm{total}$ needs to be replaced with $N_\textrm{RFI}$, which equals the total number of samples detected as RFI minus the false positives.
-\editmark{In \citet{lofar-radio-environment} the false-positives rate for the AOFlagger is estimated to be 0.5\%.}
+In \citet{lofar-radio-environment} the false-positives rate for the AOFlagger is estimated to be 0.5\%.
 
@@
 where $SS_{xx}$, $SS_{xy}$ and $SS_{yy}$ are the sums of squares, e.g., $SS_{xy}=\sum_{i=1}^n (x_i - \bar x) (y_i - \bar y)$ and $n$ is the number of samples.
-However, we found that this is not a representative error in our case, because the errors in the slope are not normally \editmark{distributed.
-%The noise in the part of the histogram over which the slope is calculated, is very small due to the large number of samples.
-Consequently, the estimated standard deviation of $\hat \alpha$ is very low.
-Nevertheless, when the slope is calculated over subsections of the original range over which the slope is calculated, one finds the line is not completely straight and the slope is changing more than is predicted by the standard error.
-Therefore}, we also introduce an error estimate $\epsilon_\alpha$ that quantifies a normalized standard deviation of the slope over the range.
+However, we found that this is not a representative error in our case, because the errors in the slope are not normally distributed.
+Therefore, we also introduce an error estimate $\epsilon_\alpha$ that quantifies a normalized standard deviation of the slope over the range.
 This error is formed by calculating the slope over $n_\alpha$ smaller sub-ranges in the histogram, creating $n_\alpha$ estimates $\alpha_i$.
-If the errors in $\alpha_i$ are normally distributed with \editmark{zero mean, an} estimate of the variance of $\hat \alpha$ can be calculated with
+If the errors in $\alpha_i$ are normally distributed with zero mean, an estimate of the variance of $\hat \alpha$ can be calculated with
 \begin{equation}
@@
 
-\editmark{The standard error of the Hill estimator of Eq.~\eqref{eq:hill} is \citep{power-law-distribution}}
+The standard error of the Hill estimator of Eq.~\eqref{eq:hill} is \citep{power-law-distribution}
 \begin{equation} \label{eq:stderr-hill}
@@
 \end{equation}
-Because the number of samples is very large ($>10^{11}$), the $\mathcal{O}$-term will be very small.
-Therefore, we will calculate the quantity without this \editmark{term.} %As with the standard error for the slope of a straight line in Eq.~\eqref{eq:stderr-slope}, the standard error for the Hill estimator yields very small quantities.
-Again, this is because it assumes the underlying power law has a fixed exponential, while in our case the power law seems to vary.
-Therefore, this value is given only for completeness.
-
-%\editmark{Because} our distributions are huge, we decided not to do goodness-of-fit tests, because these require many similar distributions to be simulated in order to reach accurate decision.
-Instead, we will try to evaluate the distributions visually.
+Because the number of samples is very large ($>10^{11}$), the $\mathcal{O}$-term will be very small. Therefore, we will calculate the quantity without this term.
 
 \section{Data description} \label{sec:dist-data}
-% TODO correct reference if van Haarlem et al is published
-We have analysed the distributions of two data sets. Both data sets are 24-h LOFAR RFI surveys and are extensively analysed in
-\citet{lofar-radio-environment}.
+We have analysed the distributions of two data sets.
+Both data sets are 24-h LOFAR RFI surveys and are extensively analysed in \citet{lofar-radio-environment}.
 We refer to \citet{lofar-overview-2013} for a full description of the capabilities of LOFAR.
@@
 Additionally, Hata's model only goes down to 150~MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies.
-Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.57;-1.49]$ for the HBA\editmark{, giving} average transmitter heights of $[0.6; 800]$ and $[3.1; 23]$~m for the LBA and HBA respectively.
+Intervals for the exponents with representative $3\sigma$ boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.57;-1.49]$ for the HBA, giving average transmitter heights of $[0.6; 800]$ and $[3.1; 23]$~m for the LBA and HBA respectively.
 Therefore, the LBA measurements are clearly not accurate enough to be conclusive.
@@
 Faint RFI could impose a fundamental limit on the attainable noise limit of long integrations.
-\editmark{We} will analyse the situation for the LOFAR EoR project.
+We will analyse the situation for the LOFAR EoR project.
 This project will use the LOFAR high-band antennas to collect on the order of 50--100 night-time observations of 6 h for a few target fields.
@@
 Therefore, the amount that time integration can decrease the flux density of RFI might be limited.
-%Additionally, some RFI sources are received by multiple stations of the array, and by multiple feeds of the individual antennas.
-Therefore, integrating data from different interferometers and data from the two polarized feeds might not lower the noise level that is caused by RFI at some point.
-In summary, RFI is unlike normal noise and might add consistent over time, interferometer and feed.
-\editmark{On the other hand}, many RFI signals observed in the LOFAR bands have a limited bandwidth, and the majority of the detected RFI sources affect only one or a few LOFAR channels of 0.76 kHz.
+On the other hand, many RFI signals observed in the LOFAR bands have a limited bandwidth, and the majority of the detected RFI sources affect only one or a few LOFAR channels of 0.76 kHz.
 Therefore, frequency averaging will lower the flux density of the RFI signal.
@@
 When integrating data, it is likely that the actual noise limit from low-level RFI will be significantly lower than the given upper limit, which was determined at highest LOFAR resolution.
-\editmark{There are several reasons for this: Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes; many RFI sources will be seen by only a few stations or are not constant over time; for the shortest baselines at 150~MHz, the far field starts around 1~km, and some RFI sources will therefore be in the near field; and finally, a large number of stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other.
+There are several reasons for this: Many RFI sources have a variable geometric phase, because they move or because their path of propagation changes; many RFI sources will be seen by only a few stations or are not constant over time; for the shortest baselines at 150~MHz, the far field starts around 1~km, and some RFI sources will therefore be in the near field; and finally, a large number of stationary RFI sources in a uniform spatial distribution will interfere both constructively and destructively with each other.
 These arguments are valid only for interferometric arrays.
-Global EoR experiments that use a single antenna will not benefit from these effects, and will still be limited by low-level RFI.}
+Global EoR experiments that use a single antenna will not benefit from these effects, and will still be limited by low-level RFI.
 
-\editmark{Fringe} stopping interferometers can partly average out RFI sources.
+Fringe stopping interferometers can partly average out RFI sources.
 Nevertheless, stationary RFI that is averaged out by fringe stopping will leave artefacts in the field centre \citep{post-correlation-filtering}.
 This is not relevant when observing the North Celestial Pole --- which is one of the LOFAR EoR fields --- because no fringe stopping is applied when observing the NCP.
-\editmark{Imaging of the data} will localize the contribution from stationary RFI near the NCP.
+Imaging of the data will localize the contribution from stationary RFI near the NCP.
 If RFI artefacts would show in the image of the NCP field, they can easily be detected and possibly be removed, or processing could ignore data near the pole.
@@
 
-\editmark{Because we have assumed 100\% of the spectrum is occupied by RFI, our given RFI constraints could be too pessimistic.
-If} only 10\% of the samples are affected by RFI, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all.
+Because we have assumed 100\% of the spectrum is occupied by RFI, our given RFI constraints could be too pessimistic.
+If only 10\% of the samples are affected by RFI, the expected value of the leaked RFI level decreases by about $25$\%, and if the detected 2.68\% true-positives contain all RFI, there is no leaked RFI at all.
 With current data, one can only speculate how much of the electromagnetic spectrum is truly occupied.
 
-\editmark{Finally, future RFI excision strategies can further enhance detection accuracy.
+Finally, future RFI excision strategies can further enhance detection accuracy.
 Once data from a large number of nights are collected, it will be possible to detect and excise RFI more accurately.
-With} the current strategy, it is likely that the LOFAR EoR project will encounter some RFI on some frequencies when averaging lots of observing nights, although this still remains to be seen.
+With the current strategy, it is likely that the LOFAR EoR project will encounter some RFI on some frequencies when averaging lots of observing nights, although this still remains to be seen.
 To mitigate this leaked RFI, the detection can be executed at higher signal-to-noise levels.

### 2013-07-16 articles/Article-RFI-distribution/article-rfi-distribution.tex
Submitted final version

@@
 
-%\usepackage[colorlinks=true,linkcolor=customlinkcolor,urlcolor=customlinkcolor,citecolor=customcitecolor,pdftex]{hyperref}
-\usepackage[pdftex]{hyperref}
+\usepackage[colorlinks=true,linkcolor=customlinkcolor,urlcolor=customlinkcolor,citecolor=customcitecolor,pdftex]{hyperref}
+%\usepackage[pdftex]{hyperref}
 
@@
 It is common knowledge that in a uniform Euclidian Universe source counts behave like power-law distributions.
-The differential distribution for sources on a flat surface is a power law with -2 exponent.
+The differential source-count distribution for sources on a flat surface is a power law with -2 exponent.
 We will derive this expected intrinsic source distribution for interfering radio sources.

### 2013-07-16 articles/Article-RFI-distribution/pack.sh

@@ -1,1 +1,1 @@
-tar -czvf rfi-distributions.tar.gz article-rfi-distribution.tex references.bib mn2e.bst mn2e.cls img/TF-L43786_L220-rfi-example-xaxistext.pdf img/transmitter-height/plot-transmitter-height-trimmed.pdf img/plot-rayleigh-and-rfi-trimmed.pdf img/plot-rayleigh-and-rfi-combined-trimmed.pdf img/explanation-lower-constraint/FindingCounts.pdf img/explanation-lower-constraint/FindingConstraints.pdf img/histograms-raw-trimmed.pdf img/plot-lba-dist-per-frequency-trimmed.pdf img/plot-hba-dist-per-frequency-trimmed.pdf img/histograms-corrected-trimmed.pdf img/histograms-with-fits-trimmed.pdf img/slopes-trimmed.pdf
+tar -czvf rfi-distributions.tar.gz article-rfi-distribution.tex references.bib mn2e.bst mn2e.cls img/TF-L43786_L220-rfi-example-xaxistext.pdf img/transmitter-height/plot-transmitter-height-trimmed.pdf img/plot-rayleigh-and-rfi-trimmed.pdf img/plot-rayleigh-and-rfi-combined-trimmed.pdf img/explanation-lower-constraint/FindingConstraints.pdf img/histograms-raw-trimmed.pdf img/plot-lba-dist-per-frequency-trimmed.pdf img/plot-hba-dist-per-frequency-trimmed.pdf img/histograms-corrected-trimmed.pdf img/histograms-with-fits-trimmed.pdf img/slopes-trimmed.pdf

### 2013-07-22 articles/Article-RFI-distribution/article-rfi-distribution.tex
Inserting correct dates
Submitted article to Arxiv

@@
 $^{32}$Department of Astrophysics/IMAPP, Radboud University Nijmegen, P.O. Box 9010, 6500 GL Nijmegen, The Netherlands \\
-$^{33}$Argelander-Institut f\"{u}r Astronomie, University of Bonn, Auf dem H\"{u}gel 71, 53121, Bonn, Germany
+$^{33}$Argelander-Institut f\"{u}r Astronomie, University of Bonn, Auf dem H\"{u}gel 71, 53121, Bonn, Germany\vspace*{-0.5cm}
 }
@@
 
-\date{Accepted [TODO date]. Received [TODO date]; in original form [TODO date]}
+\date{Accepted 2013 July 16. Received 2013 July 16; in original form 2013 March 1}
 \pagerange{\pageref{firstpage}--\pageref{lastpage}}

### 2013-07-22 articles/Article-RFI-distribution/authorlist/authorlist-arxiv.py

@@ -1,0 +1,78 @@
+#!/usr/bin/env python
+# Original version written by Ben Stappers on 19-09-2012
+
+
+import os, sys, string, math
+from optparse import OptionParser
+
+def strip_end(text, suffix):
+  if not text.endswith(suffix):
+      return text
+  return text[:-len(suffix)]
+
+class author:
+    def __init__(self, information):
+        infolist = string.split(information,"\'")
+        self.name = string.strip(infolist[1],"\'")
+        self.institutes = string.split(infolist[-1])[:-1]
+        self.printout = ""
+
+parser = OptionParser()
+
+parser.add_option("-a", "--authors", action="store", type="string", 
+		  dest="authorlist",
+		  help="Give the filename with the authors to be included on this paper. The format: \"'B~.W.~Stappers' jod yes\" where yes indicates that the author is to be included and jod indicates the institute, more than one institute can be used, yes must be the last thing on the line though")
+
+parser.add_option("-i", "--institutes", action="store", type="string", 
+		  dest="instlist",
+		  help="Give the filename with the institutes to be included on this paper, Format is address followed by institute nickname, e.g. \"jod\".")
+
+parser.add_option("-w", "--windows", action="store_true",dest="windows",
+		  help="Give a windows format output")
+
+
+
+(options, args) = parser.parse_args()
+outlist=[]
+authorfile = open(options.authorlist,'r')
+for line in authorfile.readlines():
+     if (string.split(line)[-1] == "yes"):
+         outlist.append(author(line))
+
+inpinstlist={}
+instfile = open(options.instlist,'r')
+for line in instfile.readlines():
+    inpinstlist[string.split(line.rstrip())[-1]] = strip_end(line.rstrip(), string.split(line.rstrip())[-1]) 
+#    inpinstlist.append(institute(line))
+
+outinstlist={}
+outinstlistchk=[]
+instcnt=1
+for author in outlist:
+    for inst in author.institutes:
+        if inst not in outinstlistchk:
+            instinp = [ inst, instcnt ]
+            outinstlist[inst] = instcnt
+            outinstlistchk.append(inst)
+            instcnt += 1
+
+if (options.windows):
+
+    print "----------------- Authors ---------------"
+
+    for author in outlist:
+        author.printout = str(author.name)
+        first = 1
+        for authorinst in author.institutes:
+            if first==1:
+                first = 0
+            #else:
+                #author.printout = author.printout + ', '
[…]

### 2013-07-22 articles/Article-RFI-distribution/pack-arxiv.sh

@@ -1,0 +1,1 @@
+tar -czvf rfi-distributions-arxiv.tar.gz article-rfi-distribution.tex article-rfi-distribution.bbl mn2e.bst img/TF-L43786_L220-rfi-example-xaxistext.pdf img/transmitter-height/plot-transmitter-height-trimmed.pdf img/plot-rayleigh-and-rfi-trimmed.pdf img/plot-rayleigh-and-rfi-combined-trimmed.pdf img/explanation-lower-constraint/FindingConstraints.pdf img/histograms-raw-trimmed.pdf img/plot-lba-dist-per-frequency-trimmed.pdf img/plot-hba-dist-per-frequency-trimmed.pdf img/histograms-corrected-trimmed.pdf img/histograms-with-fits-trimmed.pdf img/slopes-trimmed.pdf

### 2013-08-13 articles/Report-MWA-Beam/mwa-beam.tex
memo on the RTS beam

\documentclass[a4paper,11pt]{article} 
%\usepackage[style=nature,backend=biber]{biblatex}
\usepackage{graphicx}
\usepackage[font={it},labelfont={bf}]{caption}
\usepackage{times}
\usepackage{color}
%\usepackage{fixltx2e} % fixes placing of image positions
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{courier}

\definecolor{customhdrcolor}{rgb}{0.0,0.0,0.0}
%\definecolor{customcitecolor}{rgb}{0.0,0.25,0.25}
\definecolor{customcitecolor}{rgb}{0.0,0.5,0.75}
%\definecolor{customlinkcolor}{rgb}{0.0,0.0,1.0}
\definecolor{customlinkcolor}{rgb}{0.0,0.5,0.75}

\usepackage[colorlinks=true,linkcolor=customlinkcolor,urlcolor=customlinkcolor,citecolor=customcitecolor,pdftex]{hyperref}

\ifpdf\pdfinfo{/Title      (Memo: MWA beam-correction formulae)
               /Author     (A. R. Offringa et al.)
               /Keywords   (MWA, beam)
        }
\else\usepackage{graphics}\fi

\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\title{Memo: MWA beam-correction formulae}

\author{A.~R.~Offringa}
%\author[1,2]{A.~R.~Offringa}
%\affil[1]{RSAA, Australian National University, Mt Stromlo Observatory, via Cotter Road, Weston, ACT 2611, Australia}
%\affil[2]{ARC Centre of Excellence for All-Sky Astrophysics (CAASTRO)}
\begin{document}

\label{firstpage}
\maketitle

\section{Analytical calculation of voltage gain}
The analytical gain of a tile is calculated from the analytical gain of a dipole with ground screen, multiplied by a factor $\eta$ that represents how coherent the dipoles add together. The (complex) tile factor $\eta$ is calculated with:\footnote{Taken from the RTS (2013-08-13), file \texttt{InstrumentalCalibration.c} written by D. Mitchell}
\begin{equation}
 \eta = \frac{1}{16}\sum\limits_{a=1}^{16} e^{\frac{-2 \pi i}{\lambda} \left(\Delta x_a \sin \mathcal{Z} \sin \chi + \Delta y_a \sin \mathcal{Z} \cos \chi + \Delta z_a \cos \mathcal{Z} - \Delta \tau_a \right) },
\end{equation}
where $\lambda$ is the wavelength, $\Delta x_a, \Delta y_a$ and $\Delta z_a$ are the relative positions of dipole $a$ in the tile, $\mathcal{Z}$ is zenith angle, $\chi$ is the azimuth and $\Delta \tau_a$ is the delay applied on dipole $a$ by the beam former. The $2 \times 2$ matrix $E$ describing the electrical gains of an individual dipole\footnote{Also described here: http://mwa-lfd.haystack.mit.edu/twiki/bin/view/Main/PrimaryBeamModels by D. Mitchell} is calculated with:
\begin{equation}
\mathbf{E} = \frac{2 \sin \left( \frac{2 \pi h}{\lambda} \cos \mathcal{Z}\right)}{2 \sin \left( \frac{2 \pi h}{\lambda}\right)}
\left( \begin{array}{ccc}
\cos l \cos \delta + \sin l \sin \delta \cos \textrm{ha} & ; & -\sin l \sin \textrm{ha} \\
\sin \delta \sin \textrm{ha} & ; & \cos \textrm{ha}
\end{array} \right).
\end{equation}
Here, $l$ is latitude, $\delta$ is declanation, $\textrm{ha}$ is hour angle and $h$ is the height of tile. The factors of two cancel out, but are left there because the denominator now forms the normalization towards zenith. Finally, the Jones gain matrix $B$ of the full tile is calculated by multiplying the dipole matrix with the tile factor,
\begin{equation}
\mathbf{B} = \eta \mathbf{E}.
\end{equation}
\section{Applying beam gains in image space}
Given the $2\times 2$ complex Jones matrix $\mathbf{B}$ that represents the voltage gain of the beam in a certain direction, the RTS corrects a polarized pixel, given by the vector $\textbf{v} = \left(v_\textrm{xx}, v_\textrm{xy}, v_\textrm{yx}, v_\textrm{yy} \right)^T$, for that direction in the image plane to a beam-corrected vector containing the Stokes values with the following calculation:\footnote{Taken from the RTS (2013-08-13), file \texttt{pb\_correct.c} written by S. Ord}
\begin{equation} \label{apply-beam-image-space}
 \textbf{s} =
\mathbf{T} \left[ \left(\mathbf{B} \otimes \mathbf{B}^* \right) \left(\mathbf{B} \otimes \mathbf{B}^* \right)^*\right]^{-1} \mathbf{Y} \textbf{v},
\end{equation}
with
\begin{equation}
\textbf{s} = \left( \begin{array}{c}
s_\textrm{\tiny{I}} \\
s_\textrm{\tiny{Q}} \\
s_\textrm{\tiny{U}} \\
s_\textrm{\tiny{V}} \end{array} \right),
T = \left( \begin{array}{cccc}
1 & 0 & 0 & 1 \\
1 & 0 & 0 & -1 \\
0 & 1 & 1 & 0 \\
0 & -i & -i & 0 \end{array} \right), \mathbf{Y} = \left( \begin{array}{cccc}
1 & 0 & 0 & 0 \\
0 & 1 & i & 0 \\
0 & 1 & -i & 0 \\
0 & 0 & 0 & 1
\end{array} \right),
\end{equation}
where $\otimes$ the Kronecker product and $\mathbf{A}^*$ the conjugate transpose of matrix $\mathbf{A}$. The output $\textbf{s}$ is a vector with the Stokes parameter. The matrix $\mathbf{T}$ converts from instrumental basis (xx, xy, yx, yy) to the Stokes parameter and hence is applied last.

\textbf{Question:} I don't know why matrix $\mathbf{Y}$ (pronounced as ``why?!?'') is applied on vector $\textbf{v}$ first. $\mathbf{Y}$ has the following effect:
\begin{equation}
\mathbf{Y}\textbf{v} = \left( \begin{array}{c}
 v_\textrm{xx} \\ v_\textrm{xy} + i v_\textrm{yx} \\
 v_\textrm{xy} - i v_\textrm{yx} \\ v_\textrm{yy} \end{array}\right)
\end{equation}
Possible explanation could be that this is necessary because the beam is applied in image space, instead of on the visibilities. That would imply that $\mathbf{Y}$ can be left out if the beam is corrected in visibility space (e.g. on the model).

\section{Applying the beam more efficiently}
As a side note, Eq.~\eqref{apply-beam-image-space} involves a $4\times 4$ complex matrix inverse, which can be computationally done more efficiently by decomposing it in two $2\times 2$ complex matrix inversions:
\begin{equation} \notag
 \textbf{s} =
\mathbf{T} \left[ \left(\mathbf{B} \otimes \mathbf{B}^* \right) \left(\mathbf{B} \otimes \mathbf{B}^* \right)^*\right]^{-1} \mathbf{Y} \textbf{v}
\end{equation}
\{ Conjugate transpose is distributive over the Kronecker product  \}
\begin{equation} \notag
 \textbf{s} =
\mathbf{T} \left[ \left(\mathbf{B} \otimes \mathbf{B}^* \right) \left(\mathbf{B}^* \otimes \mathbf{B} \right)\right]^{-1} \mathbf{Y} \textbf{v}
\end{equation}
\{ Using associativity of K.p., i.e., $\left(\mathbf{A} \otimes \mathbf{B} \right) \left(\mathbf{C} \otimes \mathbf{D} \right) = (\mathbf{AC}) \otimes (\mathbf{BD})$ \}
\begin{equation} \notag
 \textbf{s} = \mathbf{T} \left[ (\mathbf{BB}^*) \otimes (\mathbf{B}^*\mathbf{B}) \right]^{-1} \mathbf{Y} \textbf{v}
\end{equation}
\{ Inverse is distributive over the Kronecker product \}
\begin{equation}
 \textbf{s} = \mathbf{T} \left[ (\mathbf{BB}^*)^{-1} \otimes (\mathbf{B}^*\mathbf{B})^{-1} \right] \mathbf{Y} \textbf{v}
\end{equation}
Note that $\mathbf{BB}^*$ and $\mathbf{B}^*\mathbf{B}$ are not equal (unless $\mathbf{B}$ is unitary, which it is not in general), but they are $2\times 2$ matrices, and their inverse thus has an analytical solution. This is much faster and easier than using a general library to do a $4 \times 4$ inversion.
\label{lastpage}

\end{document}

### 2013-08-13 articles/Report-MWA-Beam/mwa-beam.tex
Adding Mitch's explanation

@@
 where $\lambda$ is the wavelength, $\Delta x_a, \Delta y_a$ and $\Delta z_a$ are the relative positions of dipole $a$ in the tile, $\mathcal{Z}$ is zenith angle, $\chi$ is the azimuth and $\Delta \tau_a$ is the delay applied on dipole $a$ by the beam former.
-The $2 \times 2$ matrix $E$ describing the electrical gains of an individual dipole\footnote{Also described here: http://mwa-lfd.haystack.mit.edu/twiki/bin/view/Main/PrimaryBeamModels by D.
+The $2 \times 2$ matrix $\textbf{E}$ describing the electrical gains of an individual dipole\footnote{Also described here: http://mwa-lfd.haystack.mit.edu/twiki/bin/view/Main/PrimaryBeamModels by D.
 Mitchell} is calculated with:
@@
 \section{Applying beam gains in image space}
-Given the $2\times 2$ complex Jones matrix $\mathbf{B}$ that represents the voltage gain of the beam in a certain direction, the RTS corrects a polarized pixel, given by the vector $\textbf{v} = \left(v_\textrm{xx}, v_\textrm{xy}, v_\textrm{yx}, v_\textrm{yy} \right)^T$, for that direction in the image plane to a beam-corrected vector containing the Stokes values with the following calculation:\footnote{Taken from the RTS (2013-08-13), file \texttt{pb\_correct.c} written by S.
+Given the $2\times 2$ complex Jones matrix $\mathbf{B}$ that represents the voltage gain of the beam in a certain direction, the RTS corrects a polarized pixel, given by the vector $\textbf{v} = \left(v_\textrm{xx}, \textrm{real}(v_\textrm{xy}), \textrm{imag}(v_\textrm{xy}), v_\textrm{yy} \right)^T$, for that direction in the image plane to a beam-corrected vector containing the Stokes values with the following calculation:\footnote{Taken from the RTS (2013-08-13), file \texttt{pb\_correct.c} written by S.
 Ord}
@@
 
-\textbf{Question:} I don't know why matrix $\mathbf{Y}$ (pronounced as ``why?!?'') is applied on vector $\textbf{v}$ first. $\mathbf{Y}$ has the following effect:
-\begin{equation}
-\mathbf{Y}\textbf{v} = \left( \begin{array}{c}
- v_\textrm{xx} \\ v_\textrm{xy} + i v_\textrm{yx} \\
- v_\textrm{xy} - i v_\textrm{yx} \\ v_\textrm{yy} \end{array}\right)
-\end{equation}
-Possible explanation could be that this is necessary because the beam is applied in image space, instead of on the visibilities.
-That would imply that $\mathbf{Y}$ can be left out if the beam is corrected in visibility space (e.g.
-on the model).
+Matrix $\mathbf{Y}$ is applied because the RTS does not store both the xy and yx images, but instead stores one complex xy image.
+Its real values are stored in the 2nd index, and the imaginary values are stored in the 3rd index of vector $\textbf{v}$.
+Multiplying $\mathbf{v}$ with $\mathbf{Y}$ therefore has the following effect: $\mathbf{Y}\textbf{v} = \left(v_\textrm{xx},v_\textrm{xy}, \overline{v_\textrm{xy}},v_\textrm{yy}\right)^T$.
+$\mathbf{Y}$ can be left out if the input is alread xx, xy, yx, yy.
 

### 2013-08-15 articles/Report-MWA-Beam/mwa-beam.tex
Hints from Mitch

@@
 The factors of two cancel out, but are left there because the denominator now forms the normalization towards zenith.
-Finally, the Jones gain matrix $B$ of the full tile is calculated by multiplying the dipole matrix with the tile factor,
+Finally, the Jones gain matrix $\textbf{B}$ of the full tile is calculated by multiplying the dipole matrix with the tile factor,
 \begin{equation}
@@
 \section{Applying beam gains in image space}
-Given the $2\times 2$ complex Jones matrix $\mathbf{B}$ that represents the voltage gain of the beam in a certain direction, the RTS corrects a polarized pixel, given by the vector $\textbf{v} = \left(v_\textrm{xx}, \textrm{real}(v_\textrm{xy}), \textrm{imag}(v_\textrm{xy}), v_\textrm{yy} \right)^T$, for that direction in the image plane to a beam-corrected vector containing the Stokes values with the following calculation:\footnote{Taken from the RTS (2013-08-13), file \texttt{pb\_correct.c} written by S.
+Given the $2\times 2$ complex Jones matrix $\mathbf{B}$ that represents the voltage gain of the beam in a certain direction, the RTS corrects a polarized pixel, given by the vector $\textbf{v}' = \left(v_\textrm{xx}, \textrm{real}(v_\textrm{xy}), \textrm{imag}(v_\textrm{xy}), v_\textrm{yy} \right)^T$, for that direction in the image plane to a beam-corrected vector containing the Stokes values with the following calculation:\footnote{Taken from the RTS (2013-08-13), file \texttt{pb\_correct.c} written by S.
 Ord}
@@
  \textbf{s} =
-\mathbf{T} \left[ \left(\mathbf{B} \otimes \mathbf{B}^* \right) \left(\mathbf{B} \otimes \mathbf{B}^* \right)^*\right]^{-1} \mathbf{Y} \textbf{v},
+\mathbf{T} \left[ \left(\mathbf{B} \otimes \mathbf{B}^{*T} \right)^* \left(\mathbf{B} \otimes \mathbf{B}^{*T} \right)\right]^{-1} \mathbf{Y} \mathbf{v}',
+\end{equation}
+\textbf{Note:} the formula above is how the RTS seems to do it, but it may not be correct. Mitch is going to check this, as it should be:
+\begin{equation}
+ \textbf{s} =
+\mathbf{T} \left(\mathbf{B} \otimes \mathbf{B}^{*T} \right)^{-1} \mathbf{Y} \mathbf{v}',
 \end{equation}
@@
 \end{equation}
-where $\otimes$ the Kronecker product and $\mathbf{A}^*$ the conjugate transpose of matrix $\mathbf{A}$.
-The output $\textbf{s}$ is a vector with the Stokes parameter.
+where $\otimes$ the Kronecker product, $\mathbf{A}^*$ the conjugate transpose of matrix $\mathbf{A}$ and $\mathbf{A}^{*T}$ the matrix $\mathbf{A}$ where each element has been conjugated ($A_{ij}$ = $\overline{A^{*T}_{ij}}$).
+The output $\textbf{s}$ is a vector with the Stokes parameters.
 The matrix $\mathbf{T}$ converts from instrumental basis (xx, xy, yx, yy) to the Stokes parameter and hence is applied last.
@@
 Matrix $\mathbf{Y}$ is applied because the RTS does not store both the xy and yx images, but instead stores one complex xy image.
-Its real values are stored in the 2nd index, and the imaginary values are stored in the 3rd index of vector $\textbf{v}$.
-Multiplying $\mathbf{v}$ with $\mathbf{Y}$ therefore has the following effect: $\mathbf{Y}\textbf{v} = \left(v_\textrm{xx},v_\textrm{xy}, \overline{v_\textrm{xy}},v_\textrm{yy}\right)^T$.
-$\mathbf{Y}$ can be left out if the input is alread xx, xy, yx, yy.
+Its real values are stored in the 2nd index, and the imaginary values are stored in the 3rd index of vector $\mathbf{v}'$.
+Multiplying $\mathbf{v}'$ with $\mathbf{Y}$ therefore has the following effect: $\mathbf{v}=\mathbf{Y}\mathbf{v}' = \left(v_\textrm{xx},v_\textrm{xy}, \overline{v_\textrm{xy}},v_\textrm{yy}\right)^T$.
+$\mathbf{Y}$ can be left out if the input is already in the form xx, xy, yx, yy.
+
+\section{Calculating weights}
+\textbf{Note:} the following still needs to be checked, as currently it seems the RTS is doing this differently.
+
+Eq.~\ref{apply-beam-image-space} describes how to apply the beam to a single measurement.
+The RTS integrates multiple snapshots with different beam values, which requires a weighted average of the measurements.
+A measurement is beam-corrected and weighted by the square of the beam value.
+These weighted measurements are summed together and finally the total weight is divided out.
+Therefore, the final pixel value $\mathbf{s'}$ is estimated from the individual measurements $\mathbf{v}_i$ and their corresponding beam values $\mathbf{B}_i$ as:
+\begin{equation}
+\mathbf{s'} = \left( \sum\limits_{i=1}^N \textbf{W}_i \mathbf{s}_i \right) \left( \sum\limits_{i=1}^N \textbf{W}_i \right)^{-1},
+\end{equation}
+with
+\begin{equation}
+\textbf{W}_i = \mathbf{T} \left[ \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^* \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)\right].
+\end{equation}
+In this, $\mathbf{s}_i$ can be substituted with Eq.~\ref{apply-beam-image-space} to prevent having to invert the beam for every measurement:
+\begin{equation}
+\mathbf{s'} = 
+\left(
+  \sum\limits_{i=1}^N \mathbf{T} \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^* \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right) 
+  \mathbf{T} \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^{-1} \mathbf{v}
+\right)
+\left( \sum\limits_{i=1}^N \textbf{W}_i \right)^{-1},
+\end{equation}
+\{ We can weight first and calculates Stokes parameters later \}
+\begin{eqnarray}
+\notag
+\mathbf{s'} & =  &
+\left(
+  \mathbf{T} \sum\limits_{i=1}^N \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^* \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right) 
+  \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^{-1} \mathbf{v}
+\right)
+\left( \sum\limits_{i=1}^N \textbf{W}_i \right)^{-1} \\
+& = &
+\left(
+  \mathbf{T} \sum\limits_{i=1}^N \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^* \mathbf{v}
+\right)
+\left( \sum\limits_{i=1}^N \textbf{W}_i \right)^{-1},
+\end{eqnarray}
 
@@
  \textbf{s} =
-\mathbf{T} \left[ \left(\mathbf{B} \otimes \mathbf{B}^* \right) \left(\mathbf{B} \otimes \mathbf{B}^* \right)^*\right]^{-1} \mathbf{Y} \textbf{v}
+\mathbf{T} \left[ \left(\mathbf{B} \otimes \mathbf{B}^{*T} \right) \left(\mathbf{B} \otimes \mathbf{B}^{*T} \right)^*\right]^{-1} \mathbf{Y} \textbf{v}
 \end{equation}
@@
  \textbf{s} =
-\mathbf{T} \left[ \left(\mathbf{B} \otimes \mathbf{B}^* \right) \left(\mathbf{B}^* \otimes \mathbf{B} \right)\right]^{-1} \mathbf{Y} \textbf{v}
+\mathbf{T} \left[ \left(\mathbf{B} \otimes \mathbf{B}^{*T} \right) \left(\mathbf{B}^{*} \otimes \mathbf{B}^T \right)\right]^{-1} \mathbf{Y} \textbf{v}
 \end{equation}
@@
 \begin{equation} \notag
- \textbf{s} = \mathbf{T} \left[ (\mathbf{BB}^*) \otimes (\mathbf{B}^*\mathbf{B}) \right]^{-1} \mathbf{Y} \textbf{v}
+ \textbf{s} = \mathbf{T} \left[ (\mathbf{BB}^*) \otimes (\mathbf{B}^{*T}\mathbf{B}^T) \right]^{-1} \mathbf{Y} \textbf{v}
 \end{equation}
@@
 \begin{equation}
- \textbf{s} = \mathbf{T} \left[ (\mathbf{BB}^*)^{-1} \otimes (\mathbf{B}^*\mathbf{B})^{-1} \right] \mathbf{Y} \textbf{v}
+ \textbf{s} = \mathbf{T} \left[ (\mathbf{BB}^*)^{-1} \otimes (\mathbf{B}^{*T}\mathbf{B}^T)^{-1} \right] \mathbf{Y} \textbf{v}
 \end{equation}
-Note that $\mathbf{BB}^*$ and $\mathbf{B}^*\mathbf{B}$ are not equal (unless $\mathbf{B}$ is unitary, which it is not in general), but they are $2\times 2$ matrices, and their inverse thus has an analytical solution.
+\{ Using distributivity of inversion over transpose, and $\mathbf{A}^T \mathbf{B}^T=(\mathbf{BA})^T$ \}
+\begin{equation}
+ \textbf{s} = \mathbf{T} \left[ (\mathbf{BB}^*)^{-1} \otimes \left((\mathbf{B}\mathbf{B}^*)^{-1}\right)^T \right] \mathbf{Y} \textbf{v}
+\end{equation}
+$\mathbf{BB}^*$ is a $2\times 2$ matrix, and its inverse thus has an analytical solution.
 This is much faster and easier than using a general library to do a $4 \times 4$ inversion.
+(It is not completely clear Eq.~\ref{apply-beam-image-space} is correct, so this might change -- it is expected that the data should not have the square of the Mueller matrix $\mathbf{B}\otimes\mathbf{B}^{*T}$ matrix applied.
+Even then it is possible to use only 2x2 matrix inversions, because the inverse is distribute over the Kronecker product).
 \label{lastpage}

### 2013-08-18 articles/Report-MWA-Beam/mwa-beam.tex
Some changes on beam in cal
Fixing sign error in beam\!

@@
 where $\lambda$ is the wavelength, $\Delta x_a, \Delta y_a$ and $\Delta z_a$ are the relative positions of dipole $a$ in the tile, $\mathcal{Z}$ is zenith angle, $\chi$ is the azimuth and $\Delta \tau_a$ is the delay applied on dipole $a$ by the beam former.
+Antennae are 1.1 \textit{m} separated from each other within tiles and currently are all assumed to have a differential height $\Delta z_i$ of 0.
+The MWA tools pb scripts assume the antennae are 0.278~\textit{m} tall while the RTS assumes 0.3~\textit{m}.
 The $2 \times 2$ matrix $\textbf{E}$ describing the electrical gains of an individual dipole\footnote{Also described here: http://mwa-lfd.haystack.mit.edu/twiki/bin/view/Main/PrimaryBeamModels by D.

### 2013-08-20 articles/Report-MWA-Beam/mwa-beam.tex
Adding Jones notation

@@
 These weighted measurements are summed together and finally the total weight is divided out.
-Therefore, the final pixel value $\mathbf{s'}$ is estimated from the individual measurements $\mathbf{v}_i$ and their corresponding beam values $\mathbf{B}_i$ as:
-\begin{equation}
+Therefore, the final pixel value $\mathbf{s'}$ is estimated from the individual measurements $\mathbf{s}_i$ and their corresponding beam values $\mathbf{B}_i$ as:
+\begin{equation} \label{eq:weighting}
 \mathbf{s'} = \left( \sum\limits_{i=1}^N \textbf{W}_i \mathbf{s}_i \right) \left( \sum\limits_{i=1}^N \textbf{W}_i \right)^{-1},
@@
 with
-\begin{equation}
+\begin{equation} \label{eq:weight-per-value}
 \textbf{W}_i = \mathbf{T} \left[ \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^* \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)\right].
@@
   \sum\limits_{i=1}^N \mathbf{T} \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^* \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right) 
-  \mathbf{T} \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^{-1} \mathbf{v}
+  \mathbf{T} \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^{-1} \mathbf{v}_i
 \right)
@@
   \mathbf{T} \sum\limits_{i=1}^N \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^* \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right) 
-  \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^{-1} \mathbf{v}
+  \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^{-1} \mathbf{v}_i
 \right)
 \left( \sum\limits_{i=1}^N \textbf{W}_i \right)^{-1} \\
+\label{eq:weighting-final}
 & = &
 \left(
-  \mathbf{T} \sum\limits_{i=1}^N \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^* \mathbf{v}
+  \mathbf{T} \sum\limits_{i=1}^N \left(\mathbf{B}_i \otimes \mathbf{B}_i^{*T} \right)^* \mathbf{v}_i
 \right)
 \left( \sum\limits_{i=1}^N \textbf{W}_i \right)^{-1},
+\end{eqnarray}
+\section{Using Jones matrices}
+Because our optical system is 'pure', the Mueller $4\times 4$-matrix calcules can be avoided by using $2\times 2$ Jones matrices. The relationship between these two is:
+\begin{equation}
+\left(\mathbf{A} \otimes \mathbf{B}^{*T}\right) \mathbf{v} = \textbf{AVB}^*.
+\end{equation}
+The left hand side multiplies a $4\times 4$ matrix with a vector containing the visibilities ordered as $\mathbf{v}=[v_{xx}, v_{xy}, v_{yx}, v_{yy}]^T$, while the right hand side uses the $2\times 2$ matrix
+\begin{equation}
+\mathbf{V}=\left(\begin{array}{cc}
+v_{xx} & v_{xy}\\ v_{yx} & v_{yy}
+\end{array}\right).
+\end{equation}
+Using this relation, Eq.~\eqref{eq:weighting-final}, excluding conversion to Stokes, can be rewritten:
+\begin{eqnarray}
+\mathbf{R}=
+\left( \sum\limits_{i=1}^N \mathbf{B}^*\mathbf{B}\right)^{-1}
+\left( \sum\limits_{i=1}^N \mathbf{B}^*\mathbf{V}\mathbf{B} \right)
+\left( \sum\limits_{i=1}^N\mathbf{B}^*\mathbf{B}\right)^{-1}
 \end{eqnarray}

### 2013-09-20 articles/Report-D0002/report-d0002-RFI-analysis.tex
Report on D0002

\documentclass[a4paper,10pt]{article} 
%\usepackage[style=nature,backend=biber]{biblatex}
\usepackage{graphicx}
\usepackage[font={it},labelfont={bf}]{caption}
\usepackage{times}
\usepackage{color}
%\usepackage{fixltx2e} % fixes placing of image positions
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{courier}
\usepackage[round]{natbib}
\usepackage{wrapfig}

\definecolor{customhdrcolor}{rgb}{0.0,0.0,0.0}
%\definecolor{customcitecolor}{rgb}{0.0,0.25,0.25}
\definecolor{customcitecolor}{rgb}{0.0,0.5,0.75}
%\definecolor{customlinkcolor}{rgb}{0.0,0.0,1.0}
\definecolor{customlinkcolor}{rgb}{0.0,0.5,0.75}

\usepackage[colorlinks=true,linkcolor=customlinkcolor,urlcolor=customlinkcolor,citecolor=customcitecolor,pdftex]{hyperref}

\ifpdf\pdfinfo{/Title      (Report on D0002 observations: high spectral resolution RFI analysis)
               /Author     (A. R. Offringa)
               /Keywords   (MWA, RFI, spectral resolution)
        }
\else\usepackage{graphics}\fi

\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\title{Report on D0002 observations: high spectral resolution RFI analysis}

\author{A.~R.~Offringa}
%\author[1,2]{A.~R.~Offringa}
%\affil[1]{RSAA, Australian National University, Mt Stromlo Observatory, via Cotter Road, Weston, ACT 2611, Australia}
%\affil[2]{ARC Centre of Excellence for All-Sky Astrophysics (CAASTRO)}
\begin{document}

\label{firstpage}
\maketitle

\section{Goals}
MWA director's time was requested to compare RFI detection accuracy between a spectral resolution of 10~kHz and 40~kHz. High spectral resolution has been a major benefit for dealing with the RFI environment of LOFAR \citep{lofar-radio-environment}. Whether the data quality improves with higher resolution, caused by more accurate RFI detection, is relevant for optimizing future MWA observing configurations. Moreover, for future telescopes such as the SKA it is very interesting to see how frequency resolution relates to data quality, especially in the quiet radio environment of the MWA.

\section{Observations}
I received one hour of MWA data observed with a spectral resolution of 10~kHz and temporal resolution of 2~s. The observations were recorded in the night of 30-31 July 2013. The EoR bands covering 139--200~MHz with centre coarse channels 121 and 145 were selected. The first 4~min were spent on calibrator 3C444, after which the quiet EoR0 field was observed for 1h. The bands were observed intermittently in blocks of 5~min.

\section{Results}
The observations were processed with the Cotter MWA preprocessing pipeline, which flags the data using the AOFlagger RFI detector. This detector uses combinatorial thresholding \citep{post-correlation-rfi-classification} and morphological detection \citep{scale-invariant-rank-operator}.
\begin{figure*}
\begin{center}
\includegraphics[width=12.5cm]{Apparent-image-EoR0.png}
\caption{Apparent image with 16 mJy noise level, deconvolved with {\tt wsclean}. These are the data after initial transfer of calibration solutions: no self-calibration or peeling has been applied yet. I believe it is the deepest MWA image made so far. Artefacts around the brightest sources are considerably decreased by self-calibration, but I have not made a deconvolved image of the self-calibrated data yet.}
\label{fig:eor0}
\end{center}
\end{figure*}
\subsection{Imaging quality}
The observations are of good quality. The calibrator scans are calibrated with a 3C444 point source model without any problems. A full polarization calibration script developed for the spectral analysis of sources in the EoR fields was used to do this. Transferring the 3C444 solutions to the EoR0 observations results in a reasonably good initial calibration. A resulting image of this is given in Fig.~\ref{fig:eor0}. Self-calibration and peeling increases the image quality further, but I have not made a deconvolved image with all data after self-calibration and peeling. Imaging only half the data results in an increase of noise very close to $\sqrt{2}$, suggesting that the observation is not yet confusion limited. The noise level is not as deep as estimated by $t \approx 8 \times 10^4/B\sigma^2_s$, formula (4) in \citet{mwa}. That formula results in $\sigma = 0.89$ mJy for an hour of data (with $B=30.72$~MHz~$\times 22/24$ due to the missing GPU boxes), which is a factor of $\sim 18$ lower than the actually observed noise level. This deficient might be partially caused by calibration artefacts, but further self-calibration and peeling did not decrease the noise more than a few mJy. This is a quiet field, so one should expect that any field will need more time (and calibration effort) to reach the desired noise limit. It seems unlikely that RFI is the cause of this deficiency.

\subsection{RFI analysis}
\begin{wraptable}{r}{6cm}
\caption{List of most dominant RFI sources in the range 139--200 MHz. Occupancy values describe the ratio of detected RFI in a 10 kHz channel over the full hour.}\label{table:rfi-sources}\begin{center}\begin{tabular}{rrr}
\hline
\hline
\textbf{Frequency}& \textbf{Occupancy} & \textbf{Coarse ch} \\
\textbf{(MHz)}& \textbf{(\%)} & \textbf{number} \\
\hline
141.685 &  0.97 & 111 \\
145.955 &  8.01 & 114 \\
149.955 & 31.52 & 117 \\
150.025 & 29.96 & 117 \\
178.745 &  4.13 & 140 \\
186.155 &  1.88 & 145 \\
194.455 &  2.02 & 152 \\
200.015 &  6.15 & 156 \\
\hline
\end{tabular}
\end{center}
\end{wraptable}

\begin{figure*}
\begin{center}
\includegraphics[width=12cm]{rfi-spectrum.pdf}
\caption{Detected RFI ratio over frequency for all visibilities (1~h, all correlations). The visible transmitters are enumerated in Table~\ref{table:rfi-sources}. The gaps around 157 and 181~MHz are because of two broken GPU correlator boxes. The 1.28~MHz periodic structure is because of the poly-phase filter: each 1.28 MHz sub-band represents a coarse channel. }
\label{fig:rfi-spectrum}
\end{center}
\end{figure*}

The amount of RFI in these observations is relatively low -- lower than previously observed in a South Galactic Pole RFI measurement with 32T. This might be because of a different sensitivity towards the horizon due to different beam former settings, but it could also be accidental. The RFI environment in more recent EoR1 observations looks similar to these EoR0 observations. The strongest RFI contaminator with the MWA band comes from the Orbcom satellites, but these transmit outside the bandwidth of these observations. Fig.~\ref{fig:rfi-spectrum} shows the amount of detected RFI over frequency, for the full hour. Two transmitters around 146 and 150 MHz are clearly visible, but a few more can be identified. The signal at 150 MHz seems to contaminate two channels that are 70 kHz apart, so this signal could come from two transmitters. A list of channels containing a significant amount of RFI are listed in Table~\ref{table:rfi-sources}. There are a few more peaks in the data, but it is hard to distinguish the DC channel offset problem and coarse channel edges from actual low-level RFI contamination.

\begin{figure*}
\begin{center}
\includegraphics[width=12cm]{stddev-spectrum.pdf}
\caption{Standard deviation over frequency for all visibilities (1~h, all correlations). No significant RFI residuals are visible. The jump at 187~MHz is because of digital gains that are applied to the data before correlation, resulting in different noise statistics.}
\label{fig:stddev-spectrum}
\end{center}
\end{figure*}

The flagger performs an accurate rejection and flags almost all RFI. At 10 kHz, 0.83\% of all data is lost due to RFI. Fig.~\ref{fig:stddev-spectrum} shows the standard deviation of all the data after the RFI rejection. There is a small noise increase around 147 MHz. To see how much frequency resolution affects the quality, I have averaged the data to 40~kHz and reapplied the AOFlagger. The resulting standard deviation plot looks extremely similar. The only difference is that at 40 kHz the amount of detected RFI is about 1\% higher. This is hard to estimate accurately using these data, because of the DC channel and sideband edges that get into the statistics. Cotter makes sure to keep these out of the statistics, but a standalone run of the AOFlagger is required to flag the 40 kHz data, and the AOFlagger does not distinguish between DC channels and RFI. On EoR1 data, typical RFI occupancies reported by Cotter are $\sim 1.65$\%, agreeing with the fact that at 40 kHz about 1\% more data is lost due to RFI.

\subsection{Other uses of these observations}
\begin{itemize}
 \item Undergraduate student Jake Hughes has used these data to analyze whether RFI events can be located to a specific point on the sky (or horizon). Certain constant transmitters were clearly visible as peaks in the sky when the data was not flagged. He also tried to image short bursts of RFI, that were apparent in a few time steps. These only increased the noise in the image, but this affected the whole sky image, and not a specific part of the sky. It is likely that these RFI burst are therefore occurring in the near field.
 \item Because of its high quality, I have used these data extensively to develop my spectral analysis software.
 \item I have provided the observations with self-calibration and subtraction of the brightest sources to Cathryn Trott, who will use these to test her EoR pipeline.
 \item The observations can be included in the other EoR analyses.
\end{itemize}
The latter three items do of course not necessarily require the high spectral resolution. However, clearly these observations have been used extensively and were very helpful.

\section{Conclusions}
\begin{itemize}
 \item The observations were of high quality.
 \item The image noise is more than an order of magnitude worse than estimated by \citet{mwa}. Some improvement might be made by spending more effort on calibration, but I think the formula given by Tingay et al. is too optimistic.
 \item RFI is not much of a concern for MWA in the 139--200~MHz frequency range. About 1.65\% is lost due to RFI at 40~kHz, while 0.83\% is lost at 10~kHz. In general no leaked RFI is visible after applying the AOFlagger. That does not necessarily mean there is none, but it seems to be insignificant from current data.
 \item Preliminary, the only benefit that 10~kHz observations seems to have on RFI rejection, is that about 1\% less data is lost due to RFI. This should be tested at different beam former settings, different epochs and different frequencies to make sure this is the case in general.
 \item We now have a baseline of RFI measurements, which will be useful to compare to in the future. Regularly repeating such a test will be useful.
\end{itemize}

\label{lastpage}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}

### 2013-10-04 articles/Proposal-wsclean/paper-proposal-fast-widefield-imager.tex
Proposal wsclean paper

\documentclass[a4paper,10pt]{article} 
%\usepackage[style=nature,backend=biber]{biblatex}
\usepackage{graphicx}
\usepackage[font={it},labelfont={bf}]{caption}
\usepackage{times}
\usepackage{color}
%\usepackage{fixltx2e} % fixes placing of image positions
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{courier}
\usepackage[round]{natbib}
\usepackage{wrapfig}

\definecolor{customhdrcolor}{rgb}{0.0,0.0,0.0}
%\definecolor{customcitecolor}{rgb}{0.0,0.25,0.25}
\definecolor{customcitecolor}{rgb}{0.0,0.5,0.75}
%\definecolor{customlinkcolor}{rgb}{0.0,0.0,1.0}
\definecolor{customlinkcolor}{rgb}{0.0,0.5,0.75}

\usepackage[colorlinks=true,linkcolor=customlinkcolor,urlcolor=customlinkcolor,citecolor=customcitecolor,pdftex]{hyperref}

\ifpdf\pdfinfo{/Title      (MWA paper proposal Offringa)
               /Author     (A. R. Offringa)
               /Keywords   (MWA, imaging)
        }
\else\usepackage{graphics}\fi

\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

%\title{MWA Paper proposal:\\````WSClean: a fast implementation of a generic widefield imager''}

%\author{A.~R.~Offringa}
%\author[1,2]{A.~R.~Offringa}
%\affil[1]{RSAA, Australian National University, Mt Stromlo Observatory, via Cotter Road, Weston, ACT 2611, Australia}
%\affil[2]{ARC Centre of Excellence for All-Sky Astrophysics (CAASTRO)}
\begin{document}

\label{firstpage}
\section*{MWA Paper proposal: ``WSClean: a new implementation of a fast, generic wide-field imager''}
\begin{tabular}{ll}
Publication Title: & WSClean: a new implementation of a fast, generic \\
                   & wide-field imager\\
                   & \\
Principle Contact: & Dr. Andr\'e R. Offringa \\
                   & \\
Draft publication\\
\hspace{5mm}author list: & A.~R.~Offringa, B.~McKinley, N.~Hurley-Walker, \\
                         & F.~H.~Briggs, MWA commissioning team and GLEAM \\
                         & members where appropriate, MWA builders list \\
                   & \\
Intended Journal: & Probably MNRAS\\
                   & \\
Intended date of submission\\
\hspace{5mm}to collaboration: & February 2014\\
\end{tabular}

\section*{Paper summary}
We present our new imaging implementation, ``WSClean'', which is an order of magnitude faster than existing implementations and can handle MWA full-sky imaging at suitable resolution.

Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with a simple FFT based imager with an image based deconvolution procedure such as described by \citet{hogbom-clean}. Instead, the imaging algorithm needs to accomodate the ``w-term'' during inversion, which is the term that describes the deviation of the array to a perfect plane. This deviation is amplified by the field of view of the telescope, making this especially a problem for low-frequency widefield telescopes.

The most commonly used method to image observations from non-coplanar arrays is the w-projection algorithm \citep{wprojection-cornwell} that has been implemented in CASA. This is the fastest generic widefield imager available, and CASA is the only software package (known to us) that implements the w-projection algorithm. Other packages rely on the facetting algorithm \citep{facetting-cornwell}, which is about an order of magnitude slower and is less accurate \citep{wprojection-cornwell}. Despite the success of the w-projection algorithm, a new generation of widefield observatories has come on-line and is producing data sets that are orders of magnitude larger than before. Examples of such telescopes include the Murchison Widefield Array (MWA), the upgraded Jansky Very Large Array (JVA) and the Low-Frequency Array (LOFAR). Because of the large field of view or the enormous data volumes involved, observers are running into the limitations of the w-projection algorithm. At the MWA, we have seen that it can take up to tens of hours with the w-projection algorithm to image a two-minute MWA observation when the observed field is at low elevation. Moreover, the w-projection implementation in CASA can not handle images of 10 k $\times$ 10 k, which is desired for making full-sky images at the native resolution of the MWA. This is going to be an even bigger problem when the Square-Kilometre Array (SKA) begins its operation, and a considerably improvement in imaging techniques is required.

We present a new implementation of a generic widefield imager that significantly improves upon the w-projection implementation. It is approximately an order of magnitude faster or more in all tested situations, while it maintains equal accuracy and can handle MWA full-sky imaging. The w-stacking method is used for correcting the w-terms \citep{widefield-imaging-ska-cornwell}, and causes the increase in speed. Performance can be improved further by combination with another technique: the MWA telescope splits observations in data products of typically a few minutes, and the w-terms can be kept small by phasing an observation to the orthogonal of the array's best fitting plane. This is somewhat similar to the w-snapshot algorithm proposed by \citet{widefield-imaging-ska-cornwell}. Our imager uses in this case a hybrid between w-stacking and a variation of the w-snapshot algorithms. We named the new imager ``WSClean'', as an abbriviation for ``W-Stacking Clean''.

After describing the new imaging implementation, we will test its performance using various MWA observations, including observations with varying elevation and fields with diffuse emmission and point sources. We will compare these results with the results from the w-projection implementation in CASA and to other non-w-projection algorithms, thereby focussing on accuracy and performance.

Once the paper is accepted for publication in a scientific journal, the WSClean source code will be publicly released.

\label{lastpage}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}

### 2013-10-04 articles/Proposal-wsclean/paper-proposal-fast-widefield-imager.tex
Proposal

@@
 \label{firstpage}
-\section*{MWA Paper proposal: ``WSClean: a new implementation of a fast, generic wide-field imager''}
+\section*{MWA paper proposal: ``WSClean: a new implementation of a fast, generic wide-field imager''}
 \begin{tabular}{ll}
@@
 \section*{Paper summary}
-We present our new imaging implementation, ``WSClean'', which is an order of magnitude faster than existing implementations and can handle MWA full-sky imaging at suitable resolution.
+We present our new imager implementation, ``WSClean'', which we currently believe to be an order of magnitude faster than existing implementations, as well as that it can handle MWA full-sky imaging at suitable resolution.
 
-Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with a simple FFT based imager with an image based deconvolution procedure such as described by \citet{hogbom-clean}.
-Instead, the imaging algorithm needs to accomodate the ``w-term'' during inversion, which is the term that describes the deviation of the array to a perfect plane.
-This deviation is amplified by the field of view of the telescope, making this especially a problem for low-frequency widefield telescopes.
+Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with simple imagers that are based on a two-dimensional Fast Fourier Transformation (FFT).
+Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array to a perfect plane.
+This deviation is amplified by the field of view of the telescope, making this problem especially hard for low-frequency widefield telescopes.
 
@@
 Moreover, the w-projection implementation in CASA can not handle images of 10 k $\times$ 10 k, which is desired for making full-sky images at the native resolution of the MWA.
-This is going to be an even bigger problem when the Square-Kilometre Array (SKA) begins its operation, and a considerably improvement in imaging techniques is required.
+This is going to be an even bigger problem when the Square-Kilometre Array (SKA) begins its operation, and hence a considerable improvement in imaging techniques is required.
 
 We present a new implementation of a generic widefield imager that significantly improves upon the w-projection implementation.
-It is approximately an order of magnitude faster or more in all tested situations, while it maintains equal accuracy and can handle MWA full-sky imaging.
+Currently we believe it is approximately an order of magnitude faster in all situations and even more for very large images, while it maintains equal accuracy and can handle MWA full-sky imaging.
 The w-stacking method is used for correcting the w-terms \citep{widefield-imaging-ska-cornwell}, and causes the increase in speed.
-Performance can be improved further by combination with another technique: the MWA telescope splits observations in data products of typically a few minutes, and the w-terms can be kept small by phasing an observation to the orthogonal of the array's best fitting plane.
-This is somewhat similar to the w-snapshot algorithm proposed by \citet{widefield-imaging-ska-cornwell}.
-Our imager uses in this case a hybrid between w-stacking and a variation of the w-snapshot algorithms.
-We named the new imager ``WSClean'', as an abbriviation for ``W-Stacking Clean''.
+Performance can be improved further by combination with another technique: the MWA telescope splits observations in data products of typically a few minutes, and the w-terms can be kept small by phasing an observation to the orthogonal of the array's best-fitting plane.
+This is somewhat similar to the w-snapshot algorithm proposed in \citet{widefield-imaging-ska-cornwell}.
+Hence, we can describe the imager as a hybrid between w-stacking and a variation of the w-snapshot algorithms.
+It also implements the Clean deconvolution procedure (\citealt{hogbom-clean} and various variations).
+We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''.
 
-After describing the new imaging implementation, we will test its performance using various MWA observations, including observations with varying elevation and fields with diffuse emmission and point sources.
+After describing the new imaging implementation, we will test its performance using various MWA observations, including observations with varying elevation and fields (diffuse emmission, point sources).
 We will compare these results with the results from the w-projection implementation in CASA and to other non-w-projection algorithms, thereby focussing on accuracy and performance.
@@
 Once the paper is accepted for publication in a scientific journal, the WSClean source code will be publicly released.
+
+\begin{figure*}[bh]
+\begin{center}
+\includegraphics[width=12cm]{EoR0-apparent.png}
+\caption{A deep MWA image of EoR0 created from one hour of observing, imaged and deconvolved with WSClean in about an hour using 10 nodes on Fornax.
+Image has not been beam corrected.
+Noise level after correcting is 7 mJy.}
+\label{fig:stddev-spectrum}
+\end{center}
+\end{figure*}
 

### 2013-10-10 articles/Proposal-wsclean/paper-proposal-fast-widefield-imager.tex
Paper proposal

@@
 \usepackage[round]{natbib}
-\usepackage{wrapfig}
+\usepackage{hhline}
 
@@
 
-%\title{MWA Paper proposal:\\````WSClean: a fast implementation of a generic widefield imager''}
+%\title{MWA Paper proposal:\\````WSClean: a fast implementation of a generic wide-field imager''}
 
@@
                    & \\
-Draft publication\\
-\hspace{5mm}author list: & A.~R.~Offringa, B.~McKinley, N.~Hurley-Walker, \\
-                         & F.~H.~Briggs, MWA commissioning team and GLEAM \\
-                         & members where appropriate, MWA builders list \\
+Draft publication        & A.~R.~Offringa, B.~McKinley, N.~Hurley-Walker, \\
+\hspace{5mm}author list: & F.~H.~Briggs, J. Rhee, MWA commissioning team \\
+                         & and GLEAM members where appropriate, MWA \\
+                         & builders list \\
                    & \\
@@
 Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array to a perfect plane.
-This deviation is amplified by the field of view of the telescope, making this problem especially hard for low-frequency widefield telescopes.
+This deviation is amplified by the field of view of the telescope, making this problem especially hard for low-frequency wide-field telescopes.
 
 The most commonly used method to image observations from non-coplanar arrays is the w-projection algorithm \citep{wprojection-cornwell} that has been implemented in CASA.
-This is the fastest generic widefield imager available, and CASA is the only software package (known to us) that implements the w-projection algorithm.
-Other packages rely on the facetting algorithm \citep{facetting-cornwell}, which is about an order of magnitude slower and is less accurate \citep{wprojection-cornwell}.
-Despite the success of the w-projection algorithm, a new generation of widefield observatories has come on-line and is producing data sets that are orders of magnitude larger than before.
-Examples of such telescopes include the Murchison Widefield Array (MWA), the upgraded Jansky Very Large Array (JVA) and the Low-Frequency Array (LOFAR).
+This is the fastest generic wide-field imager available, and CASA is the only software package (known to us) that implements the w-projection algorithm.
+Other packages rely on the faceting algorithm \citep{facetting-cornwell}, which is about an order of magnitude slower and is less accurate \citep{wprojection-cornwell}.
+Despite the success of the w-projection algorithm, a new generation of wide-field observatories has come on-line and is producing data sets that are orders of magnitude larger than before.
+Examples of such telescopes include the Murchison Widefield Array (MWA), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR).
 Because of the large field of view or the enormous data volumes involved, observers are running into the limitations of the w-projection algorithm.
@@
 
-We present a new implementation of a generic widefield imager that significantly improves upon the w-projection implementation.
+We present a new implementation of a generic wide-field imager that significantly improves upon the w-projection implementation.
 Currently we believe it is approximately an order of magnitude faster in all situations and even more for very large images, while it maintains equal accuracy and can handle MWA full-sky imaging.
@@
 
-After describing the new imaging implementation, we will test its performance using various MWA observations, including observations with varying elevation and fields (diffuse emmission, point sources).
-We will compare these results with the results from the w-projection implementation in CASA and to other non-w-projection algorithms, thereby focussing on accuracy and performance.
+After describing the new imaging implementation, we will test its performance using various MWA observations, including observations with varying elevation and fields (diffuse emission, point sources).
+We will compare these results with the results from the w-projection implementation in CASA and to other non-w-projection algorithms, thereby focusing on accuracy and performance.
+The data sets that we currently envision to use are summarized in Table~\ref{tbl:observations}.
 
@@
 
+\begin{table}
+\caption{Data sets envisioned to be used in publication} \label{tbl:observations}
+ \begin{tabular}{|c|l|p{6cm}|}
+ \hhline{|=|=|=|}
+ \textbf{Observatory} & \textbf{Description} & \textbf{Reasons} \\
+ \textbf{+ date} & & \\
+ \hhline{|=|=|=|}
+  2013-05-02 & Decl -55          & $\bullet$ Test imaging at different frequencies\\
+  MWA 128T   &                   & $\bullet$ Show computational performance \\
+             &                   & \hspace{5mm} at elevation $\sim 60$\\
+             &                   & $\bullet$ Full-sky imaging \\
+ \hline
+  2013-05-16 & Vela and Puppis A & $\bullet$ Test cleaning of diffuse structure \\
+  MWA 128T   &                   & $\bullet$ Show computational performance \\
+             &                   & \hspace{5mm} of cleaning\\
+             &                   & $\bullet$ Pointed observations \\
+ \hline
+  2013-07-31 & EoR0              & $\bullet$ Test cleaning of point sources \\
+  MWA 128T   &                   & $\bullet$ Test deep imaging \\
+ \hline
+  ---        & Simulations       & $\bullet$ Validate photometry / astrometry \\
+             &                   & $\bullet$ Test computational performance \\
+             &                   & \hspace{5mm} at very low elevation\\
+ \hhline{|=|=|=|}
+\end{tabular}
+\end{table}
+
 \label{lastpage}

### 2013-10-15 articles/Proposal-wsclean/paper-proposal-fast-widefield-imager.tex
Comments by Frank

@@
 \section*{Paper summary}
-We present our new imager implementation, ``WSClean'', which we currently believe to be an order of magnitude faster than existing implementations, as well as that it can handle MWA full-sky imaging at suitable resolution.
+We present our new imager implementation, ``WSClean'', which we find to be an order of magnitude faster than existing implementations, as well as being capable of MWA full-sky imaging at full resolution.
 
 Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with simple imagers that are based on a two-dimensional Fast Fourier Transformation (FFT).
-Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array to a perfect plane.
-This deviation is amplified by the field of view of the telescope, making this problem especially hard for low-frequency wide-field telescopes.
+Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane.
+This deviation is amplified for telescopes with wide field of view, making this problem especially hard for low-frequency wide-field telescopes.
 
@@
 Other packages rely on the faceting algorithm \citep{facetting-cornwell}, which is about an order of magnitude slower and is less accurate \citep{wprojection-cornwell}.
-Despite the success of the w-projection algorithm, a new generation of wide-field observatories has come on-line and is producing data sets that are orders of magnitude larger than before.
+The w-projection algorithm has thus been successful so far.
+However, the available implementations are unable to cope with the data rates from the new generation of wide-field observatories, which are producing data sets that are orders of magnitude larger than before.
 Examples of such telescopes include the Murchison Widefield Array (MWA), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR).
-Because of the large field of view or the enormous data volumes involved, observers are running into the limitations of the w-projection algorithm.
 At the MWA, we have seen that it can take up to tens of hours with the w-projection algorithm to image a two-minute MWA observation when the observed field is at low elevation.
 Moreover, the w-projection implementation in CASA can not handle images of 10 k $\times$ 10 k, which is desired for making full-sky images at the native resolution of the MWA.
-This is going to be an even bigger problem when the Square-Kilometre Array (SKA) begins its operation, and hence a considerable improvement in imaging techniques is required.
+This will become an even bigger problem when the Square-Kilometre Array (SKA) begins its operation, and hence a considerable improvement in imaging techniques is required.
 
 We present a new implementation of a generic wide-field imager that significantly improves upon the w-projection implementation.
-Currently we believe it is approximately an order of magnitude faster in all situations and even more for very large images, while it maintains equal accuracy and can handle MWA full-sky imaging.
-The w-stacking method is used for correcting the w-terms \citep{widefield-imaging-ska-cornwell}, and causes the increase in speed.
+Currently it performs approximately an order of magnitude faster in all situations and even more for very large images, while it maintains equal accuracy and can handle MWA full-sky imaging.
+The implementation uses the ``w-stacking method'' for correcting the w-terms \citep{widefield-imaging-ska-cornwell} to obtain the increase in speed.
 Performance can be improved further by combination with another technique: the MWA telescope splits observations in data products of typically a few minutes, and the w-terms can be kept small by phasing an observation to the orthogonal of the array's best-fitting plane.
 This is somewhat similar to the w-snapshot algorithm proposed in \citet{widefield-imaging-ska-cornwell}.
-Hence, we can describe the imager as a hybrid between w-stacking and a variation of the w-snapshot algorithms.
+Hence, we can describe the imager as a hybrid between w-stacking and a variation of the w-snapshot algorithm.
 It also implements the Clean deconvolution procedure (\citealt{hogbom-clean} and various variations).
@@
   2013-05-02 & Decl -55          & $\bullet$ Test imaging at different frequencies\\
-  MWA 128T   &                   & $\bullet$ Show computational performance \\
+  MWA 128T   & GLEAM test data   & $\bullet$ Show computational performance \\
              &                   & \hspace{5mm} at elevation $\sim 60$\\

### 2013-10-15 articles/Proposal-wsclean/paper-proposal-fast-widefield-imager.tex
Some minor changes

--- 2013-10-10 articles/Proposal-wsclean/paper-proposal-fast-widefield-imager.tex
@@
 \begin{tabular}{ll}
-Publication Title: & WSClean: a new implementation of a fast, generic \\
+Publication title: & WSClean: a new implementation of a fast, generic \\
                    & wide-field imager\\
                    & \\
-Principle Contact: & Dr. Andr\'e R. Offringa \\
+Principle contact: & Dr. Andr\'e R. Offringa \\
                    & \\
@@
                    & \\
-Intended Journal: & Probably MNRAS\\
+Journal:           & Probably MNRAS\\
                    & \\
@@
 After describing the new imaging implementation, we will test its performance using various MWA observations, including observations with varying elevation and fields (diffuse emission, point sources).
+GMRT data and -- if available -- LOFAR data will also be used.
 We will compare these results with the results from the w-projection implementation in CASA and to other non-w-projection algorithms, thereby focusing on accuracy and performance.
@@
 Image has not been beam corrected.
-Noise level after correcting is 7 mJy.}
+Noise level after correcting is 7 mJy.
+200 sources were subtracted from the centre before imaging.}
 \label{fig:stddev-spectrum}
@@
  \hline
-  ---        & Simulations       & $\bullet$ Validate photometry / astrometry \\
-             &                   & $\bullet$ Test computational performance \\
+  2008-2009  & ZCosmos           & $\bullet$ Test relevance to GMRT data \\
+  GMRT       & (stacking)        & $\bullet$ Verify making image cubes \\
+ \hline
+  LOFAR      & To be determined  & $\bullet$ Test relevance to LOFAR data \\
+             &                   & (only if data easily obtainable) \\
+ \hline
+             & Simulations       & $\bullet$ Validate photometry / astrometry \\
+   ---       &                   & $\bullet$ Test computational performance \\
              &                   & \hspace{5mm} at very low elevation\\

### 2013-10-16 articles/Proposal-wsclean/paper-proposal-fast-widefield-imager.tex
Slight changes

--- 2013-10-15 articles/Proposal-wsclean/paper-proposal-fast-widefield-imager.tex
@@
 Draft publication        & A.~R.~Offringa, B.~McKinley, N.~Hurley-Walker, \\
-\hspace{5mm}author list: & F.~H.~Briggs, J. Rhee, MWA commissioning team \\
+\hspace{5mm}author list: & F.~H.~Briggs, J.~Rhee, L.~Feng, D.~Kaplan, \\
+                         & A.~R.~Neben, MWA commissioning team \\
                          & and GLEAM members where appropriate, MWA \\

### 2013-10-22 pbs-scripts/clear_scripts.sh
Adding Fornax pbs scripts

@@ -1,0 +1,1 @@
+rm ~/pbs_work/* -v

### 2013-10-22 pbs-scripts/killall.sh

@@ -1,0 +1,1 @@
+qselect -u $USER | xargs qdel

### 2013-10-22 pbs-scripts/queue-applycal.sh

@@ -1,0 +1,21 @@
+listfile=$1
+solfile=$2
+jobid=$3
+
+if [[ "${solfile}" == "" ]] ; then
+  echo Syntax: $0 \<listfile\> \<solution file\> \<jobid\>
+else
+  declare -i index=0
+  for ms in `cat ${listfile}` ; do
+    jobname="apply-${jobid}-${index}"
+    jobscript="${HOME}/pbs_work/${jobname}.script"
+    sed s/JOB_NAME/${jobname}/ ${HOME}/scripts/pbs-header.txt|sed s/WALLTIME/00:15:00/ > ${jobscript}
+    
+    echo "applysolutions ${ms} ${solfile}" >> ${jobscript}
+
+    index=${index}+1
+    chmod u+x ${jobscript}
+    echo "qsub ${jobscript}"
+  done
+  echo "qstat -a -t -u \${USER}"
+fi

### 2013-10-22 pbs-scripts/queue-calibrate.sh

@@ -1,0 +1,28 @@
+listfile=$1
+model=$2
+outputdir=$3
+
+if [[ "${outputdir}" == "" ]] ; then
+  echo Syntax: $0 \<listfile\> \<model\> \<solution outputdir\>
+else
+
+  mkdir -p ${outputdir}
+  declare -i index=0
+  echo To execute:
+  for ms in `cat ${listfile}` ; do
+    jobname="cal${index}-`date +%H%M%S`"
+    jobscript="${HOME}/pbs_work/${jobname}.script"
+    sed s/JOB_NAME/${jobname}/ ${HOME}/scripts/pbs-header.txt|sed s/WALLTIME/00:30:00/ > ${jobscript}
+    solfile="${ms##*/}"
+    solfile="${solfile%/}"
+    outputdir="${outputdir%/}"
+    solfile="${outputdir}/${solfile%.ms}-sol.bin"
+    
+    echo "calibrate -i 100 -a 0.00001 0.000001 -applybeam -diag -m ${model} ${ms} ${solfile} > ${HOME}/pbs_work/${jobname}-out.txt" >> ${jobscript}
+
+    index=${index}+1
+    chmod u+x ${jobscript}
+    echo "qsub ${jobscript}"
+  done
+  echo "qstat -a -t -u \${USER}"
+fi

### 2013-10-22 pbs-scripts/queue-dec18.6.sh

@@ -1,0 +1,1 @@
+./queue-imaging.sh /home/nhurleywalker/data/nhurleywalker/G0008/Dec18.6.txt /home/nhurleywalker/data/nhurleywalker/G0008 /group/astronomy556/aoffringa/dec18.6

### 2013-10-22 pbs-scripts/queue-expuvfits.sh

@@ -1,0 +1,25 @@
+listfile=$1
+outdir=$2
+
+if [[ "${outdir}" == "" ]] ; then
+  echo Syntax: $0 \<listfile\> \<outputdir\>
+else
+  declare -i index=0
+  for ms in `cat ${listfile}` ; do
+    jobname="exp-${index}"
+    jobscript="${HOME}/pbs_work/${jobname}.script"
+    sed s/JOB_NAME/${jobname}/ ${HOME}/scripts/pbs-header.txt|sed s/WALLTIME/00:15:00/ > ${jobscript}
+    
+    fitsfile="${ms##*/}"
+    fitsfile="${fitsfile%/}"
+    outdir="${outdir%/}"
+    fitsfile="${outdir}/${fitsfile%.ms}.uvfits"
+
+    echo "casapy --nologger --nologfile --nogui -c \"exportuvfits(vis='${ms}',fitsfile='${fitsfile}',datacolumn='data')\"" >> ${jobscript}
+
+    index=${index}+1
+    chmod u+x ${jobscript}
+    echo "qsub ${jobscript}"
+  done
+  echo "qstat -a -t -u \${USER}"
+fi

### 2013-10-22 pbs-scripts/queue-imaging.sh

@@ -1,0 +1,30 @@
+listfile=$1
+outputdir=$2
+
+if [[ "${outputdir}" == "" ]] ; then
+  echo Syntax: $0 \<listfile\> \<image outputdir\>
+else
+  mkdir -p ${outputdir}
+  jobname="img-`date +%H%M%S`"
+  jobscript="${HOME}/pbs_work/job-${jobname}.script"
+  sed s/JOB_NAME/${jobname}/ ${HOME}/scripts/pbs-header-arr.txt > ${jobscript}
+  declare -i index=0
+  for ms in `cat ${listfile}` ; do
+    outputprefix="${ms##*/}"
+    outputprefix="${outputprefix%/}"
+    outputdir="${outputdir%/}"
+    outputprefix="${outputdir}/${outputprefix%.ms}"
+
+    
+    echo "imgname[${index}]=${outputprefix}" >> ${jobscript}
+    echo "ms[${index}]=${ms}" >> ${jobscript}
+
+    index=${index}+1
+  done
+  declare -i maxindex=${index}-1
+  chmod u+x ${jobscript}
+
+  echo "wsclean -size 4096 4096 -scale 0.012 -nwlayers 256 -niter 20000 -gain 0.08 -mgain 0.9 -negative -gridmode kb -threshold 0.3 -weight uniform -name \${imgname[\${JOBID}]} \${ms[\${JOBID}]} >& ${HOME}/pbs_work/log-${jobname}-\${JOBID}.txt" >> ${jobscript}
+  echo To execute:
+  echo "qsub -J 0-${maxindex} ${jobscript} ; qstat -a -t -u \${USER}"
+fi

### 2013-10-22 pbs-scripts/queue-polimage.sh

@@ -1,0 +1,60 @@
+ms=$1
+
+if [[ "${ms}" == "" ]] ; then
+  echo Syntax: $0 \<ms-full-path\>
+else
+  niter=10000
+  size=2560
+  scale=0.01
+
+  outputdir="${ms%/}"
+  outputdir="${outputdir%.ms}"
+  filename=`basename ${outputdir}`
+  outputdir=`dirname ${outputdir}`
+  rootname="${outputdir}/${filename}"
+
+  polpar[0]="-pol xx -name ${rootname}-xx -niter ${niter}"
+  polpar[1]="-pol xy -name ${rootname}-xy -niter ${niter}"
+  polpar[2]="-pol yx -name ${rootname}-yx"
+  polpar[3]="-pol yy -name ${rootname}-yy"
+  polpar[4]="-pol xy -imaginarypart -name ${rootname}-xyi"
+  polpar[5]="-pol yx -imaginarypart -name ${rootname}-yxi"
+  dependencies=""
+  for (( p=0 ; p!=6 ; p++ )) ; do
+    jobname="pimg${p}"
+    jobscript="${HOME}/pbs_work/job-${jobname}-${filename}.script"
+    sed s/JOB_NAME/${jobname}/ ${HOME}/scripts/pbs-header.txt|sed s/WALLTIME/00:30:00/ > ${jobscript}
+    echo "wsclean ${polpar[p]} -size ${size} ${size} -scale ${scale} >& ${HOME}/pbs_work/log-${jobname}.txt ${ms}" >> ${jobscript}
+    chmod u+x ${jobscript}
+    echo "qsub ${jobscript}"
+    jobnum=`qsub ${jobscript}`
+    if [[ ${p} == 0 ]] ; then
+	dependencies="depend=afterok:${jobnum}"
+    else
+	dependencies="${dependencies},afterok:${jobnum}"
+    fi
+  done
+
+  jobname="pimgpb"
+  jobscript="${HOME}/pbs_work/job-${jobname}-${filename}.script"
+  sed s/JOB_NAME/${jobname}/ ${HOME}/scripts/pbs-header.txt|sed s/WALLTIME/00:15:00/ > ${jobscript}
+  echo "cd ${outputdir}" >> ${jobscript}
+  echo "beam -proto ${rootname}-xx-image.fits -ms ${ms}"  >> ${jobscript}
+  echo -e -n "pbcorrect " >> ${jobscript}
[…]

### 2013-10-22 pbs-scripts/queue-selfcal.sh

@@ -1,0 +1,29 @@
+listfile=$1
+model=$2
+outputdir=$3
+
+if [[ "${outputdir}" == "" ]] ; then
+  echo Syntax: $0 \<listfile\> \<model\> \<solution outputdir\>
+else
+
+  mkdir -p ${outputdir}
+  declare -i index=0
+  echo To execute:
+  for ms in `cat ${listfile}` ; do
+    jobname="sc${index}-`date +%H%M%S`"
+    jobscript="${HOME}/pbs_work/${jobname}.script"
+    sed s/JOB_NAME/${jobname}/ ${HOME}/scripts/pbs-header.txt|sed s/WALLTIME/04:00:00/ > ${jobscript}
+    solfile="${ms##*/}"
+    solfile="${solfile%/}"
+    outputdir="${outputdir%/}"
+    solfile="${outputdir}/${solfile%.ms}-sol.bin"
+    
+    echo "calibrate -t 2 -i 100 -a 0.00001 0.000001 -applybeam -m ${model} ${ms} ${solfile} > ${HOME}/pbs_work/${jobname}-out.txt" >> ${jobscript}
+    echo "applysolutions ${ms} ${solfile}" >> ${jobscript}
+
+    index=${index}+1
+    chmod u+x ${jobscript}
+    echo "qsub ${jobscript}"
+  done
+  echo "qstat -a -t -u \${USER}"
+fi

### 2013-10-22 pbs-scripts/queue-spectrum.sh

@@ -1,0 +1,26 @@
+listfile=$1
+model=$2
+outputdir=$3
+
+if [[ "${outputdir}" == "" ]] ; then
+  echo Syntax: $0 \<listfile\> \<model\> \<outputdir\>
+else
+  declare -i index=0
+  for ms in `cat ${listfile}` ; do
+    jobname="spectrum-${index}"
+    jobscript="${HOME}/pbs_work/${jobname}.script"
+    sed s/JOB_NAME/${jobname}/ ${HOME}/scripts/pbs-header.txt|sed s/WALLTIME/01:30:00/ > ${jobscript}
+    
+    outp="${ms##*/}"
+    outp="${outp%/}"
+    outputdir="${outputdir%/}"
+    outp="${outputdir}/${outp%.ms}-spectrum.txt"
+
+    echo "spectrum -applybeam ${model} ${outp} ${ms} > ${HOME}/pbs_work/${jobname}-out.txt " >> ${jobscript}
+
+    index=${index}+1
+    chmod u+x ${jobscript}
+    echo "qsub ${jobscript}"
+  done
+  echo "qstat -a -t -u \${USER}"
+fi

### 2013-10-22 pbs-scripts/queue-subtract.sh

@@ -1,0 +1,25 @@
+listfile=$1
+model=$2
+
+if [[ "${model}" == "" ]] ; then
+  echo Syntax: $0 \<listfile\> \<model\>
+else
+  declare -i index=0
+  for ms in `cat ${listfile}` ; do
+    jobname="sub-${index}"
+    jobscript="${HOME}/pbs_work/${jobname}.script"
+    sed s/JOB_NAME/${jobname}/ ${HOME}/scripts/pbs-header.txt|sed s/WALLTIME/01:30:00/ > ${jobscript}
+    
+    outp="${ms##*/}"
+    outp="${outp%/}"
+    outputdir="${outputdir%/}"
+    outp="${outputdir}/${outp%.ms}-spectrum.txt"
+
+    echo "subtrmodel -applybeam ${model} ${ms} > ${HOME}/pbs_work/${jobname}-out.txt " >> ${jobscript}
+
+    index=${index}+1
+    chmod u+x ${jobscript}
+    echo "qsub ${jobscript}"
+  done
+  echo "qstat -a -t -u \${USER}"
+fi

### 2013-10-22 articles/Proposal-wsclean/paper-proposal-fast-widefield-imager.tex

--- 2013-10-15 articles/Proposal-wsclean/paper-proposal-fast-widefield-imager.tex
@@
 Draft publication        & A.~R.~Offringa, B.~McKinley, N.~Hurley-Walker, \\
-\hspace{5mm}author list: & F.~H.~Briggs, J. Rhee, MWA commissioning team \\
-                         & and GLEAM members where appropriate, MWA \\
-                         & builders list \\
+\hspace{5mm}author list: & F.~H.~Briggs, J.~Rhee, L.~Feng, D.~Kaplan, \\
+                         & A.~R.~Neben, J.~D.~Hughes, MWA commissioning \\
+                         & team and GLEAM members where appropriate, \\
+                         & MWA builders list \\
                    & \\

### 2013-10-23 pbs-scripts/queue-polimage.sh
Fixing niter in YY

@@ -14,9 +14,9 @@ else
   rootname="${outputdir}/${filename}"
 
   polpar[0]="-pol xx -name ${rootname}-xx -niter ${niter}"
-  polpar[1]="-pol xy -name ${rootname}-xy -niter ${niter}"
+  polpar[1]="-pol xy -name ${rootname}-xy"
   polpar[2]="-pol yx -name ${rootname}-yx"
-  polpar[3]="-pol yy -name ${rootname}-yy"
+  polpar[3]="-pol yy -name ${rootname}-yy -niter ${niter}"
   polpar[4]="-pol xy -imaginarypart -name ${rootname}-xyi"
   polpar[5]="-pol yx -imaginarypart -name ${rootname}-yxi"
   dependencies=""

### 2013-12-01 articles/Report-Thoughts/thoughts.tex
Some equations

\documentclass[a4paper,11pt]{article} 
%\usepackage[style=nature,backend=biber]{biblatex}
\usepackage{graphicx}
\usepackage[font={it},labelfont={bf}]{caption}
\usepackage{times}
\usepackage{color}
%\usepackage{fixltx2e} % fixes placing of image positions
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{courier}

\definecolor{customhdrcolor}{rgb}{0.0,0.0,0.0}
%\definecolor{customcitecolor}{rgb}{0.0,0.25,0.25}
\definecolor{customcitecolor}{rgb}{0.0,0.5,0.75}
%\definecolor{customlinkcolor}{rgb}{0.0,0.0,1.0}
\definecolor{customlinkcolor}{rgb}{0.0,0.5,0.75}

\usepackage[colorlinks=true,linkcolor=customlinkcolor,urlcolor=customlinkcolor,citecolor=customcitecolor,pdftex]{hyperref}

\ifpdf\pdfinfo{/Title      (Thoughts on calibration)
               /Author     (A. R. Offringa)
               /Keywords   (calibration)
        }
\else\usepackage{graphics}\fi

\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\title{Thoughts on calibration}

\author{A.~R.~Offringa}
%\author[1,2]{A.~R.~Offringa}
%\affil[1]{RSAA, Australian National University, Mt Stromlo Observatory, via Cotter Road, Weston, ACT 2611, Australia}
%\affil[2]{ARC Centre of Excellence for All-Sky Astrophysics (CAASTRO)}
\begin{document}

\label{firstpage}
\maketitle

\section{Calibrating the ionospheric phase}
Mitch minimizes the following equation for calibration in the RTS:
\begin{equation}
\sum\limits_{j=1}^{N_a} \sum\limits_{k,k\neq j}^{N_a} \left\| \mathbf{M}_{j,k} - \mathbf{J}_j \mathbf{V}_{j,k} \overline{\mathbf{J}}_k \right\|^2_F,
\end{equation}
where $\mathbf{M}_{j,k}$ and $\mathbf{V}_{j,k}$ are the predicted and observed visibility matrices respectively for baseline $j,k$ and $\mathbf{J}_j$ is the Jones matrix for antenna $j$. Each of these are $2\times 2$ matrices.

For the MWA, the relevant effects within $\mathbf{J}$ can be decomposed (in order) in the ionospheric phase term $j_I$, Faraday rotation $\mathbf{J}_F$, the beam $\mathbf{J}_B$ and the instrumental gain $\mathbf{J}_G$:
\begin{equation}
 \mathbf{J} = j_I \mathbf{J}_F \mathbf{J}_B \mathbf{J}_G.
\end{equation}
$j_I$ is a scalar term and commutes with everything. $\mathbf{J}_F$ is a rotation matrix and does not commute with $\mathbf{J}_B$, which itself can perform shear, rotation and scale transforms. $\mathbf{J}_G$ is a diagonal (scaling) matrix, and does not commute with $\mathbf{J}_B$ either.

Once a solution $\mathbf{J}$ is found, one can try to decompose it into these terms. To start off, the value of $\mathbf{J}_B$ is (approximately) known. However, because it does not commute with its neighbouring terms, one cannot divide it out. This even remains the case when the model is unpolarized. Fortunately, in MWA all tiles are looking approximately through the same ionosphere, and therefore we can assume $\mathbf{J}_F$ is equal for every antenna. As long as the model is unpolarized, the model becomes a scalar term $m$ and the $\mathbf{J}_F$ term is eliminated:
\begin{equation}
 \mathbf{J}_F^{-1} m \overline{\mathbf{J}}_F^{-1} = m \left( \mathbf{J}_F^{-1} \overline{\mathbf{J}}_F^{-1} \right) = m.
\end{equation}
This implies one can divide the beam from the model, such that $\mathbf{N}_{j,k} = \mathbf{J}_B^{-1}m_{j,k}\overline{\mathbf{J}}_B^{-1}$ and one arrives at:
\begin{equation}
 \sum\limits_{j=1}^{N_a} \sum\limits_{k,k\neq j}^{N_a} \left\| \mathbf{N}_{j,k} - j_j^I \mathbf{J}^G_j \mathbf{V}_{j,k} \overline{\mathbf{J}}_k^G \overline{j}_k^I \right\|^2_F.
\end{equation}
It is clear we can not trivially distinguish between the ionospheric phase term and the instrumental (complex) gain. However, if the instrumental gain is constant over a long time while the ionosphere changes on scales of minutes but has a unity average over long term, one can solve first for the instrumental gain by taking the solution over a long interval, and perform ionospheric solutions over short intervals. Assume that $W$ are the corrected visibilities after instrumental gain calibration.
\begin{equation}
 \sum\limits_{j=1}^{N_a} \sum\limits_{k,k\neq j}^{N_a} \left\| \mathbf{N}_{j,k} - j_j^I \mathbf{W}_{j,k} \overline{j}_k^I \right\|^2_F.
\end{equation}
Since $j_j^I = \overline{j}^I_k$, this is equal to
\begin{equation}
 \sum\limits_{j=1}^{N_a} \sum\limits_{k,k\neq j}^{N_a} \left\| \mathbf{N}_{j,k} - j_I^2 \mathbf{W}_{j,k} \right\|^2_F.
\end{equation}
This can now become a scalar problem for each polarisation:
\begin{equation}
 \sum\limits_{j=1}^{N_a} \sum\limits_{k,k\neq j}^{N_a} \left( \sum\limits_{p=1}^4 \left[ n_{j,k,p} - j_I^2 w_{j,k,p} \right] \right)^2.
\end{equation}

\label{lastpage}

\end{document}

### 2013-12-05 articles/Report-Thoughts/thoughts.tex
Position fitting

@@
 
+This method works, but does not deal with ionospheric position shifting, and hence does in practice fail to do an accurate subtraction job when the source position are not found every short interval (2 min).
+
+\section{Source position finding}
+Instead of fitting for a global gain and phase, we can also fit for a global (real) gain $g$ and two parameters $\Delta l$ and $\Delta m$ that describe the movement of the source.
+We would then want to calculate the minimization
+\begin{equation}
+ \min_{g,\Delta l,\Delta m} \epsilon(g,\Delta l,\Delta m),\hspace{1cm} g\approx 1, \Delta l\approx 0, \Delta m\approx 0
+\end{equation}
+where $\epsilon$ is defined as the squared error between model and data:
+\begin{equation}
+ \epsilon(g,\Delta l,\Delta m) = \left( g e^{2\pi \left( u\Delta l + v\Delta m\right)} M(i) - V(i) \right)^2
+\end{equation}
+
 \label{lastpage}

### 2013-12-06 articles/Report-Thoughts/thoughts.tex
Iterate peeling multiple times and subtract sources correctly

--- 2013-12-01 articles/Report-Thoughts/thoughts.tex
@@
 
+This method works, but does not deal with ionospheric position shifting, and hence does in practice fail to do an accurate subtraction job when the source position are not found every short interval (2 min).
+
+\section{Source position finding}
+One can fit for a global (real) gain $g$ and two parameters $\Delta l$ and $\Delta m$ that describe the movement of the source. To do so, one calculates the optimization problem
+\begin{equation}
+ \min_{g,\Delta l,\Delta m} \epsilon(g,\Delta l,\Delta m),\hspace{1cm} g,\Delta l,\Delta m \in \mathcal{R}, g\approx 1, \Delta l\approx 0, \Delta m\approx 0
+\end{equation}
+where the error $\epsilon$ is defined as the squared magnitude of the difference between model and data:
+\begin{equation}
+ \epsilon(g,\Delta l,\Delta m) = \left| g e^{2\pi \left( u\Delta l + v\Delta m\right)} M(i) - V(i) \right|^2
+\end{equation}
+
 \label{lastpage}

### 2014-01-12 articles/2014/Article-WSClean/wsclean.tex

--- 2013-10-16 articles/Proposal-wsclean/paper-proposal-fast-widefield-imager.tex
@@
-\documentclass[a4paper,10pt]{article} 
-%\usepackage[style=nature,backend=biber]{biblatex}
+\documentclass[useAMS,usenatbib]{mn2e}
+\usepackage[utf8]{inputenc}
 \usepackage{graphicx}
-\usepackage[font={it},labelfont={bf}]{caption}
+\usepackage{amsmath}
+\usepackage{amssymb}
+%\usepackage[font={it},labelfont={bf}]{caption}
+%\usepackage{units}
 \usepackage{times}
 \usepackage{color}
-%\usepackage{fixltx2e} % fixes placing of image positions
-\usepackage{authblk}
-\usepackage{amsmath}
-\usepackage{courier}
-\usepackage[round]{natbib}
-\usepackage{hhline}
+\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
+
+\usepackage{fixltx2e} % fixes placing of image positions
 
 \definecolor{customhdrcolor}{rgb}{0.0,0.0,0.0}
-%\definecolor{customcitecolor}{rgb}{0.0,0.25,0.25}
 \definecolor{customcitecolor}{rgb}{0.0,0.5,0.75}
-%\definecolor{customlinkcolor}{rgb}{0.0,0.0,1.0}
 \definecolor{customlinkcolor}{rgb}{0.0,0.5,0.75}
@@
 
-\ifpdf\pdfinfo{/Title      (MWA paper proposal Offringa)
-               /Author     (A. R. Offringa)
-               /Keywords   (MWA, imaging)
+\ifpdf\pdfinfo{/Title      (WSClean)
+               /Author     (A. R. Offringa et al.)
+               /Keywords   (instrumentation: interferometers;methods: observational;techniques: interferometric;radio continuum: general)
         }
@@
 
-%\title{MWA Paper proposal:\\````WSClean: a fast implementation of a generic wide-field imager''}
+%\newcommand{\editmark}[1]{#1}
+%\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
 
-%\author{A.~R.~Offringa}
-%\author[1,2]{A.~R.~Offringa}
-%\affil[1]{RSAA, Australian National University, Mt Stromlo Observatory, via Cotter Road, Weston, ACT 2611, Australia}
-%\affil[2]{ARC Centre of Excellence for All-Sky Astrophysics (CAASTRO)}
+% To make Dutch ``tussenvoegsels'' work correctly in Latex such as ``de Bruyn'', we use this command
+% It fixes ordering and uppercases.
+% In text, it should be written with uppercase, as in ``De Bruyn''
+\DeclareRobustCommand{\TUSSEN}[3]{#2}
+
+\title[WSClean: a new imager]{WSClean: a new implementation of a fast, generic wide-field imager}
+
+\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{E-mail:
+\url{andre.offringa@anu.edu.au}}, B.~McKinley$^{1,2}$, N.~Hurley-Walker, 
+F.~H.~Briggs, \newauthor
+J.~Rhee, L.~Feng, D.~Kaplan, A.~R.~Neben, J.~D.~Hughes, MWA commissioning team\newauthor
+and GLEAM members where appropriate \& MWA builders list
+\\
+$^{1}$RSAA, Australian National University, Mt Stromlo Observatory, via Cotter Road, Weston, ACT 2611, Australia \\
+$^{2}$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO) \\
+}
+
 \begin{document}
 
+\date{Accepted TODO. Received TODO; in original form TODO}
+\pagerange{\pageref{firstpage}--\pageref{lastpage}}
+\pubyear{2014}
+
 \label{firstpage}
-\section*{MWA paper proposal: ``WSClean: a new implementation of a fast, generic wide-field imager''}
-\begin{tabular}{ll}
-Publication Title: & WSClean: a new implementation of a fast, generic \\
-                   & wide-field imager\\
-                   & \\
-Principle Contact: & Dr. Andr\'e R. Offringa \\
-                   & \\
-Draft publication        & A.~R.~Offringa, B.~McKinley, N.~Hurley-Walker, \\
-\hspace{5mm}author list: & F.~H.~Briggs, J.~Rhee, L.~Feng, D.~Kaplan, \\
-                         & A.~R.~Neben, MWA commissioning team \\
-                         & and GLEAM members where appropriate, MWA \\
-                         & builders list \\
-                   & \\
-Intended Journal: & Probably MNRAS\\
-                   & \\
-Intended date of submission\\
-\hspace{5mm}to collaboration: & February 2014\\
-\end{tabular}
+\maketitle
 
-\section*{Paper summary}
+\begin{abstract}
 We present our new imager implementation, ``WSClean'', which we find to be an order of magnitude faster than existing implementations, as well as being capable of MWA full-sky imaging at full resolution.
+TODO
+\end{abstract}
+
+\begin{keywords}
+instrumentation: interferometers -- methods: observational -- techniques: interferometric -- radio continuum: general
+\end{keywords}
+
+\section{Introduction}
+
+\section*{Acknowledgments}
+...
+
+% To make Dutch ``tussenvoegsels'' work correctly in Latex such as ``de Bruyn'', we use this command
+% In bibliography, it should be written with lowercase, as in ``de Bruyn''
+\DeclareRobustCommand{\TUSSEN}[3]{#3}
 
@@
 Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane.
-This deviation is amplified for telescopes with wide field of view, making this problem especially hard for low-frequency wide-field telescopes.
+This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes, that by nature are wide-field instruments.
 
@@
 After describing the new imaging implementation, we will test its performance using various MWA observations, including observations with varying elevation and fields (diffuse emission, point sources).
+GMRT data and -- if available -- LOFAR data will also be used.
 We will compare these results with the results from the w-projection implementation in CASA and to other non-w-projection algorithms, thereby focusing on accuracy and performance.
-The data sets that we currently envision to use are summarized in Table~\ref{tbl:observations}.
 
-Once the paper is accepted for publication in a scientific journal, the WSClean source code will be publicly released.
+\section{The w-stacking technique}
+In this section, we will describe the w-stacking algorithm from a mathematical point of view.
 
-\begin{figure*}[bh]
-\begin{center}
-\includegraphics[width=12cm]{EoR0-apparent.png}
-\caption{A deep MWA image of EoR0 created from one hour of observing, imaged and deconvolved with WSClean in about an hour using 10 nodes on Fornax.
-Image has not been beam corrected.
-Noise level after correcting is 7 mJy.}
-\label{fig:stddev-spectrum}
-\end{center}
-\end{figure*}
+The sky equation for an interferometer is given by
+\begin{eqnarray}\label{eq:visibility-function}
+V(u,v,w) = \iint \frac{A(l,m) I(l,m)}{\sqrt{1-l^2-m^2}} e^{-2\pi i \left(ul + vm + w(\sqrt{1-l^2-m^2}-1)\right)} dl dm,
+\end{eqnarray}
+where function $V$ is formed by the calibrated visibilities, $u,v,w$ is a coordinate giving the observed mode in the coordinate system of the array, $A$ is the primary-beam function, $I$ is the sky function and $l,m$ are cosine sky coordinates.
+Imaging consists of inverting this equation to find $I$ from $V$.
 
-\begin{table}
-\caption{Data sets envisioned to be used in publication} \label{tbl:observations}
- \begin{tabular}{|c|l|p{6cm}|}
- \hhline{|=|=|=|}
- \textbf{Observatory} & \textbf{Description} & \textbf{Reasons} \\
- \textbf{+ date} & & \\
- \hhline{|=|=|=|}
-  2013-05-02 & Decl -55          & $\bullet$ Test imaging at different frequencies\\
-  MWA 128T   & GLEAM test data   & $\bullet$ Show computational performance \\
-             &                   & \hspace{5mm} at elevation $\sim 60$\\
-             &                   & $\bullet$ Full-sky imaging \\
- \hline
-  2013-05-16 & Vela and Puppis A & $\bullet$ Test cleaning of diffuse structure \\
-  MWA 128T   &                   & $\bullet$ Show computational performance \\
-             &                   & \hspace{5mm} of cleaning\\
-             &                   & $\bullet$ Pointed observations \\
- \hline
-  2013-07-31 & EoR0              & $\bullet$ Test cleaning of point sources \\
-  MWA 128T   &                   & $\bullet$ Test deep imaging \\
- \hline
-  ---        & Simulations       & $\bullet$ Validate photometry / astrometry \\
-             &                   & $\bullet$ Test computational performance \\
-             &                   & \hspace{5mm} at very low elevation\\
- \hhline{|=|=|=|}
-\end{tabular}
-\end{table}
+For small field of views, the term $\sqrt{1-l^2-m^2}$ is approximately of unit size, transforming Eq.~\eqref{eq:visibility-function} into an ordinary 2-dimensional Fourier transform which can be inverted with an inverse Fast-Fourier Transform (FFT).
+A common rule is that this is valid when $\forall w,l,m: w\left(\sqrt{1-l^2-m^2}-1\right) \ll 1$.
+
+TODO: Algorithm; plain w-stacking and further alterations such as w-snapshotting and full-sky imaging.
+\section{The WSClean imager implementation}
+
+\section{Analysis}
+TODO: Show image quality; show time-efficiency comparisons.
+
+\bibliographystyle{mn2e}
+\bibliography{references}
 
@@
 
-\bibliographystyle{plainnat}
-\bibliography{references}
-
 \end{document}

### 2014-01-14 articles/2014/Article-WSClean/wsclean.tex
Lot's of work on wsclean article

@@
 
-\section*{Acknowledgments}
-...
-
-% To make Dutch ``tussenvoegsels'' work correctly in Latex such as ``de Bruyn'', we use this command
-% In bibliography, it should be written with lowercase, as in ``de Bruyn''
-\DeclareRobustCommand{\TUSSEN}[3]{#3}
-
 Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with simple imagers that are based on a two-dimensional Fast Fourier Transformation (FFT).
@@
 
-The sky equation for an interferometer is given by
-\begin{eqnarray}\label{eq:visibility-function}
-V(u,v,w) = \iint \frac{A(l,m) I(l,m)}{\sqrt{1-l^2-m^2}} e^{-2\pi i \left(ul + vm + w(\sqrt{1-l^2-m^2}-1)\right)} dl dm,
-\end{eqnarray}
-where function $V$ is formed by the calibrated visibilities, $u,v,w$ is a coordinate giving the observed mode in the coordinate system of the array, $A$ is the primary-beam function, $I$ is the sky function and $l,m$ are cosine sky coordinates.
-Imaging consists of inverting this equation to find $I$ from $V$.
+An interferometer samples the complex visibility function
+\begin{align}\notag
+V(u,v,w) = & \iint \frac{A(l,m) I(l,m)}{\sqrt{1-l^2-m^2}} \cdot \\ \label{eq:visibility-function}
+& e^{-2\pi i \left(ul + vm + w(\sqrt{1-l^2-m^2}-1)\right)} dl dm,
+\end{align}
+where $u,v,w$ is a baseline coordinate in the coordinate system of the array, $A$ is the primary-beam function, $I$ is the sky function and $l,m$ are cosine sky coordinates.
+We will use $I'(l,m)$ to denote the sky function before primary-beam correction, $I'(l,m)=A(l,m)I(l,m)$.
+We will not discuss calibration, and assume $V$ has been calibrated before imaging.
+In the case of a polarized measurement, the symbols $V$, $A$ and $I$ are $2\times 2$ Jones matrices, but without loss of generality we will ignore polarization and treat inversion as a scalar problem.
+Imaging consists of inverting Eq.~\eqref{eq:visibility-function}, i.e., to find $I'$ from $V$.
 
-For small field of views, the term $\sqrt{1-l^2-m^2}$ is approximately of unit size, transforming Eq.~\eqref{eq:visibility-function} into an ordinary 2-dimensional Fourier transform which can be inverted with an inverse Fast-Fourier Transform (FFT).
-A common rule is that this is valid when $\forall w,l,m: w\left(\sqrt{1-l^2-m^2}-1\right) \ll 1$.
+For small field of views, the term $\sqrt{1-l^2-m^2}$ is approximately of unit size, making Eq.~\eqref{eq:visibility-function} approximate an ordinary 2-dimensional Fourier transform, which can be inverted with an inverse Fast Fourier Transform (FFT).
+A common rule is that this is valid when
+\begin{equation}\label{eq:when-2d-is-valid}
+\forall w,l,m: w\left(\sqrt{1-l^2-m^2}-1\right) \ll 1.
+\end{equation}
+
+We will now mathematically derive the w-stacking technique. First Eq.~\eqref{eq:visibility-function} is rewritten to
+\begin{align}\notag
+V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \cdot \\ \notag
+& e^{-2\pi i \left(ul + vm\right)} dl dm.
+\end{align}
+This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
+% Intermediate step:
+%\begin{align}\notag
+%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \cdot \\ \notag
+%& e^{-2\pi i \left(ul + vm\right)} du dv,
+%\end{align}
+%which can be rewritten to
+\begin{align}\notag
+\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \cdot \\ \notag
+& e^{-2\pi i \left(ul + vm\right)} du dv.
+\end{align}
+Integrating both sides over $w_0$ to $w_e$, the minimum and maximum value of $w$, results in
+\begin{align}\notag
+\frac{I'(l,m)\left(w_e - w_0\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \cdot \\ \label{eq:wstacking}
+\iint V(u,v,w)  e^{-2\pi i \left(ul + vm\right)} du dv dw.
+\end{align}
+The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation.
+This shows that the sky function can be reconstructed by i)~taking all samples that correspond to a specific value of $w$; ii)~grid those samples on a uniform grid; iii)~calculating the inverse FFT; iv)~apply the direction dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; v)~repeat this for all $w$-values and add the results together; vi)~apply the final scaling.
+
+In practice, the final scaling will be different from $\left(w_e - w_0\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples.
+Therefore, the number of added samples and their weight have to be taken into account.
+Additionally, it might be required to divide out the effect of a possible convolution kernel and/or primary beam.
+
+\subsection{Discretization of w} \label{sec:gridding-w}
+While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of w-layers that need to be processed.
+For this, we can use a similar rule to \eqref{eq:when-2d-is-valid}, and make sure that for two subsequent discretized $w$-values, $w_A$ and $w_B$, the phase difference is less than one radian.
+This results in the constraint
+\begin{equation} \label{eq:minimum-w-distance}
+\left|\left(w_A - w_B\right) 2\pi (\sqrt{1-l^2-m^2}-1)\right| \ll 1.
+\end{equation}
+This suggest that a uniform discretization in $w$ is optimal TODO in contrast to some article?
+-- could be explained by having more samples at smaller $w$s.
+From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
+\begin{equation}
+ N_\textrm{wlayers} \gg 2\pi \left(w_e - w_0\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
+\end{equation}
+Actual values for the right-hand side can differ a lot.
+The value of $w_e - w_0$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image.
+For the MWA, a typical value of $w_e - w_0$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$.
+[TODO table of arrays with $w$ values at zenith/more angles/more freq?] For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$.
+This implies that, in the case of the MWA, tens of $w$-layers are required at zenith and hundreds at lower elevations.
+The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in section TODO.
 
 TODO: Algorithm; plain w-stacking and further alterations such as w-snapshotting and full-sky imaging.
+\subsection{Gridding}
+TODO
+
+\subsection{Time complexity of w-stacking}
+We will now derive the time complexity of w-stacking.
+The time complexity is expressed in several parameters.
+The number of samples $N_\textrm{samples}$ in the image influences the amount of samples that need to be gridded.
+Each sample belongs to exactly one $w$-layer, and thus each sample is gridded exactly once.
+For gridding, a sample is convolved with a convolution kernel to suppress FFT aliasing as well as gridding aliasing.
+If the kernel has a diameter of $N_\textrm{kernel}$, the amount of operations involved in gridding all samples scales with $N_\textrm{samples}N^2_\textrm{kernel}$.
+
+Due to memory constraints it is often necessary to perform multiple passes over the data set.
+The overhead caused by this is often negligable, but when both the number of passes and the number of samples is large, the algorithm will become dominated by seeking through the data set.
+Therefore, we add another term $(\eta+1) N_\textrm{samples}$, where $\eta$ defines the involved overhead.
+This overhead becomes higher when $N_\textrm{wlayers}$, $N_\textrm{imgsize}$ and $N_\textrm{samples}$ are higher or when the available memory is lower, but is $\sim 0$ in common MWA cases.
+
+For each $w$-layer, an inverse FFT operation is performed with a size equal to the image size.
+If we define $N_\textrm{imgsize}$ the size of the image along one axis, the number of operations involved in this scales with $N_\textrm{wlayers} N^2_\textrm{imgsize} \log N_\textrm{imgsize}$.
+After each inverse FFT, the direction dependent phase shift needs to be applied, for which a number of operations $\sim$ $N_\textrm{wlayers} N^2_\textrm{imgsize}$ is required, and the time spent on the inverse FFT normally dominates.
+Finally, the layers need to be added together and will be scaled, which adds a factor relative to $N_\textrm{samples}$ operations, which also is insignificant relative to the inverse FFTs.
+
+This implies that the total number of operations $N_\textrm{op}$ scales in the following way:
+\begin{align} \notag
+ N_\textrm{op} \sim & N_\textrm{wlayers} \left( N^2_\textrm{imgsize} \log N_\textrm{imgsize} + N^2_\textrm{imgsize} \right) + \\
+ & N^2_\textrm{imgsize} + N_\textrm{samples} \left( N^2_\textrm{kernel} + {\eta} + 1 \right).
+\end{align}
+If we ignore the possible reading overhead, and assume that the number of $w$-layers scales with the image size, a strict bound in Big-O notation is given by
+\begin{equation}
+\mathcal{O}\left(N^3_\textrm{imgsize} \log N_\textrm{imgsize} + N_\textrm{samples} N^2_\textrm{kernel}\right).
+\end{equation}
+
+We can compare this to $w$-projection algorithm \citep{wprojection-cornwell}, for which the total number of operations scales as
+\begin{align} \notag
+ N_\textrm{op-wproj} \sim & N^2_\textrm{imgsize} \log N_\textrm{imgsize} + N^2_\textrm{imgsize} + \\
+ & N_\textrm{samples} \left( N_\textrm{kernel} + N_\textrm{wkernel} \right)^2 + N_\textrm{wplanes},
+\end{align}
+where $N_\textrm{wkernel}$ is the size of the added kernel to correct the $w$-corrections in $uv$-space as a convolution, instead of as multiplication in $lm$-space.
+Each sample needs to be convolved with the combined kernel, hence the term $N_\textrm{samples} \left( N_\textrm{kernel} + N_\textrm{wkernel} \right)^2$.
+There is also time associated with calculating these kernels, and this can be substantial for the MWA.
+This scales with $N_\textrm{wplanes}$ the number of $w$-projection planes.
+The constraints on this number are equal to the constraints on $N_\textrm{wlayers}$ in the $w$-stacking algorithm, as described in \S\ref{sec:gridding-w}.
+The number of pixels in the $w$-kernel is depending on how quickly the kernel approaches zero.
+This kernel is a function in $u,v$, and hence the larger the angular size of the image, the smaller an individual $u,v$ pixel becomes, and the larger (in pixels) the kernel becomes.
+Hence, $N_\textrm{wkernel}$ scales with $N_\textrm{imgsize}$.
+The resulting bound for $w$-projection is:
+\begin{equation}
+\mathcal{O}\left(N^2_\textrm{imgsize} \log N_\textrm{imgsize} + N_\textrm{samples} N^2_\textrm{imgsize} + N_\textrm{planes} \right).
+\end{equation}
+By comparing the time complexity between $w$-stacking and $w$-projection, we can conclude that in the limiting behaviour, the $w$-stacking method will be faster when the gridding of the visibilities is the dominating cost of the algorithm.
+The $w$-projection algorithm will be faster when the inverse FFTs are dominating.
+The main parameter that changes this balance is the number of visibilities to be gridded, i.e., the size of the dataset.
+For larger datasets, $w$-stacking becomes more favourable.
+In section TODO we will determine which method is faster in practice for different parameters.
+
+\subsection{Cleaning a w-stacked image}
+TODO describe how to do
+
+It is tempting to use a very low number of $w$-layers, and define the major clean iteration threshold such that it will never clean any $w$-aliasing artefacts.
+However, predicting the visibilities from an image with a similar low number of $w$-layers will cause similar aliased errors, and thus will subtract incorrect fluxes from the visibilities.
+In the end, even though the sources causing the $w$-aliasing effects are removed, aliasing artefacts will still remain.
+
 \section{The WSClean imager implementation}
@@
 TODO: Show image quality; show time-efficiency comparisons.
+
+\section*{Acknowledgments}
+...
+
+% To make Dutch ``tussenvoegsels'' work correctly in Latex such as ``de Bruyn'', we use this command
+% In bibliography, it should be written with lowercase, as in ``de Bruyn''
+\DeclareRobustCommand{\TUSSEN}[3]{#3}
 

### 2014-01-18 articles/2014/Article-WSClean/wsclean.tex

@@
 Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane.
-This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes, that by nature are wide-field instruments.
+This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes, that by nature are wide-field instruments \citep{perley-noncoplanar-arrays}.
 
-The most commonly used method to image observations from non-coplanar arrays is the w-projection algorithm \citep{wprojection-cornwell} that has been implemented in CASA.
-This is the fastest generic wide-field imager available, and CASA is the only software package (known to us) that implements the w-projection algorithm.
-Other packages rely on the faceting algorithm \citep{facetting-cornwell}, which is about an order of magnitude slower and is less accurate \citep{wprojection-cornwell}.
-The w-projection algorithm has thus been successful so far.
-However, the available implementations are unable to cope with the data rates from the new generation of wide-field observatories, which are producing data sets that are orders of magnitude larger than before.
+There are several methods to deal with the w-term during imaging: facetting \citep{facetting-cornwell}, three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, w-projection \citep{wprojection-cornwell}, w-stacking\footnote{TODO when/where was this first mentioned in literature?} \citep{widefield-imaging-ska-cornwell} and w-snapshots \citep{widefield-imaging-ska-cornwell}.
+The w-snapshot algorithm is a hybrid between snapshot imaging and one of the other w-term correcting techniques, most efficiently w-projection or w-stacking.
+
+Probably the most commonly used software implementation to image observations from non-coplanar arrays is the w-projection algorithm \citep{wprojection-cornwell} that has been implemented in the Common Astronomy Software Applications (CASA, \citealt{casa}), because of its many features, high quality and free availability.
+New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before.
 Examples of such telescopes include the Murchison Widefield Array (MWA), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR).
-At the MWA, we have seen that it can take up to tens of hours with the w-projection algorithm to image a two-minute MWA observation when the observed field is at low elevation.
-Moreover, the w-projection implementation in CASA can not handle images of 10 k $\times$ 10 k, which is desired for making full-sky images at the native resolution of the MWA.
-This will become an even bigger problem when the Square-Kilometre Array (SKA) begins its operation, and hence a considerable improvement in imaging techniques is required.
+At the MWA we have seen that, when the observed field is at low elevation, it can take up to tens of hours with CASA's w-projection algorithm to image a two-minute MWA observation.
+The problem of wide-field imaging will become an even bigger challenge when the Square-Kilometre Array (SKA) begins its operation, and hence a considerable improvement in imaging techniques is required.
+In \citet{widefield-imaging-ska-cornwell}, it is shown that in theory the w-stacking algorithm is  the best approach for the SKA.
 
-We present a new implementation of a generic wide-field imager that significantly improves upon the w-projection implementation.
+In this article, we present a new implementation of a generic wide-field imager that significantly improves upon the w-projection implementation.
 Currently it performs approximately an order of magnitude faster in all situations and even more for very large images, while it maintains equal accuracy and can handle MWA full-sky imaging.
 The implementation uses the ``w-stacking method'' for correcting the w-terms \citep{widefield-imaging-ska-cornwell} to obtain the increase in speed.
-Performance can be improved further by combination with another technique: the MWA telescope splits observations in data products of typically a few minutes, and the w-terms can be kept small by phasing an observation to the orthogonal of the array's best-fitting plane.
+Performance can be improved further by combination with another techniques: the MWA telescope splits observations in data products of typically a few minutes, and the w-terms can be kept small by phasing an observation to the orthogonal of the array's best-fitting plane.
 This is somewhat similar to the w-snapshot algorithm proposed in \citet{widefield-imaging-ska-cornwell}.
@@
 
+TODO.
 After describing the new imaging implementation, we will test its performance using various MWA observations, including observations with varying elevation and fields (diffuse emission, point sources).
@@
 We will use $I'(l,m)$ to denote the sky function before primary-beam correction, $I'(l,m)=A(l,m)I(l,m)$.
-We will not discuss calibration, and assume $V$ has been calibrated before imaging.
+We will not discuss calibration, but assume $V$ has been calibrated before imaging.
 In the case of a polarized measurement, the symbols $V$, $A$ and $I$ are $2\times 2$ Jones matrices, but without loss of generality we will ignore polarization and treat inversion as a scalar problem.
@@
 
-We will now mathematically derive the w-stacking technique. First Eq.~\eqref{eq:visibility-function} is rewritten to
+To derive the w-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
 \begin{align}\notag
@@
 Additionally, it might be required to divide out the effect of a possible convolution kernel and/or primary beam.
+In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds.
+The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-values samples.
+In this case we can therefore calculate the image for $w<0$ from the image with $w>0$.
+That allows us to set $w_0$ to the minimum absolute $w$-value, which requires at least two times less layers.
+In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence is always performed as a complex-to-complex transform.
 
@@
 Therefore, we add another term $(\eta+1) N_\textrm{samples}$, where $\eta$ defines the involved overhead.
-This overhead becomes higher when $N_\textrm{wlayers}$, $N_\textrm{imgsize}$ and $N_\textrm{samples}$ are higher or when the available memory is lower, but is $\sim 0$ in common MWA cases.
+This overhead becomes higher when $N_\textrm{wlayers}$, $N_\textrm{pixels}$ and $N_\textrm{samples}$ are higher or when the available memory is lower, but is $\sim 0$ in common MWA cases.
 
 For each $w$-layer, an inverse FFT operation is performed with a size equal to the image size.
-If we define $N_\textrm{imgsize}$ the size of the image along one axis, the number of operations involved in this scales with $N_\textrm{wlayers} N^2_\textrm{imgsize} \log N_\textrm{imgsize}$.
-After each inverse FFT, the direction dependent phase shift needs to be applied, for which a number of operations $\sim$ $N_\textrm{wlayers} N^2_\textrm{imgsize}$ is required, and the time spent on the inverse FFT normally dominates.
-Finally, the layers need to be added together and will be scaled, which adds a factor relative to $N_\textrm{samples}$ operations, which also is insignificant relative to the inverse FFTs.
+If we define $N_\textrm{pixels}$ the number of pixels in the image along each dimension, the number of operations involved in the inverse FFTs scales with $N_\textrm{wlayers} N^2_\textrm{pixels} \log N_\textrm{pixels}$.
+After each inverse FFT, the direction dependent phase shift needs to be applied, for which a number of operations $\sim$ $N_\textrm{wlayers} N^2_\textrm{pixels}$ is required, and the time spent on the inverse FFT normally dominates.
+Finally, the layers need to be added together and will be scaled, which adds a factor relative to $N_\textrm{pixels}$ operations, which also is insignificant relative to the inverse FFTs.
 
@@
 \begin{align} \notag
- N_\textrm{op} \sim & N_\textrm{wlayers} \left( N^2_\textrm{imgsize} \log N_\textrm{imgsize} + N^2_\textrm{imgsize} \right) + \\
- & N^2_\textrm{imgsize} + N_\textrm{samples} \left( N^2_\textrm{kernel} + {\eta} + 1 \right).
+ N_\textrm{op} \sim & N_\textrm{wlayers} \left( N^2_\textrm{pixels} \log N_\textrm{pixels} + N^2_\textrm{pixels} \right) + \\
+ & N^2_\textrm{pixels} + N_\textrm{samples} \left( N^2_\textrm{kernel} + {\eta} + 1 \right).
 \end{align}
@@
 \begin{equation}
-\mathcal{O}\left(N^3_\textrm{imgsize} \log N_\textrm{imgsize} + N_\textrm{samples} N^2_\textrm{kernel}\right).
+\mathcal{O}\left(N^3_\textrm{pixels} \log N_\textrm{pixels} + N_\textrm{samples} N^2_\textrm{kernel}\right).
 \end{equation}
@@
 \begin{align} \notag
- N_\textrm{op-wproj} \sim & N^2_\textrm{imgsize} \log N_\textrm{imgsize} + N^2_\textrm{imgsize} + \\
+ N_\textrm{op-wproj} \sim & N^2_\textrm{pixels} \log N_\textrm{pixels} + N^2_\textrm{pixels} + \\
  & N_\textrm{samples} \left( N_\textrm{kernel} + N_\textrm{wkernel} \right)^2 + N_\textrm{wplanes},
@@
 There is also time associated with calculating these kernels, and this can be substantial for the MWA.
-This scales with $N_\textrm{wplanes}$ the number of $w$-projection planes.
+This scales with $N_\textrm{wplanes}$, the number of $w$-projection planes.
 The constraints on this number are equal to the constraints on $N_\textrm{wlayers}$ in the $w$-stacking algorithm, as described in \S\ref{sec:gridding-w}.
@@
 This kernel is a function in $u,v$, and hence the larger the angular size of the image, the smaller an individual $u,v$ pixel becomes, and the larger (in pixels) the kernel becomes.
-Hence, $N_\textrm{wkernel}$ scales with $N_\textrm{imgsize}$.
+Hence, $N_\textrm{wkernel}$ scales with $N_\textrm{pixels}$.
 The resulting bound for $w$-projection is:
 \begin{equation}
-\mathcal{O}\left(N^2_\textrm{imgsize} \log N_\textrm{imgsize} + N_\textrm{samples} N^2_\textrm{imgsize} + N_\textrm{planes} \right).
+\mathcal{O}\left(N^2_\textrm{pixels} \log N_\textrm{pixels} + N_\textrm{samples} N^2_\textrm{pixels} + N_\textrm{planes} \right).
 \end{equation}
@@
 
+TODO: discuss low w-values means less-direction-invariant PSF.
+
 \section{The WSClean imager implementation}
+We have implemented the $w$-stacking algorithm in the C++ language.
+Several steps are multi-threaded in this implementation: reading and writing, gridding from and to different $w$-layers, performing the Fast Fourier transforms and peak-finding and image subtraction during Cleaning.
+In addition, intrinsics were used for the latter.
+This speeds up a minor Cleaning iteration,
 
@@
 TODO: Show image quality; show time-efficiency comparisons.
+
+\section{Discussion}
+TODO: describe hybrid between w-stacking, w-projection and w-snapshots (if not yet discussed in Cornwell's papers)
 

### 2014-01-19 articles/2014/Article-WSClean/wsclean.tex
Introduction, layout

@@
 Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with simple imagers that are based on a two-dimensional Fast Fourier Transformation (FFT).
-Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane.
-This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes, that by nature are wide-field instruments \citep{perley-noncoplanar-arrays}.
+Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}.
+This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes, that by nature are wide-field instruments.
 
-There are several methods to deal with the w-term during imaging: facetting \citep{facetting-cornwell}, three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, w-projection \citep{wprojection-cornwell}, w-stacking\footnote{TODO when/where was this first mentioned in literature?} \citep{widefield-imaging-ska-cornwell} and w-snapshots \citep{widefield-imaging-ska-cornwell}.
-The w-snapshot algorithm is a hybrid between snapshot imaging and one of the other w-term correcting techniques, most efficiently w-projection or w-stacking.
+There are several methods to deal with the w-term during imaging: facetting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, w-projection \citep{wprojection-cornwell}, w-stacking\footnote{TODO when/where was this first mentioned in literature?} \citep{widefield-imaging-ska-cornwell} and w-snapshots \citep{widefield-imaging-ska-cornwell}.
+The w-snapshots algorithm is a hybrid between snapshot imaging and one of the other w-term correcting techniques, most efficiently w-projection or w-stacking.
 
-Probably the most commonly used software implementation to image observations from non-coplanar arrays is the w-projection algorithm \citep{wprojection-cornwell} that has been implemented in the Common Astronomy Software Applications (CASA, \citealt{casa}), because of its many features, high quality and free availability.
 New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before.
-Examples of such telescopes include the Murchison Widefield Array (MWA), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR).
-At the MWA we have seen that, when the observed field is at low elevation, it can take up to tens of hours with CASA's w-projection algorithm to image a two-minute MWA observation.
-The problem of wide-field imaging will become an even bigger challenge when the Square-Kilometre Array (SKA) begins its operation, and hence a considerable improvement in imaging techniques is required.
-In \citet{widefield-imaging-ska-cornwell}, it is shown that in theory the w-stacking algorithm is  the best approach for the SKA.
+Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}).
+At the MWA, some of the data is reduced with the w-projection algorithm implementation of the Common Astronomy Software Applications (CASA, \citealt{casa}).
+This is an attractive option because of its many features, high quality and free availability.
+However, at the MWA we have seen that a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's w-projection algorithm.
+Another option for reducing MWA data is the Real-Time System (RTS), which has been designed as an efficient calibrating and imaging pipeline specifically for MWA data \citep{rts-mwa}.
+It can use GPUs for improving efficiency, and uses a w-projection kernel during imaging to correct for the w-term.
+The kernel can be kept small, because the data is processed in snapshots that have been phase-centered to Zenith.
+The RTS can not perform deconvolution as of yet.
 
-In this article, we present a new implementation of a generic wide-field imager that significantly improves upon the w-projection implementation.
-Currently it performs approximately an order of magnitude faster in all situations and even more for very large images, while it maintains equal accuracy and can handle MWA full-sky imaging.
-The implementation uses the ``w-stacking method'' for correcting the w-terms \citep{widefield-imaging-ska-cornwell} to obtain the increase in speed.
-Performance can be improved further by combination with another techniques: the MWA telescope splits observations in data products of typically a few minutes, and the w-terms can be kept small by phasing an observation to the orthogonal of the array's best-fitting plane.
-This is somewhat similar to the w-snapshot algorithm proposed in \citet{widefield-imaging-ska-cornwell}.
-Hence, we can describe the imager as a hybrid between w-stacking and a variation of the w-snapshot algorithm.
-It also implements the Clean deconvolution procedure (\citealt{hogbom-clean} and various variations).
+For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure (TODO cite).
+This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data.
+The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, w-projection and w-stacking \citep{awimager-2013}.
+
+Once the Square-Kilometre Array (SKA) begins its operation, the required computational power for wide-field imaging will become an even bigger challenge.
+In \citet{widefield-imaging-ska-cornwell}, it is shown that in theory the w-snapshots algorithm is an efficient approach for the SKA.
+
+In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's w-projection implementation.
+To obtain the increase in speed, the implementation uses the w-stacking method for correcting the w-terms \citep{widefield-imaging-ska-cornwell} combined with snapshot imaging.
 We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''.
+This works improves the field of wide-field imaging in the following ways:
+\begin{itemize}
+ \item The performance of a w-stacking implementation is compared to a w-projection implementation.
+ \item The advantages of using the w-snapshots algorithm are analysed.
+ \item Details are described of how to efficiently implement the w-stacking and w-snapshots algorithms.
+ \item A new implementation is released, which we believe to be the fastest generic widefield imager available for many imaging scenarios.
+\end{itemize}
 
-TODO.
-After describing the new imaging implementation, we will test its performance using various MWA observations, including observations with varying elevation and fields (diffuse emission, point sources).
-GMRT data and -- if available -- LOFAR data will also be used.
-We will compare these results with the results from the w-projection implementation in CASA and to other non-w-projection algorithms, thereby focusing on accuracy and performance.
+The w-stacking algorithm is described in Sect.~\ref{sec:wstacking}.
+Details of implementing the w-stacking and w-snapshots algorithms are described in Sect.~\ref{sec:implementation}.
+Using various MWA observations, including observations with varying elevation, the performance will be analysed in Sect.~\ref{sec:analysis}.
 
-\section{The w-stacking technique}
+\section{The w-stacking technique} \label{sec:wstacking}
 In this section, we will describe the w-stacking algorithm from a mathematical point of view.
@@
 Therefore, the number of added samples and their weight have to be taken into account.
-Additionally, it might be required to divide out the effect of a possible convolution kernel and/or primary beam.
+Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction dependent effects.
 In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds.
@@
 
-TODO: Algorithm; plain w-stacking and further alterations such as w-snapshotting and full-sky imaging.
+TODO: Algorithm; plain w-stacking and further alterations such as w-snapshots and full-sky imaging.
 \subsection{Gridding}
@@
 TODO describe how to do
+TODO \citet{hogbom-clean}
 
@@
 
-\section{The WSClean imager implementation}
+\subsection{W-snapshots}
+TODO discuss projecting visibilities onto different tangent plane, imaging by small change to algorithm, output standard fits file by using CPIXij keyword \citet{wcs-in-fits}.
+, and the w-terms can be kept small by phasing a snapshot to the orthogonal of the array's best-fitting plane
+
+\section{The WSClean imager implementation} \label{sec:implementation}
 We have implemented the $w$-stacking algorithm in the C++ language.
@@
 
-\section{Analysis}
+\subsection{Application of WSClean to the MWA}
+The MWA correlator splits observations in data products of typically a few minutes
+TODO
+
+\section{Analysis} \label{sec:analysis}
 TODO: Show image quality; show time-efficiency comparisons.
 
-\section{Discussion}
+\section{Discussion} \label{sec:discussion}
 TODO: describe hybrid between w-stacking, w-projection and w-snapshots (if not yet discussed in Cornwell's papers)
+
+TODO: A-projection makes the psf more invariant: therefore, truely ``joined'' deconvolution might be more accurate.
 

### 2014-01-21 articles/2014/Article-WSClean/wsclean.tex
Minor changes
Adding images
Work on article

@@
 We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''.
-This works improves the field of wide-field imaging in the following ways:
+This work improves the field of wide-field imaging in the following ways:
 \begin{itemize}
@@
 While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of w-layers that need to be processed.
-For this, we can use a similar rule to \eqref{eq:when-2d-is-valid}, and make sure that for two subsequent discretized $w$-values, $w_A$ and $w_B$, the phase difference is less than one radian.
+For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian.
 This results in the constraint
@@
 TODO: Algorithm; plain w-stacking and further alterations such as w-snapshots and full-sky imaging.
-\subsection{Gridding}
-TODO
 
@@
 In addition, intrinsics were used for the latter.
-This speeds up a minor Cleaning iteration,
+This speeds up a minor Cleaning iterations,
 
@@
 TODO: Show image quality; show time-efficiency comparisons.
+
+\subsection{Gridding}
+During gridding of the visibilities in $uv$-space, it is useful to apply a gridding convolution kernel.
+A common kernel function is a windowed sinc function.
+A low-pass filter decreases the flux of sources outside the field of view, and thus helps to attenuate aliased ghost sources and sidelobes (see \citealt{post-correlation-filtering}) that arise because of the assumed periodicity during the inverse FFT.
+By supersampling, a convolution kernel also makes it possible to place samples more accurately at their $uv$ position, thereby lowering decorrelation.
+
+The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding (TODO cite).
+CASA convolves samples with a PSWF of 7 pixels total width during gridding.
+When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate\footnote{CASA uses pre-calculation tables for its fixed-size window function}.
+WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and its properties are similar to a PSWF.
+The type of window function has no effect on the gridding performance, but gridding has a quadratic time dependency on the size of the kernel.
+
+\begin{table}
+ \caption{Parameter values during benchmarks (unless the effect of the parameter was tested).}
+ \begin{tabular}{lr}
+   \textbf{Parameter} & \textbf{Value} \\
+   Image size             & 3072 \\
+   Number of visibilities & $10^9$
+ \end{tabular}
+\end{table}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{img/benchmark-zenith-angle/za}
+\caption{...}
+\label{fig:timing-zenith-angle}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{img/benchmark-nsamples/nsamples}
+\caption{...}
+\label{fig:timing-nsamples}
+\end{center}
+\end{figure}
 

### 2014-01-22 articles/2014/Article-WSClean/wsclean.tex
Made time complexity more concise, added kernel size discussion

@@
 In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's w-projection implementation.
-To obtain the increase in speed, the implementation uses the w-stacking method for correcting the w-terms \citep{widefield-imaging-ska-cornwell} combined with snapshot imaging.
+To obtain the increase in speed, the implementation uses the w-stacking method for correcting the w-terms combined with snapshot imaging.
 We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''.
@@
 From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
-\begin{equation}
+\begin{equation} \label{eq:nwlayers-bound}
  N_\textrm{wlayers} \gg 2\pi \left(w_e - w_0\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
@@
 \subsection{Time complexity of w-stacking}
-We will now derive the time complexity of w-stacking.
-The time complexity is expressed in several parameters.
-The number of samples $N_\textrm{samples}$ in the image influences the amount of samples that need to be gridded.
-Each sample belongs to exactly one $w$-layer, and thus each sample is gridded exactly once.
-For gridding, a sample is convolved with a convolution kernel to suppress FFT aliasing as well as gridding aliasing.
-If the kernel has a diameter of $N_\textrm{kernel}$, the amount of operations involved in gridding all samples scales with $N_\textrm{samples}N^2_\textrm{kernel}$.
+It is useful to analyse the time complexity of w-stacking and compare it with w-projection, to understand which algorithm performs better in a given situation.
+We will use the following symbols: $N_\textrm{wlay}$ is the number of w-layers for w-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each dimension, $N_\textrm{vis}$ is the number of visibilities and $N_\textrm{kern}$ is the size of the antialiasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the w-kernel for w-projection, $w_e$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.
 
-Due to memory constraints it is often necessary to perform multiple passes over the data set.
-The overhead caused by this is often negligable, but when both the number of passes and the number of samples is large, the algorithm will become dominated by seeking through the data set.
-Therefore, we add another term $(\eta+1) N_\textrm{samples}$, where $\eta$ defines the involved overhead.
-This overhead becomes higher when $N_\textrm{wlayers}$, $N_\textrm{pixels}$ and $N_\textrm{samples}$ are higher or when the available memory is lower, but is $\sim 0$ in common MWA cases.
+\begin{table}[h]
+ \caption{Scaling of the computational cost for various operations} \label{tbl:computational-cost-per-operation}
+ \begin{tabular}{lll}
+   \textbf{Operation} & \textbf{W-stacking} & \textbf{W-projection} \\
+   Fourier transform(s) & $N_\textrm{wlay} N^2_\textrm{pix} \log N_\textrm{pix}$ & $N^2_\textrm{pix} \log N_\textrm{pix}$ \\
+   W-term corrections   & $N_\textrm{wlay} N^2_\textrm{pix}$ & $N_\textrm{vis} N^2_\textrm{wkern}$ \\
+   Gridding & $N_\textrm{vis}N^2_\textrm{kern}$ & $N_\textrm{vis} N_\textrm{kern}^2$
+ \end{tabular}
+\end{table}
 
-For each $w$-layer, an inverse FFT operation is performed with a size equal to the image size.
-If we define $N_\textrm{pixels}$ the number of pixels in the image along each dimension, the number of operations involved in the inverse FFTs scales with $N_\textrm{wlayers} N^2_\textrm{pixels} \log N_\textrm{pixels}$.
-After each inverse FFT, the direction dependent phase shift needs to be applied, for which a number of operations $\sim$ $N_\textrm{wlayers} N^2_\textrm{pixels}$ is required, and the time spent on the inverse FFT normally dominates.
-Finally, the layers need to be added together and will be scaled, which adds a factor relative to $N_\textrm{pixels}$ operations, which also is insignificant relative to the inverse FFTs.
-
-This implies that the total number of operations $N_\textrm{op}$ scales in the following way:
-\begin{align} \notag
- N_\textrm{op} \sim & N_\textrm{wlayers} \left( N^2_\textrm{pixels} \log N_\textrm{pixels} + N^2_\textrm{pixels} \right) + \\
- & N^2_\textrm{pixels} + N_\textrm{samples} \left( N^2_\textrm{kernel} + {\eta} + 1 \right).
-\end{align}
-If we ignore the possible reading overhead, and assume that the number of $w$-layers scales with the image size, a strict bound in Big-O notation is given by
+Table~\ref{tbl:computational-cost-per-operation} shows how the computational costs scale for the operations that dominate the imaging in the w-stacking and w-projection algorithms.
+For comparison, we can assume the gridding kernel adds a constant term and the terms $N_\textrm{wlay}$ and $N_\textrm{wkern}$ follow approximately $w_e \sin \alpha_\textrm{FOV}$.
+The time complexity for w-stacking is then given by
 \begin{equation}
-\mathcal{O}\left(N^3_\textrm{pixels} \log N_\textrm{pixels} + N_\textrm{samples} N^2_\textrm{kernel}\right).
+\textrm{TC}_\textrm{wstacking}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} \sin \alpha_\textrm{FOV}+ N_\textrm{vis} \right),
 \end{equation}
-
-We can compare this to $w$-projection algorithm \citep{wprojection-cornwell}, for which the total number of operations scales as
-\begin{align} \notag
- N_\textrm{op-wproj} \sim & N^2_\textrm{pixels} \log N_\textrm{pixels} + N^2_\textrm{pixels} + \\
- & N_\textrm{samples} \left( N_\textrm{kernel} + N_\textrm{wkernel} \right)^2 + N_\textrm{wplanes},
-\end{align}
-where $N_\textrm{wkernel}$ is the size of the added kernel to correct the $w$-corrections in $uv$-space as a convolution, instead of as multiplication in $lm$-space.
-Each sample needs to be convolved with the combined kernel, hence the term $N_\textrm{samples} \left( N_\textrm{kernel} + N_\textrm{wkernel} \right)^2$.
-There is also time associated with calculating these kernels, and this can be substantial for the MWA.
-This scales with $N_\textrm{wplanes}$, the number of $w$-projection planes.
-The constraints on this number are equal to the constraints on $N_\textrm{wlayers}$ in the $w$-stacking algorithm, as described in \S\ref{sec:gridding-w}.
-The number of pixels in the $w$-kernel is depending on how quickly the kernel approaches zero.
-This kernel is a function in $u,v$, and hence the larger the angular size of the image, the smaller an individual $u,v$ pixel becomes, and the larger (in pixels) the kernel becomes.
-Hence, $N_\textrm{wkernel}$ scales with $N_\textrm{pixels}$.
-The resulting bound for $w$-projection is:
+and for $w$-projection it is
 \begin{equation}
-\mathcal{O}\left(N^2_\textrm{pixels} \log N_\textrm{pixels} + N_\textrm{samples} N^2_\textrm{pixels} + N_\textrm{planes} \right).
+\textrm{TC}_\textrm{wprojection}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} + N_\textrm{vis} w_e^2 \sin^2 \alpha_\textrm{FOV}\right).
 \end{equation}
-By comparing the time complexity between $w$-stacking and $w$-projection, we can conclude that in the limiting behaviour, the $w$-stacking method will be faster when the gridding of the visibilities is the dominating cost of the algorithm.
+From these bounds it can be concluded that in the limiting behaviour, the $w$-stacking method will be faster when the gridding of the visibilities is the dominating cost of the algorithm.
 The $w$-projection algorithm will be faster when the inverse FFTs are dominating.
-The main parameter that changes this balance is the number of visibilities to be gridded, i.e., the size of the dataset.
-For larger datasets, $w$-stacking becomes more favourable.
-In section TODO we will determine which method is faster in practice for different parameters.
+In Sect.~\ref{sec:analysis} we will determine which method is faster in practice for different parameters.
 
@@
 
+TODO discuss dealing with memory.
+
 \subsection{Application of WSClean to the MWA}
@@
 
-\subsection{Gridding}
+\subsection{Gridding} \label{sec:gridding}
 During gridding of the visibilities in $uv$-space, it is useful to apply a gridding convolution kernel.
@@
 CASA convolves samples with a PSWF of 7 pixels total width during gridding.
-When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate\footnote{CASA uses pre-calculation tables for its fixed-size window function}.
+When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate.
 WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and its properties are similar to a PSWF.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
+\caption{...}
+\label{fig:timing-kernelsize}
+\end{center}
+\end{figure}
+
 The type of window function has no effect on the gridding performance, but gridding has a quadratic time dependency on the size of the kernel.
+Fig.~\ref{fig:timing-kernelsize} shows that this quadratic dependency becomes apparent for kernel sizes $\ge 15$.
+Decreasing the kernel size to values below $7$ has little effect, because reading the samples takes more time than gridding with small kernels.
+We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged field of view.
+Therefore, WSClean uses a default of 7 pixels for the gridding kernel.
+It can be increased when necessary.
 

### 2014-01-22 articles/2014/Article-WSClean/wsclean.tex
Progress

@@
 
-TODO: Algorithm; plain w-stacking and further alterations such as w-snapshots and full-sky imaging.
-
 \subsection{Time complexity of w-stacking}
@@
 \begin{table}[h]
- \caption{Scaling of the computational cost for various operations} \label{tbl:computational-cost-per-operation}
+ \caption{Scaling of the computational cost for various imaging steps} \label{tbl:computational-cost-per-operation}
  \begin{tabular}{lll}
    \textbf{Operation} & \textbf{W-stacking} & \textbf{W-projection} \\
+   \hline\hline
    Fourier transform(s) & $N_\textrm{wlay} N^2_\textrm{pix} \log N_\textrm{pix}$ & $N^2_\textrm{pix} \log N_\textrm{pix}$ \\
    W-term corrections   & $N_\textrm{wlay} N^2_\textrm{pix}$ & $N_\textrm{vis} N^2_\textrm{wkern}$ \\
-   Gridding & $N_\textrm{vis}N^2_\textrm{kern}$ & $N_\textrm{vis} N_\textrm{kern}^2$
+   Gridding & $N_\textrm{vis}N^2_\textrm{kern}$ & $N_\textrm{vis} N_\textrm{kern}^2$ \\
+   \hline\hline
  \end{tabular}
@@
 \begin{equation}
-\textrm{TC}_\textrm{wstacking}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} \sin \alpha_\textrm{FOV}+ N_\textrm{vis} \right),
+\textrm{TC}_\textrm{wstacking}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} w_e \sin \alpha_\textrm{FOV}+ N_\textrm{vis} \right),
 \end{equation}
@@
 
-\subsection{Cleaning a w-stacked image}
-TODO describe how to do
-TODO \citet{hogbom-clean}
-
-It is tempting to use a very low number of $w$-layers, and define the major clean iteration threshold such that it will never clean any $w$-aliasing artefacts.
-However, predicting the visibilities from an image with a similar low number of $w$-layers will cause similar aliased errors, and thus will subtract incorrect fluxes from the visibilities.
-In the end, even though the sources causing the $w$-aliasing effects are removed, aliasing artefacts will still remain.
-
-TODO: discuss low w-values means less-direction-invariant PSF.
+\subsection{Full-sky imaging}
+For low-frequency telescopes, it can be very useful to do full-sky imaging.
+In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90 degrees from the pointing centre.
+Imaging these sidelobes can be useful because there might be science (e.g.
+when searching for transients), or for self-calibration or full-sky deconvolution.
+The latter is for example useful when a strong source in a sidelobe affects the sensitivity in the pointing centre.
 
@@
 Several steps are multi-threaded in this implementation: reading and writing, gridding from and to different $w$-layers, performing the Fast Fourier transforms and peak-finding and image subtraction during Cleaning.
-In addition, intrinsics were used for the latter.
+For the latter, intrinsics were used as well.
 This speeds up a minor Cleaning iterations,
@@
 
+\subsection{Cleaning a w-stacked image}
+TODO describe how to do
+TODO \citet{hogbom-clean}
+
+It is tempting to use a very low number of $w$-layers, and define the major clean iteration threshold such that it will never clean any $w$-aliasing artefacts.
+However, predicting the visibilities from an image with a similar low number of $w$-layers will cause similar aliased errors, and thus will subtract incorrect fluxes from the visibilities.
+In the end, even though the sources causing the $w$-aliasing effects are removed, aliasing artefacts will still remain.
+
+TODO: discuss low w-values means less-direction-invariant PSF.
+
 \begin{figure}

### 2014-01-23 articles/2014/Article-WSClean/wsclean.tex
Todays work on article

@@
 \frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \cdot \\ \notag
-& e^{-2\pi i \left(ul + vm\right)} du dv.
+& e^{2\pi i \left(ul + vm\right)} du dv.
 \end{align}
@@
 \frac{I'(l,m)\left(w_e - w_0\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \cdot \\ \label{eq:wstacking}
-\iint V(u,v,w)  e^{-2\pi i \left(ul + vm\right)} du dv dw.
+\iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
 \end{align}
@@
 In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence is always performed as a complex-to-complex transform.
+
+TODO: discuss visibility prediction? (inverse of imaging)
 
@@
 
+\section{The WSClean imager implementation} \label{sec:implementation}
+We have implemented the $w$-stacking algorithm in the C++ language.
+Several steps are multi-threaded in this implementation: reading and writing; gridding from and to different $w$-layers; performing the Fast Fourier transforms; and minor Clean iterations (peak-finding and image subtraction).
+For the latter, intrinsics were used as well.
+Because of these optimizations, minor Cleaning iterations are so fast that performing the Clark Clean optimization \citep{clark-clean} of considering only a subset of pixels with a reduced PSF, is not so relevant.
+However, because the PSF varies for imaging with a non-zero w-value, major iterations are still required to fix the errors made by subtracting a constant PSF during minor iterations.
+
+TODO discuss dealing with memory.
+
 \subsection{Full-sky imaging}
-For low-frequency telescopes, it can be very useful to do full-sky imaging.
+For low-frequency telescopes, it can be useful to do full-sky imaging.
 In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90 degrees from the pointing centre.
-Imaging these sidelobes can be useful because there might be science (e.g.
-when searching for transients), or for self-calibration or full-sky deconvolution.
-The latter is for example useful when a strong source in a sidelobe affects the sensitivity in the pointing centre.
+Imaging these sidelobes might be relevant because it might either be scientifically interesting (e.g.
+when searching for transients), or be useful for self-calibration or full-sky deconvolution.
+For example, the latter is useful when a strong source in a sidelobe affects the sensitivity near the pointing centre.
+
+To image the full sky, an observation is split into short snapshots, their phase centre is changed to zenith and subsequently a suitable area is imaged.
+Depending on desired resolution and dynamic range, this might require very large images.
+To make an image at the MWA resolution, the image needs to have approximately 10K pixels in width and height.
+The $w$-values will be very small at zenith, because only the differential heights of the antennas will contribute to $w$-value at zenith.
+For the MWA, the maximum height difference between tiles is $9.3$m.
+The MWA tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the $w$-value decreases to $7.2\textrm{m} / \lambda$.
+
+(TODO: these $w$ values are in 2 min obses; need shorter time interval for correct values).
 
 \subsection{W-snapshots}
-TODO discuss projecting visibilities onto different tangent plane, imaging by small change to algorithm, output standard fits file by using CPIXij keyword \citet{wcs-in-fits}.
-, and the w-terms can be kept small by phasing a snapshot to the orthogonal of the array's best-fitting plane
+To get optimal imaging performance, the $w$-values can be kept small by phase-changing towards zenith.
+As shown, all-sky imaging can be quite efficient with this technique.
+However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.
 
-\section{The WSClean imager implementation} \label{sec:implementation}
-We have implemented the $w$-stacking algorithm in the C++ language.
-Several steps are multi-threaded in this implementation: reading and writing, gridding from and to different $w$-layers, performing the Fast Fourier transforms and peak-finding and image subtraction during Cleaning.
-For the latter, intrinsics were used as well.
-This speeds up a minor Cleaning iterations,
+Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
+%When the Fourier transform is performed, the $w$ term has been corrected (either by w-stacking or by w-projection).
+Therefore, if $U(u, v)$ is the input to the 2-dimensional Fourier transform, the transform is described by
+\begin{align}\notag
+\frac{I'(\hat{l}, \hat{m})\left(w_e - w_0\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \cdot \\
+\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
+\end{align}
+In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging.
 
-TODO discuss dealing with memory.
+A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together.
+
+Another method to shift an image, is by using the periodicity of the FFT function.
+Since sources outside the field of view will be aliased back into the field, the image can be regridded with the alias in the centre.
+This complicates gridding, because the antialiasing kernel needs to have a pass-band shape instead of a low-pass shape.
+Therefore, we chose to implement the former method.
+
+TODO discuss output standard fits file by using CPIXij keyword \citet{wcs-in-fits}.
 

### 2014-01-24 articles/2014/Article-WSClean/wsclean.tex
Progress

@@
 For the latter, intrinsics were used as well.
-Because of these optimizations, minor Cleaning iterations are so fast that performing the Clark Clean optimization \citep{clark-clean} of considering only a subset of pixels with a reduced PSF, is not so relevant.
-However, because the PSF varies for imaging with a non-zero w-value, major iterations are still required to fix the errors made by subtracting a constant PSF during minor iterations.
+Because of these optimizations, minor Cleaning iterations are so fast that performing the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, is not so relevant.
+However, because the PSF varies for imaging with a non-zero w-value, major iterations are still required to fix the errors from using a constant PSF during minor iterations.
 
-TODO discuss dealing with memory.
+It will not be possible to store all $w$-layers in memory when creating large images or when many $w$-layers are used.
+For example, our 32-GB test machine can store 227 $w$-layers of $3072\times3072$ in memory.
+In more demanding imaging configurations, the implementation performs several passes over the measurement set and will grid a subset of $w$-layers in each pass.
 
@@
 \subsection{W-snapshots}
+\begin{figure*}
+\begin{center}
+\includegraphics[width=8cm]{img/vela-normal-projection}
+\includegraphics[width=8cm]{img/vela-zenith-projection}
+\caption{MWA imaging example of the Vela and Puppis A supernova remnants.
+Left: normal projection centred on Puppis A, right: zenith projection to reduce $w$-values, but recentred on Puppis A.
+TODO time difference and more discussion, replace with deeper images.}
+\label{fig:vela-projection-example}
+\end{center}
+\end{figure*}
 To get optimal imaging performance, the $w$-values can be kept small by phase-changing towards zenith.
@@
 A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together.
+An example of the difference in projection is given in Fig.~\ref{fig:vela-projection-example}.
+A recentred image can be stored in a Flexible Image Transport System (FITS) file, which allows setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}.
+Common viewers such as Kvis and DS9 support such FITS files and display their coordinates correctly.
+TODO show difference.
 
 Another method to shift an image, is by using the periodicity of the FFT function.
-Since sources outside the field of view will be aliased back into the field, the image can be regridded with the alias in the centre.
-This complicates gridding, because the antialiasing kernel needs to have a pass-band shape instead of a low-pass shape.
+Since sources outside the field of view will be aliased back into the field, the image can be transformed to have the alias in the centre.
+This complicates gridding during imaging, because the antialiasing kernel needs to have a pass-band shape instead of a low-pass shape.
 Therefore, we chose to implement the former method.
-
-TODO discuss output standard fits file by using CPIXij keyword \citet{wcs-in-fits}.
 
@@
 TODO
+
+TODO: mention full-polarization imaging
 

### 2014-01-27 articles/2014/Article-WSClean/wsclean.tex
Progress

@@
 \section{The WSClean imager implementation} \label{sec:implementation}
-We have implemented the $w$-stacking algorithm in the C++ language.
-Several steps are multi-threaded in this implementation: reading and writing; gridding from and to different $w$-layers; performing the Fast Fourier transforms; and minor Clean iterations (peak-finding and image subtraction).
-For the latter, intrinsics were used as well.
+With the purpose of testing new algorithms for the imaging of MWA data, we have written a new imager around the $w$-stacking algorithm.
+Nevertheless, the imager is not specialized for the MWA, and has been succesfully used for other telescopes (TODO menton more).
+
+The new imager, called ``WSClean'', is written in the C++ language.
+It reads visibilities from CASA measurement sets and writes output images to Flexible Image Transport System (FITS) files.
+Several steps are multi-threaded: reading and writing; gridding from and to different $w$-layers; performing the Fast Fourier transforms; and minor Clean iterations (peak-finding and image subtraction).
+For the latter, intrinsics are used as well.
 Because of these optimizations, minor Cleaning iterations are so fast that performing the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, is not so relevant.
-However, because the PSF varies for imaging with a non-zero w-value, major iterations are still required to fix the errors from using a constant PSF during minor iterations.
+However, the PSF varies for imaging with a non-zero w-value, and subtracting a constant PSF image space leads to inaccuracies.
+After a number of these minor iterations, it is therefore beneficial to invert the model back to the visibilities and subtract the model directly from the visibilities.
+This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}.
 
@@
 
-\subsection{W-snapshots}
+\subsection{W-snapshots} \label{sec:snapshot-imaging}
 \begin{figure*}
@@
 \includegraphics[width=8cm]{img/vela-zenith-projection}
-\caption{MWA imaging example of the Vela and Puppis A supernova remnants.
-Left: normal projection centred on Puppis A, right: zenith projection to reduce $w$-values, but recentred on Puppis A.
+\caption{13-min MWA observation of supernova remnants Vela and Puppis A.
+Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, shifted towards Puppis A during imaging.
+Both images have been made with WSClean using the Cotton-Schwab clean algorithm.
+No beam correction was applied.
 TODO time difference and more discussion, replace with deeper images.}
@@
 In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging.
+By doing the inverse corrections during visibility prediction, a ``shifted'' visibility set can be cleaned with major iterations similar to a non-shifted set.
 
 A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together.
-An example of the difference in projection is given in Fig.~\ref{fig:vela-projection-example}.
+An example of the difference in projection is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13~min MWA observation imaged with WSClean.
+Cleaning an image that is in a different projection yields slightly different results, but qualitativily the two images are of equal accuracy.
 A recentred image can be stored in a Flexible Image Transport System (FITS) file, which allows setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}.

### 2014-01-29 articles/2014/Article-WSClean/wsclean.tex
Progress

@@
 \usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
+\usepackage{xspace}
 
@@
 %\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
+\newcommand{\degree}{\ensuremath{^{\circ}}\xspace}
 
@@
 
-\title[WSClean: a new imager]{WSClean: a new implementation of a fast, generic wide-field imager}
+\title[WSClean: a fast, generic wide-field imager]{WSClean: an implementation of a fast, generic wide-field imager}
 
@@
 The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation.
-This shows that the sky function can be reconstructed by i)~taking all samples that correspond to a specific value of $w$; ii)~grid those samples on a uniform grid; iii)~calculating the inverse FFT; iv)~apply the direction dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; v)~repeat this for all $w$-values and add the results together; vi)~apply the final scaling.
+This shows that the sky function can be reconstructed by i)~grid samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~apply the direction dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeat this for all $w$-values and add the results together; v)~apply the final scaling.
 
@@
 
-TODO: discuss visibility prediction? (inverse of imaging)
+The inverse of imaging, i.e., to calculate the visibility from a model image, can be done by reversing the $w$-stacking algorithm: i)~inverse scale the image; ii)~copy the image to several layers; iii)~inverse apply the direction-dependent phaseshift for each layer; iv)~FFT each layer; and v)~sample a required visibility from the correct $w$-layer.
+We will refer to this operation as prediction.
 
@@
 Because of these optimizations, minor Cleaning iterations are so fast that performing the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, is not so relevant.
-However, the PSF varies for imaging with a non-zero w-value, and subtracting a constant PSF image space leads to inaccuracies.
-After a number of these minor iterations, it is therefore beneficial to invert the model back to the visibilities and subtract the model directly from the visibilities.
+However, the PSF varies for imaging with a non-zero w-value, and subtracting a constant PSF in image space leads to inaccuracies.
+After a number of these minor iterations, it is therefore beneficial to invert the model back to the visibilities and subtract the model directly from the visibilities \citep{wprojection-cornwell}.
 This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}.
@@
 For low-frequency telescopes, it can be useful to do full-sky imaging.
-In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90 degrees from the pointing centre.
+In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90\degree from the pointing centre.
 Imaging these sidelobes might be relevant because it might either be scientifically interesting (e.g.
@@
 The $w$-values will be very small at zenith, because only the differential heights of the antennas will contribute to $w$-value at zenith.
-For the MWA, the maximum height difference between tiles is $9.3$m.
-The MWA tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the $w$-value decreases to $7.2\textrm{m} / \lambda$.
-
-(TODO: these $w$ values are in 2 min obses; need shorter time interval for correct values).
+For the MWA, the maximum height difference between tiles is $8.5$m.
+The MWA tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the maximum $w$-value decreases to $5.5\textrm{m} / \lambda$.
+In the following sections, when discussing MWA zenith direction we will refer to this optimal $w$-direction.
 
@@
 Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, shifted towards Puppis A during imaging.
-Both images have been made with WSClean using the Cotton-Schwab clean algorithm.
+Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm.
 No beam correction was applied.
@@
 In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging.
-By doing the inverse corrections during visibility prediction, a ``shifted'' visibility set can be cleaned with major iterations similar to a non-shifted set.
+By doing the inverse corrections during visibility prediction, such a shifted visibility set can be cleaned with major iterations similar to a non-shifted set.
 
 A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together.
-An example of the difference in projection is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13~min MWA observation imaged with WSClean.
-Cleaning an image that is in a different projection yields slightly different results, but qualitativily the two images are of equal accuracy.
 A recentred image can be stored in a Flexible Image Transport System (FITS) file, which allows setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}.
 Common viewers such as Kvis and DS9 support such FITS files and display their coordinates correctly.
-TODO show difference.
+An example of the difference in projection is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13~min MWA observation imaged with WSClean.
+Cleaning an image that is in a different projection yields slightly different results, but qualitativily the two images are clearly of equal accuracy.
 
@@
 
-\subsection{Application of WSClean to the MWA}
-The MWA correlator splits observations in data products of typically a few minutes
-TODO
+\section{Analysis} \label{sec:analysis}
+We will now analyse the performance and accuracy of our implementation, and compare it with the w-projection implementation in CASA.
+For the analysis, we use imaging parameters which are common for MWA Imaging.
+When a specific parameter value is not mentioned, the settings from Table~\ref{tbl:default-parameters} are used.
+The number of $w$-projection planes in w-projection is kept equal to the number of $w$-layers in w-stacking, and is set to the right hand side of Eq.~\eqref{eq:nwlayers-bound}.
+This yields 97 $w$-projection planes or $w$-layers at 10\degree zenith angle.
+Some configurations with high $w$-values fail to image with CASA, presumably because the $w$-kernels become to large.
 
-TODO: mention full-polarization imaging
+TODO discuss CASA version
 
-\section{Analysis} \label{sec:analysis}
-TODO: Show image quality; show time-efficiency comparisons.
+TODO discuss desktop computer specs
+\begin{table}%
+\caption{Parameter values used during benchmarks, unless otherwise mentioned.} \label{tbl:default-parameters}%
+\begin{center}\begin{tabular}{rl}%
+\hline
+Array & MWA 128T \\
+Image size & $3072 \times 3072$ \\
+Angular resolution & $0.72'$ \\
+Number of visibilities & $3.5 \times 10^8$ \\
+Time resolution & 2~s \\
+Frequency resolution & 40~kHz \\
+Observation duration & 112 s\\
+Bandwidth & 30.72 MHz (768 channels)\\
+Central frequency & 182 MHz \\
+Zenith angle at phase centre & 10\degree \\
+Max $w$-value for phase centre & 172 $\lambda$ (283 m) \\
+Nr. polarizations in set & 4 \\
+Imaged polarization & XX \\
+Imaging mode & multi-frequency synthesis \\
+\hline
+\end{tabular}\end{center}\end{table}
 
+\subsection{Performance}
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{img/benchmark-nsamples/nsamples}
+\caption{Performance dependency on the number of imaged visibilities for imaging at zenith.
+Both methods are linearly dependent on the number of visibilities.
+WSClean is 2-3 times faster than CASA.
+Error bars show 10$\sigma$ level.}
+\label{fig:timing-nsamples}
+\end{center}
+\end{figure}
+
+In Fig.~\ref{fig:timing-nsamples} shows the time it takes to image a 2-min MWA snapshot at zenith with varying time integration.
+Time integration affects the number of visibilities to be gridded without changing the maximum $w$-value.
+Both methods have a linear time dependency on the number of visibilities.
+Because of the small $w$-values at zenith, the two methods are expected to be almost identical.
+However, the two methods show a factor of 2-3 difference, which might be attributed to different choices in optimizations.
+The linear dependency implies that gridding is dominating the cost.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{img/benchmark-zenith-angle/za}
+\caption{Effect of zenith angle (ZA) of the phase centre on performance.
+Image width and height are $2048$ for the dashed line and $3072$ for the solid line.
+For the latter, CASA fails to image observations with ZA $\ge 15$.
+ZA dramatically affects the computational cost of w-projection, and WSClean is more than an order of magnitude faster at ZA $\ge$ 10\degree with a $3072$ image size (or ZA $\ge$ 30\degree at $1536$.) }
+\label{fig:timing-zenith-angle}
+\end{center}
+\end{figure}
+
+In Fig.~\ref{fig:timing-zenith-angle}, the computational cost as function of the zenith angle is plotted.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{img/benchmark-channels/channels}
+\caption{...}
+\label{fig:timing-channels}
+\end{center}
+\end{figure}
+
+\subsection{Accuracy}
+TODO show aliasing errors for too few wlayers.
 \subsection{Gridding} \label{sec:gridding}
@@
 
-\begin{table}
- \caption{Parameter values during benchmarks (unless the effect of the parameter was tested).}
- \begin{tabular}{lr}
-   \textbf{Parameter} & \textbf{Value} \\
-   Image size             & 3072 \\
-   Number of visibilities & $10^9$
- \end{tabular}
-\end{table}
-
 \subsection{Cleaning a w-stacked image}
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=8cm]{img/benchmark-zenith-angle/za}
-\caption{...}
-\label{fig:timing-zenith-angle}
-\end{center}
-\end{figure}
+\section{Application of WSClean to the MWA}
+The MWA correlator splits observations in data products of typically a few minutes
+TODO
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=8cm]{img/benchmark-nsamples/nsamples}
-\caption{...}
-\label{fig:timing-nsamples}
-\end{center}
-\end{figure}
+TODO: calculate when to use w-snapshots
+
+TODO: mention full-polarization imaging
 

### 2014-01-29 articles/2014/Article-WSClean/wsclean.tex
Correction
Progress

@@
 The number of $w$-projection planes in w-projection is kept equal to the number of $w$-layers in w-stacking, and is set to the right hand side of Eq.~\eqref{eq:nwlayers-bound}.
-This yields 97 $w$-projection planes or $w$-layers at 10\degree zenith angle.
+This yields 195 $w$-projection planes or $w$-layers at 10\degree zenith angle.
 Some configurations with high $w$-values fail to image with CASA, presumably because the $w$-kernels become to large.
@@
 Imaging mode & multi-frequency synthesis \\
+TODO size \\
 \hline
@@
 \includegraphics[width=8cm]{img/benchmark-nsamples/nsamples}
-\caption{Performance dependency on the number of imaged visibilities for imaging at zenith.
-Both methods are linearly dependent on the number of visibilities.
-WSClean is 2-3 times faster than CASA.
-Error bars show 10$\sigma$ level.}
+\caption{Effect of data volume on number of samples for imaging at 10\degree zenith angle (solid lines) and at zenith (dashed lines).
+Error bars show 5$\sigma$ level.
+Both methods are linearly dependent on the number of visibilities for large sets.
+The right-most points represents 112 seconds of data at MWA's nominal resolution of 0.5~s / 40~kHz.
+}
 \label{fig:timing-nsamples}
@@
 
-In Fig.~\ref{fig:timing-nsamples} shows the time it takes to image a 2-min MWA snapshot at zenith with varying time integration.
-Time integration affects the number of visibilities to be gridded without changing the maximum $w$-value.
-Both methods have a linear time dependency on the number of visibilities.
-Because of the small $w$-values at zenith, the two methods are expected to be almost identical.
-However, the two methods show a factor of 2-3 difference, which might be attributed to different choices in optimizations.
-The linear dependency implies that gridding is dominating the cost.
+Fig.~\ref{fig:timing-nsamples} shows the wall-clock time to image a 112-s MWA snapshot as a function of the size of the visibility set.
+Results for imaging at 10\degree zenith angle (ZA) and at zenith are shown.
+The size of the visibility set was varied by changing the time resolution.
+This affects the number of visibilities to be gridded without changing the maximum $w$-value.
+For larger sets, both methods show a linear time dependency on the number of visibilities.
+This implies that gridding or reading dominates the cost.
+In that situation, the $w$-stacking implementation is 7.9 times faster at $\textrm{ZA}=10\degree$ and 2.9 times faster at $\textrm{ZA}=0\degree$.
+At $\textrm{ZA}=10$\degree, the performance of WSClean is approximately constant for sets with less than about $10^8$ visibilities, which corresponds with a single polarization of 16~s of MWA data in multi-frequency synthesis mode.
+At that point, the FFTs are the dominant cost in the algorithm.
+At zenith, the two methods are expected to perform almost identical.
+The factor of 2-3 difference in Fig.~\ref{fig:timing-nsamples} could be due to different choices in optimizations.
 

### 2014-01-29 articles/2014/Article-WSClean/wsclean.tex
Bla

@@
 \includegraphics[width=8cm]{img/benchmark-nsamples/nsamples}
-\caption{Effect of data volume on number of samples for imaging at 10\degree zenith angle (solid lines) and at zenith (dashed lines).
+\caption{Effect of data volume on imaging performance for imaging at 10\degree zenith angle (solid lines) and at zenith (dashed lines).
 Error bars show 5$\sigma$ level.
@@
 \includegraphics[width=8cm]{img/benchmark-zenith-angle/za}
-\caption{Effect of zenith angle (ZA) of the phase centre on performance.
+\caption{Effect of zenith angle (ZA) on imaging performance.
 Image width and height are $2048$ for the dashed line and $3072$ for the solid line.

### 2014-01-30 articles/2014/Article-WSClean/wsclean.tex
Todays work

@@
 This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data.
-The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, w-projection and w-stacking \citep{awimager-2013}.
+TODO: discuss A-Projection (CASA manual say: ``A-Projection is an algorithm to account for the effects of the antenna primary beam (PB) during imaging.
+The time-dependent effects of the PB are projected-out during the imaging phase and the PB is included in the prediction phase of the iterative image deconvolution (see Bhatnagar, Cornwell, Golap \& Uson 2008, A\&A, 487, 419'') The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, w-projection and w-stacking \citep{awimager-2013}.
 
@@
 Imaged polarization & XX \\
-Imaging mode & multi-frequency synthesis \\
-TODO size \\
+Imaging mode & Multi-frequency synthesis \\
+Weighting & uniform \\
+Data size & 18 GB \\
 \hline
@@
 
+\subsection{Accuracy}
+\begin{table}%
+\caption{Results on imaging accuracy measurements.} \label{tbl:accuracy-measurements}%
+\begin{center}\begin{tabular}{llll}%
+\hline\hline
+& \textbf{WSClean} & \textbf{WSClean}    & \textbf{CASA} \\
+&                  & \textbf{+ recentre} & \\
+\hline
+Source flux standard error & 1.31\% & & 1.34\% \\
+RMS in residual image & 0.94 mJy & & 1.90 mJy \\
+Computational time & 8 min & & forever (TODO) \\
+TODO at ZA=10\degree & & \\
+\hline\hline
+\end{tabular}\end{center}\end{table}
+
+\begin{figure*}
+\begin{center}
+\includegraphics[width=8cm]{img/residual-casa}
+\includegraphics[width=8cm]{img/residual-wsclean}
+\caption{Residual images after Cotton-Schwab cleaning of a simulated MWA observation, using similar inversion and cleaning parameters.
+Left image: w-projection using CASA; right image: w-stacking using WSClean.
+The image has been zoomed in just above the image centre, with the image centre being displayed in the bottom-centre of the panels.
+Sources in the image produced with CASA are less accurately subtracted, leading to a residual noise level of 1.9 mJy, whereas WSClean reaches 0.94 mJy.
+Since the effect is stronger away from the phase centre, we believe this is caused by the finite size of the w-kernels used in w-projection, which leads to inaccuracies.}
+\label{fig:residuals}
+\end{center}
+\end{figure*}
+
+We performed a few measurements to validate that there are no major accuracy difference between WSClean and CASA.
+To do so, we have simulated a MWA observation with hundred sources of 1~Jy in a 20 degrees area.
+A unitary primary beam is assumed.
+We image the set both with WSClean and CASA, using Cotton-Schwab cleaning to a threshold of 10~mJy.
+The two imagers calculate slightly different restoring (synthesized) beams, hence to avoid bias the restoring beams are fixed.
+Other imaging parameters are given in Table~\ref{tbl:default-parameters}.
+Aegean (TODO discuss) was subsequently used to perform source detection on the produced images.
+Even though no noise is added to the visibilities, sidelobe noise of the residual 10~mJy noises triggers a few false detections, which are ignored.
+
+Table~\ref{tbl:accuracy-measurements} lists the measured standard error of the source brightnesses as detected by Aegean, as well as the RMS in the residual image.
+
+TODO show aliasing errors for too few wlayers.
+
 \subsection{Performance}
+We measured the performance of the imagers using several MWA data sets.
+Each specific configuration is run 5 times and standard deviations are calculated over these 5 runs.
+The variations between runs is typically a few seconds.
+In each benchmark, the wall-clock time is measured that is required to produce the synthesized point-spread function and the image itself.
+No cleaning or prediction was performed.
+
 \begin{figure}
@@
 
-\subsection{Accuracy}
-TODO show aliasing errors for too few wlayers.
+TODO: for fairness, also check making 4 polarization images.
+
 \subsection{Gridding} \label{sec:gridding}

### 2014-01-30 articles/2014/Article-WSClean/wsclean.tex
Some progress

@@
 
+To reach high dynamic ranges, it can be necessary to deal with direction-dependent effects (DDEs).
+This is especially true for widefield telescopes.
+One way to handle DDEs is by using the a-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
+One such DDE is the effect of the ionosphere.
 For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure (TODO cite).
 This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data.
-TODO: discuss A-Projection (CASA manual say: ``A-Projection is an algorithm to account for the effects of the antenna primary beam (PB) during imaging.
-The time-dependent effects of the PB are projected-out during the imaging phase and the PB is included in the prediction phase of the iterative image deconvolution (see Bhatnagar, Cornwell, Golap \& Uson 2008, A\&A, 487, 419'') The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, w-projection and w-stacking \citep{awimager-2013}.
+The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, w-projection and w-stacking \citep{awimager-2013}.
 
 Once the Square-Kilometre Array (SKA) begins its operation, the required computational power for wide-field imaging will become an even bigger challenge.
-In \citet{widefield-imaging-ska-cornwell}, it is shown that in theory the w-snapshots algorithm is an efficient approach for the SKA.
+In \citet{widefield-imaging-ska-cornwell}, it is shown that in theory the w-snapshots algorithm is the most efficient approach for the SKA.
 
 In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's w-projection implementation.
-To obtain the increase in speed, the implementation uses the w-stacking method for correcting the w-terms combined with snapshot imaging.
+To obtain the increase in speed, the implementation uses the w-stacking method for correcting the w-terms, optionally combined with w-snapshot imaging.
 We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''.
@@
 \begin{itemize}
- \item The performance of a w-stacking implementation is compared to a w-projection implementation.
- \item The advantages of using the w-snapshots algorithm are analysed.
+ \item The performance of w-stacking is compared to w-projection.
+ \item Parameters that affect the w-snapshots algorithm are analysed.
  \item Details are described of how to efficiently implement the w-stacking and w-snapshots algorithms.
- \item A new implementation is released, which we believe to be the fastest generic widefield imager available for many imaging scenarios.
+ \item A new implementation is released, which we believe to be the fastest generic widefield imager available.
 \end{itemize}
@@
 Details of implementing the w-stacking and w-snapshots algorithms are described in Sect.~\ref{sec:implementation}.
-Using various MWA observations, including observations with varying elevation, the performance will be analysed in Sect.~\ref{sec:analysis}.
+The performance and accuracy will be analysed in Sect.~\ref{sec:analysis}.
 
@@
 \caption{13-min MWA observation of supernova remnants Vela and Puppis A.
-Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, shifted towards Puppis A during imaging.
+Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging.
 Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm.
@@
 In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging.
-By doing the inverse corrections during visibility prediction, such a shifted visibility set can be cleaned with major iterations similar to a non-shifted set.
+This is a method of implementing the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}.
+By doing the inverse corrections during visibility prediction, such a shifted visibility set can be cleaned with Cotton-Schwab iterations similar to a non-shifted set.
 
@@
 \hline
+\multicolumn{1}{l}{\textit{Zenith angle 0\degree}} \\
+\hline
 Source flux standard error & 1.31\% & & 1.34\% \\
 RMS in residual image & 0.94 mJy & & 1.90 mJy \\
-Computational time & 8 min & & forever (TODO) \\
-TODO at ZA=10\degree & & \\
+Computational time & 8 min & & 19.3 min \\
+\hline
+\multicolumn{1}{l}{\textit{Zenith angle 10\degree}} \\
+\hline
+Source flux standard error & 1.75\% & & 2.41 \%\\
+RMS in residual image & 0.88 mJy & & 1.04 mJy \\
+Computational time & 15.3 min & & 178.2 min \\
 \hline\hline

### 2014-01-31 articles/2014/Article-WSClean/wsclean.tex
Progress on article

@@
 \subsection{Discretization of w} \label{sec:gridding-w}
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{img/aliasing-example-annotated}
+\caption{Aliasing artefacts caused by insufficient $w$-layers.
+WSClean was set to use 12 $w$-layers.
+The centre of the image is at 10\degree zenith angle.
+This would normally require imaging with $\sim195$ $w$-layers.
+Original sources are 1~Jy, ghost sources are approximately 20\%.
+Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce ghost sources by themselves.
+}
+\label{fig:aliasing-example}
+\end{center}
+\end{figure}
 While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of w-layers that need to be processed.
@@
 This implies that, in the case of the MWA, tens of $w$-layers are required at zenith and hundreds at lower elevations.
-The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in section TODO.
+The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in the next sections.
+
+Using too few $w$-layers results in decorrelation of off-axis sources in the longer baselines.
+Additionally, $w$-aliasing errors can cause ghost sources in the image.
+An (extreme) example of this effect is shown in Fig.~\ref{fig:aliasing-example}.
+When Cotton-Schwab cleaning includes prediction with too few $w$-layers, incorrect values will be subtracted from the visibilities even when no aliased sources are cleaned during minor iterations.
+Therefore, accurate prediction is more important than accurate imaging, because aliasing artefacts are attenuated by cleaning as long as the model is subtracted accurately.
+Possible ways to allow less $w$-layers during imaging is by predicting with more $w$-layers or a direct Fourier transform.
+We have not looked into these options yet.
 
@@
 
-\begin{table}[h]
+\begin{table}
  \caption{Scaling of the computational cost for various imaging steps} \label{tbl:computational-cost-per-operation}
@@
 It reads visibilities from CASA measurement sets and writes output images to Flexible Image Transport System (FITS) files.
-Several steps are multi-threaded: reading and writing; gridding from and to different $w$-layers; performing the Fast Fourier transforms; and minor Clean iterations (peak-finding and image subtraction).
+Several steps are multi-threaded: reading and writing; gridding from and to different $w$-layers; performing the Fast Fourier transforms; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}).
 For the latter, intrinsics are used as well.
-Because of these optimizations, minor Cleaning iterations are so fast that performing the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, is not so relevant.
-However, the PSF varies for imaging with a non-zero w-value, and subtracting a constant PSF in image space leads to inaccuracies.
-After a number of these minor iterations, it is therefore beneficial to invert the model back to the visibilities and subtract the model directly from the visibilities \citep{wprojection-cornwell}.
+Because of these optimizations, minor Cleaning iterations are performed at a rate of hundreds per second at $3072\times3072$.
+This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, less relevant.
+However, the PSF varies for imaging with non-zero $w$-values, and subtracting a constant PSF in image space leads to inaccuracies.
+After a number of minor iterations, it is therefore beneficial to invert the model back to the visibilities via prediction, and subtract the model directly from the visibilities \citep{wprojection-cornwell}.
 This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}.
@@
 \subsection{Full-sky imaging}
-For low-frequency telescopes, it can be useful to do full-sky imaging.
+For low-frequency telescopes, it can be of interest to do full-sky imaging.
 In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90\degree from the pointing centre.
@@
 
-To image the full sky, an observation is split into short snapshots, their phase centre is changed to zenith and subsequently a suitable area is imaged.
+To efficiently image the full sky, an observation is split into short snapshots, their phase centre is changed to zenith and subsequently a suitable area is imaged.
 Depending on desired resolution and dynamic range, this might require very large images.
 To make an image at the MWA resolution, the image needs to have approximately 10K pixels in width and height.
-The $w$-values will be very small at zenith, because only the differential heights of the antennas will contribute to $w$-value at zenith.
+The $w$-values will be very small at zenith, because at zenith only the vertical offsets of the antennas will contribute to the $w$-value.
 For the MWA, the maximum height difference between tiles is $8.5$m.
 The MWA tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the maximum $w$-value decreases to $5.5\textrm{m} / \lambda$.
-In the following sections, when discussing MWA zenith direction we will refer to this optimal $w$-direction.
+In the following sections, when discussing the MWA zenith direction we are referring to this optimal $w$-direction.
 
@@
 \end{figure*}
-To get optimal imaging performance, the $w$-values can be kept small by phase-changing towards zenith.
-As shown, all-sky imaging can be quite efficient with this technique.
+As shown, all-sky imaging can be quite efficient with snapshot imaging towards zenith.
 However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.
@@
 In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging.
-This is a method of implementing the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}.
-By doing the inverse corrections during visibility prediction, such a shifted visibility set can be cleaned with Cotton-Schwab iterations similar to a non-shifted set.
+This essentially is a method of implementing the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}.
+By doing the inverse corrections during prediction, such a shifted visibility set can be cleaned with Cotton-Schwab iterations similar to a non-shifted set.
 
@@
 
-Another method to shift an image, is by using the periodicity of the FFT function.
+Another method to shift an image is by using the periodicity of the FFT function.
 Since sources outside the field of view will be aliased back into the field, the image can be transformed to have the alias in the centre.
@@
 The number of $w$-projection planes in w-projection is kept equal to the number of $w$-layers in w-stacking, and is set to the right hand side of Eq.~\eqref{eq:nwlayers-bound}.
-This yields 195 $w$-projection planes or $w$-layers at 10\degree zenith angle.
+This yields 195 $w$-planes/layers at 10\degree zenith angle.
 Some configurations with high $w$-values fail to image with CASA, presumably because the $w$-kernels become to large.
 
-TODO discuss CASA version
-
-TODO discuss desktop computer specs
+In all analyses CASA stable release 42.0, revision 26465 was used, which was released September 2013.
+It includes performance improvements to the \texttt{clean} task which have recently been made.
+The latest version as of February 2014 of WSClean was used.
+The tests were run on a high-end desktop, with a 3.20-GHz Intel Core i7-3930K processor and 32 GB of memory.
+The data are stored on a multi-disk array with 5 spinning hard disks, which has a combined read rate of about 450 MB/s.
 \begin{table}%
@@
 \caption{Results on imaging accuracy measurements.} \label{tbl:accuracy-measurements}%
-\begin{center}\begin{tabular}{llll}%
+\begin{center}\begin{tabular}{lrrr}%
 \hline\hline
@@
 \hline
-\multicolumn{1}{l}{\textit{Zenith angle 0\degree}} \\
+\multicolumn{3}{l}{\textit{Zenith angle 0\degree (minimum w-layers/planes)}} \\
 \hline
 Source flux standard error & 1.31\% & & 1.34\% \\
-RMS in residual image & 0.94 mJy & & 1.90 mJy \\
-Computational time & 8 min & & 19.3 min \\
+RMS in residual image & 0.94 mJy & --- & 1.90 mJy\\
+Computational time & 8.5 min & & 19.3 min\\
 \hline
-\multicolumn{1}{l}{\textit{Zenith angle 10\degree}} \\
+\multicolumn{3}{l}{\textit{Zenith angle 0\degree (128 w-layers/planes)}} \\
 \hline
-Source flux standard error & 1.75\% & & 2.41 \%\\
-RMS in residual image & 0.88 mJy & & 1.04 mJy \\
-Computational time & 15.3 min & & 178.2 min \\
+Source flux standard error & 1.39\% & & 2.08\% \\
+RMS in residual image & 0.94 mJy & --- & 0.94 mJy \\
+Computational time & 10.3 min & & 19.6 min \\
+\hline
+\multicolumn{3}{l}{\textit{Zenith angle 10\degree}} \\
+\hline
+Source flux standard error & 1.75\% & 2.92\% & 2.41 \%\\
+RMS in residual image & 0.90 mJy & 1.05 mJy & 1.07 mJy \\
+Computational time & 15.3 min & 6.6 min & 178.2 min \\
+& & recentreTODO & \\
 \hline\hline
@@
 \begin{center}
-\includegraphics[width=8cm]{img/residual-casa}
-\includegraphics[width=8cm]{img/residual-wsclean}
-\caption{Residual images after Cotton-Schwab cleaning of a simulated MWA observation, using similar inversion and cleaning parameters.
-Left image: w-projection using CASA; right image: w-stacking using WSClean.
-The image has been zoomed in just above the image centre, with the image centre being displayed in the bottom-centre of the panels.
-Sources in the image produced with CASA are less accurately subtracted, leading to a residual noise level of 1.9 mJy, whereas WSClean reaches 0.94 mJy.
-Since the effect is stronger away from the phase centre, we believe this is caused by the finite size of the w-kernels used in w-projection, which leads to inaccuracies.}
+\includegraphics[height=5cm]{img/residual-casa}\hspace{1cm}\includegraphics[height=5cm]{img/residual-wsclean}
+\caption{Residual images after Cotton-Schwab cleaning of a simulated 10\degree-zenith angle MWA observation using CASA (left) and WSClean (right), using similar inversion and cleaning parameters.
+The panels show a small part of the full images.
+The full field contains 100 simulated sources over 20\degree.
+Sources in the image produced with CASA are slightly less accurately subtracted, leading to a residual noise level of 1.07 mJy and some visible artefacts, whereas WSClean reaches 0.90 mJy.
+Since the effect is stronger away from the phase centre, it is likely that this is caused by the finite size of the w-kernels used in w-projection, which leads to inaccuracies.}
 \label{fig:residuals}
@@
 
-We performed a few measurements to validate that there are no major accuracy difference between WSClean and CASA.
-To do so, we have simulated a MWA observation with hundred sources of 1~Jy in a 20 degrees area.
+We performed measurements to assess the accuracy of WSClean and CASA.
+To do so, we have simulated a MWA observation without system noise, but with hundred sources of 1~Jy in a 20\degree diameter area.
 A unitary primary beam is assumed.
-We image the set both with WSClean and CASA, using Cotton-Schwab cleaning to a threshold of 10~mJy.
+We imaged the simulated set both with WSClean and CASA, using Cotton-Schwab cleaning to a threshold of 10~mJy.
 The two imagers calculate slightly different restoring (synthesized) beams, hence to avoid bias the restoring beams are fixed.
 Other imaging parameters are given in Table~\ref{tbl:default-parameters}.
-Aegean (TODO discuss) was subsequently used to perform source detection on the produced images.
-Even though no noise is added to the visibilities, sidelobe noise of the residual 10~mJy noises triggers a few false detections, which are ignored.
+The Aegean program \citep{aegean-hancock-2012} was used to perform source detection on the produced images.
+Sidelobe noise of the residual 10~mJy noise triggers a few false detections.
+These are ignored.
 
-Table~\ref{tbl:accuracy-measurements} lists the measured standard error of the source brightnesses as detected by Aegean, as well as the RMS in the residual image.
+Table~\ref{tbl:accuracy-measurements} lists the measured standard errors of the source brightnesses as detected by Aegean and the RMS in the residual image.
+In all tests, WSClean performs slightly more accurate.
+At zenith, CASA produces a residual RMS twice as high as WSClean.
+This is caused by the fact that we keep the number of w-projection planes in w-projection equal to the number of w-layers in w-stacking, resulting in only 12 $w$-planes at zenith.
+This evidently has a stronger effect on the w-projection algorithm.
+However, the computational performance of the w-projection algorithm is hardly affected by the number of w-projection planes, and in practical situations one would always use more w-projection planes\footnote{In the zenith case CASA actually issues a warning to use more $w$-projection planes.}.
+When 128 w-projection planes are used in CASA, its residual RMS is equal to WSClean with 12 $w$-layers.
+We do not know why this parameter needs to be higher in CASA to reach the same RMS.
+In theory it should be enough, since Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlayers}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree.
+Also unexpected is that the source flux density becomes worse by increasing the number of planes.
+For WSClean both values stay approximately the same when the number of $w$-layers is increased.
 
-TODO show aliasing errors for too few wlayers.
+In the $ZA=10\degree$ case, CASA is slightly less accurate.
+This is consistent, and can actually be observed in the images as shown in Fig.~\ref{fig:residuals}.
+TODO this might be caused by a bug; check: We have also made an image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}.
+Its accuracy is lower compared to the normal projection.
+This is due to the fact that the sources are further away from the phase centre, and sources have therefore on average higher $(l,m)$-values, thereby increasing the average $w$-correction.
+Nevertheless, the reprojection technique needs on average fewer $w$-layers to reach the same level of accuracy.
 
-\subsection{Performance}
+WSClean is faster in all tested cases.
+WSClean and CASA determine a different next major iteration threshold.
+CASA performs 5 major iterations at zenith and 8 at 10\degree angle, while WSClean always performs 5 major iterations.
+Major iterations are dominating the computing time.
+We will look more closely at the difference in performance in the next section.
+
+Many parameters affect the accuracy more than the observed differences between the methods.
+These include the number of major iterations, $w$-planes/layers, restoring beam, image size, weighting, kernel size and grid oversampling.
+Therefore, the differences are not very significant.
+The results do show that synthesis imaging using common imaging parameters results in an approximate flux-density standard error of at least $\sim1\%$.
+
+\subsection{Performance analysis}
 We measured the performance of the imagers using several MWA data sets.
@@
 Image width and height are $2048$ for the dashed line and $3072$ for the solid line.
-For the latter, CASA fails to image observations with ZA $\ge 15$.
-ZA dramatically affects the computational cost of w-projection, and WSClean is more than an order of magnitude faster at ZA $\ge$ 10\degree with a $3072$ image size (or ZA $\ge$ 30\degree at $1536$.) }
+CASA fails to image observations with ZA $\ge 15$ with a $3072$ image size.
+ZA dramatically affects the computational cost of w-projection.
+}
 \label{fig:timing-zenith-angle}
@@
 
-\subsection{Cleaning a w-stacked image}
-TODO describe how to do
-TODO \citet{hogbom-clean}
-
-It is tempting to use a very low number of $w$-layers, and define the major clean iteration threshold such that it will never clean any $w$-aliasing artefacts.
-However, predicting the visibilities from an image with a similar low number of $w$-layers will cause similar aliased errors, and thus will subtract incorrect fluxes from the visibilities.
-In the end, even though the sources causing the $w$-aliasing effects are removed, aliasing artefacts will still remain.
-
-TODO: discuss low w-values means less-direction-invariant PSF.
-
 \section{Application of WSClean to the MWA}
@@
 \section{Discussion} \label{sec:discussion}
+WSClean has been build out of necessacity to be able to make images with the MWA.
+
 TODO: describe hybrid between w-stacking, w-projection and w-snapshots (if not yet discussed in Cornwell's papers)
@@
 TODO: A-projection makes the psf more invariant: therefore, truely ``joined'' deconvolution might be more accurate.
+
+TODO: discuss low w-values means less-direction-invariant PSF.
 

### 2014-02-02 articles/2014/Article-WSClean/wsclean.tex
Work on article

@@
 This is an attractive option because of its many features, high quality and free availability.
-However, at the MWA we have seen that a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's w-projection algorithm.
+However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's w-projection algorithm.
 Another option for reducing MWA data is the Real-Time System (RTS), which has been designed as an efficient calibrating and imaging pipeline specifically for MWA data \citep{rts-mwa}.
@@
 \includegraphics[width=8cm]{img/aliasing-example-annotated}
-\caption{Aliasing artefacts caused by insufficient $w$-layers.
+\caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field.
 WSClean was set to use 12 $w$-layers.
-The centre of the image is at 10\degree zenith angle.
-This would normally require imaging with $\sim195$ $w$-layers.
+The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers.
 Original sources are 1~Jy, ghost sources are approximately 20\%.
-Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce ghost sources by themselves.
+Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce more ghost sources.
 }
@@
 
-\subsection{Time complexity of w-stacking}
+\subsection{Time complexity of w-stacking} \label{sec:time-complexity}
 It is useful to analyse the time complexity of w-stacking and compare it with w-projection, to understand which algorithm performs better in a given situation.
@@
 With the purpose of testing new algorithms for the imaging of MWA data, we have written a new imager around the $w$-stacking algorithm.
-Nevertheless, the imager is not specialized for the MWA, and has been succesfully used for other telescopes (TODO menton more).
+The imager is not specialized for the MWA, and has been succesfully used for other telescopes (TODO menton more).
 
@@
 For the latter, intrinsics are used as well.
-Because of these optimizations, minor Cleaning iterations are performed at a rate of hundreds per second at $3072\times3072$.
+Because of these optimizations, minor Cleaning iterations with an image size of $3072\times3072$ are performed at a rate of hundreds per second.
 This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, less relevant.
@@
 \end{figure*}
-As shown, all-sky imaging can be quite efficient with snapshot imaging towards zenith.
+As discussed, all-sky snapshot imaging with zenith as phase centre is quite efficient.
 However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.
@@
 
-In all analyses CASA stable release 42.0, revision 26465 was used, which was released September 2013.
+The used software version of CASA is ``stable release 42.0, revision 26465'', which was released September 2013.
 It includes performance improvements to the \texttt{clean} task which have recently been made.
-The latest version as of February 2014 of WSClean was used.
-The tests were run on a high-end desktop, with a 3.20-GHz Intel Core i7-3930K processor and 32 GB of memory.
+For WSClean, the latest version as of February 2014 was used.
+The tests were run on a high-end desktop, with a 3.20-GHz Intel Core i7-3930K processor with 6 cores and 32 GB of memory.
 The data are stored on a multi-disk array with 5 spinning hard disks, which has a combined read rate of about 450 MB/s.
@@
 
-We performed measurements to assess the accuracy of WSClean and CASA.
-To do so, we have simulated a MWA observation without system noise, but with hundred sources of 1~Jy in a 20\degree diameter area.
+To asses the accuracy of WSClean and CASA, we have simulated a MWA observation without system noise, but with hundred sources of 1~Jy in a 20\degree diameter area.
 A unitary primary beam is assumed.
@@
 Other imaging parameters are given in Table~\ref{tbl:default-parameters}.
-The Aegean program \citep{aegean-hancock-2012} was used to perform source detection on the produced images.
-Sidelobe noise of the residual 10~mJy noise triggers a few false detections.
+The Aegean program \citep{aegean-hancock-2012} is used to perform source detection on the produced images.
+Sidelobe noise of the residual 10~mJy source structures triggers a few false detections.
 These are ignored.
@@
 
-In the $ZA=10\degree$ case, CASA is slightly less accurate.
-This is consistent, and can actually be observed in the images as shown in Fig.~\ref{fig:residuals}.
-TODO this might be caused by a bug; check: We have also made an image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}.
+In the $\textrm{ZA}=10\degree$ case, CASA is slightly less accurate.
+We have noticed this consistently at higher zenith angles, and it can actually be observed in the produced images, as shown in Fig.~\ref{fig:residuals}.
+TODO the following might be caused by a bug; check: We have also made an image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}.
 Its accuracy is lower compared to the normal projection.
@@
 These include the number of major iterations, $w$-planes/layers, restoring beam, image size, weighting, kernel size and grid oversampling.
-Therefore, the differences are not very significant.
+Therefore, the differences in accuracy are not very significant.
 The results do show that synthesis imaging using common imaging parameters results in an approximate flux-density standard error of at least $\sim1\%$.
@@
 This implies that gridding or reading dominates the cost.
-In that situation, the $w$-stacking implementation is 7.9 times faster at $\textrm{ZA}=10\degree$ and 2.9 times faster at $\textrm{ZA}=0\degree$.
-At $\textrm{ZA}=10$\degree, the performance of WSClean is approximately constant for sets with less than about $10^8$ visibilities, which corresponds with a single polarization of 16~s of MWA data in multi-frequency synthesis mode.
-At that point, the FFTs are the dominant cost in the algorithm.
+In that situation, the $w$-stacking implementation is 7.9 times faster at $\textrm{ZA}=10\degree$ and 2.6 times faster at $\textrm{ZA}=0\degree$.
+With small data volumes, the FFTs start to dominate the cost.
+This is visible from Fig.~\ref{fig:timing-nsamples} by the flattening towards the left.
+At that point, WSClean is in both cases approximately 3 times faster.
 At zenith, the two methods are expected to perform almost identical.
@@
 \caption{Effect of zenith angle (ZA) on imaging performance.
-Image width and height are $2048$ for the dashed line and $3072$ for the solid line.
-CASA fails to image observations with ZA $\ge 15$ with a $3072$ image size.
-ZA dramatically affects the computational cost of w-projection.
+Image pixel size is $2048$ for the dashed lines and $3072$ for the solid lines.
+CASA fails to image observations with ZA $\ge 15$ with a $3072$ pixel image size.
 }
@@
 
-In Fig.~\ref{fig:timing-zenith-angle}, the computational cost as function of the zenith angle is plotted.
+In Fig.~\ref{fig:timing-zenith-angle}, the computational cost as function of zenith angle is plotted.
+It shows that, as expected from \S\ref{sec:time-complexity}, the $w$-projection algorithm is more affected by the increased $w$-values, leading to differences of more than an order of magnitude at zenith angles of $\gtrapprox 20$\degree.
+Additionally, at higher zenith angles, the $w$-kernels become too large to be able to make images of 3072 pixels or larger.
+Using the recentering technique to decrease the $w$-values makes the computational cost approximately constant.
+However, the additional time required to rephase and regrid the measurement set makes it only worthwhile for images $\ge3072$ pixels and $\textrm{ZA}\ge15\degree$.
+TODO: this is quite heavily depending on the speed of \texttt{chgcentre}, and that might still be improvable...
 
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{img/benchmark-resolution/resolution}
+\caption{...}
+\label{fig:timing-resolution}
+\end{center}
+\end{figure}
+
 TODO: for fairness, also check making 4 polarization images.
@@
 \subsection{Gridding} \label{sec:gridding}
-During gridding of the visibilities in $uv$-space, it is useful to apply a gridding convolution kernel.
-A common kernel function is a windowed sinc function.
-A low-pass filter decreases the flux of sources outside the field of view, and thus helps to attenuate aliased ghost sources and sidelobes (see \citealt{post-correlation-filtering}) that arise because of the assumed periodicity during the inverse FFT.
+A gridding convolution kernel improves the accuracy of gridding in $uv$-space.
+A common kernel function is a windowed sinc function, which acts as a low-pass filter.
+This decreases the flux of sources outside the field of view, and thus helps to attenuate aliased ghost sources and sidelobes that arise because of the assumed periodicity during the inverse FFT \citep{post-correlation-filtering}.
 By supersampling, a convolution kernel also makes it possible to place samples more accurately at their $uv$ position, thereby lowering decorrelation.
 
-The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding (TODO cite).
+The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}.
 CASA convolves samples with a PSWF of 7 pixels total width during gridding.
 When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate.
-WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and its properties are similar to a PSWF.
+WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation for the PSWF \citep{fourier-kernel-selection-1991}.
 
@@
 The type of window function has no effect on the gridding performance, but gridding has a quadratic time dependency on the size of the kernel.
-Fig.~\ref{fig:timing-kernelsize} shows that this quadratic dependency becomes apparent for kernel sizes $\ge 15$.
-Decreasing the kernel size to values below $7$ has little effect, because reading the samples takes more time than gridding with small kernels.
+Fig.~\ref{fig:timing-kernelsize} shows that this quadratic dependency becomes apparent for kernel sizes $\gtrapprox 15$.
+Decreasing the kernel size to values below $7$ has little effect on performance, because with small kernels reading the samples takes more time than gridding.
 We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged field of view.
@@
 TODO: describe hybrid between w-stacking, w-projection and w-snapshots (if not yet discussed in Cornwell's papers)
+
+TODO: discuss at what is the optimal time to split measurement sets into when using recentering technique.
 

### 2014-02-02 articles/2014/Article-WSClean/img/fit.cpp

@@ -1,0 +1,674 @@
+#include <iostream>
+#include <string>
+#include <cmath>
+#include <fstream>
+#include <sstream>
+#include <cstdlib>
+#include <vector>
+
+#include <gsl/gsl_vector.h>
+#include <gsl/gsl_multifit_nlin.h>
+
+bool fixDW = false, fixE = false;
+
+double sinZA(double za)
+{
+	return (2900.0*sin(za*(M_PI/180.0)) + 5.5*cos(za*(M_PI/180.0))) / 2900.0;
+}
+
+double fovFact(double fov)
+{
+	double fovRad = 0.5 * fov * (M_PI/180.0);
+	double fov2 = fovRad * fovRad / 2.0;
+	// trueFov = 2.0 * atan(0.5*fov * (M_PI/180.0));
+	return fov2 < 1.0 ? (1.0 - sqrt(1.0 - fov2)) : 1.0;
+}
+
+struct TimingRecord {
+	double average, stddev;
+	size_t n;
+	double sum, sumSq;
+	
+	void Recalculate()
+	{
+		average = sum / double(n);
+		double sumMeanSquared = sum * sum / n;
+		stddev = sqrt((sumSq - sumMeanSquared) / n);
+	}
+	
+	std::string ToLine(const std::string& description, const double val)
+	{
+		std::ostringstream newLine;
+		newLine.precision(11);
+		newLine << description << '\t' << val
+			<< '\t' << average << '\t' << stddev
+			<< '\t' << n << '\t' << sum << '\t' << sumSq;
+		return newLine.str();
+	}
+};
+
+bool parseLine(const std::string& line, std::string& description, double& value, TimingRecord& record)
+{
+	std::string tokens[7];
+	size_t curPos = 0;
+	for(size_t i=0; i!=6; ++i)
+	{
+		size_t delimitor = line.find('\t', curPos);
+		if(delimitor == std::string::npos)
+			return false;
+		tokens[i] = line.substr(curPos, delimitor - curPos);
+		curPos = delimitor+1;
+	}
+	tokens[6] = line.substr(curPos);
+	
+	description = tokens[0];
+	value = atof(tokens[1].c_str());
+	record.average = atof(tokens[2].c_str());
+	record.stddev = atof(tokens[3].c_str());
+	record.n = atof(tokens[4].c_str());
+	record.sum = atof(tokens[5].c_str());
[…]

### 2014-02-03 articles/2014/Article-WSClean/wsclean.tex
Work on benchmarks and article

@@
 \usepackage{xspace}
+\usepackage{subcaption}
 
@@
 
+\begin{figure*}%
+\begin{subfigure}{.5\linewidth}%
+\includegraphics[width=\linewidth]{img/benchmark-nsamples/nsamples}%
+\caption{}\label{fig:timing-nsamples}%
+\end{subfigure}%
+\begin{subfigure}{.5\linewidth}%
+\includegraphics[width=\linewidth]{img/benchmark-zenith-angle/za}%
+\caption{}\label{fig:timing-zenith-angle}%
+\end{subfigure}\\
+\begin{subfigure}{.3\linewidth}%
+\includegraphics[width=\linewidth]{img/benchmark-resolution/resolution}
+\caption{}
+\label{fig:timing-resolution}
+\end{subfigure}%
+\begin{subfigure}{.3\linewidth}%
+\includegraphics[width=\linewidth]{img/benchmark-channels/channels}
+\caption{}
+\label{fig:timing-channels}
+\end{subfigure}%
+\begin{subfigure}{.3\linewidth}%
+\includegraphics[width=\linewidth]{img/benchmark-fov/fov}
+\caption{}
+\label{fig:timing-fov}
+\end{subfigure}%
+\caption{Imaging performance as a function of several configuration parameters. Error bars show 5$\sigma$ level. }\label{fig:timings}
+\end{figure*}
+
 \subsection{Performance analysis}
+
 We measured the performance of the imagers using several MWA data sets.
-Each specific configuration is run 5 times and standard deviations are calculated over these 5 runs.
+Each specific configuration is run 5 times and standard deviations are calculated.
 The variations between runs is typically a few seconds.
@@
 No cleaning or prediction was performed.
+The results are given in Fig.~\ref{fig:timings}.
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=8cm]{img/benchmark-nsamples/nsamples}
-\caption{Effect of data volume on imaging performance for imaging at 10\degree zenith angle (solid lines) and at zenith (dashed lines).
-Error bars show 5$\sigma$ level.
-Both methods are linearly dependent on the number of visibilities for large sets.
-The right-most points represents 112 seconds of data at MWA's nominal resolution of 0.5~s / 40~kHz.
-}
-\label{fig:timing-nsamples}
-\end{center}
-\end{figure}
-
-Fig.~\ref{fig:timing-nsamples} shows the wall-clock time to image a 112-s MWA snapshot as a function of the size of the visibility set.
-Results for imaging at 10\degree zenith angle (ZA) and at zenith are shown.
-The size of the visibility set was varied by changing the time resolution.
-This affects the number of visibilities to be gridded without changing the maximum $w$-value.
+Panel~\ref{fig:timing-nsamples} shows the dependency on the size of the visibility set.
+Results for imaging at zenith and 10\degree zenith angle (ZA) are shown.
+The size of the visibility set was varied by changing the time resolution, which affects the number of visibilities to be gridded without changing the maximum $w$-value.
 For larger sets, both methods show a linear time dependency on the number of visibilities.
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=8cm]{img/benchmark-zenith-angle/za}
-\caption{Effect of zenith angle (ZA) on imaging performance.
-Image pixel size is $2048$ for the dashed lines and $3072$ for the solid lines.
-CASA fails to image observations with ZA $\ge 15$ with a $3072$ pixel image size.
-}
-\label{fig:timing-zenith-angle}
-\end{center}
-\end{figure}
-
-In Fig.~\ref{fig:timing-zenith-angle}, the computational cost as function of zenith angle is plotted.
+In Panel~\ref{fig:timing-zenith-angle}, the computational cost as function of zenith angle is plotted.
 It shows that, as expected from \S\ref{sec:time-complexity}, the $w$-projection algorithm is more affected by the increased $w$-values, leading to differences of more than an order of magnitude at zenith angles of $\gtrapprox 20$\degree.
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=8cm]{img/benchmark-channels/channels}
-\caption{...}
-\label{fig:timing-channels}
-\end{center}
-\end{figure}
+In Panel~\ref{fig:timing-resolution}, the number of pixels in the image is changed without changing the field of view.
+The performance of $w$-stacking is more affected by the size of the image, but is still significantly faster in making 12.8K images at 10\degree ZA than $w$-projection.
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=8cm]{img/benchmark-resolution/resolution}
-\caption{...}
-\label{fig:timing-resolution}
-\end{center}
-\end{figure}
+Panel~\ref{fig:timing-channels} shows performance versus the output frequency resolution.
+The cost of imaging at multiple frequencies is the cost of making an images times the number of frequencies.
+However, each separate imaging contains a part of the total number of visibilities.
+As shown before, FFTs become the dominant cost for small visibility sets, and performance and number of frequency images therefore have a linear relation, which is clearly visible from the plot.
 
 TODO: for fairness, also check making 4 polarization images.
+
+\subsection{Derivation of computing cost formulae}
+Based on the expected cost terms described in Sec.~\ref{sec:time-complexity}, we fit analytical functions to the CASA and WSClean measurements.
+Several functions with different free parameters were tested, and formulae with minimum amount of parameters are selected that still follow the trend of the measurements and have reasonably small errors.
+The measurements are weighted with the inverse standard deviation.
+The single outlying measurement of CASA in Fig.~\ref{fig:timing-nsamples} is removed.
+
+For WSClean's time cost $t_\textrm{WSClean}$ we find 
+\begin{align}\notag
+t_\textrm{WSClean}& (w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = \\ \notag
+& 0.000326 N_\textrm{pix}^2 \log N_\textrm{pix} (w_e + 0.000212) +\\
+& 0.252 N_\textrm{vis} - 7.64
+\end{align}
+and for CASA we find
+\begin{equation}
+t_\textrm{CASA}(w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = .
+\end{equation}
+Parameter $w_e$ is the maximum $w$ value and is esimated with
+\begin{equation}
+ w_e = \frac{\max |\mathbf{x}| \sin za + \max z \cos za}{\lambda \max |\mathbf{x}|} \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right),
+\end{equation}
+where $\max |\mathbf{x}|$ is the maximum baseline length, $\max z$ is the maximum height difference between antennas and $\lambda$ is the wavelength.
 

### 2014-02-03 articles/2014/Article-WSClean/img/fit.cpp

@@ -9,7 +9,7 @@
 #include <gsl/gsl_vector.h>
 #include <gsl/gsl_multifit_nlin.h>
 
-bool fixDW = false, fixE = false;
+bool fixDW = false, fixConstant = false;
 
 double sinZA(double za)
 {
@@ -18,9 +18,8 @@ double sinZA(double za)
 
 double fovFact(double fov)
 {
-	double fovRad = 0.5 * fov * (M_PI/180.0);
-	double fov2 = fovRad * fovRad / 2.0;
-	// trueFov = 2.0 * atan(0.5*fov * (M_PI/180.0));
+	double fovRad = 0.5 * fov * (M_PI/180.0); // this is the max l
+	double fov2 = fovRad * fovRad;
 	return fov2 < 1.0 ? (1.0 - sqrt(1.0 - fov2)) : 1.0;
 }
 
@@ -99,15 +98,15 @@ struct FittingInfo
 	Sample *samples;
 };
 
-double eval(FittingFunc func, double za, double nVis, double nPix, double fov, double a, double b, double c, double d, double e)
+double eval(FittingFunc func, double za, double nVis, double nPix, double fov, double a, double b, double c, double d, double e, double f)
 {
 	if(func == WSCleanFunc) {
 		double we = sinZA(za) * fovFact(fov);
 		return a * (we+b)*nPix*nPix*log2(nPix) + c * nVis + d;
 	}
 	else {
-		double we = sinZA(za) * (fovFact(fov) + d);
-		return a * nPix*nPix*log2(nPix) + nVis * b * we*we + e;
+		double we = (sinZA(za) * fovFact(fov) + d);
+		return a * nPix*nPix*log2(nPix) + b * nVis * we * we + f;
 	}
 }
 
@@ -125,7 +124,7 @@ int fitFunc(const gsl_vector *xvec, void *data, gsl_vector *f)
 		double value, y, w;
 		if(fittingInfo.func == WSCleanFunc || fittingInfo.func == CASAFunc)
 		{
-			const double e = gsl_vector_get(xvec, 4);
+			const double e = gsl_vector_get(xvec, 4), f = gsl_vector_get(xvec, 5);
 			y = fittingInfo.samples[i].time;
 			w = fittingInfo.samples[i].weight;
 			const double
@@ -133,8 +132,7 @@ int fitFunc(const gsl_vector *xvec, void *data, gsl_vector *f)
[…]

### 2014-02-03 articles/2014/Article-WSClean/img/fit.cpp
Article

@@ -515,6 +515,11 @@ void calculateStdError(std::vector<Sample>& samples, FittingFunc func, double a,
 		<< "Unweighted mean: " << unweightedSum/samples.size() << " unweighted stderr: " << stddev(unweightedSum, unweightedSumSq, samples.size()) << '\n';
 }
 
+double optimalDeltaW(FittingFunc func, double a, double b, double c, double d, double e, double f, double za, double nPix, double nVis, double fov)
+{
+	
+}
+
 void fitAllWSClean()
 {
 	std::vector<Sample> samples;

### 2014-02-04 articles/2014/Article-WSClean/wsclean.tex
Article
Work on article

@@
 
+\subsection{Gridding} \label{sec:gridding}
+A gridding convolution kernel improves the accuracy of gridding in $uv$-space.
+A common kernel function is a windowed sinc function, which acts as a low-pass filter.
+This decreases the flux of sources outside the field of view, and thus helps to attenuate aliased ghost sources and sidelobes that arise because of the assumed periodicity during the inverse FFT \citep{post-correlation-filtering}.
+By supersampling, a convolution kernel also makes it possible to place samples more accurately at their $uv$ position, thereby lowering decorrelation.
+
+The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}.
+CASA convolves samples with a PSWF of 7 pixels total width during gridding.
+When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate.
+WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation for the PSWF \citep{fourier-kernel-selection-1991}.
+After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that 7 pixels is a good compromise between performance and accuracy, and selected this as the default for WSClean.
+
 \section{Analysis} \label{sec:analysis}
@@
 \begin{figure*}%
-\begin{subfigure}{.5\linewidth}%
+\begin{subfigure}{.36\linewidth}%
 \includegraphics[width=\linewidth]{img/benchmark-nsamples/nsamples}%
@@
 \end{subfigure}%
-\begin{subfigure}{.5\linewidth}%
+\begin{subfigure}{.36\linewidth}%
 \includegraphics[width=\linewidth]{img/benchmark-zenith-angle/za}%
@@
 \end{subfigure}\\
-\begin{subfigure}{.3\linewidth}%
+\begin{subfigure}{.36\linewidth}%
 \includegraphics[width=\linewidth]{img/benchmark-resolution/resolution}
-\caption{}
-\label{fig:timing-resolution}
+\caption{}\label{fig:timing-resolution}%
 \end{subfigure}%
-\begin{subfigure}{.3\linewidth}%
+\hspace{-.05\linewidth}\begin{subfigure}{.36\linewidth}%
 \includegraphics[width=\linewidth]{img/benchmark-channels/channels}
-\caption{}
-\label{fig:timing-channels}
+\caption{}\label{fig:timing-channels}%
 \end{subfigure}%
-\begin{subfigure}{.3\linewidth}%
+\hspace{-.05\linewidth}\begin{subfigure}{.36\linewidth}%
 \includegraphics[width=\linewidth]{img/benchmark-fov/fov}
@@
 \subsection{Derivation of computing cost formulae}
-Based on the expected cost terms described in Sec.~\ref{sec:time-complexity}, we fit analytical functions to the CASA and WSClean measurements.
+Based on the expected cost terms described in Sec.~\ref{sec:time-complexity}, we derive analytical functions for the CASA and WSClean measurements using least-squares fitting.
 Several functions with different free parameters were tested, and formulae with minimum amount of parameters are selected that still follow the trend of the measurements and have reasonably small errors.
@@
 \end{equation}
-Parameter $w_e$ is the maximum $w$ value and is esimated with
+Parameter $w_e$ is the maximum $w$ value and is estimated with
 \begin{equation}
@@
 where $\max |\mathbf{x}|$ is the maximum baseline length, $\max z$ is the maximum height difference between antennas and $\lambda$ is the wavelength.
-
-\subsection{Gridding} \label{sec:gridding}
-A gridding convolution kernel improves the accuracy of gridding in $uv$-space.
-A common kernel function is a windowed sinc function, which acts as a low-pass filter.
-This decreases the flux of sources outside the field of view, and thus helps to attenuate aliased ghost sources and sidelobes that arise because of the assumed periodicity during the inverse FFT \citep{post-correlation-filtering}.
-By supersampling, a convolution kernel also makes it possible to place samples more accurately at their $uv$ position, thereby lowering decorrelation.
-
-The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}.
-CASA convolves samples with a PSWF of 7 pixels total width during gridding.
-When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate.
-WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation for the PSWF \citep{fourier-kernel-selection-1991}.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
-\caption{...}
-\label{fig:timing-kernelsize}
-\end{center}
-\end{figure}
-
-The type of window function has no effect on the gridding performance, but gridding has a quadratic time dependency on the size of the kernel.
-Fig.~\ref{fig:timing-kernelsize} shows that this quadratic dependency becomes apparent for kernel sizes $\gtrapprox 15$.
-Decreasing the kernel size to values below $7$ has little effect on performance, because with small kernels reading the samples takes more time than gridding.
-We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged field of view.
-Therefore, WSClean uses a default of 7 pixels for the gridding kernel.
-It can be increased when necessary.
 
@@
 ...
+\appendix
+
+\section{Optimal gridding kernel size} \label{sec:gridding-appendix}
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
+\caption{...}
+\label{fig:timing-kernelsize}
+\end{center}
+\end{figure}
+As discussed in \S\ref{sec:gridding}, WSClean uses a Kaiser-Bessel window against aliasing.
+The type of window function has no effect on the gridding performance, but the size of the kernel affects gridding performance quadratically.
+Fig.~\ref{fig:timing-kernelsize} shows that this quadratic relation becomes significant for kernel sizes $\gtrapprox 10$.
+Decreasing the kernel size to values below $7$ has little effect on performance, because for small kernels the visibility-reading rate is lower than the gridding rate.
+We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged field of view.
+Therefore, WSClean uses a default of 7 pixels for the gridding kernel.
+It can be increased when necessary.
+Different hardware might have a slightly different optimum because of different reading and calculation performance.
 

### 2014-02-04 articles/2014/Article-WSClean/img/fit.cpp

@@ -5,6 +5,8 @@
 #include <sstream>
 #include <cstdlib>
 #include <vector>
+#include <stdexcept>
+#include <limits>
 
 #include <gsl/gsl_vector.h>
 #include <gsl/gsl_multifit_nlin.h>
@@ -82,12 +84,13 @@ enum FittingFunc
 	NZASqFunc2,
 	NZASqFunc3,
 	WSCleanFunc,
+	WSSCleanFunc,
 	CASAFunc
 };
 
 struct Sample
 {
-	double time, weight, za, nPix, nVis, fov;
+	double time, weight, za, nPix, nVis, fov, nChan;
 };
 
 struct FittingInfo
@@ -98,15 +101,45 @@ struct FittingInfo
 	Sample *samples;
 };
 
-double eval(FittingFunc func, double za, double nVis, double nPix, double fov, double a, double b, double c, double d, double e, double f)
+double eval(FittingFunc func, double za, double nVis, double nPix, double fov, double nChan, double a, double b, double c, double d, double e, double f)
 {
 	if(func == WSCleanFunc) {
 		double we = sinZA(za) * fovFact(fov);
-		return a * (we+b)*nPix*nPix*log2(nPix) + c * nVis + d;
+		return nChan * (a * (we+b)*nPix*nPix*log2(nPix) + c * nVis/nChan) + d;
+	}
+	else if(func == WSSCleanFunc) {
+		return nChan * (a *nPix*nPix*log2(nPix) + b * nVis/nChan) + c;
 	}
 	else {
 		double we = (sinZA(za) * fovFact(fov) + d);
-		return a * nPix*nPix*log2(nPix) + b * nVis * we * we + f;
+		return nChan * (a * nPix*nPix*log2(nPix) + b * nVis/nChan * we * we) + f;
+	}
+}
+
+double eval(FittingFunc func, double za, double nVis, double nPix, double fov, double nChan)
+{
+	switch(func)
+	{
+		case WSCleanFunc:
+			return eval(func, za, nVis, nPix, fov, nChan, 0.00013703, 0.000432297, 0.241851, 0.0, 0.0, 0.0);
+		case CASAFunc:
+			return eval(func, za, nVis, nPix, fov, nChan, 1.97854e-07, 47825.5, 0.0, 0.000473665, 0.0, 235.922);
+		default:
+			return 0.0;
+	}
+}
+
[…]

### 2014-02-04 articles/2014/Article-WSClean/wsclean.tex
update
Work on article

@@
 & 0.000326 N_\textrm{pix}^2 \log N_\textrm{pix} (w_e + 0.000212) +\\
+\label{eq:wsclean-computing-cost}%
 & 0.252 N_\textrm{vis} - 7.64
@@
 t_\textrm{CASA}(w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = .
+\label{eq:casa-computing-cost}%
 \end{equation}
@@
 where $\max |\mathbf{x}|$ is the maximum baseline length, $\max z$ is the maximum height difference between antennas and $\lambda$ is the wavelength.
+
+TODO: actually, $w_e$ should not be normalized by baseline length $\max z$: fix in fits too.
+
+It is possible to combine the $w$-projection and $w$-stacking algorithms by gridding on several $w$-layers to which the large $w$-terms are corrected after the FFT, and using a $w$-correcting kernel for making small $w$-corrections, thereby projecting them on the nearest $w$-layer.
+This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory.
+The AWImager has been applied like this, which let to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}.
+For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more general the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_e$ value and smaller $N_\textrm{vis}$.
+If $\Delta w$ is the distance between $w$-layers, then
+\begin{equation}
+t(\Delta w) = \frac{w_e}{\Delta w} t_\textrm{CASA}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
+\end{equation}
+with $N'_\textrm{vis} = \frac{N_\textrm{vis}\Delta w}{w_e}$ and $w'_e = \Delta w$.
 

### 2014-02-05 articles/2014/Article-WSClean/wsclean.tex
Improved WSSC benchmark and channel benchmark, work on article

@@
 
+TODO: derive function for $t_\textrm{CASA} = t_\textrm{WSClean}$.
+
 It is possible to combine the $w$-projection and $w$-stacking algorithms by gridding on several $w$-layers to which the large $w$-terms are corrected after the FFT, and using a $w$-correcting kernel for making small $w$-corrections, thereby projecting them on the nearest $w$-layer.
@@
 \section{Application of WSClean to the MWA}
-The MWA correlator splits observations in data products of typically a few minutes
-TODO
+For technical reasons, the MWA correlator splits observations in data products of typically a few minutes.
+This simplifies $w$-snapshot imaging, which can be done by phase rotating each snapshot to zenith with a possibly additional phase-centre shift.
 
 TODO: calculate when to use w-snapshots
+TODO: calculate optimum splitting time?
 
-TODO: mention full-polarization imaging
+\subsection{Beam correction}
+Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix.
+Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}.
+When assuming all individual antennas of the tiles are working properly, all tiles have the same beam.
+The beam can then be corrected in image space for all tiles at once with
+\begin{equation}\label{eq:beam-correction}
+ I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
+\end{equation}
+where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose.
+Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are often referred to as $X$ and $Y$ because many software treats the two polarizations as such.
+However, this is not correct because $X$ is officially defined as being parallel to the celestial equator.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters.
+The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry.
+To make proper MWA beam correction possible, WSClean can output real valued $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images.
+Each polarization requires a separate run of the $w$-stacking algorithm.
+Once these six images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}.
+This method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam.
+To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm.
+The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.
 
 \section{Discussion} \label{sec:discussion}
-WSClean has been build out of necessacity to be able to make images with the MWA.
+WSClean has been build out of necessacity to be able to make images with the MWA, because no
 
@@
 \includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
-\caption{...}
+\caption{Effect of antialiasing kernel size on performance.}
 \label{fig:timing-kernelsize}
@@
 It can be increased when necessary.
-Different hardware might have a slightly different optimum because of different reading and calculation performance.
+Of course, different hardware might give slightly different results because of different reading and calculation performance.
 

### 2014-02-05 articles/2014/Article-WSClean/img/fit.cpp

@@ -91,6 +91,8 @@ enum FittingFunc
 struct Sample
 {
 	double time, weight, za, nPix, nVis, fov, nChan;
+	
+	bool sameConfiguration(const Sample& s) const { return s.za==za && s.nPix==nPix && s.nVis==nVis && s.fov==fov && s.nChan==nChan; }
 };
 
 struct FittingInfo
@@ -595,6 +597,19 @@ void calculateStdError(std::vector<Sample>& samples, FittingFunc func, double a,
 		<< "Unweighted mean: " << unweightedSum/samples.size() << " unweighted stderr: " << stddev(unweightedSum, unweightedSumSq, samples.size()) << '\n';
 }
 
+void printDuplicates(const std::vector<Sample>& samples)
+{
+	for(std::vector<Sample>::const_iterator s1=samples.begin(); s1!=samples.end(); ++s1)
+	{
+		for(std::vector<Sample>::const_iterator s2=s1+1; s2!=samples.end(); ++s2)
+		{
+			if(s1->sameConfiguration(*s2))
+				std::cout << "Duplicate configuration: " << s1->za << ", " << s1->nVis << ", " << s1->nPix << ", " << s1->fov << ", " << s1->nChan <<
+					"\n Timings: " << s1->time << " vs " << s2->time << '\n';
+		}
+	}
+}
+
 void optimalDeltaW(FittingFunc func, double a, double b, double c, double d, double e, double f, double za, double nVis, double nPix, double fov, double nChan, bool zeroConstant, double& dw, double& casaGain, double& wscleanGain)
 {
 	double we = sinZA(za) * fovFact(fov);
@@ -663,6 +678,7 @@ void fitAllWSClean()
 	readFOVFile(samples, "benchmark-fov/timings-zenith-fov-wsc.txt", 0.0, 3072, 349.6, 1, 1.0);
 	readNChanFile(samples, "benchmark-channels/timings-channels-wsc.txt", 0.0, 3072, 349.6, 24.576, 1.0);
 	std::cout << "Read " << samples.size() << " values.\n";
+	printDuplicates(samples);
 	FittingInfo info;
 	info.n = samples.size();
 	info.samples = samples.data();
@@ -687,16 +703,17 @@ void fitAllWSSClean()
[…]

### 2014-02-05 articles/2014/Article-WSClean/wsclean.tex
Some work on article

@@
 
-It is possible to combine the $w$-projection and $w$-stacking algorithms by gridding on several $w$-layers to which the large $w$-terms are corrected after the FFT, and using a $w$-correcting kernel for making small $w$-corrections, thereby projecting them on the nearest $w$-layer.
+It is possible to combine the $w$-projection and $w$-stacking algorithms by gridding onto several $w$-layers to which large $w$-terms are corrected after the FFT, and using a $w$-correcting kernel for making small $w$-corrections.
 This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory.

### 2014-02-05 articles/2014/Article-WSClean/img/fit.cpp

@@ -667,6 +667,28 @@ void optimalDeltaW(const std::vector<Sample>& samples, FittingFunc func, double 
 	std::cout << ", mean dw=" << sumDW/count << '\n';
 }
 
+void optimalDeltaT(FittingFunc func, double a, double b, double c, double d, double e, double f, double za, double nVis, double nPix, double fov, double nChan, bool zeroConstant, double& dt)
+{
+	double minY = std::numeric_limits<double>::max(), minDT = 0.0;
+	double dtTrial = 1.0;
+	double fz = zeroConstant ? 0.0 : f;
+	while(dtTrial < 300000.0)
+	{
+		double trialFOV = fov + 2.0 * (dtTrial - 112.0) * (360.0 / 60.0 / 60.0 / 24.0);
+		if(trialFOV < 0.0) trialFOV = 0.0;
+		double y = (112.0/dtTrial) * eval(func, za, nVis * dtTrial/112.0, nPix, trialFOV, nChan, a, b, c, d, e, fz);
+		//std::cout << "za=" << trialFOV<< ", dt=" << dtTrial << ", y=" << y << '\n';
+		if(y < minY) {
+			minDT = dtTrial;
+			minY = y;
+		}
+		dtTrial *= 1.01;
+	}
+	dt = minDT;
+	double tNormal = eval(func, za, nVis, nPix, fov, nChan, a, b, c, d, e, fz);
+	std::cout << "Optimal snapshot time: " << minDT << " (time=" << minY << " vs " << tNormal << ")\n";
+}
+
 void fitAllWSClean()
 {
 	std::vector<Sample> samples;
@@ -696,6 +718,9 @@ void fitAllWSClean()
 	writeFOVFit("benchmark-fov/fit-ZA010-fov-wsc.txt", info.func, a, b, c, d, e, f, 10.0, 3072, 349.6, 1);
 	writeNChanFit("benchmark-channels/fit-zenith-wsc.txt", info.func, a, b, c, d, e, f, 0.0, 3072, 349.6, 36.864);
 	calculateStdError(samples, info.func, a, b, c, d, e, f);
+	double dt;
+	optimalDeltaT(info.func, a, b, c, d, e, f, 10.0, 349.6, 3072, 36.864, 1, false, dt);
+	optimalDeltaT(info.func, a, b, c, d, e, f, 0.0, 349.6, 3072, 36.864, 1, false, dt);
 }
 
[…]

### 2014-02-05 chgcentre/main.cpp

@@ -289,6 +289,30 @@ MDirection ZenithDirection(MeasurementSet& set)
 	return casa::MDirection::Convert(zenithAzEl, j2000Ref)();
 }
 
+MDirection ZenithDirectionStart(MeasurementSet& set)
+{
+	casa::MPosition arrayPos = ArrayCentroid(set);
+	casa::MEpoch::ROScalarColumn timeColumn(set, set.columnName(casa::MSMainEnums::TIME));
+	casa::MEpoch time = timeColumn(0);
+	casa::MeasFrame frame(arrayPos, time);
+	const casa::MDirection::Ref azelgeoRef(casa::MDirection::AZELGEO, frame);
+	const casa::MDirection::Ref j2000Ref(casa::MDirection::J2000, frame);
+	casa::MDirection zenithAzEl(casa::MVDirection(0.0, 0.0, 1.0), azelgeoRef);
+	return casa::MDirection::Convert(zenithAzEl, j2000Ref)();
+}
+
+MDirection ZenithDirectionEnd(MeasurementSet& set)
+{
+	casa::MPosition arrayPos = ArrayCentroid(set);
+	casa::MEpoch::ROScalarColumn timeColumn(set, set.columnName(casa::MSMainEnums::TIME));
+	casa::MEpoch time = timeColumn(set.nrow()-1);
+	casa::MeasFrame frame(arrayPos, time);
+	const casa::MDirection::Ref azelgeoRef(casa::MDirection::AZELGEO, frame);
+	const casa::MDirection::Ref j2000Ref(casa::MDirection::J2000, frame);
+	casa::MDirection zenithAzEl(casa::MVDirection(0.0, 0.0, 1.0), azelgeoRef);
+	return casa::MDirection::Convert(zenithAzEl, j2000Ref)();
+}
+
 MDirection MinWDirection(MeasurementSet& set)
 {
 	MPosition centroid = ArrayCentroid(set);
@@ -353,7 +377,7 @@ void printPhaseDir(const std::string &filename)
 		std::cout << dirToString(phaseDirection) << '\n';
 	}
 	
-	std::cout << "Zenith is at:\n  " << dirToString(zenith) << '\n';
+	std::cout << "Zenith is at:\n  " << dirToString(zenith) << " (" << dirToString(ZenithDirectionStart(set)) << " - " << dirToString(ZenithDirectionEnd(set)) << '\n';
 	std::cout << "Min-w direction is at:\n  " << dirToString(MinWDirection(set)) << '\n';
 }
 

### 2014-02-06 articles/2014/Article-WSClean/wsclean.tex
New results and article almost finished.

@@
 The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers.
-Original sources are 1~Jy, ghost sources are approximately 20\%.
+Sources are 1~Jy (red circles), ghost sources are approximately 0.2~Jy.
 Each source produces two ghost sources, but because they re-appear after a major cleaning cycle, they are eventually cleaned and produce more ghost sources.
@@
 \subsection{Full-sky imaging}
-For low-frequency telescopes, it can be of interest to do full-sky imaging.
+For low-frequency telescopes, it can be of interest to do full-sky (i.e., horizon to horizon) imaging.
 In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90\degree from the pointing centre.
-Imaging these sidelobes might be relevant because it might either be scientifically interesting (e.g.
-when searching for transients), or be useful for self-calibration or full-sky deconvolution.
-For example, the latter is useful when a strong source in a sidelobe affects the sensitivity near the pointing centre.
+Imaging these sidelobes might be relevant because they are scientifically of interest (e.g.
+when searching for transients).
+Full-sky imaging can also be useful for self-calibration or deconvolution.
+For example, self-calibration using full-sky clean components has been found to give good results in imaging the resolved FR-II radio source Fornax~A (McKinley et al., in prep.).
 
@@
 WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation for the PSWF \citep{fourier-kernel-selection-1991}.
-After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that 7 pixels is a good compromise between performance and accuracy, and selected this as the default for WSClean.
+After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that 7 pixels is a good compromise between performance and accuracy, and this is the default for WSClean.
 
@@
 
-TODO: for fairness, also check making 4 polarization images.
+It is important to point out that all these measurement were performed by imaging a single polarization.
+Making full-polarization images with WSClean currently requires four runs.
+Imaging four Stokes parameters with CASA costs slightly less than four times the time of imaging a single polarization.
+With the settings of Table~\ref{tbl:default-parameters}, the difference is a factor of $3.5$, and imaging XX and YY together costs $1.9$ times imaging just XX.
 
@@
 Several functions with different free parameters were tested, and formulae with minimum amount of parameters are selected that still follow the trend of the measurements and have reasonably small errors.
-The measurements are weighted with the inverse standard deviation.
+Measurements are weighted with the inverse standard deviation.
 The single outlying measurement of CASA in Fig.~\ref{fig:timing-nsamples} is removed.
 
-For WSClean's time cost $t_\textrm{WSClean}$ we find 
-\begin{align}\notag
-t_\textrm{WSClean}& (w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = \\ \notag
-& 0.000326 N_\textrm{pix}^2 \log N_\textrm{pix} (w_e + 0.000212) +\\
-\label{eq:wsclean-computing-cost}%
-& 0.252 N_\textrm{vis} - 7.64
+For the WSClean time cost function $t_\textrm{WSClean}$ we find 
+\begin{align}\label{eq:wsclean-computing-cost}
+& t_\textrm{WSClean} (w_e,N_\textrm{Mvis},N_\textrm{freq},N_\textrm{kpix}) = \\ \notag%
+& N_\textrm{chan} \left( 0.452 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} (w_e + 0.611) + 244 N_\textrm{Mvis} \right)
 \end{align}
 and for CASA we find
-\begin{equation}
-t_\textrm{CASA}(w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = .
+\begin{align}\label{eq:casa-computing-cost}
+& t_\textrm{CASA}(w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = \\ \notag%
+& N_\textrm{chan} \left(1.03 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} + 5.93 N_\textrm{Mvis} (w_e + 1.71)^2\right) + 172.
 \label{eq:casa-computing-cost}%
-\end{equation}
+\end{align}
 Parameter $w_e$ is the maximum $w$ value and is estimated with
 \begin{equation}
- w_e = \frac{\max |\mathbf{x}| \sin za + \max z \cos za}{\lambda \max |\mathbf{x}|} \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right),
+ w_e = \frac{1}{\lambda}\left[\max |\mathbf{x}| \sin \textrm{ZA} + \max z \cos \textrm{ZA} \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right)\right],
 \end{equation}
@@
 
-TODO: actually, $w_e$ should not be normalized by baseline length $\max z$: fix in fits too.
-
-TODO: derive function for $t_\textrm{CASA} = t_\textrm{WSClean}$.
-
-It is possible to combine the $w$-projection and $w$-stacking algorithms by gridding onto several $w$-layers to which large $w$-terms are corrected after the FFT, and using a $w$-correcting kernel for making small $w$-corrections.
+The $w$-projection and $w$-stacking algorithms can be combined by gridding onto several $w$-layers to which large $w$-terms are corrected after the FFTs, and using a $w$-correcting kernel for making small $w$-corrections before the FFTs.
 This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory.
 The AWImager has been applied like this, which let to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}.
-For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more general the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_e$ value and smaller $N_\textrm{vis}$.
+For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more generally the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_e$ value and smaller $N_\textrm{vis}$.
 If $\Delta w$ is the distance between $w$-layers, then
@@
 
+TODO fill in formulae for common surveys, e.g. ESKAP and LOFAR.
+
 \section{Application of WSClean to the MWA}
@@
 
-TODO: calculate when to use w-snapshots
+TODO: calculate when to use w-snapshots: discuss at what is the optimal time to split measurement sets into when using recentering technique.
 TODO: calculate optimum splitting time?
@@
 where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose.
-Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are often referred to as $X$ and $Y$ because many software treats the two polarizations as such.
-However, this is not correct because $X$ is officially defined as being parallel to the celestial equator.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters.
+Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such.
+However, this is not correct, because $X$ is defined as being parallel to the celestial equator.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters.
 The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry.
-To make proper MWA beam correction possible, WSClean can output real valued $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images.
+To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images.
 Each polarization requires a separate run of the $w$-stacking algorithm.
@@
 
-\section{Discussion} \label{sec:discussion}
-WSClean has been build out of necessacity to be able to make images with the MWA, because no
+\section{Discussion \& conclusions} \label{sec:discussion}
+We have shown that the $w$-stacking algorithm is well suited for imaging MWA observations.
+The WSClean $w$-stacking implementation is faster than CASA's $w$-projection algorithm in all common MWA imaging configurations, and already an order of magnitude faster at a relatively small zenith angle of $10$\degree.
+At the same time, it is slightly more accurate.
+For approximately zenith angles $>15\degree$ or field of views $>35\degree$, rephasing visibility sets and performing snapshot imaging becomes worthwhile.
+This can lead to performance improvements of a factor of $3$ for the MWA.
+For arrays with longer baselines and higher resolution, snapshot imaging leads to even higher performance improvements.
+This agrees with \citet{widefield-imaging-ska-cornwell}.
+Additionally, we have shown that to implement $w$-snapshot imaging, a pure $w$-stacking implementation is likely more efficient than pure $w$-projection.
+A hybrid between the two might improve things by a few factors.
+TODO should we calculate some timings with SKA's configuration?
 
-TODO: describe hybrid between w-stacking, w-projection and w-snapshots (if not yet discussed in Cornwell's papers)
-
-TODO: discuss at what is the optimal time to split measurement sets into when using recentering technique.
-
-TODO: A-projection makes the psf more invariant: therefore, truely ``joined'' deconvolution might be more accurate.
-
-TODO: discuss low w-values means less-direction-invariant PSF.
+It remains challenging to joinedly deconvolve observations that are long enough to encur significant changes in the beam shape.
+It is possible to clean such an observation without beam corrections, but then a constant amount of flux is subtracted over time, which leads to inaccuracies.
+Alternatively, beam corrections can be made, either by integrating beam-corrected snapshots or by performing $a$-projection.
+This however leads to a stronger spatially-varying point-spread function, and would allow fewer image-based minor iterations in a major cycle.
+When it is required to calculate the beam on short time scales, calculating $a$-projection kernels or performing snapshot imaging also becomes much more expensive.
+For a few hundred components, an approach with direct Fourier transforms is accurate and affordable, but this becomes intractable for thousands of components.
+Images with diffuse emmission typically require cleaning of more than ten-thousand components.
+Joined deconvolution is not an issue for achieving low to intermediate dynamic range with the MWA.
+Because of the excellent instanteneous $uv$-coverage and near-confusion noise level, snapshots can be cleaned individually to deep levels.
+For example, cleaning to $5\sigma$ level results in a cleaning threshold of approximately 100~mJy in 112~s observations.
+However, MWA's EoR observations (TODO cite) require more advanced deconvolution techniques.
 

### 2014-02-06 articles/2014/Article-WSClean/img/fit.cpp

@@ -15,7 +15,7 @@ bool fixDW = false, fixConstant = false;
 
 double sinZA(double za)
 {
-	return (2900.0*sin(za*(M_PI/180.0)) + 5.5*cos(za*(M_PI/180.0))) / 2900.0;
+	return (2900.0*sin(za*(M_PI/180.0)) + 5.5*cos(za*(M_PI/180.0)));
 }
 
 double fovFact(double fov)
@@ -403,8 +403,10 @@ void addSample(std::vector<Sample>& samples, const Sample& sample)
 {
 	if(!std::isfinite(sample.weight) || sample.weight==0.0)
 		std::cout << "Removing sample with 0 stddev\n";
-	else
+	else {
 		samples.push_back(sample);
+		samples.back().nPix *= 0.001;
+	}
 }
 
 void readNVisFile(std::vector<Sample>& samples, const std::string& filename, double za, double nPix, double fov, double nChan, double w)
@@ -532,7 +534,7 @@ void writeNPixFit(const std::string& filename, FittingFunc func, double a, doubl
 	while(nPix < xEnd)
 	{
 		nPix *= 1.01;
-		file << nPix << '\t' << eval(func, za, nVis, nPix, fov, nChan, a, b, c, d, e, f) << '\n';
+		file << nPix << '\t' << eval(func, za, nVis, nPix/1000.0, fov, nChan, a, b, c, d, e, f) << '\n';
 	}
 }
 
@@ -632,7 +634,7 @@ void optimalDeltaW(FittingFunc func, double a, double b, double c, double d, dou
 	if(minDW == we)
 		dw = std::numeric_limits<double>::quiet_NaN();
 	else
-		dw = minDW*2900.0;
+		dw = minDW;
 	casaGain = eval(CASAFunc, za, nVis, nPix, fov, nChan)/minY;
 	wscleanGain = eval(WSCleanFunc, za, nVis, nPix, fov, nChan)/minY;
 }
@@ -709,18 +711,18 @@ void fitAllWSClean()
 	double a = 1.0, b = 0.0, c = 1.0, d = 0.0, e = 0.0, f = 0.0;
 	fit(info, a, b, c, d, e, f);
[…]

### 2014-02-06 articles/2014/Article-WSClean/wsclean.tex
Article

@@
 \begin{equation}
- w_e = \frac{1}{\lambda}\left[\max |\mathbf{x}| \sin \textrm{ZA} + \max z \cos \textrm{ZA} \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right)\right],
+ w_e = \frac{1}{\lambda}\left[\max |\Delta \mathbf{x}| \sin \textrm{ZA} + \max \Delta z \cos \textrm{ZA} \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right)\right],
 \end{equation}
-where $\max |\mathbf{x}|$ is the maximum baseline length, $\max z$ is the maximum height difference between antennas and $\lambda$ is the wavelength.
+where $\max |\Delta \mathbf{x}|$ is the maximum baseline length, $\max \Delta z$ is the maximum height difference between antennas and $\lambda$ is the wavelength.
+
+\begin{table*}
+\begin{tabular}{lrrrrrrr}
+ Survey   & $\lambda$ & FWHM  & Resolution & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. time & B/$\delta \nu$ \\
+\hline
+ GLEAM    &           &       & 2'     & 2.9~km & 5~m & 2~s & 32~MHz/40~kHz\\
+ EMU wide & 20 cm  & 1\degree & 10''    & 6~km &  0~m &     & ?/?\\
+ MSSS low &  5 m   & 9.8\degree & 100'' & 5~km &      &10~s & 16~MHz/4~khZ\\
+ MSSS high&  2 m   & 3.8\degree & 120'' & 5~km &      &10~s & 16~MHz/4~khZ\\
+\end{tabular}
+\end{table*}
 

### 2014-02-07 articles/2014/Article-WSClean/wsclean.tex
Article almost done

@@
 Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}).
-At the MWA, some of the data is reduced with the w-projection algorithm implementation of the Common Astronomy Software Applications (CASA, \citealt{casa}).
+At the MWA, some of the data is reduced with the w-projection algorithm implementation of the Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}).
 This is an attractive option because of its many features, high quality and free availability.
@@
 It reads visibilities from CASA measurement sets and writes output images to Flexible Image Transport System (FITS) files.
-Several steps are multi-threaded: reading and writing; gridding from and to different $w$-layers; performing the Fast Fourier transforms; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}).
+Several steps are multi-threaded using the threading module of the C++11 standard library.
+These are: reading and writing; gridding from and to different $w$-layers; performing the Fast Fourier transforms; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}).
 For the latter, intrinsics are used as well.
@@
 & t_\textrm{WSClean} (w_e,N_\textrm{Mvis},N_\textrm{freq},N_\textrm{kpix}) = \\ \notag%
-& N_\textrm{chan} \left( 0.452 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} (w_e + 0.611) + 244 N_\textrm{Mvis} \right)
+& N_\textrm{freq} \left( 0.452 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} (w_e + 0.611) + 244 N_\textrm{Mvis} \right)
 \end{align}
@@
 & t_\textrm{CASA}(w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = \\ \notag%
-& N_\textrm{chan} \left(1.03 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} + 5.93 N_\textrm{Mvis} (w_e + 1.71)^2\right) + 172.
+& N_\textrm{freq} \left(1.03 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} + 5.93 N_\textrm{Mvis} (w_e + 1.71)^2\right) + 172.
 \label{eq:casa-computing-cost}%
@@
 \begin{table*}
-\begin{tabular}{lrrrrrrr}
- Survey   & $\lambda$ & FWHM  & Resolution & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. time & B/$\delta \nu$ \\
+\caption{Predicted imaging times for various surveys, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree.
+Predictions are for $w$-projection in CASA, $w$-stacking with WSClean and $w$-stacked 12$\times$5-min snapshots with WSClean, with approximately equal accuracy.}
+\begin{tabular}{lrrrrrrrrrr|rrr}
+ Survey   & $\lambda$ & FWHM  & Beams & Ant & Res & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ & $w$-proj & $w$-stack & $w$-snaps \\
+          &           &       &       &     &     & (km)                       &                 & (s)    & (MHz) &  (kHz) &  &  & \\
 \hline
- GLEAM    &           &       & 2'     & 2.9~km & 5~m & 2~s & 32~MHz/40~kHz\\
- EMU wide & 20 cm  & 1\degree & 10''    & 6~km &  0~m &     & ?/?\\
- MSSS low &  5 m   & 9.8\degree & 100'' & 5~km &      &10~s & 16~MHz/4~khZ\\
- MSSS high&  2 m   & 3.8\degree & 120'' & 5~km &      &10~s & 16~MHz/4~khZ\\
+ GLEAM    &  2 m   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5~m & 2 & 32 & 40     &14h 02m & 1h 52 m & 1h 23m \\
+ EMU      & 20 cm  & 1\degree   & 36   &  36 & 10''  & 6 &  0~m &10 & 300 & 20    &19h 49m &14h 12 m & 14h 13m\\
+ MSSS low &  5 m   & 9.8\degree & 5    &  20 & 100'' & 5 &      &10 & 16 & 16     & 2h 30m & 0h 11 m & 0h 11m\\
+ MSSS high&  2 m   & 3.8\degree & 5    &  40 & 120'' & 5 &      &10 & 16 & 16     & 2h 31m & 0h 38 m & 0h 39m\\
+ VLSS     & 4.1 m  & 14\degree  & 1    &  27 &  80'' &11.1& 0~m &10 & 1.56 & 12.2 & 0h 32m & 0h 12 m & 0h 02m\\
 \end{tabular}
@@
 
+\subsection{Optimal snapshot time}
+The computing time of snapshot imaging can be expressed in either of the $w$-correction techniques.
+Snapshot imaging effectively removes the dependency on ZA from $w_e$.
+Given the snapshot time $\Delta t_\textrm{snapshot}$ and the total time $\Delta t_\textrm{total}$, the total time cost of inversion becomes
+\begin{equation}
+t(\Delta t_\textrm{snapshot}) = \frac{\Delta t_\textrm{total}}{\Delta t_\textrm{snapshot}} t_\textrm{imaging}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
+\end{equation}
+with
+\begin{equation}
+w'_e = \frac{1}{\lambda}\max \Delta z \left(1.0 - \cos(\frac{1}{2}(\textrm{FOV} + \angle t_\textrm{snapshot})\right)
+\end{equation}
+and $N'_\textrm{vis} = \frac{N_\textrm{vis}\Delta t_\textrm{total}}{t_\textrm{snapshot}}$.
+This function can be minimized to find the optimal snapshot time.
+For WSClean with MWA parameters, we find this function does not reach its minimum within 24~h, suggesting that the optimal snapshot time is as long as possible.
+This implies that the change in in performance due to higher $w$-values because of Earth rotation during an individual snapshot is less significant than the cost of the additional FFTs.
+For $w$-projection in CASA we find an optimal snapshot time of 122~seconds.
+
+\subsection{Optimal $w$-project/$w$-stacking combination}
 The $w$-projection and $w$-stacking algorithms can be combined by gridding onto several $w$-layers to which large $w$-terms are corrected after the FFTs, and using a $w$-correcting kernel for making small $w$-corrections before the FFTs.
@@
 \end{equation}
-with $N'_\textrm{vis} = \frac{N_\textrm{vis}\Delta w}{w_e}$ and $w'_e = \Delta w$.
-
-TODO fill in formulae for common surveys, e.g. ESKAP and LOFAR.
+with $N'_\textrm{vis} = \frac{N_\textrm{vis}\Delta w}{w_e}$ and $w'_e = \Delta w.
 

### 2014-02-07 articles/2014/Article-WSClean/img/fit.cpp

@@ -13,9 +13,9 @@
 
 bool fixDW = false, fixConstant = false;
 
-double sinZA(double za)
+double sinZA(double za, double lambda = 1.64, double maxBaselineM = 2900.0, double maxHeightDiffM = 5.5)
 {
-	return (2900.0*sin(za*(M_PI/180.0)) + 5.5*cos(za*(M_PI/180.0)));
+	return (maxBaselineM*sin(za*(M_PI/180.0)) + maxHeightDiffM*cos(za*(M_PI/180.0))) / lambda;
 }
 
 double fovFact(double fov)
@@ -85,6 +85,7 @@ enum FittingFunc
 	NZASqFunc3,
 	WSCleanFunc,
 	WSSCleanFunc,
+	WSSCleanExtrapolatedFunc,
 	CASAFunc
 };
 
@@ -103,18 +104,22 @@ struct FittingInfo
 	Sample *samples;
 };
 
-double eval(FittingFunc func, double za, double nVis, double nPix, double fov, double nChan, double a, double b, double c, double d, double e, double f)
+double eval(FittingFunc func, double za, double nVis, double nPix, double fov, double nChan, double a, double b, double c, double d, double e, double f, double lambda = 1.64, double maxBaselineM = 2900.0, double maxHeightDiffM = 5.5)
 {
 	if(func == WSCleanFunc) {
-		double we = sinZA(za) * fovFact(fov);
+		double we = sinZA(za, lambda, maxBaselineM, maxHeightDiffM) * fovFact(fov);
 		return nChan * (a * (we+b)*nPix*nPix*log2(nPix) + c * nVis/nChan) + d;
 	}
 	else if(func == WSSCleanFunc) {
 		return nChan * (a *nPix*nPix*log2(nPix) + b * nVis/nChan) + c;
 	}
-	else {
-		double we = (sinZA(za) * fovFact(fov) + d);
+	else if(func == CASAFunc) {
+		double we = (sinZA(za, lambda, maxBaselineM, maxHeightDiffM) * fovFact(fov) + d);
 		return nChan * (a * nPix*nPix*log2(nPix) + b * nVis/nChan * we * we) + f;
+	}
+	else {
+		double we = maxHeightDiffM/lambda * fovFact(fov);
[…]

### 2014-02-09 articles/2014/Article-WSClean/wsclean.tex
Better fits

@@
 
+TODO say something about error of analytical function vs measurements.
+
 \begin{table*}
@@
 \begin{tabular}{lrrrrrrrrrr|rrr}
- Survey   & $\lambda$ & FWHM  & Beams & Ant & Res & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ & $w$-proj & $w$-stack & $w$-snaps \\
-          &           &       &       &     &     & (km)                       &                 & (s)    & (MHz) &  (kHz) &  &  & \\
+ Configuration&$\lambda$&FWHM  & Beams & Ant & Res & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ & $w$-proj & $w$-stack & $w$-snaps \\
+          & (m)       &       &       &     &     & (km)                       & (m)             & (s)    & (MHz) &  (kHz) &  &  & \\
 \hline
- GLEAM    &  2 m   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5~m & 2 & 32 & 40     &14h 02m & 1h 52 m & 1h 23m \\
- EMU      & 20 cm  & 1\degree   & 36   &  36 & 10''  & 6 &  0~m &10 & 300 & 20    &19h 49m &14h 12 m & 14h 13m\\
- MSSS low &  5 m   & 9.8\degree & 5    &  20 & 100'' & 5 &      &10 & 16 & 16     & 2h 30m & 0h 11 m & 0h 11m\\
- MSSS high&  2 m   & 3.8\degree & 5    &  40 & 120'' & 5 &      &10 & 16 & 16     & 2h 31m & 0h 38 m & 0h 39m\\
- VLSS     & 4.1 m  & 14\degree  & 1    &  27 &  80'' &11.1& 0~m &10 & 1.56 & 12.2 & 0h 32m & 0h 12 m & 0h 02m\\
+ GLEAM    &  2   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5 & 2 & 32 & 40     &14h 02m & 1h 52 m & 1h 23m \\
+ EMU      & 0.2  & 1\degree   & 30   &  36 & 10''  & 6 &  0 &10 & 300 & 20    &19h 49m &14h 12 m & 14h 13m\\
+ MSSS low &  5   & 9.8\degree & 5    &  20 & 100'' & 5 &    &10 & 16 & 16     & 2h 30m & 0h 11 m & 0h 11m\\
+ MSSS high&  2   & 3.8\degree & 5    &  40 & 120'' & 5 &    &10 & 16 & 16     & 2h 31m & 0h 38 m & 0h 39m\\
+ AARTFAAC &  5   & 45\degree  & 1    & 288 &  20'  & 0.3& 1 & 1 &  7 & 24     & & & \\
+ VLSS     & 4.1  & 14\degree  & 1    &  27 &  80'' &11.1& 0 &10 & 1.56 & 12.2 & 0h 32m & 0h 12 m & 0h 02m\\
+ SKA-P1 AA inner&3& 1.2\degree& 480  &  35 & 140'' &2.5& 2 &1.2 & 380 & 1     &       &         &      \\
+ SKA-P1 AA mid&3 & 1.2\degree & 480  &  50 & 3.5'' &100& 50 &1.2 & 380 & 1     &       &         &      \\
 \end{tabular}
@@
 
-\subsection{Optimal snapshot time}
-The computing time of snapshot imaging can be expressed in either of the $w$-correction techniques.
+\subsection{Optimal snapshot duration}
+Snapshot imaging can in theory be implemented with either $w$-projection or $w$-stacking.
+Using the derived formula, we can estimate the cost of making snapshots with both techniques.
 Snapshot imaging effectively removes the dependency on ZA from $w_e$.
-Given the snapshot time $\Delta t_\textrm{snapshot}$ and the total time $\Delta t_\textrm{total}$, the total time cost of inversion becomes
-\begin{equation}
-t(\Delta t_\textrm{snapshot}) = \frac{\Delta t_\textrm{total}}{\Delta t_\textrm{snapshot}} t_\textrm{imaging}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
+Given the snapshot duration $\Delta \tau_\textrm{snapshot}$ and the total integration time $\Delta \tau_\textrm{total}$, the total time cost of inversion becomes
+\begin{equation} \label{eq:snapshot-cost}
+t_\textrm{snapshot}(\Delta \tau_\textrm{snapshot}) = \frac{\Delta \tau_\textrm{total}}{\Delta \tau_\textrm{snapshot}} t_\textrm{imaging}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
 \end{equation}
@@
 \begin{equation}
-w'_e = \frac{1}{\lambda}\max \Delta z \left(1.0 - \cos(\frac{1}{2}(\textrm{FOV} + \angle t_\textrm{snapshot})\right)
+w'_e = \frac{1}{\lambda}\max \Delta z \left(1.0 - \cos(\frac{1}{2}(\textrm{FOV} + \omega_E \tau_\textrm{snapshot})\right),
 \end{equation}
-and $N'_\textrm{vis} = \frac{N_\textrm{vis}\Delta t_\textrm{total}}{t_\textrm{snapshot}}$.
-This function can be minimized to find the optimal snapshot time.
-For WSClean with MWA parameters, we find this function does not reach its minimum within 24~h, suggesting that the optimal snapshot time is as long as possible.
-This implies that the change in in performance due to higher $w$-values because of Earth rotation during an individual snapshot is less significant than the cost of the additional FFTs.
-For $w$-projection in CASA we find an optimal snapshot time of 122~seconds.
+$\omega_E$ the rotational speed of the Earth and $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta \tau_\textrm{total}}{\tau_\textrm{snapshot}}$.
+We have excluded the cost of phase shifting the visibilities and gridding.
+The cost for regridding with simple nearest neighbour or bilinair interpolation is indeed negligable, although more accurate interpolation (e.g.
+Lanczos interpolation) can be expensive.
+Also, our current implementation of the phase-changing program does take a non-negligable time, but this implementation is not optimized and can in theory be implemented in a preprocessing pipeline or on the fly during imaging.
 
-\subsection{Optimal $w$-project/$w$-stacking combination}
-The $w$-projection and $w$-stacking algorithms can be combined by gridding onto several $w$-layers to which large $w$-terms are corrected after the FFTs, and using a $w$-correcting kernel for making small $w$-corrections before the FFTs.
+Function $t_\textrm{snapshot}$ can be minimized to find the optimal snapshot duration.
+For WSClean with MWA parameters, we find that this function decreases but no minimum is reached for $\Delta \tau_\textrm{snapshot}<24$~h.
+Therefore, from a performance perspective, the snapshot duration should be as large as possible.
+For $w$-projection in CASA we find an optimal snapshot duration of 122~s for MWA observations.
+
+\subsection{Combination of $w$-projection and $w$-stacking}
+To combine the $w$-projection and $w$-stacking algorithms, small $w$-corrections are made before the FFTs using a $w$-correcting kernel and large $w$-terms are corrected after the FFTs by gridding onto several $w$-layers.
 This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory.
@@
 \end{equation}
-with $N'_\textrm{vis} = \frac{N_\textrm{vis}\Delta w}{w_e}$ and $w'_e = \Delta w.
+with $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta w}{w_e}$ and $w'_e = \Delta w$.
 
@@
 
+TODO discuss fact that w-projection needs layers for multi-threading.
+
 \section*{Acknowledgments}
@@
 \includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
-\caption{Effect of antialiasing kernel size on performance.}
+\caption{Effect of antialiasing-kernel size on performance.}
 \label{fig:timing-kernelsize}

### 2014-02-09 articles/2014/Article-WSClean/img/fit.cpp

@@ -108,18 +108,18 @@ double eval(FittingFunc func, double za, double nVis, double nPix, double fov, d
 {
 	if(func == WSCleanFunc) {
 		double we = sinZA(za, lambda, maxBaselineM, maxHeightDiffM) * fovFact(fov);
-		return nChan * (a * (we+b)*nPix*nPix*log2(nPix) + c * nVis/nChan) + d;
+		return nChan * (a * (we+b)*nPix*nPix*log2(nPix) + d) + c * nVis;
 	}
 	else if(func == WSSCleanFunc) {
-		return nChan * (a *nPix*nPix*log2(nPix) + b * nVis/nChan) + c;
+		return nChan * a *nPix*nPix*log2(nPix) + b * nVis + c;
 	}
 	else if(func == CASAFunc) {
-		double we = (sinZA(za, lambda, maxBaselineM, maxHeightDiffM) * fovFact(fov) + d);
-		return nChan * (a * nPix*nPix*log2(nPix) + b * nVis/nChan * we * we) + f;
+		double we = sinZA(za, lambda, maxBaselineM, maxHeightDiffM) * fovFact(fov);
+		return nChan * a * nPix*nPix*log2(nPix) + b * nVis * (we * we + d) + f;
 	}
 	else {
 		double we = maxHeightDiffM/lambda * fovFact(fov);
-		return nChan * (a * (we+b)*nPix*nPix*log2(nPix) + c * nVis/nChan) + d;
+		return nChan * a * (we+b)*nPix*nPix*log2(nPix) + c * nVis + d;
 	}
 }
 
@@ -134,47 +134,6 @@ double eval(FittingFunc func, double za, double nVis, double nPix, double fov, d
 		default:
 			return 0.0;
 	}
-}
-
-void evalSurveyConfig(const std::string& desc, double lambda, double fwhm, size_t beams, double intTimeSec, double bandWidth, double freqRes, size_t antennas, double angRes, double maxBaseline, double maxDiffHeight, FittingFunc func, double a, double b, double c, double d, double e, double f)
-{
-	const double duration = 10.0 * 60.0;
-	const double za1 = 10.0, za2 = 20.0;
[…]

### 2014-02-10 articles/2014/Article-WSClean/wsclean.tex
Updating snapshot example image with vectorized one

@@
 
-There are several methods to deal with the w-term during imaging: facetting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, w-projection \citep{wprojection-cornwell}, w-stacking\footnote{TODO when/where was this first mentioned in literature?} \citep{widefield-imaging-ska-cornwell} and w-snapshots \citep{widefield-imaging-ska-cornwell}.
+There are several methods to deal with the w-term during imaging: facetting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, w-projection \citep{wprojection-cornwell}, w-stacking \citep{widefield-imaging-ska-cornwell} and w-snapshots \citep{widefield-imaging-ska-cornwell}.
 The w-snapshots algorithm is a hybrid between snapshot imaging and one of the other w-term correcting techniques, most efficiently w-projection or w-stacking.
@@
 Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}).
-At the MWA, some of the data is reduced with the w-projection algorithm implementation of the Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}).
-This is an attractive option because of its many features, high quality and free availability.
+The Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}) has an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution.
 However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's w-projection algorithm.
+Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.
+
 Another option for reducing MWA data is the Real-Time System (RTS), which has been designed as an efficient calibrating and imaging pipeline specifically for MWA data \citep{rts-mwa}.
@@
 One such DDE is the effect of the ionosphere.
-For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure (TODO cite).
+For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure \citep{lonsdale-calibration-approaches-ionosphere}.
 This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data.
@@
 Once the Square-Kilometre Array (SKA) begins its operation, the required computational power for wide-field imaging will become an even bigger challenge.
-In \citet{widefield-imaging-ska-cornwell}, it is shown that in theory the w-snapshots algorithm is the most efficient approach for the SKA.
+\citet{widefield-imaging-ska-cornwell} shows that in theory the w-snapshots algorithm is the most efficient approach for the SKA.
 
-In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's w-projection implementation.
+In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's w-projection implementation, and is able to succesfully image any MWA observation.
 To obtain the increase in speed, the implementation uses the w-stacking method for correcting the w-terms, optionally combined with w-snapshot imaging.
 We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''.
-This work improves the field of wide-field imaging in the following ways:
-\begin{itemize}
- \item The performance of w-stacking is compared to w-projection.
- \item Parameters that affect the w-snapshots algorithm are analysed.
- \item Details are described of how to efficiently implement the w-stacking and w-snapshots algorithms.
- \item A new implementation is released, which we believe to be the fastest generic widefield imager available.
-\end{itemize}
+Our new imaging implementation, which we believe to be the fastest generic widefield imager available, is publicly released\footnote{TODO add URL}.
 
-The w-stacking algorithm is described in Sect.~\ref{sec:wstacking}.
-Details of implementing the w-stacking and w-snapshots algorithms are described in Sect.~\ref{sec:implementation}.
+This paper is structured as follows: The $w$-stacking algorithm is described in Sect.~\ref{sec:wstacking}.
+Details of implementing the $w$-stacking and $w$-snapshots algorithms are described in Sect.~\ref{sec:implementation}.
 The performance and accuracy will be analysed in Sect.~\ref{sec:analysis}.
@@
 \section{The w-stacking technique} \label{sec:wstacking}
-In this section, we will describe the w-stacking algorithm from a mathematical point of view.
+In this section, we will describe the $w$-stacking algorithm from a mathematical point of view.
 
@@
 We will not discuss calibration, but assume $V$ has been calibrated before imaging.
-In the case of a polarized measurement, the symbols $V$, $A$ and $I$ are $2\times 2$ Jones matrices, but without loss of generality we will ignore polarization and treat inversion as a scalar problem.
+In the case of a polarized measurement, the symbols $V$, $A$ and $I$ are $2\times 2$ matrices, but without loss of generality we will ignore polarization and treat inversion as a scalar problem.
 Imaging consists of inverting Eq.~\eqref{eq:visibility-function}, i.e., to find $I'$ from $V$.
@@
 \end{equation}
-This suggest that a uniform discretization in $w$ is optimal TODO in contrast to some article?
--- could be explained by having more samples at smaller $w$s.
+This suggest that a uniform discretization in $w$ is optimal (TODO this is in contrast to what Cornwell 2008 says -- mailed Cornwell to ask why).
 From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
@@
 For the MWA, a typical value of $w_e - w_0$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$.
-[TODO table of arrays with $w$ values at zenith/more angles/more freq?] For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$.
-This implies that, in the case of the MWA, tens of $w$-layers are required at zenith and hundreds at lower elevations.
+For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$.
+This implies that tens of $w$-layers are required at zenith and hundreds at lower elevations.
 The number of $w$-layers has a large effect on the performance of the $w$-stacking algorithm, and will be discussed further in the next sections.
@@
 With the purpose of testing new algorithms for the imaging of MWA data, we have written a new imager around the $w$-stacking algorithm.
-The imager is not specialized for the MWA, and has been succesfully used for other telescopes (TODO menton more).
+As will be shown in Sect.~\ref{sec:application-to-nonmwa}, the imager is not specialized for the MWA, and has been succesfully used for imaging data from other telescopes.
 
@@
 No beam correction was applied.
-TODO time difference and more discussion, replace with deeper images.}
+TODO time difference and more discussion.}
 \label{fig:vela-projection-example}
@@
 
-Table~\ref{tbl:accuracy-measurements} lists the measured standard errors of the source brightnesses as detected by Aegean and the RMS in the residual image.
+Table~\ref{tbl:accuracy-measurements} lists the measured RMS in the residual image and the standard errors of the source brightnesses as detected by Aegean.
 In all tests, WSClean performs slightly more accurate.
@@
 
+\section{Application of WSClean to other telescopes} \label{sec:application-to-nonmwa}
+TODO
+
 \section{Discussion \& conclusions} \label{sec:discussion}
@@
 TODO discuss fact that w-projection needs layers for multi-threading.
+
+TODO discuss optimalisations with GPUs
 

### 2014-02-10 articles/2014/Article-WSClean/wsclean.tex
Work on article

@@
 No beam correction was applied.
-TODO time difference and more discussion.}
+Computing cost was 60 min for the normal projected image and 41 min for the recentred image.}
 \label{fig:vela-projection-example}
@@
 Cleaning an image that is in a different projection yields slightly different results, but qualitativily the two images are clearly of equal accuracy.
+The wall-clock time for imaging is 60 min and 41 min for respectively normal projection and the recentred image.
+Of the 41 min, 3 min is spent on phase rotating the visibilities.
 
@@
 Therefore, we chose to implement the former method.
+
+\subsection{Beam correction}
+Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix.
+Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}.
+When assuming all individual antennas of the tiles are working properly, all tiles have the same beam.
+The beam can then be corrected in image space for all tiles at once with
+\begin{equation}\label{eq:beam-correction}
+ I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
+\end{equation}
+where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose.
+Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such.
+However, this is not correct, because $X$ is defined as being parallel to the celestial equator.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters.
+The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry.
+To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images.
+Each polarization requires a separate run of the $w$-stacking algorithm.
+Once these six images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}.
+This method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam.
+To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm excessively unless an a-projection kernel is used.
+The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.
 
@@
 
-Panel~\ref{fig:timing-channels} shows performance versus the output frequency resolution.
-The cost of imaging at multiple frequencies is the cost of making an images times the number of frequencies.
-However, each separate imaging contains a part of the total number of visibilities.
-As shown before, FFTs become the dominant cost for small visibility sets, and performance and number of frequency images therefore have a linear relation, which is clearly visible from the plot.
+Panel~\ref{fig:timing-channels} shows performance versus the spectral output resolution.
+The cost of imaging multiple frequencies is the cost of making an image from fewer visibilities times the number of frequencies.
+FFTs become the dominant cost for such small visibility sets, and the number of imaged frequencies affect performance linearly.
 
-It is important to point out that all these measurement were performed by imaging a single polarization.
+%It is important to point out that all these measurement were performed by imaging a single polarization.
 Making full-polarization images with WSClean currently requires four runs.
@@
 \begin{table*}
-\caption{Predicted imaging times for various surveys, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree.
+\caption{Predicted computational cost for various configurations, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree.
 Predictions are for $w$-projection in CASA, $w$-stacking with WSClean and $w$-stacked 12$\times$5-min snapshots with WSClean, with approximately equal accuracy.}
@@
 \hline
- GLEAM    &  2   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5 & 2 & 32 & 40     &14h 02m & 1h 52 m & 1h 23m \\
- EMU      & 0.2  & 1\degree   & 30   &  36 & 10''  & 6 &  0 &10 & 300 & 20    &19h 49m &14h 12 m & 14h 13m\\
- MSSS low &  5   & 9.8\degree & 5    &  20 & 100'' & 5 &    &10 & 16 & 16     & 2h 30m & 0h 11 m & 0h 11m\\
- MSSS high&  2   & 3.8\degree & 5    &  40 & 120'' & 5 &    &10 & 16 & 16     & 2h 31m & 0h 38 m & 0h 39m\\
- AARTFAAC &  5   & 45\degree  & 1    & 288 &  20'  & 0.3& 1 & 1 &  7 & 24     & & & \\
- VLSS     & 4.1  & 14\degree  & 1    &  27 &  80'' &11.1& 0 &10 & 1.56 & 12.2 & 0h 32m & 0h 12 m & 0h 02m\\
- SKA-P1 AA inner&3& 1.2\degree& 480  &  35 & 140'' &2.5& 2 &1.2 & 380 & 1     &       &         &      \\
- SKA-P1 AA mid&3 & 1.2\degree & 480  &  50 & 3.5'' &100& 50 &1.2 & 380 & 1     &       &         &      \\
+ GLEAM    &  2   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5 & 2 & 32 & 40     &14h 02m & 1h 52m & 1h 23m \\
+ EMU      & 0.2  & 1\degree   & 30   &  36 & 10''  & 6 &  0 &10 & 300 & 20    &19h 49m &14h 12m & 14h 13m\\
+ MSSS low &  5   & 9.8\degree & 5    &  20 & 100'' & 5 &    &10 & 16 & 16     & 2h 30m & 0h 11m & 0h 11m\\
+ MSSS high&  2   & 3.8\degree & 5    &  40 & 120'' & 5 &    &10 & 16 & 16     & 2h 31m & 0h 38m & 0h 39m\\
+ AARTFAAC &  5   & 45\degree  & 1    & 288 &  20'  & 0.3& 1 & 1 &  7 & 24     & 10h 47m & 4h 58m & 4h 58m \\
+ VLSS     & 4.1  & 14\degree  & 1    &  27 &  80'' &11.1& 0 &10 & 1.56 & 12.2 & 0h 32m & 0h 12m & 0h 02m\\
+ SKA-P1 AA inner&3& 1.2\degree& 480  &  35 & 140'' &2.5& 2 &1.2 & 380 & 1     & 3218 d & 1865 d  & 1865 d \\
+ SKA-P1 AA mid&3 & 1.2\degree & 480  &  50 & 3.5'' &100& 50 &1.2 & 380 & 1     & 6658 d & 3843 d & 3844 d \\
 \end{tabular}
@@
 with $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta w}{w_e}$ and $w'_e = \Delta w$.
-
-\section{Application of WSClean to the MWA}
-For technical reasons, the MWA correlator splits observations in data products of typically a few minutes.
-This simplifies $w$-snapshot imaging, which can be done by phase rotating each snapshot to zenith with a possibly additional phase-centre shift.
-
-TODO: calculate when to use w-snapshots: discuss at what is the optimal time to split measurement sets into when using recentering technique.
-TODO: calculate optimum splitting time?
-
-\subsection{Beam correction}
-Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix.
-Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}.
-When assuming all individual antennas of the tiles are working properly, all tiles have the same beam.
-The beam can then be corrected in image space for all tiles at once with
-\begin{equation}\label{eq:beam-correction}
- I(l, m) = J(l,m)^{-1} I'(l, m) \left(J(l,m)^*\right)^{-1},
-\end{equation}
-where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose.
-Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such.
-However, this is not correct, because $X$ is defined as being parallel to the celestial equator.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters.
-The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry.
-To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images.
-Each polarization requires a separate run of the $w$-stacking algorithm.
-Once these six images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}.
-This method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam.
-To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm.
-The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.
 

### 2014-02-10 articles/2014/Article-WSClean/img/fit.cpp

@@ -11,7 +11,7 @@
 #include <gsl/gsl_vector.h>
 #include <gsl/gsl_multifit_nlin.h>
 
-bool fixDW = false, fixConstant = false;
+bool fixDW = false, fixConstant = false, fixZAOffset = false;
 
 double sinZA(double za, double lambda = 1.64, double maxBaselineM = 2900.0, double maxHeightDiffM = 5.5)
 {
@@ -114,7 +114,7 @@ double eval(FittingFunc func, double za, double nVis, double nPix, double fov, d
 		return nChan * a *nPix*nPix*log2(nPix) + b * nVis + c;
 	}
 	else if(func == CASAFunc) {
-		double we = sinZA(za, lambda, maxBaselineM, maxHeightDiffM) * fovFact(fov);
+		double we = (sinZA(za, lambda, maxBaselineM, maxHeightDiffM) + e) * fovFact(fov);
 		return nChan * a * nPix*nPix*log2(nPix) + b * nVis * (we * we + d) + f;
 	}
 	else {
@@ -145,8 +145,8 @@ double evalWE(FittingFunc func, double we, double nVis, double nPix, double nCha
 		return nChan * (a *nPix*nPix*log2(nPix) + b * nVis/nChan) + c;
 	}
 	else {
-		we = we + d;
-		return nChan * (a * nPix*nPix*log2(nPix) + b * nVis/nChan * we * we) + f;
+		we = we + e;
+		return nChan * a * nPix*nPix*log2(nPix) + b * nVis * (we * we + d) + f;
 	}
 }
 
@@ -253,13 +253,14 @@ int fitFuncDeriv(const gsl_vector *xvec, void *data, gsl_matrix *J)
 				dc = fixConstant ? 0.0 : 1.0;
 			}
 			else { // nChan * (a * nPix*nPix*log2(nPix) + b * nVis/nChan * (we * we + d)) + f;
-				double we = sinZA(za) * fovFact(fov);
+				double e = gsl_vector_get(xvec, 4);
+				double we = (sinZA(za) + e) * fovFact(fov);
 				da = nChan * nPix*nPix*log2(nPix);
 				db = nVis * 2.0 * (we * we + d);
 				dc = 0.0;
 //				dd = fixDW ? 0.0 : nChan * b * 2.0 * (d + sinZA(za) * fovFact(fov));
[…]

### 2014-02-11 articles/2014/Article-WSClean/wsclean.tex
Work on article

@@
 \end{equation}
-This suggest that a uniform discretization in $w$ is optimal (TODO this is in contrast to what Cornwell 2008 says -- mailed Cornwell to ask why).
+This suggest that a uniform discretization in $w$ is optimal (TODO this is in contrast to what Cornwell 2008 says -- Cornwell mailed me he made a mistake...
+Do we need to explicitly point that out here :-/ ?).
 From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
@@
 \hline
-\multicolumn{3}{l}{\textit{Zenith angle 0\degree (minimum w-layers/planes)}} \\
+\multicolumn{3}{l}{\textit{Zenith angle 0\degree (6 w-layers/planes)}} \\
 \hline
@@
 \hline
-Source flux standard error & 1.75\% & 2.92\% & 2.41 \%\\
-RMS in residual image & 0.90 mJy & 1.05 mJy & 1.07 mJy \\
+Source flux standard error & 1.75\% & 1.40\% & 2.41 \%\\
+RMS in residual image & 0.90 mJy & 1.03 mJy & 1.07 mJy \\
 Computational time & 15.3 min & 6.6 min & 178.2 min \\
-& & recentreTODO & \\
 \hline\hline
@@
 
-To asses the accuracy of WSClean and CASA, we have simulated a MWA observation without system noise, but with hundred sources of 1~Jy in a 20\degree diameter area.
+To assess the accuracy of WSClean and CASA, we simulate a MWA observation with hundred sources of 1~Jy in a 20\degree diameter area, without adding system noise.
 A unitary primary beam is assumed.
-We imaged the simulated set both with WSClean and CASA, using Cotton-Schwab cleaning to a threshold of 10~mJy.
+We image the simulated set with WSClean and CASA using Cotton-Schwab cleaning to a threshold of 10~mJy.
 The two imagers calculate slightly different restoring (synthesized) beams, hence to avoid bias the restoring beams are fixed.
@@
 At zenith, CASA produces a residual RMS twice as high as WSClean.
-This is caused by the fact that we keep the number of w-projection planes in w-projection equal to the number of w-layers in w-stacking, resulting in only 12 $w$-planes at zenith.
-This evidently has a stronger effect on the w-projection algorithm.
-However, the computational performance of the w-projection algorithm is hardly affected by the number of w-projection planes, and in practical situations one would always use more w-projection planes\footnote{In the zenith case CASA actually issues a warning to use more $w$-projection planes.}.
-When 128 w-projection planes are used in CASA, its residual RMS is equal to WSClean with 12 $w$-layers.
-We do not know why this parameter needs to be higher in CASA to reach the same RMS.
-In theory it should be enough, since Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlayers}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree.
+This is caused by the fact that we keep the number of $w$-projection planes in w-projection equal to the number of w-layers in w-stacking, resulting in only 12 $w$-planes at zenith.
+This evidently has a stronger effect on the $w$-projection algorithm.
+However, the computational performance of the $w$-projection algorithm is hardly affected by the number of $w$-projection planes, and in practical situations one would always use more w-projection planes\footnote{In the zenith case CASA issues a warning to use more $w$-projection planes.}.
+When 128 $w$-projection planes are used in CASA, its residual RMS is equal to WSClean with 12 $w$-layers, but the flux density measurements are less accurate.
+We do not know why this parameter needs to be higher in CASA to reach the same RMS.\footnote{TODO: do I want to find out this is because of invalid sqrt(w) gridding?} Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlayers}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree.
 Also unexpected is that the source flux density becomes worse by increasing the number of planes.
@@
 In the $\textrm{ZA}=10\degree$ case, CASA is slightly less accurate.
-We have noticed this consistently at higher zenith angles, and it can actually be observed in the produced images, as shown in Fig.~\ref{fig:residuals}.
-TODO the following might be caused by a bug; check: We have also made an image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}.
-Its accuracy is lower compared to the normal projection.
-This is due to the fact that the sources are further away from the phase centre, and sources have therefore on average higher $(l,m)$-values, thereby increasing the average $w$-correction.
-Nevertheless, the reprojection technique needs on average fewer $w$-layers to reach the same level of accuracy.
+The extra noise can be observed in the produced images, as shown in Fig.~\ref{fig:residuals}.
+Also made is an image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}.
+Its accuracy is higher compared to the normal projection, but its residual noise is also higher.
+The reprojection technique needs on average fewer $w$-layers to reach the same level of accuracy.
 
@@
 
-Many parameters affect the accuracy more than the observed differences between the methods.
-These include the number of major iterations, $w$-planes/layers, restoring beam, image size, weighting, kernel size and grid oversampling.
-Therefore, the differences in accuracy are not very significant.
-The results do show that synthesis imaging using common imaging parameters results in an approximate flux-density standard error of at least $\sim1\%$.
-
 \begin{figure*}%
-\begin{subfigure}{.36\linewidth}%
+\begin{subfigure}{.5\linewidth}%
 \includegraphics[width=\linewidth]{img/benchmark-nsamples/nsamples}%
@@
 \end{subfigure}%
-\begin{subfigure}{.36\linewidth}%
+\hspace{-.05\linewidth}\begin{subfigure}{.5\linewidth}%
 \includegraphics[width=\linewidth]{img/benchmark-zenith-angle/za}%
@@
 \end{subfigure}\\
-\begin{subfigure}{.36\linewidth}%
+\begin{subfigure}{.5\linewidth}%
 \includegraphics[width=\linewidth]{img/benchmark-resolution/resolution}
@@
 \end{subfigure}%
-\hspace{-.05\linewidth}\begin{subfigure}{.36\linewidth}%
+\hspace{-.05\linewidth}\begin{subfigure}{.5\linewidth}%
 \includegraphics[width=\linewidth]{img/benchmark-channels/channels}
 \caption{}\label{fig:timing-channels}%
+\end{subfigure}\\
+\begin{subfigure}{.5\linewidth}%
+\includegraphics[width=\linewidth]{img/benchmark-fov/fov}
+\caption{}\label{fig:timing-fov}%
 \end{subfigure}%
-\hspace{-.05\linewidth}\begin{subfigure}{.36\linewidth}%
-\includegraphics[width=\linewidth]{img/benchmark-fov/fov}
-\caption{}
-\label{fig:timing-fov}
-\end{subfigure}%
-\caption{Imaging performance as a function of several configuration parameters. Error bars show 5$\sigma$ level. }\label{fig:timings}
+\caption{Imaging performance as a function of several parameters. Error bars show 5$\sigma$ level. Unfinished lines indicate the imager could not run succesfully.}\label{fig:timings}
 \end{figure*}
@@
 However, the additional time required to rephase and regrid the measurement set makes it only worthwhile for images $\ge3072$ pixels and $\textrm{ZA}\ge15\degree$.
-TODO: this is quite heavily depending on the speed of \texttt{chgcentre}, and that might still be improvable...
+This is dependent on the speed of rotating the phase centre and regridding.
+Our program to change the phase centre is currently not multi-threaded, so its performance might be improvable.
 
@@
 
+The cost of spectral imaging, i.e.
+imaging multiple frequencies, is the cost of making an image from fewer visibilities times the number of desired frequencies.
 Panel~\ref{fig:timing-channels} shows performance versus the spectral output resolution.
-The cost of imaging multiple frequencies is the cost of making an image from fewer visibilities times the number of frequencies.
-FFTs become the dominant cost for such small visibility sets, and the number of imaged frequencies affect performance linearly.
+As can be expected from the previous results, FFTs become the dominant cost for such small visibility sets, and the number of imaged frequencies affect performance linearly.
+
+As can be seen in Panel~\ref{fig:timing-fov}, the field of view (FOV) has a large affect on performance when imaging off-zenith.
+This plot was created by varying the size of the pixel in the output image, such that the number of pixels in the image did not change.
+The FOV is calculated as angle between the left and right most pixel in the image.
+CASA is not able to make off-zenith images larger than $\sim30$\degree, and its imaging cost increases much more rapidly compared to the cost of WSClean, which implies that gridding is the major cost in this scenario.
+The image recentering technique is clearly beneficial for FOVs larger than $\sim30$\degree, and can even make a difference of an order of magnitude at very large FOVs of $\sim90$\degree.
 
@@
 Measurements are weighted with the inverse standard deviation.
-The single outlying measurement of CASA in Fig.~\ref{fig:timing-nsamples} is removed.
+The single outlying measurement for $N_\textrm{pix}=1792$ with CASA in Fig.~\ref{fig:timing-resolution} is removed.
 
-For the WSClean time cost function $t_\textrm{WSClean}$ we find 
+For the WSClean time-cost function $t_\textrm{WSClean}$ we find 
 \begin{align}\label{eq:wsclean-computing-cost}
@@
 where $\max |\Delta \mathbf{x}|$ is the maximum baseline length, $\max \Delta z$ is the maximum height difference between antennas and $\lambda$ is the wavelength.
-
-TODO say something about error of analytical function vs measurements.
-
-\begin{table*}
-\caption{Predicted computational cost for various configurations, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree.
-Predictions are for $w$-projection in CASA, $w$-stacking with WSClean and $w$-stacked 12$\times$5-min snapshots with WSClean, with approximately equal accuracy.}
-\begin{tabular}{lrrrrrrrrrr|rrr}
- Configuration&$\lambda$&FWHM  & Beams & Ant & Res & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ & $w$-proj & $w$-stack & $w$-snaps \\
-          & (m)       &       &       &     &     & (km)                       & (m)             & (s)    & (MHz) &  (kHz) &  &  & \\
-\hline
- GLEAM    &  2   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5 & 2 & 32 & 40     &14h 02m & 1h 52m & 1h 23m \\
- EMU      & 0.2  & 1\degree   & 30   &  36 & 10''  & 6 &  0 &10 & 300 & 20    &19h 49m &14h 12m & 14h 13m\\
- MSSS low &  5   & 9.8\degree & 5    &  20 & 100'' & 5 &    &10 & 16 & 16     & 2h 30m & 0h 11m & 0h 11m\\
- MSSS high&  2   & 3.8\degree & 5    &  40 & 120'' & 5 &    &10 & 16 & 16     & 2h 31m & 0h 38m & 0h 39m\\
- AARTFAAC &  5   & 45\degree  & 1    & 288 &  20'  & 0.3& 1 & 1 &  7 & 24     & 10h 47m & 4h 58m & 4h 58m \\
- VLSS     & 4.1  & 14\degree  & 1    &  27 &  80'' &11.1& 0 &10 & 1.56 & 12.2 & 0h 32m & 0h 12m & 0h 02m\\
- SKA-P1 AA inner&3& 1.2\degree& 480  &  35 & 140'' &2.5& 2 &1.2 & 380 & 1     & 3218 d & 1865 d  & 1865 d \\
- SKA-P1 AA mid&3 & 1.2\degree & 480  &  50 & 3.5'' &100& 50 &1.2 & 380 & 1     & 6658 d & 3843 d & 3844 d \\
-\end{tabular}
-\end{table*}
+These functions follow the trend of the measurements well, with an absolute error of 14.7\% and 20.7\% for respectively the WSClean and CASA functions.
 
@@
 
+\begin{table*}
+\caption{Predicted computational cost for various configurations, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree.
+Predictions are for $w$-projection in CASA, $w$-stacking with WSClean and $w$-stacked snapshots with WSClean and optimal snapshot duration, all with approximately equal accuracy.}
+\begin{tabular}{lrrrrrrrrrr|rrr}
+ Configuration&$\lambda$& FOV  & Beams & Ant & Res & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ & $w$-proj & $w$-stack & $w$-snaps \\
+          & (m)       & (FWHM)&       &     &     & (km)                       & (m)             & (s)    & (MHz) &  (kHz) &  &  & \\
+\hline
+ GLEAM    &  2   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5 & 2 & 32 & 40     &11h 20m & 1h 49m & 1h 23m \\
+ EMU      & 0.2  & 1\degree   & 30   &  36 & 10''  & 6 &  0 &10 & 300 & 20    &23h 52m &11h 52m & 11h 49m\\
+ MSSS low &  5   & 9.8\degree & 5    &  20 & 100'' & 5 &    &10 & 16 & 16     & 0h 53m & 0h 12m & 0h 11m\\
+ MSSS high&  2   & 3.8\degree & 5    &  40 & 120'' & 5 &    &10 & 16 & 16     & 1h 41m & 0h 39m & 0h 39m\\
+ AARTFAAC &  5   & 45\degree  & 1    & 288 &  20'  & 0.3& 1 & 1 &  7 & 24     & 12h 01m & 4h 58m & 4h 58m \\
+ VLSS     & 4.1  & 14\degree  & 1    &  27 &  80'' &11.1& 0 &10 & 1.56 & 12.2 & 0h 10m & 0h 11m & 0h 01m\\
+ SKA-P1 AA inner&3& 1.2\degree& 480  &  35 & 140'' &2.5& 2 &1.2 & 380 & 1     & 3219 d & 1865 d  & 1865 d \\
+ SKA-P1 AA mid&3 & 1.2\degree & 480  &  50 & 3.5'' &100& 50 &1.2 & 380 & 1     &  6740 d & 3845 d & 3844 d \\
+\end{tabular}
+\end{table*}
+
+\subsection{Estimated computing cost for other telescopes} \label{sec:application-to-nonmwa}
+We use the derived functions to estimate the computational cost of the algorithms for configurations of several telescopes.
+We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multifrequency Snapshot Sky Survey (MSSS)\footnote{See \url{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; the Amsterdam—ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \url{http://www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); and the low-frequency phase~1 aperture arrays of the Square-Kilometre Array, with inner (2.5~km) and mid (2.5~km) configurations \citep{ska-phase1-2010}.
+In multi-beam configurations such as the SKA and EMU configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams.
+
 \subsection{Combination of $w$-projection and $w$-stacking}
@@
 with $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta w}{w_e}$ and $w'_e = \Delta w$.
-
-\section{Application of WSClean to other telescopes} \label{sec:application-to-nonmwa}
-TODO
 
@@
 TODO discuss optimalisations with GPUs
+
+TODO The results do show that synthesis imaging using common imaging parameters results in an approximate flux-density standard error of at least $\sim1\%$.
 

### 2014-02-11 articles/2014/Article-WSClean/img/fit.cpp

@@ -594,6 +594,7 @@ double stddev(double sum, double sumSq, double sumweight)
 void calculateStdError(std::vector<Sample>& samples, FittingFunc func, double a, double b, double c, double d, double e, double f)
 {
 	double weights = 0.0, weightedSum = 0.0, unweightedSum = 0.0, weightedSumSq = 0.0, unweightedSumSq = 0.0;
+	double absMeanFrac = 0.0;
 	for(size_t i=0; i!=samples.size(); ++i)
 	{
 		const Sample& s = samples[i];
@@ -604,9 +605,11 @@ void calculateStdError(std::vector<Sample>& samples, FittingFunc func, double a,
 		weightedSumSq += dwy*dwy;
 		unweightedSum += dy;
 		unweightedSumSq += dy*dy;
+		absMeanFrac += fabs(dy) / fabs(y);
 	}
 	std::cout << "Mean error: " << weightedSum/weights << "  standard error: " << stddev(weightedSum, weightedSumSq, weights) << '\n'
 		<< "Unweighted mean: " << unweightedSum/samples.size() << " unweighted stderr: " << stddev(unweightedSum, unweightedSumSq, samples.size()) << '\n';
+	std::cout << "Mean absolute error: " << 100.0*absMeanFrac / samples.size() << "%\n";
 }
 
 void printDuplicates(const std::vector<Sample>& samples)
@@ -679,17 +682,18 @@ void optimalDeltaW(const std::vector<Sample>& samples, FittingFunc func, double 
 	std::cout << ", mean dw=" << sumDW/count << '\n';
 }
 
-void optimalDeltaT(FittingFunc func, const std::string& filename, double a, double b, double c, double d, double e, double f, double za, double nVis, double nPix, double fov, double nChan, bool zeroConstant, double& dt)
+void optimalDeltaT(FittingFunc func, const std::string& filename, double a, double b, double c, double d, double e, double f, double za, double nVis, double nPix, double fov, double nChan, bool zeroConstant, double& dt, size_t beamsTimesIters, double lambda, double maxBaseline, double maxHeight)
 {
 	std::ofstream file(filename);
[…]

### 2014-02-12 articles/2014/Article-WSClean/wsclean.tex
Work on article

@@
 For WSClean, the latest version as of February 2014 was used.
-The tests were run on a high-end desktop, with a 3.20-GHz Intel Core i7-3930K processor with 6 cores and 32 GB of memory.
+The tests were run on a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with 6 cores that can perform 138 giga-floating point operations per second (GFLOPS).
 The data are stored on a multi-disk array with 5 spinning hard disks, which has a combined read rate of about 450 MB/s.
@@
 \begin{table*}
-\caption{Predicted computational cost for various configurations, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree.
-Predictions are for $w$-projection in CASA, $w$-stacking with WSClean and $w$-stacked snapshots with WSClean and optimal snapshot duration, all with approximately equal accuracy.}
-\begin{tabular}{lrrrrrrrrrr|rrr}
- Configuration&$\lambda$& FOV  & Beams & Ant & Res & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ & $w$-proj & $w$-stack & $w$-snaps \\
-          & (m)       & (FWHM)&       &     &     & (km)                       & (m)             & (s)    & (MHz) &  (kHz) &  &  & \\
+\caption{Configurations for which the computational cost are predicted.} \label{tbl:computational-cost-configurations}
+\begin{tabular}{l|rrrrrrrrrr}
+ Configuration&$\lambda$& FOV  & Beams & Ant & Res & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ \\
+          & (m)       & (FWHM)&       &     &     & (km)                       & (m)             & (s)    & (MHz) &  (kHz) \\
 \hline
- GLEAM    &  2   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5 & 2 & 32 & 40     &11h 20m & 1h 49m & 1h 23m \\
- EMU      & 0.2  & 1\degree   & 30   &  36 & 10''  & 6 &  0 &10 & 300 & 20    &23h 52m &11h 52m & 11h 49m\\
- MSSS low &  5   & 9.8\degree & 5    &  20 & 100'' & 5 &    &10 & 16 & 16     & 0h 53m & 0h 12m & 0h 11m\\
- MSSS high&  2   & 3.8\degree & 5    &  40 & 120'' & 5 &    &10 & 16 & 16     & 1h 41m & 0h 39m & 0h 39m\\
- AARTFAAC &  5   & 45\degree  & 1    & 288 &  20'  & 0.3& 1 & 1 &  7 & 24     & 12h 01m & 4h 58m & 4h 58m \\
- VLSS     & 4.1  & 14\degree  & 1    &  27 &  80'' &11.1& 0 &10 & 1.56 & 12.2 & 0h 10m & 0h 11m & 0h 01m\\
- SKA-P1 AA inner&3& 1.2\degree& 480  &  35 & 140'' &2.5& 2 &1.2 & 380 & 1     & 3219 d & 1865 d  & 1865 d \\
- SKA-P1 AA mid&3 & 1.2\degree & 480  &  50 & 3.5'' &100& 50 &1.2 & 380 & 1     &  6740 d & 3845 d & 3844 d \\
+ GLEAM    &  2   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5 & 2 & 32 & 40 \\
+ EMU      & 0.2  & 1\degree   & 30   &  36 & 10''  & 6 &  0 &10 & 300 & 20\\
+ MSSS low &  5   & 9.8\degree & 5    &  20 & 100'' & 5 &    &10 & 16 & 16 \\
+ MSSS high&  2   & 3.8\degree & 5    &  40 & 120'' & 5 &    &10 & 16 & 16 \\
+ AARTFAAC &  5   & 45\degree  & 1    & 288 &  20'  & 0.3& 1 & 1 &  7 & 24 \\
+ VLSS     & 4.1  & 14\degree  & 1    &  27 &  80'' &11.1& 0 &10 & 1.56 & 12.2 \\
+ SKA-P1 AA inner&3& 1.2\degree& 480  &  35 & 140'' &2.5& 2 &  8 & 380 & 80 \\
+ SKA-P1 AA +mid&3 & 1.2\degree & 480  &  50 & 3.5'' &100& 50 & 0.2 & 380 & 2 \\
+\end{tabular}
+\end{table*}
+
+\begin{table*}
+\caption{Predicted computational cost for configurations listed in Table~\ref{tbl:computational-cost-configurations}, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree.
+Predictions are for $w$-projection with CASA; $w$-stacking with WSClean; $w$-stacked snapshots with WSClean using optimal snapshot duration; and a hybrid between $w$-projection and $w$-stacking using CASA with optimal $\Delta w$.
+All have approximately equal accuracy.} \label{tbl:computational-cost-predictions}
+\begin{tabular}{l|rrrr|}
+               & \multicolumn{4}{c|}{Predicted computing cost on test computer} \\
+ Configuration & $w$-projection & $w$-stacking & $w$-snapshot & hybrid \\
+ \hline
+ GLEAM    & 11h 20m & 1h 49m & 1h 23m  & 3h 07m   \\
+ EMU      & 23h 52m &11h 52m & 11h 49m &23h 52m   \\
+ MSSS low & 0h 53m & 0h 12m & 0h 11m   & 0h 53m   \\
+ MSSS high& 1h 41m & 0h 39m & 0h 39m   & 1h 20m   \\
+ AARTFAAC & 12h 01m & 4h 58m & 4h 58m  & 8h 38m   \\
+ VLSS     & 0h 10m & 0h 11m & 0h 01m   & 0h 10m   \\
+ SKA-P1 AA inner&7.3 d &2.9 d &        &        \\
+ SKA-P1 AA mid& 16839 d & 9605 d &        &        \\
 \end{tabular}
@@
 
-\section{Discussion \& conclusions} \label{sec:discussion}
+\section{Conclusions} \label{sec:discussion}
 We have shown that the $w$-stacking algorithm is well suited for imaging MWA observations.
-The WSClean $w$-stacking implementation is faster than CASA's $w$-projection algorithm in all common MWA imaging configurations, and already an order of magnitude faster at a relatively small zenith angle of $10$\degree.
-At the same time, it is slightly more accurate.
-For approximately zenith angles $>15\degree$ or field of views $>35\degree$, rephasing visibility sets and performing snapshot imaging becomes worthwhile.
-This can lead to performance improvements of a factor of $3$ for the MWA.
-For arrays with longer baselines and higher resolution, snapshot imaging leads to even higher performance improvements.
-This agrees with \citet{widefield-imaging-ska-cornwell}.
-Additionally, we have shown that to implement $w$-snapshot imaging, a pure $w$-stacking implementation is likely more efficient than pure $w$-projection.
-A hybrid between the two might improve things by a few factors.
-TODO should we calculate some timings with SKA's configuration?
+The WSClean $w$-stacking implementation is faster than CASA's $w$-projection algorithm in all common MWA imaging configurations, leading up to an order of magnitude difference already at a relatively small zenith angle of $10$\degree.
+It is also slightly more accurate.
+For approximately zenith angles $>15\degree$ or field-of-view diameters $>35\degree$, snapshot imaging becomes faster.
+This can lead to performance improvements of a factor of $3$ for the MWA, but only in the most expensive imaging configurations that are less commonly used.
+Considering the extra regridding step required, which complicates issues such as calculating the integrated beam shape, recentering snapshots might not be worthwhile for the MWA.
+Extrapolation of the computing cost predicts that this holds for most arrays, including SKA low.
+A hybrid between $w$-stacking and $w$-projection might improve things by a few factors.
 
-It remains challenging to joinedly deconvolve observations that are long enough to encur significant changes in the beam shape.
-It is possible to clean such an observation without beam corrections, but then a constant amount of flux is subtracted over time, which leads to inaccuracies.
+It remains challenging to deconvolve observations that are long enough to encur significant changes in the beam shape.
+It is possible to clean an observation integrated over time without beam corrections, but this will subtract a constant (absolute) flux over time.
 Alternatively, beam corrections can be made, either by integrating beam-corrected snapshots or by performing $a$-projection.
@@
 When it is required to calculate the beam on short time scales, calculating $a$-projection kernels or performing snapshot imaging also becomes much more expensive.
-For a few hundred components, an approach with direct Fourier transforms is accurate and affordable, but this becomes intractable for thousands of components.
-Images with diffuse emmission typically require cleaning of more than ten-thousand components.
-Joined deconvolution is not an issue for achieving low to intermediate dynamic range with the MWA.
-Because of the excellent instanteneous $uv$-coverage and near-confusion noise level, snapshots can be cleaned individually to deep levels.
+For a few thousand components, an approach with direct Fourier transforms is accurate and affordable, but this becomes intractable for tens of thousands of components.
+Images with diffuse emmission typically require cleaning that many components.
+
+Deconvolution is not an issue for achieving low to intermediate dynamic range with the MWA.
+Because of sufficient instanteneous $uv$-coverage and near-confusion snapshot noise level, snapshots can be cleaned individually to deep levels.
 For example, cleaning to $5\sigma$ level results in a cleaning threshold of approximately 100~mJy in 112~s observations.
@@
 
-TODO discuss fact that w-projection needs layers for multi-threading.
+By using the $w$-stacking algorithms, some computational cost is transferred from gridding to performing FFTs.
+WSClean uses the FFTW library for calculating the FFTs \citep{fftw-2005}.
+Further performance improvement can be made by using a FFT library that is executed on graphical processing units (GPUs).
 
-TODO discuss optimalisations with GPUs
-
-TODO The results do show that synthesis imaging using common imaging parameters results in an approximate flux-density standard error of at least $\sim1\%$.
+TODO The results show that synthesis imaging using common imaging parameters results in an approximate flux-density standard error of at least $\sim1\%$.
 

### 2014-02-12 articles/2014/Article-WSClean/img/fit.cpp

@@ -136,17 +136,22 @@ double eval(FittingFunc func, double za, double nVis, double nPix, double fov, d
 	}
 }
 
-double evalWE(FittingFunc func, double we, double nVis, double nPix, double nChan, double a, double b, double c, double d, double e, double f)
+double evalWE(FittingFunc func, double za, double fov, double fact, double nVis, double nPix, double nChan, double a, double b, double c, double d, double e, double f, double lambda, double maxBaselineM, double maxHeightDiffM)
 {
 	if(func == WSCleanFunc) {
-		return nChan * (a * (we+b)*nPix*nPix*log2(nPix) + c * nVis/nChan) + d;
+		double we = sinZA(za, lambda, maxBaselineM, maxHeightDiffM) * fovFact(fov) / fact;
+		return nChan * (a * (we+b)*nPix*nPix*log2(nPix) + d) + c * nVis;
 	}
 	else if(func == WSSCleanFunc) {
-		return nChan * (a *nPix*nPix*log2(nPix) + b * nVis/nChan) + c;
+		return nChan * a *nPix*nPix*log2(nPix) + b * nVis + c;
+	}
+	else if(func == CASAFunc) {
+		double we = (sinZA(za, lambda, maxBaselineM, maxHeightDiffM) / sqrt(fact) + e) * fovFact(fov) / sqrt(fact);
+		return nChan * a * nPix*nPix*log2(nPix) + b * nVis * (we * we + d) + f;
 	}
 	else {
-		we = we + e;
-		return nChan * a * nPix*nPix*log2(nPix) + b * nVis * (we * we + d) + f;
+		double we = maxHeightDiffM/lambda * fovFact(fov) / fact;
+		return nChan * a * (we+b)*nPix*nPix*log2(nPix) + c * nVis + d;
 	}
 }
 
@@ -625,34 +630,32 @@ void printDuplicates(const std::vector<Sample>& samples)
 	}
 }
 
-void optimalDeltaW(FittingFunc func, double a, double b, double c, double d, double e, double f, double za, double nVis, double nPix, double fov, double nChan, bool zeroConstant, double& dw, double& casaGain, double& wscleanGain)
[…]

### 2014-02-13 articles/2014/Article-WSClean/wsclean.tex
First draft of article finished.

@@
 \begin{abstract}
-We present our new imager implementation, ``WSClean'', which we find to be an order of magnitude faster than existing implementations, as well as being capable of MWA full-sky imaging at full resolution.
-TODO
+We present a new imager implementation that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm.
+On MWA data, we find it to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution.
+The performance dependencies of CASA's $w$-projection and WSClean are analysed and analytical functions are derived that describe the required computing cost for both imagers.
+The difference in computing cost is mainly caused by the efficiency of our new imager on MWA's 27\degree field of view.
+We predict the computing costs for several other configurations and estimate that our imager is approximately a factor of 2--3 faster on other arrays.
+Especially imaging SKA observations will become challenging, for which we predict that our new algorithm is still at least 30 times slower than is budgetted.
+We find that the $w$-snapshot has only a minor benefit.
+The source code of our new imager is publicly released.
 \end{abstract}
@@
 
-Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with simple imagers that are based on a two-dimensional Fast Fourier Transformation (FFT).
+Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with a two-dimensional Fast Fourier Transformation (FFT).
 Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}.
-This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes, that by nature are wide-field instruments.
+This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes that by nature are wide-field instruments.
 
 There are several methods to deal with the w-term during imaging: facetting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, w-projection \citep{wprojection-cornwell}, w-stacking \citep{widefield-imaging-ska-cornwell} and w-snapshots \citep{widefield-imaging-ska-cornwell}.
-The w-snapshots algorithm is a hybrid between snapshot imaging and one of the other w-term correcting techniques, most efficiently w-projection or w-stacking.
 
@@
 Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}).
-The Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}) has an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution.
+The Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution.
 However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's w-projection algorithm.
@@
 
-Another option for reducing MWA data is the Real-Time System (RTS), which has been designed as an efficient calibrating and imaging pipeline specifically for MWA data \citep{rts-mwa}.
+Another option exists for imaging MWA data: the Real-Time System (RTS, \citealt{rts-mwa}).
+This has been designed as an efficient calibration and imaging pipeline specifically for MWA data.
 It can use GPUs for improving efficiency, and uses a w-projection kernel during imaging to correct for the w-term.
-The kernel can be kept small, because the data is processed in snapshots that have been phase-centered to Zenith.
+The kernel can be kept small, because the data is processed in snapshots that have been phase-centered to zenith.
 The RTS can not perform deconvolution as of yet.
@@
 Once the Square-Kilometre Array (SKA) begins its operation, the required computational power for wide-field imaging will become an even bigger challenge.
-\citet{widefield-imaging-ska-cornwell} shows that in theory the w-snapshots algorithm is the most efficient approach for the SKA.
+\citet{widefield-imaging-ska-cornwell} argues that the w-snapshots algorithm is the most efficient approach for the SKA.
 
@@
 The performance and accuracy will be analysed in Sect.~\ref{sec:analysis}.
+The conclusions are presented in Sect.~\ref{sec:conclusions}.
 
@@
 \begin{align}\notag
-V(u,v,w) = & \iint \frac{A(l,m) I(l,m)}{\sqrt{1-l^2-m^2}} \cdot \\ \label{eq:visibility-function}
+V(u,v,w) = & \iint \frac{A(l,m) I(l,m)}{\sqrt{1-l^2-m^2}} \times \\ \label{eq:visibility-function}
 & e^{-2\pi i \left(ul + vm + w(\sqrt{1-l^2-m^2}-1)\right)} dl dm,
@@
 We will not discuss calibration, but assume $V$ has been calibrated before imaging.
-In the case of a polarized measurement, the symbols $V$, $A$ and $I$ are $2\times 2$ matrices, but without loss of generality we will ignore polarization and treat inversion as a scalar problem.
+In the case of a polarized measurement, the symbols become $2\times 2$ matrices, but without loss of generality we will ignore polarization and treat inversion as a scalar problem.
 Imaging consists of inverting Eq.~\eqref{eq:visibility-function}, i.e., to find $I'$ from $V$.
 
-For small field of views, the term $\sqrt{1-l^2-m^2}$ is approximately of unit size, making Eq.~\eqref{eq:visibility-function} approximate an ordinary 2-dimensional Fourier transform, which can be inverted with an inverse Fast Fourier Transform (FFT).
+For small field of views, the term $\sqrt{1-l^2-m^2}$ is approximately of unit size, making Eq.~\eqref{eq:visibility-function} approximate an ordinary 2-dimensional Fourier transform that can be inverted with an inverse Fast Fourier Transform (FFT).
 A common rule is that this is valid when
@@
 \begin{align}\notag
-V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \cdot \\ \notag
+V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \times \\ \notag
 & e^{-2\pi i \left(ul + vm\right)} dl dm.
@@
 %\begin{align}\notag
-%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \cdot \\ \notag
+%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \times \\ \notag
 %& e^{-2\pi i \left(ul + vm\right)} du dv,
@@
 \begin{align}\notag
-\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \cdot \\ \notag
+\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \times \\ \notag
 & e^{2\pi i \left(ul + vm\right)} du dv.
@@
 \begin{align}\notag
-\frac{I'(l,m)\left(w_e - w_0\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \cdot \\ \label{eq:wstacking}
+\frac{I'(l,m)\left(w_e - w_0\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \times \\ \label{eq:wstacking}
 \iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
@@
 In this case we can therefore calculate the image for $w<0$ from the image with $w>0$.
-That allows us to set $w_0$ to the minimum absolute $w$-value, which requires at least two times less layers.
-In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence is always performed as a complex-to-complex transform.
+That allows us to set $w_0$ to the minimum absolute $w$-value, which requires two times fewer layers.
+In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence the inverse FFT is always performed as a complex-to-complex transform.
 
@@
 Therefore, accurate prediction is more important than accurate imaging, because aliasing artefacts are attenuated by cleaning as long as the model is subtracted accurately.
-Possible ways to allow less $w$-layers during imaging is by predicting with more $w$-layers or a direct Fourier transform.
-We have not looked into these options yet.
 
@@
 \begin{align}\notag
-\frac{I'(\hat{l}, \hat{m})\left(w_e - w_0\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \cdot \\
+\frac{I'(\hat{l}, \hat{m})\left(w_e - w_0\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \times \\
 \iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
@@
 A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together.
-A recentred image can be stored in a Flexible Image Transport System (FITS) file, which allows setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}.
+A recentred image can be stored in a Flexible Image Transport System (FITS) file with the slant orthographic (SIN) projection normally used in synthesis imaging, by setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}.
 Common viewers such as Kvis and DS9 support such FITS files and display their coordinates correctly.
@@
 The used software version of CASA is ``stable release 42.0, revision 26465'', which was released September 2013.
-It includes performance improvements to the \texttt{clean} task which have recently been made.
 For WSClean, the latest version as of February 2014 was used.
@@
 This evidently has a stronger effect on the $w$-projection algorithm.
-However, the computational performance of the $w$-projection algorithm is hardly affected by the number of $w$-projection planes, and in practical situations one would always use more w-projection planes\footnote{In the zenith case CASA issues a warning to use more $w$-projection planes.}.
+However, the computational performance of the $w$-projection algorithm is hardly affected by the number of $w$-projection planes, and in practical situations one would always use more w-projection planes.
 When 128 $w$-projection planes are used in CASA, its residual RMS is equal to WSClean with 12 $w$-layers, but the flux density measurements are less accurate.
-We do not know why this parameter needs to be higher in CASA to reach the same RMS.\footnote{TODO: do I want to find out this is because of invalid sqrt(w) gridding?} Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlayers}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree.
+We do not know why this parameter needs to be higher in CASA to reach the same RMS (TODO: do I want to find out this is because of invalid sqrt(w) gridding?).
+Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlayers}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree.
 Also unexpected is that the source flux density becomes worse by increasing the number of planes.
@@
 The extra noise can be observed in the produced images, as shown in Fig.~\ref{fig:residuals}.
-Also made is an image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}.
+An image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}, is also made and analyzed.
 Its accuracy is higher compared to the normal projection, but its residual noise is also higher.
@@
 WSClean is faster in all tested cases.
-WSClean and CASA determine a different next major iteration threshold.
-CASA performs 5 major iterations at zenith and 8 at 10\degree angle, while WSClean always performs 5 major iterations.
-Major iterations are dominating the computing time.
+Both imagers have performed 5 major iterations for these results, and major iterations are dominating the computing time.
 We will look more closely at the difference in performance in the next section.
@@
 These functions follow the trend of the measurements well, with an absolute error of 14.7\% and 20.7\% for respectively the WSClean and CASA functions.
+(TODO $w_e$ is slightly different for CASA)
 
@@
 Therefore, from a performance perspective, the snapshot duration should be as large as possible.
-For $w$-projection in CASA we find an optimal snapshot duration of 122~s for MWA observations.
+For $w$-projection in CASA we find an optimal snapshot duration of $\sim2$ min for MWA observations.
 
@@
  GLEAM    &  2   & 24.7\degree& 1    & 128 & 2'   & 2.9 & 5 & 2 & 32 & 40 \\
- EMU      & 0.2  & 1\degree   & 30   &  36 & 10''  & 6 &  0 &10 & 300 & 20\\
- MSSS low &  5   & 9.8\degree & 5    &  20 & 100'' & 5 &    &10 & 16 & 16 \\
- MSSS high&  2   & 3.8\degree & 5    &  40 & 120'' & 5 &    &10 & 16 & 16 \\
- AARTFAAC &  5   & 45\degree  & 1    & 288 &  20'  & 0.3& 1 & 1 &  7 & 24 \\
- VLSS     & 4.1  & 14\degree  & 1    &  27 &  80'' &11.1& 0 &10 & 1.56 & 12.2 \\
+ EMU      & 0.2  & 1\degree   & 30   &  36 & 10''  & 6 & 0.2 &10 & 300 & 20\\
+ MSSS low &  5   & 9.8\degree & 5    &  20 & 100'' & 5 & 2  &10 & 16 & 16 \\
+ MSSS high&  2   & 3.8\degree & 5    &  40 & 120'' & 5 & 2  &10 & 16 & 16 \\
+ LOFAR LBA NL& 5 & 4.9\degree & 1    &  38 &   3'' &180& 20 & 1 & 96 & 1 \\
+ AARTFAAC &  5   & 45\degree  & 1    & 288 &  20'  & 0.3& 0.5 & 1 &  7 & 24 \\
+ VLSS     & 4.1  & 14\degree  & 1    &  27 &  80'' &11.1& 0.2 &10 & 1.56 & 12.2 \\
+ MeerKAT  & 0.2  &  1\degree  & 1    &  64 &   6'' &  8 & 1   & 0.5 & 750  & 50 \\
  SKA-P1 AA inner&3& 1.2\degree& 480  &  35 & 140'' &2.5& 2 &  8 & 380 & 80 \\
@@
 \begin{table*}
-\caption{Predicted computational cost for configurations listed in Table~\ref{tbl:computational-cost-configurations}, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree.
+\caption{Predicted computational cost for configurations listed in Table~\ref{tbl:computational-cost-configurations}, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree with 1 polarization.
 Predictions are for $w$-projection with CASA; $w$-stacking with WSClean; $w$-stacked snapshots with WSClean using optimal snapshot duration; and a hybrid between $w$-projection and $w$-stacking using CASA with optimal $\Delta w$.
 All have approximately equal accuracy.} \label{tbl:computational-cost-predictions}
-\begin{tabular}{l|rrrr|}
-               & \multicolumn{4}{c|}{Predicted computing cost on test computer} \\
- Configuration & $w$-projection & $w$-stacking & $w$-snapshot & hybrid \\
+\begin{tabular}{l|rrrr|r}
+               & \multicolumn{4}{c|}{Predicted computing cost on test computer} & max \\
+ Configuration & $w$-projection & $w$-stacking & $w$-snapshot & hybrid & FLOPS/float \\
  \hline
- GLEAM    & 11h 20m & 1h 49m & 1h 23m  & 3h 07m   \\
- EMU      & 23h 52m &11h 52m & 11h 49m &23h 52m   \\
- MSSS low & 0h 53m & 0h 12m & 0h 11m   & 0h 53m   \\
- MSSS high& 1h 41m & 0h 39m & 0h 39m   & 1h 20m   \\
- AARTFAAC & 12h 01m & 4h 58m & 4h 58m  & 8h 38m   \\
- VLSS     & 0h 10m & 0h 11m & 0h 01m   & 0h 10m   \\
- SKA-P1 AA inner&7.3 d &2.9 d &        &        \\
- SKA-P1 AA mid& 16839 d & 9605 d &        &        \\
+ GLEAM    &   67h 7m & 8h 31m &  8h 05m & 15h 18m & $6.9 \times 10^2$    \\
+ EMU      &    5.2 d &  2.9 d &   2.9 d & 5.2 d   & $2.1 \times 10^4$      \\
+ MSSS low &  2h 18m  & 0h 59m &  0h 58m & 2h 16m  & $3.5 \times 10^3$   \\
+ MSSS high&  7h 16m  & 3h 52m &  3h 52m & 7h 16m  & $3.4 \times 10^3$   \\
+ LOFAR NL &  50.6 d  &   10 d &   7.2 d & 15.9 d & $4.5 \times 10^3$ \\
+ AARTFAAC &  71h 25m &29h 50m & 29h 50m & 51h 34m & $6.8 \times 10^2$   \\
+ VLSS     &  0h 12m  & 0h 12m &  0h 02m &  0h 12m & $12.1 \times 10^3$     \\
+ MeerKAT  &   10.8 d &  6.2 d &   6.2 d &  10.8 d & $6.8 \times 10^2$ \\
+ SKA-P1 AA inner&7.2 d& 2.9 d &   2.9 d &   3.7 d & $3.3 \times 10^5$   \\
+ SKA-P1 AA mid&100,996 d&57,605 d& 57,605 d & 99,402 d & $3.3 \times 10^5$ \\
 \end{tabular}
 \end{table*}
-
-\subsection{Estimated computing cost for other telescopes} \label{sec:application-to-nonmwa}
-We use the derived functions to estimate the computational cost of the algorithms for configurations of several telescopes.
-We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multifrequency Snapshot Sky Survey (MSSS)\footnote{See \url{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; the Amsterdam—ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \url{http://www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); and the low-frequency phase~1 aperture arrays of the Square-Kilometre Array, with inner (2.5~km) and mid (2.5~km) configurations \citep{ska-phase1-2010}.
-In multi-beam configurations such as the SKA and EMU configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams.
 
@@
 with $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta w}{w_e}$ and $w'_e = \Delta w$.
+For MWA observations, the optimal value for $\Delta w$ is about unity, and improves the speed of $w$-projection with about a factor of four.
+However, the performance of the hybrid method is still about a factor of two lower than our pure-stacking implementation.
+This could again be the difference in optimisation choices between CASA and WSClean, but it does show that the effort of implementing a hybrid over pure stacking might not be worthwhile.
 
-\section{Conclusions} \label{sec:discussion}
+An optimisation suggested by \citet{awimager-2013} is to not grid visibilities with $w$-values larger than some value, because this is where most of the computational cost resides when using $w$-projection, while for LOFAR there is little benefit in gridding these samples.
+In $w$-stacking, the speed gain associated with this optimisation is less significant.
+Limitting the $w$-values also lowers the snapshots resolution in one direction significantly, because limitting all the long baselines in one direction are no longer gridded, which is often not desirable for the MWA.
+
+\subsection{Estimated computing cost for other telescopes} \label{sec:application-to-nonmwa}
+We use the derived functions to estimate the computational cost of the algorithms for configurations of several telescopes.
+We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multifrequency Snapshot Sky Survey (MSSS)\footnote{See \url{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam—ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \url{http://www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The Meer Karoo Array Telescope (MeerKAT); and the low-frequency phase~1 aperture arrays of the Square-Kilometre Array, with inner (2.5~km) and mid (2.5~km) configurations \citep{ska-phase1-2010}.
+The configurations are summarized in Table~\ref{tbl:computational-cost-configurations}.
+In multi-beam configurations such as the SKA and EMU configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams.
+
+The results are summarized in Table~\ref{tbl:computational-cost-predictions}.
+WSClean is in most situations predicted to be 2--3 times faster than CASA.
+The MWA is somewhat of an exception, where WSClean is 8$\times$ faster.
+The main cause for this is the wide field of view of the MWA.
+The $w$-snapshot method does not improve much over normal WSClean operation in any of the cases, except the VLSS case.
+We think this is because the VLSS had a modest data rates compared to the other configurations, making the constant cost per $w$-plane/layer stand out.
+The $w$-snapshot might become more valuable at higher zenith angles (TODO do I want to predict that too?) The hybrid method is in all situations appromixately a factor of two slower than WSClean.
+
+\section{Conclusions} \label{sec:conclusions}
 We have shown that the $w$-stacking algorithm is well suited for imaging MWA observations.
@@
 
-It remains challenging to deconvolve observations that are long enough to encur significant changes in the beam shape.
-It is possible to clean an observation integrated over time without beam corrections, but this will subtract a constant (absolute) flux over time.
-Alternatively, beam corrections can be made, either by integrating beam-corrected snapshots or by performing $a$-projection.
-This however leads to a stronger spatially-varying point-spread function, and would allow fewer image-based minor iterations in a major cycle.
-When it is required to calculate the beam on short time scales, calculating $a$-projection kernels or performing snapshot imaging also becomes much more expensive.
-For a few thousand components, an approach with direct Fourier transforms is accurate and affordable, but this becomes intractable for tens of thousands of components.
-Images with diffuse emmission typically require cleaning that many components.
+In \citet{ska-phase1-2010}, the SKA computing power is budgetted by assuming an imaging speed of $10^4$ FLOPS/float.
+Extrapolation of our results shows that the current imagers perform $30$--$70$ times slower than this on SKA data.
+It was suggested that $w$-snapshots would improve the imaging speed for the SKA situation \citep{widefield-imaging-ska-cornwell}, but our results do not confirm this.
+This is mainly because the primary beam of a SKA station is not very large (1.2\degree), which causes the $w$-terms to not have a significant effect on performance: the bulk of the time spent is on gridding the antialiasing kernels instead of the $w$-kernels.
+Clearly it remains challenging to perform widefield imaging with acceptable performance.
+Nevertheless, some optimisations could be made for the SKA.
+For example, a performance gain of a few factors can already be achieved by averaging shorter baselines to their maximum integration time and bandwidth resolution before imaging.
 
-Deconvolution is not an issue for achieving low to intermediate dynamic range with the MWA.
+Deconvolution is not an issue for achieving low to intermediate dynamic ranges with the MWA.
 Because of sufficient instanteneous $uv$-coverage and near-confusion snapshot noise level, snapshots can be cleaned individually to deep levels.
 For example, cleaning to $5\sigma$ level results in a cleaning threshold of approximately 100~mJy in 112~s observations.
-However, MWA's EoR observations (TODO cite) require more advanced deconvolution techniques.
+However, MWA's EoR observations \citep{bowman-science-with-the-mwa-2013} require more advanced deconvolution techniques.
+
+So far, we have corrected the (full-polarization) beam in image space, which is possible because of the homogenuity of the MWA tiles.
+This is cheap for our current snapshot time of 112s, but this might be unfeasible when it is required to calculate the beam on very short time scales.
+Accurate MWA beam models are still being developed.
+When beam updates are required often, calculating $a$-projection kernels or performing snapshot imaging also becomes more expensive, and it would be interesting to find out where the balance lies between these methods.
+When only a few thousand components need to be deconvolved, an approach with direct Fourier transforms is accurate and affordable.
+For wide-field arrays this can be combined with calibration of the direction dependence, e.g.
+with the SAGECAL \citep{sage-calibration-ii} or RTS peeling \citep{rts-mwa} calibration techniques.
+This is not possible for fields with diffuse emmission for which no model is known a-priory.
 
@@
 WSClean uses the FFTW library for calculating the FFTs \citep{fftw-2005}.
-Further performance improvement can be made by using a FFT library that is executed on graphical processing units (GPUs).
-
-TODO The results show that synthesis imaging using common imaging parameters results in an approximate flux-density standard error of at least $\sim1\%$.
+Further performance improvement can be made by using one of the available FFT libraries that make use of graphical processing units (GPUs).
 

### 2014-02-13 articles/2014/Article-WSClean/img/fit.cpp

@@ -63,7 +63,7 @@ std::string timeToStr(double secComputingRequired, double secObservation, double
 		str << round(gFlopsRequired*10.0)/10.0 << " GFLOPS";
 	str << ", ";
 	double perVis = secComputingRequired * BATTLESTAR_GFLOPS_PER_SECOND / nMVis;
-	str << perVis << " or " << round(10.0*perVis / exp10(floor(log10(perVis))))/10.0 << " x 10^" << floor(log10(perVis)) << " FLOPS/vis";
+	str << perVis << " or " << round(20.0*perVis / exp10(floor(log10(perVis))))/10.0 << " x 10^" << floor(log10(perVis)) << " FLOPS/float";
 	
 	return str.str();
 }
@@ -659,7 +659,8 @@ void optimalDeltaW(FittingFunc func, double a, double b, double c, double d, dou
 	double i = 1;
 	double fz = zeroConstant ? 0.0 : f, fnz = zeroConstant ? f : 0.0;
 	double we = sinZA(za) * fovFact(fov);
-	while(i < 100.0)
+	const double totalTime = 60.0 * 60.0;
+	while(i < 1000.0)
 	{
 		double y = double(i) * evalWE(func, za, fov, double(i), nVis / double(i), nPix, nChan, a, b, c, d, e, fz, lambda, maxBaseline, maxHeight) + fnz;
 		double dw = we / double(i);
@@ -667,12 +668,12 @@ void optimalDeltaW(FittingFunc func, double a, double b, double c, double d, dou
 			minDW = dw;
 			minTime = y;
 		}
-		i *= 1.01;
+		i *= 1.001;
 	}
 	
-	double tNormal = beamsTimesIters * eval(func, za, nVis, nPix, fov, nChan, a, b, c, d, e, fz, lambda, maxBaseline, maxHeight);
+	double tNormal = beamsTimesIters * (eval(func, za, nVis, nPix, fov, nChan, a, b, c, d, e, fz, lambda, maxBaseline, maxHeight) + fnz);
 	minTime *= beamsTimesIters;
[…]

### 2014-02-14 articles/2014/Article-WSClean/wsclean.tex
Article

@@
 Especially imaging SKA observations will become challenging, for which we predict that our new algorithm is still at least 30 times slower than is budgetted.
-We find that the $w$-snapshot has only a minor benefit.
+We find that the $w$-snapshot algorithm has only a minor benefit.
 The source code of our new imager is publicly released.
@@
 \begin{table*}
-\caption{Predicted computational cost for configurations listed in Table~\ref{tbl:computational-cost-configurations}, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree with 1 polarization.
+\caption{Predicted computational costs for configurations listed in Table~\ref{tbl:computational-cost-configurations}, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree with 1 polarization.
 Predictions are for $w$-projection with CASA; $w$-stacking with WSClean; $w$-stacked snapshots with WSClean using optimal snapshot duration; and a hybrid between $w$-projection and $w$-stacking using CASA with optimal $\Delta w$.

### 2014-02-14 cotter/cotter.cpp
Adding version information and some meta data to MS and UVFits outputs

@@ -5,11 +5,13 @@
 #include "fitswriter.h"
 #include "geometry.h"
 #include "mswriter.h"
+#include "mwafits.h"
 #include "mwams.h"
 #include "subbandpassband.h"
 #include "progressbar.h"
 #include "threadedwriter.h"
 #include "radeccoord.h"
+#include "version.h"
 
 #include <aoflagger.h>
 
@@ -515,8 +517,13 @@ void Cotter::processOneContiguousBand(const std::string& outputFilename, size_t 
 	
 	if(_outputFormat == MSOutputFormat)
 	{
-		std::cout << "Writing MWA fields to file...\n";
-		writeMWAFields(outputFilename, _mwaConfig.Header().nScans/partCount);
+		std::cout << "Writing MWA fields to measurement set...\n";
+		writeMWAFieldsToMS(outputFilename, _mwaConfig.Header().nScans/partCount);
+	}
+	else if(_outputFormat == FitsOutputFormat)
+	{
+		std::cout << "Writing MWA fields to UVFits file...\n";
+		writeMWAFieldsToUVFits(outputFilename);
 	}
 	
 	_writeWatch.Pause();
@@ -1326,7 +1333,7 @@ void Cotter::writeAlignmentScans()
 	}
 }
 
-void Cotter::writeMWAFields(const std::string& outputFilename, size_t flagWindowSize)
+void Cotter::writeMWAFieldsToMS(const std::string& outputFilename, size_t flagWindowSize)
 {
 	MWAMS mwaMs(outputFilename);
 	mwaMs.InitializeMWAFields();
@@ -1371,5 +1378,11 @@ void Cotter::writeMWAFields(const std::string& outputFilename, size_t flagWindow
 	for(int i=0; i!=24; ++i)
 		mwaMs.WriteMWASubbandInfo(i, _mwaConfig.HeaderExt().subbandGains[i], false);
 	
-	mwaMs.WriteMWAKeywords(_mwaConfig.HeaderExt().fiberFactor, 0);
+	mwaMs.WriteMWAKeywords(_mwaConfig.HeaderExt().fiberFactor, _mwaConfig.HeaderExt().metaDataVersion, _mwaConfig.HeaderExt().mwaPyVersion, COTTER_VERSION_STR, COTTER_VERSION_DATE);
+}
+
+void Cotter::writeMWAFieldsToUVFits(const std::string& outputFilename)
+{
[…]

### 2014-02-14 cotter/cotter.h

@@ -147,7 +147,8 @@ class Cotter : private UVWCalculater
 		void reorderSubbands(aoflagger::ImageSet& imageSet) const;
 		void initializeSbOrder();
 		void writeAlignmentScans();
-		void writeMWAFields(const std::string& outputFilename, size_t flagWindowSize);
+		void writeMWAFieldsToMS(const std::string& outputFilename, size_t flagWindowSize);
+		void writeMWAFieldsToUVFits(const std::string& outputFilename);
 		size_t rowsPerTimescan() const
 		{
 			if(_removeFlaggedAntennae && _removeAutoCorrelations)

### 2014-02-14 cotter/metafitsfile.cpp

@@ -261,6 +261,14 @@ void MetaFitsFile::parseKeyword(MWAHeader &header, MWAHeaderExt &headerExt, cons
 		; //parseFitsDate(keyValue, header.year, header.month, header.day, header.refHour, header.refMinute, header.refSecond);
 	else if(keyName == "DATE")
 		; // Date that metafits was created; ignored.
+	else if(keyName == "VERSION")
+		headerExt.metaDataVersion = keyValue.c_str();
+	else if(keyName == "MWAVER")
+		headerExt.mwaPyVersion = parseFitsString(keyValue.c_str());
+	else if(keyName == "MWADATE")
+		headerExt.mwaPyDate = parseFitsString(keyValue.c_str());
+	else if(keyName == "TELESCOP")
+		; // Ignore; will always be set to 'MWA'
 	else if(keyName == "EXPOSURE" || keyName == "MJD" || keyName == "LST" || keyName == "HA" || keyName == "AZIMUTH" || keyName == "ALTITUDE" || keyName == "SUN-DIST" || keyName == "MOONDIST" || keyName == "JUP-DIST" || keyName == "GRIDNUM" || keyName == "RECVRS" || keyName == "CHANNELS" || keyName == "SUN-ALT" || keyName == "TILEFLAG" || keyName == "NAV_FREQ" || keyName == "FINECHAN" || keyName == "TIMEOFF")
 		; // Ignore these fields, they can be derived from others.
 	else

### 2014-02-14 cotter/mwaconfig.h

@@ -52,7 +52,7 @@ struct MWAHeaderExt
 	MWAHeaderExt();
 	
 	int gpsTime;
-	std::string observerName, projectName, gridName, mode, filename;
+	std::string observerName, projectName, gridName, mode, filename, mwaPyVersion, mwaPyDate, metaDataVersion;
 	int delays[16], subbandGains[24], subbandNumbers[24];
 	bool hasCalibrator, hasGlobalSubbandGains;
 	int centreSBNumber;

### 2014-02-14 cotter/mwafits.cpp

@@ -1,0 +1,82 @@
+#include "mwafits.h"
+#include "fitsuser.h"
+
+#include <sstream>
+
+#include <fitsio.h>
+
+struct MWAFitsData : private FitsUser
+{
+	MWAFitsData(const std::string& filename)
+	{
+		int status = 0;
+		fits_open_file(&_fitsPtr, filename.c_str(), READWRITE, &status);
+		checkStatus(status);
+	}
+	
+	~MWAFitsData()
+	{
+		int status = 0;
+		fits_close_file(_fitsPtr, &status);
+		checkStatus(status);
+	}
+	
+	void writeKey(const char* name, const char* value, const char* comment)
+	{
+		int status = 0;
+		fits_write_key(_fitsPtr, TSTRING, name, (void*) value, comment, &status);
+		checkStatus(status);
+	}
+	
+	void writeKey(const char* name, const std::string& value, const char* comment)
+	{
+		writeKey(name, value.c_str(), comment);
+	}
+	
+	void writeKey(const char* name, double value, const char* comment)
+	{
+		std::ostringstream str;
+		str << value;
+		writeKey(name, str.str(), comment);
+	}
+	
+private:
+	fitsfile *_fitsPtr;
+};
+
+const char *MWAFits::_keywordNames[] = {
+	"FIBRFACT", "METAVER",
+	"MWAPYVER",
+	"COTVER", "COTVDATE"
+};
+
+const char *MWAFits::_keywordComments[] = {
+	"Fiber velocity factor", "METAFITS version number",
+	"MWAPY version number",
+	"Cotter version", "Cotter version date"
+};
+
+MWAFits::MWAFits(const std::string& filename) : _filename(filename),
+	_data(new MWAFitsData(filename))
+{
+}
+
+MWAFits::~MWAFits()
+{
+	delete _data;
+}
+
+template<typename T>
+void MWAFits::writeKey(MWAFitsEnums::MWAKeywords keyword, const T& value)
+{
+	_data->writeKey(keywordName(keyword), value, keywordComment(keyword));
+}
+
+void MWAFits::WriteMWAKeywords(double fibreVelFactor, const std::string& metaDataVersion, const std::string& mwaPyVersion, const std::string& cotterVersion, const std::string& cotterVersionDate)
+{
[…]

### 2014-02-14 cotter/mwams.cpp

@@ -53,7 +53,9 @@ const std::string MWAMS::_columnNames[] = {
 };
 
 const std::string MWAMS::_keywordNames[] = {
-	"MWA_FIBER_VEL_FACTOR", "MWA_METADATA_VERSION"
+	"MWA_FIBER_VEL_FACTOR", "MWA_METADATA_VERSION",
+	"MWA_MWAPY_VERSION",
+	"MWA_COTTER_VERSION", "MWA_COTTER_VERSION_DATE"
 };
 
 MWAMS::MWAMS(const string& filename) : _filename(filename),
@@ -323,8 +325,11 @@ void MWAMS::WriteMWASubbandInfo(int number, double gain, bool isFlagged)
 	flagRowCol.put(index, isFlagged);
 }
 
-void MWAMS::WriteMWAKeywords(double fibreVelFactor, int metaDataVersion)
+void MWAMS::WriteMWAKeywords(double fibreVelFactor, const std::string& metaDataVersion, const std::string& mwaPyVersion, const std::string& cotterVersion, const std::string& cotterVersionDate)
 {
 	_data->_measurementSet.rwKeywordSet().define(keywordName(MWAMSEnums::MWA_FIBER_VEL_FACTOR), fibreVelFactor);
 	_data->_measurementSet.rwKeywordSet().define(keywordName(MWAMSEnums::MWA_METADATA_VERSION), metaDataVersion);
+	_data->_measurementSet.rwKeywordSet().define(keywordName(MWAMSEnums::MWA_MWAPY_VERSION), mwaPyVersion);
+	_data->_measurementSet.rwKeywordSet().define(keywordName(MWAMSEnums::MWA_COTTER_VERSION), cotterVersion);
+	_data->_measurementSet.rwKeywordSet().define(keywordName(MWAMSEnums::MWA_COTTER_VERSION_DATE), cotterVersionDate);
 }

### 2014-02-14 cotter/mwams.h

@@ -32,7 +32,9 @@ struct MWAMSEnums
 	};
 	enum MWAKeywords
 	{
-		MWA_FIBER_VEL_FACTOR, MWA_METADATA_VERSION
+		MWA_FIBER_VEL_FACTOR, MWA_METADATA_VERSION,
+		MWA_MWAPY_VERSION,
+		MWA_COTTER_VERSION, MWA_COTTER_VERSION_DATE
 	};
 };
 
@@ -78,7 +80,7 @@ class MWAMS
 		
 		void WriteMWASubbandInfo(int number, double gain, bool isFlagged);
 		
-		void WriteMWAKeywords(double fibreVelFactor, int metaDataVersion);
+		void WriteMWAKeywords(double fibreVelFactor, const std::string& metaDataVersion, const std::string& mwaPyVersion, const std::string& cotterVersion, const std::string& cotterVersionDate);
 		
 	private:
 		void addMWAAntennaFields();

### 2014-02-14 cotter/version.h

@@ -1,0 +1,9 @@
+#ifndef COTTER_VERSION_H
+#define COTTER_VERSION_H
+
+#define COTTER_VERSION_STR  "1.0"
+#define COTTER_VERSION_MAJOR 1
+#define COTTER_VERSION_MINOR 0
+#define COTTER_VERSION_DATE "2014-02-14"
+
+#endif

### 2014-02-17 articles/2014/Article-WSClean/wsclean.tex
Work on article

@@
 
-\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{E-mail:
-\url{andre.offringa@anu.edu.au}}, B.~McKinley$^{1,2}$, N.~Hurley-Walker, 
-F.~H.~Briggs, \newauthor
-J.~Rhee, L.~Feng, D.~Kaplan, A.~R.~Neben, J.~D.~Hughes, MWA commissioning team\newauthor
-and GLEAM members where appropriate \& MWA builders list
+\def\ANU{$^{1}$}
+\def\CAASTRO{$^{2}$}
+\def\ASU{$^{3}$}
+\def\CASS{$^{4}$}
+\def\Curtin{$^{5}$}
+\def\CfA{$^{6}$}
+\def\Haystack{$^{7}$}
+\def\MIT{$^{8}$}
+\def\NRAO{$^{9}$}
+\def\RRI{$^{10}$}
+\def\SKASA{$^{11}$}
+\def\Tata{$^{12}$}
+\def\UMelbourne{$^{13}$}
+\def\USydney{$^{14}$}
+\def\UW{$^{15}$}
+\def\UWisc{$^{16}$}
+\def\Victoria{$^{17}$}
+
+\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\hspace{2cm}E-mail:
+\url{andre.offringa@anu.edu.au}},
+B.~McKinley\ANU$^,$\CAASTRO,
+N.~Hurley-Walker\Curtin,
+F.~H.~Briggs$^{1,2}$, \newauthor
+J.~Rhee\ANU,
+L.~Feng,
+D.~L.~Kaplan\UWisc,
+A.~R.~Neben,
+R.~B.~Wayth\Curtin$^,$\CAASTRO, \newauthor
+J.~D.~Hughes\ANU,
+G.~Bernardi\SKASA,
+J.~D.~Bowman\ASU,
+F.~Briggs\ANU,
+R.~J.~Cappallo\Haystack,
+B.~E.~Corey\Haystack, \newauthor
+A.~A.~Deshpande\RRI,
+D.~Emrich\Curtin,
+B.~M.~Gaensler\USydney$^,$\CAASTRO,
+R.~Goeke\MIT,
+L.~J.~Greenhill\CfA, \newauthor
+B.~J.~Hazelton\UW,
+J.~N.~Hewitt\MIT,
+M.~Johnston-Hollitt\Victoria,
+J.~C.~Kasper\CfA, \newauthor
+E.~Kratzenberg\Haystack,
+C.~J.~Lonsdale\Haystack,
+M.~J.~Lynch\Curtin,
+S.~R.~McWhirter\Haystack,
+D.~A.~Mitchell\CASS$^,$\CAASTRO, \newauthor
+M.~F.~Morales\UW,
+E.~Morgan\MIT,
+D.~Oberoi\Tata,
+S.~M.~Ord\Curtin$^,$\CAASTRO,
+T.~Prabu\RRI, \newauthor
+A.~E.~E.~Rogers\Haystack,
+D.~A.~Roshi\NRAO \footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
+N.~Udaya~Shankar\RRI,
+K.~S.~Srivani\RRI,
+R.~Subrahmanyan\RRI$^,$\CAASTRO, \newauthor
+S.~J.~Tingay\Curtin$^,$\CAASTRO, 
+M.~Waterson\Curtin$^,$\ANU,
+R.~L.~Webster\UMelbourne$^,$\CAASTRO,
+A.~R.~Whitney\Haystack, \newauthor
+A.~Williams\Curtin,
+C.~L.~Williams\MIT
 \\
-$^{1}$RSAA, Australian National University, Mt Stromlo Observatory, via Cotter Road, Weston, ACT 2611, Australia \\
-$^{2}$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO) \\
+$^{1}$Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
+$^{2}$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
+$^{3}$School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
+$^{4}$CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
+$^{5}$International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
+$^{6}$Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
+$^{7}$MIT Haystack Observatory, Westford, MA 01886, USA\\
+$^{8}$Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
+$^{9}$National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
+$^{10}$Raman Research Institute, Bangalore 560080, India\\
+$^{11}$Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
+$^{12}$National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
+$^{13}$School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
+$^{14}$Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
+$^{15}$Department of Physics, University of Washington, Seattle, WA 98195, USA\\
+$^{16}$Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
+$^{17}$School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
 }
@@
 We predict the computing costs for several other configurations and estimate that our imager is approximately a factor of 2--3 faster on other arrays.
-Especially imaging SKA observations will become challenging, for which we predict that our new algorithm is still at least 30 times slower than is budgetted.
-We find that the $w$-snapshot algorithm has only a minor benefit.
+Imaging SKA observations is challenging, and SKA1 low will require 3--60 PetaFLOPS of computing power with current techniques.
+We find that the $w$-snapshot algorithm does not further improve computing requirements.
 The source code of our new imager is publicly released.
@@
 Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with a two-dimensional Fast Fourier Transformation (FFT).
-Instead, the imaging algorithm needs to account for the ``w-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}.
+Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}.
 This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes that by nature are wide-field instruments.
 
-There are several methods to deal with the w-term during imaging: facetting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, w-projection \citep{wprojection-cornwell}, w-stacking \citep{widefield-imaging-ska-cornwell} and w-snapshots \citep{widefield-imaging-ska-cornwell}.
+There are several methods to deal with the $w$-term during imaging: facetting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, $w$-projection \citep{wprojection-cornwell}, $w$-stacking \citep{ska-memo-regridding-2011} and $w$-snapshots \citep{widefield-imaging-ska-cornwell}.
 
@@
 The Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution.
-However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's w-projection algorithm.
+However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's $w$-projection algorithm.
 Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.
@@
 This has been designed as an efficient calibration and imaging pipeline specifically for MWA data.
-It can use GPUs for improving efficiency, and uses a w-projection kernel during imaging to correct for the w-term.
+It can use GPUs to improve efficiency, and uses a $w$-projection kernel during imaging to correct for the $w$-term.
 The kernel can be kept small, because the data is processed in snapshots that have been phase-centered to zenith.
@@
 This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data.
-The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, w-projection and w-stacking \citep{awimager-2013}.
+The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, $w$-projection and $w$-stacking \citep{awimager-2013}.
 
 Once the Square-Kilometre Array (SKA) begins its operation, the required computational power for wide-field imaging will become an even bigger challenge.
-\citet{widefield-imaging-ska-cornwell} argues that the w-snapshots algorithm is the most efficient approach for the SKA.
+\citet{widefield-imaging-ska-cornwell} argues that the $w$-snapshots algorithm is the most efficient approach for the SKA.
 
-In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's w-projection implementation, and is able to succesfully image any MWA observation.
-To obtain the increase in speed, the implementation uses the w-stacking method for correcting the w-terms, optionally combined with w-snapshot imaging.
-We named the new imager ``WSClean'', as an abbreviation for ``W-Stacking Clean''.
+In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's $w$-projection implementation, and is able to succesfully image any MWA observation.
+To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with $w$-snapshot imaging.
+We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''.
 Our new imaging implementation, which we believe to be the fastest generic widefield imager available, is publicly released\footnote{TODO add URL}.
@@
 
-\section{The w-stacking technique} \label{sec:wstacking}
+\section{The $w$-stacking technique} \label{sec:wstacking}
 In this section, we will describe the $w$-stacking algorithm from a mathematical point of view.
@@
 
-To derive the w-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
+To derive the $w$-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
 \begin{align}\notag
@@
 \end{align}
-Integrating both sides over $w_0$ to $w_e$, the minimum and maximum value of $w$, results in
+Integrating both sides over $w_{\min}$ to $w_{\max}$, the minimum and maximum value of $w$, results in
 \begin{align}\notag
-\frac{I'(l,m)\left(w_e - w_0\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \times \\ \label{eq:wstacking}
+\frac{I'(l,m)\left(w_{\max} - w_{\min}\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \times \\ \label{eq:wstacking}
 \iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
@@
 
-In practice, the final scaling will be different from $\left(w_e - w_0\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples.
+In practice, the final scaling will be different from $\left(w_{\max} - w_{\min}\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples.
 Therefore, the number of added samples and their weight have to be taken into account.
@@
 In this case we can therefore calculate the image for $w<0$ from the image with $w>0$.
-That allows us to set $w_0$ to the minimum absolute $w$-value, which requires two times fewer layers.
+That allows us to set $w_{\min}$ to the minimum absolute $w$-value, which requires two times fewer layers.
 In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence the inverse FFT is always performed as a complex-to-complex transform.
@@
 
-\subsection{Discretization of w} \label{sec:gridding-w}
+\subsection{Discretization of $w$} \label{sec:gridding-w}
 \begin{figure}
@@
 \end{figure}
-While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of w-layers that need to be processed.
+While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of $w$-layers that need to be processed.
 For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian.
@@
 \begin{equation} \label{eq:nwlayers-bound}
- N_\textrm{wlayers} \gg 2\pi \left(w_e - w_0\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
+ N_\textrm{wlayers} \gg 2\pi \left(w_{\max} - w_{\min}\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
 \end{equation}
 Actual values for the right-hand side can differ a lot.
-The value of $w_e - w_0$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image.
-For the MWA, a typical value of $w_e - w_0$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$.
+The value of $w_{\max} - w_{\min}$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image.
+For the MWA, a typical value of $w_{\max} - w_{\min}$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$.
 For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$.
@@
 
-\subsection{Time complexity of w-stacking} \label{sec:time-complexity}
-It is useful to analyse the time complexity of w-stacking and compare it with w-projection, to understand which algorithm performs better in a given situation.
-We will use the following symbols: $N_\textrm{wlay}$ is the number of w-layers for w-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each dimension, $N_\textrm{vis}$ is the number of visibilities and $N_\textrm{kern}$ is the size of the antialiasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the w-kernel for w-projection, $w_e$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.
+\subsection{Time complexity of $w$-stacking} \label{sec:time-complexity}
+It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation.
+We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each dimension, $N_\textrm{vis}$ is the number of visibilities and $N_\textrm{kern}$ is the size of the antialiasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.
 
@@
  \begin{tabular}{lll}
-   \textbf{Operation} & \textbf{W-stacking} & \textbf{W-projection} \\
+   \textbf{Operation} & \textbf{$w$-stacking} & \textbf{$w$-projection} \\
    \hline\hline
    Fourier transform(s) & $N_\textrm{wlay} N^2_\textrm{pix} \log N_\textrm{pix}$ & $N^2_\textrm{pix} \log N_\textrm{pix}$ \\
-   W-term corrections   & $N_\textrm{wlay} N^2_\textrm{pix}$ & $N_\textrm{vis} N^2_\textrm{wkern}$ \\
+   $w$-term corrections   & $N_\textrm{wlay} N^2_\textrm{pix}$ & $N_\textrm{vis} N^2_\textrm{wkern}$ \\
    Gridding & $N_\textrm{vis}N^2_\textrm{kern}$ & $N_\textrm{vis} N_\textrm{kern}^2$ \\
@@
 
-Table~\ref{tbl:computational-cost-per-operation} shows how the computational costs scale for the operations that dominate the imaging in the w-stacking and w-projection algorithms.
-For comparison, we can assume the gridding kernel adds a constant term and the terms $N_\textrm{wlay}$ and $N_\textrm{wkern}$ follow approximately $w_e \sin \alpha_\textrm{FOV}$.
-The time complexity for w-stacking is then given by
+Table~\ref{tbl:computational-cost-per-operation} shows how the computational costs scale for the operations that dominate the imaging in the $w$-stacking and $w$-projection algorithms.
+For comparison, we can assume the gridding kernel adds a constant term and the terms $N_\textrm{wlay}$ and $N_\textrm{wkern}$ follow approximately $w_{\max} \sin \alpha_\textrm{FOV}$.
+The time complexity for $w$-stacking is then given by
 \begin{equation}
-\textrm{TC}_\textrm{wstacking}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} w_e \sin \alpha_\textrm{FOV}+ N_\textrm{vis} \right),
+\textrm{TC}_\textrm{wstacking}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} w_{\max} \sin \alpha_\textrm{FOV}+ N_\textrm{vis} \right),
 \end{equation}
@@
 \begin{equation}
-\textrm{TC}_\textrm{wprojection}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} + N_\textrm{vis} w_e^2 \sin^2 \alpha_\textrm{FOV}\right).
+\textrm{TC}_\textrm{wprojection}=\mathcal{O}\left(N^2_\textrm{pix} \log N_\textrm{pix} + N_\textrm{vis} w_{\max}^2 \sin^2 \alpha_\textrm{FOV}\right).
 \end{equation}
@@
 
-\subsection{W-snapshots} \label{sec:snapshot-imaging}
+\subsection{$w$-snapshots} \label{sec:snapshot-imaging}
 \begin{figure*}
@@
 Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
-%When the Fourier transform is performed, the $w$ term has been corrected (either by w-stacking or by w-projection).
+%When the Fourier transform is performed, the $w$ term has been corrected (either by $w$-stacking or by $w$-projection).
 Therefore, if $U(u, v)$ is the input to the 2-dimensional Fourier transform, the transform is described by
 \begin{align}\notag
-\frac{I'(\hat{l}, \hat{m})\left(w_e - w_0\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \times \\
+\frac{I'(\hat{l}, \hat{m})\left(w_{\max} - w_{\min}\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \times \\
 \iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
@@
 A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together.
-A recentred image can be stored in a Flexible Image Transport System (FITS) file with the slant orthographic (SIN) projection normally used in synthesis imaging, by setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}.
+A recentred image can be stored in a Flexible Image Transport System (FITS) file with the slant orthographic (SIN) projection normally used in interferometric imaging \citep{fits-coordinates-2002}, by setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}.
 Common viewers such as Kvis and DS9 support such FITS files and display their coordinates correctly.
@@
 \section{Analysis} \label{sec:analysis}
-We will now analyse the performance and accuracy of our implementation, and compare it with the w-projection implementation in CASA.
+We will now analyse the performance and accuracy of our implementation, and compare it with the $w$-projection implementation in CASA.
 For the analysis, we use imaging parameters which are common for MWA Imaging.
 When a specific parameter value is not mentioned, the settings from Table~\ref{tbl:default-parameters} are used.
-The number of $w$-projection planes in w-projection is kept equal to the number of $w$-layers in w-stacking, and is set to the right hand side of Eq.~\eqref{eq:nwlayers-bound}.
+The number of $w$-projection planes in $w$-projection is kept equal to the number of $w$-layers in $w$-stacking, and is set to the right hand side of Eq.~\eqref{eq:nwlayers-bound}.
 This yields 195 $w$-planes/layers at 10\degree zenith angle.
@@
 Nr. polarizations in set & 4 \\
-Imaged polarization & XX \\
+Imaged polarization & $pp$ ($\sim$XX) \\
 Imaging mode & Multi-frequency synthesis \\
@@
 \hline
-\multicolumn{3}{l}{\textit{Zenith angle 0\degree (6 w-layers/planes)}} \\
+\multicolumn{3}{l}{\textit{Zenith angle 0\degree (6 $w$-layers/planes)}} \\
 \hline
@@
 \hline
-\multicolumn{3}{l}{\textit{Zenith angle 0\degree (128 w-layers/planes)}} \\
+\multicolumn{3}{l}{\textit{Zenith angle 0\degree (128 $w$-layers/planes)}} \\
 \hline
@@
 Sources in the image produced with CASA are slightly less accurately subtracted, leading to a residual noise level of 1.07 mJy and some visible artefacts, whereas WSClean reaches 0.90 mJy.
-Since the effect is stronger away from the phase centre, it is likely that this is caused by the finite size of the w-kernels used in w-projection, which leads to inaccuracies.}
+Since the effect is stronger away from the phase centre, it is likely that this is caused by the finite size of the $w$-kernels used in $w$-projection, which leads to inaccuracies.}
 \label{fig:residuals}
@@
 At zenith, CASA produces a residual RMS twice as high as WSClean.
-This is caused by the fact that we keep the number of $w$-projection planes in w-projection equal to the number of w-layers in w-stacking, resulting in only 12 $w$-planes at zenith.
+This is caused by the fact that we keep the number of $w$-projection planes in $w$-projection equal to the number of $w$-layers in $w$-stacking, resulting in only 12 $w$-planes at zenith.
 This evidently has a stronger effect on the $w$-projection algorithm.
-However, the computational performance of the $w$-projection algorithm is hardly affected by the number of $w$-projection planes, and in practical situations one would always use more w-projection planes.
+However, the computational performance of the $w$-projection algorithm is hardly affected by the number of $w$-projection planes, and in practical situations one would always use more $w$-projection planes.
 When 128 $w$-projection planes are used in CASA, its residual RMS is equal to WSClean with 12 $w$-layers, but the flux density measurements are less accurate.
@@
 WSClean is faster in all tested cases.
-Both imagers have performed 5 major iterations for these results, and major iterations are dominating the computing time.
+Both imagers have performed 5 major iterations for these results, and the inversions and predictions are dominating the computing time.
 We will look more closely at the difference in performance in the next section.
@@
 Additionally, at higher zenith angles, the $w$-kernels become too large to be able to make images of 3072 pixels or larger.
-Using the recentering technique to decrease the $w$-values makes the computational cost approximately constant.
+Using the recentering technique with WSClean to decrease the $w$-values makes the computational cost approximately constant.
 However, the additional time required to rephase and regrid the measurement set makes it only worthwhile for images $\ge3072$ pixels and $\textrm{ZA}\ge15\degree$.
@@
 Our program to change the phase centre is currently not multi-threaded, so its performance might be improvable.
+Furthermore, the relative cost of changing the phase centre is lower when performing multiple major iterations.
 
@@
 Measurements are weighted with the inverse standard deviation.
+While it is normal to weight by inverse variance instead, we found that this results in too much weight on fast configurations, leading to functions that represent the general trend less well.
 The single outlying measurement for $N_\textrm{pix}=1792$ with CASA in Fig.~\ref{fig:timing-resolution} is removed.
@@
 \begin{align}\label{eq:wsclean-computing-cost}
-& t_\textrm{WSClean} (w_e,N_\textrm{Mvis},N_\textrm{freq},N_\textrm{kpix}) = \\ \notag%
-& N_\textrm{freq} \left( 0.452 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} (w_e + 0.611) + 244 N_\textrm{Mvis} \right)
+& t_\textrm{WSClean} (w_{\max},N_\textrm{Mvis},N_\textrm{freq},N_\textrm{kpix}) = \\ \notag%
+& N_\textrm{freq} \left( 0.452 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} (w_{\max} + 0.611) + 244 N_\textrm{Mvis} \right)
 \end{align}
@@
 \begin{align}\label{eq:casa-computing-cost}
-& t_\textrm{CASA}(w_e,N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = \\ \notag%
-& N_\textrm{freq} \left(1.03 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} + 5.93 N_\textrm{Mvis} (w_e + 1.71)^2\right) + 172.
+& t_\textrm{CASA}(w_{\max},N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = \\ \notag%
+& N_\textrm{freq} \left(1.03 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} + 5.93 N_\textrm{Mvis} (w_{\max} + 1.71)^2\right) + 172.
 \label{eq:casa-computing-cost}%
 \end{align}
-Parameter $w_e$ is the maximum $w$ value and is estimated with
+Parameter $w_{\max}$ is the maximum $w$ value and is estimated with
 \begin{equation}
- w_e = \frac{1}{\lambda}\left[\max |\Delta \mathbf{x}| \sin \textrm{ZA} + \max \Delta z \cos \textrm{ZA} \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right)\right],
+ w_{\max} = \frac{1}{\lambda}\left[D \sin \textrm{ZA} + \max \Delta z \cos \textrm{ZA} \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right)\right],
 \end{equation}
-where $\max |\Delta \mathbf{x}|$ is the maximum baseline length, $\max \Delta z$ is the maximum height difference between antennas and $\lambda$ is the wavelength.
+where $D$ is the maximum baseline length, $\max \Delta z$ is the maximum height difference between antennas and $\lambda$ is the wavelength, all in meters.
 These functions follow the trend of the measurements well, with an absolute error of 14.7\% and 20.7\% for respectively the WSClean and CASA functions.
-(TODO $w_e$ is slightly different for CASA)
+(TODO $w_{\max}$ is slightly different for CASA)
 
 \subsection{Optimal snapshot duration}
-Snapshot imaging can in theory be implemented with either $w$-projection or $w$-stacking.
+TODO: need to recalculate this with proper FOV and ZA corrections.
+Snapshot imaging, such as the recentring technique described in \S\ref{sec:snapshot-imaging}, can in theory be implemented with either $w$-projection or $w$-stacking.
 Using the derived formula, we can estimate the cost of making snapshots with both techniques.
-Snapshot imaging effectively removes the dependency on ZA from $w_e$.
+Snapshot imaging effectively removes the dependency on ZA from $w_{\max}$.
 Given the snapshot duration $\Delta \tau_\textrm{snapshot}$ and the total integration time $\Delta \tau_\textrm{total}$, the total time cost of inversion becomes
 \begin{equation} \label{eq:snapshot-cost}
-t_\textrm{snapshot}(\Delta \tau_\textrm{snapshot}) = \frac{\Delta \tau_\textrm{total}}{\Delta \tau_\textrm{snapshot}} t_\textrm{imaging}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
+t_\textrm{snapshot}(\Delta \tau_\textrm{snapshot}) = \frac{\Delta \tau_\textrm{total}}{\Delta \tau_\textrm{snapshot}} t_\textrm{imaging}(w'_{\max},N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
 \end{equation}
@@
 \begin{equation}
-w'_e = \frac{1}{\lambda}\max \Delta z \left(1.0 - \cos(\frac{1}{2}(\textrm{FOV} + \omega_E \tau_\textrm{snapshot})\right),
+w'_{\max} = \frac{1}{\lambda}\max \Delta z \left(1.0 - \cos(\frac{1}{2}(\textrm{FOV} + \omega_E \tau_\textrm{snapshot})\right),
 \end{equation}
@@
 \begin{tabular}{l|rrrrrrrrrr}
- Configuration&$\lambda$& FOV  & Beams & Ant & Res & $\max |\Delta \mathbf{x}|$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ \\
+ Configuration&$\lambda$& FOV  & Beams & Ant & Res & $D$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ \\
           & (m)       & (FWHM)&       &     &     & (km)                       & (m)             & (s)    & (MHz) &  (kHz) \\
@@
  MeerKAT  & 0.2  &  1\degree  & 1    &  64 &   6'' &  8 & 1   & 0.5 & 750  & 50 \\
- SKA-P1 AA inner&3& 1.2\degree& 480  &  35 & 140'' &2.5& 2 &  8 & 380 & 80 \\
- SKA-P1 AA +mid&3 & 1.2\degree & 480  &  50 & 3.5'' &100& 50 & 0.2 & 380 & 2 \\
+% SKA1 AA inner&2 &  5\degree  & 1    & 433 &  15'' &0.6& 0.5 &  10 & 250 & 1 \\
+ SKA1 AA core&2  &  5\degree  & 1    & 866 &    3' &  3& 5 & 10 & 250 & 1 \\
+ SKA1 AA full&2  &  5\degree  & 1    & 911 &   5'' &100& 50 & 0.6& 250 & 1 \\
 \end{tabular}
@@
 All have approximately equal accuracy.} \label{tbl:computational-cost-predictions}
-\begin{tabular}{l|rrrr|r}
-               & \multicolumn{4}{c|}{Predicted computing cost on test computer} & max \\
- Configuration & $w$-projection & $w$-stacking & $w$-snapshot & hybrid & FLOPS/float \\
+\begin{tabular}{l|rrrr|rr}
+               & \multicolumn{4}{c|}{Predicted computing cost on test computer} & min & best \\
+ Configuration & $w$-projection & $w$-stacking & $w$-snapshot & hybrid & computing & FLOPS/float \\
  \hline
- GLEAM    &   67h 7m & 8h 31m &  8h 05m & 15h 18m & $6.9 \times 10^2$    \\
- EMU      &    5.2 d &  2.9 d &   2.9 d & 5.2 d   & $2.1 \times 10^4$      \\
- MSSS low &  2h 18m  & 0h 59m &  0h 58m & 2h 16m  & $3.5 \times 10^3$   \\
- MSSS high&  7h 16m  & 3h 52m &  3h 52m & 7h 16m  & $3.4 \times 10^3$   \\
- LOFAR NL &  50.6 d  &   10 d &   7.2 d & 15.9 d & $4.5 \times 10^3$ \\
- AARTFAAC &  71h 25m &29h 50m & 29h 50m & 51h 34m & $6.8 \times 10^2$   \\
- VLSS     &  0h 12m  & 0h 12m &  0h 02m &  0h 12m & $12.1 \times 10^3$     \\
- MeerKAT  &   10.8 d &  6.2 d &   6.2 d &  10.8 d & $6.8 \times 10^2$ \\
- SKA-P1 AA inner&7.2 d& 2.9 d &   2.9 d &   3.7 d & $3.3 \times 10^5$   \\
- SKA-P1 AA mid&100,996 d&57,605 d& 57,605 d & 99,402 d & $3.3 \times 10^5$ \\
+ GLEAM    &   67h 7m & 8h 31m &  8h 05m & 15h 18m & 1.2 TFLOPS & $6.9 \times 10^2$    \\
+ EMU      &    5.2 d &  2.9 d &   2.9 d & 5.2 d   & 10 TFLOPS & $2.1 \times 10^4$      \\
+ MSSS low &  2h 18m  & 0h 59m &  0h 58m & 2h 16m  & 140 GFLOPS & $3.5 \times 10^3$   \\
+ MSSS high&  7h 16m  & 3h 52m &  3h 52m & 7h 16m  & 530 GFLOPS & $3.4 \times 10^3$   \\
+ LOFAR NL &  50.6 d  &   10 d &   7.2 d & 15.9 d & 33 TFLOPS & $4.5 \times 10^3$ \\
+ AARTFAAC &  71h 25m &29h 50m & 29h 50m & 51h 34m & 4.1 TFLOPS & $6.8 \times 10^2$   \\
+ VLSS     &  0h 12m  & 0h 12m &  0h 02m &  0h 12m & 4.1 GFLOPS & $12.1 \times 10^3$     \\
+ MeerKAT  &   10.8 d &  6.2 d &   6.2 d &  10.8 d & 21 TFLOPS & $6.8 \times 10^2$ \\
+% SKA1 AA inner&392 d &  228 d &   228 d &   392 d & & $6.8 \times 10^2$   \\
+ SKA1 AA core&1581 d &  911 d &   911 d &  1571 d &  3 PFLOPS & $6.8 \times 10^2$ \\
+ SKA1 AA full& 633 yr& 48.8 yr & 48.8 yr& 84.2 yr & 59 PFLOPS & $6.8 \times 10^2$ \\
 \end{tabular}
@@
 The AWImager has been applied like this, which let to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}.
-For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more generally the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_e$ value and smaller $N_\textrm{vis}$.
+For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more generally the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_{\max}$ value and smaller $N_\textrm{vis}$.
 If $\Delta w$ is the distance between $w$-layers, then
 \begin{equation}
-t(\Delta w) = \frac{w_e}{\Delta w} t_\textrm{CASA}(w'_e,N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
+t(\Delta w) = \frac{w_{\max}}{\Delta w} t_\textrm{CASA}(w'_{\max},N'_\textrm{vis},N_\textrm{freq},N_\textrm{pix}),
 \end{equation}
-with $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta w}{w_e}$ and $w'_e = \Delta w$.
+with $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta w}{w_{\max}}$ and $w'_{\max} = \Delta w$.
 For MWA observations, the optimal value for $\Delta w$ is about unity, and improves the speed of $w$-projection with about a factor of four.
@@
 We use the derived functions to estimate the computational cost of the algorithms for configurations of several telescopes.
-We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multifrequency Snapshot Sky Survey (MSSS)\footnote{See \url{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam—ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \url{http://www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The Meer Karoo Array Telescope (MeerKAT); and the low-frequency phase~1 aperture arrays of the Square-Kilometre Array, with inner (2.5~km) and mid (2.5~km) configurations \citep{ska-phase1-2010}.
+We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multifrequency Snapshot Sky Survey (MSSS)\footnote{See \url{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam—ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \url{http://www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The Meer Karoo Array Telescope (MeerKAT); and the low-frequency phase~1 aperture arrays of the Square-Kilometre Array, with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}.
 The configurations are summarized in Table~\ref{tbl:computational-cost-configurations}.
 In multi-beam configurations such as the SKA and EMU configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams.
+The image size is determined by 5 times the field of view divided by resolution.
+The factor of five is to allow better cleaning accuracy and source-position fitting, as is common.
+The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).
 
@@
 
+Imaging the full field of view with the full SKA becomes very expensive due to the resolution of $18\textrm{k} \times 18$k pixels. This translates to a computing power of $\sim60$ PetaFLOPS.
+
 \section{Conclusions} \label{sec:conclusions}
@@
 This can lead to performance improvements of a factor of $3$ for the MWA, but only in the most expensive imaging configurations that are less commonly used.
-Considering the extra regridding step required, which complicates issues such as calculating the integrated beam shape, recentering snapshots might not be worthwhile for the MWA.
+Considering the extra regridding step required, which complicates issues such as calculating the integrated beam shape, recentering snapshots is worthwhile for the MWA only at very low elevations or large fields of view.
 Extrapolation of the computing cost predicts that this holds for most arrays, including SKA low.
-A hybrid between $w$-stacking and $w$-projection might improve things by a few factors.
+A hybrid between $w$-stacking and $w$-projection seems not to improve performance over a pure $w$-stacking implementation, but does improve a pure $w$-projection implementation significantly.
 
-In \citet{ska-phase1-2010}, the SKA computing power is budgetted by assuming an imaging speed of $10^4$ FLOPS/float.
-Extrapolation of our results shows that the current imagers perform $30$--$70$ times slower than this on SKA data.
-It was suggested that $w$-snapshots would improve the imaging speed for the SKA situation \citep{widefield-imaging-ska-cornwell}, but our results do not confirm this.
-This is mainly because the primary beam of a SKA station is not very large (1.2\degree), which causes the $w$-terms to not have a significant effect on performance: the bulk of the time spent is on gridding the antialiasing kernels instead of the $w$-kernels.
-Clearly it remains challenging to perform widefield imaging with acceptable performance.
-Nevertheless, some optimisations could be made for the SKA.
-For example, a performance gain of a few factors can already be achieved by averaging shorter baselines to their maximum integration time and bandwidth resolution before imaging.
+The available SKA computing power is estimated to be around $\sim$100 PetaFLOPS.
+Extrapolation of our results shows that the current imagers require $3$--$60$ petaFLOPS for SKA1 low alone.
+It was suggested that $w$-snapshots would improve the imaging speed for the SKA situation \citep{widefield-imaging-ska-cornwell}, but our results show this is not the case when already using the $w$-stacking algorithm.
+Clearly it remains challenging to perform widefield imaging with acceptable performance, but some optimisations could be made for the SKA.
+For example, a performance gain of a few factors can be achieved by averaging shorter baselines to their maximum integration time and bandwidth resolution before imaging.
+Moreover, although longer baselines make the imaging more expensive, it is likely that it is not often required to image the full field of view when using the longest baselines.
 
@@
 \section*{Acknowledgments}
-...
+This scientific work makes use of the Murchison Radio-astronomy Observatory, operated by CSIRO.
+We acknowledge the Wajarri Yamatji people as the traditional owners of the Observatory site.
+Support for the MWA comes from the U.S.
+National Science Foundation (grants AST-0457585, PHY-0835713, CAREER-0847753, and AST-0908884), the Australian Research Council (LIEF grants LE0775621 and LE0882938), the U.S.
+Air Force Office of Scientic Research (grant FA9550-0510247), and the Centre for All-sky Astrophysics (an Australian Research Council Centre of Excellence funded by grant CE110001020).
+Support is also provided by the Smithsonian Astrophysical Observatory, the MIT School of Science, the Raman Research Institute, the Australian National University, and the Victoria University of Wellington (via grant MED-E1799 from the New Zealand Ministry of Economic Development and an IBM Shared University Research Grant).
+The Australian Federal government provides additional support via the Commonwealth Scientific and Industrial Research Organisation (CSIRO), National Collaborative Research Infrastructure Strategy, Education Investment Fund, and the Australia India Strategic Research Fund, and Astronomy Australia Limited, under contract to Curtin University.
+We acknowledge the iVEC Petabyte Data Store, the Initiative in Innovative Computing and the CUDA Center for Excellence sponsored by NVIDIA at Harvard University, and the International Centre for Radio Astronomy Research (ICRAR), a Joint Venture of Curtin University and The University of Western Australia, funded by the Western Australian State government.
 \appendix

### 2014-02-17 articles/2014/Article-WSClean/img/fit.cpp

@@ -62,8 +62,8 @@ std::string timeToStr(double secComputingRequired, double secObservation, double
 	else
 		str << round(gFlopsRequired*10.0)/10.0 << " GFLOPS";
 	str << ", ";
-	double perVis = secComputingRequired * BATTLESTAR_GFLOPS_PER_SECOND / nMVis;
-	str << perVis << " or " << round(20.0*perVis / exp10(floor(log10(perVis))))/10.0 << " x 10^" << floor(log10(perVis)) << " FLOPS/float";
+	double perFloat = 2.0 * secComputingRequired * BATTLESTAR_GFLOPS_PER_SECOND / nMVis;
+	str << perFloat << " or " << round(10.0*perFloat / exp10(floor(log10(perFloat))))/10.0 << " x 10^" << floor(log10(perFloat)) << " FLOPS/float";
 	
 	return str.str();
 }
@@ -709,7 +709,7 @@ void optimalDeltaW(FittingFunc func, double a, double b, double c, double d, dou
 void optimalDeltaT(FittingFunc func, const std::string& filename, double a, double b, double c, double d, double e, double f, double za, double nVis, double nPix, double fov, double nChan, bool zeroConstant, double& dt, size_t beamsTimesIters, double lambda, double maxBaseline, double maxHeight)
 {
 	std::ofstream file(filename);
-	double minY = std::numeric_limits<double>::max(), minDT = 0.0;
+	double minY = std::numeric_limits<double>::max(), minDT = 0.0, minFOV = 0.0;
 	double dtTrial = 1.0, prevTrial = 1.0;
 	double fz = zeroConstant ? 0.0 : f;
 	const double totalTime = 60.0*60.0;
@@ -722,6 +722,7 @@ void optimalDeltaT(FittingFunc func, const std::string& filename, double a, doub
 		if(y < minY) {
 			minDT = dtTrial;
 			minY = y;
+			minFOV = trialFOV;
 		}
 		file << dtTrial << '\t' << y << '\n';
 		prevTrial = dtTrial;
@@ -730,7 +731,7 @@ void optimalDeltaT(FittingFunc func, const std::string& filename, double a, doub
 	dt = minDT;
[…]

### 2014-02-17 articles/2014/Article-WSClean/wsclean.tex
Correcting affiliations

@@
 \def\CAASTRO{$^{2}$}
-\def\ASU{$^{3}$}
-\def\CASS{$^{4}$}
-\def\Curtin{$^{5}$}
-\def\CfA{$^{6}$}
-\def\Haystack{$^{7}$}
-\def\MIT{$^{8}$}
-\def\NRAO{$^{9}$}
-\def\RRI{$^{10}$}
-\def\SKASA{$^{11}$}
-\def\Tata{$^{12}$}
-\def\UMelbourne{$^{13}$}
-\def\USydney{$^{14}$}
-\def\UW{$^{15}$}
-\def\UWisc{$^{16}$}
-\def\Victoria{$^{17}$}
+\def\Curtin{$^{3}$}
+\def\UWisc{$^{4}$}
+\def\MIT{$^{5}$}
+\def\SKASA{$^{6}$}
+\def\ASU{$^{7}$}
+\def\Haystack{$^{8}$}
+\def\RRI{$^{9}$}
+\def\USydney{$^{10}$}
+\def\CfA{$^{11}$}
+\def\UW{$^{12}$}
+\def\Victoria{$^{13}$}
+\def\CASS{$^{14}$}
+\def\Tata{$^{15}$}
+\def\NRAO{$^{16}$}
+\def\UMelbourne{$^{17}$}
 
-\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\hspace{2cm}E-mail:
+\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{E-mail:
 \url{andre.offringa@anu.edu.au}},
@@
 F.~H.~Briggs$^{1,2}$, \newauthor
+R.~B.~Wayth\Curtin$^,$\CAASTRO, 
+D.~L.~Kaplan\UWisc,
+L.~Feng\MIT,
+A.~R.~Neben\MIT,
+J.~D.~Hughes\ANU, \newauthor
 J.~Rhee\ANU,
-L.~Feng,
-D.~L.~Kaplan\UWisc,
-A.~R.~Neben,
-R.~B.~Wayth\Curtin$^,$\CAASTRO, \newauthor
-J.~D.~Hughes\ANU,
 G.~Bernardi\SKASA,
 J.~D.~Bowman\ASU,
-F.~Briggs\ANU,
 R.~J.~Cappallo\Haystack,
@@
 S.~M.~Ord\Curtin$^,$\CAASTRO,
-T.~Prabu\RRI, \newauthor
-A.~E.~E.~Rogers\Haystack,
+T.~Prabu\RRI, 
+A.~E.~E.~Rogers\Haystack, \newauthor
 D.~A.~Roshi\NRAO \footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
@@
 \\
-$^{1}$Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
-$^{2}$ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
-$^{3}$School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
-$^{4}$CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
-$^{5}$International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
-$^{6}$Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
-$^{7}$MIT Haystack Observatory, Westford, MA 01886, USA\\
-$^{8}$Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
-$^{9}$National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
-$^{10}$Raman Research Institute, Bangalore 560080, India\\
-$^{11}$Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
-$^{12}$National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
-$^{13}$School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
-$^{14}$Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
-$^{15}$Department of Physics, University of Washington, Seattle, WA 98195, USA\\
-$^{16}$Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
-$^{17}$School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
+\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
+\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
+\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
+\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
+\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
+\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town 7405, South Africa\\
+\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
+\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
+\RRI{}Raman Research Institute, Bangalore 560080, India\\
+\USydney{}Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
+\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
+\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
+\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
+\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
+\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
+\NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
+\UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
 }
@@
 \begin{abstract}
-We present a new imager implementation that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm.
+We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm.
 On MWA data, we find it to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution.
@@
 We predict the computing costs for several other configurations and estimate that our imager is approximately a factor of 2--3 faster on other arrays.
-Imaging SKA observations is challenging, and SKA1 low will require 3--60 PetaFLOPS of computing power with current techniques.
+Imaging SKA observations is challenging, as the full SKA1 low will require 60 PetaFLOPS of computing power with current techniques.
 We find that the $w$-snapshot algorithm does not further improve computing requirements.
@@
 TODO: need to recalculate this with proper FOV and ZA corrections.
-Snapshot imaging, such as the recentring technique described in \S\ref{sec:snapshot-imaging}, can in theory be implemented with either $w$-projection or $w$-stacking.
+Snapshot imaging, such as the recentring technique described in \S\ref{sec:snapshot-imaging}, can be implemented with either $w$-projection or $w$-stacking.
 Using the derived formula, we can estimate the cost of making snapshots with both techniques.
@@
 \begin{table*}
-\caption{Configurations for which the computational cost are predicted.} \label{tbl:computational-cost-configurations}
+\caption{Configurations for which the computational cost is predicted.} \label{tbl:computational-cost-configurations}
 \begin{tabular}{l|rrrrrrrrrr}
@@
 In $w$-stacking, the speed gain associated with this optimisation is less significant.
-Limitting the $w$-values also lowers the snapshots resolution in one direction significantly, because limitting all the long baselines in one direction are no longer gridded, which is often not desirable for the MWA.
+Limiting the $w$-values also lowers the snapshots resolution in one direction significantly, because the long baselines in one direction are no longer gridded, which is often not desirable for the MWA.
 
@@
 In multi-beam configurations such as the SKA and EMU configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams.
-The image size is determined by 5 times the field of view divided by resolution.
+The image size is set to 5 times the FOV width divided by the resolution.
 The factor of five is to allow better cleaning accuracy and source-position fitting, as is common.

### 2014-02-17 articles/2014/Article-WSClean/img/fit.cpp

@@ -715,9 +715,11 @@ void optimalDeltaT(FittingFunc func, const std::string& filename, double a, doub
 	const double totalTime = 60.0*60.0;
 	while(dtTrial < totalTime)
 	{
-		double trialFOV = fov + 2.0 * (dtTrial - 112.0) * (360.0 / 60.0 / 60.0 / 24.0);
+		double tAngle = dtTrial * (360.0 / 60.0 / 60.0 / 24.0);
+		double trialFOV = fov + sin(za) * 2.0;
+		double trialZA = tAngle * 0.5;
 		if(trialFOV < 0.0) trialFOV = 0.0;
-		double y = (totalTime/dtTrial) * eval(func, 0.0, nVis * dtTrial/totalTime, nPix, trialFOV, nChan, a, b, c, d, e, fz, lambda, maxBaseline, maxHeight);
+		double y = (totalTime/dtTrial) * eval(func, trialZA, nVis * dtTrial/totalTime, nPix, trialFOV, nChan, a, b, c, d, e, fz, lambda, maxBaseline, maxHeight);
 		//std::cout << "za=" << trialFOV<< ", dt=" << dtTrial << ", y=" << y << '\n';
 		if(y < minY) {
 			minDT = dtTrial;

### 2014-02-17 articles/2014/Article-WSClean/wsclean.tex
Some corrections

@@
 
-\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{E-mail:
-\url{andre.offringa@anu.edu.au}},
+\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\hspace{2cm}E-mail: \url{andre.offringa@anu.edu.au}},
 B.~McKinley\ANU$^,$\CAASTRO,
@@
 We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm.
-On MWA data, we find it to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution.
+On MWA data, we find it to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with full polarization correction.
 The performance dependencies of CASA's $w$-projection and WSClean are analysed and analytical functions are derived that describe the required computing cost for both imagers.
@@
 We predict the computing costs for several other configurations and estimate that our imager is approximately a factor of 2--3 faster on other arrays.
-Imaging SKA observations is challenging, as the full SKA1 low will require 60 PetaFLOPS of computing power with current techniques.
+Imaging SKA observations is challenging, as we estimate the computing cost of full SKA1 low imaging to be 60 PetaFLOPS with current techniques.
 We find that the $w$-snapshot algorithm does not further improve computing requirements.
@@
 
-Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with a two-dimensional Fast Fourier Transformation (FFT).
+Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with a two-dimensional fast Fourier transform (FFT).
 Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}.
@@
 
-For small field of views, the term $\sqrt{1-l^2-m^2}$ is approximately of unit size, making Eq.~\eqref{eq:visibility-function} approximate an ordinary 2-dimensional Fourier transform that can be inverted with an inverse Fast Fourier Transform (FFT).
+For small field of views, the term $\sqrt{1-l^2-m^2}$ is approximately of unit size, making Eq.~\eqref{eq:visibility-function} approximate an ordinary 2-dimensional Fourier transform that can be inverted with an inverse FFT.
 A common rule is that this is valid when
@@
 The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation.
-This shows that the sky function can be reconstructed by i)~grid samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~apply the direction dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeat this for all $w$-values and add the results together; v)~apply the final scaling.
+This shows that the sky function can be reconstructed by i)~grid samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~apply the direction-dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeat this for all $w$-values and add the results together; v)~apply the final scaling.
 
@@
 Therefore, the number of added samples and their weight have to be taken into account.
-Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction dependent effects.
+Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction-dependent effects.
 In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds.
@@
 It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation.
-We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each dimension, $N_\textrm{vis}$ is the number of visibilities and $N_\textrm{kern}$ is the size of the antialiasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.
+We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each dimension, $N_\textrm{vis}$ is the number of visibilities, $N_\textrm{kern}$ is the size of the antialiasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.
 
@@
 Several steps are multi-threaded using the threading module of the C++11 standard library.
-These are: reading and writing; gridding from and to different $w$-layers; performing the Fast Fourier transforms; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}).
+These are: reading and writing; gridding from and to different $w$-layers; performing the FFTs; and H\"ogbom Clean iterations (peak-finding and image subtraction, \citealt{hogbom-clean}).
 For the latter, intrinsics are used as well.
@@
 Depending on desired resolution and dynamic range, this might require very large images.
-To make an image at the MWA resolution, the image needs to have approximately 10K pixels in width and height.
+To make an image at the MWA resolution, the image needs to have approximately 10k pixels in width and height.
 The $w$-values will be very small at zenith, because at zenith only the vertical offsets of the antennas will contribute to the $w$-value.
@@
 In the following sections, when discussing the MWA zenith direction we are referring to this optimal $w$-direction.
+
+TODO: add big-ass full-sky image
 
@@
 
+TODO: I just realized that the original definition of $w$-snapshots might be Natasha's way of fixing $w$-values in MWACS. In that case, these methods are new and it might be good to distinguish this.
+
 \subsection{Beam correction}
@@
 Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such.
-However, this is not correct, because $X$ is defined as being parallel to the celestial equator.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters.
+However, this is incorrect because i) $X$ is normally defined as being parallel to the horizon; and ii) ($2\times$) Stokes I$\neq pp + qq$.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters.
 The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry.
@@
 The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.
+
+TODO: add Stokes I and V images of polarized source (e.g. Martin's source, if he agrees)
 
@@
 When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate.
-WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation for the PSWF \citep{fourier-kernel-selection-1991}.
+WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation of the PSWF \citep{fourier-kernel-selection-1991}.
 After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that 7 pixels is a good compromise between performance and accuracy, and this is the default for WSClean.
@@
 As can be seen in Panel~\ref{fig:timing-fov}, the field of view (FOV) has a large affect on performance when imaging off-zenith.
-This plot was created by varying the size of the pixel in the output image, such that the number of pixels in the image did not change.
-The FOV is calculated as angle between the left and right most pixel in the image.
-CASA is not able to make off-zenith images larger than $\sim30$\degree, and its imaging cost increases much more rapidly compared to the cost of WSClean, which implies that gridding is the major cost in this scenario.
-The image recentering technique is clearly beneficial for FOVs larger than $\sim30$\degree, and can even make a difference of an order of magnitude at very large FOVs of $\sim90$\degree.
+This plot was created by varying the size of a pixel in the output image, such that the number of pixels in the image did not change.
+The FOV is calculated as angle between the left- and right-most pixel in the image.
+CASA is not able to make off-zenith images larger than $\sim$30\degree, and its imaging cost increases much more rapidly compared to the cost of WSClean, which implies that gridding is the major cost in this scenario.
+The image recentering technique is clearly beneficial for FOVs larger than $\sim$30\degree, and can even make a difference of an order of magnitude at very large FOVs of $\sim$90\degree.
 
@@
 \subsection{Optimal snapshot duration}
-TODO: need to recalculate this with proper FOV and ZA corrections.
 Snapshot imaging, such as the recentring technique described in \S\ref{sec:snapshot-imaging}, can be implemented with either $w$-projection or $w$-stacking.
-Using the derived formula, we can estimate the cost of making snapshots with both techniques.
+Using the derived formulae, we can estimate the cost of making snapshots with both techniques.
 Snapshot imaging effectively removes the dependency on ZA from $w_{\max}$.

### 2014-02-18 articles/2014/Article-WSClean/wsclean.tex
Slight updates

@@
 
-TODO: add big-ass full-sky image
+TODO: add big full-sky image
 
@@
 Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
-%When the Fourier transform is performed, the $w$ term has been corrected (either by $w$-stacking or by $w$-projection).
-Therefore, if $U(u, v)$ is the input to the 2-dimensional Fourier transform, the transform is described by
 \begin{align}\notag
@@
 This essentially is a method of implementing the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}.
-By doing the inverse corrections during prediction, such a shifted visibility set can be cleaned with Cotton-Schwab iterations similar to a non-shifted set.
+However, instead of creating warped images by performing the FFT over a tilted plane in $uvw$-space, our method transforms planes with constant $w$-values.
+The benefit of our method is that the resulting image has a circular synthesized beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning.
+Our method is also relatively easy to implement.
+It has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on zenith angle, field of view and the non-coplanarity.
+
+By doing the inverse corrections during prediction, a recentred visibility set can be cleaned with Cotton-Schwab iterations similar to a non-recentred set.
+The PSF used during minor cleaning iterations is created by multiplying the weights with the recentering corrections, such that the PSF represents a source in the middle of the image.
 
@@
 Common viewers such as Kvis and DS9 support such FITS files and display their coordinates correctly.
-An example of the difference in projection is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13~min MWA observation imaged with WSClean.
+Unlike warped snapshots, recentred images do not need the generalized-SIN-projection keywords \texttt{PV2\_1} and \texttt{PV2\_2}.
+An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13~min MWA observation imaged with WSClean.
 Cleaning an image that is in a different projection yields slightly different results, but qualitativily the two images are clearly of equal accuracy.
@@
 Therefore, we chose to implement the former method.
-
-TODO: I just realized that the original definition of $w$-snapshots might be Natasha's way of fixing $w$-values in MWACS. In that case, these methods are new and it might be good to distinguish this.
 
@@
 \end{align}
-Parameter $w_{\max}$ is the maximum $w$ value and is estimated with
-\begin{equation}
- w_{\max} = \frac{1}{\lambda}\left[D \sin \textrm{ZA} + \max \Delta z \cos \textrm{ZA} \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right)\right],
-\end{equation}
-where $D$ is the maximum baseline length, $\max \Delta z$ is the maximum height difference between antennas and $\lambda$ is the wavelength, all in meters.
+Parameter $w_{\max}$ is the maximum $w$-value and is estimated with
+\begin{equation}\resizebox{0.45\textwidth}{!}{$
+ w_{\max} = \frac{1}{\lambda}\left[(D \sin \textrm{ZA} + \Delta z_{\max} \cos \textrm{ZA} + \xi) \left(1.0 - \cos(\frac{1}{2}\textrm{FOV})\right)\right]$},
+\end{equation}%
+where $D$ is the maximum baseline length, $\Delta z_{\max}$ is the maximum height difference between antennas and $\lambda$ is the wavelength, all in meters, and $\xi$ is an extra parameter for fitting the CASA measurements, $\xi_\textrm{CASA}=28.4$ and $\xi_\textrm{WSClean}=0$.
 These functions follow the trend of the measurements well, with an absolute error of 14.7\% and 20.7\% for respectively the WSClean and CASA functions.
-(TODO $w_{\max}$ is slightly different for CASA)
 

### 2014-02-21 articles/2014/Article-WSClean/wsclean.tex
Added Martin's images of Stokes V pulsar detection

@@
 D.~L.~Kaplan\UWisc,
+M.~E.~Bell\USydney$^,$\CAASTRO,
 L.~Feng\MIT,
@@
 Imaging SKA observations is challenging, as we estimate the computing cost of full SKA1 low imaging to be 60 PetaFLOPS with current techniques.
-We find that the $w$-snapshot algorithm does not further improve computing requirements.
+We find that the $w$-snapshot algorithm does not further improve computing requirements significantly.
 The source code of our new imager is publicly released.
@@
 \subsection{Beam correction}
+\begin{figure*}
+\begin{center}
+\includegraphics[width=6cm]{img/pulsar-stokesi}\hspace{1cm}\includegraphics[width=6cm]{img/pulsar-stokesv}
+\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with WSClean.
+The pulsar (centre of image) displays 17\% circular polarization.}
+\label{fig:pulsar-stokes-iv}
+\end{center}
+\end{figure*}
 Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix.
@@
 Once these six images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}.
-This method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam.
+The method is demonstrated in Fig.~\ref{fig:pulsar-stokes-iv}, which displays a detection of pulsar J0437-4715 in Stokes V.
+The image was made with an initial pipeline for the radio-sky monitor project that uses the MWA to search for transients, including exoplanets (TODO cite).
+The pipeline includes WSClean to make the Stokes-parameter images.
+
+Our image-based beam correction method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam.
 To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm excessively unless an a-projection kernel is used.
 The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.
-
-TODO: add Stokes I and V images of polarized source (e.g. Martin's source, if he agrees)
 

### 2014-02-21 articles/2014/Article-WSClean/wsclean.tex
Small details, and adding joined polarization deconvolution

@@
 This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}.
+WSClean allows cleaning individual polarizations, or can jointly deconvolve the polarizations.
+In the latter case, peaks are searched in $pp^2+2pq \overline{(pq)} + qq^2$ space, where $\overline{pq}=qp$ is the complex conjugate of $pq$, and the PSF is subtracted from the polarizations with different factors.
 
@@
 Image size & $3072 \times 3072$ \\
-Angular resolution & $0.72'$ \\
+Angular pixel size & $0.72'$ \\
 Number of visibilities & $3.5 \times 10^8$ \\
@@
 Therefore, from a performance perspective, the snapshot duration should be as large as possible.
+In practise, the beam needs to be corrected on small timescales.
+A snapshot duration of more than a few minutes is therefore not possible.
 For $w$-projection in CASA we find an optimal snapshot duration of $\sim2$ min for MWA observations.

### 2014-02-25 articles/2014/Article-WSClean/wsclean.tex
Small changes

@@
 \end{equation}
-This suggest that a uniform discretization in $w$ is optimal (TODO this is in contrast to what Cornwell 2008 says -- Cornwell mailed me he made a mistake...
-Do we need to explicitly point that out here :-/ ?).
+This suggest that a uniform discretization in $w$ is optimal.
+This is in contrast to \citet{wprojection-cornwell} where $\sqrt{w}$ tabulation is suggested.
 From Eq.~\eqref{eq:minimum-w-distance}, the required number of layers can be derived and is given by
@@
 Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such.
-However, this is incorrect because i) $X$ is normally defined as being parallel to the horizon; and ii) ($2\times$) Stokes I$\neq pp + qq$.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters.
+However, this is incorrect because they are not necessarily orthogonal, i.e., ($2\times$) Stokes I$\neq pp + qq$.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters.
 The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry.
 To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images.
-Each polarization requires a separate run of the $w$-stacking algorithm.
-Once these six images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}.
+This requires four runs of the algorithm.
+Once these images are created, a separate program is used to perform the correction of Eq.~\eqref{eq:beam-correction}.
 The method is demonstrated in Fig.~\ref{fig:pulsar-stokes-iv}, which displays a detection of pulsar J0437-4715 in Stokes V.
-The image was made with an initial pipeline for the radio-sky monitor project that uses the MWA to search for transients, including exoplanets (TODO cite).
+The image was made with an initial pipeline for the radio-sky monitor project that uses the MWA to search for transients and exoplanets.
 The pipeline includes WSClean to make the Stokes-parameter images.
 
-Our image-based beam correction method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam.
+Our image-based beam-correction method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam.
 To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm excessively unless an a-projection kernel is used.
@@
 When 128 $w$-projection planes are used in CASA, its residual RMS is equal to WSClean with 12 $w$-layers, but the flux density measurements are less accurate.
-We do not know why this parameter needs to be higher in CASA to reach the same RMS (TODO: do I want to find out this is because of invalid sqrt(w) gridding?).
+We do not know why this parameter needs to be higher in CASA to reach the same RMS.
 Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlayers}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree.

### 2014-02-25 articles/2014/Article-WSClean/wsclean.tex
Small updates

@@
 Several functions with different free parameters were tested, and formulae with minimum amount of parameters are selected that still follow the trend of the measurements and have reasonably small errors.
-Measurements are weighted with the inverse standard deviation.
-While it is normal to weight by inverse variance instead, we found that this results in too much weight on fast configurations, leading to functions that represent the general trend less well.
+Measurements are weighted with the inverse standard deviation instead of the variance, because we found that the latter results in too much weight on fast configurations, leading to functions that represent the general trend less well.
 The single outlying measurement for $N_\textrm{pix}=1792$ with CASA in Fig.~\ref{fig:timing-resolution} is removed.
@@
 \begin{equation}
-w'_{\max} = \frac{1}{\lambda}\max \Delta z \left(1.0 - \cos(\frac{1}{2}(\textrm{FOV} + \omega_E \tau_\textrm{snapshot})\right),
+w'_{\max} = \frac{1}{\lambda}\max \Delta z \left(1.0 - \cos(\frac{1}{2}(\textrm{FOV} + \omega_E \Delta \tau_\textrm{snapshot})\right),
 \end{equation}
@@
 Accurate MWA beam models are still being developed.
-When beam updates are required often, calculating $a$-projection kernels or performing snapshot imaging also becomes more expensive, and it would be interesting to find out where the balance lies between these methods.
+When beam corrections are required on short time scales, calculating $a$-projection kernels or performing snapshot imaging also becomes more expensive, and it would be interesting to find out where the balance lies between these methods.
 When only a few thousand components need to be deconvolved, an approach with direct Fourier transforms is accurate and affordable.
@@
 with the SAGECAL \citep{sage-calibration-ii} or RTS peeling \citep{rts-mwa} calibration techniques.
-This is not possible for fields with diffuse emmission for which no model is known a-priory.
+This is not possible for fields with diffuse emmission or faint point sources when no model is known a-priory.
 

### 2014-02-27 articles/2014/Article-WSClean/wsclean.tex
Work on article
Final details paper

@@
 We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm.
-On MWA data, we find it to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with full polarization correction.
-The performance dependencies of CASA's $w$-projection and WSClean are analysed and analytical functions are derived that describe the required computing cost for both imagers.
-The difference in computing cost is mainly caused by the efficiency of our new imager on MWA's 27\degree field of view.
-We predict the computing costs for several other configurations and estimate that our imager is approximately a factor of 2--3 faster on other arrays.
+The performance dependencies of CASA's $w$-projection and our new imager are analysed and analytical functions are derived that describe the required computing cost for both imagers.
+On MWA data, we find our new method to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with full polarization correction.
+We predict the computing costs for several other arrays and estimate that our imager is a factor of 2--12 faster, depending on the array configuration.
 Imaging SKA observations is challenging, as we estimate the computing cost of full SKA1 low imaging to be 60 PetaFLOPS with current techniques.
-We find that the $w$-snapshot algorithm does not further improve computing requirements significantly.
+We find that combining $w$-stacking with the $w$-snapshot algorithm does not significantly improve computing requirements.
 The source code of our new imager is publicly released.
@@
 New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before.
-Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}).
+Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa-design-2009}, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}).
 The Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution.
@@
 
-TODO: add big full-sky image
+TODO: add big full-sky image (PKS J2358-6054 is visible)
 
@@
 No beam correction was applied.
-Computing cost was 60 min for the normal projected image and 41 min for the recentred image.}
+Imaging computing cost was 60 min for the normal projected image and 41 min for the recentred image.}
 \label{fig:vela-projection-example}
@@
 Unlike warped snapshots, recentred images do not need the generalized-SIN-projection keywords \texttt{PV2\_1} and \texttt{PV2\_2}.
+
 An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13~min MWA observation imaged with WSClean.
+The processing steps performed for this image were i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the AOFlagger \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibrate on Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) image each of the seven 112 s intervals separately.
 Cleaning an image that is in a different projection yields slightly different results, but qualitativily the two images are clearly of equal accuracy.
@@
 A common kernel function is a windowed sinc function, which acts as a low-pass filter.
-This decreases the flux of sources outside the field of view, and thus helps to attenuate aliased ghost sources and sidelobes that arise because of the assumed periodicity during the inverse FFT \citep{post-correlation-filtering}.
+This decreases the flux of sources outside the field of view, and thus helps to attenuate aliased ghost sources and sidelobes \citep{post-correlation-filtering}.
 By supersampling, a convolution kernel also makes it possible to place samples more accurately at their $uv$ position, thereby lowering decorrelation.
@@
 
+\subsection{Decreasing inversion resolution} \label{sec:decreasing-inversion-resolution}
+Typically, for cleaning it is desired that images have a pixel size at least 5 times smaller than the synthesized-beam width.
+This improves the accuracy of the clean algorithm, because the positions of image maxima will be closer to the actual source positions.
+This factor of $\sim$5 is normally taken into account in the overal imaging resolution, i.e., the image size is increased during inversion.
+For the $w$-projection method, the cost of inversion with an increased resolution is small, because it does not increase the size of the $w$-kernels.
+It will affect the inverse FFT, but the relative cost of this step is negligable in the total cost.
+The $w$-stacking method is however significant affected by the image resolution, because it performs many inverse FFTs.
+
+The spatial frequencies in the output image are band limited by the synthesized beam.
+Therefore, it is not required to perform the inversion at higher resolution.
+As long as the $uv$-plane is sampled with at least the Nyquist frequency, a high resolution image can be perfectly reconstructed from an inversion at lower resolution.
+After an upscaled image has been cleaned, a model is created at the same (high) resolution.
+Because the model consists of delta functions, the spatial frequencies in the model are not band limited.
+Therefore, downscaling the model will remove information from the model.
+However, only the low spatial frequencies of this image will be used in the prediction.
+As long as downscaling does not modify the low-frequency components, the output of the prediction step will not change.
+This requires low-pass filtering the model before decimation.
+
+We have implemented an option in WSClean to automatically decrease the inversion resolution to the Nyquist limit,
+\begin{equation}\label{eq:nyquist-resolution}
+ N_\textrm{Nyquist} = 2 \frac{N_\textrm{pix} S_\textrm{pix}}{S_\textrm{synth beam}},
+\end{equation}
+where $N_\textrm{Nyquist}$ the resolution in pixels used during inversion, $N_\textrm{pix}$ the requested number of pixels of the image along one side, $S_\textrm{pix}$ the requested angular pixel scale and $S_\textrm{synth beam}$ the minimum angular size of the synthesized beam.
+Before cleaning, the low-resolution image is interpolated with a procedure consisting of i) a low-resolution FFT; ii) zero padding; and iii) a high-resolution inverse FFT.
+Such interpolation assumes the image to be periodic, but this is already assumed during the inversion process.
+After cleaning, the model is decimated by the inverse procedure, consisting of i) a high-resolution FFT; ii) truncation; and iii) a low-resolution inverse FFT.
+With common settings, this procedure decreases the inversion imaging resolution with a factor of 2.5.
+Without further correction, such a decrease would lower the positional accuracy with which visibilities are gridded onto the $uv$-plane.
+This can be corrected by increasing the oversampling rate, which has almost no affect on performance.
+
+Recreating Fig.~\ref{fig:vela-projection-example} using both the snapshot method and this optimization lowers the computational cost from 41~min to 24~min.
+Of the 24~min, 77\% is spent on cleaning approximately 100,000 components.
+The outputs with and without rescaling do not visibly differ, but have an RMS difference of 20 mJy.
+We think this can be explained by the non-linear behaviour of clean.
+
 \section{Analysis} \label{sec:analysis}
@@
 This yields 195 $w$-planes/layers at 10\degree zenith angle.
-Some configurations with high $w$-values fail to image with CASA, presumably because the $w$-kernels become to large.
+Several configurations with large $w$-values fail to image with CASA, presumably because the $w$-kernels become to large.
 
@@
 The reprojection technique needs on average fewer $w$-layers to reach the same level of accuracy.
+We have performed tests with and without the optimization of \S\ref{sec:decreasing-inversion-resolution}.
+They yield identical numbers.
 
@@
 \end{subfigure}%
-\caption{Imaging performance as a function of several parameters. Error bars show 5$\sigma$ level. Unfinished lines indicate the imager could not run succesfully.}\label{fig:timings}
+\caption{Imaging performance as a function of several parameters.
+Error bars show 5$\sigma$ level.
+Unfinished lines indicate the imager could not run the specific configuration succesfully.}\label{fig:timings}
 \end{figure*}
@@
 In each benchmark, the wall-clock time is measured that is required to produce the synthesized point-spread function and the image itself.
-No cleaning or prediction was performed.
+No cleaning or prediction was performed, and the optimization of \S\ref{sec:decreasing-inversion-resolution} is not used.
 The results are given in Fig.~\ref{fig:timings}.
@@
 In Panel~\ref{fig:timing-resolution}, the number of pixels in the image is changed without changing the field of view.
-The performance of $w$-stacking is more affected by the size of the image, but is still significantly faster in making 12.8K images at 10\degree ZA than $w$-projection.
+The performance of $w$-stacking is more affected by the size of the image, but is still significantly faster in making 12.8K images at 0\degree ZA than $w$-projection.
+Imaging the primary MWA beam requires approximately a resolution of 3k for cleaning.
+If the small-inversion optimization of \S\ref{sec:decreasing-inversion-resolution} is used, this decreases to about 1.5k.
+At ZA=10\degree, this saves about a factor of 2 in computing cost.
+The benefit of the optimization increases with resolution.
+At 10k resolution, cost is decreased by an order of magnitude.
 
@@
  \hline
- GLEAM    &   67h 7m & 8h 31m &  8h 05m & 15h 18m & 1.2 TFLOPS & $6.9 \times 10^2$    \\
- EMU      &    5.2 d &  2.9 d &   2.9 d & 5.2 d   & 10 TFLOPS & $2.1 \times 10^4$      \\
- MSSS low &  2h 18m  & 0h 59m &  0h 58m & 2h 16m  & 140 GFLOPS & $3.5 \times 10^3$   \\
- MSSS high&  7h 16m  & 3h 52m &  3h 52m & 7h 16m  & 530 GFLOPS & $3.4 \times 10^3$   \\
- LOFAR NL &  50.6 d  &   10 d &   7.2 d & 15.9 d & 33 TFLOPS & $4.5 \times 10^3$ \\
- AARTFAAC &  71h 25m &29h 50m & 29h 50m & 51h 34m & 4.1 TFLOPS & $6.8 \times 10^2$   \\
- VLSS     &  0h 12m  & 0h 12m &  0h 02m &  0h 12m & 4.1 GFLOPS & $12.1 \times 10^3$     \\
+ GLEAM    &  65h 25m & 8h 04m &  8h 03m & 14h 10m   & 1.1 TFLOPS & $6.9 \times 10^2$    \\
+ EMU      &    5.2 d &  2.9 d &   2.9 d & 4.4 d     & 9.7 TFLOPS & $2.1 \times 10^4$      \\
+ MSSS low &  2h 16m  & 0h 57m &  0h 57m & 2h 10m    & 130 GFLOPS & $3.5 \times 10^3$   \\
+ MSSS high&  7h 16m  & 3h 52m &  3h 52m & 5h 44m    & 530 GFLOPS & $3.4 \times 10^3$   \\
+ LOFAR NL &  50.2 d  &  7.3 d &   7.2 d & 12.9 d    &  24 TFLOPS & $4.5 \times 10^3$ \\
+ AARTFAAC &  2.5 d   &  1.2 d &   1.2 d &   2.5 d   & 4.1 TFLOPS & $6.8 \times 10^2$   \\
+ VLSS     &  0h 09m  & 0h 01m &  0h 01m &  0h 08m   & 2.2 GFLOPS & $12.1 \times 10^3$     \\
  MeerKAT  &   10.8 d &  6.2 d &   6.2 d &  10.8 d   &  21 TFLOPS & $6.8 \times 10^2$ \\
 % SKA1 AA inner&392 d &  228 d &   228 d &   392 d & & $6.8 \times 10^2$   \\
- SKA1 AA core&1581 d &  911 d &   911 d &  1571 d &  3 PFLOPS & $6.8 \times 10^2$ \\
- SKA1 AA full& 633 yr& 48.8 yr & 48.8 yr& 84.2 yr & 59 PFLOPS & $6.8 \times 10^2$ \\
+ SKA1 AA core&1581 d &  911 d &   911 d &  1570.8   & 3.0 PFLOPS & $6.8 \times 10^2$ \\
+ SKA1 AA full& 643 yr& 48.8 yr & 48.8 yr& 84.1 yr   &  59 PFLOPS & $6.8 \times 10^2$ \\
 \end{tabular}
@@
 This could again be the difference in optimisation choices between CASA and WSClean, but it does show that the effort of implementing a hybrid over pure stacking might not be worthwhile.
+All configurations except VLSS and GLEAM have optimal values for $\Delta w$ much smaller than one, suggesting all $w$-corrections should be done by stacking instead of projection.
 
@@
 In multi-beam configurations such as the SKA and EMU configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams.
-The image size is set to 5 times the FOV width divided by the resolution.
-The factor of five is to allow better cleaning accuracy and source-position fitting, as is common.
+The image size is set to 2 times the FOV width divided by the resolution, as described by Eq.~\ref{eq:nyquist-resolution}.
 The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).
@@
 WSClean is in most situations predicted to be 2--3 times faster than CASA.
-The MWA is somewhat of an exception, where WSClean is 8$\times$ faster.
-The main cause for this is the wide field of view of the MWA.
+WSClean has the largest benefit on full LOFAR, MWA and the full SKA, where WSClean is respectively 7, 8$\times$ and 12 times faster.
 The $w$-snapshot method does not improve much over normal WSClean operation in any of the cases, except the VLSS case.
 We think this is because the VLSS had a modest data rates compared to the other configurations, making the constant cost per $w$-plane/layer stand out.
-The $w$-snapshot might become more valuable at higher zenith angles (TODO do I want to predict that too?) The hybrid method is in all situations appromixately a factor of two slower than WSClean.
+The $w$-snapshot might become more valuable at higher zenith angles.
+The hybrid method is in all situations appromixately a factor of two slower than WSClean.
 
-Imaging the full field of view with the full SKA becomes very expensive due to the resolution of $18\textrm{k} \times 18$k pixels. This translates to a computing power of $\sim60$ PetaFLOPS.
+Imaging the full field of view with the full SKA becomes very expensive due to the high frequency and time resolution, and image resolution of $7.2\textrm{k} \times 7.2$k pixels.
+This translates to a computing power requirement of $\sim60$ PetaFLOPS.
 
@@
 A hybrid between $w$-stacking and $w$-projection seems not to improve performance over a pure $w$-stacking implementation, but does improve a pure $w$-projection implementation significantly.
+Optimizing image resolution during inversion and prediction increases performance by a factor of 2--10 and has no noticable effect on the accuracy.
 
@@
 For example, a performance gain of a few factors can be achieved by averaging shorter baselines to their maximum integration time and bandwidth resolution before imaging.
-Moreover, although longer baselines make the imaging more expensive, it is likely that it is not often required to image the full field of view when using the longest baselines.
+Moreover, although longer baselines make the imaging more expensive, it is likely that it is not always required to image the full field of view when using the longest baselines.
 
@@
 This is not possible for fields with diffuse emmission or faint point sources when no model is known a-priory.
+
+WSClean is currently not able to perform multi-scale clean.
+Results of applying CASA's multi-scale clean on MWA data show that especially the Galactic plane is significantly better deconvolved using multi-scale clean, but it is very expensive.
+Imaging a one-minute observation takes 30 hours of computational time without $w$-projection.
+We plan to implement some form of multi-scale cleaning in WSClean.
+It is likely that additional optimizations need to be made to be able to do this with acceptable performance.
 

### 2014-02-27 articles/2014/Article-WSClean/img/fit.cpp

@@ -13,6 +13,13 @@
 
 bool fixDW = false, fixConstant = false, fixZAOffset = false;
 
+const double majorIterationsInversionCount = 10.0; // total nr of inversions + predictions for configuration evaluations
+
+// number of times to sample higher than resolution
+// For cleaning, one would normally use ~5.0, but with the WSClean resampling optimization it is possible to
+// sample at Nyquist resolution (i.e., 0.5 beam size) and upscale the result.
+const double overSamplingResolution = 2.0; 
+
 double sinZA(double za, double lambda = 1.64, double maxBaselineM = 2900.0, double maxHeightDiffM = 5.5)
 {
 	return (maxBaselineM*sin(za*(M_PI/180.0)) + maxHeightDiffM*cos(za*(M_PI/180.0))) / lambda;
@@ -746,15 +753,14 @@ void evalSurveyConfig(const std::string& desc, double lambda, double fwhm, size_
 	const double duration = 60.0 * 60.0;
 	const double za = 20.0;
 	double nMVis = bandWidth / freqRes * (antennas * (antennas-1))/2 * (duration/intTimeSec) * 1e-6;
-	const double nKPix = 5.0 * fov/(angRes*1000.0);
+	const double nKPix = overSamplingResolution * fov/(angRes*1000.0);
 	const double nChan = 1;
 	if(func == WSSCleanExtrapolatedFunc)
 	{
 		nMVis *= 300.0 / duration;
 	}
 	const double we = sinZA(za, lambda, maxBaseline, maxDiffHeight) * fovFact(fov);
-	const double majorIterationCount = 10;
-	double time = majorIterationCount * double(beams) * eval(func, za, nMVis, nKPix, fov, nChan, a, b, c, d, e, f, lambda, maxBaseline, maxDiffHeight);
+	double time = majorIterationsInversionCount * double(beams) * eval(func, za, nMVis, nKPix, fov, nChan, a, b, c, d, e, f, lambda, maxBaseline, maxDiffHeight);
 	if(func == WSSCleanExtrapolatedFunc)
 	{
 		time *= duration / 300.0;
@@ -767,9 +773,8 @@ void evalSurveyConfig(const std::string& desc, double lambda, double fwhm, size_
 	{
 		double dt, dw, minDWTime;
 		std::string filename = func == WSCleanFunc ? "dt-wsclean.txt" : "dt-casa.txt";
[…]

### 2014-02-27 articles/2014/Article-WSClean/wsclean.tex
Adding full-sky image

@@
 
-TODO: add big full-sky image (PKS J2358-6054 is visible)
+\begin{figure*}
+\begin{center}
+\includegraphics[width=13.5cm]{img/fullsky-together}
+\caption{A beam-corrected MWA image of a 112~s zenith observation that covers almost the full sky.
+The lower image shows a zoom-in on the southern side lobe.
+PKS J2358-6054 ($\sim100$ Jy) is visible in the centre of the southern side lobe and resolved.
+The noise level in the side lobe is about 200 mJy.}
+\label{fig:full-sky-example}
+\end{center}
+\end{figure*}
 

### 2014-02-27 articles/2014/Article-WSClean/wsclean.tex
Add reference to new image

@@
 \subsection{Full-sky imaging}
-For low-frequency telescopes, it can be of interest to do full-sky (i.e., horizon to horizon) imaging.
-In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90\degree from the pointing centre.
-Imaging these sidelobes might be relevant because they are scientifically of interest (e.g.
-when searching for transients).
-Full-sky imaging can also be useful for self-calibration or deconvolution.
-For example, self-calibration using full-sky clean components has been found to give good results in imaging the resolved FR-II radio source Fornax~A (McKinley et al., in prep.).
-
-To efficiently image the full sky, an observation is split into short snapshots, their phase centre is changed to zenith and subsequently a suitable area is imaged.
-Depending on desired resolution and dynamic range, this might require very large images.
-To make an image at the MWA resolution, the image needs to have approximately 10k pixels in width and height.
-The $w$-values will be very small at zenith, because at zenith only the vertical offsets of the antennas will contribute to the $w$-value.
-For the MWA, the maximum height difference between tiles is $8.5$m.
-The MWA tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the maximum $w$-value decreases to $5.5\textrm{m} / \lambda$.
-In the following sections, when discussing the MWA zenith direction we are referring to this optimal $w$-direction.
 
@@
 \end{figure*}
+
+For low-frequency telescopes, it can be of interest to do full-sky (i.e., horizon to horizon) imaging.
+In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90\degree from the pointing centre.
+Imaging these sidelobes might be relevant because they are scientifically of interest (e.g.
+when searching for transients).
+Full-sky imaging can also be useful for self-calibration or deconvolution.
+For example, self-calibration using full-sky clean components has been found to give good results in imaging the resolved FR-II radio source Fornax~A (McKinley et al., in prep.).
+An example full-sky MWA image is shown in Fig.~\ref{fig:full-sky-example}.
+
+To efficiently image the full sky, an observation is split into short snapshots, their phase centre is changed to zenith and subsequently a suitable area is imaged.
+Depending on desired resolution and dynamic range, this might require very large images.
+To make an image at the MWA resolution, the image needs to have approximately 10k pixels in width and height.
+The $w$-values will be very small at zenith, because at zenith only the vertical offsets of the antennas will contribute to the $w$-value.
+For the MWA, the maximum height difference between tiles is $8.5$m.
+The tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the maximum $w$-value decreases to $5.5\textrm{m} / \lambda$.
+In the following sections, when discussing the MWA zenith direction we are referring to this optimal $w$-direction.
 

### 2014-02-28 articles/2014/Article-WSClean/wsclean.tex
Replace noise measurement with better one, final touches to article before submit to collaboration

@@
 
-\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\hspace{2cm}E-mail: \url{andre.offringa@anu.edu.au}},
+\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{E-mail: \url{andre.offringa@anu.edu.au}},
 B.~McKinley\ANU$^,$\CAASTRO,
@@
 A.~E.~E.~Rogers\Haystack, \newauthor
-D.~A.~Roshi\NRAO \footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
+D.~A.~Roshi\NRAO, %\footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
 N.~Udaya~Shankar\RRI,
@@
 We predict the computing costs for several other arrays and estimate that our imager is a factor of 2--12 faster, depending on the array configuration.
-Imaging SKA observations is challenging, as we estimate the computing cost of full SKA1 low imaging to be 60 PetaFLOPS with current techniques.
+We estimate the computing cost for imaging full SKA1-low observations to be 60 PetaFLOPS with current techniques.
 We find that combining $w$-stacking with the $w$-snapshot algorithm does not significantly improve computing requirements.
@@
 
-There are several methods to deal with the $w$-term during imaging: facetting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, $w$-projection \citep{wprojection-cornwell}, $w$-stacking \citep{ska-memo-regridding-2011} and $w$-snapshots \citep{widefield-imaging-ska-cornwell}.
+There are several methods to deal with the $w$-term during imaging: faceting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, $w$-projection \citep{wprojection-cornwell}, $w$-stacking \citep{ska-memo-regridding-2011} and $w$-snapshots \citep{widefield-imaging-ska-cornwell}.
 
@@
 It can use GPUs to improve efficiency, and uses a $w$-projection kernel during imaging to correct for the $w$-term.
-The kernel can be kept small, because the data is processed in snapshots that have been phase-centered to zenith.
+The kernel can be kept small, because the data is processed in snapshots.
 The RTS can not perform deconvolution as of yet.
@@
 
-In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's $w$-projection implementation, and is able to succesfully image any MWA observation.
-To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with $w$-snapshot imaging.
+In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's $w$-projection implementation, and is able to successfully image any MWA observation.
+To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with a new technique for $w$-snapshot imaging.
 We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''.
-Our new imaging implementation, which we believe to be the fastest generic widefield imager available, is publicly released\footnote{TODO add URL}.
+Our new imaging implementation, which we believe to be the fastest generic widefield imager available, is publicly released\footnote{The WSClean source code can be found at:\\\url{http://sourceforge.net/p/wsclean}}.
 
@@
 We will not discuss calibration, but assume $V$ has been calibrated before imaging.
-In the case of a polarized measurement, the symbols become $2\times 2$ matrices, but without loss of generality we will ignore polarization and treat inversion as a scalar problem.
+In the case of a polarized measurement, the symbols become $2\times 2$ matrices and beam correction is more complicated, but without loss of generality we will ignore polarization and treat inversion as a scalar problem.
 Imaging consists of inverting Eq.~\eqref{eq:visibility-function}, i.e., to find $I'$ from $V$.
@@
 The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation.
-This shows that the sky function can be reconstructed by i)~grid samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~apply the direction-dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeat this for all $w$-values and add the results together; v)~apply the final scaling.
+This shows that the sky function can be reconstructed by i)~gridding samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~apply the direction-dependent phase shift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeat this for all $w$-values and add the results together; v)~apply the final scaling.
 
@@
 
-The inverse of imaging, i.e., to calculate the visibility from a model image, can be done by reversing the $w$-stacking algorithm: i)~inverse scale the image; ii)~copy the image to several layers; iii)~inverse apply the direction-dependent phaseshift for each layer; iv)~FFT each layer; and v)~sample a required visibility from the correct $w$-layer.
+The inverse of imaging, i.e., to calculate the visibility from a model image, can be done by reversing the $w$-stacking algorithm: i)~inverse scale the image; ii)~copy the image to several layers; iii)~inverse apply the direction-dependent phase shift for each layer; iv)~FFT each layer; and v)~sample a required visibility from the correct $w$-layer.
 We will refer to this operation as prediction.
@@
 It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation.
-We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each dimension, $N_\textrm{vis}$ is the number of visibilities, $N_\textrm{kern}$ is the size of the antialiasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.
+We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each side, $N_\textrm{vis}$ is the number of visibilities, $N_\textrm{kern}$ is the size of the anti-aliasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.
 
@@
 With the purpose of testing new algorithms for the imaging of MWA data, we have written a new imager around the $w$-stacking algorithm.
-As will be shown in Sect.~\ref{sec:application-to-nonmwa}, the imager is not specialized for the MWA, and has been succesfully used for imaging data from other telescopes.
+The imager is not specialized for the MWA, and has been successfully used for imaging data from other telescopes.
 
@@
 This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, less relevant.
-However, the PSF varies for imaging with non-zero $w$-values, and subtracting a constant PSF in image space leads to inaccuracies.
+For very large images this optimization might still be useful, but we have not implemented it in WSClean.
+Because the PSF varies for imaging with non-zero $w$-values, subtracting a constant PSF in image space leads to inaccuracies.
 After a number of minor iterations, it is therefore beneficial to invert the model back to the visibilities via prediction, and subtract the model directly from the visibilities \citep{wprojection-cornwell}.
@@
 PKS J2358-6054 ($\sim100$ Jy) is visible in the centre of the southern side lobe and resolved.
-The noise level in the side lobe is about 200 mJy.}
+The noise level in the side lobe is about 200 mJy.
+PKS J2358-6054 has been cleaned, but some artefacts remain because no direction-dependent calibration has been performed.}
 \label{fig:full-sky-example}
@@
 Depending on desired resolution and dynamic range, this might require very large images.
-To make an image at the MWA resolution, the image needs to have approximately 10k pixels in width and height.
+To make an image at the MWA resolution, the image needs to have approximately 10k pixels along each side.
 The $w$-values will be very small at zenith, because at zenith only the vertical offsets of the antennas will contribute to the $w$-value.
-For the MWA, the maximum height difference between tiles is $8.5$m.
+For the 128-tile MWA, the maximum differential elevation between tiles is $8.5$m.
 The tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the maximum $w$-value decreases to $5.5\textrm{m} / \lambda$.
@@
 
-An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13~min MWA observation imaged with WSClean.
+An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13-min MWA observation imaged with WSClean.
 The processing steps performed for this image were i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the AOFlagger \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibrate on Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) image each of the seven 112 s intervals separately.
-Cleaning an image that is in a different projection yields slightly different results, but qualitativily the two images are clearly of equal accuracy.
+Cleaning an image that is in a different projection yields slightly different results, but qualitatively the two images are clearly of equal accuracy.
 The wall-clock time for imaging is 60 min and 41 min for respectively normal projection and the recentred image.
@@
 Since sources outside the field of view will be aliased back into the field, the image can be transformed to have the alias in the centre.
-This complicates gridding during imaging, because the antialiasing kernel needs to have a pass-band shape instead of a low-pass shape.
+This complicates gridding during imaging, because the anti-aliasing kernel needs to have a pass-band shape instead of a low-pass shape.
 Therefore, we chose to implement the former method.
@@
 \end{figure*}
-Because the MWA consists of fixed, beamformed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix.
+Because the MWA consists of fixed, beam-formed dipole antennas, the voltage beam of a MWA tile is described by a complex, non-diagonal Jones matrix.
 Alternatively, Mueller matrices can be used as well, but in general these result in more computations \citep{revisiting-me-i}.
@@
 Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such.
-However, this is incorrect because they are not necessarily orthogonal, i.e., ($2\times$) Stokes I$\neq pp + qq$.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters.
+However, this is somewhat confusing because they are not necessarily orthogonal.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters.
 The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry.
@@
 This improves the accuracy of the clean algorithm, because the positions of image maxima will be closer to the actual source positions.
-This factor of $\sim$5 is normally taken into account in the overal imaging resolution, i.e., the image size is increased during inversion.
+This factor of $\sim$5 is normally taken into account in the overall imaging resolution, i.e., the image size is increased during inversion.
 For the $w$-projection method, the cost of inversion with an increased resolution is small, because it does not increase the size of the $w$-kernels.
-It will affect the inverse FFT, but the relative cost of this step is negligable in the total cost.
+It will affect the inverse FFT, but the relative cost of this step is negligible in the total cost.
 The $w$-stacking method is however significant affected by the image resolution, because it performs many inverse FFTs.
@@
 As long as the $uv$-plane is sampled with at least the Nyquist frequency, a high resolution image can be perfectly reconstructed from an inversion at lower resolution.
-After an upscaled image has been cleaned, a model is created at the same (high) resolution.
+After an up-scaled image has been cleaned, a model is created at the same (high) resolution.
 Because the model consists of delta functions, the spatial frequencies in the model are not band limited.
@@
 The extra noise can be observed in the produced images, as shown in Fig.~\ref{fig:residuals}.
-An image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}, is also made and analyzed.
+An image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}, is also made and analysed.
 Its accuracy is higher compared to the normal projection, but its residual noise is also higher.
@@
 Error bars show 5$\sigma$ level.
-Unfinished lines indicate the imager could not run the specific configuration succesfully.}\label{fig:timings}
+Unfinished lines indicate the imager could not run the specific configuration successfully.}\label{fig:timings}
 \end{figure*}
@@
 where $D$ is the maximum baseline length, $\Delta z_{\max}$ is the maximum height difference between antennas and $\lambda$ is the wavelength, all in meters, and $\xi$ is an extra parameter for fitting the CASA measurements, $\xi_\textrm{CASA}=28.4$ and $\xi_\textrm{WSClean}=0$.
-These functions follow the trend of the measurements well, with an absolute error of 14.7\% and 20.7\% for respectively the WSClean and CASA functions.
+These functions follow the trend of the measurements well, with an absolute error of 14.8\% and 20.7\% for respectively the WSClean and CASA functions.
 
@@
 We have excluded the cost of phase shifting the visibilities and gridding.
-The cost for regridding with simple nearest neighbour or bilinair interpolation is indeed negligable, although more accurate interpolation (e.g.
+The cost for regridding with simple nearest neighbour or bilinair interpolation is indeed negligible, although more accurate interpolation (e.g.
 Lanczos interpolation) can be expensive.
-Also, our current implementation of the phase-changing program does take a non-negligable time, but this implementation is not optimized and can in theory be implemented in a preprocessing pipeline or on the fly during imaging.
+Also, our current implementation of the phase-changing program does take a non-negligible time, but this implementation is not optimized and can in theory be implemented in a preprocessing pipeline or on the fly during imaging.
 
@@
 Therefore, from a performance perspective, the snapshot duration should be as large as possible.
-In practise, the beam needs to be corrected on small timescales.
+In practise, the beam needs to be corrected on small time scales.
 A snapshot duration of more than a few minutes is therefore not possible.
@@
 We use the derived functions to estimate the computational cost of the algorithms for configurations of several telescopes.
-We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multifrequency Snapshot Sky Survey (MSSS)\footnote{See \url{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam—ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \url{http://www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The Meer Karoo Array Telescope (MeerKAT); and the low-frequency phase~1 aperture arrays of the Square-Kilometre Array, with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}.
+We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multi-frequency Snapshot Sky Survey (MSSS)\footnote{See \url{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam—ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \url{http://www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The Meer Karoo Array Telescope (MeerKAT); and the low-frequency phase~1 aperture arrays of the Square-Kilometre Array, with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}.
 The configurations are summarized in Table~\ref{tbl:computational-cost-configurations}.
@@
 The $w$-snapshot might become more valuable at higher zenith angles.
-The hybrid method is in all situations appromixately a factor of two slower than WSClean.
+The hybrid method is in all situations approximately a factor of two slower than WSClean.
 
@@
 This can lead to performance improvements of a factor of $3$ for the MWA, but only in the most expensive imaging configurations that are less commonly used.
-Considering the extra regridding step required, which complicates issues such as calculating the integrated beam shape, recentering snapshots is worthwhile for the MWA only at very low elevations or large fields of view.
+Considering the extra regridding step required, which complicates issues such as calculating the integrated beam shape, recentering snapshots is worthwhile for the MWA only at very low elevations or with large fields of view.
 Extrapolation of the computing cost predicts that this holds for most arrays, including SKA low.
-A hybrid between $w$-stacking and $w$-projection seems not to improve performance over a pure $w$-stacking implementation, but does improve a pure $w$-projection implementation significantly.
-Optimizing image resolution during inversion and prediction increases performance by a factor of 2--10 and has no noticable effect on the accuracy.
+A hybrid between $w$-stacking and $w$-projection does not improve performance over a pure $w$-stacking implementation, but does improve a pure $w$-projection implementation significantly.
+Our optimization of lowering the image resolution during inversion and prediction increases performance by a factor of 2--10 and has no noticeable effect on the accuracy.
 
@@
 Deconvolution is not an issue for achieving low to intermediate dynamic ranges with the MWA.
-Because of sufficient instanteneous $uv$-coverage and near-confusion snapshot noise level, snapshots can be cleaned individually to deep levels.
+Because of sufficient instantaneous $uv$-coverage and near-confusion snapshot noise level, snapshots can be cleaned individually to deep levels.
 For example, cleaning to $5\sigma$ level results in a cleaning threshold of approximately 100~mJy in 112~s observations.
@@
 
-So far, we have corrected the (full-polarization) beam in image space, which is possible because of the homogenuity of the MWA tiles.
-This is cheap for our current snapshot time of 112s, but this might be unfeasible when it is required to calculate the beam on very short time scales.
+So far, we have corrected the (full-polarization) beam in image space, which is possible because of the homogeneity of the MWA tiles.
+This is cheap for our current snapshot time of 112s, but this might be infeasible when it is required to calculate the beam on very short time scales.
 Accurate MWA beam models are still being developed.
@@
 with the SAGECAL \citep{sage-calibration-ii} or RTS peeling \citep{rts-mwa} calibration techniques.
-This is not possible for fields with diffuse emmission or faint point sources when no model is known a-priory.
+This is not possible for fields with diffuse emission or faint point sources when no model is known a-priory.
 
@@
 National Science Foundation (grants AST-0457585, PHY-0835713, CAREER-0847753, and AST-0908884), the Australian Research Council (LIEF grants LE0775621 and LE0882938), the U.S.
-Air Force Office of Scientic Research (grant FA9550-0510247), and the Centre for All-sky Astrophysics (an Australian Research Council Centre of Excellence funded by grant CE110001020).
+Air Force Office of Scientific Research (grant FA9550-0510247), and the Centre for All-sky Astrophysics (an Australian Research Council Centre of Excellence funded by grant CE110001020).
 Support is also provided by the Smithsonian Astrophysical Observatory, the MIT School of Science, the Raman Research Institute, the Australian National University, and the Victoria University of Wellington (via grant MED-E1799 from the New Zealand Ministry of Economic Development and an IBM Shared University Research Grant).
@@
 \includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
-\caption{Effect of antialiasing-kernel size on performance.}
+\caption{Effect of anti-aliasing kernel size on performance.}
 \label{fig:timing-kernelsize}

### 2014-02-28 articles/2014/Article-WSClean/wsclean.tex
Version as submitted to MWA collaborative review

@@
 \def\UWisc{$^{4}$}
-\def\MIT{$^{5}$}
-\def\SKASA{$^{6}$}
-\def\ASU{$^{7}$}
-\def\Haystack{$^{8}$}
-\def\RRI{$^{9}$}
-\def\USydney{$^{10}$}
+\def\USydney{$^{5}$}
+\def\MIT{$^{6}$}
+\def\SKASA{$^{7}$}
+\def\ASU{$^{8}$}
+\def\Haystack{$^{9}$}
+\def\RRI{$^{10}$}
 \def\CfA{$^{11}$}
@@
 J.~Rhee\ANU,
+T.~Murphy\USydney$^,$\CAASTRO,
 G.~Bernardi\SKASA,
@@
 \UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
+\USydney{}Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
 \MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
@@
 \RRI{}Raman Research Institute, Bangalore 560080, India\\
-\USydney{}Sydney Institute for Astronomy, School of Physics, The University of Sydney, NSW 2006, Australia\\
 \CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
@@
 To reach high dynamic ranges, it can be necessary to deal with direction-dependent effects (DDEs).
-This is especially true for widefield telescopes.
+This is especially true for wide-field telescopes.
 One way to handle DDEs is by using the a-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
@@
 We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''.
-Our new imaging implementation, which we believe to be the fastest generic widefield imager available, is publicly released\footnote{The WSClean source code can be found at:\\\url{http://sourceforge.net/p/wsclean}}.
+Our new imaging implementation, which we believe to be the fastest generic wide-field imager available, is publicly released\footnote{The WSClean source code can be found at:\\\url{http://sourceforge.net/p/wsclean}}.
 
@@
 The method is demonstrated in Fig.~\ref{fig:pulsar-stokes-iv}, which displays a detection of pulsar J0437-4715 in Stokes V.
-The image was made with an initial pipeline for the radio-sky monitor project that uses the MWA to search for transients and exoplanets.
+The image was made with an initial pipeline for the MWA radio-sky monitor project (PI: Bell) that uses the MWA to search for transient and variable sources.
 The pipeline includes WSClean to make the Stokes-parameter images.
@@
 It was suggested that $w$-snapshots would improve the imaging speed for the SKA situation \citep{widefield-imaging-ska-cornwell}, but our results show this is not the case when already using the $w$-stacking algorithm.
-Clearly it remains challenging to perform widefield imaging with acceptable performance, but some optimisations could be made for the SKA.
+Clearly it remains challenging to perform wide-field imaging with acceptable performance, but some optimisations could be made for the SKA.
 For example, a performance gain of a few factors can be achieved by averaging shorter baselines to their maximum integration time and bandwidth resolution before imaging.

### 2014-02-28 articles/2014/Article-WSClean/img/fit.cpp
Updating formula

@@ -832,7 +832,8 @@ void fitAllWSClean()
 	fit(info, a, b, c, d, e, f);
 	fixConstant = false;
 	fit(info, a, b, c, d, e, f);
-	std::cout << "Nchan (" << a << " Npix^2 log Npix fovFact sin ZA + " << b << ") + " << c << " Nvis) + " << d << "\n";
+	// nChan * (a * (we+b)*nPix*nPix*log2(nPix) + d) + c * nVis;
+	std::cout << "Nchan (" << a << " Npix^2 log Npix (w_max + " << b << ") + " << d << ") + " << c << " Nvis\n";
 	writeNVisFit("benchmark-nsamples/fit-zenith-wsc.txt", info.func, a, b, c, d, e, f, 0.0, 3.072, 36.864, 1);
 	writeNVisFit("benchmark-nsamples/fit-ZA010-wsc.txt", info.func, a, b, c, d, e, f, 10.0, 3.072, 36.864, 1);
 	writeNPixFit("benchmark-resolution/fit-zenith-wsclean.txt", info.func, a, b, c, d, e, f, 0.0, 349.6, 36.864, 1);
@@ -911,7 +912,8 @@ void fitAllCASA()
 	fit(info, a, b, c, d, e, f);
 	fixDW = false;
 	fit(info, a, b, c, d, e, f);
-	std::cout << "t = Nchan (" << a << " Npix^2 log Npix + " << b << " Nvis ((sin ZA + " << e << ") * fov)^2 + " << b*d << " Nvis + " << f << '\n';
+	// nChan * a * nPix*nPix*log2(nPix) + b * nVis * (we * we + d) + f;
+	std::cout << "t = Nchan " << a << " Npix^2 log Npix + " << b << " Nvis ((sin ZA + " << e << ") * fov)^2 + " << d << ") + " << f << '\n';
 	writeNVisFit("benchmark-nsamples/fit-zenith-casa.txt", info.func, a, b, c, d, e, f, 0.0, 3.072, 36.864, 1);
 	writeNVisFit("benchmark-nsamples/fit-ZA010-casa.txt", info.func, a, b, c, d, e, f, 10.0, 3.072, 36.864, 1);
 	writeNPixFit("benchmark-resolution/fit-zenith-casa.txt", info.func, a, b, c, d, e, f, 0.0, 349.6, 36.864, 1);

### 2014-02-28 articles/2014/Article-WSClean/wsclean.tex
Slight updates to formulae

@@
 & t_\textrm{WSClean} (w_{\max},N_\textrm{Mvis},N_\textrm{freq},N_\textrm{kpix}) = \\ \notag%
-& N_\textrm{freq} \left( 0.452 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} (w_{\max} + 0.611) + 244 N_\textrm{Mvis} \right)
+& N_\textrm{freq} \left( 0.526 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} (w_{\max} + 0.715) + 0.535 \right) + 0.248 N_\textrm{Mvis}
 \end{align}
@@
 & t_\textrm{CASA}(w_{\max},N_\textrm{vis},N_\textrm{freq},N_\textrm{pix}) = \\ \notag%
-& N_\textrm{freq} \left(1.03 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} + 5.93 N_\textrm{Mvis} (w_{\max} + 1.71)^2\right) + 172.
+& N_\textrm{freq} \left(0.965 N_\textrm{kpix}^2 \log_2 N_\textrm{kpix} + 0.0106 N_\textrm{Mvis} (w_{\max}^2 + 40.1)\right) + 40.8.
 \label{eq:casa-computing-cost}%
@@
 with the SAGECAL \citep{sage-calibration-ii} or RTS peeling \citep{rts-mwa} calibration techniques.
-This is not possible for fields with diffuse emission or faint point sources when no model is known a-priory.
+This is not possible for fields with diffuse emission or faint point sources when no model is known \textit{a priori}.
 

### 2014-02-28 articles/2014/Article-WSClean/img/fit.cpp

@@ -833,7 +833,7 @@ void fitAllWSClean()
 	fixConstant = false;
 	fit(info, a, b, c, d, e, f);
 	// nChan * (a * (we+b)*nPix*nPix*log2(nPix) + d) + c * nVis;
-	std::cout << "Nchan (" << a << " Npix^2 log Npix (w_max + " << b << ") + " << d << ") + " << c << " Nvis\n";
+	std::cout << "Nchan (" << a << " Nkpix^2 log Nkpix (w_max + " << b << ") + " << d << ") + " << c << " NMvis\n";
 	writeNVisFit("benchmark-nsamples/fit-zenith-wsc.txt", info.func, a, b, c, d, e, f, 0.0, 3.072, 36.864, 1);
 	writeNVisFit("benchmark-nsamples/fit-ZA010-wsc.txt", info.func, a, b, c, d, e, f, 10.0, 3.072, 36.864, 1);
 	writeNPixFit("benchmark-resolution/fit-zenith-wsclean.txt", info.func, a, b, c, d, e, f, 0.0, 349.6, 36.864, 1);
@@ -913,7 +913,7 @@ void fitAllCASA()
 	fixDW = false;
 	fit(info, a, b, c, d, e, f);
 	// nChan * a * nPix*nPix*log2(nPix) + b * nVis * (we * we + d) + f;
-	std::cout << "t = Nchan " << a << " Npix^2 log Npix + " << b << " Nvis ((sin ZA + " << e << ") * fov)^2 + " << d << ") + " << f << '\n';
+	std::cout << "t = Nchan " << a << " Nkpix^2 log Nkpix + " << b << " NMvis ((sin ZA + " << e << ") * fov)^2 + " << d << ") + " << f << '\n';
 	writeNVisFit("benchmark-nsamples/fit-zenith-casa.txt", info.func, a, b, c, d, e, f, 0.0, 3.072, 36.864, 1);
 	writeNVisFit("benchmark-nsamples/fit-ZA010-casa.txt", info.func, a, b, c, d, e, f, 10.0, 3.072, 36.864, 1);
 	writeNPixFit("benchmark-resolution/fit-zenith-casa.txt", info.func, a, b, c, d, e, f, 0.0, 349.6, 36.864, 1);

### 2014-02-28 articles/2014/Article-WSClean/wsclean.tex
Implemented Natasha's comments

@@
 
-\ifpdf\pdfinfo{/Title      (WSClean)
+\ifpdf\pdfinfo{/Title      (WSClean: an implementation of a fast, generic wide-field imager)
                /Author     (A. R. Offringa et al.)
@@
 We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm.
-The performance dependencies of CASA's $w$-projection and our new imager are analysed and analytical functions are derived that describe the required computing cost for both imagers.
-On MWA data, we find our new method to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with full polarization correction.
+The performance dependencies of \textsc{CASA}'s $w$-projection and our new imager are analysed and analytical functions are derived that describe the required computing cost for both imagers.
+On data from the Murchison Widefield Array, we find our new method to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with correct polarization correction.
 We predict the computing costs for several other arrays and estimate that our imager is a factor of 2--12 faster, depending on the array configuration.
-We estimate the computing cost for imaging full SKA1-low observations to be 60 PetaFLOPS with current techniques.
+We estimate the computing cost for imaging the low-frequency Square-Kilometre Array observations to be 60 PetaFLOPS with current techniques.
 We find that combining $w$-stacking with the $w$-snapshot algorithm does not significantly improve computing requirements.
@@
 \section{Introduction}
-
-Observations from non-coplanar interferometric radio telescopes that image large fractions of the sky at once can not be imaged with a two-dimensional fast Fourier transform (FFT).
+Visibility data from non-coplanar interferometric radio telescopes which observe large fractions of the sky at once can not be accurately imaged with a two-dimensional fast Fourier transform (FFT).
 Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}.
-This deviation is amplified for telescopes with wide field of view, making this especially an issue for low-frequency telescopes that by nature are wide-field instruments.
+This deviation is amplified for telescopes with wide field of view (FOV), making this a particularly important issue for low-frequency telescopes that by nature are wide-field instruments.
 
@@
 Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa-design-2009}, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}).
-The Common Astronomy Software Applications (CASA, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution.
-However, at the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with CASA's $w$-projection algorithm.
+The Common Astronomy Software Applications (\textsc{CASA}, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution.
+However, with the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with \textsc{CASA}'s $w$-projection algorithm.
 Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.
@@
 
-In this article, we present a new implementation of a generic wide-field imager that is significantly faster than CASA's $w$-projection implementation, and is able to successfully image any MWA observation.
+In this article, we present a new implementation of a generic wide-field imager that is significantly faster than \textsc{CASA}'s $w$-projection implementation, and is able to successfully image any MWA observation.
 To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with a new technique for $w$-snapshot imaging.
-We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''.
-Our new imaging implementation, which we believe to be the fastest generic wide-field imager available, is publicly released\footnote{The WSClean source code can be found at:\\\url{http://sourceforge.net/p/wsclean}}.
+We named the new imager ``\textsc{WSClean}'', as an abbreviation for ``$w$-Stacking Clean''.
+Our new imaging implementation, which we believe to be the fastest generic wide-field imager available, is publicly released\footnote{The \textsc{WSClean} source code can be found at:\\\href{http://sourceforge.net/p/wsclean}{http://sourceforge.net/p/wsclean}}.
 
@@
 
-For small field of views, the term $\sqrt{1-l^2-m^2}$ is approximately of unit size, making Eq.~\eqref{eq:visibility-function} approximate an ordinary 2-dimensional Fourier transform that can be inverted with an inverse FFT.
+For small FOVs, the term $\sqrt{1-l^2-m^2}$ is approximately of unit size, making Eq.~\eqref{eq:visibility-function} approximately an ordinary invertable two-dimensional Fourier transform.
 A common rule is that this is valid when
@@
 The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation.
-This shows that the sky function can be reconstructed by i)~gridding samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~apply the direction-dependent phase shift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeat this for all $w$-values and add the results together; v)~apply the final scaling.
+This shows that the sky function can be reconstructed by i)~gridding samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~applying the direction-dependent phase shift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeating this for all $w$-values and adding the results together; v)~applying the final scaling.
 
@@
 In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds.
-The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-values samples.
+The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-value samples.
 In this case we can therefore calculate the image for $w<0$ from the image with $w>0$.
-That allows us to set $w_{\min}$ to the minimum absolute $w$-value, which requires two times fewer layers.
+That allows us to set $w_{\min}$ to the minimum absolute $w$-value, which requires a factor of two fewer layers.
 In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence the inverse FFT is always performed as a complex-to-complex transform.
@@
 \caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field.
-WSClean was set to use 12 $w$-layers.
+\textsc{WSClean} was set to use 12 $w$-layers.
 The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers.
@@
 \begin{equation} \label{eq:nwlayers-bound}
- N_\textrm{wlayers} \gg 2\pi \left(w_{\max} - w_{\min}\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
+ N_\textrm{wlay} \gg 2\pi \left(w_{\max} - w_{\min}\right) \max_{l,m} \left(1 - \sqrt{1-l^2-m^2}\right).
 \end{equation}
-Actual values for the right-hand side can differ a lot.
+Actual values for the right-hand side can be very different, depending on the observation.
 The value of $w_{\max} - w_{\min}$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image.
-For the MWA, a typical value of $w_{\max} - w_{\min}$ is $\sim 10$ at zenith, but is already $\sim 400$ at a zenith angle of $30^{\circ}$.
+For the MWA, a typical value of $w_{\max} - w_{\min}$ is $\sim 10$ at zenith, and reaches $\sim 400$ at a zenith angle of $30^{\circ}$.
 For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$.
@@
 It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation.
-We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each side, $N_\textrm{vis}$ is the number of visibilities, $N_\textrm{kern}$ is the size of the anti-aliasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging field of view.
+We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each side, $N_\textrm{vis}$ is the number of visibilities, $N_\textrm{kern}$ is the size of the anti-aliasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging FOV.
 
@@
 From these bounds it can be concluded that in the limiting behaviour, the $w$-stacking method will be faster when the gridding of the visibilities is the dominating cost of the algorithm.
-The $w$-projection algorithm will be faster when the inverse FFTs are dominating.
+The $w$-projection algorithm will be faster when the inverse FFTs are the dominant expense.
 In Sect.~\ref{sec:analysis} we will determine which method is faster in practice for different parameters.
 
-\section{The WSClean imager implementation} \label{sec:implementation}
+\section{The \textsc{WSClean} imager implementation} \label{sec:implementation}
 With the purpose of testing new algorithms for the imaging of MWA data, we have written a new imager around the $w$-stacking algorithm.
@@
 
-The new imager, called ``WSClean'', is written in the C++ language.
-It reads visibilities from CASA measurement sets and writes output images to Flexible Image Transport System (FITS) files.
+The new imager, called ``\textsc{WSClean}'', is written in the C++ language.
+It reads visibilities from \textsc{CASA} measurement sets and writes output images to Flexible Image Transport System (FITS) files.
 Several steps are multi-threaded using the threading module of the C++11 standard library.
@@
 This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, less relevant.
-For very large images this optimization might still be useful, but we have not implemented it in WSClean.
+For very large images this optimization might still be useful, but we have not implemented it in \textsc{WSClean}.
 Because the PSF varies for imaging with non-zero $w$-values, subtracting a constant PSF in image space leads to inaccuracies.
@@
 This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}.
-WSClean allows cleaning individual polarizations, or can jointly deconvolve the polarizations.
+\textsc{WSClean} allows cleaning individual polarizations, or can jointly deconvolve the polarizations.
 In the latter case, peaks are searched in $pp^2+2pq \overline{(pq)} + qq^2$ space, where $\overline{pq}=qp$ is the complex conjugate of $pq$, and the PSF is subtracted from the polarizations with different factors.
@@
 It will not be possible to store all $w$-layers in memory when creating large images or when many $w$-layers are used.
-For example, our 32-GB test machine can store 227 $w$-layers of $3072\times3072$ in memory.
+For example, our test machine, with 32 GB of memory, can store 227 $w$-layers of $3072\times3072$.
 In more demanding imaging configurations, the implementation performs several passes over the measurement set and will grid a subset of $w$-layers in each pass.
@@
 
-To efficiently image the full sky, an observation is split into short snapshots, their phase centre is changed to zenith and subsequently a suitable area is imaged.
+To efficiently image the full sky, an observation is split into short snapshots, each of which is phase-rotated to zenith and subsequently imaged, with an appropriate number of pixels and resolution.
 Depending on desired resolution and dynamic range, this might require very large images.
@@
 Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging.
-Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm.
+Both Stokes-I images have been made with \textsc{WSClean} using the Cotton-Schwab clean algorithm.
 No beam correction was applied.
@@
 Our method is also relatively easy to implement.
-It has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on zenith angle, field of view and the non-coplanarity.
+It has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on zenith angle, FOV and the non-coplanarity.
 
@@
 A recentred image can be stored in a Flexible Image Transport System (FITS) file with the slant orthographic (SIN) projection normally used in interferometric imaging \citep{fits-coordinates-2002}, by setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}.
-Common viewers such as Kvis and DS9 support such FITS files and display their coordinates correctly.
+Common viewers such as \textsc{Kvis} and \textsc{DS9} support such FITS files and display their coordinates correctly.
 Unlike warped snapshots, recentred images do not need the generalized-SIN-projection keywords \texttt{PV2\_1} and \texttt{PV2\_2}.
 
-An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13-min MWA observation imaged with WSClean.
-The processing steps performed for this image were i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the AOFlagger \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibrate on Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) image each of the seven 112 s intervals separately.
+An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13-min MWA observation imaged with \textsc{WSClean}.
+The processing steps performed for this image were i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the AOFlagger \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibration using Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) imaging of the seven 112 s intervals separately.
 Cleaning an image that is in a different projection yields slightly different results, but qualitatively the two images are clearly of equal accuracy.
@@
 Another method to shift an image is by using the periodicity of the FFT function.
-Since sources outside the field of view will be aliased back into the field, the image can be transformed to have the alias in the centre.
+Since sources outside the FOV will be aliased back into the field, the image can be transformed to have the alias in the centre.
 This complicates gridding during imaging, because the anti-aliasing kernel needs to have a pass-band shape instead of a low-pass shape.
@@
 \includegraphics[width=6cm]{img/pulsar-stokesi}\hspace{1cm}\includegraphics[width=6cm]{img/pulsar-stokesv}
-\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with WSClean.
+\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with \textsc{WSClean}.
 The pulsar (centre of image) displays 17\% circular polarization.}
@@
 where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose.
-Because $J$ is complex and non-diagonal, all 4 combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such.
+Because $J$ is complex and non-diagonal, all four combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such.
 However, this is somewhat confusing because they are not necessarily orthogonal.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters.
 The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry.
-To make proper MWA beam correction possible, WSClean can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images.
+To make proper MWA beam correction possible, \textsc{WSClean} can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images.
 This requires four runs of the algorithm.
@@
 The image was made with an initial pipeline for the MWA radio-sky monitor project (PI: Bell) that uses the MWA to search for transient and variable sources.
-The pipeline includes WSClean to make the Stokes-parameter images.
+The pipeline includes \textsc{WSClean} to make the Stokes-parameter images.
 
@@
 A common kernel function is a windowed sinc function, which acts as a low-pass filter.
-This decreases the flux of sources outside the field of view, and thus helps to attenuate aliased ghost sources and sidelobes \citep{post-correlation-filtering}.
+This decreases the flux of sources outside the FOV, and thus helps to attenuate aliased ghost sources and sidelobes \citep{post-correlation-filtering}.
 By supersampling, a convolution kernel also makes it possible to place samples more accurately at their $uv$ position, thereby lowering decorrelation.
@@
 The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}.
-CASA convolves samples with a PSWF of 7 pixels total width during gridding.
+\textsc{CASA} convolves samples with a PSWF of seven pixels total width during gridding.
 When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate.
-WSClean currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation of the PSWF \citep{fourier-kernel-selection-1991}.
-After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that 7 pixels is a good compromise between performance and accuracy, and this is the default for WSClean.
+\textsc{WSClean} currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation of the PSWF \citep{fourier-kernel-selection-1991}.
+After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that seven pixels is a good compromise between performance and accuracy, and this is the default for \textsc{WSClean}.
 
 \subsection{Decreasing inversion resolution} \label{sec:decreasing-inversion-resolution}
-Typically, for cleaning it is desired that images have a pixel size at least 5 times smaller than the synthesized-beam width.
+Typically, for cleaning it is desired that images have a pixel size at least five times smaller than the synthesized-beam width.
 This improves the accuracy of the clean algorithm, because the positions of image maxima will be closer to the actual source positions.
-This factor of $\sim$5 is normally taken into account in the overall imaging resolution, i.e., the image size is increased during inversion.
+This factor of five is normally taken into account in the overall imaging resolution, i.e., the image size is increased during inversion.
 For the $w$-projection method, the cost of inversion with an increased resolution is small, because it does not increase the size of the $w$-kernels.
 It will affect the inverse FFT, but the relative cost of this step is negligible in the total cost.
-The $w$-stacking method is however significant affected by the image resolution, because it performs many inverse FFTs.
+The $w$-stacking method is however significantly affected by the image resolution, because it performs many inverse FFTs.
 
-The spatial frequencies in the output image are band limited by the synthesized beam.
+The spatial frequencies in the output image are band-limited by the synthesized beam.
 Therefore, it is not required to perform the inversion at higher resolution.
@@
 After an up-scaled image has been cleaned, a model is created at the same (high) resolution.
-Because the model consists of delta functions, the spatial frequencies in the model are not band limited.
-Therefore, downscaling the model will remove information from the model.
+Because the model consists of delta functions, the spatial frequencies in the model are not band-limited.
+Consequently, downscaling the model will remove information from the model.
 However, only the low spatial frequencies of this image will be used in the prediction.
-As long as downscaling does not modify the low-frequency components, the output of the prediction step will not change.
+Therefore, inversion at higher resolution is not required: as long as long as downscaling does not modify the low-frequency components, the output of the prediction step will not change.
 This requires low-pass filtering the model before decimation.
 
-We have implemented an option in WSClean to automatically decrease the inversion resolution to the Nyquist limit,
+We have implemented an option in \textsc{WSClean} to automatically decrease the inversion resolution to the Nyquist limit,
 \begin{equation}\label{eq:nyquist-resolution}
@@
 Without further correction, such a decrease would lower the positional accuracy with which visibilities are gridded onto the $uv$-plane.
-This can be corrected by increasing the oversampling rate, which has almost no affect on performance.
+This can be corrected by increasing the oversampling rate, which has almost no effect on performance.
 
@@
 \section{Analysis} \label{sec:analysis}
-We will now analyse the performance and accuracy of our implementation, and compare it with the $w$-projection implementation in CASA.
-For the analysis, we use imaging parameters which are common for MWA Imaging.
+We will now analyse the performance and accuracy of our implementation, and compare it with the $w$-projection implementation in \textsc{CASA}.
+For the analysis, we use imaging parameters which are common for MWA imaging.
 When a specific parameter value is not mentioned, the settings from Table~\ref{tbl:default-parameters} are used.
@@
 This yields 195 $w$-planes/layers at 10\degree zenith angle.
-Several configurations with large $w$-values fail to image with CASA, presumably because the $w$-kernels become to large.
+Several configurations with large $w$-values fail to image with \textsc{CASA}, presumably because the $w$-kernels become too large.
 
-The used software version of CASA is ``stable release 42.0, revision 26465'', which was released September 2013.
-For WSClean, the latest version as of February 2014 was used.
-The tests were run on a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with 6 cores that can perform 138 giga-floating point operations per second (GFLOPS).
-The data are stored on a multi-disk array with 5 spinning hard disks, which has a combined read rate of about 450 MB/s.
+The used software version of \textsc{CASA} is ``stable release 42.0, revision 26465'', which was released September 2013.
+For \textsc{WSClean}, the latest version as of February 2014 was used.
+The tests were run on a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores that can perform 138 giga-floating point operations per second (GFLOPS).
+The data are stored on a multi-disk array with five spinning hard disks, which has a combined read rate of about 450 MB/s.
 \begin{table}%
@@
 \hline
-Array & MWA 128T \\
+Array & MWA \\
+Number of elements & 128 \\
 Image size & $3072 \times 3072$ \\
@@
 Max $w$-value for phase centre & 172 $\lambda$ (283 m) \\
-Nr. polarizations in set & 4 \\
+Number of polarizations in set & 4 \\
 Imaged polarization & $pp$ ($\sim$XX) \\
-Imaging mode & Multi-frequency synthesis \\
+Imaging mode & multi-frequency synthesis \\
 Weighting & uniform \\
@@
 \hline\hline
-& \textbf{WSClean} & \textbf{WSClean}    & \textbf{CASA} \\
+& \textbf{\textsc{WSClean}} & \textbf{\textsc{WSClean}}    & \textbf{\textsc{CASA}} \\
 &                  & \textbf{+ recentre} & \\
@@
 \hline
-\multicolumn{3}{l}{\textit{Zenith angle 10\degree}} \\
+\multicolumn{3}{l}{\textit{Zenith angle 10\degree (195 $w$-layers/planes)}} \\
 \hline
@@
 \begin{center}
-\includegraphics[height=5cm]{img/residual-casa}\hspace{1cm}\includegraphics[height=5cm]{img/residual-wsclean}
-\caption{Residual images after Cotton-Schwab cleaning of a simulated 10\degree-zenith angle MWA observation using CASA (left) and WSClean (right), using similar inversion and cleaning parameters.
+\includegraphics[height=5cm]{img/residual-casa-fixed-grayscale}\hspace{1cm}\includegraphics[height=5cm]{img/residual-wsclean}
+\caption{Residual images after Cotton-Schwab cleaning of a simulated 10\degree-zenith angle MWA observation using \textsc{CASA} (left) and \textsc{WSClean} (right), using similar inversion and cleaning parameters.
 The panels show a small part of the full images.
 The full field contains 100 simulated sources over 20\degree.
-Sources in the image produced with CASA are slightly less accurately subtracted, leading to a residual noise level of 1.07 mJy and some visible artefacts, whereas WSClean reaches 0.90 mJy.
+Sources in the image produced with \textsc{CASA} are slightly less accurately subtracted, leading to a residual noise level of 1.07 mJy and some visible artefacts, whereas \textsc{WSClean} reaches 0.90 mJy.
 Since the effect is stronger away from the phase centre, it is likely that this is caused by the finite size of the $w$-kernels used in $w$-projection, which leads to inaccuracies.}
@@
 
-To assess the accuracy of WSClean and CASA, we simulate a MWA observation with hundred sources of 1~Jy in a 20\degree diameter area, without adding system noise.
+To assess the accuracy of \textsc{WSClean} and \textsc{CASA}, we simulate a MWA observation with 100 sources of 1~Jy in a 20\degree diameter area, without adding system noise.
 A unitary primary beam is assumed.
-We image the simulated set with WSClean and CASA using Cotton-Schwab cleaning to a threshold of 10~mJy.
+We image the simulated set with \textsc{WSClean} and \textsc{CASA} using Cotton-Schwab cleaning to a threshold of 10~mJy.
 The two imagers calculate slightly different restoring (synthesized) beams, hence to avoid bias the restoring beams are fixed.
 Other imaging parameters are given in Table~\ref{tbl:default-parameters}.
-The Aegean program \citep{aegean-hancock-2012} is used to perform source detection on the produced images.
+The \textsc{Aegean} program \citep{aegean-hancock-2012} is used to perform source detection on the produced images.
 Sidelobe noise of the residual 10~mJy source structures triggers a few false detections.
@@
 
-Table~\ref{tbl:accuracy-measurements} lists the measured RMS in the residual image and the standard errors of the source brightnesses as detected by Aegean.
-In all tests, WSClean performs slightly more accurate.
-At zenith, CASA produces a residual RMS twice as high as WSClean.
+Table~\ref{tbl:accuracy-measurements} lists the measured root mean square (RMS) in the residual image and the standard errors of the source brightnesses as detected by \textsc{Aegean}.
+In all tests, \textsc{WSClean} performs slightly more accurately.
+At zenith, \textsc{CASA} produces a residual RMS twice as large as \textsc{WSClean}.
 This is caused by the fact that we keep the number of $w$-projection planes in $w$-projection equal to the number of $w$-layers in $w$-stacking, resulting in only 12 $w$-planes at zenith.
@@
 However, the computational performance of the $w$-projection algorithm is hardly affected by the number of $w$-projection planes, and in practical situations one would always use more $w$-projection planes.
-When 128 $w$-projection planes are used in CASA, its residual RMS is equal to WSClean with 12 $w$-layers, but the flux density measurements are less accurate.
-We do not know why this parameter needs to be higher in CASA to reach the same RMS.
-Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlayers}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree.
+When 128 $w$-projection planes are used in \textsc{CASA}, the residual RMS is equal to \textsc{WSClean} with 12 $w$-layers, but the flux density measurements are less accurate.
+We do not know why this parameter needs to be higher in \textsc{CASA} to reach the same RMS.
+We are using enough $w$-planes to cover the sources: Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlay}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree.
 Also unexpected is that the source flux density becomes worse by increasing the number of planes.
-For WSClean both values stay approximately the same when the number of $w$-layers is increased.
+For \textsc{WSClean} both values stay approximately the same when the number of $w$-layers is increased.
 
-In the $\textrm{ZA}=10\degree$ case, CASA is slightly less accurate.
+In the $\textrm{ZA}=10\degree$ case, \textsc{CASA} is slightly less accurate.
 The extra noise can be observed in the produced images, as shown in Fig.~\ref{fig:residuals}.
@@
 
-WSClean is faster in all tested cases.
-Both imagers have performed 5 major iterations for these results, and the inversions and predictions are dominating the computing time.
+\textsc{WSClean} is faster in all tested cases.
+Both imagers perform five major iterations for these results, and the inversions and predictions are dominating the computing time.
 We will look more closely at the difference in performance in the next section.
@@
 
-We measured the performance of the imagers using several MWA data sets.
-Each specific configuration is run 5 times and standard deviations are calculated.
-The variations between runs is typically a few seconds.
+We measure the performance of the imagers using several MWA data sets.
+Each specific configuration is run five times and standard deviations are calculated.
+Variation between runs is typically a few seconds.
 In each benchmark, the wall-clock time is measured that is required to produce the synthesized point-spread function and the image itself.
-No cleaning or prediction was performed, and the optimization of \S\ref{sec:decreasing-inversion-resolution} is not used.
+No cleaning or prediction is performed, and the optimization of \S\ref{sec:decreasing-inversion-resolution} is not used.
 The results are given in Fig.~\ref{fig:timings}.
@@
 In that situation, the $w$-stacking implementation is 7.9 times faster at $\textrm{ZA}=10\degree$ and 2.6 times faster at $\textrm{ZA}=0\degree$.
-With small data volumes, the FFTs start to dominate the cost.
-This is visible from Fig.~\ref{fig:timing-nsamples} by the flattening towards the left.
-At that point, WSClean is in both cases approximately 3 times faster.
+With small data volumes, the FFTs start to dominate the cost, visible in Fig.~\ref{fig:timing-nsamples} as a flattening towards the left.
+At that point, \textsc{WSClean} is in both cases approximately three times faster.
 At zenith, the two methods are expected to perform almost identical.
@@
 Additionally, at higher zenith angles, the $w$-kernels become too large to be able to make images of 3072 pixels or larger.
-Using the recentering technique with WSClean to decrease the $w$-values makes the computational cost approximately constant.
+Using the recentering technique with \textsc{WSClean} to decrease the $w$-values makes the computational cost approximately constant.
 However, the additional time required to rephase and regrid the measurement set makes it only worthwhile for images $\ge3072$ pixels and $\textrm{ZA}\ge15\degree$.
 This is dependent on the speed of rotating the phase centre and regridding.
-Our program to change the phase centre is currently not multi-threaded, so its performance might be improvable.
+Our program to change the phase centre is currently not multi-threaded, so its performance should be improvable.
 Furthermore, the relative cost of changing the phase centre is lower when performing multiple major iterations.
 
-In Panel~\ref{fig:timing-resolution}, the number of pixels in the image is changed without changing the field of view.
+In Panel~\ref{fig:timing-resolution}, the number of pixels in the image is changed without changing the FOV.
 The performance of $w$-stacking is more affected by the size of the image, but is still significantly faster in making 12.8K images at 0\degree ZA than $w$-projection.
-Imaging the primary MWA beam requires approximately a resolution of 3k for cleaning.
+Imaging the MWA primary beam requires approximately an image size of 3k pixels for cleaning.
 If the small-inversion optimization of \S\ref{sec:decreasing-inversion-resolution} is used, this decreases to about 1.5k.
@@
 The benefit of the optimization increases with resolution.
-At 10k resolution, cost is decreased by an order of magnitude.
+At an image size of 10k pixels, cost is decreased by an order of magnitude.
 
@@
 
-As can be seen in Panel~\ref{fig:timing-fov}, the field of view (FOV) has a large affect on performance when imaging off-zenith.
+As can be seen in Panel~\ref{fig:timing-fov}, the FOV has a large affect on performance when imaging off-zenith.
 This plot was created by varying the size of a pixel in the output image, such that the number of pixels in the image did not change.
 The FOV is calculated as angle between the left- and right-most pixel in the image.
-CASA is not able to make off-zenith images larger than $\sim$30\degree, and its imaging cost increases much more rapidly compared to the cost of WSClean, which implies that gridding is the major cost in this scenario.
+\textsc{CASA} is not able to make off-zenith images larger than $\sim$30\degree, and its imaging cost increases much more rapidly compared to the cost of \textsc{WSClean}, which implies that gridding is the major cost in this scenario.
 The image recentering technique is clearly beneficial for FOVs larger than $\sim$30\degree, and can even make a difference of an order of magnitude at very large FOVs of $\sim$90\degree.
@@
 %It is important to point out that all these measurement were performed by imaging a single polarization.
-Making full-polarization images with WSClean currently requires four runs.
-Imaging four Stokes parameters with CASA costs slightly less than four times the time of imaging a single polarization.
+Making full-polarization images with \textsc{WSClean} currently requires four runs.
+Imaging four Stokes parameters with \textsc{CASA} costs slightly less than four times the time of imaging a single polarization.
 With the settings of Table~\ref{tbl:default-parameters}, the difference is a factor of $3.5$, and imaging XX and YY together costs $1.9$ times imaging just XX.
@@
 \subsection{Derivation of computing cost formulae}
-Based on the expected cost terms described in Sec.~\ref{sec:time-complexity}, we derive analytical functions for the CASA and WSClean measurements using least-squares fitting.
+Based on the expected cost terms described in Sec.~\ref{sec:time-complexity}, we derive analytical functions for the \textsc{CASA} and \textsc{WSClean} measurements using least-squares fitting.
 Several functions with different free parameters were tested, and formulae with minimum amount of parameters are selected that still follow the trend of the measurements and have reasonably small errors.
 Measurements are weighted with the inverse standard deviation instead of the variance, because we found that the latter results in too much weight on fast configurations, leading to functions that represent the general trend less well.
-The single outlying measurement for $N_\textrm{pix}=1792$ with CASA in Fig.~\ref{fig:timing-resolution} is removed.
+The single outlying measurement for $N_\textrm{pix}=1792$ with \textsc{CASA} in Fig.~\ref{fig:timing-resolution} is removed.
 
-For the WSClean time-cost function $t_\textrm{WSClean}$ we find 
+For the \textsc{WSClean} time-cost function $t_\textrm{\textsc{WSClean}}$ we find 
 \begin{align}\label{eq:wsclean-computing-cost}
@@
 \end{align}
-and for CASA we find
+and for \textsc{CASA} we find
 \begin{align}\label{eq:casa-computing-cost}
@@
 \end{equation}%
-where $D$ is the maximum baseline length, $\Delta z_{\max}$ is the maximum height difference between antennas and $\lambda$ is the wavelength, all in meters, and $\xi$ is an extra parameter for fitting the CASA measurements, $\xi_\textrm{CASA}=28.4$ and $\xi_\textrm{WSClean}=0$.
-These functions follow the trend of the measurements well, with an absolute error of 14.8\% and 20.7\% for respectively the WSClean and CASA functions.
+where $D$ is the maximum baseline length, $\Delta z_{\max}$ is the maximum height difference between antennas and $\lambda$ is the wavelength, all in meters, and $\xi$ is an extra parameter for fitting the \textsc{CASA} measurements, $\xi_\textrm{CASA}=28.4$ and $\xi_\textrm{WSClean}=0$.
+These functions follow the trend of the measurements well, with an absolute error of 14.8\% and 20.7\% for respectively the \textsc{WSClean} and \textsc{CASA} functions.
 
@@
 Lanczos interpolation) can be expensive.
-Also, our current implementation of the phase-changing program does take a non-negligible time, but this implementation is not optimized and can in theory be implemented in a preprocessing pipeline or on the fly during imaging.
+Also, our current implementation of the phase-changing program does take a non-negligible time, but this implementation is not optimized and can in theory be implemented in a preprocessing pipeline or on-the-fly during imaging.
 
-Function $t_\textrm{snapshot}$ can be minimized to find the optimal snapshot duration.
-For WSClean with MWA parameters, we find that this function decreases but no minimum is reached for $\Delta \tau_\textrm{snapshot}<24$~h.
+The function $t_\textrm{snapshot}$ can be minimized to find the optimal snapshot duration.
+For \textsc{WSClean} with MWA parameters, we find that this function decreases but no minimum is reached for $\Delta \tau_\textrm{snapshot}<24$~h.
 Therefore, from a performance perspective, the snapshot duration should be as large as possible.
@@
 A snapshot duration of more than a few minutes is therefore not possible.
-For $w$-projection in CASA we find an optimal snapshot duration of $\sim2$ min for MWA observations.
+For $w$-projection in \textsc{CASA} we find an optimal snapshot duration of $\sim2$ min for MWA observations.
 
 \begin{table*}
-\caption{Configurations for which the computational cost is predicted.} \label{tbl:computational-cost-configurations}
+\caption{Configurations for which the computational cost is predicted.
+Columns: $\lambda$=wavelength; FOV=field of view; Beams=number of beams; Ant=number of elements; Res=angular resolution; $D$=maximum baseline; $\max \Delta z$=maximum differential elevation; $\Delta t$=correlator dump time; BW=bandwidth; and $\Delta \nu$=correlator frequency resolution.} \label{tbl:computational-cost-configurations}
 \begin{tabular}{l|rrrrrrrrrr}
- Configuration&$\lambda$& FOV  & Beams & Ant & Res & $D$ & $\max \Delta z$ & int. t & BW & $\Delta \nu$ \\
+ Configuration&$\lambda$& FOV  & Beams & Ant & Res & $D$ & $\max \Delta z$ & $\Delta t$ & BW & $\Delta \nu$ \\
           & (m)       & (FWHM)&       &     &     & (km)                       & (m)             & (s)    & (MHz) &  (kHz) \\
@@
 \begin{table*}
-\caption{Predicted computational costs for configurations listed in Table~\ref{tbl:computational-cost-configurations}, based on multi-frequency synthesis with 5 major iterations observing for 1 hour at ZA=20\degree with 1 polarization.
-Predictions are for $w$-projection with CASA; $w$-stacking with WSClean; $w$-stacked snapshots with WSClean using optimal snapshot duration; and a hybrid between $w$-projection and $w$-stacking using CASA with optimal $\Delta w$.
+\caption{Predicted computational costs for configurations listed in Table~\ref{tbl:computational-cost-configurations}, based on multi-frequency synthesis with five major iterations observing for 1~h at ZA=20\degree with 1 polarization.
+Predictions are for $w$-projection with \textsc{CASA}; $w$-stacking with \textsc{WSClean}; $w$-stacked snapshots with \textsc{WSClean} using optimal snapshot duration; and a hybrid between $w$-projection and $w$-stacking using \textsc{CASA} with optimal $\Delta w$.
 All have approximately equal accuracy.} \label{tbl:computational-cost-predictions}
@@
 This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory.
-The AWImager has been applied like this, which let to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}.
+The AWImager has been applied in this way, which led to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}.
 For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more generally the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_{\max}$ value and smaller $N_\textrm{vis}$.
@@
 However, the performance of the hybrid method is still about a factor of two lower than our pure-stacking implementation.
-This could again be the difference in optimisation choices between CASA and WSClean, but it does show that the effort of implementing a hybrid over pure stacking might not be worthwhile.
-All configurations except VLSS and GLEAM have optimal values for $\Delta w$ much smaller than one, suggesting all $w$-corrections should be done by stacking instead of projection.
+This could again be the difference in optimisation choices between \textsc{CASA} and \textsc{WSClean}, but it does show that the effort of implementing a hybrid over pure stacking might not be worthwhile.
 
@@
 In $w$-stacking, the speed gain associated with this optimisation is less significant.
-Limiting the $w$-values also lowers the snapshots resolution in one direction significantly, because the long baselines in one direction are no longer gridded, which is often not desirable for the MWA.
+Limiting the $w$-values also lowers the snapshot resolution in one direction significantly, because the long baselines in one direction are no longer gridded, which is often not desirable for the MWA.
 
@@
 We use the derived functions to estimate the computational cost of the algorithms for configurations of several telescopes.
-We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multi-frequency Snapshot Sky Survey (MSSS)\footnote{See \url{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam—ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \url{http://www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The Meer Karoo Array Telescope (MeerKAT); and the low-frequency phase~1 aperture arrays of the Square-Kilometre Array, with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}.
+We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multi-frequency Snapshot Sky Survey (MSSS)\footnote{See~\href{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}{www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam--ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \href{http://www.aartfaac.org}{www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The Meer Karoo Array Telescope (MeerKAT); and the low-frequency Phase~1 aperture arrays of the Square-Kilometre Array (SKA), with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}.
 The configurations are summarized in Table~\ref{tbl:computational-cost-configurations}.
 In multi-beam configurations such as the SKA and EMU configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams.
-The image size is set to 2 times the FOV width divided by the resolution, as described by Eq.~\ref{eq:nyquist-resolution}.
+The image size is set to two times the FOV width divided by the resolution, as described by Eq.~\ref{eq:nyquist-resolution}.
 The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).
@@
 The results are summarized in Table~\ref{tbl:computational-cost-predictions}.
-WSClean is in most situations predicted to be 2--3 times faster than CASA.
-WSClean has the largest benefit on full LOFAR, MWA and the full SKA, where WSClean is respectively 7, 8$\times$ and 12 times faster.
-The $w$-snapshot method does not improve much over normal WSClean operation in any of the cases, except the VLSS case.
+\textsc{WSClean} is in most situations predicted to be 2--3 times faster than \textsc{CASA}.
+\textsc{WSClean} has the largest benefit on full LOFAR, MWA and the full SKA, where \textsc{WSClean} is respectively 7, 8$\times$ and 12 times faster.
+The $w$-snapshot method does not improve much over normal \textsc{WSClean} operation in any of the cases, except the VLSS case.
 We think this is because the VLSS had a modest data rates compared to the other configurations, making the constant cost per $w$-plane/layer stand out.
 The $w$-snapshot might become more valuable at higher zenith angles.
-The hybrid method is in all situations approximately a factor of two slower than WSClean.
+The hybrid method is in all situations approximately a factor of two slower than \textsc{WSClean}.
+All configurations except VLSS and GLEAM have optimal values for $\Delta w$ much smaller than one, suggesting the $w$-corrections should be done entirely by $w$-stacking instead of the $w$-stacking/projection hybrid method.
 
-Imaging the full field of view with the full SKA becomes very expensive due to the high frequency and time resolution, and image resolution of $7.2\textrm{k} \times 7.2$k pixels.
+Imaging the full FOV with the full SKA becomes very expensive due to the high frequency and time resolution, and image resolution of $7.2\textrm{k} \times 7.2$k pixels.
 This translates to a computing power requirement of $\sim60$ PetaFLOPS.
@@
 \section{Conclusions} \label{sec:conclusions}
-We have shown that the $w$-stacking algorithm is well suited for imaging MWA observations.
-The WSClean $w$-stacking implementation is faster than CASA's $w$-projection algorithm in all common MWA imaging configurations, leading up to an order of magnitude difference already at a relatively small zenith angle of $10$\degree.
+We have shown that the $w$-stacking algorithm is well-suited for imaging MWA observations.
+The \textsc{WSClean} $w$-stacking implementation is faster than \textsc{CASA}'s $w$-projection algorithm in all common MWA imaging configurations, causing up to an order of magnitude increase in speed at a relatively small zenith angle of $10$\degree.
 It is also slightly more accurate.
-For approximately zenith angles $>15\degree$ or field-of-view diameters $>35\degree$, snapshot imaging becomes faster.
+For approximately zenith angles $>15\degree$ or FOVs $>35\degree$, snapshot imaging becomes faster.
 This can lead to performance improvements of a factor of $3$ for the MWA, but only in the most expensive imaging configurations that are less commonly used.
@@
 For example, a performance gain of a few factors can be achieved by averaging shorter baselines to their maximum integration time and bandwidth resolution before imaging.
-Moreover, although longer baselines make the imaging more expensive, it is likely that it is not always required to image the full field of view when using the longest baselines.
+Moreover, although longer baselines make the imaging more expensive, imaging the full FOV will likely not (always) be required when using the longest baselines.
 
@@
 
-WSClean is currently not able to perform multi-scale clean.
-Results of applying CASA's multi-scale clean on MWA data show that especially the Galactic plane is significantly better deconvolved using multi-scale clean, but it is very expensive.
+\textsc{WSClean} is currently not able to perform multi-scale clean.
+Results of applying \textsc{CASA}'s multi-scale clean on MWA data show that especially the Galactic plane is significantly better deconvolved using multi-scale clean, but it is very expensive.
 Imaging a one-minute observation takes 30 hours of computational time without $w$-projection.
-We plan to implement some form of multi-scale cleaning in WSClean.
+We plan to implement some form of multi-scale cleaning in \textsc{WSClean}.
 It is likely that additional optimizations need to be made to be able to do this with acceptable performance.
@@
 By using the $w$-stacking algorithms, some computational cost is transferred from gridding to performing FFTs.
-WSClean uses the FFTW library for calculating the FFTs \citep{fftw-2005}.
+\textsc{WSClean} uses the FFTW library for calculating the FFTs \citep{fftw-2005}.
 Further performance improvement can be made by using one of the available FFT libraries that make use of graphical processing units (GPUs).
@@
 \end{figure}
-As discussed in \S\ref{sec:gridding}, WSClean uses a Kaiser-Bessel window against aliasing.
+As discussed in \S\ref{sec:gridding}, \textsc{WSClean} uses a Kaiser-Bessel window against aliasing.
 The type of window function has no effect on the gridding performance, but the size of the kernel affects gridding performance quadratically.
@@
 Decreasing the kernel size to values below $7$ has little effect on performance, because for small kernels the visibility-reading rate is lower than the gridding rate.
-We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged field of view.
-Therefore, WSClean uses a default of 7 pixels for the gridding kernel.
+We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged FOV.
+Therefore, \textsc{WSClean} uses a default of 7 pixels for the gridding kernel.
 It can be increased when necessary.

### 2014-03-01 articles/2014/Article-WSClean/wsclean.tex
Implemented Melanie's comments

@@
 We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm.
-The performance dependencies of \textsc{CASA}'s $w$-projection and our new imager are analysed and analytical functions are derived that describe the required computing cost for both imagers.
+The performance dependencies of \textsc{casa}'s $w$-projection and our new imager are analysed and analytical functions are derived that describe the required computing cost for both imagers.
 On data from the Murchison Widefield Array, we find our new method to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with correct polarization correction.
@@
 Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa-design-2009}, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}).
-The Common Astronomy Software Applications (\textsc{CASA}, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution.
-However, with the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with \textsc{CASA}'s $w$-projection algorithm.
+The Common Astronomy Software Applications (\textsc{casa}, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution.
+However, with the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with \textsc{casa}'s $w$-projection algorithm.
 Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.
@@
 This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data.
-The \texttt{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, $w$-projection and $w$-stacking \citep{awimager-2013}.
+The \textsc{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, $w$-projection and $w$-stacking \citep{awimager-2013}.
 
@@
 
-In this article, we present a new implementation of a generic wide-field imager that is significantly faster than \textsc{CASA}'s $w$-projection implementation, and is able to successfully image any MWA observation.
+In this article, we present a new implementation of a generic wide-field imager that is significantly faster than \textsc{casa}'s $w$-projection implementation, and is able to successfully image any MWA observation.
 To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with a new technique for $w$-snapshot imaging.
-We named the new imager ``\textsc{WSClean}'', as an abbreviation for ``$w$-Stacking Clean''.
-Our new imaging implementation, which we believe to be the fastest generic wide-field imager available, is publicly released\footnote{The \textsc{WSClean} source code can be found at:\\\href{http://sourceforge.net/p/wsclean}{http://sourceforge.net/p/wsclean}}.
+We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''.
+Our new imaging implementation, which we believe to be the fastest generic wide-field imager available, is publicly released\footnote{The \textsc{wsclean} source code can be found at:\\\href{http://sourceforge.net/p/wsclean}{http://sourceforge.net/p/wsclean}}.
 
@@
 \caption{Aliasing artefacts caused by insufficient $w$-layers in a simulated field.
-\textsc{WSClean} was set to use 12 $w$-layers.
+\textsc{wsclean} was set to use 12 $w$-layers.
 The centre of the image is at 10\degree zenith angle, which would normally require $\sim195$ $w$-layers.
@@
 Actual values for the right-hand side can be very different, depending on the observation.
-The value of $w_{\max} - w_{\min}$ is influenced by the coplanarity of the array, the zenith angle and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image.
-For the MWA, a typical value of $w_{\max} - w_{\min}$ is $\sim 10$ at zenith, and reaches $\sim 400$ at a zenith angle of $30^{\circ}$.
+The value of $w_{\max} - w_{\min}$ is influenced by the coplanarity of the array, the zenith angle (ZA) and the wavelength, while the value of the $\max_{l,m}$ term is influenced by the angular size of the image.
+For the MWA, a typical value of $w_{\max} - w_{\min}$ is $\sim 10$ at zenith, and reaches $\sim 400$ at a ZA of $30^{\circ}$.
 For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$.
@@
 
-\section{The \textsc{WSClean} imager implementation} \label{sec:implementation}
+\section{The \textsc{wsclean} imager implementation} \label{sec:implementation}
 With the purpose of testing new algorithms for the imaging of MWA data, we have written a new imager around the $w$-stacking algorithm.
-The imager is not specialized for the MWA, and has been successfully used for imaging data from other telescopes.
+The imager is not specialized for the MWA, and has been successfully used for imaging VLA and GMRT data.
 
-The new imager, called ``\textsc{WSClean}'', is written in the C++ language.
-It reads visibilities from \textsc{CASA} measurement sets and writes output images to Flexible Image Transport System (FITS) files.
+The new imager, called ``\textsc{wsclean}'', is written in the C++ language.
+It reads visibilities from \textsc{casa} measurement sets and writes output images to Flexible Image Transport System (FITS) files.
 Several steps are multi-threaded using the threading module of the C++11 standard library.
@@
 Because of these optimizations, minor Cleaning iterations with an image size of $3072\times3072$ are performed at a rate of hundreds per second.
-This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, less relevant.
-For very large images this optimization might still be useful, but we have not implemented it in \textsc{WSClean}.
+This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced point-spread function (PSF), less relevant.
+For very large images this optimization might still be useful, but we have not implemented it in \textsc{wsclean}.
 Because the PSF varies for imaging with non-zero $w$-values, subtracting a constant PSF in image space leads to inaccuracies.
@@
 This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}.
-\textsc{WSClean} allows cleaning individual polarizations, or can jointly deconvolve the polarizations.
+\textsc{wsclean} allows cleaning individual polarizations, or can jointly deconvolve the polarizations.
 In the latter case, peaks are searched in $pp^2+2pq \overline{(pq)} + qq^2$ space, where $\overline{pq}=qp$ is the complex conjugate of $pq$, and the PSF is subtracted from the polarizations with different factors.
@@
 Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging.
-Both Stokes-I images have been made with \textsc{WSClean} using the Cotton-Schwab clean algorithm.
+Both Stokes-I images have been made with \textsc{wsclean} using the Cotton-Schwab clean algorithm.
 No beam correction was applied.
@@
 Our method is also relatively easy to implement.
-It has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on zenith angle, FOV and the non-coplanarity.
+It has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on ZA, FOV and the non-coplanarity.
 
@@
 A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together.
-A recentred image can be stored in a Flexible Image Transport System (FITS) file with the slant orthographic (SIN) projection normally used in interferometric imaging \citep{fits-coordinates-2002}, by setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}.
-Common viewers such as \textsc{Kvis} and \textsc{DS9} support such FITS files and display their coordinates correctly.
+A recentred image can be stored in a FITS file with the slant orthographic (SIN) projection normally used in interferometric imaging \citep{fits-coordinates-2002}, by setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}.
+Common viewers such as \textsc{kvis} \citep{karma-1996} and \textsc{ds9} support such FITS files and display their coordinates correctly.
 Unlike warped snapshots, recentred images do not need the generalized-SIN-projection keywords \texttt{PV2\_1} and \texttt{PV2\_2}.
 
-An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13-min MWA observation imaged with \textsc{WSClean}.
-The processing steps performed for this image were i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the AOFlagger \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibration using Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) imaging of the seven 112 s intervals separately.
+An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13-min MWA observation imaged with \textsc{wsclean}.
+The processing steps performed for this image were i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibration using Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) imaging of the seven 112 s intervals separately.
 Cleaning an image that is in a different projection yields slightly different results, but qualitatively the two images are clearly of equal accuracy.
@@
 \includegraphics[width=6cm]{img/pulsar-stokesi}\hspace{1cm}\includegraphics[width=6cm]{img/pulsar-stokesv}
-\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with \textsc{WSClean}.
-The pulsar (centre of image) displays 17\% circular polarization.}
+\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with \textsc{wsclean}.
+The pulsar (centre of image) displays 17\% circular polarization.
+Due to inaccuracies in the current beam model, other (unpolarized) sources show $\sim$1\% leakage into Stokes~V.}
 \label{fig:pulsar-stokes-iv}
@@
 The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry.
-To make proper MWA beam correction possible, \textsc{WSClean} can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images.
+To make proper MWA beam correction possible, \textsc{wsclean} can output both real $pp$, $pq$, $qp$, $qq$ images and imaginary $pq$ and $qp$ images.
 This requires four runs of the algorithm.
@@
 The image was made with an initial pipeline for the MWA radio-sky monitor project (PI: Bell) that uses the MWA to search for transient and variable sources.
-The pipeline includes \textsc{WSClean} to make the Stokes-parameter images.
+The pipeline includes \textsc{wsclean} to make the Stokes-parameter images.
 
@@
 To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm excessively unless an a-projection kernel is used.
-The AWImager uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.
+The \textsc{awimager} uses an intermediate method, and corrects a common dipole factor in image space and the phased-array beam factor in $uv$-space, which allows the gridding kernel to be smaller compared to correcting both in $uv$-space \citep{awimager-2013}.
 
@@
 The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}.
-\textsc{CASA} convolves samples with a PSWF of seven pixels total width during gridding.
+\textsc{casa} convolves samples with a PSWF of seven pixels total width during gridding.
 When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate.
-\textsc{WSClean} currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation of the PSWF \citep{fourier-kernel-selection-1991}.
-After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that seven pixels is a good compromise between performance and accuracy, and this is the default for \textsc{WSClean}.
+\textsc{wsclean} currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation of the PSWF \citep{fourier-kernel-selection-1991}.
+After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that seven pixels is a good compromise between performance and accuracy, and this is the default for \textsc{wsclean}.
 
@@
 
-We have implemented an option in \textsc{WSClean} to automatically decrease the inversion resolution to the Nyquist limit,
+We have implemented an option in \textsc{wsclean} to automatically decrease the inversion resolution to the Nyquist limit,
 \begin{equation}\label{eq:nyquist-resolution}
@@
 \section{Analysis} \label{sec:analysis}
-We will now analyse the performance and accuracy of our implementation, and compare it with the $w$-projection implementation in \textsc{CASA}.
+We will now analyse the performance and accuracy of our implementation, and compare it with the $w$-projection implementation in \textsc{casa}.
 For the analysis, we use imaging parameters which are common for MWA imaging.
@@
 The number of $w$-projection planes in $w$-projection is kept equal to the number of $w$-layers in $w$-stacking, and is set to the right hand side of Eq.~\eqref{eq:nwlayers-bound}.
-This yields 195 $w$-planes/layers at 10\degree zenith angle.
-Several configurations with large $w$-values fail to image with \textsc{CASA}, presumably because the $w$-kernels become too large.
+This yields 195 $w$-planes/layers at 10\degree ZA.
+Several configurations with large $w$-values fail to image with \textsc{casa}, presumably because the $w$-kernels become too large.
 
-The used software version of \textsc{CASA} is ``stable release 42.0, revision 26465'', which was released September 2013.
-For \textsc{WSClean}, the latest version as of February 2014 was used.
+The used software version of \textsc{casa} is ``stable release 42.0, revision 26465'', which was released September 2013.
+For \textsc{wsclean}, the latest version as of February 2014 was used.
 The tests were run on a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores that can perform 138 giga-floating point operations per second (GFLOPS).
@@
 \hline\hline
-& \textbf{\textsc{WSClean}} & \textbf{\textsc{WSClean}}    & \textbf{\textsc{CASA}} \\
+& \textbf{\textsc{wsclean}} & \textbf{\textsc{wsclean}}    & \textbf{\textsc{casa}} \\
 &                  & \textbf{+ recentre} & \\
@@
 \includegraphics[height=5cm]{img/residual-casa-fixed-grayscale}\hspace{1cm}\includegraphics[height=5cm]{img/residual-wsclean}
-\caption{Residual images after Cotton-Schwab cleaning of a simulated 10\degree-zenith angle MWA observation using \textsc{CASA} (left) and \textsc{WSClean} (right), using similar inversion and cleaning parameters.
+\caption{Residual images after Cotton-Schwab cleaning of a simulated 10\degree-zenith angle MWA observation using \textsc{casa} (left) and \textsc{wsclean} (right), using similar inversion and cleaning parameters.
 The panels show a small part of the full images.
 The full field contains 100 simulated sources over 20\degree.
-Sources in the image produced with \textsc{CASA} are slightly less accurately subtracted, leading to a residual noise level of 1.07 mJy and some visible artefacts, whereas \textsc{WSClean} reaches 0.90 mJy.
+Sources in the image produced with \textsc{casa} are slightly less accurately subtracted, leading to a residual noise level of 1.07 mJy and some visible artefacts, whereas \textsc{wsclean} reaches 0.90 mJy.
 Since the effect is stronger away from the phase centre, it is likely that this is caused by the finite size of the $w$-kernels used in $w$-projection, which leads to inaccuracies.}
@@
 
-To assess the accuracy of \textsc{WSClean} and \textsc{CASA}, we simulate a MWA observation with 100 sources of 1~Jy in a 20\degree diameter area, without adding system noise.
+To assess the accuracy of \textsc{wsclean} and \textsc{casa}, we simulate a MWA observation with 100 sources of 1~Jy in a 20\degree diameter area, without adding system noise.
 A unitary primary beam is assumed.
-We image the simulated set with \textsc{WSClean} and \textsc{CASA} using Cotton-Schwab cleaning to a threshold of 10~mJy.
+We image the simulated set with \textsc{wsclean} and \textsc{casa} using Cotton-Schwab cleaning to a threshold of 10~mJy.
 The two imagers calculate slightly different restoring (synthesized) beams, hence to avoid bias the restoring beams are fixed.
@@
 Table~\ref{tbl:accuracy-measurements} lists the measured root mean square (RMS) in the residual image and the standard errors of the source brightnesses as detected by \textsc{Aegean}.
-In all tests, \textsc{WSClean} performs slightly more accurately.
-At zenith, \textsc{CASA} produces a residual RMS twice as large as \textsc{WSClean}.
+In all tests, \textsc{wsclean} performs slightly more accurately.
+At zenith, \textsc{casa} produces a residual RMS twice as large as \textsc{wsclean}.
 This is caused by the fact that we keep the number of $w$-projection planes in $w$-projection equal to the number of $w$-layers in $w$-stacking, resulting in only 12 $w$-planes at zenith.
@@
 However, the computational performance of the $w$-projection algorithm is hardly affected by the number of $w$-projection planes, and in practical situations one would always use more $w$-projection planes.
-When 128 $w$-projection planes are used in \textsc{CASA}, the residual RMS is equal to \textsc{WSClean} with 12 $w$-layers, but the flux density measurements are less accurate.
-We do not know why this parameter needs to be higher in \textsc{CASA} to reach the same RMS.
+When 128 $w$-projection planes are used in \textsc{casa}, the residual RMS is equal to \textsc{wsclean} with 12 $w$-layers, but the flux density measurements are less accurate.
+We do not know why this parameter needs to be higher in \textsc{casa} to reach the same RMS.
 We are using enough $w$-planes to cover the sources: Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlay}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree.
 Also unexpected is that the source flux density becomes worse by increasing the number of planes.
-For \textsc{WSClean} both values stay approximately the same when the number of $w$-layers is increased.
+For \textsc{wsclean} both values stay approximately the same when the number of $w$-layers is increased.
 
-In the $\textrm{ZA}=10\degree$ case, \textsc{CASA} is slightly less accurate.
+In the $\textrm{ZA}=10\degree$ case, \textsc{casa} is slightly less accurate.
 The extra noise can be observed in the produced images, as shown in Fig.~\ref{fig:residuals}.
@@
 
-\textsc{WSClean} is faster in all tested cases.
+\textsc{wsclean} is faster in all tested cases.
 Both imagers perform five major iterations for these results, and the inversions and predictions are dominating the computing time.
@@
 Panel~\ref{fig:timing-nsamples} shows the dependency on the size of the visibility set.
-Results for imaging at zenith and 10\degree zenith angle (ZA) are shown.
+Results for imaging at zenith and 10\degree ZA are shown.
 The size of the visibility set was varied by changing the time resolution, which affects the number of visibilities to be gridded without changing the maximum $w$-value.
@@
 With small data volumes, the FFTs start to dominate the cost, visible in Fig.~\ref{fig:timing-nsamples} as a flattening towards the left.
-At that point, \textsc{WSClean} is in both cases approximately three times faster.
+At that point, \textsc{wsclean} is in both cases approximately three times faster.
 At zenith, the two methods are expected to perform almost identical.
@@
 
-In Panel~\ref{fig:timing-zenith-angle}, the computational cost as function of zenith angle is plotted.
-It shows that, as expected from \S\ref{sec:time-complexity}, the $w$-projection algorithm is more affected by the increased $w$-values, leading to differences of more than an order of magnitude at zenith angles of $\gtrapprox 20$\degree.
-Additionally, at higher zenith angles, the $w$-kernels become too large to be able to make images of 3072 pixels or larger.
-Using the recentering technique with \textsc{WSClean} to decrease the $w$-values makes the computational cost approximately constant.
+In Panel~\ref{fig:timing-zenith-angle}, the computational cost as function of ZA is plotted.
+It shows that, as expected from \S\ref{sec:time-complexity}, the $w$-projection algorithm is more affected by the increased $w$-values, leading to differences of more than an order of magnitude at ZAs of $\gtrapprox 20$\degree.
+Additionally, at higher ZAs, the $w$-kernels become too large to be able to make images of 3072 pixels or larger.
+Using the recentering technique with \textsc{wsclean} to decrease the $w$-values makes the computational cost approximately constant.
 However, the additional time required to rephase and regrid the measurement set makes it only worthwhile for images $\ge3072$ pixels and $\textrm{ZA}\ge15\degree$.
@@
 The FOV is calculated as angle between the left- and right-most pixel in the image.
-\textsc{CASA} is not able to make off-zenith images larger than $\sim$30\degree, and its imaging cost increases much more rapidly compared to the cost of \textsc{WSClean}, which implies that gridding is the major cost in this scenario.
+\textsc{casa} is not able to make off-zenith images larger than $\sim$30\degree, and its imaging cost increases much more rapidly compared to the cost of \textsc{wsclean}, which implies that gridding is the major cost in this scenario.
 The image recentering technique is clearly beneficial for FOVs larger than $\sim$30\degree, and can even make a difference of an order of magnitude at very large FOVs of $\sim$90\degree.
@@
 %It is important to point out that all these measurement were performed by imaging a single polarization.
-Making full-polarization images with \textsc{WSClean} currently requires four runs.
-Imaging four Stokes parameters with \textsc{CASA} costs slightly less than four times the time of imaging a single polarization.
+Making full-polarization images with \textsc{wsclean} currently requires four runs.
+Imaging four Stokes parameters with \textsc{casa} costs slightly less than four times the time of imaging a single polarization.
 With the settings of Table~\ref{tbl:default-parameters}, the difference is a factor of $3.5$, and imaging XX and YY together costs $1.9$ times imaging just XX.
@@
 \subsection{Derivation of computing cost formulae}
-Based on the expected cost terms described in Sec.~\ref{sec:time-complexity}, we derive analytical functions for the \textsc{CASA} and \textsc{WSClean} measurements using least-squares fitting.
+Based on the expected cost terms described in Sec.~\ref{sec:time-complexity}, we derive analytical functions for the \textsc{casa} and \textsc{wsclean} measurements using least-squares fitting.
 Several functions with different free parameters were tested, and formulae with minimum amount of parameters are selected that still follow the trend of the measurements and have reasonably small errors.
 Measurements are weighted with the inverse standard deviation instead of the variance, because we found that the latter results in too much weight on fast configurations, leading to functions that represent the general trend less well.
-The single outlying measurement for $N_\textrm{pix}=1792$ with \textsc{CASA} in Fig.~\ref{fig:timing-resolution} is removed.
+The single outlying measurement for $N_\textrm{pix}=1792$ with \textsc{casa} in Fig.~\ref{fig:timing-resolution} is removed.
 
-For the \textsc{WSClean} time-cost function $t_\textrm{\textsc{WSClean}}$ we find 
+For the \textsc{wsclean} time-cost function $t_\textrm{\textsc{wsclean}}$ we find 
 \begin{align}\label{eq:wsclean-computing-cost}
@@
 \end{align}
-and for \textsc{CASA} we find
+and for \textsc{casa} we find
 \begin{align}\label{eq:casa-computing-cost}
@@
 \end{equation}%
-where $D$ is the maximum baseline length, $\Delta z_{\max}$ is the maximum height difference between antennas and $\lambda$ is the wavelength, all in meters, and $\xi$ is an extra parameter for fitting the \textsc{CASA} measurements, $\xi_\textrm{CASA}=28.4$ and $\xi_\textrm{WSClean}=0$.
-These functions follow the trend of the measurements well, with an absolute error of 14.8\% and 20.7\% for respectively the \textsc{WSClean} and \textsc{CASA} functions.
+where $D$ is the maximum baseline length, $\Delta z_{\max}$ is the maximum height difference between antennas and $\lambda$ is the wavelength, all in meters, and $\xi$ is an extra parameter for fitting the \textsc{casa} measurements, $\xi_\textrm{CASA}=28.4$ and $\xi_\textrm{WSClean}=0$.
+These functions follow the trend of the measurements well, with an absolute error of 14.8\% and 20.7\% for respectively the \textsc{wsclean} and \textsc{casa} functions.
 
@@
 The function $t_\textrm{snapshot}$ can be minimized to find the optimal snapshot duration.
-For \textsc{WSClean} with MWA parameters, we find that this function decreases but no minimum is reached for $\Delta \tau_\textrm{snapshot}<24$~h.
+For \textsc{wsclean} with MWA parameters, we find that this function decreases but no minimum is reached for $\Delta \tau_\textrm{snapshot}<24$~h.
 Therefore, from a performance perspective, the snapshot duration should be as large as possible.
@@
 A snapshot duration of more than a few minutes is therefore not possible.
-For $w$-projection in \textsc{CASA} we find an optimal snapshot duration of $\sim2$ min for MWA observations.
+For $w$-projection in \textsc{casa} we find an optimal snapshot duration of $\sim2$ min for MWA observations.
 
@@
 \caption{Predicted computational costs for configurations listed in Table~\ref{tbl:computational-cost-configurations}, based on multi-frequency synthesis with five major iterations observing for 1~h at ZA=20\degree with 1 polarization.
-Predictions are for $w$-projection with \textsc{CASA}; $w$-stacking with \textsc{WSClean}; $w$-stacked snapshots with \textsc{WSClean} using optimal snapshot duration; and a hybrid between $w$-projection and $w$-stacking using \textsc{CASA} with optimal $\Delta w$.
+Predictions are for $w$-projection with \textsc{casa}; $w$-stacking with \textsc{wsclean}; $w$-stacked snapshots with \textsc{wsclean} using optimal snapshot duration; and a hybrid between $w$-projection and $w$-stacking using \textsc{casa} with optimal $\Delta w$.
 All have approximately equal accuracy.} \label{tbl:computational-cost-predictions}
@@
 This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory.
-The AWImager has been applied in this way, which led to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}.
+The \textsc{awimager} has been applied in this way, which led to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}.
 For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more generally the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_{\max}$ value and smaller $N_\textrm{vis}$.
@@
 However, the performance of the hybrid method is still about a factor of two lower than our pure-stacking implementation.
-This could again be the difference in optimisation choices between \textsc{CASA} and \textsc{WSClean}, but it does show that the effort of implementing a hybrid over pure stacking might not be worthwhile.
+This could again be the difference in optimisation choices between \textsc{casa} and \textsc{wsclean}, but it does show that the effort of implementing a hybrid over pure stacking might not be worthwhile.
 
@@
 In multi-beam configurations such as the SKA and EMU configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams.
+The selected wavelengths
+are approximately the central wavelength available for each instrument.
 The image size is set to two times the FOV width divided by the resolution, as described by Eq.~\ref{eq:nyquist-resolution}.
@@
 The results are summarized in Table~\ref{tbl:computational-cost-predictions}.
-\textsc{WSClean} is in most situations predicted to be 2--3 times faster than \textsc{CASA}.
-\textsc{WSClean} has the largest benefit on full LOFAR, MWA and the full SKA, where \textsc{WSClean} is respectively 7, 8$\times$ and 12 times faster.
-The $w$-snapshot method does not improve much over normal \textsc{WSClean} operation in any of the cases, except the VLSS case.
-We think this is because the VLSS had a modest data rates compared to the other configurations, making the constant cost per $w$-plane/layer stand out.
-The $w$-snapshot might become more valuable at higher zenith angles.
-The hybrid method is in all situations approximately a factor of two slower than \textsc{WSClean}.
+\textsc{wsclean} is in most situations predicted to be 2--3 times faster than \textsc{casa}.
+\textsc{wsclean} has the largest benefit on the full LOFAR, MWA and the full SKA, where \textsc{wsclean} is respectively 7, 8 and 12 times faster.
+The $w$-snapshot method does not improve much over normal \textsc{wsclean} operation in any of the cases.
+The $w$-snapshot might become more valuable at higher ZAs.
+The hybrid method is in all situations approximately a factor of two slower than \textsc{wsclean}.
 All configurations except VLSS and GLEAM have optimal values for $\Delta w$ much smaller than one, suggesting the $w$-corrections should be done entirely by $w$-stacking instead of the $w$-stacking/projection hybrid method.
@@
 We have shown that the $w$-stacking algorithm is well-suited for imaging MWA observations.
-The \textsc{WSClean} $w$-stacking implementation is faster than \textsc{CASA}'s $w$-projection algorithm in all common MWA imaging configurations, causing up to an order of magnitude increase in speed at a relatively small zenith angle of $10$\degree.
-It is also slightly more accurate.
-For approximately zenith angles $>15\degree$ or FOVs $>35\degree$, snapshot imaging becomes faster.
+The \textsc{wsclean} $w$-stacking implementation is faster than \textsc{casa}'s $w$-projection algorithm in all common MWA imaging configurations, causing up to an order of magnitude increase in speed at a relatively small ZA of $10$\degree, and is slightly more accurate.
+Roughly speaking, snapshot imaging becomes faster for ZAs $>15\degree$ or FOVs $>35\degree$.
 This can lead to performance improvements of a factor of $3$ for the MWA, but only in the most expensive imaging configurations that are less commonly used.
@@
 So far, we have corrected the (full-polarization) beam in image space, which is possible because of the homogeneity of the MWA tiles.
-This is cheap for our current snapshot time of 112s, but this might be infeasible when it is required to calculate the beam on very short time scales.
+This is computationally cheap for our current snapshot time of 112s, but this might be infeasible when it is required to calculate the beam on very short time scales.
 Accurate MWA beam models are still being developed.
@@
 
-\textsc{WSClean} is currently not able to perform multi-scale clean.
-Results of applying \textsc{CASA}'s multi-scale clean on MWA data show that especially the Galactic plane is significantly better deconvolved using multi-scale clean, but it is very expensive.
+\textsc{wsclean} is currently not able to perform multi-scale clean.
+Results of applying \textsc{casa}'s multi-scale clean on MWA data show that especially the Galactic plane is significantly better deconvolved using multi-scale clean, but it is very expensive.
 Imaging a one-minute observation takes 30 hours of computational time without $w$-projection.
-We plan to implement some form of multi-scale cleaning in \textsc{WSClean}.
+We plan to implement some form of multi-scale cleaning in \textsc{wsclean}.
 It is likely that additional optimizations need to be made to be able to do this with acceptable performance.
@@
 By using the $w$-stacking algorithms, some computational cost is transferred from gridding to performing FFTs.
-\textsc{WSClean} uses the FFTW library for calculating the FFTs \citep{fftw-2005}.
+\textsc{wsclean} uses the FFTW library for calculating the FFTs \citep{fftw-2005}.
 Further performance improvement can be made by using one of the available FFT libraries that make use of graphical processing units (GPUs).
@@
 \end{figure}
-As discussed in \S\ref{sec:gridding}, \textsc{WSClean} uses a Kaiser-Bessel window against aliasing.
+As discussed in \S\ref{sec:gridding}, \textsc{wsclean} uses a Kaiser-Bessel window against aliasing.
 The type of window function has no effect on the gridding performance, but the size of the kernel affects gridding performance quadratically.
-Fig.~\ref{fig:timing-kernelsize} shows that this quadratic relation becomes significant for kernel sizes $\gtrapprox 10$.
-Decreasing the kernel size to values below $7$ has little effect on performance, because for small kernels the visibility-reading rate is lower than the gridding rate.
+Fig.~\ref{fig:timing-kernelsize} shows that this quadratic relation becomes significant for kernel sizes $\gtrapprox 10$ pixels.
+Decreasing the kernel size to values below seven pixels has little effect on performance, because for small kernels the visibility-reading rate is lower than the gridding rate.
 We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged FOV.
-Therefore, \textsc{WSClean} uses a default of 7 pixels for the gridding kernel.
+Therefore, \textsc{wsclean} uses a default of seven pixels for the gridding kernel.
 It can be increased when necessary.

### 2014-03-01 articles/2014/Article-WSClean/wsclean.tex
Add acknowledgement

@@
 \section*{Acknowledgments}
+Offringa would like to thank O.M. Smirnov for testing \textsc{wsclean} on VLA data.
 This scientific work makes use of the Murchison Radio-astronomy Observatory, operated by CSIRO.

### 2014-03-01 articles/2014/Article-WSClean/wsclean.tex
Comment Ord

@@
 
-Another option exists for imaging MWA data: the Real-Time System (RTS, \citealt{rts-mwa}).
+Another option exists for imaging MWA data: the Real-Time System (RTS, \citealt{rts-mwa,mwa-interferometric-imaging}).
 This has been designed as an efficient calibration and imaging pipeline specifically for MWA data.
-It can use GPUs to improve efficiency, and uses a $w$-projection kernel during imaging to correct for the $w$-term.
-The kernel can be kept small, because the data is processed in snapshots.
+It can use GPUs to improve efficiency.
+It performs snapshot imaging to deal with the $w$-term, which implies that slight variations in tile elevation cause some decorrelation on the longer baselines.
 The RTS can not perform deconvolution as of yet.

### 2014-03-01 articles/2014/Article-WSClean/wsclean.tex
nwlayer 6->12
Slight fixes

@@
 \hline
-\multicolumn{3}{l}{\textit{Zenith angle 0\degree (6 $w$-layers/planes)}} \\
+\multicolumn{3}{l}{\textit{Zenith angle 0\degree (12 $w$-layers/planes)}} \\
 \hline
@@
  \hline
- GLEAM    &  65h 25m & 8h 04m &  8h 03m & 14h 10m   & 1.1 TFLOPS & $6.9 \times 10^2$    \\
+ GLEAM    &  65h 25m & 8h 04m &  8h 03m & 14h 10m   & 1.1 TFLOPS & $6.8 \times 10^2$    \\
  EMU      &    5.2 d &  2.9 d &   2.9 d & 4.4 d     & 9.7 TFLOPS & $2.1 \times 10^4$      \\
- MSSS low &  2h 16m  & 0h 57m &  0h 57m & 2h 10m    & 130 GFLOPS & $3.5 \times 10^3$   \\
+ MSSS low &  2h 16m  & 0h 57m &  0h 57m & 2h 10m    & 130 GFLOPS & $3.4 \times 10^3$   \\
  MSSS high&  7h 16m  & 3h 52m &  3h 52m & 5h 44m    & 530 GFLOPS & $3.4 \times 10^3$   \\
- LOFAR NL &  50.2 d  &  7.3 d &   7.2 d & 12.9 d    &  24 TFLOPS & $4.5 \times 10^3$ \\
+ LOFAR NL &  50.2 d  &  7.3 d &   7.2 d & 12.9 d    &  24 TFLOPS & $7.2 \times 10^2$ \\
  AARTFAAC &  2.5 d   &  1.2 d &   1.2 d &   2.5 d   & 4.1 TFLOPS & $6.8 \times 10^2$   \\
- VLSS     &  0h 09m  & 0h 01m &  0h 01m &  0h 08m   & 2.2 GFLOPS & $12.1 \times 10^3$     \\
+ VLSS     &  0h 09m  & 0h 01m &  0h 01m &  0h 08m   & 2.2 GFLOPS & $1.1 \times 10^3$     \\
  MeerKAT  &   10.8 d &  6.2 d &   6.2 d &  10.8 d   &  21 TFLOPS & $6.8 \times 10^2$ \\
@@
 The configurations are summarized in Table~\ref{tbl:computational-cost-configurations}.
-In multi-beam configurations such as the SKA and EMU configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams.
+In multi-beam configurations such as the EMU and LOFAR configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams.
 The selected wavelengths
@@
 \textsc{wsclean} is currently not able to perform multi-scale clean.
-Results of applying \textsc{casa}'s multi-scale clean on MWA data show that especially the Galactic plane is significantly better deconvolved using multi-scale clean, but it is very expensive.
+Results of applying \textsc{casa}'s multi-scale clean \citep{multiscale-clean-cornwell-2008} on MWA data show that especially the Galactic plane is significantly better deconvolved using multi-scale clean, but it is very expensive.
 Imaging a one-minute observation takes 30 hours of computational time without $w$-projection.

### 2014-03-11 articles/2014/Article-WSClean/wsclean.tex
Implemented Randall's comments

@@
 Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}.
-This deviation is amplified for telescopes with wide field of view (FOV), making this a particularly important issue for low-frequency telescopes that by nature are wide-field instruments.
+The image degradation effects of the $w$-term are amplified for telescopes with wide field of view (FOV), making this a particularly important issue for low-frequency telescopes that by nature are wide-field instruments.
 
@@
 The Common Astronomy Software Applications (\textsc{casa}, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution.
-However, with the MWA we have seen that imaging a 2-min snapshot observation at low-elevation can take up to tens of hours with \textsc{casa}'s $w$-projection algorithm.
+However, with the MWA we have seen that imaging a 2-min snapshot observation away from zenith can take up to tens of hours with \textsc{casa}'s $w$-projection algorithm.
 Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.
@@
 Therefore, the number of added samples and their weight have to be taken into account.
+Therefore, each pixel is divided by the weighted number of samples.
 Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction-dependent effects.
@@
 
-The inverse of imaging, i.e., to calculate the visibility from a model image, can be done by reversing the $w$-stacking algorithm: i)~inverse scale the image; ii)~copy the image to several layers; iii)~inverse apply the direction-dependent phase shift for each layer; iv)~FFT each layer; and v)~sample a required visibility from the correct $w$-layer.
+The inverse of imaging, i.e., to calculate the visibility from a model image, can be done by reversing the $w$-stacking algorithm: i)~multiply the image with the appropriate factor; ii)~copy the image to several layers; iii)~inverse apply the direction-dependent phase shift for each layer; iv)~FFT each layer; and v)~sample a required visibility from the correct $w$-layer.
 We will refer to this operation as prediction.
@@
 
-Using too few $w$-layers results in decorrelation of off-axis sources in the longer baselines.
+Using too few $w$-layers results in decorrelation of sources far from the phase centre in the longer baselines.
 Additionally, $w$-aliasing errors can cause ghost sources in the image.
@@
 Table~\ref{tbl:computational-cost-per-operation} shows how the computational costs scale for the operations that dominate the imaging in the $w$-stacking and $w$-projection algorithms.
-For comparison, we can assume the gridding kernel adds a constant term and the terms $N_\textrm{wlay}$ and $N_\textrm{wkern}$ follow approximately $w_{\max} \sin \alpha_\textrm{FOV}$.
+For comparison, we can assume the antialising kernel can be neglected and the terms $N_\textrm{wlay}$ and $N_\textrm{wkern}$ follow approximately $w_{\max} \sin \alpha_\textrm{FOV}$.
 The time complexity for $w$-stacking is then given by
@@
 \includegraphics[width=13.5cm]{img/fullsky-together}
-\caption{A beam-corrected MWA image of a 112~s zenith observation that covers almost the full sky.
+\caption{A beam-corrected MWA image of a 112~s zenith observation at 180~MHz that covers almost the full sky.
 The lower image shows a zoom-in on the southern side lobe.
@@
 
-A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together.
-A recentred image can be stored in a FITS file with the slant orthographic (SIN) projection normally used in interferometric imaging \citep{fits-coordinates-2002}, by setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}.
+A recentred image is in the projection of the zenith tangent plane and, like warped snapshots, will need an additional regridding step before multiple snapshots can be added together.
+A recentred image can be stored in a FITS file with the orthographic (SIN) projection normally used in interferometric imaging \citep{fits-coordinates-2002}, by setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}.
 Common viewers such as \textsc{kvis} \citep{karma-1996} and \textsc{ds9} support such FITS files and display their coordinates correctly.
@@
 \end{equation}
-where each term is a $2\times2$ complex matrices, with $J$ the voltage beam and $^*$ denoting the conjugate transpose.
+where each term is a $2\times2$ complex matrix, with $J$ the voltage beam and $^*$ denoting the conjugate transpose.
 Because $J$ is complex and non-diagonal, all four combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such.
@@
 \subsection{Gridding} \label{sec:gridding}
-A gridding convolution kernel improves the accuracy of gridding in $uv$-space.
+A gridding convolution kernel improves the accuracy of gridding in $uv$-space \citep{optimal-gridding-schwab-1983}.
 A common kernel function is a windowed sinc function, which acts as a low-pass filter.
@@
 An image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}, is also made and analysed.
-Its accuracy is higher compared to the normal projection, but its residual noise is also higher.
+As can be seen in Table~\ref{tbl:accuracy-measurements}, the source fluxes in the recentred image have smaller errors compared to the normal projection, but the residual noise is higher.
 The reprojection technique needs on average fewer $w$-layers to reach the same level of accuracy.
@@
 At that point, \textsc{wsclean} is in both cases approximately three times faster.
-At zenith, the two methods are expected to perform almost identical.
+At zenith, the two methods are expected to perform almost identically.
 The factor of 2-3 difference in Fig.~\ref{fig:timing-nsamples} could be due to different choices in optimizations.
@@
 The performance of $w$-stacking is more affected by the size of the image, but is still significantly faster in making 12.8K images at 0\degree ZA than $w$-projection.
-Imaging the MWA primary beam requires approximately an image size of 3k pixels for cleaning.
-If the small-inversion optimization of \S\ref{sec:decreasing-inversion-resolution} is used, this decreases to about 1.5k.
+Imaging the MWA primary beam requires approximately an image size of 3072 pixels for cleaning.
+If the small-inversion optimization of \S\ref{sec:decreasing-inversion-resolution} is used, this decreases to about 1500.
 At ZA=10\degree, this saves about a factor of 2 in computing cost.
 The benefit of the optimization increases with resolution.
-At an image size of 10k pixels, cost is decreased by an order of magnitude.
+At an image size of 10,000 pixels, cost is decreased by an order of magnitude.
 
@@
 The hybrid method is in all situations approximately a factor of two slower than \textsc{wsclean}.
-All configurations except VLSS and GLEAM have optimal values for $\Delta w$ much smaller than one, suggesting the $w$-corrections should be done entirely by $w$-stacking instead of the $w$-stacking/projection hybrid method.
+All configurations listed in Table~\ref{tbl:computational-cost-configurations}, with the exception of VLSS and GLEAM, have optimal values for $\Delta w$ much smaller than one, suggesting the $w$-corrections should be done entirely by $w$-stacking instead of the $w$-stacking/projection hybrid method.
 

### 2014-03-11 articles/2014/Article-WSClean/wsclean.tex
Comments of David Kaplan

@@
 \begin{abstract}
-We present a new interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm.
+Widefield interferometric imaging with existing imager implementations is computationally expensive, especially for the large data volumes created by modern non-coplanar many-element arrays.
+We present a new widefield interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm.
 The performance dependencies of \textsc{casa}'s $w$-projection and our new imager are analysed and analytical functions are derived that describe the required computing cost for both imagers.
@@
 We estimate the computing cost for imaging the low-frequency Square-Kilometre Array observations to be 60 PetaFLOPS with current techniques.
-We find that combining $w$-stacking with the $w$-snapshot algorithm does not significantly improve computing requirements.
+We find that combining $w$-stacking with the $w$-snapshot algorithm does not significantly improve computing requirements over pure $w$-stacking.
 The source code of our new imager is publicly released.
@@
 The Common Astronomy Software Applications (\textsc{casa}, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution.
-However, with the MWA we have seen that imaging a 2-min snapshot observation away from zenith can take up to tens of hours with \textsc{casa}'s $w$-projection algorithm.
-Larger images or higher zenith angles can be impossible to image because the size and number of $w$-kernels become too large to hold in memory.
+However, with the MWA we have seen that imaging a 2-min snapshot observation away from zenith can take up to tens of hours with \textsc{casa}'s $w$-projection algorithm, because of the larger $w$-terms for off-zenith observations.
+Imaging with larger image sizes or at higher zenith angles can be impossible because the size and number of $w$-kernels become too large to hold in memory.
 
@@
 It performs snapshot imaging to deal with the $w$-term, which implies that slight variations in tile elevation cause some decorrelation on the longer baselines.
-The RTS can not perform deconvolution as of yet.
+However, the RTS can not perform deconvolution as of yet.
 
@@
 One way to handle DDEs is by using the a-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
-One such DDE is the effect of the ionosphere.
+One particular DDE is the effect of the ionosphere.
 For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure \citep{lonsdale-calibration-approaches-ionosphere}.
@@
 In this case we can therefore calculate the image for $w<0$ from the image with $w>0$.
-That allows us to set $w_{\min}$ to the minimum absolute $w$-value, which requires a factor of two fewer layers.
+That allows us to set $w_{\min}$ to the minimum absolute $w$-value, which requires half the number of layers.
 In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence the inverse FFT is always performed as a complex-to-complex transform.
@@
 For the MWA, a typical value of $w_{\max} - w_{\min}$ is $\sim 10$ at zenith, and reaches $\sim 400$ at a ZA of $30^{\circ}$.
-For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75''$ size, the $\max_{l,m}$ term is $0.68$.
+For a typical full-field-of-view image of a MWA observation of $3072\times 3072$ pixels of $0.75'$ size, the $\max_{l,m}$ term is $0.68$.
 This implies that tens of $w$-layers are required at zenith and hundreds at lower elevations.
@@
 \textsc{wsclean} allows cleaning individual polarizations, or can jointly deconvolve the polarizations.
-In the latter case, peaks are searched in $pp^2+2pq \overline{(pq)} + qq^2$ space, where $\overline{pq}=qp$ is the complex conjugate of $pq$, and the PSF is subtracted from the polarizations with different factors.
+In the latter case, peaks are searched in $pp^2+2pq \overline{(pq)} + qq^2$ space, where $p$ and $q$ are the two polarizations and $\overline{pq}=qp$ is the complex conjugate of $pq$.
+After a peak has been found, the PSF is subtracted from the individual polarizations with different factors.
 
@@
 \caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with \textsc{wsclean}.
+This observation was performed in drift-scan mode.
 The pulsar (centre of image) displays 17\% circular polarization.
@@
 As long as the $uv$-plane is sampled with at least the Nyquist frequency, a high resolution image can be perfectly reconstructed from an inversion at lower resolution.
-After an up-scaled image has been cleaned, a model is created at the same (high) resolution.
+After an enlarged image has been cleaned, a model is created at the same (high) resolution.
 Because the model consists of delta functions, the spatial frequencies in the model are not band-limited.
-Consequently, downscaling the model will remove information from the model.
+Consequently, shrinking the model image will remove information from the model.
 However, only the low spatial frequencies of this image will be used in the prediction.
@@
 Of the 24~min, 77\% is spent on cleaning approximately 100,000 components.
-The outputs with and without rescaling do not visibly differ, but have an RMS difference of 20 mJy.
+The outputs with and without rescaling do not visibly differ, but have an RMS difference of 20 mJy, while the total RMS is 227 mJy.
 We think this can be explained by the non-linear behaviour of clean.
@@
 The full field contains 100 simulated sources over 20\degree.
-Sources in the image produced with \textsc{casa} are slightly less accurately subtracted, leading to a residual noise level of 1.07 mJy and some visible artefacts, whereas \textsc{wsclean} reaches 0.90 mJy.
+Sources in the image produced with \textsc{casa} are slightly less accurately subtracted, leading to a residual noise level of 1.07 mJy and some visible artefacts, whereas \textsc{wsclean} reaches 0.90 mJy RMS noise.
 Since the effect is stronger away from the phase centre, it is likely that this is caused by the finite size of the $w$-kernels used in $w$-projection, which leads to inaccuracies.}

### 2014-03-14 articles/2014/Article-WSClean/wsclean.tex
Steven's comments

@@
 Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}.
-The image degradation effects of the $w$-term are amplified for telescopes with wide field of view (FOV), making this a particularly important issue for low-frequency telescopes that by nature are wide-field instruments.
+The image degradation effects of the $w$-term are amplified for telescopes with wide fields of view (FOV), making this a significant issue for low-frequency telescopes that by nature are wide-field instruments.
 
-There are several methods to deal with the $w$-term during imaging: faceting \citep{facetting-cornwell}, a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}, $w$-projection \citep{wprojection-cornwell}, $w$-stacking \citep{ska-memo-regridding-2011} and $w$-snapshots \citep{widefield-imaging-ska-cornwell}.
+There are several methods to deal with the $w$-term during imaging: faceting \citep{facetting-cornwell}; a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}; $w$-projection \citep{wprojection-cornwell}; $w$-stacking \citep{ska-memo-regridding-2011}; and $w$-snapshots \citep{widefield-imaging-ska-cornwell}.
 
-New generation of wide-field observatories are producing data sets that are orders of magnitude larger than before.
+A new generation of wide-field observatories is producing data sets that are orders of magnitude larger than before.
 Examples of such telescopes include the Murchison Widefield Array (MWA, \citealt{mwa-design-2009}, \citealt{mwa}), the upgraded Jansky Very Large Array (JVLA) and the Low-Frequency Array (LOFAR, \citealt{lofar-2013}).
 The Common Astronomy Software Applications (\textsc{casa}, \citealt{casa-2007}, \citealt{casa-2008}) have an efficient implementation of the $w$-projection algorithm, with many available features such as multi-scale clean and spectral-shape fitting during deconvolution.
-However, with the MWA we have seen that imaging a 2-min snapshot observation away from zenith can take up to tens of hours with \textsc{casa}'s $w$-projection algorithm, because of the larger $w$-terms for off-zenith observations.
+However, with the MWA we have seen that imaging a 2-min snapshot observation away from zenith can take up to tens of wall-clock hours with \textsc{casa}'s $w$-projection algorithm, because of the larger $w$-terms for off-zenith observations.
 Imaging with larger image sizes or at higher zenith angles can be impossible because the size and number of $w$-kernels become too large to hold in memory.
@@
 One particular DDE is the effect of the ionosphere.
-For the MWA it can be assumed that the ionosphere has the same effect on all tiles, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure \citep{lonsdale-calibration-approaches-ionosphere}.
+For the MWA it can be assumed that the ionosphere has the same effect on all antennas, because the maximum baseline length is relatively small (2.9~km) and smaller than the typical size of ionospheric structure \citep{lonsdale-calibration-approaches-ionosphere}.
 This is not the case for LOFAR, making it necessary to correct the direction-dependent ionospheric effects per station before gridding the data.
-The \textsc{awimager} has been written to perform these corrections, and uses a hybrid of a-projection, $w$-projection and $w$-stacking \citep{awimager-2013}.
+The \textsc{awimager} \citep{awimager-2013} has been written to perform these corrections, and uses a hybrid of a-projection, $w$-projection and $w$-stacking.
 
@@
 The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation.
-This shows that the sky function can be reconstructed by i)~gridding samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~applying the direction-dependent phase shift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeating this for all $w$-values and adding the results together; v)~applying the final scaling.
+This shows that the sky function can be reconstructed by: i)~gridding samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~applying the direction-dependent phase shift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeating this for all $w$-values and adding the results together; v)~applying the final scaling.
 
@@
 \end{equation}
-This suggest that a uniform discretization in $w$ is optimal.
+This suggests that a uniform discretization in $w$ is optimal.
 This is in contrast to \citet{wprojection-cornwell} where $\sqrt{w}$ tabulation is suggested.
@@
 \subsection{Time complexity of $w$-stacking} \label{sec:time-complexity}
-It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation.
-We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each side, $N_\textrm{vis}$ is the number of visibilities, $N_\textrm{kern}$ is the size of the anti-aliasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging FOV.
-
 \begin{table}
- \caption{Scaling of the computational cost for various imaging steps} \label{tbl:computational-cost-per-operation}
+ \caption{Scaling of the computational cost for various imaging steps, with $N_\textrm{wlay}$ the number of $w$-layers, $N_\textrm{pix}$ the number of pixels along each side, $N_\textrm{vis}$ the number of visibilities, $N_\textrm{kern}$ the size of the anti-aliasing kernel, $N_\textrm{wkern}$ the size of the $w$-kernel, $w_{\max}$ the maximum $w$ value and $\alpha_\textrm{FOV}$ the imaging FOV.} \label{tbl:computational-cost-per-operation}
  \begin{tabular}{lll}
@@
 \end{table}
+
+It is useful to analyse the time complexity of $w$-stacking and compare it with $w$-projection, to understand which algorithm performs better in a given situation.
+We will use the following symbols: $N_\textrm{wlay}$ is the number of $w$-layers for $w$-stacking, $N_\textrm{pix}$ is the number of pixels in the image along each side, $N_\textrm{vis}$ is the number of visibilities, $N_\textrm{kern}$ is the size of the anti-aliasing kernel (see \S\ref{sec:gridding}), $N_\textrm{wkern}$ is the size of the $w$-kernel for $w$-projection, $w_{\max}$ is the maximum $w$ value and $\alpha_\textrm{FOV}$ is the imaging FOV.
 
@@
 
-To efficiently image the full sky, an observation is split into short snapshots, each of which is phase-rotated to zenith and subsequently imaged, with an appropriate number of pixels and resolution.
+To efficiently image the full sky, an observation is split into short snapshots, their phase centres are changed to zenith and the snapshots are subsequently imaged, with an appropriate number of pixels and resolution.
 Depending on desired resolution and dynamic range, this might require very large images.
@@
 An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13-min MWA observation imaged with \textsc{wsclean}.
-The processing steps performed for this image were i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibration using Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) imaging of the seven 112 s intervals separately.
+The processing steps performed for this image were: i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibration using Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) imaging of the seven 112 s intervals separately.
 Cleaning an image that is in a different projection yields slightly different results, but qualitatively the two images are clearly of equal accuracy.
-The wall-clock time for imaging is 60 min and 41 min for respectively normal projection and the recentred image.
+The wall-clock time for imaging is 60 min and 41 min for normal projection and the recentred image, respectively.
 Of the 41 min, 3 min is spent on phase rotating the visibilities.
@@
 \textsc{casa} convolves samples with a PSWF of seven pixels total width during gridding.
-When using a variable kernel size, a PSWF is quite complicated and computational expensive to calculate.
+When using a variable kernel size, a PSWF is quite complicated and computationally expensive to calculate.
 \textsc{wsclean} currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation of the PSWF \citep{fourier-kernel-selection-1991}.
@@
 \subsection{Decreasing inversion resolution} \label{sec:decreasing-inversion-resolution}
-Typically, for cleaning it is desired that images have a pixel size at least five times smaller than the synthesized-beam width.
+Typically, for cleaning it is desired that images have a pixel size (side length) at least five times smaller than the synthesized-beam width.
 This improves the accuracy of the clean algorithm, because the positions of image maxima will be closer to the actual source positions.
@@
 Because the model consists of delta functions, the spatial frequencies in the model are not band-limited.
-Consequently, shrinking the model image will remove information from the model.
+Consequently, decimating the model image will remove information from the model.
 However, only the low spatial frequencies of this image will be used in the prediction.
-Therefore, inversion at higher resolution is not required: as long as long as downscaling does not modify the low-frequency components, the output of the prediction step will not change.
-This requires low-pass filtering the model before decimation.
+Therefore, inversion at higher resolution is not required: as long as the decimation does not modify the low-frequency components, the output of the prediction step will not change.
+This requires low-pass filtering of the model before decimation.
 
@@
 \end{equation}
-where $N_\textrm{Nyquist}$ the resolution in pixels used during inversion, $N_\textrm{pix}$ the requested number of pixels of the image along one side, $S_\textrm{pix}$ the requested angular pixel scale and $S_\textrm{synth beam}$ the minimum angular size of the synthesized beam.
-Before cleaning, the low-resolution image is interpolated with a procedure consisting of i) a low-resolution FFT; ii) zero padding; and iii) a high-resolution inverse FFT.
+where $N_\textrm{Nyquist}$ is the resolution in pixels used during inversion, $N_\textrm{pix}$ is the requested number of pixels of the image along one side, $S_\textrm{pix}$ is the requested angular pixel scale and $S_\textrm{synth beam}$ is the minimum angular size of the synthesized beam.
+Before cleaning, the low-resolution image is interpolated with a procedure consisting of: i) a low-resolution FFT; ii) zero padding; and iii) a high-resolution inverse FFT.
 Such interpolation assumes the image to be periodic, but this is already assumed during the inversion process.
-After cleaning, the model is decimated by the inverse procedure, consisting of i) a high-resolution FFT; ii) truncation; and iii) a low-resolution inverse FFT.
-With common settings, this procedure decreases the inversion imaging resolution with a factor of 2.5.
+After cleaning, the model is decimated by the inverse procedure, consisting of: i) a high-resolution FFT; ii) truncation; and iii) a low-resolution inverse FFT.
+With common settings, this procedure decreases the inversion imaging resolution by a factor of 2.5.
 Without further correction, such a decrease would lower the positional accuracy with which visibilities are gridded onto the $uv$-plane.
@@
 The outputs with and without rescaling do not visibly differ, but have an RMS difference of 20 mJy, while the total RMS is 227 mJy.
-We think this can be explained by the non-linear behaviour of clean.
+Because the difference is mostly noise like, the difference could be caused by the non-linear behaviour of clean.
 
@@
 We will now analyse the performance and accuracy of our implementation, and compare it with the $w$-projection implementation in \textsc{casa}.
-For the analysis, we use imaging parameters which are common for MWA imaging.
+For the analysis, we use imaging parameters common for MWA imaging.
 When a specific parameter value is not mentioned, the settings from Table~\ref{tbl:default-parameters} are used.
@@
 This yields 195 $w$-planes/layers at 10\degree ZA.
-Several configurations with large $w$-values fail to image with \textsc{casa}, presumably because the $w$-kernels become too large.
+Several configurations with large $w$-values fail to image with \textsc{casa}, because \textsc{casa} crashes during the imaging, presumably because the $w$-kernels become too large.
 
-The used software version of \textsc{casa} is ``stable release 42.0, revision 26465'', which was released September 2013.
-For \textsc{wsclean}, the latest version as of February 2014 was used.
+The software version of \textsc{casa} is ``stable release 42.0, revision 26465'', which was released September 2013.
+For \textsc{wsclean}, version 1.0 from February 2014 was used.
 The tests were run on a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores that can perform 138 giga-floating point operations per second (GFLOPS).
@@
 Table~\ref{tbl:accuracy-measurements} lists the measured root mean square (RMS) in the residual image and the standard errors of the source brightnesses as detected by \textsc{Aegean}.
-In all tests, \textsc{wsclean} performs slightly more accurately.
-At zenith, \textsc{casa} produces a residual RMS twice as large as \textsc{wsclean}.
-This is caused by the fact that we keep the number of $w$-projection planes in $w$-projection equal to the number of $w$-layers in $w$-stacking, resulting in only 12 $w$-planes at zenith.
+\textsc{wsclean} is more accurate: it shows 2-33\% lower errors in the source fluxes and produces 0-49\% lower RMS noise compared to \textsc{casa}.
+The large residual RMS twice for \textsc{casa} at zenith is caused by the fact that we keep the number of $w$-projection planes in $w$-projection equal to the number of $w$-layers in $w$-stacking, resulting in only 12 $w$-planes at zenith.
 This evidently has a stronger effect on the $w$-projection algorithm.
@@
 In the $\textrm{ZA}=10\degree$ case, \textsc{casa} is slightly less accurate.
-The extra noise can be observed in the produced images, as shown in Fig.~\ref{fig:residuals}.
-An image with the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}, is also made and analysed.
+Extra noise can be seen in the images, as shown in Fig.~\ref{fig:residuals}.
+An image resulting from the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}, is also made and analysed.
 As can be seen in Table~\ref{tbl:accuracy-measurements}, the source fluxes in the recentred image have smaller errors compared to the normal projection, but the residual noise is higher.
@@
 Both imagers perform five major iterations for these results, and the inversions and predictions are dominating the computing time.
-We will look more closely at the difference in performance in the next section.
+We will look more closely at the differences in performance in the next section.
 
@@
 Each specific configuration is run five times and standard deviations are calculated.
-Variation between runs is typically a few seconds.
+The variation in duration between runs is typically a few seconds.
 In each benchmark, the wall-clock time is measured that is required to produce the synthesized point-spread function and the image itself.
@@
 
-In Panel~\ref{fig:timing-zenith-angle}, the computational cost as function of ZA is plotted.
+In Panel~\ref{fig:timing-zenith-angle}, the computational cost as a function of ZA is plotted.
 It shows that, as expected from \S\ref{sec:time-complexity}, the $w$-projection algorithm is more affected by the increased $w$-values, leading to differences of more than an order of magnitude at ZAs of $\gtrapprox 20$\degree.
@@
 This is dependent on the speed of rotating the phase centre and regridding.
-Our program to change the phase centre is currently not multi-threaded, so its performance should be improvable.
+Our program to change the phase centre is currently not multi-threaded, so its performance could be improved.
 Furthermore, the relative cost of changing the phase centre is lower when performing multiple major iterations.
@@
 Based on the expected cost terms described in Sec.~\ref{sec:time-complexity}, we derive analytical functions for the \textsc{casa} and \textsc{wsclean} measurements using least-squares fitting.
-Several functions with different free parameters were tested, and formulae with minimum amount of parameters are selected that still follow the trend of the measurements and have reasonably small errors.
+Several functions with different free parameters were tested, and formulae with minimum number of parameters are selected that still follow the trend of the measurements and have reasonably small errors.
 Measurements are weighted with the inverse standard deviation instead of the variance, because we found that the latter results in too much weight on fast configurations, leading to functions that represent the general trend less well.
@@
 where $D$ is the maximum baseline length, $\Delta z_{\max}$ is the maximum height difference between antennas and $\lambda$ is the wavelength, all in meters, and $\xi$ is an extra parameter for fitting the \textsc{casa} measurements, $\xi_\textrm{CASA}=28.4$ and $\xi_\textrm{WSClean}=0$.
-These functions follow the trend of the measurements well, with an absolute error of 14.8\% and 20.7\% for respectively the \textsc{wsclean} and \textsc{casa} functions.
+These functions follow the trend of the measurements well, with an absolute error of 14.8\% and 20.7\% for the \textsc{wsclean} and \textsc{casa} functions, respectively.
 
@@
 We have excluded the cost of phase shifting the visibilities and gridding.
-The cost for regridding with simple nearest neighbour or bilinair interpolation is indeed negligible, although more accurate interpolation (e.g.
+The cost for regridding with simple nearest neighbour or bilinear interpolation is indeed negligible, although more accurate interpolation (e.g.
 Lanczos interpolation) can be expensive.
@@
 To combine the $w$-projection and $w$-stacking algorithms, small $w$-corrections are made before the FFTs using a $w$-correcting kernel and large $w$-terms are corrected after the FFTs by gridding onto several $w$-layers.
-This allows using small $w$-kernels and limits at the same time the number of FFTs and required memory.
+This allows using small $w$-kernels during the $w$-projection stage and limits at the same time the number of FFTs and required memory that pure $w$-stacking would require.
 The \textsc{awimager} has been applied in this way, which led to better results compared to $w$-projection without $w$-stacking \citep{awimager-2013}.
-For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more generally the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with correspondingly smaller $w_{\max}$ value and smaller $N_\textrm{vis}$.
+For this scenario, Eq.~\eqref{eq:casa-computing-cost} can be used to determine the optimal number of $w$-layers, or more generally the optimal distance between layers, by assuming that the cost of calculating a single $w$-layer equals the cost of imaging the data set with a correspondingly smaller $w_{\max}$ value and smaller $N_\textrm{vis}$.
 If $\Delta w$ is the distance between $w$-layers, then
@@
 with $N'_\textrm{vis} = N_\textrm{vis}\frac{\Delta w}{w_{\max}}$ and $w'_{\max} = \Delta w$.
-For MWA observations, the optimal value for $\Delta w$ is about unity, and improves the speed of $w$-projection with about a factor of four.
-However, the performance of the hybrid method is still about a factor of two lower than our pure-stacking implementation.
+For MWA observations, the optimal value for $\Delta w$ is about unity, and improves the speed of $w$-projection by a factor of four.
+However, the performance of the hybrid method is still approximately a factor of two lower than our pure-stacking implementation.
 This could again be the difference in optimisation choices between \textsc{casa} and \textsc{wsclean}, but it does show that the effort of implementing a hybrid over pure stacking might not be worthwhile.
@@
 \textsc{wsclean} is in most situations predicted to be 2--3 times faster than \textsc{casa}.
-\textsc{wsclean} has the largest benefit on the full LOFAR, MWA and the full SKA, where \textsc{wsclean} is respectively 7, 8 and 12 times faster.
+\textsc{wsclean} has the largest benefit on the full LOFAR, MWA and the full SKA, where \textsc{wsclean} is 7, 8 and 12 times faster, respectively.
 The $w$-snapshot method does not improve much over normal \textsc{wsclean} operation in any of the cases.
@@
 We have shown that the $w$-stacking algorithm is well-suited for imaging MWA observations.
-The \textsc{wsclean} $w$-stacking implementation is faster than \textsc{casa}'s $w$-projection algorithm in all common MWA imaging configurations, causing up to an order of magnitude increase in speed at a relatively small ZA of $10$\degree, and is slightly more accurate.
-Roughly speaking, snapshot imaging becomes faster for ZAs $>15\degree$ or FOVs $>35\degree$.
+The \textsc{wsclean} $w$-stacking implementation is faster than \textsc{casa}'s $w$-projection algorithm in all common MWA imaging configurations, giving up to an order of magnitude increase in speed at a relatively small ZA of $10$\degree, and is slightly more accurate.
+Roughly speaking, for ZAs $>15\degree$ or FOVs $>35\degree$ snapshot imaging becomes faster.
 This can lead to performance improvements of a factor of $3$ for the MWA, but only in the most expensive imaging configurations that are less commonly used.
@@
 Clearly it remains challenging to perform wide-field imaging with acceptable performance, but some optimisations could be made for the SKA.
-For example, a performance gain of a few factors can be achieved by averaging shorter baselines to their maximum integration time and bandwidth resolution before imaging.
+For example, a performance gain of factors of a few can be achieved by averaging shorter baselines to their maximum integration time and bandwidth resolution before imaging.
 Moreover, although longer baselines make the imaging more expensive, imaging the full FOV will likely not (always) be required when using the longest baselines.
@@
 Because of sufficient instantaneous $uv$-coverage and near-confusion snapshot noise level, snapshots can be cleaned individually to deep levels.
-For example, cleaning to $5\sigma$ level results in a cleaning threshold of approximately 100~mJy in 112~s observations.
+For example, cleaning to a $5\sigma$ level results in a cleaning threshold of approximately 100~mJy in 112~s observations.
 However, MWA's EoR observations \citep{bowman-science-with-the-mwa-2013} require more advanced deconvolution techniques.
@@
 This is computationally cheap for our current snapshot time of 112s, but this might be infeasible when it is required to calculate the beam on very short time scales.
-Accurate MWA beam models are still being developed.
+Accurate MWA beam models are still being developed (Sutinjo et al., in prep.).
 When beam corrections are required on short time scales, calculating $a$-projection kernels or performing snapshot imaging also becomes more expensive, and it would be interesting to find out where the balance lies between these methods.

### 2014-03-14 articles/2014/Article-WSClean/wsclean.tex
Replacing fullsky image with better axes
Spelling error

@@
 \begin{center}
-\includegraphics[width=13.5cm]{img/fullsky-together}
+\includegraphics[height=22.5cm]{img/fullsky-together}
 \caption{A beam-corrected MWA image of a 112~s zenith observation at 180~MHz that covers almost the full sky.
@@
 An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13-min MWA observation imaged with \textsc{wsclean}.
-The processing steps performed for this image were: i) preprocessing of the observation with the Cotter preprocessing pipeline, which include time averaging and RFI flagging with the \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibration using Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) imaging of the seven 112 s intervals separately.
+The processing steps performed for this image were: i) preprocessing of the observation with the Cotter preprocessing pipeline, which includes time averaging and RFI flagging with the \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibration using Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) imaging of the seven 112 s intervals separately.
 Cleaning an image that is in a different projection yields slightly different results, but qualitatively the two images are clearly of equal accuracy.

### 2014-03-15 articles/2014/Article-WSClean/wsclean.tex
Comments made by Mitch

@@
 
-There are several methods to deal with the $w$-term during imaging: faceting \citep{facetting-cornwell}; a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}; $w$-projection \citep{wprojection-cornwell}; $w$-stacking \citep{ska-memo-regridding-2011}; and $w$-snapshots \citep{widefield-imaging-ska-cornwell}.
+There are several methods to deal with the $w$-term during imaging: faceting \citep{facetting-cornwell}; a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}; $w$-projection \citep{wprojection-cornwell}; $w$-stacking \citep{ska-memo-regridding-2011}; and warped snapshots \citep{perley-noncoplanar-arrays}.
+Hybrid methods are sometimes useful, such as with the $w$-snapshots method \citep{widefield-imaging-ska-cornwell}.
 
@@
 It can use GPUs to improve efficiency.
-It performs snapshot imaging to deal with the $w$-term, which implies that slight variations in tile elevation cause some decorrelation on the longer baselines.
-However, the RTS can not perform deconvolution as of yet.
+Snapshot imaging is performed to deal with the $w$-term, which implies that slight variations in tile elevation cause some decorrelation on the longer baselines.
+As the RTS was designed as a single-pass stream processor, standard iterative deconvolution algorithms are not available.
+Compact emission can be subtracted and peeled from visibilities using a sky model and calibration updates, but updates to the sky model need to be realised using separate forward-modelling routines \citep{forward-modelling-bernardi-2011, pindor-subtracting-sources-2011}.
 
@@
 This is especially true for wide-field telescopes.
-One way to handle DDEs is by using the a-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
+One way to handle DDEs is by using the $a$-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
 One particular DDE is the effect of the ionosphere.
@@
 The benefit of our method is that the resulting image has a circular synthesized beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning.
-Our method is also relatively easy to implement.
-It has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on ZA, FOV and the non-coplanarity.
+Regridding a recentred image is more straightforward compared to regridding a warped snapshot, because in the case of warped snapshots there is no analytic solution to the coordinate conversion \citep{perley-noncoplanar-arrays}.
+A benefit of warped snapshots is that the $w$-term errors are zero at image centre and get worse with the distance from image centre.
+With our approach, the effect gets worse with the distance from zenith.
+Our technique has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on ZA, FOV and the non-coplanarity.
 

### 2014-03-17 articles/2014/Article-WSClean/wsclean.tex
Adding reference to Rau and Cornwell
Removing appendix as suggested by Tingay

@@
 For example, self-calibration using full-sky clean components has been found to give good results in imaging the resolved FR-II radio source Fornax~A (McKinley et al., in prep.).
-An example full-sky MWA image is shown in Fig.~\ref{fig:full-sky-example}.
+An example of a full-sky MWA image is shown in Fig.~\ref{fig:full-sky-example}.
 
@@
 The benefit of our method is that the resulting image has a circular synthesized beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning.
-Regridding a recentred image is more straightforward compared to regridding a warped snapshot, because in the case of warped snapshots there is no analytic solution to the coordinate conversion \citep{perley-noncoplanar-arrays}.
+Regridding a recentred image is also more straightforward compared to regridding a warped snapshot, because in the latter case there is no analytic solution to the coordinate conversion \citep{perley-noncoplanar-arrays}.
 A benefit of warped snapshots is that the $w$-term errors are zero at image centre and get worse with the distance from image centre.
@@
 \textsc{wsclean} currently uses a Kaiser-Bessel (KB) window function, which is easy and fast to compute, and is a good approximation of the PSWF \citep{fourier-kernel-selection-1991}.
-After some benchmarks (see Appendix~\ref{sec:gridding-appendix}) we found that seven pixels is a good compromise between performance and accuracy, and this is the default for \textsc{wsclean}.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
+\caption{Effect of anti-aliasing kernel size on performance.}
+\label{fig:timing-kernelsize}
+\end{center}
+\end{figure}
+
+The type of window function has no effect on the gridding performance, but the size of the kernel affects gridding performance quadratically.
+Fig.~\ref{fig:timing-kernelsize} shows that this quadratic relation becomes significant for kernel sizes $\gtrapprox 10$ pixels.
+Decreasing the kernel size to values below seven pixels has little effect on performance, because for small kernels the visibility-reading rate is lower than the gridding rate.
+We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged FOV.
+Therefore, \textsc{wsclean} uses a default of seven pixels for the gridding kernel.
+It can be increased when necessary.
+Of course, different hardware might give slightly different results because of different reading and calculation performance.
 
@@
 It is likely that additional optimizations need to be made to be able to do this with acceptable performance.
+Combining multi-scale with wide-band deconvolution techniques is a possible further improvement \citep{rau-msmfs-2011}.
 
@@
 We acknowledge the iVEC Petabyte Data Store, the Initiative in Innovative Computing and the CUDA Center for Excellence sponsored by NVIDIA at Harvard University, and the International Centre for Radio Astronomy Research (ICRAR), a Joint Venture of Curtin University and The University of Western Australia, funded by the Western Australian State government.
-\appendix
-
-\section{Optimal gridding kernel size} \label{sec:gridding-appendix}
-\begin{figure}
-\begin{center}
-\includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
-\caption{Effect of anti-aliasing kernel size on performance.}
-\label{fig:timing-kernelsize}
-\end{center}
-\end{figure}
-As discussed in \S\ref{sec:gridding}, \textsc{wsclean} uses a Kaiser-Bessel window against aliasing.
-The type of window function has no effect on the gridding performance, but the size of the kernel affects gridding performance quadratically.
-Fig.~\ref{fig:timing-kernelsize} shows that this quadratic relation becomes significant for kernel sizes $\gtrapprox 10$ pixels.
-Decreasing the kernel size to values below seven pixels has little effect on performance, because for small kernels the visibility-reading rate is lower than the gridding rate.
-We have not noticed much benefit of larger kernels except in rare cases where a bright source lies just outside the imaged FOV.
-Therefore, \textsc{wsclean} uses a default of seven pixels for the gridding kernel.
-It can be increased when necessary.
-Of course, different hardware might give slightly different results because of different reading and calculation performance.
 

### 2014-03-17 articles/2014/Article-WSClean/wsclean.tex
Slight rewording

@@
 \textsc{wsclean} is more accurate: it shows 2-33\% lower errors in the source fluxes and produces 0-49\% lower RMS noise compared to \textsc{casa}.
-The large residual RMS twice for \textsc{casa} at zenith is caused by the fact that we keep the number of $w$-projection planes in $w$-projection equal to the number of $w$-layers in $w$-stacking, resulting in only 12 $w$-planes at zenith.
+The large residual RMS for \textsc{casa} at zenith is caused by the fact that we keep the number of $w$-projection planes in $w$-projection equal to the number of $w$-layers in $w$-stacking, resulting in only 12 $w$-planes at zenith.
 This evidently has a stronger effect on the $w$-projection algorithm.
@@
 We do not know why this parameter needs to be higher in \textsc{casa} to reach the same RMS.
-We are using enough $w$-planes to cover the sources: Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlay}\gg 6$ for the top-left image pixel that is 26\degree away from the centre, while the sources are all within 10\degree.
+We are using enough $w$-planes to cover the sources: Eq.~\eqref{eq:nwlayers-bound} results in $N_\textrm{wlay}\gg 1$ for the source furthest from the phase centre.
 Also unexpected is that the source flux density becomes worse by increasing the number of planes.

### 2014-03-29 articles/2014/Article-WSClean/wsclean.tex

@@
 
-\title[WSClean: a fast, generic wide-field imager]{WSClean: an implementation of a fast, generic wide-field imager}
+\title[WSClean: a fast, generic wide-field imager]{WSClean: an implementation of a fast, generic wide-field imager for radio astronomy}
 
@@
 \def\Tata{$^{15}$}
-\def\NRAO{$^{16}$}
-\def\UMelbourne{$^{17}$}
+\def\UMelbourne{$^{16}$}
+\def\NRAO{$^{17}$}
+\def\UMichigan{$^{18}$}
 
@@
 N.~Hurley-Walker\Curtin,
-F.~H.~Briggs$^{1,2}$, \newauthor
-R.~B.~Wayth\Curtin$^,$\CAASTRO, 
+F.~H.~Briggs$^{1,2}$,
+R.~B.~Wayth\Curtin$^,$\CAASTRO, \newauthor
 D.~L.~Kaplan\UWisc,
@@
 A.~R.~Neben\MIT,
-J.~D.~Hughes\ANU, \newauthor
-J.~Rhee\ANU,
+J.~D.~Hughes\ANU, 
+J.~Rhee\ANU, \newauthor
 T.~Murphy\USydney$^,$\CAASTRO,
+R.~Bhat\Curtin,
 G.~Bernardi\SKASA,
@@
 D.~Emrich\Curtin,
+A.~Ewall-Wice\MIT,
 B.~M.~Gaensler\USydney$^,$\CAASTRO,
-R.~Goeke\MIT,
-L.~J.~Greenhill\CfA, \newauthor
+R.~Goeke\MIT, \newauthor
+L.~J.~Greenhill\CfA, 
 B.~J.~Hazelton\UW, 
@@
 M.~Johnston-Hollitt\Victoria,
-J.~C.~Kasper\CfA, \newauthor
+D.~C.~Jacobs\ASU, \newauthor
+J.~C.~Kasper\UMichigan$^,$\CfA,
 E.~Kratzenberg\Haystack, 
+E.~Lenc\USydney,
 C.~J.~Lonsdale\Haystack,
 M.~J.~Lynch\Curtin,
-S.~R.~McWhirter\Haystack,
-D.~A.~Mitchell\CASS$^,$\CAASTRO, \newauthor
+S.~R.~McWhirter\Haystack, \newauthor
+D.~A.~Mitchell\CASS$^,$\CAASTRO,
 M.~F.~Morales\UW, 
 E.~Morgan\MIT,
-D.~Oberoi\Tata,
+N.~Kudryavtseva\Curtin,
+D.~Oberoi\Tata, \newauthor
 S.~M.~Ord\Curtin$^,$\CAASTRO,
+P.~Procopio\UMelbourne,
 T.~Prabu\RRI,
-A.~E.~E.~Rogers\Haystack, \newauthor
-D.~A.~Roshi\NRAO, %\footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
+J.~Riding\UMelbourne, 
+A.~E.~E.~Rogers\Haystack,
+D.~A.~Roshi\NRAO, \newauthor %\footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
 N.~Udaya~Shankar\RRI,
 K.~S.~Srivani\RRI,
-R.~Subrahmanyan\RRI$^,$\CAASTRO, \newauthor
+R.~Subrahmanyan\RRI$^,$\CAASTRO,
 S.~J.~Tingay\Curtin$^,$\CAASTRO, 
-M.~Waterson\Curtin$^,$\ANU,
+M.~Waterson\Curtin$^,$\ANU, \newauthor
 R.~L.~Webster\UMelbourne$^,$\CAASTRO, 
-A.~R.~Whitney\Haystack, \newauthor
+A.~R.~Whitney\Haystack, 
 A.~Williams\Curtin,
@@
 \Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India\\
+\UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
 \NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
-\UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
 }
@@
 \begin{abstract}
-Widefield interferometric imaging with existing imager implementations is computationally expensive, especially for the large data volumes created by modern non-coplanar many-element arrays.
+Widefield interferometric imaging of radio data with existing imager implementations is computationally expensive, especially for the large data volumes created by modern non-coplanar many-element arrays.
 We present a new widefield interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm.
@@
 Extrapolation of our results shows that the current imagers require $3$--$60$ petaFLOPS for SKA1 low alone.
+Although we have measured our performance in FLOPS, it is likely that memory bandwidth will be a limiting factor, because memory bandwidth is increasing less quickly compared to the floating point performance (e.g., \citealt{gpu-gridding-romein-2012}).
 It was suggested that $w$-snapshots would improve the imaging speed for the SKA situation \citep{widefield-imaging-ska-cornwell}, but our results show this is not the case when already using the $w$-stacking algorithm.

### 2014-03-29 articles/2014/Article-WSClean/wsclean.tex
Adding Luke as coauthor, correcting Emil's affiliation

@@
 T.~Murphy\USydney$^,$\CAASTRO,
-R.~Bhat\Curtin,
+N.~D.~R.~Bhat\Curtin,
 G.~Bernardi\SKASA,
@@
 J.~N.~Hewitt\MIT,
-M.~Johnston-Hollitt\Victoria,
-D.~C.~Jacobs\ASU, \newauthor
+L.~Hindson\Victoria,
+M.~Johnston-Hollitt\Victoria, \newauthor
+D.~C.~Jacobs\ASU,
 J.~C.~Kasper\UMichigan$^,$\CfA,
 E.~Kratzenberg\Haystack, 
-E.~Lenc\USydney,
+E.~Lenc\USydney$^,$\CAASTRO,
 C.~J.~Lonsdale\Haystack,
-M.~J.~Lynch\Curtin,
-S.~R.~McWhirter\Haystack, \newauthor
+M.~J.~Lynch\Curtin, \newauthor
+S.~R.~McWhirter\Haystack,
 D.~A.~Mitchell\CASS$^,$\CAASTRO,
@@
 E.~Morgan\MIT,
-N.~Kudryavtseva\Curtin,
-D.~Oberoi\Tata, \newauthor
+N.~Kudryavtseva\Curtin, \newauthor
+D.~Oberoi\Tata,
 S.~M.~Ord\Curtin$^,$\CAASTRO,
@@
 J.~Riding\UMelbourne, 
-A.~E.~E.~Rogers\Haystack,
-D.~A.~Roshi\NRAO, \newauthor %\footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
+A.~E.~E.~Rogers\Haystack, \newauthor
+D.~A.~Roshi\NRAO, %\footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
 N.~Udaya~Shankar\RRI,
@@
 R.~Subrahmanyan\RRI$^,$\CAASTRO,
-S.~J.~Tingay\Curtin$^,$\CAASTRO, 
-M.~Waterson\Curtin$^,$\ANU, \newauthor
+S.~J.~Tingay\Curtin$^,$\CAASTRO, \newauthor
+M.~Waterson\Curtin$^,$\ANU,
 R.~L.~Webster\UMelbourne$^,$\CAASTRO, 

### 2014-03-29 articles/2014/Article-WSClean/wsclean.tex
Emils comments

@@
 
-In this article, we present a new implementation of a generic wide-field imager that is significantly faster than \textsc{casa}'s $w$-projection implementation, and is able to successfully image any MWA observation.
+In this article, we present a new implementation of a generic wide-field imager that is significantly faster than \textsc{casa}'s $w$-projection implementation.
 To obtain the increase in speed, the implementation uses the $w$-stacking method for correcting the $w$-terms, optionally combined with a new technique for $w$-snapshot imaging.
@@
 
-Imaging the full FOV with the full SKA becomes very expensive due to the high frequency and time resolution, and image resolution of $7.2\textrm{k} \times 7.2$k pixels.
+Imaging the full FOV with the full SKA becomes very expensive due to the high frequency and time resolution, and image size of $7.2\textrm{k} \times 7.2$k pixels.
 This translates to a computing power requirement of $\sim60$ PetaFLOPS.
@@
 We have shown that the $w$-stacking algorithm is well-suited for imaging MWA observations.
-The \textsc{wsclean} $w$-stacking implementation is faster than \textsc{casa}'s $w$-projection algorithm in all common MWA imaging configurations, giving up to an order of magnitude increase in speed at a relatively small ZA of $10$\degree, and is slightly more accurate.
+The \textsc{wsclean} $w$-stacking implementation is faster than \textsc{casa}'s $w$-projection algorithm in all common MWA imaging configurations, giving up to an order of magnitude increase in speed at a relatively small ZA of $10$\degree, and results in slightly lower imaging errors.
 Roughly speaking, for ZAs $>15\degree$ or FOVs $>35\degree$ snapshot imaging becomes faster.
@@
 \textsc{wsclean} is currently not able to perform multi-scale clean.
-Results of applying \textsc{casa}'s multi-scale clean \citep{multiscale-clean-cornwell-2008} on MWA data show that especially the Galactic plane is significantly better deconvolved using multi-scale clean, but it is very expensive.
+Results of applying \textsc{casa}'s multi-scale clean \citep{multiscale-clean-cornwell-2008} on MWA data show that especially the Galactic plane is significantly better deconvolved using multi-scale clean, but it is very computationally expensive.
 Imaging a one-minute observation takes 30 hours of computational time without $w$-projection.

### 2014-03-29 articles/2014/Article-WSClean/wsclean.tex
Renaming sdf to editmodel

@@
 In practice, the final scaling will be different from $\left(w_{\max} - w_{\min}\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples.
-Therefore, the number of added samples and their weight have to be taken into account.
 Therefore, each pixel is divided by the weighted number of samples.
 Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction-dependent effects.
-In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds.
+In equally-polarized baselines (e.g., XX, YY, LL or RR), a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds.
 The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-value samples.
@@
 \section*{Acknowledgments}
-Offringa would like to thank O.M. Smirnov for testing \textsc{wsclean} on VLA data.
+Offringa would like to thank O.~M.~Smirnov for testing \textsc{wsclean} on VLA data.
 This scientific work makes use of the Murchison Radio-astronomy Observatory, operated by CSIRO.

### 2014-03-31 articles/2014/Article-WSClean/wsclean.tex
Comments by Bryan G

--- 2014-03-29 articles/2014/Article-WSClean/wsclean.tex
@@
 \begin{abstract}
-Widefield interferometric imaging of radio data with existing imager implementations is computationally expensive, especially for the large data volumes created by modern non-coplanar many-element arrays.
+Astronomical widefield imaging of interferometric radio data is computationally expensive, especially for the large data volumes created by modern non-coplanar many-element arrays.
 We present a new widefield interferometric imager that uses the $w$-stacking algorithm and can make use of the $w$-snapshot algorithm.
 The performance dependencies of \textsc{casa}'s $w$-projection and our new imager are analysed and analytical functions are derived that describe the required computing cost for both imagers.
-On data from the Murchison Widefield Array, we find our new method to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with correct polarization correction.
+On data from the Murchison Widefield Array, we find our new method to be an order of magnitude faster than $w$-projection, as well as being capable of full-sky imaging at full resolution and with correct polarisation correction.
 We predict the computing costs for several other arrays and estimate that our imager is a factor of 2--12 faster, depending on the array configuration.
@@
 \section{Introduction}
-Visibility data from non-coplanar interferometric radio telescopes which observe large fractions of the sky at once can not be accurately imaged with a two-dimensional fast Fourier transform (FFT).
+Visibility data from non-coplanar interferometric radio telescopes that observe large fractions of the sky at once can not be accurately imaged with a two-dimensional fast Fourier transform (FFT).
 Instead, the imaging algorithm needs to account for the ``$w$-term'' during inversion, which is the term that describes the deviation of the array from a perfect plane \citep{perley-noncoplanar-arrays}.
@@
 We will not discuss calibration, but assume $V$ has been calibrated before imaging.
-In the case of a polarized measurement, the symbols become $2\times 2$ matrices and beam correction is more complicated, but without loss of generality we will ignore polarization and treat inversion as a scalar problem.
+In the case of a polarised measurement, the symbols become $2\times 2$ matrices and beam correction is more complicated, but without loss of generality we will ignore polarisation and treat inversion as a scalar problem.
 Imaging consists of inverting Eq.~\eqref{eq:visibility-function}, i.e., to find $I'$ from $V$.
@@
 In practice, the final scaling will be different from $\left(w_{\max} - w_{\min}\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples.
-Therefore, the number of added samples and their weight have to be taken into account.
+Therefore, the number of added samples and their weights have to be taken into account.
 Therefore, each pixel is divided by the weighted number of samples.
 Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction-dependent effects.
-In XX, YY, LL or RR polarizations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds.
+In XX, YY, LL or RR polarisations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds.
 The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-value samples.
@@
 That allows us to set $w_{\min}$ to the minimum absolute $w$-value, which requires half the number of layers.
-In any case, the input to the two-dimensional inverse FFT is not generally an Hermitian-symmetric function, and hence the inverse FFT is always performed as a complex-to-complex transform.
+In any case, the input to the two-dimensional inverse FFT is not generally a Hermitian-symmetric function, and hence the inverse FFT is always performed as a complex-to-complex transform.
 
@@
 
-\subsection{Discretization of $w$} \label{sec:gridding-w}
+\subsection{Discretisation of $w$} \label{sec:gridding-w}
 \begin{figure}
@@
 \end{figure}
-While the discretization of $u$ and $v$ is similar to conventional imaging, the discretization of $w$ defines the number of $w$-layers that need to be processed.
-For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretized $w$-values, $w_A$ and $w_B$,  is less than one radian.
+While the discretisation of $u$ and $v$ is similar to conventional imaging, the discretisation of $w$ defines the number of $w$-layers that need to be processed.
+For this, one can use a rule similar to \eqref{eq:when-2d-is-valid}, and make sure that the phase difference for two subsequent discretised $w$-values, $w_A$ and $w_B$,  is less than one radian.
 This results in the constraint
@@
 \end{equation}
-This suggests that a uniform discretization in $w$ is optimal.
+This suggests that a uniform discretisation in $w$ is optimal.
 This is in contrast to \citet{wprojection-cornwell} where $\sqrt{w}$ tabulation is suggested.
@@
 With the purpose of testing new algorithms for the imaging of MWA data, we have written a new imager around the $w$-stacking algorithm.
-The imager is not specialized for the MWA, and has been successfully used for imaging VLA and GMRT data.
+The imager is not specialised for the MWA, and has been successfully used for imaging VLA and GMRT data.
 
@@
 For the latter, intrinsics are used as well.
-Because of these optimizations, minor Cleaning iterations with an image size of $3072\times3072$ are performed at a rate of hundreds per second.
-This is fast enough to make the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced point-spread function (PSF), less relevant.
-For very large images this optimization might still be useful, but we have not implemented it in \textsc{wsclean}.
+Because of these optimisations, minor Cleaning iterations with an image size of $3072\times3072$ are performed at a rate of hundreds per second.
+This is fast enough to make the Clark Clean optimisation \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced point-spread function (PSF), less relevant.
+For very large images this optimisation might still be useful, but we have not implemented it in \textsc{wsclean}.
 Because the PSF varies for imaging with non-zero $w$-values, subtracting a constant PSF in image space leads to inaccuracies.
@@
 This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}.
-\textsc{wsclean} allows cleaning individual polarizations, or can jointly deconvolve the polarizations.
-In the latter case, peaks are searched in $pp^2+2pq \overline{(pq)} + qq^2$ space, where $p$ and $q$ are the two polarizations and $\overline{pq}=qp$ is the complex conjugate of $pq$.
-After a peak has been found, the PSF is subtracted from the individual polarizations with different factors.
+\textsc{wsclean} allows cleaning individual polarisations, or can jointly deconvolve the polarisations.
+In the latter case, peaks are searched in $pp^2+2pq \overline{(pq)} + qq^2$ space, where $p$ and $q$ are the two polarisations and $\overline{pq}=qp$ is the complex conjugate of $pq$.
+After a peak has been found, the PSF is subtracted from the individual polarisations with different factors.
 
@@
 PKS J2358-6054 ($\sim100$ Jy) is visible in the centre of the southern side lobe and resolved.
-The noise level in the side lobe is about 200 mJy.
+The noise level in the side lobe is about 200 mJy/beam.
 PKS J2358-6054 has been cleaned, but some artefacts remain because no direction-dependent calibration has been performed.}
@@
 \includegraphics[width=8cm]{img/vela-zenith-projection}
-\caption{13-min MWA observation of supernova remnants Vela and Puppis A.
-Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging.
+\caption{13-min MWA observation of supernova remnants Vela and Puppis A with a centre frequency of 149~MHz.
+Left: normal projection centred on Puppis~A, right: phase-rotated to zenith to reduce $w$-values, recentred on Puppis A during imaging.
 Both Stokes-I images have been made with \textsc{wsclean} using the Cotton-Schwab clean algorithm.
@@
 
-Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
+Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\rightarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
 \begin{align}\notag
@@
 \end{align}
-In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging.
+In words, recentring an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging.
 This essentially is a method of implementing the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}.
 However, instead of creating warped images by performing the FFT over a tilted plane in $uvw$-space, our method transforms planes with constant $w$-values.
-The benefit of our method is that the resulting image has a circular synthesized beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning.
+The benefit of our method is that the resulting image has a circular synthesised beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning.
 Regridding a recentred image is also more straightforward compared to regridding a warped snapshot, because in the latter case there is no analytic solution to the coordinate conversion \citep{perley-noncoplanar-arrays}.
@@
 By doing the inverse corrections during prediction, a recentred visibility set can be cleaned with Cotton-Schwab iterations similar to a non-recentred set.
-The PSF used during minor cleaning iterations is created by multiplying the weights with the recentering corrections, such that the PSF represents a source in the middle of the image.
+The PSF used during minor cleaning iterations is created by multiplying the weights with the recentring corrections, such that the PSF represents a source in the middle of the image.
 
@@
 Common viewers such as \textsc{kvis} \citep{karma-1996} and \textsc{ds9} support such FITS files and display their coordinates correctly.
-Unlike warped snapshots, recentred images do not need the generalized-SIN-projection keywords \texttt{PV2\_1} and \texttt{PV2\_2}.
+Unlike warped snapshots, recentred images do not need the generalised-SIN-projection keywords \texttt{PV2\_1} and \texttt{PV2\_2}.
 
 An example of the difference in projection between non-recentred and recentred imaging is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13-min MWA observation imaged with \textsc{wsclean}.
-The processing steps performed for this image were: i) preprocessing of the observation with the Cotter preprocessing pipeline, which includes time averaging and RFI flagging with the \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibration using Hydra~A without direction dependence using a custom implementation of the RTS full-polarization calibration algorithm \citep{rts-mwa}; and iii) imaging of the seven 112 s intervals separately.
+The processing steps performed for this image were: i) preprocessing of the observation with the Cotter preprocessing pipeline, which includes time averaging and RFI flagging with the \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}; ii) calibration using Hydra~A without direction dependence using a custom implementation of the RTS full-polarisation calibration algorithm \citep{rts-mwa}; and iii) imaging of the seven 112 s intervals separately.
 Cleaning an image that is in a different projection yields slightly different results, but qualitatively the two images are clearly of equal accuracy.
@@
 This complicates gridding during imaging, because the anti-aliasing kernel needs to have a pass-band shape instead of a low-pass shape.
-Therefore, we chose to implement the former method.
+Therefore, we chose to implement the former method of recentring the phase centre before imaging.
 
@@
 \includegraphics[width=6cm]{img/pulsar-stokesi}\hspace{1cm}\includegraphics[width=6cm]{img/pulsar-stokesv}
-\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarization) images of PSR J0437-4715, demonstrating the full-polarization beam correction capability for MWA observations with \textsc{wsclean}.
-This observation was performed in drift-scan mode.
-The pulsar (centre of image) displays 17\% circular polarization.
-Due to inaccuracies in the current beam model, other (unpolarized) sources show $\sim$1\% leakage into Stokes~V.}
+\caption{Stokes I (left panel, total power) and Stokes V (right panel, circular polarisation) images of PSR J0437-4715, demonstrating the full-polarisation beam correction capability for MWA observations with \textsc{wsclean}.
+This observation was performed in drift-scan mode with a centre frequency of 154~MHz.
+The pulsar (centre of image) displays 17\% circular polarisation.
+Due to inaccuracies in the current beam model, other (unpolarised) sources show $\sim$1\% leakage into Stokes~V.}
 \label{fig:pulsar-stokes-iv}
@@
 where each term is a $2\times2$ complex matrix, with $J$ the voltage beam and $^*$ denoting the conjugate transpose.
-Because $J$ is complex and non-diagonal, all four combinations of the cross-correlated polarizations $p$ and $q$\footnote{The instrumental polarizations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarizations as such.
-However, this is somewhat confusing because they are not necessarily orthogonal.} including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters.
+Because $J$ is complex and non-diagonal, all four combinations of the cross-correlated polarisations\footnote{The instrumental polarisations of the MWA are informally often referred to as $X$ and $Y$, mainly because many software treats the two polarisations as such.
+However, this is somewhat confusing because they are not necessarily orthogonal.} $p$ and $q$ including the imaginary part of the $pq$ and $qp$ are required to calculate any of the Stokes parameters.
 The $pp$ and $qq$ have no imaginary part due to the $uv$-symmetry.
@@
 The method is demonstrated in Fig.~\ref{fig:pulsar-stokes-iv}, which displays a detection of pulsar J0437-4715 in Stokes V.
-The image was made with an initial pipeline for the MWA radio-sky monitor project (PI: Bell) that uses the MWA to search for transient and variable sources.
+The image was made with an initial pipeline for the MWA radio-sky monitor project that uses the MWA to search for transient and variable sources.
 The pipeline includes \textsc{wsclean} to make the Stokes-parameter images.
@@
 \subsection{Decreasing inversion resolution} \label{sec:decreasing-inversion-resolution}
-Typically, for cleaning it is desired that images have a pixel size (side length) at least five times smaller than the synthesized-beam width.
+Typically, for cleaning it is desired that images have a pixel size (side length) at least five times smaller than the synthesised-beam width.
 This improves the accuracy of the clean algorithm, because the positions of image maxima will be closer to the actual source positions.
@@
 
-The spatial frequencies in the output image are band-limited by the synthesized beam.
+The spatial frequencies in the output image are band-limited by the synthesised beam.
 Therefore, it is not required to perform the inversion at higher resolution.
@@
 \end{equation}
-where $N_\textrm{Nyquist}$ is the resolution in pixels used during inversion, $N_\textrm{pix}$ is the requested number of pixels of the image along one side, $S_\textrm{pix}$ is the requested angular pixel scale and $S_\textrm{synth beam}$ is the minimum angular size of the synthesized beam.
+where $N_\textrm{Nyquist}$ is the resolution in pixels used during inversion, $N_\textrm{pix}$ is the requested number of pixels of the image along one side, $S_\textrm{pix}$ is the requested angular pixel scale and $S_\textrm{synth beam}$ is the minimum angular size of the synthesised beam.
 Before cleaning, the low-resolution image is interpolated with a procedure consisting of: i) a low-resolution FFT; ii) zero padding; and iii) a high-resolution inverse FFT.
@@
 
-Recreating Fig.~\ref{fig:vela-projection-example} using both the snapshot method and this optimization lowers the computational cost from 41~min to 24~min.
+Recreating Fig.~\ref{fig:vela-projection-example} using both the snapshot method and this optimisation lowers the computational cost from 41~min to 24~min.
 Of the 24~min, 77\% is spent on cleaning approximately 100,000 components.
-The outputs with and without rescaling do not visibly differ, but have an RMS difference of 20 mJy, while the total RMS is 227 mJy.
+The outputs with and without rescaling do not visibly differ, but have an RMS difference of 20 mJy/beam, while the total RMS is 227 mJy/beam.
 Because the difference is mostly noise like, the difference could be caused by the non-linear behaviour of clean.
@@
 Max $w$-value for phase centre & 172 $\lambda$ (283 m) \\
-Number of polarizations in set & 4 \\
-Imaged polarization & $pp$ ($\sim$XX) \\
+Number of polarisations in set & 4 \\
+Imaged polarisation & $pp$ ($\sim$XX) \\
 Imaging mode & multi-frequency synthesis \\
@@
 Source flux standard error & 1.31\% & & 1.34\% \\
-RMS in residual image & 0.94 mJy & --- & 1.90 mJy\\
+RMS in residual image & 0.94 mJy/b & --- & 1.90 mJy/b\\
 Computational time & 8.5 min & & 19.3 min\\
@@
 Source flux standard error & 1.39\% & & 2.08\% \\
-RMS in residual image & 0.94 mJy & --- & 0.94 mJy \\
+RMS in residual image & 0.94 mJy/b & --- & 0.94 mJy/b \\
 Computational time & 10.3 min & & 19.6 min \\
@@
 Source flux standard error & 1.75\% & 1.40\% & 2.41 \%\\
-RMS in residual image & 0.90 mJy & 1.03 mJy & 1.07 mJy \\
+RMS in residual image & 0.90 mJy/b & 1.03 mJy/b & 1.07 mJy/b \\
 Computational time & 15.3 min & 6.6 min & 178.2 min \\
@@
 The full field contains 100 simulated sources over 20\degree.
-Sources in the image produced with \textsc{casa} are slightly less accurately subtracted, leading to a residual noise level of 1.07 mJy and some visible artefacts, whereas \textsc{wsclean} reaches 0.90 mJy RMS noise.
-Since the effect is stronger away from the phase centre, it is likely that this is caused by the finite size of the $w$-kernels used in $w$-projection, which leads to inaccuracies.}
+Sources in the image produced with \textsc{casa} are slightly less accurately subtracted, leading to a residual noise level of 1.07 mJy/beam and some visible artefacts, whereas \textsc{wsclean} reaches 0.90 mJy/beam RMS noise.
+Since the effect is stronger away from the phase centre, it is likely that this is caused by the finite size of the $w$-kernels used in $w$-projection, which leads to inaccuracies.
+The ring-shaped residuals are caused by imperfect deconvolution.}
 \label{fig:residuals}
@@
 
-To assess the accuracy of \textsc{wsclean} and \textsc{casa}, we simulate a MWA observation with 100 sources of 1~Jy in a 20\degree diameter area, without adding system noise.
+To assess the accuracy of \textsc{wsclean} and \textsc{casa}'s clean task, we simulate a MWA observation with 100 sources of 1~Jy in a 20\degree diameter area, without adding system noise.
 A unitary primary beam is assumed.
 We image the simulated set with \textsc{wsclean} and \textsc{casa} using Cotton-Schwab cleaning to a threshold of 10~mJy.
-The two imagers calculate slightly different restoring (synthesized) beams, hence to avoid bias the restoring beams are fixed.
+The two imagers calculate slightly different restoring (synthesised) beams, hence to avoid bias the restoring beams are fixed.
 Other imaging parameters are given in Table~\ref{tbl:default-parameters}.
@@
 The reprojection technique needs on average fewer $w$-layers to reach the same level of accuracy.
-We have performed tests with and without the optimization of \S\ref{sec:decreasing-inversion-resolution}.
+We have performed tests with and without the optimisation of \S\ref{sec:decreasing-inversion-resolution}.
 They yield identical numbers.
@@
 \textsc{wsclean} is faster in all tested cases.
-Both imagers perform five major iterations for these results, and the inversions and predictions are dominating the computing time.
+Both imagers perform five major iterations for these results, and the inversions and predictions dominate the computing time.
 We will look more closely at the differences in performance in the next section.
@@
 The variation in duration between runs is typically a few seconds.
-In each benchmark, the wall-clock time is measured that is required to produce the synthesized point-spread function and the image itself.
-No cleaning or prediction is performed, and the optimization of \S\ref{sec:decreasing-inversion-resolution} is not used.
+In each benchmark, the wall-clock time is measured that is required to produce the synthesised point-spread function and the image itself.
+No cleaning or prediction is performed, and the optimisation of \S\ref{sec:decreasing-inversion-resolution} is not used.
 The results are given in Fig.~\ref{fig:timings}.
@@
 This implies that gridding or reading dominates the cost.
-In that situation, the $w$-stacking implementation is 7.9 times faster at $\textrm{ZA}=10\degree$ and 2.6 times faster at $\textrm{ZA}=0\degree$.
+In that situation, the $w$-stacking implementation is 7.9 times faster than \textsc{casa} at $\textrm{ZA}=10\degree$ and 2.6 times faster at $\textrm{ZA}=0\degree$.
 With small data volumes, the FFTs start to dominate the cost, visible in Fig.~\ref{fig:timing-nsamples} as a flattening towards the left.
@@
 At zenith, the two methods are expected to perform almost identically.
-The factor of 2-3 difference in Fig.~\ref{fig:timing-nsamples} could be due to different choices in optimizations.
+The factor of 2-3 difference in Fig.~\ref{fig:timing-nsamples} could be due to different choices in optimisations.
 
 In Panel~\ref{fig:timing-zenith-angle}, the computational cost as a function of ZA is plotted.
-It shows that, as expected from \S\ref{sec:time-complexity}, the $w$-projection algorithm is more affected by the increased $w$-values, leading to differences of more than an order of magnitude at ZAs of $\gtrapprox 20$\degree.
+It shows that, as expected from \S\ref{sec:time-complexity}, compared to $w$-stacking the $w$-projection algorithm is more affected by the increased $w$-values, leading to differences of more than an order of magnitude at ZAs of $\gtrapprox 20$\degree.
 Additionally, at higher ZAs, the $w$-kernels become too large to be able to make images of 3072 pixels or larger.
-Using the recentering technique with \textsc{wsclean} to decrease the $w$-values makes the computational cost approximately constant.
-However, the additional time required to rephase and regrid the measurement set makes it only worthwhile for images $\ge3072$ pixels and $\textrm{ZA}\ge15\degree$.
+Using the recentring technique with \textsc{wsclean} to decrease the $w$-values makes the computational cost approximately constant.
+However, the additional time required to rephase and regrid the measurement set makes recentring only worthwhile for images $\ge3072$ pixels and $\textrm{ZA}\ge15\degree$.
 This is dependent on the speed of rotating the phase centre and regridding.
-Our program to change the phase centre is currently not multi-threaded, so its performance could be improved.
+Our code to change the phase centre is currently not multi-threaded, so its performance can be improved.
 Furthermore, the relative cost of changing the phase centre is lower when performing multiple major iterations.
@@
 Imaging the MWA primary beam requires approximately an image size of 3072 pixels for cleaning.
-If the small-inversion optimization of \S\ref{sec:decreasing-inversion-resolution} is used, this decreases to about 1500.
+If the small-inversion optimisation of \S\ref{sec:decreasing-inversion-resolution} is used, this decreases to about 1500.
 At ZA=10\degree, this saves about a factor of 2 in computing cost.
-The benefit of the optimization increases with resolution.
+The benefit of the optimisation increases with resolution.
 At an image size of 10,000 pixels, cost is decreased by an order of magnitude.
@@
 The cost of spectral imaging, i.e.
-imaging multiple frequencies, is the cost of making an image from fewer visibilities times the number of desired frequencies.
+imaging multiple frequencies, is the cost of making an image from fewer visibilities multiplied by the number of desired frequencies.
 Panel~\ref{fig:timing-channels} shows performance versus the spectral output resolution.
@@
 This plot was created by varying the size of a pixel in the output image, such that the number of pixels in the image did not change.
-The FOV is calculated as angle between the left- and right-most pixel in the image.
+The FOV is calculated as the angle subtended between the left- and right-most pixel in the image.
 \textsc{casa} is not able to make off-zenith images larger than $\sim$30\degree, and its imaging cost increases much more rapidly compared to the cost of \textsc{wsclean}, which implies that gridding is the major cost in this scenario.
-The image recentering technique is clearly beneficial for FOVs larger than $\sim$30\degree, and can even make a difference of an order of magnitude at very large FOVs of $\sim$90\degree.
+The image recentring technique is clearly beneficial for FOVs larger than $\sim$30\degree, and can even make a difference of an order of magnitude at very large FOVs of $\sim$90\degree.
 
-%It is important to point out that all these measurement were performed by imaging a single polarization.
-Making full-polarization images with \textsc{wsclean} currently requires four runs.
-Imaging four Stokes parameters with \textsc{casa} costs slightly less than four times the time of imaging a single polarization.
+%It is important to point out that all these measurement were performed by imaging a single polarisation.
+Making full-polarisation images with \textsc{wsclean} currently requires four runs.
+Imaging four Stokes parameters with \textsc{casa} costs slightly less than four times the time of imaging a single polarisation.
 With the settings of Table~\ref{tbl:default-parameters}, the difference is a factor of $3.5$, and imaging XX and YY together costs $1.9$ times imaging just XX.
@@
 Lanczos interpolation) can be expensive.
-Also, our current implementation of the phase-changing program does take a non-negligible time, but this implementation is not optimized and can in theory be implemented in a preprocessing pipeline or on-the-fly during imaging.
+Also, our current implementation of the phase-changing program does take a non-negligible time, but this implementation is not optimised and can in theory be implemented in a preprocessing pipeline or on-the-fly during imaging.
 
-The function $t_\textrm{snapshot}$ can be minimized to find the optimal snapshot duration.
+The function $t_\textrm{snapshot}$ can be minimised to find the optimal snapshot duration.
 For \textsc{wsclean} with MWA parameters, we find that this function decreases but no minimum is reached for $\Delta \tau_\textrm{snapshot}<24$~h.
@@
 \begin{table*}
-\caption{Predicted computational costs for configurations listed in Table~\ref{tbl:computational-cost-configurations}, based on multi-frequency synthesis with five major iterations observing for 1~h at ZA=20\degree with 1 polarization.
+\caption{Predicted computational costs for configurations listed in Table~\ref{tbl:computational-cost-configurations}, based on multi-frequency synthesis with five major iterations observing for 1~h at ZA=20\degree with 1 polarisation.
 Predictions are for $w$-projection with \textsc{casa}; $w$-stacking with \textsc{wsclean}; $w$-stacked snapshots with \textsc{wsclean} using optimal snapshot duration; and a hybrid between $w$-projection and $w$-stacking using \textsc{casa} with optimal $\Delta w$.
@@
 We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multi-frequency Snapshot Sky Survey (MSSS)\footnote{See~\href{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}{www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam--ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \href{http://www.aartfaac.org}{www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The Meer Karoo Array Telescope (MeerKAT); and the low-frequency Phase~1 aperture arrays of the Square-Kilometre Array (SKA), with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}.
-The configurations are summarized in Table~\ref{tbl:computational-cost-configurations}.
+The configurations are summarised in Table~\ref{tbl:computational-cost-configurations}.
 In multi-beam configurations such as the EMU and LOFAR configurations, we assume each beam is imaged separately, i.e., the FOV of a single beam is used and the computing cost is multiplied with the number of beams.
@@
 
-The results are summarized in Table~\ref{tbl:computational-cost-predictions}.
+The results are summarised in Table~\ref{tbl:computational-cost-predictions}.
 \textsc{wsclean} is in most situations predicted to be 2--3 times faster than \textsc{casa}.
@@
 This can lead to performance improvements of a factor of $3$ for the MWA, but only in the most expensive imaging configurations that are less commonly used.
-Considering the extra regridding step required, which complicates issues such as calculating the integrated beam shape, recentering snapshots is worthwhile for the MWA only at very low elevations or with large fields of view.
+Considering the extra regridding step required, which complicates issues such as calculating the integrated beam shape, recentring snapshots is worthwhile for the MWA only at very low elevations or with large fields of view.
 Extrapolation of the computing cost predicts that this holds for most arrays, including SKA low.
 A hybrid between $w$-stacking and $w$-projection does not improve performance over a pure $w$-stacking implementation, but does improve a pure $w$-projection implementation significantly.
-Our optimization of lowering the image resolution during inversion and prediction increases performance by a factor of 2--10 and has no noticeable effect on the accuracy.
+Our optimisation of lowering the image resolution during inversion and prediction increases performance by a factor of 2--10 and has no noticeable effect on the accuracy.
 
@@
 
-So far, we have corrected the (full-polarization) beam in image space, which is possible because of the homogeneity of the MWA tiles.
+So far, we have corrected the (full-polarisation) beam in image space, which is possible because of the homogeneity of the MWA tiles.
 This is computationally cheap for our current snapshot time of 112s, but this might be infeasible when it is required to calculate the beam on very short time scales.
@@
 We plan to implement some form of multi-scale cleaning in \textsc{wsclean}.
-It is likely that additional optimizations need to be made to be able to do this with acceptable performance.
+It is likely that additional optimisations need to be made to be able to do this with acceptable performance.
 Combining multi-scale with wide-band deconvolution techniques is a possible further improvement \citep{rau-msmfs-2011}.

### 2014-03-31 articles/2014/Article-WSClean/wsclean.tex

@@
 In practice, the final scaling will be different from $\left(w_{\max} - w_{\min}\right)/\sqrt{1-l^2-m^2}$ suggested by Eq.~\eqref{eq:wstacking}, because the individual $w$-layers will not be completely filled with samples.
-Therefore, the number of added samples and their weights have to be taken into account.
 Therefore, each pixel is divided by the weighted number of samples.
 Additionally, it might be required to divide out the effect of a possible convolution kernel, the primary beam and correct for other direction-dependent effects.
-In XX, YY, LL or RR polarisations, a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds.
+In equally-polarized baselines (e.g., XX, YY, LL or RR), a correlated baseline is the complex conjugate of the reversed baseline, and the relation $V(u,v,w)=\overline{V(-u,-v,-w)}$ holds.
 The right-hand side of Eq.\eqref{eq:wstacking} with only positive $w$-value samples then becomes the complex conjugate of the one with only negative $w$-value samples.

### 2014-04-01 articles/2014/Article-WSClean/wsclean.tex
Version as submitted\!

@@
 
-\ifpdf\pdfinfo{/Title      (WSClean: an implementation of a fast, generic wide-field imager)
+\ifpdf\pdfinfo{/Title      (WSClean: an implementation of a fast, generic wide-field imager for radio astronomy)
                /Author     (A. R. Offringa et al.)
@@
 
-%It is important to point out that all these measurement were performed by imaging a single polarisation.
-Making full-polarisation images with \textsc{wsclean} currently requires four runs.
-Imaging four Stokes parameters with \textsc{casa} costs slightly less than four times the time of imaging a single polarisation.
-With the settings of Table~\ref{tbl:default-parameters}, the difference is a factor of $3.5$, and imaging XX and YY together costs $1.9$ times imaging just XX.
-
 \subsection{Derivation of computing cost formulae}
@@
  MeerKAT  & 0.2  &  1\degree  & 1    &  64 &   6'' &  8 & 1   & 0.5 & 750  & 50 \\
-% SKA1 AA inner&2 &  5\degree  & 1    & 433 &  15'' &0.6& 0.5 &  10 & 250 & 1 \\
  SKA1 AA core&2  &  5\degree  & 1    & 866 &    3' &  3& 5 & 10 & 250 & 1 \\
@@
  MeerKAT  &   10.8 d &  6.2 d &   6.2 d &  10.8 d   &  21 TFLOPS & $6.8 \times 10^2$ \\
-% SKA1 AA inner&392 d &  228 d &   228 d &   392 d & & $6.8 \times 10^2$   \\
  SKA1 AA core&1581 d &  911 d &   911 d &  1570.8   & 3.0 PFLOPS & $6.8 \times 10^2$ \\

### 2014-05-26 articles/2014/Article-WSClean/wsclean.tex
Implemented suggestions made by referee

@@
 
-%\newcommand{\editmark}[1]{#1}
+\newcommand{\editmark}[1]{#1}
 %\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
+
 \newcommand{\degree}{\ensuremath{^{\circ}}\xspace}
@@
 
-\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{E-mail: \url{andre.offringa@anu.edu.au}},
+\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\editmark{Corresponding author.} E-mail: \url{andre.offringa@anu.edu.au}},
 B.~McKinley\ANU$^,$\CAASTRO,
@@
 J.~D.~Hughes\ANU, 
-J.~Rhee\ANU, \newauthor
+J.~Rhee\ANU$^,$\CAASTRO, \newauthor
 T.~Murphy\USydney$^,$\CAASTRO,
@@
 S.~M.~Ord\Curtin$^,$\CAASTRO,
+\editmark{B.~Pindor\UMelbourne,}
 P.~Procopio\UMelbourne,
 T.~Prabu\RRI,
-J.~Riding\UMelbourne, 
-A.~E.~E.~Rogers\Haystack, \newauthor
+J.~Riding\UMelbourne,  \newauthor
+%A.~E.~E.~Rogers\Haystack,
 D.~A.~Roshi\NRAO, %\footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
@@
 
-\date{Accepted TODO. Received TODO; in original form TODO}
+\date{Accepted TODO. Received TODO; in original form 01-04-2014}
 \pagerange{\pageref{firstpage}--\pageref{lastpage}}
@@
 This is especially true for wide-field telescopes.
-One way to handle DDEs is by using the $a$-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
+One way to \editmark{correct for known} DDEs is by using the $a$-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
 One particular DDE is the effect of the ionosphere.
@@
 The \textsc{awimager} \citep{awimager-2013} has been written to perform these corrections, and uses a hybrid of a-projection, $w$-projection and $w$-stacking.
+\editmark{The $a$-projection technique can only be applied directly for deterministic effects such as the correction of the primary beam.
+Effects like the ionosphere require separate calibration or estimation before $a$-projection can be applied.}
 
@@
 We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''.
-Our new imaging implementation, which we believe to be the fastest generic wide-field imager available, is publicly released\footnote{The \textsc{wsclean} source code can be found at:\\\href{http://sourceforge.net/p/wsclean}{http://sourceforge.net/p/wsclean}}.
+Our new imaging implementation, which \editmark{in our experience is anywhere from 2 to 12 times faster than the CASA $w$-projection imager}, is publicly released\footnote{The \textsc{wsclean} source code can be found at:\\\href{http://sourceforge.net/p/wsclean}{http://sourceforge.net/p/wsclean}}.
 
@@
 In this section, we will describe the $w$-stacking algorithm from a mathematical point of view.
+\editmark{Instead of applying a convolution in $uv$-space, the $w$-stacking method grids visibilities on different $w$-layers and performs the $w$-corrections after the inverse Fourier transforms \citep{ska-memo-regridding-2011}.}
 
@@
 
-Using too few $w$-layers results in decorrelation of sources far from the phase centre in the longer baselines.
-Additionally, $w$-aliasing errors can cause ghost sources in the image.
+\editmark{To grid the visibilities, the $w$-values are rounded to the $w$-value of the nearest $w$-layer.
+This discretisation can cause noticeable aliasing when using too few $w$-layers, and results in decorrelation of sources far from the phase centre in the longer baselines.
+Additionally, this $w$-aliasing can cause ghost sources to appear} in the image.
 An (extreme) example of this effect is shown in Fig.~\ref{fig:aliasing-example}.
@@
 
+\subsection{\editmark{$W$-snapshot imaging}} \label{sec:snapshot-imaging-theory}
+\editmark{$W$-snapshot imaging is a technique that combines warped-snapshot imaging with a $w$-correcting technique, such as $w$-projection or $w$-stacking \citep{widefield-imaging-ska-cornwell}.
+In the warped-snapshot imaging technique, the $w$-term is neglected, which results in an image with distorted coordinates \citep{perley-noncoplanar-arrays,mwa-interferometric-imaging}.
+Additionally, if the positions of the array elements are not perfectly planar or multiple timesteps are integrated to the same grid, visibilities will decorrelate and this can cause imaging artefacts.
+The original $w$-snapshot algorithm as described in \citet{widefield-imaging-ska-cornwell} corrects such artefacts by gridding visibilities on a tilted best-fit plane in $uvw$-space, and performs $w$-corrections towards that plane using $w$-projection or $w$-stacking.}
+
+\editmark{We have looked into implementing snapshot imaging with $w$-corrections in a slightly different way.
+Instead of performing transforms of tilted planes, our method consists of phase-rotating the visibilities such that the phase centre of the observation is towards the zenith direction, i.e., the direction with minimal $w$-values.
+During imaging, the $w$-layers are recentred from zenith to the direction of interest by phase-shifting the visibilities, thereby translating the image over the tangent plane.
+This method results in $w$-corrected snapshots which need to be regridded before further integration, similar to the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}.
+However, instead of creating warped images by performing the FFT over a tilted plane in $uvw$-space, this method transforms planes with constant $w$-values and produces images in zenith projection.}
+
+\editmark{An} image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\rightarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
+\begin{align}\notag
+\frac{I'(\hat{l}, \hat{m})\left(w_{\max} - w_{\min}\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \times \\
+\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
+\end{align}
+In words, recentring an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging.
+\editmark{By doing the inverse corrections during prediction, a recentred visibility set can be cleaned with Cotton-Schwab iterations similar to a non-recentred set.}
+
+\editmark{Our main reason for developing this method is that it is easier to implement, because no changes are required to the performance-critical gridding step.
+Another} benefit of our method is that the resulting image has a circular synthesised beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning.
+Regridding a recentred image is also more straightforward compared to regridding a warped snapshot, because in the latter case there is no analytic solution to the coordinate conversion \citep{perley-noncoplanar-arrays}.
+A benefit of warped snapshots is that the $w$-term errors are zero at image centre and get worse with the distance from image centre.
+With our approach, the effect gets worse with the distance from zenith.
+Our technique has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on ZA, FOV and the non-coplanarity.
+
+Another method to shift an image is by using the periodicity of the FFT function.
+Since sources outside the FOV will be aliased back into the field, the image can be transformed to have the alias in the centre.
+This complicates gridding during imaging, because the anti-aliasing kernel needs to have a pass-band shape instead of a low-pass shape.
+Therefore, we chose to implement the former method of recentring the phase centre before imaging.
+
 \section{The \textsc{wsclean} imager implementation} \label{sec:implementation}
@@
 \textsc{wsclean} allows cleaning individual polarisations, or can jointly deconvolve the polarisations.
-In the latter case, peaks are searched in $pp^2+2pq \overline{(pq)} + qq^2$ space, where $p$ and $q$ are the two polarisations and $\overline{pq}=qp$ is the complex conjugate of $pq$.
+\editmark{In the latter case, peak finding can be performed in the sum of squared Stokes parameters, $I^2+Q^2+U^2+V^2$, or in} $pp^2+2pq \overline{(pq)} + qq^2$ space, where $p$ and $q$ are the two polarisations and $\overline{pq}=qp$ is the complex conjugate of $pq$.
 After a peak has been found, the PSF is subtracted from the individual polarisations with different factors.
@@
 Full-sky imaging can also be useful for self-calibration or deconvolution.
-For example, self-calibration using full-sky clean components has been found to give good results in imaging the resolved FR-II radio source Fornax~A (McKinley et al., in prep.).
+For example, self-calibration using full-sky clean components has been found to give good results in imaging the resolved FR-II radio source Fornax~A (McKinley et al., \editmark{submitted}).
 An example of a full-sky MWA image is shown in Fig.~\ref{fig:full-sky-example}.
@@
 
-\subsection{$w$-snapshots} \label{sec:snapshot-imaging}
+\subsection{Implementation of $w$-snapshot imaging} \label{sec:snapshot-imaging-implementation}
 \begin{figure*}
@@
 However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.
-
-Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\rightarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
-\begin{align}\notag
-\frac{I'(\hat{l}, \hat{m})\left(w_{\max} - w_{\min}\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_{\min}}^{w_{\max}} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \times \\
-\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
-\end{align}
-In words, recentring an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging.
-This essentially is a method of implementing the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}.
-However, instead of creating warped images by performing the FFT over a tilted plane in $uvw$-space, our method transforms planes with constant $w$-values.
-The benefit of our method is that the resulting image has a circular synthesised beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning.
-Regridding a recentred image is also more straightforward compared to regridding a warped snapshot, because in the latter case there is no analytic solution to the coordinate conversion \citep{perley-noncoplanar-arrays}.
-A benefit of warped snapshots is that the $w$-term errors are zero at image centre and get worse with the distance from image centre.
-With our approach, the effect gets worse with the distance from zenith.
-Our technique has a similar computational cost compared to the $w$-snapshot algorithm, although the required number of $w$-layers has a different dependency on ZA, FOV and the non-coplanarity.
-
-By doing the inverse corrections during prediction, a recentred visibility set can be cleaned with Cotton-Schwab iterations similar to a non-recentred set.
-The PSF used during minor cleaning iterations is created by multiplying the weights with the recentring corrections, such that the PSF represents a source in the middle of the image.
+\editmark{For these cases, the recentring technique described in \S\ref{sec:snapshot-imaging-theory} was implemented in \textsc{wsclean}.
+This allows making smaller images with minimal $w$-values that are recentred on the direction of interest.
+The implementation is generic and supports interferometric data from any telescope.}
 
@@
 Unlike warped snapshots, recentred images do not need the generalised-SIN-projection keywords \texttt{PV2\_1} and \texttt{PV2\_2}.
+
+\editmark{Our implementation supports cleaning of recentred images in the same modes that normal images can be cleaned.} The PSF used during minor cleaning iterations is created by multiplying the weights with the recentring corrections, such that the PSF represents a source in the middle of the image.
 
@@
 
-Another method to shift an image is by using the periodicity of the FFT function.
-Since sources outside the FOV will be aliased back into the field, the image can be transformed to have the alias in the centre.
-This complicates gridding during imaging, because the anti-aliasing kernel needs to have a pass-band shape instead of a low-pass shape.
-Therefore, we chose to implement the former method of recentring the phase centre before imaging.
-
-\subsection{Beam correction}
+\subsection{Beam correction \editmark{for the MWA}}
 \begin{figure*}
@@
 
-Our image-based beam-correction method avoids expensive beam-correcting kernels during gridding, but only works because all tiles have the same beam.
+Our image-based beam-correction method avoids expensive beam-correcting kernels during gridding, but \editmark{requires snapshot imaging, because the MWA beam changes over time because of Earth rotation, and} only works because all tiles have the same beam.
 To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm excessively unless an a-projection kernel is used.
@@
 The prolate spheroidal wave function (PSWF) is generally considered to be the optimal windowing function for gridding \citep{fourier-kernel-selection-1991}.
-\textsc{casa} convolves samples with a PSWF of seven pixels total width during gridding.
+\textsc{casa}'s gridder implementation convolves samples with a PSWF of seven pixels total width during gridding.
 When using a variable kernel size, a PSWF is quite complicated and computationally expensive to calculate.
@@
 \includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
-\caption{Effect of anti-aliasing kernel size on performance.}
+\caption{\editmark{Gridding kernel size plotted against imaging time, using common MWA settings.}}
 \label{fig:timing-kernelsize}
@@
 
-\subsection{Decreasing inversion resolution} \label{sec:decreasing-inversion-resolution}
+\subsection{\editmark{Lowering the resolution of inversion}} \label{sec:decreasing-inversion-resolution}
 Typically, for cleaning it is desired that images have a pixel size (side length) at least five times smaller than the synthesised-beam width.
@@
 The spatial frequencies in the output image are band-limited by the synthesised beam.
-Therefore, it is not required to perform the inversion at higher resolution.
-As long as the $uv$-plane is sampled with at least the Nyquist frequency, a high resolution image can be perfectly reconstructed from an inversion at lower resolution.
-After an enlarged image has been cleaned, a model is created at the same (high) resolution.
+Therefore, as long as the $uv$-plane is sampled with at least the Nyquist frequency, a high-resolution image can be perfectly reconstructed from an inversion at lower resolution.
+\editmark{Cleaning can be performed on the high-resolution image that is reconstructed from the low-resolution image}.
+After \editmark{the high-resolution} image has been cleaned, a model is created at the same (high) resolution.
 Because the model consists of delta functions, the spatial frequencies in the model are not band-limited.
-Consequently, decimating the model image will remove information from the model.
-However, only the low spatial frequencies of this image will be used in the prediction.
-Therefore, inversion at higher resolution is not required: as long as the decimation does not modify the low-frequency components, the output of the prediction step will not change.
-This requires low-pass filtering of the model before decimation.
+Consequently, \editmark{lowering the resolution of the model image will remove information from the model.
+However, only the low spatial frequencies of this image will be used in the prediction, because in $uv$-space only visibilities up to the corresponding maximum baseline length will be sampled.
+Therefore, as long as lowering the resolution does} not modify the low-frequency components, the output of the prediction step will not change.
 
@@
 Of the 24~min, 77\% is spent on cleaning approximately 100,000 components.
-The outputs with and without rescaling do not visibly differ, but have an RMS difference of 20 mJy/beam, while the total RMS is 227 mJy/beam.
+The outputs with and without \editmark{lowering the inversion and prediction resolution} do not visibly differ, but \editmark{the difference between the residual images has an RMS of 18 mJy/beam.
+This can be compared to a noise level of 67 mJy/beam in the original residual image and a peak flux in the restored image of 10.3 Jy/beam}.
 Because the difference is mostly noise like, the difference could be caused by the non-linear behaviour of clean.
@@
 Extra noise can be seen in the images, as shown in Fig.~\ref{fig:residuals}.
-An image resulting from the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging}, is also made and analysed.
+An image resulting from the technique of recentring a zenith phase-centred visibility set, as described in Sect.~\ref{sec:snapshot-imaging-theory}, is also made and analysed.
 As can be seen in Table~\ref{tbl:accuracy-measurements}, the source fluxes in the recentred image have smaller errors compared to the normal projection, but the residual noise is higher.
-The reprojection technique needs on average fewer $w$-layers to reach the same level of accuracy.
-We have performed tests with and without the optimisation of \S\ref{sec:decreasing-inversion-resolution}.
+The recentring technique needs on average fewer $w$-layers to reach the same level of accuracy.
+We have performed the tests with and without the optimisation of \S\ref{sec:decreasing-inversion-resolution}.
 They yield identical numbers.
@@
 Error bars show 5$\sigma$ level.
-Unfinished lines indicate the imager could not run the specific configuration successfully.}\label{fig:timings}
+Unfinished lines indicate the imager could not run the specific configuration successfully.
+\editmark{The label ``R+\textsc{wsclean}'' refers to using WSClean with the recentring technique described in \S\ref{sec:snapshot-imaging-theory}.}}\label{fig:timings}
 \end{figure*}
@@
 Imaging the MWA primary beam requires approximately an image size of 3072 pixels for cleaning.
-If the small-inversion optimisation of \S\ref{sec:decreasing-inversion-resolution} is used, this decreases to about 1500.
+If the small-inversion optimisation of \S\ref{sec:decreasing-inversion-resolution} is used, \editmark{the inversion can be performed at an image size of 1500 pixels}.
 At ZA=10\degree, this saves about a factor of 2 in computing cost.
-The benefit of the optimisation increases with resolution.
-At an image size of 10,000 pixels, cost is decreased by an order of magnitude.
+The benefit of the optimisation increases with resolution: At an image size of 10,000 pixels, cost is decreased by an order of magnitude.
 
@@
 \subsection{Optimal snapshot duration}
-Snapshot imaging, such as the recentring technique described in \S\ref{sec:snapshot-imaging}, can be implemented with either $w$-projection or $w$-stacking.
+Snapshot imaging, such as the recentring technique described in \S\ref{sec:snapshot-imaging-theory}, can be implemented with either $w$-projection or $w$-stacking.
 Using the derived formulae, we can estimate the cost of making snapshots with both techniques.
@@
 The image size is set to two times the FOV width divided by the resolution, as described by Eq.~\ref{eq:nyquist-resolution}.
-The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).
+\editmark{Therefore, this assumes that the optimisation of \S\ref{sec:decreasing-inversion-resolution} to inverse at lower resolution is used.} The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).
 
@@
 \textsc{wsclean} has the largest benefit on the full LOFAR, MWA and the full SKA, where \textsc{wsclean} is 7, 8 and 12 times faster, respectively.
-The $w$-snapshot method does not improve much over normal \textsc{wsclean} operation in any of the cases.
-The $w$-snapshot might become more valuable at higher ZAs.
-The hybrid method is in all situations approximately a factor of two slower than \textsc{wsclean}.
+The $w$-snapshot method does not improve the performance much over normal \textsc{wsclean} operation in any of the cases.
+\editmark{This might seem to differ from some of the results in Fig.~\ref{fig:timings}, where the $w$-snapshot method does show improvement in certain cases.
+This is because Fig.~\ref{fig:timings} tests somewhat more extreme parameters, in which the $w$-snapshot method shows more benefit.
+The $w$-snapshot might become more valuable at higher ZAs or when the image size needs to be larger than the half-power beam width.} The hybrid method is in all situations approximately a factor of two slower than \textsc{wsclean}.
 All configurations listed in Table~\ref{tbl:computational-cost-configurations}, with the exception of VLSS and GLEAM, have optimal values for $\Delta w$ much smaller than one, suggesting the $w$-corrections should be done entirely by $w$-stacking instead of the $w$-stacking/projection hybrid method.
@@
 A hybrid between $w$-stacking and $w$-projection does not improve performance over a pure $w$-stacking implementation, but does improve a pure $w$-projection implementation significantly.
-Our optimisation of lowering the image resolution during inversion and prediction increases performance by a factor of 2--10 and has no noticeable effect on the accuracy.
+Our optimisation of lowering the image resolution during inversion and prediction increases performance by a factor of 2--10 and has no noticeable effect on the accuracy \editmark{in either our test simulations, which reach approximately a dynamic range of 1:1000, or on the complicated field of Fig.~\ref{fig:vela-projection-example}}.
 
@@
 Although we have measured our performance in FLOPS, it is likely that memory bandwidth will be a limiting factor, because memory bandwidth is increasing less quickly compared to the floating point performance (e.g., \citealt{gpu-gridding-romein-2012}).
-It was suggested that $w$-snapshots would improve the imaging speed for the SKA situation \citep{widefield-imaging-ska-cornwell}, but our results show this is not the case when already using the $w$-stacking algorithm.
-Clearly it remains challenging to perform wide-field imaging with acceptable performance, but some optimisations could be made for the SKA.
-For example, a performance gain of factors of a few can be achieved by averaging shorter baselines to their maximum integration time and bandwidth resolution before imaging.
+\editmark{\citet{widefield-imaging-ska-cornwell} suggests that the $w$-snapshots algorithm improves the imaging speed for the SKA situation, but our results show that $w$-snapshot imaging does not improve the SKA imaging performance when the $w$-stacking algorithm is used.} Clearly it remains challenging to perform wide-field imaging with acceptable performance, but some optimisations could be made for the SKA.
+For example, a performance gain of factors of a few can be achieved by averaging shorter baselines to their \editmark{lowest time and bandwidth resolution} before imaging.
 Moreover, although longer baselines make the imaging more expensive, imaging the full FOV will likely not (always) be required when using the longest baselines.

### 2014-05-26 articles/2014/Article-WSClean/img/fit.cpp

@@ -713,13 +713,13 @@ void optimalDeltaW(FittingFunc func, double a, double b, double c, double d, dou
 	std::cout << ", mean dw=" << sumDW/count << '\n';
 }*/
 
-void optimalDeltaT(FittingFunc func, const std::string& filename, double a, double b, double c, double d, double e, double f, double za, double nVis, double nPix, double fov, double nChan, bool zeroConstant, double& dt, size_t beamsTimesIters, double lambda, double maxBaseline, double maxHeight)
+void optimalDeltaT(FittingFunc func, const std::string& filename, double a, double b, double c, double d, double e, double f, double za, double nVis, double nPix, double fov, double nChan, bool zeroConstant, double& dt, size_t beamsTimesIters, double lambda, double maxBaseline, double maxHeight, double duration)
 {
 	std::ofstream file(filename);
 	double minY = std::numeric_limits<double>::max(), minDT = 0.0, minFOV = 0.0;
-	double dtTrial = 1.0, prevTrial = 1.0;
+	double dtTrial = 1.0;
 	double fz = zeroConstant ? 0.0 : f;
-	const double totalTime = 60.0*60.0;
+	const double totalTime = duration; //60.0*60.0;
 	while(dtTrial < totalTime)
 	{
 		double tAngle = dtTrial * (360.0 / 60.0 / 60.0 / 24.0);
@@ -734,7 +734,6 @@ void optimalDeltaT(FittingFunc func, const std::string& filename, double a, doub
 			minFOV = trialFOV;
 		}
 		file << dtTrial << '\t' << y << '\n';
-		prevTrial = dtTrial;
 		dtTrial *= 1.001;
 	}
 	dt = minDT;
@@ -773,7 +772,41 @@ void evalSurveyConfig(const std::string& desc, double lambda, double fwhm, size_
 	{
 		double dt, dw, minDWTime;
 		std::string filename = func == WSCleanFunc ? "dt-wsclean.txt" : "dt-casa.txt";
-		optimalDeltaT(func, filename, a, b, c, d, e, f, za, nMVis, nKPix, fov, nChan, false, dt, beams*majorIterationsInversionCount, lambda, maxBaseline, maxDiffHeight);
[…]

### 2014-07-01 articles/2014/Article-WSClean/Response-to-referee.txt
Adding some lost files from WSClean article

I have asked the director of the MWA board, S. Tingay, to respond to the editor's question on the appropriateness of the large number of coauthors. He asked me to forward the following answer:

"The authorship of this paper, and other similar papers published by the MWA consortium, contain a significant number of authors.  The authorship consists of people who have made a significant intellectual contribution to the establishment of the MWA (development of the instrumentation; deployment, testing, and commissioning of the instrument) as well as people who have undertaken the substantial observations and data analysis, and the interpretation of the data (in the case of this paper, as processed using the WSClean software.  The authors are people either funded to contribute to the MWA or are people who work on the MWA project in the research allocations of their positions.  All contributions are meaningful and are audited by the MWA Director and Board.  These contributions should be justly recognised in publications such as this."

We have marked the first author as a corresponding author, as suggested by the editor.

Additionally, we had taken off one co-author and added co-author B. Pindor, who had accidentially been left off.

We would like to thank the reviewer for his thorough review. We agree on almost all his comments, have changed the paper accordingly, and we think that his improved the clarity of the paper. We will provide detailed answers to his comments below. We have marked all changes with bold and red in the revised article.

p. 2 col. 1: “To reach high dynamic ranges, it can be necessary to deal with DDEs…” As worded, the paragraph seems to imply that A-projection is the complete answer to the DDE problem. A-projection will only correct for DDEs that are known. This can only be done directly for deterministic effects like primary beam rotation; effects like the ionosphere require separate calibration or estimation before A-projection can be applied.

Answer: We've added a sentence to clarify this.

p.2 col. 2: “”which we believe to be the fastest generic wide-field imager” such statements are good for talks, but not for papers. At least two GPU-based imagers purport to be faster, at which point one starts splitting hairs about terms like “generic” and “available”. I’d be more comfortable with a statement like “Our experience suggests that WSClean is anywhere from 2 to 12 times faster than mainstream w-projection imagers (CASA, lwimager) on the same datasets.” This is already impressive enough!

Answer: we have changed this to: "Our new imaging implementation, which in our experience is anywhere from 2 to 12 times faster than the CASA w-projection imager, is publicly released\footnote..."

Section 2: this needs a reference to Cornwell, or whoever suggested the w-stacking approach first.

Answer: We've added a citation to "Analysis of Convolutional Resampling Algorithm Performance" by B. Humphreys and T. J. Cornwell.

p. 3 col. 2: “Using too few w-layers…” The term w-aliasing is introduced unexplained (or is the paragraph meant to be the explanation?) In any case a better explanation of the effect is needed.

Answer: We've reworded this.

p. 4 Sect 3.2. This section needs more clarity, especially the last paragraph of p. 4. There needs to be a clear comparison between the Cornwell et al. 2012 definition of w-snapshots, and the method proposed here. I recommend splitting Sect. 3.2 in two, and moving the theoretical description of w-snapshots to Section 2, while leaving implementation details in Section 3.

Answer: We have moved the theoretical description to a new subsection in Section 2 as suggested, and added some text to explain the method more clearly.

Section 3 in general: recommend clearly splitting into generic implementation details, and MWA-specific details, as it is currently not clear how much is MWA-specific and whether this limits functionality in any way.

Answer: We have added a sentence to Sect 3.2 to explain that the method is generic, and renamed Sect 3.3 to include "for the MWA".

p. 7: “Our image-based beam-correction…” This can only work if the beam is also constant in time. How is sky rotation accounted for?

Answer: we account for this by doing snapshot imaging, and assuming a constant beam over the snapshot interval. We have reworded this sentence to say: "Our image-based beam-correction method avoids expensive beam-correcting kernels during gridding, but requires snapshot imaging, because the MWA beam changes over time because of Earth rotation, and only works because all tiles have the same beam."

p.7. “CASA convolves samples” better “CASA’s gridder implementation convolve samples”.

Answer: Reworded.

p.7 Sect 3.5.: “Decreasing inversion resolution” quite a mouthful, suggest revising title. “Lowering the resolution of reverse transforms”, perhaps?

Answer: Renamed to "Lowering the resolution of inversion"

p. 7 Figure 5: This figure needs labels, including axis labels.

Answer: We reworded the axis labels and reworded the figure caption.

p. 7 Sect. 3.5.: suggest use of “PSF” rather than “synthesized-beam” to avoid confusion with primary beam.

Answer: Not changed, we think that 'synthesized beam' should be a known term for readers of the article, and we think that talking about 'synthesized-beam width' makes a bit more sense than talking about 'PSF width'.

p. 7 Sect 3.5: “Therefore inversion at high resolution is not required…” and until end of paragraph, please improve clarity.

Answer: We have reworded this paragraph.

p. 7 “The outputs with and without rescaling”. The term “rescaling” is dropped in unannounced, does this simply refer to the resolution-lowering optimization described in this section? If so, this should be clearly defined.

Answer: Changed to: "The outputs with and without lowering the inversion and prediction resolution do not visibly differ, but ..."

p. 8 end of Sect 3.5: RMS comparison per se is somewhat meaningless without referencing the source flux (i.e. effectively the DR of the image). Do I interpret the colour bars of Figure 3 correctly, that this is a 10:1 DR image? Have the authors explored whether/how the lowered resolution affects things in mid- to high-DR situations? Given the nature of the optimization, I would expect the errors to scale with model flux, i.e. to be more noticeable at high DR. If so, this is an important caveat that should be noted.

Answer: We have added the peak flux level in the image (10 Jy) and clarified that the quoted noise level was in the residual image. If DR is defined as peak flux / noise level, then this image is a 1:150 DR image. In Section 4.1, a 1:1000 simulated scenario with a simpler model is tested, and as described there, no difference is visible between images with and without the method.

Figure 7: What does R-WSCLEAN refer to? I assume “recentering”, but this should be made clear. Or is it “rescaling”? If the former, and since “recentering” seems to be equivalent to w-snapshots (as Sect 3.2. implies, but could make clearer), perhaps this should be made clear in the figure caption.

Answer: Clarified in the caption; it refers indeed to recentring.

p.10 col. 2: “Imaging the MWA primary beam requires 3072 pixels… decreases to about 1500…” I thought the low-res optimization does not apply to the clean model image per se, but only kicks in during the reverse transform? if so, this sentence is confusing.

Answer: Reworded to: "If the small-inversion optimisation of \S3.5 is used, the inversion can be performed at an image size of 1500 pixels."

p. 12 Table 5: If R-WSCLEAN is equivalent to w-snapshot, why are we seeing almost exactly the same performance figures for w-stacking and w-snapshot? This seems to contradict the curves of Figure 7.

Answer: This is because Fig. 7 tests larger image sizes with more costly scenarios, which makes the w-snapshot is more effective. However, with the scenarios from Table 4, there is no benefit. We have added the following explanation to clarify this:
After "The w-snapshot method does not improve the performance much over normal WSCLEAN operation in any of the cases."
We added: "This might seem to differ from some of the results in Fig. 7, where the w-snapshot method does show improvement in certain cases. This is because Fig. 7 tests somewhat more extreme parameters, in which the w-snapshot method shows more benefit. The w-snapshot might become more valuable at higher ZAs or when the image size needs to be larger than the half-power beam width."

p. 12 Table 5: is the lowered-resolution optimization on for these performance comparisons? This should be stated. If yes, then the DR should be stated -- the caveat is, do we know how the optimization behaves at high DRs?

Answer: We have clarified in 4.6 that this assumes the optimization. We have added to Sect 5 the DR of the simulated observation (1:1000).

p. 12 col. 1: “The image size is set to two times the FOV width divided by the resolution”. Is this justified, given that the authors discuss the need for x5 oversampling for CLEAN earlier on?

Answer: This is justified by the fact that the small-inversion size optimization can be used, which we have no clarified.

p. 14 col. 1: “Our optimisation of lowering the image resolution … has no noticeable effect…” See remarks above, has this been tested at high DR?

Answer: Clarified, see earlier answer.

p. 14 col. 1: “It was suggested that w-snapshots would improve…” I do not understand the sentence, please rephrase.

Answer: Reworded.

### 2014-07-01 articles/2014/Article-WSClean/pack.sh

@@ -1,0 +1,10 @@
+tar -czvf wsclean.tar.gz wsclean.tex references.bib mn2e.* img/aliasing-example-annotated.pdf img/fullsky-together.pdf img/vela-normal-projection.pdf img/vela-zenith-projection.pdf img/pulsar-stokesi.pdf img/pulsar-stokesv.pdf img/benchmark-kernelsize/kernel.pdf img/residual-casa-fixed-grayscale.pdf img/residual-wsclean.pdf img/benchmark-nsamples/nsamples.pdf img/benchmark-zenith-angle/za.pdf img/benchmark-resolution/resolution.pdf img/benchmark-channels/channels.pdf img/benchmark-fov/fov.pdf
+mkdir test
+cd test
+tar -xzvf ../wsclean.tar.gz
+pdflatex wsclean.tex
+bibtex wsclean.aux
+pdflatex wsclean.tex
+bibtex wsclean.aux
+pdflatex wsclean.tex
+okular wsclean.pdf

### 2014-07-03 articles/2014/Article-WSClean/wsclean.tex
Removing editmarks

@@
 
-\newcommand{\editmark}[1]{#1}
-%\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
+%\newcommand{\editmark}[1]{#1}
+\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
 
@@
 
-\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\editmark{Corresponding author.} E-mail: \url{andre.offringa@anu.edu.au}},
+\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{Corresponding author. E-mail: \url{andre.offringa@anu.edu.au}},
 B.~McKinley\ANU$^,$\CAASTRO,
@@
 S.~M.~Ord\Curtin$^,$\CAASTRO,
-\editmark{B.~Pindor\UMelbourne,}
+B.~Pindor\UMelbourne,
 P.~Procopio\UMelbourne,
@@
 This is especially true for wide-field telescopes.
-One way to \editmark{correct for known} DDEs is by using the $a$-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
+One way to correct for known DDEs is by using the $a$-projection technique, which convolves the data during gridding with a kernel that corrects the DDEs \citep{aprojection-2008}.
 One particular DDE is the effect of the ionosphere.
@@
 The \textsc{awimager} \citep{awimager-2013} has been written to perform these corrections, and uses a hybrid of a-projection, $w$-projection and $w$-stacking.
-\editmark{The $a$-projection technique can only be applied directly for deterministic effects such as the correction of the primary beam.
-Effects like the ionosphere require separate calibration or estimation before $a$-projection can be applied.}
+The $a$-projection technique can only be applied directly for deterministic effects such as the correction of the primary beam.
+Effects like the ionosphere require separate calibration or estimation before $a$-projection can be applied.
 
@@
 We named the new imager ``WSClean'', as an abbreviation for ``$w$-Stacking Clean''.
-Our new imaging implementation, which \editmark{in our experience is anywhere from 2 to 12 times faster than the CASA $w$-projection imager}, is publicly released\footnote{The \textsc{wsclean} source code can be found at:\\\href{http://sourceforge.net/p/wsclean}{http://sourceforge.net/p/wsclean}}.
+Our new imaging implementation, which in our experience is anywhere from 2 to 12 times faster than the CASA $w$-projection imager, is publicly released\footnote{The \textsc{wsclean} source code can be found at:\\\href{http://sourceforge.net/p/wsclean}{http://sourceforge.net/p/wsclean}}.
 
@@
 In this section, we will describe the $w$-stacking algorithm from a mathematical point of view.
-\editmark{Instead of applying a convolution in $uv$-space, the $w$-stacking method grids visibilities on different $w$-layers and performs the $w$-corrections after the inverse Fourier transforms \citep{ska-memo-regridding-2011}.}
+Instead of applying a convolution in $uv$-space, the $w$-stacking method grids visibilities on different $w$-layers and performs the $w$-corrections after the inverse Fourier transforms \citep{ska-memo-regridding-2011}.
 
@@
 
-\editmark{To grid the visibilities, the $w$-values are rounded to the $w$-value of the nearest $w$-layer.
+To grid the visibilities, the $w$-values are rounded to the $w$-value of the nearest $w$-layer.
 This discretisation can cause noticeable aliasing when using too few $w$-layers, and results in decorrelation of sources far from the phase centre in the longer baselines.
-Additionally, this $w$-aliasing can cause ghost sources to appear} in the image.
+Additionally, this $w$-aliasing can cause ghost sources to appear in the image.
 An (extreme) example of this effect is shown in Fig.~\ref{fig:aliasing-example}.
@@
 
-\subsection{\editmark{$W$-snapshot imaging}} \label{sec:snapshot-imaging-theory}
-\editmark{$W$-snapshot imaging is a technique that combines warped-snapshot imaging with a $w$-correcting technique, such as $w$-projection or $w$-stacking \citep{widefield-imaging-ska-cornwell}.
+\subsection{$W$-snapshot imaging} \label{sec:snapshot-imaging-theory}
+$W$-snapshot imaging is a technique that combines warped-snapshot imaging with a $w$-correcting technique, such as $w$-projection or $w$-stacking \citep{widefield-imaging-ska-cornwell}.
 In the warped-snapshot imaging technique, the $w$-term is neglected, which results in an image with distorted coordinates \citep{perley-noncoplanar-arrays,mwa-interferometric-imaging}.
 Additionally, if the positions of the array elements are not perfectly planar or multiple timesteps are integrated to the same grid, visibilities will decorrelate and this can cause imaging artefacts.
-The original $w$-snapshot algorithm as described in \citet{widefield-imaging-ska-cornwell} corrects such artefacts by gridding visibilities on a tilted best-fit plane in $uvw$-space, and performs $w$-corrections towards that plane using $w$-projection or $w$-stacking.}
+The original $w$-snapshot algorithm as described in \citet{widefield-imaging-ska-cornwell} corrects such artefacts by gridding visibilities on a tilted best-fit plane in $uvw$-space, and performs $w$-corrections towards that plane using $w$-projection or $w$-stacking.
 
-\editmark{We have looked into implementing snapshot imaging with $w$-corrections in a slightly different way.
+We have looked into implementing snapshot imaging with $w$-corrections in a slightly different way.
 Instead of performing transforms of tilted planes, our method consists of phase-rotating the visibilities such that the phase centre of the observation is towards the zenith direction, i.e., the direction with minimal $w$-values.
@@
 This method results in $w$-corrected snapshots which need to be regridded before further integration, similar to the $w$-snapshots algorithm as described by \citet{widefield-imaging-ska-cornwell}.
-However, instead of creating warped images by performing the FFT over a tilted plane in $uvw$-space, this method transforms planes with constant $w$-values and produces images in zenith projection.}
+However, instead of creating warped images by performing the FFT over a tilted plane in $uvw$-space, this method transforms planes with constant $w$-values and produces images in zenith projection.
 
-\editmark{An} image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\rightarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
+An image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\rightarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
 \begin{align}\notag
@@
 In words, recentring an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging.
-\editmark{By doing the inverse corrections during prediction, a recentred visibility set can be cleaned with Cotton-Schwab iterations similar to a non-recentred set.}
+By doing the inverse corrections during prediction, a recentred visibility set can be cleaned with Cotton-Schwab iterations similar to a non-recentred set.
 
-\editmark{Our main reason for developing this method is that it is easier to implement, because no changes are required to the performance-critical gridding step.
-Another} benefit of our method is that the resulting image has a circular synthesised beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning.
+Our main reason for developing this method is that it is easier to implement, because no changes are required to the performance-critical gridding step.
+Another benefit of our method is that the resulting image has a circular synthesised beam, i.e., the resolution in $l$ and $m$ directions matches the intrinsic resolution of the instrument, which is desirable for cleaning.
 Regridding a recentred image is also more straightforward compared to regridding a warped snapshot, because in the latter case there is no analytic solution to the coordinate conversion \citep{perley-noncoplanar-arrays}.
@@
 \textsc{wsclean} allows cleaning individual polarisations, or can jointly deconvolve the polarisations.
-\editmark{In the latter case, peak finding can be performed in the sum of squared Stokes parameters, $I^2+Q^2+U^2+V^2$, or in} $pp^2+2pq \overline{(pq)} + qq^2$ space, where $p$ and $q$ are the two polarisations and $\overline{pq}=qp$ is the complex conjugate of $pq$.
+In the latter case, peak finding can be performed in the sum of squared Stokes parameters, $I^2+Q^2+U^2+V^2$, or in $pp^2+2pq \overline{(pq)} + qq^2$ space, where $p$ and $q$ are the two polarisations and $\overline{pq}=qp$ is the complex conjugate of $pq$.
 After a peak has been found, the PSF is subtracted from the individual polarisations with different factors.
@@
 Full-sky imaging can also be useful for self-calibration or deconvolution.
-For example, self-calibration using full-sky clean components has been found to give good results in imaging the resolved FR-II radio source Fornax~A (McKinley et al., \editmark{submitted}).
+For example, self-calibration using full-sky clean components has been found to give good results in imaging the resolved FR-II radio source Fornax~A (McKinley et al., submitted).
 An example of a full-sky MWA image is shown in Fig.~\ref{fig:full-sky-example}.
@@
 However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.
-\editmark{For these cases, the recentring technique described in \S\ref{sec:snapshot-imaging-theory} was implemented in \textsc{wsclean}.
+For these cases, the recentring technique described in \S\ref{sec:snapshot-imaging-theory} was implemented in \textsc{wsclean}.
 This allows making smaller images with minimal $w$-values that are recentred on the direction of interest.
-The implementation is generic and supports interferometric data from any telescope.}
+The implementation is generic and supports interferometric data from any telescope.
 
@@
 
-\editmark{Our implementation supports cleaning of recentred images in the same modes that normal images can be cleaned.} The PSF used during minor cleaning iterations is created by multiplying the weights with the recentring corrections, such that the PSF represents a source in the middle of the image.
+Our implementation supports cleaning of recentred images in the same modes that normal images can be cleaned.
+The PSF used during minor cleaning iterations is created by multiplying the weights with the recentring corrections, such that the PSF represents a source in the middle of the image.
 
@@
 
-\subsection{Beam correction \editmark{for the MWA}}
+\subsection{Beam correction for the MWA}
 \begin{figure*}
@@
 
-Our image-based beam-correction method avoids expensive beam-correcting kernels during gridding, but \editmark{requires snapshot imaging, because the MWA beam changes over time because of Earth rotation, and} only works because all tiles have the same beam.
+Our image-based beam-correction method avoids expensive beam-correcting kernels during gridding, but requires snapshot imaging, because the MWA beam changes over time because of Earth rotation, and only works because all tiles have the same beam.
 To apply the same method on heterogeneous arrays such as LOFAR, each set of correlations with a different combination of station beams will have to be imaged separately, which will increase the cost of the algorithm excessively unless an a-projection kernel is used.
@@
 \includegraphics[width=8cm]{img/benchmark-kernelsize/kernel}
-\caption{\editmark{Gridding kernel size plotted against imaging time, using common MWA settings.}}
+\caption{Gridding kernel size plotted against imaging time, using common MWA settings.}
 \label{fig:timing-kernelsize}
@@
 
-\subsection{\editmark{Lowering the resolution of inversion}} \label{sec:decreasing-inversion-resolution}
+\subsection{Lowering the resolution of inversion} \label{sec:decreasing-inversion-resolution}
 Typically, for cleaning it is desired that images have a pixel size (side length) at least five times smaller than the synthesised-beam width.
@@
 Therefore, as long as the $uv$-plane is sampled with at least the Nyquist frequency, a high-resolution image can be perfectly reconstructed from an inversion at lower resolution.
-\editmark{Cleaning can be performed on the high-resolution image that is reconstructed from the low-resolution image}.
-After \editmark{the high-resolution} image has been cleaned, a model is created at the same (high) resolution.
+Cleaning can be performed on the high-resolution image that is reconstructed from the low-resolution image.
+After the high-resolution image has been cleaned, a model is created at the same (high) resolution.
 Because the model consists of delta functions, the spatial frequencies in the model are not band-limited.
-Consequently, \editmark{lowering the resolution of the model image will remove information from the model.
+Consequently, lowering the resolution of the model image will remove information from the model.
 However, only the low spatial frequencies of this image will be used in the prediction, because in $uv$-space only visibilities up to the corresponding maximum baseline length will be sampled.
-Therefore, as long as lowering the resolution does} not modify the low-frequency components, the output of the prediction step will not change.
+Therefore, as long as lowering the resolution does not modify the low-frequency components, the output of the prediction step will not change.
 
@@
 Of the 24~min, 77\% is spent on cleaning approximately 100,000 components.
-The outputs with and without \editmark{lowering the inversion and prediction resolution} do not visibly differ, but \editmark{the difference between the residual images has an RMS of 18 mJy/beam.
-This can be compared to a noise level of 67 mJy/beam in the original residual image and a peak flux in the restored image of 10.3 Jy/beam}.
+The outputs with and without lowering the inversion and prediction resolution do not visibly differ, but the difference between the residual images has an RMS of 18 mJy/beam.
+This can be compared to a noise level of 67 mJy/beam in the original residual image and a peak flux in the restored image of 10.3 Jy/beam.
 Because the difference is mostly noise like, the difference could be caused by the non-linear behaviour of clean.
@@
 Unfinished lines indicate the imager could not run the specific configuration successfully.
-\editmark{The label ``R+\textsc{wsclean}'' refers to using WSClean with the recentring technique described in \S\ref{sec:snapshot-imaging-theory}.}}\label{fig:timings}
+The label ``R+\textsc{wsclean}'' refers to using WSClean with the recentring technique described in \S\ref{sec:snapshot-imaging-theory}.}\label{fig:timings}
 \end{figure*}
@@
 Imaging the MWA primary beam requires approximately an image size of 3072 pixels for cleaning.
-If the small-inversion optimisation of \S\ref{sec:decreasing-inversion-resolution} is used, \editmark{the inversion can be performed at an image size of 1500 pixels}.
+If the small-inversion optimisation of \S\ref{sec:decreasing-inversion-resolution} is used, the inversion can be performed at an image size of 1500 pixels.
 At ZA=10\degree, this saves about a factor of 2 in computing cost.
@@
 The image size is set to two times the FOV width divided by the resolution, as described by Eq.~\ref{eq:nyquist-resolution}.
-\editmark{Therefore, this assumes that the optimisation of \S\ref{sec:decreasing-inversion-resolution} to inverse at lower resolution is used.} The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).
+Therefore, this assumes that the optimisation of \S\ref{sec:decreasing-inversion-resolution} to inverse at lower resolution is used.
+The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).
 
@@
 The $w$-snapshot method does not improve the performance much over normal \textsc{wsclean} operation in any of the cases.
-\editmark{This might seem to differ from some of the results in Fig.~\ref{fig:timings}, where the $w$-snapshot method does show improvement in certain cases.
+This might seem to differ from some of the results in Fig.~\ref{fig:timings}, where the $w$-snapshot method does show improvement in certain cases.
 This is because Fig.~\ref{fig:timings} tests somewhat more extreme parameters, in which the $w$-snapshot method shows more benefit.
-The $w$-snapshot might become more valuable at higher ZAs or when the image size needs to be larger than the half-power beam width.} The hybrid method is in all situations approximately a factor of two slower than \textsc{wsclean}.
+The $w$-snapshot might become more valuable at higher ZAs or when the image size needs to be larger than the half-power beam width.
+The hybrid method is in all situations approximately a factor of two slower than \textsc{wsclean}.
 All configurations listed in Table~\ref{tbl:computational-cost-configurations}, with the exception of VLSS and GLEAM, have optimal values for $\Delta w$ much smaller than one, suggesting the $w$-corrections should be done entirely by $w$-stacking instead of the $w$-stacking/projection hybrid method.
@@
 A hybrid between $w$-stacking and $w$-projection does not improve performance over a pure $w$-stacking implementation, but does improve a pure $w$-projection implementation significantly.
-Our optimisation of lowering the image resolution during inversion and prediction increases performance by a factor of 2--10 and has no noticeable effect on the accuracy \editmark{in either our test simulations, which reach approximately a dynamic range of 1:1000, or on the complicated field of Fig.~\ref{fig:vela-projection-example}}.
+Our optimisation of lowering the image resolution during inversion and prediction increases performance by a factor of 2--10 and has no noticeable effect on the accuracy in either our test simulations, which reach approximately a dynamic range of 1:1000, or on the complicated field of Fig.~\ref{fig:vela-projection-example}.
 
@@
 Although we have measured our performance in FLOPS, it is likely that memory bandwidth will be a limiting factor, because memory bandwidth is increasing less quickly compared to the floating point performance (e.g., \citealt{gpu-gridding-romein-2012}).
-\editmark{\citet{widefield-imaging-ska-cornwell} suggests that the $w$-snapshots algorithm improves the imaging speed for the SKA situation, but our results show that $w$-snapshot imaging does not improve the SKA imaging performance when the $w$-stacking algorithm is used.} Clearly it remains challenging to perform wide-field imaging with acceptable performance, but some optimisations could be made for the SKA.
-For example, a performance gain of factors of a few can be achieved by averaging shorter baselines to their \editmark{lowest time and bandwidth resolution} before imaging.
+\citet{widefield-imaging-ska-cornwell} suggests that the $w$-snapshots algorithm improves the imaging speed for the SKA situation, but our results show that $w$-snapshot imaging does not improve the SKA imaging performance when the $w$-stacking algorithm is used.
+Clearly it remains challenging to perform wide-field imaging with acceptable performance, but some optimisations could be made for the SKA.
+For example, a performance gain of factors of a few can be achieved by averaging shorter baselines to their lowest time and bandwidth resolution before imaging.
 Moreover, although longer baselines make the imaging more expensive, imaging the full FOV will likely not (always) be required when using the longest baselines.

### 2014-07-03 articles/2014/Article-WSClean/wsclean.tex
Fixing suggested edits by reviewer

@@
 
-\date{Accepted TODO. Received TODO; in original form 01-04-2014}
+\date{Accepted TODO. Received TODO; in original form 1 April 2014}
 \pagerange{\pageref{firstpage}--\pageref{lastpage}}
@@
 
-\subsection{Time complexity of $w$-stacking} \label{sec:time-complexity}
+\subsection{\editmark{Computational} complexity of $w$-stacking} \label{sec:time-complexity}
 \begin{table}
@@
 
-As can be seen in Panel~\ref{fig:timing-fov}, the FOV has a large affect on performance when imaging off-zenith.
+As can be seen in Panel~\ref{fig:timing-fov}, the FOV has a large \editmark{effect} on performance when imaging off-zenith.
 This plot was created by varying the size of a pixel in the output image, such that the number of pixels in the image did not change.
@@
 Therefore, from a performance perspective, the snapshot duration should be as large as possible.
-In practise, the beam needs to be corrected on small time scales.
+In \editmark{practice}, the beam needs to be corrected on small time scales.
 A snapshot duration of more than a few minutes is therefore not possible.
@@
 We use the derived functions to estimate the computational cost of the algorithms for configurations of several telescopes.
-We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multi-frequency Snapshot Sky Survey (MSSS)\footnote{See~\href{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}{www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam--ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \href{http://www.aartfaac.org}{www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The Meer Karoo Array Telescope (MeerKAT); and the low-frequency Phase~1 aperture arrays of the Square-Kilometre Array (SKA), with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}.
+We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multi-frequency Snapshot Sky Survey (MSSS)\footnote{See~\href{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}{www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam--ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \href{http://www.aartfaac.org}{www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The \editmark{MeerKAT}; and the low-frequency Phase~1 aperture arrays of the Square-Kilometre Array (SKA), with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}.
 The configurations are summarised in Table~\ref{tbl:computational-cost-configurations}.
@@
 The image size is set to two times the FOV width divided by the resolution, as described by Eq.~\ref{eq:nyquist-resolution}.
-Therefore, this assumes that the optimisation of \S\ref{sec:decreasing-inversion-resolution} to inverse at lower resolution is used.
+Therefore, this assumes that the optimisation of \S\ref{sec:decreasing-inversion-resolution} to \editmark{compute the} inverse at lower resolution is used.
 The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).
@@
 Although we have measured our performance in FLOPS, it is likely that memory bandwidth will be a limiting factor, because memory bandwidth is increasing less quickly compared to the floating point performance (e.g., \citealt{gpu-gridding-romein-2012}).
-\citet{widefield-imaging-ska-cornwell} suggests that the $w$-snapshots algorithm improves the imaging speed for the SKA situation, but our results show that $w$-snapshot imaging does not improve the SKA imaging performance when the $w$-stacking algorithm is used.
+\citet{widefield-imaging-ska-cornwell} suggests that the $w$-snapshots algorithm improves the imaging speed for the SKA situation, but our results show that \editmark{$w$-snapshot imaging does not improve SKA imaging performance over $w$-stacking}.
 Clearly it remains challenging to perform wide-field imaging with acceptable performance, but some optimisations could be made for the SKA.

### 2014-07-04 articles/2014/Article-WSClean/wsclean.tex
Fixing figures as requested by editor: adding quantity and units where necessary, making line widths correct in legends
Mention 3 author-papers in full in first mention

@@
 \newcommand{\degree}{\ensuremath{^{\circ}}\xspace}
+
+% By changing this definition from {} to {-bw}, the grayscale plots are used in Figure 7.
+%\newcommand{\bwimgextension}{-bw}
+\newcommand{\bwimgextension}{}
 
@@
 
-There are several methods to deal with the $w$-term during imaging: faceting \citep{facetting-cornwell}; a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}; $w$-projection \citep{wprojection-cornwell}; $w$-stacking \citep{ska-memo-regridding-2011}; and warped snapshots \citep{perley-noncoplanar-arrays}.
-Hybrid methods are sometimes useful, such as with the $w$-snapshots method \citep{widefield-imaging-ska-cornwell}.
+There are several methods to deal with the $w$-term during imaging: faceting \citep{facetting-cornwell}; a three-dimensional Fourier transform \citep{perley-noncoplanar-arrays}; $w$-projection \citep*{wprojection-cornwell}; $w$-stacking \citep{ska-memo-regridding-2011}; and warped snapshots \citep{perley-noncoplanar-arrays}.
+Hybrid methods are sometimes useful, such as with the $w$-snapshots method \citep*{widefield-imaging-ska-cornwell}.
 
@@
 \begin{subfigure}{.5\linewidth}%
-\includegraphics[width=\linewidth]{img/benchmark-nsamples/nsamples}%
+\includegraphics[width=\linewidth]{img/benchmark-nsamples/nsamples\bwimgextension}%
 \caption{}\label{fig:timing-nsamples}%
@@
 \hspace{-.05\linewidth}\begin{subfigure}{.5\linewidth}%
-\includegraphics[width=\linewidth]{img/benchmark-zenith-angle/za}%
+\includegraphics[width=\linewidth]{img/benchmark-zenith-angle/za\bwimgextension}%
 \caption{}\label{fig:timing-zenith-angle}%
@@
 \begin{subfigure}{.5\linewidth}%
-\includegraphics[width=\linewidth]{img/benchmark-resolution/resolution}
+\includegraphics[width=\linewidth]{img/benchmark-resolution/resolution\bwimgextension}
 \caption{}\label{fig:timing-resolution}%
@@
 \hspace{-.05\linewidth}\begin{subfigure}{.5\linewidth}%
-\includegraphics[width=\linewidth]{img/benchmark-channels/channels}
+\includegraphics[width=\linewidth]{img/benchmark-channels/channels\bwimgextension}
 \caption{}\label{fig:timing-channels}%
@@
 \begin{subfigure}{.5\linewidth}%
-\includegraphics[width=\linewidth]{img/benchmark-fov/fov}
+\includegraphics[width=\linewidth]{img/benchmark-fov/fov\bwimgextension}
 \caption{}\label{fig:timing-fov}%

### 2014-07-04 articles/2014/Article-WSClean/pack.sh
Correcting ska memo reference

@@ -1,4 +1,4 @@
-tar -czvf wsclean.tar.gz wsclean.tex references.bib mn2e.* img/aliasing-example-annotated.pdf img/fullsky-together.pdf img/vela-normal-projection.pdf img/vela-zenith-projection.pdf img/pulsar-stokesi.pdf img/pulsar-stokesv.pdf img/benchmark-kernelsize/kernel.pdf img/residual-casa-fixed-grayscale.pdf img/residual-wsclean.pdf img/benchmark-nsamples/nsamples.pdf img/benchmark-zenith-angle/za.pdf img/benchmark-resolution/resolution.pdf img/benchmark-channels/channels.pdf img/benchmark-fov/fov.pdf
+tar -czvf wsclean.tar.gz wsclean.tex references.bib mn2e.* img/aliasing-example-annotated.pdf img/fullsky-together.pdf img/vela-normal-projection.pdf img/vela-zenith-projection.pdf img/pulsar-stokesi.pdf img/pulsar-stokesv.pdf img/benchmark-kernelsize/kernel.pdf img/residual-casa-fixed-grayscale.pdf img/residual-wsclean.pdf img/benchmark-nsamples/nsamples{.pdf,-bw.pdf} img/benchmark-zenith-angle/za{.pdf,-bw.pdf} img/benchmark-resolution/resolution{.pdf,-bw.pdf} img/benchmark-channels/channels{.pdf,-bw.pdf} img/benchmark-fov/fov{.pdf,-bw.pdf}
 mkdir test
 cd test
 tar -xzvf ../wsclean.tar.gz

### 2014-07-04 ps/pstransform.cpp

@@ -58,7 +58,7 @@ void averageAnuli(FFT2D* fft)
 		width = fft->Width(), midX = width/2;
 	size_t channelIndex;
 	ao::uvector<double> weights(perpGrid.size());
-	ao::uvector<std::complex<double>> sums(perpGrid.size());
+	ao::uvector<double> sums(perpGrid.size());
 	
 	while(transformTasks.read(channelIndex))
 	{
@@ -93,13 +93,13 @@ void averageAnuli(FFT2D* fft)
 						perpIndex = before;
 				}
 				size_t perpIntIndex = perpIndex->second;
-				sums[perpIntIndex] += rowIn[sourceX];
+				sums[perpIntIndex] += std::norm(rowIn[sourceX]);
 				++weights[perpIntIndex];
 			}
 		}
 		for(size_t i=0; i!=perpGrid.size(); ++i)
 		{
-			output[i] = std::norm(sums[i]) / weights[i];
+			output[i] = sums[i] / weights[i];
 		}
 		
 		delete[] input;
@@ -199,16 +199,21 @@ int main(int argc, char* argv[])
 	transformTasks.clear();
 	threads.clear();
 	
-	size_t perGridSize = 250;
+	size_t perGridSize = 500;
 	double maxVal = double(width)*0.5*M_SQRT2;
-	double minVal = 1e-4 * maxVal;
-	double eMin = exp(minVal), eRange = exp(maxVal)-eMin;
-	for(size_t dist=0; dist!=perGridSize; ++dist)
+	double minVal = 1e-3 * maxVal;
+	double nmin = 0, nmax = perGridSize;
+	double eRange = (log(maxVal)-log(minVal))/(nmax-nmin);
+	std::cout << "eRange="<< eRange << "(" << minVal << ',' << maxVal << ")\n";
+	perpGrid.insert(std::make_pair(0.0, 0));
+	for(size_t dist=0; dist!=perGridSize-1; ++dist)
 	{
-		double logVal = log(double(dist)*eRange/perGridSize+eMin); //dist*maxVal/perGridSize
+		double x = dist;
+		double logVal = minVal * exp(eRange * x) - nmin; //dist*maxVal/perGridSize
 		std::cout << logVal << ' ';
-		perpGrid.insert(std::make_pair(logVal, dist));
+		perpGrid.insert(std::make_pair(logVal, dist+1));
 	}
+	std::cout << '\n';
 	psData.resize(dataCube.size());
 	
 	std::cout << "Averaging anuli...\n";

### 2014-07-04 articles/2014/Article-WSClean/wsclean.tex
Some comments
Fixing affiliation list and removing Hewitt

@@
 
-%\newcommand{\editmark}[1]{#1}
-\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
+% Changes between revision are marked like \editmark{this is a change}, which becomes
+% bold and red. To turn this off, uncomment the next line and comment the subsequent line.
+\newcommand{\editmark}[1]{#1}
+%\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
 
@@
 \def\Victoria{$^{13}$}
-\def\CASS{$^{14}$}
-\def\Tata{$^{15}$}
-\def\UMelbourne{$^{16}$}
-\def\NRAO{$^{17}$}
-\def\UMichigan{$^{18}$}
+\def\UMichigan{$^{14}$}
+\def\CASS{$^{15}$}
+\def\Tata{$^{16}$}
+\def\UMelbourne{$^{17}$}
+\def\NRAO{$^{18}$}
 
@@
 B.~J.~Hazelton\UW, 
-J.~N.~Hewitt\MIT,
 L.~Hindson\Victoria,
-M.~Johnston-Hollitt\Victoria, \newauthor
-D.~C.~Jacobs\ASU,
+M.~Johnston-Hollitt\Victoria,
+D.~C.~Jacobs\ASU, \newauthor
 J.~C.~Kasper\UMichigan$^,$\CfA,
@@
 \Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
+\UMichigan{}University of Michigan, Ann Harbor, MI 48109, USA\\
 \CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
@@
 \UMelbourne{}School of Physics, The University of Melbourne, Parkville, VIC 3010, Australia\\
-\NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA\\
+\NRAO{}National Radio Astronomy Observatory, Charlottesville and Greenbank, USA
 }
@@
 % To make Dutch ``tussenvoegsels'' work correctly in Latex such as ``de Bruyn'', we use this command
-% In bibliography, it should be written with lowercase, as in ``de Bruyn''
+% In bibliography, it should be written with lowercase, as in ``de Bruyn''.
+% This comment switches to the bibliography style.
 \DeclareRobustCommand{\TUSSEN}[3]{#3}

### 2014-07-08 articles/2014/Article-WSClean/wsclean.tex
Removing editmarks for submission to Arxiv

@@
 \usepackage{amssymb}
-%\usepackage[font={it},labelfont={bf}]{caption}
-%\usepackage{units}
 \usepackage{times}
@@
 % bold and red. To turn this off, uncomment the next line and comment the subsequent line.
-\newcommand{\editmark}[1]{#1}
+%\newcommand{\editmark}[1]{#1}
 %\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
@@
 J.~Riding\UMelbourne,  \newauthor
-%A.~E.~E.~Rogers\Haystack,
-D.~A.~Roshi\NRAO, %\footnote{The National Radio Astronomy Observatory is a facility of the National Science Foundation operated under cooperative agreement by Associated Universities, Inc.},
+D.~A.~Roshi\NRAO,
 N.~Udaya~Shankar\RRI,
@@
 
-\date{Accepted TODO. Received TODO; in original form 1 April 2014}
+\date{Accepted 2014 July 6.  Received 2014 July 4; in original form 2014 April 1.}
 \pagerange{\pageref{firstpage}--\pageref{lastpage}}
@@
 
-\subsection{\editmark{Computational} complexity of $w$-stacking} \label{sec:time-complexity}
+\subsection{Computational complexity of $w$-stacking} \label{sec:time-complexity}
 \begin{table}
@@
 
-As can be seen in Panel~\ref{fig:timing-fov}, the FOV has a large \editmark{effect} on performance when imaging off-zenith.
+As can be seen in Panel~\ref{fig:timing-fov}, the FOV has a large effect on performance when imaging off-zenith.
 This plot was created by varying the size of a pixel in the output image, such that the number of pixels in the image did not change.
@@
 Therefore, from a performance perspective, the snapshot duration should be as large as possible.
-In \editmark{practice}, the beam needs to be corrected on small time scales.
+In practice, the beam needs to be corrected on small time scales.
 A snapshot duration of more than a few minutes is therefore not possible.
@@
 We use the derived functions to estimate the computational cost of the algorithms for configurations of several telescopes.
-We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multi-frequency Snapshot Sky Survey (MSSS)\footnote{See~\href{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}{www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam--ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \href{http://www.aartfaac.org}{www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The \editmark{MeerKAT}; and the low-frequency Phase~1 aperture arrays of the Square-Kilometre Array (SKA), with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}.
+We estimate the cost of imaging an hour of ZA=20\degree data for the following surveys or array configurations: The Galactic and Extragalactic MWA survey (GLEAM); the Evolutionary Map of the Universe (EMU) ASKAP survey \citep{emu-norris-2011}; the low and high bands of the LOFAR Multi-frequency Snapshot Sky Survey (MSSS)\footnote{See~\href{https://www.astron.nl/radio-observatory/lofar-msss/lofar-msss}{www.astron.nl/radio-observatory/lofar-msss/lofar-msss}}; all Dutch LOFAR stations \citep{lofar-2013}; the Amsterdam--ASTRON Radio Transients Facility and Analysis Centre (AARTFAAC) project\footnote{See \href{http://www.aartfaac.org}{www.aartfaac.org}}; The VLA Low-Frequency Sky Survey (VLSS, \citealt{vlss-2007}); The MeerKAT; and the low-frequency Phase~1 aperture arrays of the Square-Kilometre Array (SKA), with the full core (3~km) and the core+arms configurations \citep{ska-phase1-2013}.
 The configurations are summarised in Table~\ref{tbl:computational-cost-configurations}.
@@
 The image size is set to two times the FOV width divided by the resolution, as described by Eq.~\ref{eq:nyquist-resolution}.
-Therefore, this assumes that the optimisation of \S\ref{sec:decreasing-inversion-resolution} to \editmark{compute the} inverse at lower resolution is used.
+Therefore, this assumes that the optimisation of \S\ref{sec:decreasing-inversion-resolution} to compute the inverse at lower resolution is used.
 The total required computing power is calculated by multiplying the estimated computing time on our test machine with the performance of the machine (138 GFLOPS).
@@
 Although we have measured our performance in FLOPS, it is likely that memory bandwidth will be a limiting factor, because memory bandwidth is increasing less quickly compared to the floating point performance (e.g., \citealt{gpu-gridding-romein-2012}).
-\citet{widefield-imaging-ska-cornwell} suggests that the $w$-snapshots algorithm improves the imaging speed for the SKA situation, but our results show that \editmark{$w$-snapshot imaging does not improve SKA imaging performance over $w$-stacking}.
+\citet{widefield-imaging-ska-cornwell} suggests that the $w$-snapshots algorithm improves the imaging speed for the SKA situation, but our results show that $w$-snapshot imaging does not improve SKA imaging performance over $w$-stacking.
 Clearly it remains challenging to perform wide-field imaging with acceptable performance, but some optimisations could be made for the SKA.

### 2014-07-08 ps/plot/pswidget.cpp
Fixing some bugs in axes and using correct units for y-axis.

@@ -27,6 +27,8 @@ PSWidget::PSWidget() :
 	_showZAxisDescription(true),
 	_showTitle(true),
 	_max(1.0), _min(0.0),
+	_xAxisMin(1.0), _xAxisMax(10.0),
+	_yAxisMin(1.0), _yAxisMax(10.0),
 	_range(MinMax),
 	_cairoFilter(Cairo::FILTER_BEST),
 	_manualTitle(false),
@@ -154,11 +156,11 @@ void PSWidget::update(Cairo::RefPtr<Cairo::Context> cairo, unsigned width, unsig
 	}
 	if(_showXYAxes)
 	{
-		_vertScale->InitializeLogarithmicTicks(0.1, 10.0);
-		_vertScale->SetUnitsCaption("k\u2225 (h/Mpc)\u00B3");
+		_vertScale->InitializeLogarithmicTicks(_yAxisMin, _yAxisMax);
+		_vertScale->SetUnitsCaption("k\u2225 (h/Mpc)"); // \u00B3 = ^3
 	
-		_horiScale->InitializeLogarithmicTicks(0.1, 10.0);
-		_horiScale->SetUnitsCaption("k\u22A5 (h/Mpc)\u00B3"); // \u27C2 is official perp symbol
+		_horiScale->InitializeLogarithmicTicks(_xAxisMin, _xAxisMax);
+		_horiScale->SetUnitsCaption("k\u22A5 (h/Mpc)"); // \u27C2 is official perp symbol
 		
 		if(_manualXAxisDescription)
 			_horiScale->SetUnitsCaption(_xAxisDescription);
@@ -170,7 +172,7 @@ void PSWidget::update(Cairo::RefPtr<Cairo::Context> cairo, unsigned width, unsig
 		if(_manualZAxisDescription)
 			_colorScale->SetUnitsCaption(_zAxisDescription);
 		else
-			_colorScale->SetUnitsCaption("mK");
+			_colorScale->SetUnitsCaption("Power spectrum value (mK\u00B2[Mpc/h]\u00B3)");
 		if(_logZScale)
 			_colorScale->InitializeLogarithmicTicks(min, max);
 		else
@@ -333,12 +335,17 @@ void PSWidget::findMinMax(double& min, double& max)
 		min -= 1.0;
 		max += 1.0;
 	}
-	if(_logZScale && min<=0.0)
+	if(_logZScale)
 	{
-		if(max <= 0.0)
[…]

### 2014-07-08 ps/plot/pswidget.h

@@ -178,6 +178,10 @@ class PSWidget : public Gtk::DrawingArea {
 			_yStart = 0;
 			_yEnd = height;
 		}
+		void SetXAxisMin(double xAxisMin) { _xAxisMin = xAxisMin; }
+		void SetXAxisMax(double xAxisMax) { _xAxisMax = xAxisMax; }
+		void SetYAxisMin(double yAxisMin) { _yAxisMin = yAxisMin; }
+		void SetYAxisMax(double yAxisMax) { _yAxisMax = yAxisMax; }
 	private:
 		void findMinMax(double& min, double& max);
 		ColorMap* createColorMap();
@@ -214,6 +218,7 @@ class PSWidget : public Gtk::DrawingArea {
 		bool _showZAxisDescription;
 		bool _showTitle;
 		double _max, _min;
+		double _xAxisMin, _xAxisMax, _yAxisMin, _yAxisMax;
 		std::string _titleText, _manualTitleText;
 		enum Range _range;
 		Cairo::Filter _cairoFilter;

### 2014-07-08 ps/plot/tickset.h

@@ -197,11 +197,11 @@ class LogarithmicTickSet : public TickSet
 					return;
 				if(tickEnd > tickStart)
 				{
-					const unsigned distance = (unsigned) log10(tickEnd / tickStart);
+					const unsigned distance = (unsigned) round(log10(tickEnd / tickStart));
 					const unsigned step = (distance + sizeRequest - 1) / sizeRequest;
 					const double factor = exp10((double) step);
 					double pos = tickStart * factor;
-					while(pos <= tickEnd && _ticks.size() < sizeRequest)
+					while(pos <= _max && _ticks.size() < sizeRequest)
 					{
 						_ticks.push_back(pos);
 						pos *= factor;
@@ -227,20 +227,6 @@ class LogarithmicTickSet : public TickSet
 					double base = tickStart / 10.0;
 					do {
 						for(double i=2.0;i<9.0;i+=2.0)
-						{
-							double val = base * i;
-							if(val >= _min && val <= _max)
-								_ticks.push_back(val);
-						}
-						base *= 10.0;
-					} while(base < _max);
-				}
-				// can we add two and five?
-				else if((_ticks.size()+1)*3 < sizeRequest)
-				{
-					double base = tickStart / 10.0;
-					do {
-						for(double i=2.0;i<6.0;i+=3.0)
 						{
 							double val = base * i;
 							if(val >= _min && val <= _max)

### 2014-07-08 ps/plot/verticalplotscale.cpp

@@ -18,6 +18,7 @@
  *   59 Temple Place - Suite 330, Boston, MA  02111-1307, USA.             *
  ***************************************************************************/
 #include "verticalplotscale.h"
+#include <iostream>
 
 VerticalPlotScale::VerticalPlotScale()
 	: _plotWidth(0), _plotHeight(0), _metricsAreInitialized(false), _tickSet(0), _isLogarithmic(false), _drawWithDescription(true), _unitsCaption("y"), _descriptionFontSize(14), _tickValuesFontSize(14)

### 2014-07-08 ps/psplot.cpp

@@ -19,6 +19,23 @@ class PSWindow : public Gtk::Window {
 		{
 			FitsReader reader(filename);
 			ao::uvector<double> image(reader.ImageWidth() * reader.ImageHeight());
+			double minX, minY, maxX, maxY;
+			bool kwExists = reader.ReadDoubleKeyIfExists("PSXMIN", minX);
+			if(!reader.ReadDoubleKeyIfExists("PSYMIN", minY))
+				kwExists = false;
+			if(!reader.ReadDoubleKeyIfExists("PSXMAX", maxX))
+				kwExists = false;
+			if(!reader.ReadDoubleKeyIfExists("PSYMAX", maxY))
+				kwExists = false;
+			if(!kwExists)
+				std::cout << "Warning: PS range keywords missing -- axes won't have proper units.\n";
+			else {
+				std::cout << "Y: " << minY << '-' << maxY << '\n';
+				_psWidget.SetXAxisMin(minX);
+				_psWidget.SetYAxisMin(minY);
+				_psWidget.SetXAxisMax(maxX);
+				_psWidget.SetYAxisMax(maxY);
+			}
 			reader.Read(image.data());
 			_psWidget.SetImage(image.data(), reader.ImageWidth(), reader.ImageHeight());
 		}

### 2014-07-08 ps/pstransform.cpp

@@ -8,6 +8,7 @@
 #include "fitswriter.h"
 
 #include "aocommon/uvector.h"
+#include "universe.h"
 #include "windowfunction.h"
 
 fftw_plan fftPlan;
@@ -197,6 +198,12 @@ int main(int argc, char* argv[])
 	}
 	fft.Finish();
 	fft.SaveUV(dataCube.front(), "UV-first-image.fits");
+
+	double redshiftLo = Universe::HIRedshift(frequencies.back()), redshiftHi = Universe::HIRedshift(frequencies.front());
+	std::cout << "- Redshift range: " << redshiftLo << " - " << redshiftHi << '\n';
+	std::cout << "- Distance: " << Universe::ComovingDistanceInMPC(redshiftLo) << " - " << Universe::ComovingDistanceInMPC(redshiftHi) << " Mpc\n";
+	double distanceRange = (Universe::ComovingDistanceInMPC(redshiftHi)-Universe::ComovingDistanceInMPC(redshiftLo))/dataCube.size();
+	std::cout << "- Grid step: " << distanceRange << " Mpc\n";
 	
 	std::complex<double>* inputData = reinterpret_cast<std::complex<double>*>(fftw_malloc(dataCube.size() * sizeof(std::complex<double>)));
 	std::complex<double>* outputData = reinterpret_cast<std::complex<double>*>(fftw_malloc(dataCube.size() * sizeof(std::complex<double>)));
@@ -290,6 +297,10 @@ int main(int argc, char* argv[])
 		}
 	}
 	FitsWriter writer;
+	writer.SetExtraKeyword("PSXMIN", perpGrid.begin()->first);
+	writer.SetExtraKeyword("PSXMAX", perpGrid.rbegin()->first);
+	writer.SetExtraKeyword("PSYMIN", 0);
+	writer.SetExtraKeyword("PSYMAX", distanceRange);
 	writer.SetImageDimensions(perpGrid.size(), dataCube.size());
 	writer.Write("ps-linear.fits", psLinearImage.data());
 	
@@ -357,5 +368,10 @@ int main(int argc, char* argv[])
 		}
 	}
 	writer.SetImageDimensions(perpGrid.size(), logGridSize);
+	// TODO need to fix X axis min so that it is more accurate
+	writer.SetExtraKeyword("PSXMIN", (++perpGrid.begin())->first);
[…]

### 2014-07-08 ps/universe.h
Adding file with cosm. pars

@@ -1,0 +1,90 @@
+#ifndef UNIVERSE_H
+#define UNIVERSE_H
+
+#include <cmath>
+
+class Universe
+{
+public:
+	/**
+	 * Mass density parameter
+	 */
+	static double OmegaM()
+	{
+		return 0.27;
+	}
+	
+	/**
+	 * Curvature
+	 */
+	static double OmegaK()
+	{
+		return 1.0 - OmegaM() - OmegaL();
+	}
+	
+	/**
+	 * Dark energy density parameter
+	 */
+	static double OmegaL()
+	{
+		return 0.73;
+	}
+	
+	
+	static double H0()
+	{
+		return 0.72;
+	}
+	
+	/**
+	 * Hubble distance
+	 */
+	static double DH()
+	{
+		return C() / H0();
+	}
+	
+	static double DHinMPC()
+	{
+		return 4228.0;
+	}
+	
+	/**
+	 * Speed of light
+	 */
+	static double C()
+	{
+		return 299792458.0;
+	}
+	
+	/**
+	 * Hubble parameter
+	 */
+	static double E(double z)
+	{
+		return sqrt(OmegaM()*(1.0+z)*(1.0+z)*(1.0+z) + OmegaK()*(1.0+z)*(1.0+z) + OmegaL());
+	}
+	
+	static double ComovingDistanceInMPC(double z)
+	{
+		double d = 0.0;
+		for(size_t i=0; i!=10001; ++i)
+		{
+			d += 1.0 / E(double(i)*z/10000.0);
+		}
+		// We integrate over 10001 samples :
+		return d * DHinMPC() * z / 10001.0;
+	}
+	
+	static double HIRestFrequencyInHz()
+	{
+		return 1420405752.0;
+	}
+	
+	static double HIRedshift(double frequencyInHz)
+	{
+		return HIRestFrequencyInHz()/frequencyInHz - 1.0;
+	}
+};
+
+#endif

### 2014-08-07 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
More examples in article

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage[font={it},labelfont={bf}]{caption}
%\usepackage{units}
\usepackage{times}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{xspace}
%\usepackage{subcaption}

\usepackage{fixltx2e} % fixes placing of image positions

\definecolor{customhdrcolor}{rgb}{0.0,0.0,0.0}
\definecolor{customcitecolor}{rgb}{0.0,0.5,0.75}
\definecolor{customlinkcolor}{rgb}{0.0,0.5,0.75}

\usepackage[colorlinks=true,linkcolor=customlinkcolor,urlcolor=customlinkcolor,citecolor=customcitecolor,pdftex]{hyperref}

\ifpdf\pdfinfo{/Title      (Analysis of the Murchison radio environment for low-frequency astronomy)
               /Author     (A. R. Offringa et al.)
               /Keywords   (instrumentation: interferometers;methods: observational;techniques: interferometric;radio continuum: general)
        }
\else\usepackage{graphics}\fi

\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\newcommand{\editmark}[1]{#1}
%\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}

\newcommand{\degree}{\ensuremath{^{\circ}}\xspace}

% To make Dutch ``tussenvoegsels'' work correctly in Latex such as ``de Bruyn'', we use this command
% It fixes ordering and uppercases.
% In text, it should be written with uppercase, as in ``De Bruyn''
\DeclareRobustCommand{\TUSSEN}[3]{#2}

\title[The Murchison radio environment]{The low-frequency Murchison radio environment:\\interference analysis and mitigation}

\def\ANU{$^{1}$}
\def\CAASTRO{$^{2}$}
\def\Curtin{$^{3}$}

\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\editmark{Corresponding author.} E-mail: \url{andre.offringa@anu.edu.au}},
R.~B.~Wayth\Curtin$^,$\CAASTRO, ... , %\newauthor
builders list
\\
\ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
\Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
}

\begin{document}

\date{Accepted TODO. Received TODO; in original form TODO}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\pubyear{2014}

\label{firstpage}
\maketitle

\begin{abstract}
...
\end{abstract}

\begin{keywords}
instrumentation: interferometers -- methods: observational -- techniques: interferometric -- radio continuum: general
\end{keywords}

\section{Introduction}
In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of new radio observatories, such as the Atacama Large Millimeter Array (ALMA), the Green Bank Telescope (GBT), the Jansky Very Large Array (JVLA), the Low-Frequency Array (LOFAR), the Murchison Widefield Array (MWA), the Parkes Observatory and the Precision Array for Probing the Epoch of Reionization (PAPER). Spectrum allocation management and radio-quiet zones help to limit the interference, but do not solve interference from air-based transmitters or accidential electromagnetic radiation, for example from cars or wind turbines.

While some success has been seen in filtering interference from observations \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, in which the data samples are cleaned from the contribution of interference, such an approach is often not feasible. Therefore, the most common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{post-correlation-rfi-classification}. The consequence of the latter approach is that a certain amount of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed. It is important to analyse the results of interference detection, as improving the understanding of RFI contamination can aid in several ways: understanding the effects on the science; improved observation scheduling; designing hardware and location of future telescopes with a maximal cost/benefit approach; and enforcing spectrum management strategies that are most beneficial for the observatories.

In this article, we will look specifically at the RFI situation of the MWA \citep{mwa}. The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 80--300~MHz with a 31~MHz instantaneous bandwidth. Its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located in the Murchison region in Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be build near the location of the MWA.

The interferometric arrays LOFAR, PAPER and the Long Wavelength Array (LWA) are all designed to perform 21-cm EoR science, and all observe in approximately the same frequency range as the MWA. Comparing the MWA to these arrays therefore allows the comparison of the observatory site and hardware design with respect to the RFI mitigation effectivity. In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose any issues on regular observations despite that, unlike the MWA, LOFAR is build in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors analyse the first imaging of Epoch of Reionization long-integration observations with LOFAR. They reach near-thermal noise sensitivity, and argue that RFI is not a limitation for further increasing the sensitivity. Further analysis showed that RFI models for LOFAR can estimate the leakage of the RFI detection \citep{offringa-rfi-distributions}. This analysis concluded that with sufficient precautions, such as good receiver design and accurate detection methods, the leaked RFI is weak and averages down in longer integrations like Gaussian noise.

TODO paper structure.

\section{Method}
RFI detection (``data flagging'') is one of the first steps in the data processing. Several observatories or projects have designed pipelines that include RFI mitigation. In some projects, simple thresholding is used to mitigate the worst interference. For example, in \citet{parsons-paper-eorlimit-2014} a $3\sigma$ threshold is used for reducing PAPER data. However, more advanced techniques such as the \textsc{sumthreshold} technique have been shown to be more accurate \citep{post-correlation-rfi-classification}. A few pipelines that include RFI detection are \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR but used for various telescopes; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}).

It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern -- in particular for a many-element array such as the MWA. Because of the high accuracy and optimized algorithms of LOFAR's \textsc{aoflagger}, it is also a good candidate for flagging MWA data.

\subsection{The AOFlagger}

\begin{figure}
\begin{center}\hspace*{-0.8cm}\includegraphics[width=10.5cm]{img/bandpass}
\caption{Correlator output RMS over frequency in an MWA high-frequency observation, calculated over all baselines and 112 seconds of data. In this band, the band-pass shows two large discontinuities over frequency, because the MWA receivers apply different digital gains at different frequencies to minimize quantization noise. TODO explain subband bandpass somewhere %TODO
}
\label{fig:bandpass}
\end{center}
\end{figure}

To detect RFI in MWA observations, the \textsc{aoflagger} software is used. The \textsc{aoflagger} allows customizing the flagging strategies, such as the threshold levels and expected smoothness of good data. Optimal values vary from telescope to telescope because of a different band-pass shape, different time and frequency resolutions and varying field of views. Specific strategies for several telescopes are implemented, including an MWA specific strategy which is used in this work. All \textsc{aoflagger} strategies are variations on the original LOFAR strategy. In this strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency direction. Subsequently, line-shaped features are flagged with the \textsc{sumthreshold} method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

The MWA and LOFAR strategies differ on one aspect. For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation. This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimize quantization noise. These can not be fully corrected TODO why not? %TODO
An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}. The bandpass corrections are not permanently applied to the data, but only applied during flagging. Normally, a per-channel calibration is performed after flagging for getting accurate flux calibration, which corrects the gain discontinuities permanently.

\textsc{aoflagger} version 2.6 released on 26 June 2014 is used in this work.

\subsection{\textsc{cotter}: the MWA preprocessing pipeline}

The \textsc{aoflagger} software provides a C++ library\footnote{\url{http://aoflagger.sourceforge.net/doc/api/}} that can be integrated in a pipeline, thereby minimizing reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. Additionally, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume. It will also collect some statistics and write these into a measurement set with the same format as defined for LOFAR\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Various tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot these over various dimensions in an interactive manner.

Currently, the \textsc{cotter} preprocessing pipeline is ran by the scientist that calibrates and images the data. MWA observations are split into snapshots of a few minutes by the correlator. The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID. The first step for the scientist is to download the raw files for a given observation ID. There are various ways of processing MWA data afterwards. For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is ran in a special mode such that it only flags the data, and does not correct or convert the raw correlator files. After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS. For more traditional data processing with e.g. \textsc{casa} or \textsc{miriad}, the \textsc{cotter} output is set to either the \textsc{casa} or uvfits format, and is subsequently imported into the astronomical software.

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The \textsc{sumthreshold} technique and the scale-invariant rank operator that are used by the flagging strategy, use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency ranges that are handed over in one part to the flagger are as large as possible. However, a typical snapshot is about 50 GB in size, and without increasing reading and writing, it requires thus 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time chunks and flags those independently. This has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. Therefore, we are working on a platform with sufficient memory that pre-runs \textsc{cotter} on all observations.

\begin{itemize}
 \item Do we mention miriad and CASA's flagging tools?
\end{itemize}

\section{Data description}
\begin{itemize}
 \item Briefly describe GLEAM \& EoR projects (or in introduction?)
 \item List obs dates
\end{itemize}

\section{Detection results}

\noindent\begin{figure*}
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/amateur-2m-band-example}\includegraphics[width=9.1cm]{img/amateur-2m-band-flagged}
\caption{Strong RFI contamination found in the Amateur-band (2 m, 146 MHz), which falls in the lower band of the MWA EoR observations. These panels show the Stokes I amplitudes. In the right panel, the result of RFI detection is shown with purple. The thin horizontal lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively. Two more transmitters are detected in these data: at 150.17 and 156.7 Mhz, which are both one 40 kHz channel wide. These transmitters are not distinguishable in this plot, but are shown in the next figures.}
\label{fig:amateur-2m}
\end{center}
\end{figure*}

\noindent\begin{figure*}
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/150_2_mhz_example}\includegraphics[width=9.1cm]{img/150_2_mhz_flagged}
\caption{Short RFI burst of 2 s centred on 150.17 MHz}
\label{fig:150_2}
\end{center}
\end{figure*}

\noindent\begin{figure*}
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}
\caption{A very weakly observed transmitter centred on 156.66 MHz.}
\label{fig:156_7}
\end{center}
\end{figure*}

\begin{itemize}
 \item An example of RFI found (e.g., the EoR DVB, 
 \item Percentages per 
 \item Performance
\end{itemize}

\section{Discussion}
\begin{itemize}
 \item SKA -- how to implement a Cotter for SKA?
\end{itemize}

\section{Conclusions}
\begin{itemize}
 \item Much less RFI than LOFAR (but at a cost)
 \item About half the observations don't need flagging at all. 
 \item FM not completely empty
\end{itemize}

\section*{Acknowledgments}
This scientific work makes use of the Murchison Radio-astronomy Observatory, operated by CSIRO. We acknowledge the Wajarri Yamatji people as the traditional owners of the Observatory site. Support for the MWA comes from the U.S. National Science Foundation (grants AST-0457585, PHY-0835713, CAREER-0847753, and AST-0908884), the Australian Research Council (LIEF grants LE0775621 and LE0882938), the U.S. Air Force Office of Scientific Research (grant FA9550-0510247), and the Centre for All-sky Astrophysics (an Australian Research Council Centre of Excellence funded by grant CE110001020). Support is also provided by the Smithsonian Astrophysical Observatory, the MIT School of Science, the Raman Research Institute, the Australian National University, and the Victoria University of Wellington (via grant MED-E1799 from the New Zealand Ministry of Economic Development and an IBM Shared University Research Grant). The Australian Federal government provides additional support via the Commonwealth Scientific and Industrial Research Organisation (CSIRO), National Collaborative Research Infrastructure Strategy, Education Investment Fund, and the Australia India Strategic Research Fund, and Astronomy Australia Limited, under contract to Curtin University. We acknowledge the iVEC Petabyte Data Store, the Initiative in Innovative Computing and the CUDA Center for Excellence sponsored by NVIDIA at Harvard University, and the International Centre for Radio Astronomy Research (ICRAR), a Joint Venture of Curtin University and The University of Western Australia, funded by the Western Australian State government.

% To make Dutch ``tussenvoegsels'' work correctly in Latex such as ``de Bruyn'', we use this command
% In bibliography, it should be written with lowercase, as in ``de Bruyn''
\DeclareRobustCommand{\TUSSEN}[3]{#3}

\bibliographystyle{mn2e}
\bibliography{references}

\label{lastpage}

\end{document}

### 2014-08-07 articles/2014/Article-WSClean/pack-arxiv.sh
Corrections/scripts for WSClean article

@@ -1,0 +1,10 @@
+tar -czvf wsclean-arxiv.tar.gz wsclean.tex wsclean.bbl img/aliasing-example-annotated.pdf img/fullsky-together.pdf img/vela-normal-projection.pdf img/vela-zenith-projection.pdf img/pulsar-stokesi.pdf img/pulsar-stokesv.pdf img/benchmark-kernelsize/kernel.pdf img/residual-casa-fixed-grayscale.pdf img/residual-wsclean.pdf img/benchmark-nsamples/nsamples.pdf img/benchmark-zenith-angle/za.pdf img/benchmark-resolution/resolution.pdf img/benchmark-channels/channels.pdf img/benchmark-fov/fov.pdf
+rm test -rf
+mkdir test
+cd test
+cp ../mn2e.{cls,bst} .
+tar -xzvf ../wsclean-arxiv.tar.gz
+pdflatex wsclean.tex
+pdflatex wsclean.tex
+pdflatex wsclean.tex
+okular wsclean.pdf

### 2014-08-07 cotter/cotter.cpp
First changes to make writing flag files faster

@@ -245,13 +245,12 @@ void Cotter::processOneContiguousBand(const std::string& outputFilename, size_t 
 			_writer = new FlagWriter(outputFilename, _mwaConfig.HeaderExt().gpsTime, _mwaConfig.Header().nScans, _subbandCount, _subbandOrder);
 			break;
 		case FitsOutputFormat:
-			_writer = new FitsWriter(outputFilename);
+			_writer = new ThreadedWriter(new FitsWriter(outputFilename));
 			break;
 		case MSOutputFormat:
-			_writer = new MSWriter(outputFilename);
+			_writer = new ThreadedWriter(new MSWriter(outputFilename));
 			break;
 	}
-	_writer = new ThreadedWriter(_writer);
 	
 	if(freqAvgFactor != 1 || timeAvgFactor != 1)
 	{
@@ -769,6 +768,44 @@ void Cotter::processAndWriteTimestep(size_t timeIndex)
 	#endif
 				
 				_writer->WriteRow(dateMJD*86400.0, dateMJD*86400.0, antenna1, antenna2, u, v, w, _mwaConfig.Header().integrationTime, _outputData, _outputFlags, _outputWeights);
+			}
+		}
+	}
+}
+
+void Cotter::processAndWriteTimestepFlagsOnly(size_t timeIndex)
+{
+	const size_t antennaCount = _mwaConfig.NAntennae();
+	const size_t nChannels = nChannelsInCurSBRange();
+	const double dateMJD = _mwaConfig.Header().dateFirstScanMJD + timeIndex * _mwaConfig.Header().integrationTime/86400.0;
+	
+	_writer->AddRows(rowsPerTimescan());
+	
+	initializeWeights(_outputWeights);
+	for(size_t antenna1=0; antenna1!=antennaCount; ++antenna1)
+	{
+		for(size_t antenna2=antenna1; antenna2!=antennaCount; ++antenna2)
+		{
+			if(outputBaseline(antenna1, antenna2))
+			{
+				const FlagMask &flagMask = *_flagBuffers.find(std::pair<size_t, size_t>(antenna1, antenna2))->second;
+				
+				const size_t flagStride = flagMask.HorizontalStride();
+				
+				size_t bufferIndex = timeIndex - _curChunkStart;
+				for(size_t p=0; p!=4; ++p)
+				{
+					const bool *flagPtr = flagMask.Buffer()+bufferIndex;
[…]

### 2014-08-07 cotter/cotter.h

@@ -137,6 +137,7 @@ class Cotter : private UVWCalculater
 		void createReader(const std::vector<std::string> &curFileset);
 		void initializeReader();
 		void processAndWriteTimestep(size_t timeIndex);
+		void processAndWriteTimestepFlagsOnly(size_t timeIndex);
 		void baselineProcessThreadFunc();
 		void processBaseline(size_t antenna1, size_t antenna2, aoflagger::QualityStatistics &statistics);
 		void correctConjugated(aoflagger::ImageSet& imageSet, size_t imageIndex) const;

### 2014-08-07 cotter/cotter.cpp
Enabling faster writing routine for flag-only output mode

@@ -516,7 +516,10 @@ void Cotter::processOneContiguousBand(const std::string& outputFilename, size_t 
 			for(size_t t=_curChunkStart; t!=_curChunkEnd; ++t)
 			{
 				_progressBar->SetProgress(t-_curChunkStart, _curChunkEnd-_curChunkStart);
-				processAndWriteTimestep(t);
+				if(_outputFormat == FlagsOutputFormat)
+					processAndWriteTimestepFlagsOnly(t);
+				else
+					processAndWriteTimestep(t);
 			}
 			free(_outputData);
 			free(_outputWeights);

### 2014-08-07 cotter/version.h

@@ -1,9 +1,9 @@
 #ifndef COTTER_VERSION_H
 #define COTTER_VERSION_H
 
-#define COTTER_VERSION_STR  "2.3"
+#define COTTER_VERSION_STR  "2.4"
 #define COTTER_VERSION_MAJOR 2
-#define COTTER_VERSION_MINOR 3
-#define COTTER_VERSION_DATE "2014-07-14"
+#define COTTER_VERSION_MINOR 4
+#define COTTER_VERSION_DATE "2014-08-07"
 
 #endif

### 2014-08-13 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Small changes

@@
 \begin{center}\hspace*{-0.8cm}\includegraphics[width=10.5cm]{img/bandpass}
-\caption{Correlator output RMS over frequency in an MWA high-frequency observation, calculated over all baselines and 112 seconds of data.
+\caption{Correlator output RMS over frequency in an MWA high-frequency observation, calculated over all cross-correlated baselines and 112 seconds of data.
 In this band, the band-pass shows two large discontinuities over frequency, because the MWA receivers apply different digital gains at different frequencies to minimize quantization noise.
@@
 \begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/amateur-2m-band-example}\includegraphics[width=9.1cm]{img/amateur-2m-band-flagged}
-\caption{Strong RFI contamination found in the Amateur-band (2 m, 146 MHz), which falls in the lower band of the MWA EoR observations.
+\caption{Strong RFI contamination found in the 2 m amateur band (146 MHz), which falls in the lower band of the MWA EoR observations.
 These panels show the Stokes I amplitudes.

### 2014-08-26 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
More plots etc

@@
 
-While some success has been seen in filtering interference from observations \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, in which the data samples are cleaned from the contribution of interference, such an approach is often not feasible.
-Therefore, the most common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{post-correlation-rfi-classification}.
+While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI.
+Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{post-correlation-rfi-classification}.
 The consequence of the latter approach is that a certain amount of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed.
@@
 In this article, we will look specifically at the RFI situation of the MWA \citep{mwa}.
-The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 80--300~MHz with a 31~MHz instantaneous bandwidth.
+The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 80--300~MHz with a 31-MHz instantaneous bandwidth.
 Its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
@@
 Further analysis showed that RFI models for LOFAR can estimate the leakage of the RFI detection \citep{offringa-rfi-distributions}.
-This analysis concluded that with sufficient precautions, such as good receiver design and accurate detection methods, the leaked RFI is weak and averages down in longer integrations like Gaussian noise.
+Those analyses concluded that with sufficient precautions, such as good receiver design and accurate detection methods, the leaked RFI is weak and averages down similar to Gaussian noise.
 
@@
 \section{Method}
-RFI detection (``data flagging'') is one of the first steps in the data processing.
-Several observatories or projects have designed pipelines that include RFI mitigation.
+RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing.
 In some projects, simple thresholding is used to mitigate the worst interference.
 For example, in \citet{parsons-paper-eorlimit-2014} a $3\sigma$ threshold is used for reducing PAPER data.
-However, more advanced techniques such as the \textsc{sumthreshold} technique have been shown to be more accurate \citep{post-correlation-rfi-classification}.
-A few pipelines that include RFI detection are \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR but used for various telescopes; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}).
+For more accurate detection, advanced techniques such as the \textsc{sumthreshold} technique have been developed \citep{post-correlation-rfi-classification}.
+Several observatories or projects have designed pipelines that include advanced RFI mitigation.
+Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR but used for various telescopes; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}).
 
@@
 
-To detect RFI in MWA observations, the \textsc{aoflagger} software is used.
-The \textsc{aoflagger} allows customizing the flagging strategies, such as the threshold levels and expected smoothness of good data.
-Optimal values vary from telescope to telescope because of a different band-pass shape, different time and frequency resolutions and varying field of views.
-Specific strategies for several telescopes are implemented, including an MWA specific strategy which is used in this work.
-All \textsc{aoflagger} strategies are variations on the original LOFAR strategy.
-In this strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency direction.
-Subsequently, line-shaped features are flagged with the \textsc{sumthreshold} method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}.
+To detect RFI in MWA observations, we have used the \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data.
+The \textsc{aoflagger} allows customizing its flagging strategy, such as changing the threshold levels and expected smoothness of good data.
+Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions and varying field of views.
+Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA specific strategy which is used in this work.
+The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citet{lofar-radio-environment}.
+In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency direction.
+Subsequently, line-shaped features are detected by the \textsc{sumthreshold} method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}.
 After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask.
@@
 Additionally, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume.
-It will also collect some statistics and write these into a measurement set with the same format as defined for LOFAR\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}.
+It also collects some statistics and write these into a measurement set with the same format as used by LOFAR\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}.
 Various tools are available to analyse these statistics, e.g.
@@
 
-Currently, the \textsc{cotter} preprocessing pipeline is ran by the scientist that calibrates and images the data.
 MWA observations are split into snapshots of a few minutes by the correlator.
 The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID.
+For more details about the correlator, see TODO fix reference Ord et al.
+(2014).
+Currently, the \textsc{cotter} preprocessing pipeline is ran by the scientist that calibrates and images the data.
 The first step for the scientist is to download the raw files for a given observation ID.
@@
 For more traditional data processing with e.g.
-\textsc{casa} or \textsc{miriad}, the \textsc{cotter} output is set to either the \textsc{casa} or uvfits format, and is subsequently imported into the astronomical software.
+\textsc{casa} or \textsc{miriad}, the \textsc{cotter} output is set to either the \textsc{casa} or uvfits format.
+The output is directly readable by the astronomical software.
 
@@
 Therefore, we are working on a platform with sufficient memory that pre-runs \textsc{cotter} on all observations.
-
-\begin{itemize}
- \item Do we mention miriad and CASA's flagging tools?
-\end{itemize}
+The partitioning of data to overcome the time-ordering problem is also use by the \textsc{ndppp} software used by LOFAR (TODO reference?).
 
@@
 \begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}
-\caption{A very weakly observed transmitter centred on 156.66 MHz.}
+\caption{A weakly-observed transmitter centred on 156.66 MHz.}
 \label{fig:156_7}
+\end{center}
+\end{figure*}
+
+\noindent\begin{figure*}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
+\caption{..}
+\label{fig:stddev-per-set}
+\end{center}
+\end{figure*}
+
+\noindent\begin{figure*}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set}\vspace{-1cm}
+\caption{..}
+\label{fig:rfi-per-set}
 \end{center}

### 2014-09-01 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Plots and data description

@@
 \section{Introduction}
-In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of new radio observatories, such as the Atacama Large Millimeter Array (ALMA), the Green Bank Telescope (GBT), the Jansky Very Large Array (JVLA), the Low-Frequency Array (LOFAR), the Murchison Widefield Array (MWA), the Parkes Observatory and the Precision Array for Probing the Epoch of Reionization (PAPER).
+In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of radio observatories.
 Spectrum allocation management and radio-quiet zones help to limit the interference, but do not solve interference from air-based transmitters or accidential electromagnetic radiation, for example from cars or wind turbines.
@@
 The consequence of the latter approach is that a certain amount of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed.
-It is important to analyse the results of interference detection, as improving the understanding of RFI contamination can aid in several ways: understanding the effects on the science; improved observation scheduling; designing hardware and location of future telescopes with a maximal cost/benefit approach; and enforcing spectrum management strategies that are most beneficial for the observatories.
+Analysing the results of interference detection and improving the understanding of RFI contamination can aid in several ways: understanding the effects on the science; improved observation scheduling; designing hardware and location of future telescopes with a maximal cost/benefit approach; and enforcing spectrum management strategies that are most beneficial for the observatories.
 
@@
 In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose any issues on regular observations despite that, unlike the MWA, LOFAR is build in a populated area.
-This is confirmed by \citet{ncp-eor-yatawatta}, where the authors analyse the first imaging of Epoch of Reionization long-integration observations with LOFAR.
+This is confirmed by \citet{ncp-eor-yatawatta}, where the authors analyse the first imaging of EoR long-integration observations with LOFAR.
 They reach near-thermal noise sensitivity, and argue that RFI is not a limitation for further increasing the sensitivity.
@@
 RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing.
-In some projects, simple thresholding is used to mitigate the worst interference.
-For example, in \citet{parsons-paper-eorlimit-2014} a $3\sigma$ threshold is used for reducing PAPER data.
-For more accurate detection, advanced techniques such as the \textsc{sumthreshold} technique have been developed \citep{post-correlation-rfi-classification}.
-Several observatories or projects have designed pipelines that include advanced RFI mitigation.
-Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR but used for various telescopes; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}).
-
 It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}.
-Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern -- in particular for a many-element array such as the MWA.
-Because of the high accuracy and optimized algorithms of LOFAR's \textsc{aoflagger}, it is also a good candidate for flagging MWA data.
+Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern -- in particular for a many-element arrays.
+In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate.
+For example, a $3\sigma$ threshold is used for reducing PAPER data in \citet{parsons-paper-eorlimit-2014}.
+Several observatories or projects have designed pipelines that include more advanced RFI mitigation.
+Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}).
+LOFAR's \textsc{aoflagger} has shown to have a good accuracy, is fast and has a library interface, which allows it to be integrated in a pipeline.
+It is therefore also used for MWA's RFI detection.
 
@@
 This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimize quantization noise.
-These can not be fully corrected TODO why not?
-%TODO
 An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}.
@@
 Normally, a per-channel calibration is performed after flagging for getting accurate flux calibration, which corrects the gain discontinuities permanently.
+Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but this step was still included in this work.
 
-\textsc{aoflagger} version 2.6 released on 26 June 2014 is used in this work.
+In this work, \textsc{aoflagger} version 2.6 released on 26 June 2014 is used.
 
@@
 Additionally, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume.
-It also collects some statistics and write these into a measurement set with the same format as used by LOFAR\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}.
+It also collects some statistics and writes these into a measurement set with the same format as used by LOFAR\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}.
 Various tools are available to analyse these statistics, e.g.
@@
 Currently, the \textsc{cotter} preprocessing pipeline is ran by the scientist that calibrates and images the data.
-The first step for the scientist is to download the raw files for a given observation ID.
-There are various ways of processing MWA data afterwards.
-For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is ran in a special mode such that it only flags the data, and does not correct or convert the raw correlator files.
+Once the scientist has downloaded the raw files for a given observation ID, there are various ways of processing MWA data.
+For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is ran in a special mode such that it only flags the data, and does not convert the raw correlator files.
 After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS.
@@
 \textsc{casa} or \textsc{miriad}, the \textsc{cotter} output is set to either the \textsc{casa} or uvfits format.
-The output is directly readable by the astronomical software.
+The output is then directly readable by the astronomical software.
 
@@
 The \textsc{sumthreshold} technique and the scale-invariant rank operator that are used by the flagging strategy, use statistics over large time and frequency ranges.
-Therefore, accuracy is improved when the time-frequency ranges that are handed over in one part to the flagger are as large as possible.
+Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible.
 However, a typical snapshot is about 50 GB in size, and without increasing reading and writing, it requires thus 50 GB of memory to perform flagging on the full data.
 When less memory is available, \textsc{cotter} will split the observation in different time chunks and flags those independently.
-This has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy.
-Therefore, we are working on a platform with sufficient memory that pre-runs \textsc{cotter} on all observations.
-The partitioning of data to overcome the time-ordering problem is also use by the \textsc{ndppp} software used by LOFAR (TODO reference?).
+The \textsc{ndppp} software used by LOFAR (TODO reference?) uses a similar partitioning of data to overcome the time-ordering problem.
+Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy.
+Because not all astronomers might have easily access to high-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations.
 
 \section{Data description}
+
+\begin{table}
+\caption{..}\label{tbl:obs-list}
+\begin{tabular}{|l|l|r|r|}
+\hline
+\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} \\
+\hline
+EoR   & 2013-08-23 & 138.9--197.7 & 3h\\
+GLEAM & 2013-08-25 & 72.3--133.8, 138.8--230.8 & 7h \\
+EoR   & 2013-08-26 & 138.9--197.7 & 6h \\
+GLEAM & 2013-11-25 & 72.3--133.8, 138.8--230.8 & 7h \\
+GLEAM & 2014-03-16 & 72.3--133.8, 138.8--230.8 & 7h \\
+EoR   & 2014-04-10 & 138.9--197.7 & 6h \\
+GLEAM & 2013-11-05 & 72.3--133.8, 138.8--230.8 & 7h \\
+EoR   & 2014-02-05 & 138.9--197.7 & 6h\\
+GLEAM & 2014-03-17 & 72.3--133.8, 138.8--230.8 & 7h\\
+GLEAM & 2014-06-18 & 72.3--133.8, 138.8--230.8 & 7h \\
+RFI test&2014-08-27& 131.2--161.9 & 2$\times$8 m \\
+\hline
+\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} \\
+\hline
+\end{tabular}
+\end{table}
+
+We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall).
+The observations are listen in Table~\ref{tbl:obs-list}.
+
+The MWA EoR project observes primarily in the 138.9--197.7~MHz range, covering the HI 21-cm line with redshift 6.1--9.2.
+Because the MWA has a 30.72~MHz instantaneous bandwidth, this range has been split up so that each night observes with a central frequency of either 154.2 or 182.4~MHz, with an overlap of 2.6 MHz.
+In these observations, the beamformers track a selected EoR field such that the sensitivity of the primary beam is maximal towards the field.
+
+The GLEAM survey covers 72.3--230.8 MHz split into 5 bands of 30.72 MHz. Because the GLEAM observations are made in drift-scan mode, 5 subsequent snapshots observe all 5 bands.
+
+Both projects avoid the subbands that cover the frequency range 133.8--138.8~MHz, because the ORBCOMM low Earth orbitting (LEO) satellites transmit in these subbands.
+When these frequencies are observed, the subbands are often so strongly contaminated that it affects imaging sensitivity.
+For completeness, we include an 16-min observation that covers this frequency range.
+
 \begin{itemize}
- \item Briefly describe GLEAM \& EoR projects (or in introduction?)
  \item List obs dates

### 2014-09-05 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Work on article

@@
 
-\begin{table}
-\caption{..}\label{tbl:obs-list}
-\begin{tabular}{|l|l|r|r|}
+\begin{table*}
+\caption{List of observations used in the analyses}\label{tbl:obs-list}
+\begin{tabular}{|l|l|r|r|r|c|}
 \hline
-\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} \\
+\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DVB? \\
 \hline
-EoR   & 2013-08-23 & 138.9--197.7 & 3h\\
-GLEAM & 2013-08-25 & 72.3--133.8, 138.8--230.8 & 7h \\
-EoR   & 2013-08-26 & 138.9--197.7 & 6h \\
-GLEAM & 2013-11-25 & 72.3--133.8, 138.8--230.8 & 7h \\
-GLEAM & 2014-03-16 & 72.3--133.8, 138.8--230.8 & 7h \\
-EoR   & 2014-04-10 & 138.9--197.7 & 6h \\
-GLEAM & 2013-11-05 & 72.3--133.8, 138.8--230.8 & 7h \\
-EoR   & 2014-02-05 & 138.9--197.7 & 6h\\
-GLEAM & 2014-03-17 & 72.3--133.8, 138.8--230.8 & 7h\\
-GLEAM & 2014-06-18 & 72.3--133.8, 138.8--230.8 & 7h \\
-RFI test&2014-08-27& 131.2--161.9 & 2$\times$8 m \\
+EoR   & 2013-08-23 & 138.8--197.7              & 3h & 0.53\% & ---\\
+GLEAM & 2013-08-25 & 72.3--133.8, 138.8--230.8 & 7h & 0.94\% & no \\
+EoR   & 2013-08-26 & 138.8--197.7              & 6h & 0.81\% & ---\\
+GLEAM & 2013-11-05 & 72.3--133.8, 138.8--230.8 & 7h &  1.3\% & no \\
+GLEAM & 2013-11-25 & 72.3--133.8, 138.8--230.8 & 7h & 0.69\% & no \\
+EoR   & 2014-02-05 & 138.8--197.7              & 6h & 0.54\% & -- \\
+GLEAM & 2014-03-16 & 72.3--133.8, 138.8--230.8 & 7h & 0.79\% & no \\
+GLEAM & 2014-03-17 & 72.3--133.8, 138.8--230.8 & 7h & 1.64\% & yes \\
+EoR   & 2014-04-10 & 138.8--197.7              & 6h & 0.68\% & ---\\
+GLEAM & 2014-06-18 & 72.3--133.8, 138.8--230.8 & 7h & 0.98\% & no \\
+RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
 \hline
-\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} \\
+\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.97\%} & \textbf{1/6}\\
 \hline
 \end{tabular}
-\end{table}
+\end{table*}
 
@@
 Because the MWA has a 30.72~MHz instantaneous bandwidth, this range has been split up so that each night observes with a central frequency of either 154.2 or 182.4~MHz, with an overlap of 2.6 MHz.
-In these observations, the beamformers track a selected EoR field such that the sensitivity of the primary beam is maximal towards the field.
+In these observations, the selected EoR field is tracked byt the primary beam, such that the sensitivity towards the field is maximal.
 
@@
 For completeness, we include an 16-min observation that covers this frequency range.
-
-\begin{itemize}
- \item List obs dates
-\end{itemize}
 
@@
 \begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/150_2_mhz_example}\includegraphics[width=9.1cm]{img/150_2_mhz_flagged}
-\caption{Short RFI burst of 2 s centred on 150.17 MHz}
+\caption{Short RFI burst of 2 s centred on 150.17 MHz. This shows that flagging at high time and frequency resolution is necessary to accurately detect RFI.}
 \label{fig:150_2}
@@
 \begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}
-\caption{A weakly-observed transmitter centred on 156.66 MHz.}
+\caption{A weakly-observed transmitter centred on 156.66 MHz.
+Although it is hard to identify in such a time-frequency plot, the same frequency is flagged in several correlations, and it can therefore be assumed this is real RFI.
+This is an example in which the AOFlagger detect an event that would not be detected by regular thresholding.
+}
 \label{fig:156_7}
@@
 \end{figure*}
+
+\noindent\begin{figure}
+% obs data: 1061583576, 2013-08-26.
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/dvb_example}
+\caption{RFI from Digital Video Broadcasting (DVB).
+Because of its broadband nature, this kind of RFI is not well flagged in the initial AOFlagger step.
+It is detected by a later step by checking the global statistics.
+This type of RFI makes most of the 30.72~MHz bandwidth unusable.
+It is not seen very often (1 out of 6 observed nights).
+}
+\label{fig:dvb_example}
+\end{center}
+\end{figure}
 
@@
 
+\noindent\begin{figure*}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/2014-03-17-dvb-stddevs}\vspace{-1cm}
+\caption{..}
+\label{fig:rfi-per-set}
+\end{center}
+\end{figure*}
+
 \begin{itemize}
- \item An example of RFI found (e.g., the EoR DVB, 
- \item Percentages per 
+ \item Total RFI statistics
+ \item Percentages + stddev per frequency + discuss
+ \item An example of RFI found (e.g., the EoR DVB, ..)
  \item Performance
+ \item Log N log S Histograms
 \end{itemize}

### 2014-09-05 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Work

@@
 \begin{table*}
-\caption{List of observations used in the analyses}\label{tbl:obs-list}
+\caption{List of observations used in the analyses.
+The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has detected (which includes false positives).
+The DVB column displays the fraction of visibilities that were unusable because of the presence of DVB signals.
+Because the DVB frequencies are only observed in GLEAM observations, this fraction is only given for those.
+}\label{tbl:obs-list}
 \begin{tabular}{|l|l|r|r|r|c|}
 \hline
-\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DVB? \\
+\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DVB \\
 \hline
 EoR   & 2013-08-23 & 138.8--197.7              & 3h & 0.53\% & ---\\
-GLEAM & 2013-08-25 & 72.3--133.8, 138.8--230.8 & 7h & 0.94\% & no \\
+GLEAM & 2013-08-25 & 72.3--133.8, 138.8--230.8 & 7h & 0.94\% & 0\% \\
 EoR   & 2013-08-26 & 138.8--197.7              & 6h & 0.81\% & ---\\
-GLEAM & 2013-11-05 & 72.3--133.8, 138.8--230.8 & 7h &  1.3\% & no \\
-GLEAM & 2013-11-25 & 72.3--133.8, 138.8--230.8 & 7h & 0.69\% & no \\
+GLEAM & 2013-11-05 & 72.3--133.8, 138.8--230.8 & 7h & 1.30\% & 0\% \\
+GLEAM & 2013-11-25 & 72.3--133.8, 138.8--230.8 & 7h & 0.69\% & 0\% \\
 EoR   & 2014-02-05 & 138.8--197.7              & 6h & 0.54\% & -- \\
-GLEAM & 2014-03-16 & 72.3--133.8, 138.8--230.8 & 7h & 0.79\% & no \\
-GLEAM & 2014-03-17 & 72.3--133.8, 138.8--230.8 & 7h & 1.64\% & yes \\
+GLEAM & 2014-03-16 & 72.3--133.8, 138.8--230.8 & 7h & 0.79\% & 0\% \\
+GLEAM & 2014-03-17 & 72.3--133.8, 138.8--230.8 & 7h & 1.64\% & 3\% \\
 EoR   & 2014-04-10 & 138.8--197.7              & 6h & 0.68\% & ---\\
-GLEAM & 2014-06-18 & 72.3--133.8, 138.8--230.8 & 7h & 0.98\% & no \\
+GLEAM & 2014-06-18 & 72.3--133.8, 138.8--230.8 & 7h & 0.98\% & 0\% \\
 RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
 \hline
-\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.97\%} & \textbf{1/6}\\
+\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.97\%} & \textbf{0.5\%}\\
 \hline
@@
 \section{Detection results}
+
+\noindent\begin{figure*}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
+\caption{..}
+\label{fig:stddev-per-set}
+\end{center}
+\end{figure*}
 
@@
 \end{figure}
-
-\noindent\begin{figure*}
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
-\caption{..}
-\label{fig:stddev-per-set}
-\end{center}
-\end{figure*}
 

### 2014-09-08 articles/2014/Abstract-BASP/abstract.tex
BASP abstract beginning

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[9pt,conference,a4paper]{IEEEtran}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% correct bad hyphenation here
%\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}

\title{Radio interferometric imaging for the SKA and its pathfinders}

\author{%
\IEEEauthorblockN{
Andr\'e R. Offringa\IEEEauthorrefmark{1}}
%
\IEEEauthorblockA{\IEEEauthorrefmark{1} 
Research School of Astronomy \& Astrophysics,
Australian National University,
Weston Creek, 2611 ACT
Australia.}
%
} % end of author

\maketitle

\begin{abstract}
Please add a very brief abstract here (max 150 words).
\end{abstract}

% Either use sections (but not subsections please) or add a 9pt space
% here.  If you use sections then please remove the 9pt space.
%\vspace*{9pt}
\section{Introduction}
To make maps of the sky with a radio interferometer, the radio data has to be processed with an imaging algorithm, and the SKA pathfinders show that this imaging step is a considerable challenge \cite{awimager-2013,offringa-wsclean-2014}. The imaging complexity scales with the number of elements, maximum baseline length, angular field of view, data volume and dynamic range. The SKA will be larger by orders of magnitude on all these aspects \cite{ska-phase1-2013}, and will therefore require new advances in solving the imaging problem. I will give an overview of the challenges involved, and summarize the currently available approaches.

\section{W-term correction}
The first step in imaging is the mathematical inversion of the visibility function -- the function that maps the sky brightness to the complex correlation products. For simple arrays, this function simplifies into a two-dimensional Fourier transform, but arrays such as the SKA that will have a wide field of view and non-coplanar elements require correction for the '$w$-terms', which describe the distance of an element to the plane. Larger $w$-terms increase the cost of the inversion. Two algorithms to correct for the $w$-terms are $w$-projection \cite{wprojection-cornwell} and $w$-stacking \cite{ska-memo-regridding-2011}. With common imaging parameters, these algorithms have a squared and linear dependency on the size of the $w$-term respectively, but the faster $w$-stacking algorithm requires more memory \cite{offringa-wsclean-2014}. Further improvement can be made by combining either of these techniques with snapshot imaging. This process, called $w$-snapshot imaging, decreases the size of the $w$-term, but requires regridding of the snapshot images before they can be integrated or deconvolved \cite{widefield-imaging-ska-cornwell}. An implementation of this algorithm in \textsc{wsclean} has shown that this is effective only for very off-zenith imaging or for imaging very large fields of view \cite{offringa-wsclean-2014}.

\section{Correction of directional effects}
Direction-dependent effects (DDEs) that change over time, such as a changing primary beam or ionosphere -- which will be the case for the SKA -- have to be corrected for before images from different epochs are added together. Pathfinders have shown this can be corrected during the inversion step using the $a$-projection technique \cite{awimager-2013,aprojection-2008} or, when the changes are uniform over the elements, by correcting snapshots in image space \cite{offringa-wsclean-2014,mwa-interferometric-imaging}. A hybrid of these two can be used to reduce the cost.

\section{Deconvolution}

\section{Parallellism}

\section{\textsc{wsclean}}
WSClean is a new and fast widefield imager developed in particular for imaging MWA data, but it is a generic imager able to deconvolve, produce beam-corrected images for fixed tile arrays and perform wideband deconvolution. Since recently it also has a novel multi-resolution multi-scale clean algorithm \cite{offringa-wsclean-2014}.

\DeclareRobustCommand{\TUSSEN}[3]{#3}

\bibliographystyle{IEEEtran}
\bibliography{bib}

\end{document}

### 2014-09-10 articles/2014/Abstract-BASP/abstract.tex
BASP abstract

@@
 \begin{abstract}
-Please add a very brief abstract here (max 150 words).
-\end{abstract}
+To make maps of the sky with a radio interferometer, the radio data has to be processed with an imaging algorithm.
+The SKA pathfinders show that this imaging step can be a considerable challenge.
+The imaging complexity scales with the number of elements, maximum baseline length, angular field of view, data volume and dynamic range.
+The SKA will be larger by orders of magnitude on all these aspects, and will therefore require new advances in solving the imaging problem.
+I will give an overview of the challenges involved, and summarize the currently available approaches.\end{abstract}
 
@@
 \section{Introduction}
-To make maps of the sky with a radio interferometer, the radio data has to be processed with an imaging algorithm, and the SKA pathfinders show that this imaging step is a considerable challenge \cite{awimager-2013,offringa-wsclean-2014}.
-The imaging complexity scales with the number of elements, maximum baseline length, angular field of view, data volume and dynamic range.
-The SKA will be larger by orders of magnitude on all these aspects \cite{ska-phase1-2013}, and will therefore require new advances in solving the imaging problem.
-I will give an overview of the challenges involved, and summarize the currently available approaches.
+Many algorithms are available for imaging radio data, that each have their own benefits and performance dependencies.
+In practice, many of these algorithms are combined, making a full imaging program very complex.
+In the following sections, I will describe the commonly used algorithms and their use for the SKA.
 
-\section{W-term correction}
-The first step in imaging is the mathematical inversion of the visibility function -- the function that maps the sky brightness to the complex correlation products.
-For simple arrays, this function simplifies into a two-dimensional Fourier transform, but arrays such as the SKA that will have a wide field of view and non-coplanar elements require correction for the '$w$-terms', which describe the distance of an element to the plane.
-Larger $w$-terms increase the cost of the inversion.
-Two algorithms to correct for the $w$-terms are $w$-projection \cite{wprojection-cornwell} and $w$-stacking \cite{ska-memo-regridding-2011}.
-With common imaging parameters, these algorithms have a squared and linear dependency on the size of the $w$-term respectively, but the faster $w$-stacking algorithm requires more memory \cite{offringa-wsclean-2014}.
-Further improvement can be made by combining either of these techniques with snapshot imaging.
-This process, called $w$-snapshot imaging, decreases the size of the $w$-term, but requires regridding of the snapshot images before they can be integrated or deconvolved \cite{widefield-imaging-ska-cornwell}.
-An implementation of this algorithm in \textsc{wsclean} has shown that this is effective only for very off-zenith imaging or for imaging very large fields of view \cite{offringa-wsclean-2014}.
+\section{$w$-term correction}
+The first step in imaging is the mathematical inversion of the function that maps the sky brightness to the correlation products.
+The performance of this step often dominates the total imaging time.
+Arrays such as the SKA that will have a wide field of view and non-coplanar elements require expensive correction for the '$w$-terms', which describe the distance of an element to the plane.
+The fastests algorithms to correct for the $w$-terms are $w$-projection \cite{wprojection-cornwell} and $w$-stacking \cite{ska-memo-regridding-2011}.
+These algorithms have a squared and linear dependency on the size of the $w$-term respectively, although the faster $w$-stacking algorithm requires more memory \cite{offringa-wsclean-2014}.
+Further improvements can be made by combining either of these techniques with snapshot imaging.
+This decreases the size of the $w$-term, but requires regridding of the snapshot images before they can be integrated or deconvolved \cite{widefield-imaging-ska-cornwell}.
+This algorithm has recently been implemented in \textsc{wsclean}, and is useful for MWA imaging of very off-zenith or very large fields of view \cite{offringa-wsclean-2014}.
 
 \section{Correction of directional effects}
-Direction-dependent effects (DDEs) that change over time, such as a changing primary beam or ionosphere -- which will be the case for the SKA -- have to be corrected for before images from different epochs are added together.
-Pathfinders have shown this can be corrected during the inversion step using the $a$-projection technique \cite{awimager-2013,aprojection-2008} or, when the changes are uniform over the elements, by correcting snapshots in image space \cite{offringa-wsclean-2014,mwa-interferometric-imaging}.
-A hybrid of these two can be used to reduce the cost.
+Direction-dependent effects (DDEs) such as the ionosphere or a changing or heterogenous primary beam -- as will be the case for the SKA -- have to be corrected before images from different epochs are added together.
+It has been shown this can be corrected during the inversion step using the $a$-projection technique \cite{aprojection-2008} or, when the changes are uniform over the elements, by correcting snapshots in image space \cite{offringa-wsclean-2014}.
+For certain effects, a hybrid of these two can be used to reduce cost.
 
 \section{Deconvolution}
+An image created by inversion is convolved with the point-spread function (PSF) of the instrument, and therefore needs to be deconvolved to improve the dynamic range.
+For a wide-field array such as the SKA, the PSF varies spatially, which complicates this process.
+The Cotton-Schwab algorithm \cite{cotton-schwab-clean} corrects this: Minor iterations assume a constant PSF and produce a model for the image, while major iterations calculate and subtract the full effect of the changing PSF.
+The simplest algorithm for implementing minor iterations is the H\"ogbom algorithm \cite{hogbom-clean}.
+In more complex fields, multi-scale (MS) cleaning improves the deconvolution accuracy \cite{multiscale-clean-cornwell-2008}, but the application of \textsc{casa}'s MS algorithm on MWA data shows the algorithm is too slow for practical use with a wide-field array with so many elements.
+A fast multi-resolution multi-scale algorithm has recently been implemented in \textsc{wsclean} (Offringa et al., in prep.).
+Recently, deconvolution using compressed sensing (CS) have shown promising results \cite{wenger-compressed-sensing-2010}.
+CS improves the deconvolution accuracy over H\"ogbom cleaning, but its performance and robustness have not been extensively validated using real data.
 
-\section{Parallellism}
+Deconvolving multi-frequency synthesis (MFS) images is complicated further when a large bandwidth is imaged, making it necessary to account for spectral variation of sources.
+The Sault-Wieringa algorithm \cite{sault-wieringa-1994} takes spectral variation into account during deconvolution, and produces the MFS map as well as a spectral-index map.
+A different approach is taken by \textsc{wsclean}, which splits the bandwidth in subbands and deconvolves them joinedly.
+Further variation can be corrected during the major iterations.
 
-\section{\textsc{wsclean}}
-WSClean is a new and fast widefield imager developed in particular for imaging MWA data, but it is a generic imager able to deconvolve, produce beam-corrected images for fixed tile arrays and perform wideband deconvolution.
-Since recently it also has a novel multi-resolution multi-scale clean algorithm \cite{offringa-wsclean-2014}.
+\section{Parallellism \& data format}
+In practical situations, a hybrid of these algorithms is often used, although different science goals will need different subsets of algorithms.
+Software packages such as \textsc{casa} and \textsc{wsclean} already implement many of these algorithms.
+Adaptions will be necessary for SKA's data volumes, such as parallellization over subband, short snapshot or image subset (mosaicing).
+The data will need to be available in such a way that each node can work on a subset of the data.
 

### 2014-09-11 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Adding log N log S images
Small changes

@@
 \caption{List of observations used in the analyses.
-The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has detected (which includes false positives).
-The DVB column displays the fraction of visibilities that were unusable because of the presence of DVB signals.
+The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (which includes false positives).
+The DVB column displays the fraction of visibilities that were unusable because of the presence of DVB signals, and which the flagger does not flag.
 Because the DVB frequencies are only observed in GLEAM observations, this fraction is only given for those.
@@
 
-The GLEAM survey covers 72.3--230.8 MHz split into 5 bands of 30.72 MHz. Because the GLEAM observations are made in drift-scan mode, 5 subsequent snapshots observe all 5 bands.
+The GLEAM survey covers 72.3--230.8 MHz split into 5 bands of 30.72 MHz. The GLEAM observations are made in drift-scan mode, and the 5 bands are observed one by one in 5 subsequent snapshots.
 
@@
 \caption{..}
-\label{fig:rfi-per-set}
+\label{fig:dvb-stddevs}
 \end{center}
 \end{figure*}
+
+\noindent\begin{figure}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/lognlogs-residual}
+\caption{Log N log S for residuals}
+\label{fig:lognlogs-residuals}
+\end{center}
+\end{figure}
+
+\noindent\begin{figure}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/lognlogs-rfi}
+\caption{Log N log S for RFI }
+\label{fig:lognlogs-rfi}
+\end{center}
+\end{figure}
 

### 2014-09-11 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Adding seperated plots

--- 2014-09-05 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
@@
 \noindent\begin{figure*}
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set}\vspace{-1cm}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set-eor}\vspace{-1cm}
 \caption{..}
-\label{fig:rfi-per-set}
+\label{fig:rfi-per-set-eor}
+\end{center}
+\end{figure*}
+
+\noindent\begin{figure*}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set-gleam}\vspace{-1cm}
+\caption{..}
+\label{fig:rfi-per-set-gleam}
 \end{center}

### 2014-09-11 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Work

--- 2014-09-11 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
@@
 \noindent\begin{figure*}
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set}\vspace{-1cm}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set-eor}\vspace{-1cm}
 \caption{..}
-\label{fig:rfi-per-set}
+\label{fig:rfi-per-set-eor}
 \end{center}
@@
 \noindent\begin{figure*}
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/2014-03-17-dvb-stddevs}\vspace{-1cm}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set-gleam}\vspace{-1cm}
+\caption{..}
+\label{fig:rfi-per-set-gleam}
+\end{center}
+\end{figure*}
+
+\noindent\begin{figure}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/2014-03-17-dvb-stddevs}
 \caption{..}
@@
 \end{center}
-\end{figure*}
+\end{figure}
 

### 2014-09-11 altcotter/averagingwriter.h
Removing altcotter branch, all its features are now available in Cotter

@@ -1,252 +1,0 @@
-#ifndef AVERAGING_MS_WRITER_H
-#define AVERAGING_MS_WRITER_H
-
-#include "writer.h"
-
-#include <iostream>
-
-class UVWCalculater
-{
-	public:
-		virtual void CalculateUVW(double date, size_t antenna1, size_t antenna2, double &u, double &v, double &w) = 0;
-};
-
-class AveragingWriter : public Writer
-{
-	public:
-		AveragingWriter(Writer *writer, size_t timeCount, size_t freqAvgFactor, UVWCalculater &uvwCalculater)
-		: _writer(writer), _timeAvgFactor(timeCount), _freqAvgFactor(freqAvgFactor), _rowsAdded(0),
-		_originalChannelCount(0), _avgChannelCount(0), _antennaCount(0), _uvwCalculater(uvwCalculater)
-		{
-		}
-		
-		virtual ~AveragingWriter()
-		{
-			destroyBuffers();
-			
-			delete _writer;
-		}
-		
-		virtual void WriteBandInfo(const std::string &name, const std::vector<Writer::ChannelInfo> &channels, double refFreq, double totalBandwidth, bool flagRow)
-		{
-			if(channels.size()%_freqAvgFactor != 0)
-			{
-				std::cout << " Warning: channels averaging factor is not a multiply of total number of channels. Last channel(s) will be left out.\n";
-			}
-			
-			_avgChannelCount = channels.size() / _freqAvgFactor;
-			_originalChannelCount = channels.size();
-			
-			std::vector<Writer::ChannelInfo> avgChannels(_avgChannelCount);
-			for(size_t ch=0; ch!=_avgChannelCount; ++ch)
-			{
-				Writer::ChannelInfo channel;
-				channel.chanFreq = 0.0;
-				channel.chanWidth = 0.0;
-				channel.effectiveBW = 0.0;
-				channel.resolution = 0.0;
-				for(size_t i=0; i!=_freqAvgFactor; ++i)
-				{
-					const Writer::ChannelInfo& curChannel = channels[ch*_freqAvgFactor + i];
-					channel.chanFreq += curChannel.chanFreq;
-					channel.chanWidth += curChannel.chanWidth;
-					channel.effectiveBW += curChannel.effectiveBW;
-					channel.resolution += curChannel.resolution;
-				}
-				
-				channel.chanFreq /= (double) _freqAvgFactor;
-				
[…]

### 2014-09-11 altcotter/baselinebuffer.h

@@ -1,43 +1,0 @@
-#ifndef BASELINE_BUFFER_H
-#define BASELINE_BUFFER_H
-
-#include <cstring>
-
-class BaselineBuffer
-{
-	public:
-		BaselineBuffer() :
-			nElementsPerRow(0)
-		{
-			for(size_t p=0; p!=4; ++p)
-			{
-				real[p] = 0; imag[p] = 0;
-			}
-		}
-		
-		BaselineBuffer(const BaselineBuffer &source) :
-			nElementsPerRow(source.nElementsPerRow)
-		{
-			for(size_t p=0; p!=4; ++p)
-			{
-				real[p] = source.real[p];
-				imag[p] = source.imag[p];
-			}
-		}
-		
-		BaselineBuffer& operator=(const BaselineBuffer &source)
-		{
-			nElementsPerRow = source.nElementsPerRow;
-			for(size_t p=0; p!=4; ++p)
-			{
-				real[p] = source.real[p];
-				imag[p] = source.imag[p];
-			}
-			return *this;
-		}
-		
-		float *real[4], *imag[4];
-		size_t nElementsPerRow;
-};
-
-#endif

### 2014-09-11 altcotter/data/convert_avgxyyx_to_code.cpp

@@ -1,24 +1,0 @@
-#include <iostream>
-#include <iomanip>
-#include <fstream>
-
-using namespace std;
-
-int main(int argc, char *argv[])
-{
-  const char *filename = argv[1];
-  ifstream stream(filename);
-  cout << "double passband[128] = {\n  ";
-  cout << setprecision(15);
-  for(size_t i=0; i!=128; ++i)
-  {
-    unsigned channelNo;
-    double xx, xy, yx, yy;
-    stream >> channelNo >> xx >> xy >> yx >> yy;
-    cout << ((xy + yx) * 0.5);
-    if(i != 127) cout << ", ";
-    if(i%4 == 3 && i!=127)
-      cout << "\n  ";
-  }
-  std::cout << "\n};\n";
-}

### 2014-09-11 altcotter/data/make_code.sh

@@ -1,2 +1,0 @@
-g++ -Wall -O3 -o convert_avgxyyx_to_code convert_avgxyyx_to_code.cpp
-./convert_avgxyyx_to_code subband-passband.txt

### 2014-09-11 altcotter/fitsuser.cpp

@@ -1,35 +1,0 @@
-#include "fitsuser.h"
-
-#include <fitsio.h>
-
-#include <stdexcept>
-#include <sstream>
-
-void FitsUser::checkStatus(int status)
-{
-	if(status != 0)
-	{
-		throwError(status);
-	}
-}
-
-void FitsUser::throwError(int status, const std::string &msg)
-{
-	std::stringstream msgStr;
-	
-	if(!msg.empty()) msgStr << msg << ". ";
-	
-	char statusStr[FLEN_STATUS], errMsg[FLEN_ERRMSG];
-	fits_get_errstatus(status, statusStr);
-	msgStr << "CFitsio reported an error while reading fits files. Status = " << status << ": " << statusStr;
-
-	if(fits_read_errmsg(errMsg))
-	{
-		msgStr << ". Error message stack: " << errMsg << '.';
-
-		while( fits_read_errmsg(errMsg) )
-			msgStr << ' ' << errMsg << '.';
-	}
-	throw std::runtime_error(msgStr.str());
-}
-

### 2014-09-11 altcotter/fitswriter.h

@@ -1,115 +1,0 @@
-#ifndef FITSWRITER_H
-#define FITSWRITER_H
-
-#include "fitsuser.h"
-#include "writer.h"
-
-#include <fitsio.h>
-
-#include <complex>
-#include <vector>
-#include <string>
-
-class FitsWriter : public Writer, private FitsUser
-{
-	public:
-		FitsWriter(const std::string& filename);
-		virtual ~FitsWriter();
-		
-		virtual void WriteBandInfo(const std::string& name, const std::vector<ChannelInfo>& channels, double refFreq, double totalBandwidth, bool flagRow);
-		virtual void WriteAntennae(const std::vector<AntennaInfo>& antennae, double time);
-		virtual void WritePolarizationForLinearPols(bool flagRow);
-		virtual void WriteField(const FieldInfo& field);
-		virtual void WriteSource(const SourceInfo &source);
-		virtual void WriteObservation(const std::string& telescopeName, double startTime, double endTime, const std::string& observer, const std::string& scheduleType, const std::string& project, double releaseDate, bool flagRow);
-		virtual void WriteHistoryItem(const std::string &commandLine, const std::string &application, const std::vector<std::string> &params);
-		virtual void SetArrayLocation(double x, double y, double z)
-		{
-			_arrayX = x;
-			_arrayY = y;
-			_arrayZ = z;
-		}
-		
-		virtual void AddRows(size_t count);
-		virtual void WriteRow(double time, double timeCentroid, size_t antenna1, size_t antenna2, double u, double v, double w, double interval, const std::complex<float>* data, const bool* flags, const float *weights);
-		virtual bool AreAntennaPositionsLocal() const { return true; }
-		
-	private:
-		void initGroupHeader();
-		void writeAntennaTable();
-		
-		void setKeywordToDouble(const char *keywordName, double value) const
-		{
-			int status = 0;
-			if(fits_update_key(_fptr, TDOUBLE, keywordName, &value, NULL, &status))
-				throwError(status, "Could not write keyword");
-		}
-		
-		void setKeywordToFloat(const char *keywordName, float value) const
-		{
-			int status = 0;
-			if(fits_update_key(_fptr, TFLOAT, keywordName, &value, NULL, &status))
-				throwError(status, "Could not write keyword");
-		}
-		
-		void setKeywordToInt(const char *keywordName, int value) const
-		{
[…]

### 2014-09-11 altcotter/flagwriter.cpp

@@ -1,161 +1,0 @@
-#include "flagwriter.h"
-#include <stdexcept>
-
-const uint16_t
-	FlagWriter::VERSION_MINOR = 0,
-	FlagWriter::VERSION_MAJOR = 1;
-
-FlagWriter::FlagWriter(const std::string &filename, int gpsTime, size_t timestepCount, size_t gpuBoxCount, const std::vector<size_t>& subbandToGPUBoxFileIndex) :
-	_timestepCount(timestepCount),
-	_antennaCount(0),
-	_channelCount(0),
-	_channelsPerGPUBox(0),
-	_polarizationCount(0),
-	_rowStride(0),
-	_rowCount(0),
-	_gpuBoxCount(gpuBoxCount),
-	_gpsTime(gpsTime),
-	_files(gpuBoxCount)
-{
-	if(gpuBoxCount == 0)
-		throw std::runtime_error("Flagwriter was initialized with zero gpuboxes");
-		
-	size_t numberPos = filename.find("%%");
-	if(numberPos == std::string::npos)
-		throw std::runtime_error("When writing flag files, multiple files will be written. Therefore, the name of the flagfile should contain two percent symbols (\"%%\"), e.g. \"Flagfile%%.mwaf\". These will be replaced by the gpubox(/subband) number.");
-	std::string name(filename);
-	for(size_t i=0; i!=gpuBoxCount; ++i)
-	{
-		size_t gpuBoxIndex = subbandToGPUBoxFileIndex[i] + 1;
-		name[numberPos] = (char) ('0' + (gpuBoxIndex/10));
-		name[numberPos+1] = (char) ('0' + (gpuBoxIndex%10));
-		_files[i] = new std::ofstream(name.c_str(), std::ios_base::out | std::ios_base::binary | std::ios_base::trunc);
-		if(!_files[i]->good())
-			throw std::runtime_error(std::string("Could not open flag file \"") + name + "\" for writing.");
-	}
-}
-
-FlagWriter::~FlagWriter()
-{
-	for(std::vector<std::ofstream*>::iterator i=_files.begin(); i!=_files.end(); ++i)
-		delete *i;
-}
-
-void FlagWriter::writeHeader()
-{
-	struct Header header;
-	header.fileIdentifier[0] = 'M';
-	header.fileIdentifier[1] = 'W';
-	header.fileIdentifier[2] = 'A';
-	header.fileIdentifier[3] = 'F';
-	header.versionMajor = VERSION_MAJOR;
[…]

### 2014-09-11 altcotter/flagwriter.h

@@ -1,92 +1,0 @@
-#ifndef FLAG_WRITER_H
-#define FLAG_WRITER_H
-
-#include "writer.h"
-
-#include <stdint.h>
-
-#include <fstream>
-#include <iostream>
-
-class FlagWriter : public Writer
-{
-	public:
-		FlagWriter(const std::string &filename, int gpsTime, size_t timestepCount, size_t gpuBoxCount, const std::vector<size_t>& subbandToGPUBoxFileIndex);
-		
-		~FlagWriter();
-		
-		void WriteBandInfo(const std::string &name, const std::vector<Writer::ChannelInfo> &channels, double refFreq, double totalBandwidth, bool flagRow)
-		{
-			_channelCount = channels.size();
-			setStride();
-		}
-		
-		void WriteAntennae(const std::vector<Writer::AntennaInfo> &antennae, double time)
-		{
-			_antennaCount = antennae.size();
-			setStride();
-		}
-		
-		void WritePolarizationForLinearPols(bool flagRow)
-		{
-			_polarizationCount = 4;
-			setStride();
-		}
-		
-		void WriteSource(const Writer::SourceInfo &source)
-		{
-		}
-		
-		void WriteField(const Writer::FieldInfo& field)
-		{
-		}
-		
-		void WriteObservation(const std::string& telescopeName, double startTime, double endTime, const std::string& observer, const std::string& scheduleType, const std::string& project, double releaseDate, bool flagRow)
-		{
-		}
-		
-		void AddRows(size_t rowCount)
-		{
-			if(_rowCount == 0)
-				writeHeader();
-			_rowCount += rowCount;
-		}
-		
-		void WriteRow(double time, double timeCentroid, size_t antenna1, size_t antenna2, double u, double v, double w, double interval, const std::complex<float>* data, const bool* flags, const float *weights)
-		{
-			writeRow(antenna1, antenna2, flags);
-		}
-		
-		void WriteHistoryItem(const std::string &commandLine, const std::string &application, const std::vector<std::string> &params)
-		{
-		}
-		
-		bool IsTimeAligned(size_t antenna1, size_t antenna2)
-		{
-			return true;
-		}
-	private:
-		void writeHeader();
-		void writeRow(size_t antenna1, size_t antenna2, const bool* flags);
-		void setStride();
-		static void pack(unsigned char *output, const unsigned char *input, size_t count);
[…]

### 2014-09-11 altcotter/gpufilereader.cpp

@@ -1,444 +1,0 @@
-#include "gpufilereader.h"
-#include "progressbar.h"
-
-#include <boost/thread/thread.hpp>
-
-#include <complex>
-#include <iostream>
-#include <sstream>
-#include <stdexcept>
-
-void GPUFileReader::openFiles()
-{
-	int status = 0;
-	bool hasStartTime = false, hasWarnedAboutDifferentTimes = false;
-	std::vector<long> startTimePerFile(_filenames.size());
-	for(size_t i=0; i!=_filenames.size(); ++i)
-	{
-		const std::string &curFilename = _filenames[i];
-		fitsfile *fptr = 0;
-		if(curFilename.empty())
-		{
-			std::cout << "(Skipping unavailable file)\n";
-			_fitsFiles.push_back(0);
-			_fitsHDUCounts.push_back(0);
-		}
-		else if(!fits_open_file(&fptr, curFilename.c_str(), READONLY, &status))
-		{
-			_fitsFiles.push_back(fptr);
-			
-			int hduCount;
-			fits_get_num_hdus(fptr, &hduCount, &status);
-			checkStatus(status);
-			
-			_fitsHDUCounts.push_back(hduCount);
-			std::cout << "There are " << hduCount << " HDUs in file " << _filenames[i] << '\n';
-			
-			long thisFileTime;
-			fits_read_key(fptr, TLONG, "TIME", &thisFileTime, 0, &status);
-			checkStatus(status);
-			
-			if(!hasStartTime) 
-			{
-				_startTime = thisFileTime;
-				hasStartTime = true;
-			}
-			startTimePerFile[i] = thisFileTime;
-			if(_startTime != thisFileTime)
-			{
-				if(!hasWarnedAboutDifferentTimes || thisFileTime > _startTime)
-				{
-					std::cout << "WARNING: file number " << (i+1) << " of current time range has different start time!\n"
-						"Current file start time: " << thisFileTime << " previous file had: " << _startTime << ".\n";
-				}
-				if(thisFileTime > _startTime)
-				{
-					_startTime = thisFileTime;
-					if(_doAlign)
-						std::cout << "Using start time of " << _startTime << " and aligning other files accordingly.\n";
-					else
-						std::cout << "Using start time of " << _startTime << " and NOT aligning files accordingly as requested:\n"
[…]

### 2014-09-11 altcotter/gpufilereader.h

@@ -1,136 +1,0 @@
-#include <string>
-#include <vector>
-#include <ctime>
-#include <complex>
-
-#include <fitsio.h>
-#include <stdexcept>
-
-#include "baselinebuffer.h"
-#include "fitsuser.h"
-#include "lane.h"
-
-/**
- * The GPU file reader, that can read the files produced by the MWA correlator.
- * Format based on reader in build_lfiles.c by sord.
- * 
- * To use this class:
- * - Construct it
- * - Add all GPU files to the reader that belong to the observation with equal time range
- *   by calling AddFile()
- * - Then, call Initialize() to read the required metadata, such as the antenna count.
- * - Allocate destination buffers and set them with SetDestBaselineBuffer()
- * - Set the input-to-antenna mapping for all inputs with SetCorrInputToOutput().
- * - Finally, call Read() to start reading all data.
- * 
- * The MWA correlator produces multiple GPU files that need to be combined. Each
- * GPU file contains a selected number of channels of each scan.
- * The elements inside each scan are ordered by
- * polarization2, polarization1, station2, station1, channel, where polarization2
- * is the index that is moving fastest.
- * station2 is the non-conjugated station within a baseline.
- * 
- * @author André Offringa
- */
-class GPUFileReader : private FitsUser
-{
-	public:
-		GPUFileReader(size_t nAntenna, size_t nChannelsInTotal, size_t threadCount) :
-			_shuffleTasks(threadCount),
-			_availableGPUMatrixBuffers(threadCount),
-			_isOpen(false),
-			_nAntenna(nAntenna),
-			_nChannelsInTotal(nChannelsInTotal),
-			_bufferSize(0),
-			_currentHDU(0),
-			_stopHDU(0),
-			_threadCount(threadCount),
-			_integrationTime(0.0),
-			_doAlign(true)
-		{ }
-		~GPUFileReader() { closeFiles(); }
-		
-		void AddFile(const char *filename) { _filenames.push_back(std::string(filename)); }
-		
-		void Initialize(double integrationTime, bool doAlign) {
-			_buffers.resize(_nAntenna * _nAntenna);
-			_mappedBuffers.resize(_nAntenna * _nAntenna);
[…]

### 2014-09-11 altcotter/lane.h

@@ -1,209 +1,0 @@
-#ifndef LANE_H
-#define LANE_H
-
-#include <cstring>
-#include <deque>
-
-#include <boost/thread/mutex.hpp>
-#include <boost/thread/condition.hpp>
-
-template<typename T>
-class lane : public boost::noncopyable
-{
-	public:
-		typedef T data_type;
-		
-		lane(size_t capacity)
-		:
-			_capacity(capacity),
-			_buffer(new T[capacity]),
-			_write_position(0),
-			_free_write_space(_capacity),
-			_status(status_normal)
-		{
-		}
-		
-		~lane()
-		{
-			delete[] _buffer;
-		}
-		
-		void reset()
-		{
-			_write_position = 0;
-			_free_write_space = _capacity;
-			_status = status_normal;
-		}
-		
-		void write(const data_type &element)
-		{
-			write(&element, 1);
-		}
-		
-		void write(const data_type *elements, size_t n)
-		{
-			boost::mutex::scoped_lock lock(_mutex);
-			
-			if(_status == status_normal)
-			{
-				size_t write_size = _free_write_space > n ? n : _free_write_space;
-				immediate_write(elements, write_size);
-				n -= write_size;
-				
-				while(n != 0) {
-					elements += write_size;
-				
-					do {
-						_has_work_condition.wait(lock);
-					} while(_free_write_space == 0 && _status == status_normal);
-					
-					write_size = _free_write_space > n ? n : _free_write_space;
-					immediate_write(elements, write_size);
-					n -= write_size;
-				} while(n != 0);
-			}
-		}
-		
-		bool read(data_type &destination)
-		{
-			return read(&destination, 1) == 1;
-		}
-		
-		size_t read(data_type *destinations, size_t n)
-		{
-			size_t n_left = n;
-			
-			boost::mutex::scoped_lock lock(_mutex);
-			
-			size_t free_space = free_read_space();
-			size_t read_size = free_space > n ? n : free_space;
-			immediate_read(destinations, read_size);
-			n_left -= read_size;
-			
-			while(n_left != 0 && _status == status_normal)
-			{
-				destinations += read_size;
-				
-				do {
-					_has_work_condition.wait(lock);
-				} while(free_read_space() == 0 && _status == status_normal);
-				
-				free_space = free_read_space();
-				read_size = free_space > n_left ? n_left : free_space;
-				immediate_read(destinations, read_size);
-				n_left -= read_size;
-			}
-			return n - n_left;
-		}
-		
-		void write_end()
-		{
[…]

### 2014-09-11 altcotter/main.cpp

@@ -1,400 +1,0 @@
-#include "cotter.h"
-#include "numberlist.h"
-#include "radeccoord.h"
-#include "version.h"
-
-#include <aoflagger.h>
-
-#include <boost/algorithm/string.hpp>
-
-#include <iostream>
-#include <vector>
-
-bool isDigit(char c)
-{
-	return c >= '0' && c <= '9';
-}
-
-void checkInputFilename(const std::string &filename)
-{
-	if(filename.size() < 5 || filename.substr(filename.size()-5)!=".fits")
-		throw std::runtime_error("The raw GPU input files should have a .fits extension");
-	if(filename.size() < 10 || filename[filename.size()-8] != '_')
-		throw std::runtime_error("The raw GPU input filenames should end in '...nn_nn.fits'");
-	if(!isDigit(filename[filename.size()-9]) || !isDigit(filename[filename.size()-10]))
-		throw std::runtime_error("Could not parse gpu box number -- the raw GPU input filenames should end in '...nn_nn.fits'");
-	if(!isDigit(filename[filename.size()-6]) || !isDigit(filename[filename.size()-7]))
-		throw std::runtime_error("Could not parse time step number -- the raw GPU input filenames should end in '...nn_nn.fits'");
-}
-
-size_t timeStepNumberFromFilename(const std::string &filename)
-{
-	return (filename[filename.size()-7]-'0')*10 + (filename[filename.size()-6]-'0');
-}
-
-size_t gpuBoxNumberFromFilename(const std::string &filename)
-{
-	size_t num = (filename[filename.size()-10]-'0')*10 + (filename[filename.size()-9]-'0');
-	if(num == 0)
-		throw std::runtime_error("Could not parse gpu box number in filename (it was zero)");
-	return num-1;
-}
-
-bool isFitsFile(const std::string &filename)
-{
-	if(filename.size() > 7)
-	{
-		return boost::to_upper_copy(filename.substr(filename.size()-7)) == ".UVFITS";
-	}
-	return false;
-}
-
-bool isMWAFlagFile(const std::string &filename)
-{
-	if(filename.size() > 5)
-	{
-		return boost::to_upper_copy(filename.substr(filename.size()-5)) == ".MWAF";
-	}
-	return false;
-}
-
-void usage()
-{
-	std::cout << "usage: cotter [options] <gpufiles> \n"
-	"Options:\n"
[…]

### 2014-09-11 altcotter/metafitsfile.cpp

@@ -1,355 +1,0 @@
-#include "metafitsfile.h"
-#include "mwaconfig.h"
-#include "geometry.h"
-
-#include <iostream>
-#include <stdexcept>
-#include <sstream>
-#include <cmath>
-
-// Bug in cfitsio < 3.31
-#ifndef __CINT__
-#ifdef __cplusplus
-extern "C" {
-#endif
-#endif
-int fffree(void *value, int *status); 
-#ifndef __CINT__
-#ifdef __cplusplus
-}
-#endif
-#endif
-
-MetaFitsFile::MetaFitsFile(const char* filename)
-{
-	int status = 0;
-	if(fits_open_file(&_fptr, filename, READONLY, &status))
-		throwError(status, std::string("Cannot open file ") + filename);
-	
-	int hduCount;
-	fits_get_num_hdus(_fptr, &hduCount, &status);
-	checkStatus(status);
-	
-	if(hduCount < 2)
-		throw std::runtime_error("At least two HDUs are expected to be in the meta fitsfile");
-}
-
-MetaFitsFile::~MetaFitsFile()
-{
-	int status = 0;
-	if(fits_close_file(_fptr, &status))
-		throwError(status, "Could not close metafits file");
-}
-
-void MetaFitsFile::ReadHeader(MWAHeader& header, MWAHeaderExt &headerExt)
-{
-	int status = 0, hduType;
-	fits_movabs_hdu(_fptr, 1, &hduType, &status);
-	checkStatus(status);
-	
-	int keywordCount;
-	fits_get_hdrspace(_fptr, &keywordCount, 0, &status);
-	checkStatus(status);
-	
-	headerExt.tilePointingDecRad = -99.0;
-	headerExt.tilePointingRARad = -99.0;
-	header.raHrs = -99.0;
-	header.decDegs = -99.0;
-	
-	for(int i=0; i!=keywordCount; ++i)
-	{
-		char keyName[80], keyValue[80], *keyValueLong;
-		fits_read_keyn(_fptr, i+1, keyName, keyValue, 0, &status);
-		checkStatus(status);
-		
-		std::string keyValueStr = keyValue;
-		if(keyValueStr.size()>=3  && keyValueStr[0]=='\'' && keyValueStr[keyValueStr.size()-1]=='\'' && keyValueStr[keyValueStr.size()-2]=='&')
-		{
-			fits_read_key_longstr(_fptr, keyName, &keyValueLong, NULL, &status);
[…]

### 2014-09-11 altcotter/mswriter.cpp

@@ -1,490 +1,0 @@
-#include "mswriter.h"
-
-#include <ms/MeasurementSets/MeasurementSet.h>
-
-#include <tables/Tables/ArrayColumn.h>
-#include <tables/Tables/ArrColDesc.h>
-#include <tables/Tables/ScalarColumn.h>
-#include <tables/Tables/SetupNewTab.h>
-
-#include <measures/TableMeasures/TableMeasDesc.h>
-
-#include <measures/Measures/MFrequency.h>
-
-using namespace casa;
-
-class MSWriterData
-{
-	public:
-		MeasurementSet *ms;
-		
-		ScalarColumn<double> *_timeCol, *_timeCentroidCol;
-		ScalarColumn<int> *_antenna1Col, *_antenna2Col;
-		ScalarColumn<int> *_dataDescIdCol;
-		ArrayColumn<double> *_uvwCol;
-		ScalarColumn<double> *_intervalCol, *_exposureCol;
-		ScalarColumn<int> *_processorIdCol, *_scanNumberCol, *_stateIdCol;
-		ArrayColumn<std::complex<float> > *_dataCol;
-		ArrayColumn<bool> *_flagCol;
-		ArrayColumn<float> *_sigmaCol;
-		ArrayColumn<float> *_weightCol;
-		ArrayColumn<float> *_weightSpectrumCol;
-		
-		casa::Vector<double> *_uvwVec;
-		casa::Vector<float> *_sigmaArr;
-		casa::Array<std::complex<float> > *_dataArr;
-		casa::Array<bool> *_flagArr;
-		casa::Array<float> *_weightSpectrumArr;
-		casa::Vector<float> *_weightsArr;
-};
-
-MSWriter::MSWriter(const std::string& filename) :
-	_data(new MSWriterData),
-	_rowIndex(0),
-	_nChannels(0)
-{
-	TableDesc tableDesc = MS::requiredTableDesc();
-	ArrayColumnDesc<std::complex<float> > dataColumnDesc = ArrayColumnDesc<std::complex<float> >(MS::columnName(casa::MSMainEnums::DATA));
-	tableDesc.addColumn(dataColumnDesc);
-	ArrayColumnDesc<float> weightSpectrumColumnDesc = ArrayColumnDesc<float>(MS::columnName(casa::MSMainEnums::WEIGHT_SPECTRUM));
-	tableDesc.addColumn(weightSpectrumColumnDesc);
-	SetupNewTable newTab(filename, tableDesc, Table::New);
[…]

### 2014-09-11 altcotter/mwaconfig.h

@@ -1,150 +1,0 @@
-#ifndef MWA_CONFIG_H
-#define MWA_CONFIG_H
-
-#include <vector>
-#include <cstring>
-#include <string>
-#include <map>
-
-#include "mwainput.h"
-
-struct MWAHeader
-{
-	public:
-		MWAHeader();
-		
-		size_t nInputs;                 // the number of inputs to the correlator
-		size_t nScans;                  // the number of time samples
-		size_t nChannels;               // the number of spectral channels
-		enum CorrType { None, CrossCorrelation, AutoCorrelation, BothCorrelations } correlationType;
-		double integrationTime;         // per time sample, in seconds
-		double centralFrequencyMHz;     // observing central frequency and 
-		double bandwidthMHz;            // bandwidth (MHz)
-		double raHrs, decDegs;          // ra,dec of phase centre.
-		double haHrsStart;              // the HA of the phase center at the start of the integration
-		double refEl, refAz;            // the el/az of the normal to the plane of the array (radian)
-		int    year, month, day;        // date/time in UTC.
-		int    refHour, refMinute;
-		double refSecond;
-		//bool   invertFrequency;         // flag to indicate that freq decreases with increasing channel number.
-		bool   conjugate;               // conjugate the final vis to correct for any sign errors
-		bool   geomCorrection;          // apply geometric phase correction
-		std::string fieldName;
-		std::string polProducts;
-		
-		double dateFirstScanMJD;        // The central time of the first time step. Not in header file, derived from it.
-		
-		double GetDateLastScanMJD() const
-		{
-			return dateFirstScanMJD + (integrationTime/86400.0)*nScans;
-		}
-		double GetStartDateMJD() const;
-		double GetDateFirstScanFromFields() const;
-		
-		void Validate(bool lockPointing);
-	private:
-		MWAHeader(const MWAHeader &) { }
-		void operator=(const MWAHeader &) { }
-};
-
-struct MWAHeaderExt
-{
-	MWAHeaderExt();
-	
-	int gpsTime;
-	std::string observerName, projectName, gridName, mode, filename, mwaPyVersion, mwaPyDate, metaDataVersion;
-	int delays[16], subbandGains[24], subbandNumbers[24];
-	bool hasCalibrator, hasGlobalSubbandGains;
[…]

### 2014-09-11 altcotter/mwafits.h

@@ -1,48 +1,0 @@
-#ifndef MWA_FITS_H
-#define MWA_FITS_H
-
-#include <string>
-
-/**
- * This module writes the extended MWA keywords into uvfits files
- */
-
-struct MWAFitsEnums
-{
-	enum MWAKeywords
-	{
-		MWA_FIBER_VEL_FACTOR, MWA_METADATA_VERSION,
-		MWA_MWAPY_VERSION,
-		MWA_COTTER_VERSION, MWA_COTTER_VERSION_DATE
-	};
-};
-
-class MWAFits
-{
-	public:
-		MWAFits(const std::string &filename);
-		
-		~MWAFits();
-		
-		void WriteMWAKeywords(double fibreVelFactor, const std::string& metaDataVersion, const std::string& mwaPyVersion, const std::string& cotterVersion, const std::string& cotterVersionDate);
-		
-	private:
-		static const char* keywordName(enum MWAFitsEnums::MWAKeywords keyword)
-		{
-			return _keywordNames[(int) keyword];
-		}
-		static const char* keywordComment(enum MWAFitsEnums::MWAKeywords keyword)
-		{
-			return _keywordComments[(int) keyword];
-		}
-		template<typename T>
-		void writeKey(enum MWAFitsEnums::MWAKeywords keyword, const T& value);
-		
-		static const char* _keywordNames[];
-		static const char* _keywordComments[];
-		
-		const std::string _filename;
-		struct MWAFitsData *_data;
-};
-
-#endif

### 2014-09-11 altcotter/progressbar.cpp

@@ -1,40 +1,0 @@
-#include "progressbar.h"
-
-#include <iostream>
-
-ProgressBar::ProgressBar(const std::string& taskDescription) :
-	_taskDescription(taskDescription),
-	_displayedDots(0)
-{
-	std::cout << taskDescription << ":";
-	if(taskDescription.size() < 40)
-		std::cout << " 0%" << std::flush;
-	else
-		std::cout << "\n 0%" << std::flush;
-}
-
-ProgressBar::~ProgressBar()
-{
-	SetProgress(1,1);
-}
-
-void ProgressBar::SetProgress(size_t taskIndex, size_t taskCount)
-{
-	unsigned progress = (taskIndex * 100 / taskCount);
-	unsigned dots = progress / 2;
-	
-	if(dots > _displayedDots)
-	{
-		while(dots != _displayedDots)
-		{
-			++_displayedDots;
-			if(_displayedDots % 5 == 0)
-				std::cout << ((_displayedDots/5)*10) << '%';
-			else
-				std::cout << '.';
-		}
-		if(progress == 100)
-			std::cout << '\n';
-		std::cout << std::flush;
-	}
-}

### 2014-09-11 altcotter/slalib.h

@@ -1,500 +1,0 @@
-#ifndef SLALIBHDEF
-#define SLALIBHDEF
-
-#ifdef __cplusplus
-extern "C" {
-#endif
-
-#include <math.h>
-
-/*
-**  - - - - - - - - -
-**   s l a l i b . h
-**  - - - - - - - - -
-**
-**  Prototype function declarations for slalib library.
-**
-**  Last revision:   5 April 2005
-**
-**  Copyright P.T.Wallace.  All rights reserved.
-*/
-
-void slaAddet ( double rm, double dm, double eq, double *rc, double *dc );
-
-void slaAfin ( char *string, int *iptr, float *a, int *j );
-
-double slaAirmas ( double zd );
-
-void slaAltaz ( double ha, double dec, double phi,
-                double *az, double *azd, double *azdd,
-                double *el, double *eld, double *eldd,
-                double *pa, double *pad, double *padd );
-
-void slaAmp ( double ra, double da, double date, double eq,
-              double *rm, double *dm );
-
-void slaAmpqk ( double ra, double da, double amprms[21],
-                double *rm, double *dm );
-
-void slaAop ( double rap, double dap, double date, double dut,
-              double elongm, double phim, double hm, double xp,
-              double yp, double tdk, double pmb, double rh,
-              double wl, double tlr,
-              double *aob, double *zob, double *hob,
-              double *dob, double *rob );
-
-void slaAoppa ( double date, double dut, double elongm, double phim,
-                double hm, double xp, double yp, double tdk, double pmb,
-                double rh, double wl, double tlr, double aoprms[14] );
-
-void slaAoppat ( double date, double aoprms[14] );
-
-void slaAopqk ( double rap, double dap, double aoprms[14],
-                double *aob, double *zob, double *hob,
-                double *dob, double *rob );
-
-void slaAtmdsp ( double tdk, double pmb, double rh, double wl1,
-                 double a1, double b1, double wl2, double *a2, double *b2 );
-
-void slaAv2m ( float axvec[3], float rmat[3][3] );
-
-float slaBear ( float a1, float b1, float a2, float b2 );
-
-void slaCaf2r ( int ideg, int iamin, float asec, float *rad, int *j );
-
-void slaCaldj ( int iy, int im, int id, double *djm, int *j );
-
-void slaCalyd ( int iy, int im, int id, int *ny, int *nd, int *j );
-
[…]

### 2014-09-11 altcotter/stopwatch.h

@@ -1,61 +1,0 @@
-/***************************************************************************
- *   Copyright (C) 2008 by A.R. Offringa   *
- *   offringa@astro.rug.nl   *
- *                                                                         *
- *   This program is free software; you can redistribute it and/or modify  *
- *   it under the terms of the GNU General Public License as published by  *
- *   the Free Software Foundation; either version 2 of the License, or     *
- *   (at your option) any later version.                                   *
- *                                                                         *
- *   This program is distributed in the hope that it will be useful,       *
- *   but WITHOUT ANY WARRANTY; without even the implied warranty of        *
- *   MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the         *
- *   GNU General Public License for more details.                          *
- *                                                                         *
- *   You should have received a copy of the GNU General Public License     *
- *   along with this program; if not, write to the                         *
- *   Free Software Foundation, Inc.,                                       *
- *   59 Temple Place - Suite 330, Boston, MA  02111-1307, USA.             *
- ***************************************************************************/
-#ifndef STOPWATCH_H
-#define STOPWATCH_H
-
-#include <boost/date_time/posix_time/posix_time_types.hpp>
-
-#include <string>
-
-#ifndef DOXYGEN_SHOULD_SKIP_THIS
-
-/**
-	@author A.R. Offringa <offringa@astro.rug.nl>
-*/
-class Stopwatch{
-	public:
-		Stopwatch();
-		Stopwatch(bool start);
-		~Stopwatch();
-		void Start();
-		void Pause();
-		void Reset();
-
-		std::string ToString() const;
-		std::string ToShortString() const;
-		long double Seconds() const;
-		
-		std::string ToDaysString() const;
-		std::string ToHoursString() const;
-		std::string ToMinutesString() const;
-		std::string ToSecondsString() const;
-		std::string ToMilliSecondsString() const;
-		std::string ToMicroSecondsString() const;
-		std::string ToNanoSecondsString() const;
-	private:
-		bool _running;
-		boost::posix_time::ptime _startTime;
-		boost::posix_time::time_duration _sum;
-};
-
-// end of Doxygen skip
-#endif
-
-#endif

### 2014-09-11 altcotter/subbandpassband.h

@@ -1,23 +1,0 @@
-#ifndef SUBBAND_PASSBAND_H
-#define SUBBAND_PASSBAND_H
-
-#include <vector>
-
-class SubbandPassband
-{
-	public:
-		static void GetSubbandPassband(std::vector<double> &passband, std::size_t channelsPerSubband);
-		
-	private:
-		/**
-		 * See the cpp file for explanation of the difference between these bandpasses.
-		 */
-		const static double _sb128ChannelSubbandValue2013[128];
-		const static double _sb128ChannelSubbandValue2014FromMemo[128];
-		
-		const static double *_sb128ChannelSubbandValue;
-		
-		static double getSubbandValue(std::size_t channel, std::size_t channelCount);
-};
-
-#endif

### 2014-09-11 altcotter/threadedwriter.cpp

@@ -1,100 +1,0 @@
-#include "threadedwriter.h"
-
-#include <boost/mem_fn.hpp>
-
-ThreadedWriter::ThreadedWriter(Writer *parentWriter) : ForwardingWriter(parentWriter),
-	_isWriterReady(false),
-	_isBufferReady(false),
-	_isFinishing(false),
-	_bufferedData(0),
-	_bufferedFlags(0),
-	_bufferedWeights(0),
-	_thread(&ThreadedWriter::writerThreadFunc, this)
-{
-}
-
-ThreadedWriter::~ThreadedWriter()
-{
-	boost::mutex::scoped_lock lock(_mutex);
-	_isFinishing = true;
-	lock.unlock();
-	
-	_bufferChangeCondition.notify_all();
-	_thread.join();
-	
-	delete[] _bufferedData;
-	delete[] _bufferedFlags;
-	delete[] _bufferedWeights;
-}
-
-void ThreadedWriter::WriteBandInfo(const std::string &name, const std::vector<Writer::ChannelInfo> &channels, double refFreq, double totalBandwidth, bool flagRow)
-{
-	_arraySize = channels.size() * 4;
-	_bufferedData = new std::complex<float>[_arraySize];
-	_bufferedFlags = new bool[_arraySize];
-	_bufferedWeights = new float[_arraySize];
-	
-	ForwardingWriter::WriteBandInfo(name, channels, refFreq, totalBandwidth, flagRow);
-}
-
-void ThreadedWriter::AddRows(size_t rowCount)
-{
-	boost::mutex::scoped_lock lock(_mutex);
-	
-	// Wait until the writer is ready AND the buffer is empty
-	while(!_isWriterReady || _isBufferReady)
-		_bufferChangeCondition.wait(lock);
-	
-	// Just keep mutex locked (might take time, but this method is not called so often...)
-	ParentWriter().AddRows(rowCount);
-}
-
-void ThreadedWriter::WriteRow(double time, double timeCentroid, size_t antenna1, size_t antenna2, double u, double v, double w, double interval, const std::complex<float>* data, const bool* flags, const float *weights)
-{
-	boost::mutex::scoped_lock lock(_mutex);
-	
-	// Wait until the writer is ready AND the buffer is empty (=not ready)
-	while(!_isWriterReady || _isBufferReady)
-		_bufferChangeCondition.wait(lock);
-	
[…]

### 2014-09-11 altcotter/version.h

@@ -1,9 +1,0 @@
-#ifndef COTTER_VERSION_H
-#define COTTER_VERSION_H
-
-#define COTTER_VERSION_STR  "1.6"
-#define COTTER_VERSION_MAJOR 1
-#define COTTER_VERSION_MINOR 6
-#define COTTER_VERSION_DATE "2014-05-07"
-
-#endif

### 2014-09-11 altcotter/writer.h

@@ -1,83 +1,0 @@
-#ifndef WRITER_H
-#define WRITER_H
-
-#include <string>
-#include <vector>
-#include <complex>
-
-class Writer
-{
-	public:
-		struct AntennaInfo
-		{
-			AntennaInfo() { }
-			AntennaInfo(const AntennaInfo& source);
-			AntennaInfo &operator=(const AntennaInfo& source);
-			
-			std::string name, station;
-			std::string type, mount;
-			double x, y, z;
-			double diameter;
-			bool flag;
-		};
-		
-		struct ChannelInfo
-		{
-			double chanFreq, chanWidth, effectiveBW, resolution;
-		};
-		
-		struct SourceInfo
-		{
-			SourceInfo() { }
-			int sourceId;
-			double time, interval;
-			int spectralWindowId, numLines;
-			std::string name;
-			int calibrationGroup;
-			std::string code;
-			double directionRA, directionDec;
-			double properMotion[2];
-		private:
-			SourceInfo(const SourceInfo&) { }
-			void operator=(const SourceInfo&) { }
-		};
-		
-		struct FieldInfo
-		{
-			FieldInfo() { }
-			std::string name, code;
-			double time;
-			int numPoly;
-			double delayDirRA, delayDirDec;
-			double phaseDirRA, phaseDirDec;
-			double referenceDirRA, referenceDirDec;
-			int sourceId;
-			bool flagRow;
-		private:
-			FieldInfo(const FieldInfo&) { }
-			void operator=(const FieldInfo&) { }
-		};
-		
-		virtual ~Writer() { }
-		virtual void SetArrayLocation(double x, double y, double z) { }
-		virtual void WriteBandInfo(const std::string &name, const std::vector<ChannelInfo> &channels, double refFreq, double totalBandwidth, bool flagRow) = 0;
-		virtual void WriteAntennae(const std::vector<AntennaInfo> &antennae, double time) = 0;
-		virtual void WritePolarizationForLinearPols(bool flagRow) = 0;
-		virtual void WriteSource(const SourceInfo &source) = 0;
-		virtual void WriteField(const FieldInfo& field) = 0;
-		virtual void WriteObservation(const std::string& telescopeName, double startTime, double endTime, const std::string& observer, const std::string& scheduleType, const std::string& project, double releaseDate, bool flagRow) = 0;
-		virtual void WriteHistoryItem(const std::string &commandLine, const std::string &application, const std::vector<std::string> &params) = 0;
-		
-		virtual void AddRows(size_t count) = 0;
[…]

### 2014-09-11 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Work

@@
 Additionally, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume.
-It also collects some statistics and writes these into a measurement set with the same format as used by LOFAR\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}.
+It also collects some statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}.
 Various tools are available to analyse these statistics, e.g.
-the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot these over various dimensions in an interactive manner.
+the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.
 
@@
 Because not all astronomers might have easily access to high-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations.
+
+When time or frequency averaging is requested, \textsc{cotter} averages samples together that have not been flagged.
+When all input samples are flagged, the average of all input visibilities is stored into the the output sample, and the output sample is flagged.
+This method makes it possible to superficially analyse flagged samples in the output, even though information is lost in the averaging.
+\textsc{cotter} stores a weight for each visibility in the output file, which is scaled to the number of unflagged input samples that were used for the output sample.
+Because of this, when no averaging is requested, the output is 50\% larger than the input (one extra float per complex float visibility).
+In practice, most observations are recorded at a resolution of 0.5 s and 40 kHz, and averaged to a resolution of 2--4 s and 40--80 kHz to reduce the data volume.
+This does not significantly cause any time or bandwidth decorrelation up to the first beam nul.
+\textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $uvw$ values for the central time and frequency of the output sample.
+This helps to prevent time/frequency decorrelation.
 

### 2014-09-15 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Tada
Stats... etc

@@
 \usepackage{xspace}
-%\usepackage{subcaption}
+\usepackage{subcaption}
 
@@
 LOFAR's \textsc{aoflagger} has shown to have a good accuracy, is fast and has a library interface, which allows it to be integrated in a pipeline.
-It is therefore also used for MWA's RFI detection.
+It is therefore used for MWA's RFI detection.
 
@@
 EoR   & 2013-08-26 & 138.8--197.7              & 6h & 0.81\% & ---\\
-GLEAM & 2013-11-05 & 72.3--133.8, 138.8--230.8 & 7h & 1.30\% & 0\% \\
+GLEAM & 2013-11-05 & 72.3--133.8, 138.8--230.8 & 7h & 1.27\% & 0\% \\
 GLEAM & 2013-11-25 & 72.3--133.8, 138.8--230.8 & 7h & 0.69\% & 0\% \\
@@
 \hline
-\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.97\%} & \textbf{0.5\%}\\
+\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{...\%} & \textbf{0.5\%}\\
 \hline
@@
 Because the MWA has a 30.72~MHz instantaneous bandwidth, this range has been split up so that each night observes with a central frequency of either 154.2 or 182.4~MHz, with an overlap of 2.6 MHz.
-In these observations, the selected EoR field is tracked byt the primary beam, such that the sensitivity towards the field is maximal.
+In these observations, the selected EoR field is tracked by the primary beam, such that the sensitivity towards the field is maximal.
 
@@
 
-Both projects avoid the subbands that cover the frequency range 133.8--138.8~MHz, because the ORBCOMM low Earth orbitting (LEO) satellites transmit in these subbands.
+Both projects avoid the subbands that cover the frequency range 133.8--138.9~MHz, because the ORBCOMM low Earth orbitting (LEO) satellites transmit in these subbands.
 When these frequencies are observed, the subbands are often so strongly contaminated that it affects imaging sensitivity.
@@
 \noindent\begin{figure*}
+\subcaptionbox{RFI contamination found in the 2 m amateur band (146 MHz).}[\linewidth]{%
 % obs data: 1061583576, 2013-08-26.
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/amateur-2m-band-example}\includegraphics[width=9.1cm]{img/amateur-2m-band-flagged}
-\caption{Strong RFI contamination found in the 2 m amateur band (146 MHz), which falls in the lower band of the MWA EoR observations.
+\includegraphics[width=9.1cm]{img/amateur-2m-band-example}\includegraphics[width=9.1cm]{img/amateur-2m-band-flagged}
+\label{fig:amateur-2m}
+}\\\vspace{3mm}%
+\subcaptionbox{Short RFI burst of 2 s centred on 150.17 MHz. This shows that flagging at high time and frequency resolution is necessary to accurately detect RFI.}[\linewidth]{%
+% obs data: 1061583576, 2013-08-26.
+\includegraphics[width=9.1cm]{img/150_2_mhz_example}\includegraphics[width=9.1cm]{img/150_2_mhz_flagged}
+\label{fig:150_2}
+}\\\vspace{3mm}%
+\subcaptionbox{A weakly-observed transmitter centred on 156.66 MHz.
+This same frequency is flagged in several correlations, and it can therefore be assumed this is real RFI.
+This is an example in which the AOFlagger detects an event that would not be detected by regular thresholding.
+}[\linewidth]{%
+% obs data: 1061583576, 2013-08-26.
+\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}
+\label{fig:156_7}
+}%
+\caption{Examples of RFI events found in the EoR low band (138.9--197.7 MHz) in a 2-min snapshot with relatively high RFI contamination.
 These panels show the Stokes I amplitudes.
-In the right panel, the result of RFI detection is shown with purple.
-The thin horizontal lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.
-Two more transmitters are detected in these data: at 150.17 and 156.7 Mhz, which are both one 40 kHz channel wide.
-These transmitters are not distinguishable in this plot, but are shown in the next figures.}
-\label{fig:amateur-2m}
-\end{center}
-\end{figure*}
-
-\noindent\begin{figure*}
-% obs data: 1061583576, 2013-08-26.
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/150_2_mhz_example}\includegraphics[width=9.1cm]{img/150_2_mhz_flagged}
-\caption{Short RFI burst of 2 s centred on 150.17 MHz. This shows that flagging at high time and frequency resolution is necessary to accurately detect RFI.}
-\label{fig:150_2}
-\end{center}
-\end{figure*}
-
-\noindent\begin{figure*}
-% obs data: 1061583576, 2013-08-26.
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}
-\caption{A weakly-observed transmitter centred on 156.66 MHz.
-Although it is hard to identify in such a time-frequency plot, the same frequency is flagged in several correlations, and it can therefore be assumed this is real RFI.
-This is an example in which the AOFlagger detect an event that would not be detected by regular thresholding.
-}
-\label{fig:156_7}
-\end{center}
+In the right figures, the result of RFI detection is shown with purple.
+The horizontally flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
 \end{figure*}
@@
 \label{fig:dvb-stddevs}
+\end{center}
+\end{figure}
+
+\noindent\begin{figure}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/rfi-per-band}
+\caption{RFI per subband}
+\label{fig:rfi-per-band}
 \end{center}

### 2014-09-28 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Updating data in log-n log-s plots and some other minor changes

@@
 \caption{RFI from Digital Video Broadcasting (DVB).
-Because of its broadband nature, this kind of RFI is not well flagged in the initial AOFlagger step.
-It is detected by a later step by checking the global statistics.
+Because of its broadband nature, this kind of RFI is not well flagged in the initial AOFlagger step, but is detectable in the global statistics.
 This type of RFI makes most of the 30.72~MHz bandwidth unusable.
-It is not seen very often (1 out of 6 observed nights).
+It is visible for about 2 h in 1 out of the 6 nights that observed this band.
 }

### 2014-09-28 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Finishing log N log S plots with new data

@@
 Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}).
-LOFAR's \textsc{aoflagger} has shown to have a good accuracy, is fast and has a library interface, which allows it to be integrated in a pipeline.
-It is therefore used for MWA's RFI detection.
+For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used.
+This flagger has shown to have a good accuracy, is fast and has a library interface, which allows it to be integrated in a pipeline.
 
@@
 When these frequencies are observed, the subbands are often so strongly contaminated that it affects imaging sensitivity.
-For completeness, we include an 16-min observation that covers this frequency range.
+For completeness, we include a 16-min observation that covers this frequency range.
 

### 2014-10-02 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Updates

@@
 Additionally, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume.
-It also collects some statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}.
+It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}.
 Various tools are available to analyse these statistics, e.g.

### 2014-10-03 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Article

@@
 \author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\editmark{Corresponding author.} E-mail: \url{andre.offringa@anu.edu.au}},
-R.~B.~Wayth\Curtin$^,$\CAASTRO, ... , %\newauthor
+R.~B.~Wayth\Curtin$^,$\CAASTRO,
+N.~Hurley-Walker\Curtin,
+D.~Kaplan\Curtin,
+... , %\newauthor
 builders list
@@
 It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}.
-Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern -- in particular for a many-element arrays.
+Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern -- in particular for many-element arrays.
 In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate.
@@
 \hline
-EoR   & 2013-08-23 & 138.8--197.7              & 3h & 0.53\% & ---\\
-GLEAM & 2013-08-25 & 72.3--133.8, 138.8--230.8 & 7h & 0.94\% & 0\% \\
-EoR   & 2013-08-26 & 138.8--197.7              & 6h & 0.81\% & ---\\
-GLEAM & 2013-11-05 & 72.3--133.8, 138.8--230.8 & 7h & 1.27\% & 0\% \\
-GLEAM & 2013-11-25 & 72.3--133.8, 138.8--230.8 & 7h & 0.69\% & 0\% \\
-EoR   & 2014-02-05 & 138.8--197.7              & 6h & 0.54\% & -- \\
-GLEAM & 2014-03-16 & 72.3--133.8, 138.8--230.8 & 7h & 0.79\% & 0\% \\
-GLEAM & 2014-03-17 & 72.3--133.8, 138.8--230.8 & 7h & 1.64\% & 3\% \\
-EoR   & 2014-04-10 & 138.8--197.7              & 6h & 0.68\% & ---\\
-GLEAM & 2014-06-18 & 72.3--133.8, 138.8--230.8 & 7h & 0.98\% & 0\% \\
+EoR high& 2013-08-23 & 167.0--197.7              & 3h & 0.53\% & ---\\
+GLEAM & 2013-08-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.94\% & 0\% \\
+EoR low& 2013-08-26 & 138.9--169.6              & 6h & 0.81\% & ---\\
+GLEAM & 2013-11-05 & 72.3--133.8, 138.9--230.8 & 7h & 1.27\% & 0\% \\
+GLEAM & 2013-11-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.69\% & 0\% \\
+EoR high& 2014-02-05 & 167.0--197.7              & 6h & 0.54\% & -- \\
+GLEAM & 2014-03-16 & 72.3--133.8, 138.9--230.8 & 7h & 0.79\% & 0\% \\
+GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 3\% \\
+EoR high& 2014-04-10 & 167.0--197.7              & 6h & 0.68\% & ---\\
+GLEAM & 2014-06-18 & 72.3--133.8, 138.9--230.8 & 7h & 0.98\% & 0\% \\
 RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
@@
 
-\noindent\begin{figure*}
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set-eor}\vspace{-1cm}
-\caption{..}
-\label{fig:rfi-per-set-eor}
-\end{center}
-\end{figure*}
+%\noindent\begin{figure*}
+%\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set-eor}\vspace{-1cm}
+%\caption{..}
+%\label{fig:rfi-per-set-eor}
+%\end{center}
+%\end{figure*}
 
-\noindent\begin{figure*}
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set-gleam}\vspace{-1cm}
-\caption{..}
-\label{fig:rfi-per-set-gleam}
-\end{center}
-\end{figure*}
+%\noindent\begin{figure*}
+%\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set-gleam}\vspace{-1cm}
+%\caption{..}
+%\label{fig:rfi-per-set-gleam}
+%\end{center}
+%\end{figure*}
 
@@
 
-\noindent\begin{figure}
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/rfi-per-band}
+\noindent\begin{figure*}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
 \caption{RFI per subband}
@@
 \end{center}
-\end{figure}
+\end{figure*}
 
-\noindent\begin{figure}
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/lognlogs-residual}
-\caption{Log N log S for residuals}
-\label{fig:lognlogs-residuals}
+\noindent\begin{figure*}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/lognlogs-multiplot}
+\caption{Log N log S}
+\label{fig:lognlogs}
 \end{center}
-\end{figure}
-
-\noindent\begin{figure}
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/lognlogs-rfi}
-\caption{Log N log S for RFI }
-\label{fig:lognlogs-rfi}
-\end{center}
-\end{figure}
+\end{figure*}
 

### 2014-10-03 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Converted to PASA format

@@
-\documentclass[useAMS,usenatbib]{mn2e}
-\usepackage[utf8]{inputenc}
+\documentclass{pasa}
+\bibliographystyle{apj}
+%\usepackage[utf8]{inputenc}
 \usepackage{graphicx}
@@
 \usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
-\usepackage{xspace}
+%\usepackage{xspace}
 \usepackage{subcaption}
 
-\usepackage{fixltx2e} % fixes placing of image positions
+%\usepackage{fixltx2e} % fixes placing of image positions
 
@@
 \else\usepackage{graphics}\fi
-
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
+\usepackage{graphics}
 
@@
 N.~Hurley-Walker\Curtin,
-D.~Kaplan\Curtin,
+D.~Kaplan...,
 ... , %\newauthor
@@
 }
+\jid{PASA}
+\doi{10.1017/pas.\the\year.xxx}
+\jyear{\the\year}
+%Citations may be made using the natbib commands \citet{},\citep{} etc.
+\usepackage[authoryear]{natbib}
+\bibpunct{(}{)}{;}{a}{}{,}
+\setlength{\bibsep}{0.3mm}
 
 \begin{document}
+\maketitle
 
-\date{Accepted TODO. Received TODO; in original form TODO}
-\pagerange{\pageref{firstpage}--\pageref{lastpage}}
-\pubyear{2014}
+%\date{Accepted TODO. Received TODO; in original form TODO}
+%\pagerange{\pageref{firstpage}--\pageref{lastpage}}
+%\pubyear{2014}
 
 \label{firstpage}
-\maketitle
 
@@
 
-The \textsc{aoflagger} software provides a C++ library\footnote{\url{http://aoflagger.sourceforge.net/doc/api/}} that can be integrated in a pipeline, thereby minimizing reading/writing of data.
+The \textsc{aoflagger} software provides a C++ library%TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
+that can be integrated in a pipeline, thereby minimizing reading/writing of data.
 We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection.
@@
 \hline
-\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{...\%} & \textbf{0.5\%}\\
+\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.96\%} & \textbf{0.5\%}\\
 \hline
@@
 In these observations, the selected EoR field is tracked by the primary beam, such that the sensitivity towards the field is maximal.
-
-The GLEAM survey covers 72.3--230.8 MHz split into 5 bands of 30.72 MHz. The GLEAM observations are made in drift-scan mode, and the 5 bands are observed one by one in 5 subsequent snapshots.
+The GLEAM survey covers 72.3--230.8 MHz split into 5 bands of 30.72 MHz.
+The GLEAM observations are made in drift-scan mode, and the 5 bands are observed one by one in 5 subsequent snapshots.
 
@@
 
+The poly-phase filter bank that performs the first separation into subbands, introduces a 1.28~MHz periodic spectral signature.
+This shape is e.g.
+visible in Fig.~\ref{fig:bandpass}.
+The poly-phase filter band-pass shape is hard to take out completely, because the filters are imperfect and each sub-band is effected by leakage from signals in its adjacent sub-bands.
+This causes a direction-dependent band-pass.
+To solve this, the bordering 80~kHz on both sides of a 1.28~Mhz subband are normally flagged by \textsc{cotter}.
+This implies that in normal observations, 13\% of the data is lost due to the poly-phase filter.
+However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels.
+Therefore, for the analyses in this paper we increase the number of removed edge channels to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.
+
 \section{Detection results}
+The second last column of Table~\ref{tbl:obs-list} list for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger}.
 
@@
 
-\bibliographystyle{mn2e}
 \bibliography{references}

### 2014-10-03 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Work on article

@@
 
-\subsection{The AOFlagger}
+\subsection{The \textsc{aoflagger} RFI detector}
 
@@
 In this band, the band-pass shows two large discontinuities over frequency, because the MWA receivers apply different digital gains at different frequencies to minimize quantization noise.
-TODO explain subband bandpass somewhere %TODO
+The 1.28~MHz pattern is due to the imperfect poly-phase filter.
 }
@@
 The DVB column displays the fraction of visibilities that were unusable because of the presence of DVB signals, and which the flagger does not flag.
-Because the DVB frequencies are only observed in GLEAM observations, this fraction is only given for those.
+Not all observations cover the DVB frequencies.
 }\label{tbl:obs-list}
+\begin{center}%
 \begin{tabular}{|l|l|r|r|r|c|}
@@
 \hline
-EoR high& 2013-08-23 & 167.0--197.7              & 3h & 0.53\% & ---\\
+EoR high& 2013-08-23 & 167.0--197.7              & 3h & 0.53\% & 0\% \\
 GLEAM & 2013-08-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.94\% & 0\% \\
@@
 GLEAM & 2013-11-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.69\% & 0\% \\
-EoR high& 2014-02-05 & 167.0--197.7              & 6h & 0.54\% & -- \\
+EoR high& 2014-02-05 & 167.0--197.7              & 6h & 0.54\% & 0\%\\
 GLEAM & 2014-03-16 & 72.3--133.8, 138.9--230.8 & 7h & 0.79\% & 0\% \\
 GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 3\% \\
-EoR high& 2014-04-10 & 167.0--197.7              & 6h & 0.68\% & ---\\
+EoR high& 2014-04-10 & 167.0--197.7              & 6h & 0.68\% & 0\%\\
 GLEAM & 2014-06-18 & 72.3--133.8, 138.9--230.8 & 7h & 0.98\% & 0\% \\
-RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
+Orbcomm RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
 \hline
-\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.96\%} & \textbf{0.5\%}\\
+\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.96\%} & \textbf{0.33\%}\\
 \hline
 \end{tabular}
+\end{center}
 \end{table*}
@@
 \section{Detection results}
-The second last column of Table~\ref{tbl:obs-list} list for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger}.
+\noindent\begin{figure*}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
+\caption{RFI per subband}
+\label{fig:rfi-per-band}
+\end{center}
+\end{figure*}
+
+The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger}.
+This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% detections.
+Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI.
+This is mostly caused by the FM bands, which are only observed in GLEAM observations.
+The EoR low-band observation shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%.
+
+Fig.~\ref{fig:rfi-per-band} shows the overal detected RFI ratio per sub-band. The FM bands around 100~MHz and the Orbcomm bands around 138~MHz are clearly present in the data.
 
@@
 This same frequency is flagged in several correlations, and it can therefore be assumed this is real RFI.
-This is an example in which the AOFlagger detects an event that would not be detected by regular thresholding.
+This is an example in which the \textsc{aoflagger} detects an event that would not be detected by regular thresholding.
 }[\linewidth]{%
@@
 \caption{RFI from Digital Video Broadcasting (DVB).
-Because of its broadband nature, this kind of RFI is not well flagged in the initial AOFlagger step, but is detectable in the global statistics.
+Because of its broadband nature, this kind of RFI is not well flagged in the initial \textsc{aoflagger} step, but is detectable in the global statistics.
 This type of RFI makes most of the 30.72~MHz bandwidth unusable.
@@
 \end{figure}
-
-\noindent\begin{figure*}
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
-\caption{RFI per subband}
-\label{fig:rfi-per-band}
-\end{center}
-\end{figure*}
 

### 2014-10-08 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Lots of article updates

@@
 GLEAM & 2014-03-16 & 72.3--133.8, 138.9--230.8 & 7h & 0.79\% & 0\% \\
-GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 3\% \\
+GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 30\% \\
 EoR high& 2014-04-10 & 167.0--197.7              & 6h & 0.68\% & 0\%\\
@@
 \hline
-\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.96\%} & \textbf{0.33\%}\\
+\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.96\%} & \textbf{3\%}\\
 \hline
@@
 The poly-phase filter band-pass shape is hard to take out completely, because the filters are imperfect and each sub-band is effected by leakage from signals in its adjacent sub-bands.
-This causes a direction-dependent band-pass.
+The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent.
 To solve this, the bordering 80~kHz on both sides of a 1.28~Mhz subband are normally flagged by \textsc{cotter}.
@@
 \begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
-\caption{RFI per subband}
+\caption{RFI occupancy per subband, calculated over all observations except GLEAM 2014-03-17. The latter has been left out because it is affected by DVB.}
 \label{fig:rfi-per-band}
@@
 
-Fig.~\ref{fig:rfi-per-band} shows the overal detected RFI ratio per sub-band. The FM bands around 100~MHz and the Orbcomm bands around 138~MHz are clearly present in the data.
+Fig.~\ref{fig:rfi-per-band} shows the overal detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17.
+The latter is the only observation affected by interference from digital video broadcasting (DVB), and we will analyse that observation later.
+The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data.
+Excluding the RFI from DVB, the EoR high band is again slightly cleaner than the EoR low band, as its worst subband at 188.2 MHz has 1.03\% occupancy.
+The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy.
+TODO what are these transmitters?
 
@@
 \begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
-\caption{..}
+\caption{Observed noise in the high-resolution channels for each observation.
+This is the residual noise in the visibilities after RFI detection.
+Observation `GLEAM 2014-03-17' shows DVB contamination around 180~MHz and higher, and a few channels in the FM and 2-m amateur bands show a deviating standard deviation.}
 \label{fig:stddev-per-set}
@@
 \end{figure*}
+
+The residual noise levels after flagging can be used to validate whether the flagged data is free of RFI.
+In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations.
+Observation `GLEAM 2014-03-17' shows residual DVB interference, and it is clear that this RFI has not been adequately flagged.
+Therefore, observations have to be cleaned of DVB with another method.
+Some channels in the FM and 2-m amateur bands show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and RFI leakage.
+The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the large loss of data and residual RFI.
+Other than that, the RFI leakage is minimal and the RFI mitigation employed in \textsc{cotter} is sufficient to allow calibration and imaging without further effort.
+This includes the FM bands.
+This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.
+
+\noindent\begin{figure*}%
+% obs data: 1061583576, 2013-08-26.
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=8cm]{img/dvb_example}\includegraphics[width=8cm]{img/2014-03-17-dvb-stddevs}
+\end{center}
+\caption{RFI from Digital Video Broadcasting (DVB), which is detected in 1 of the 9 observations (GLEAM~2014-03-17) that cover the DVB frequencies.
+Left plot: dynamic spectrum for the worst affected snapshot; right plot: visibility RMS within each of the DVB bands for the affected night after flagging.
+Because of its broadband nature, this kind of RFI is not well flagged in the initial \textsc{aoflagger} step, but it is detectable in the global statistics.
+When observing at 182~MHz, this type of RFI makes most of the 30.72~MHz bandwidth unusable when it is visible.
+}
+\label{fig:dvb}
+\end{figure*}
+
+Because significant DVB interference is left in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively.
+The DVB transmitters are terrestrial, and the fact that only this night observes the DVB, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity.
+As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth.
+Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive for steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive for RFI that is this broad.
+
+Despite not being detecting by the flagger, the right plot of Fig.~\ref{fig:dvb} shows that it is easy to recognize DVB interference from the data statistics, for example by testing whether the RMS in any of the occupied DVB bands exceeds the RMS in DVB band 4, which is not used for broadcasting.
+When DVB is detected in a snapshot, in most cases the best action will be to remove the entire snapshot, as it is not worth having special mechanisms in place to salvage the small amount of residual good data within an affected snapshot.
+With 2 out of 7 hours affected in 1 out of 9 observed nights, a rough probability of 3\% can be given for observing DVB, as is detailed in the last column of Table~\ref{tbl:obs-list}.
 
@@
 
-\noindent\begin{figure}
-% obs data: 1061583576, 2013-08-26.
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/dvb_example}
-\caption{RFI from Digital Video Broadcasting (DVB).
-Because of its broadband nature, this kind of RFI is not well flagged in the initial \textsc{aoflagger} step, but is detectable in the global statistics.
-This type of RFI makes most of the 30.72~MHz bandwidth unusable.
-It is visible for about 2 h in 1 out of the 6 nights that observed this band.
-}
-\label{fig:dvb_example}
-\end{center}
-\end{figure}
-
 %\noindent\begin{figure*}
@@
 
-\noindent\begin{figure}
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/2014-03-17-dvb-stddevs}
-\caption{..}
-\label{fig:dvb-stddevs}
-\end{center}
-\end{figure}
-
 \noindent\begin{figure*}
@@
 \begin{itemize}
- \item Total RFI statistics
- \item Percentages + stddev per frequency + discuss
  \item An example of RFI found (e.g., the EoR DVB, ..)

### 2014-10-10 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Todays work

@@
 \hline
+\multicolumn{4}{|r|}{\textbf{With uniform channel weighting}} & \textbf{1.13\%} & \\
+\hline
 \end{tabular}
@@
 We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall).
-The observations are listen in Table~\ref{tbl:obs-list}.
+The observation nights are listed in Table~\ref{tbl:obs-list}.
 
@@
 The EoR low-band observation shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%.
+The total RFI occupancy in all observations is 0.96\%.
+Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global expected RFI ratio in the 72.3--230.8 MHz range.
 
@@
 Therefore, observations have to be cleaned of DVB with another method.
-Some channels in the FM and 2-m amateur bands show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and RFI leakage.
-The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the large loss of data and residual RFI.
-Other than that, the RFI leakage is minimal and the RFI mitigation employed in \textsc{cotter} is sufficient to allow calibration and imaging without further effort.
-This includes the FM bands.
+Additionally, some channels in the FM and 2-m amateur bands show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and because of RFI leakage, but because the effect is small these frequencies can be calibrated and imaged without a problem.
+The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI.
+Other than that, the RFI leakage is minimal.
+With exception of the ORBCOMM frequencies and DVB affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts.
 This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.
@@
 % obs data: 1061583576, 2013-08-26.
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=8cm]{img/dvb_example}\includegraphics[width=8cm]{img/2014-03-17-dvb-stddevs}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=10cm]{img/dvb_example}\includegraphics[width=5cm]{img/2014-03-17-dvb-stddevs}
 \end{center}
-\caption{RFI from Digital Video Broadcasting (DVB), which is detected in 1 of the 9 observations (GLEAM~2014-03-17) that cover the DVB frequencies.
-Left plot: dynamic spectrum for the worst affected snapshot; right plot: visibility RMS within each of the DVB bands for the affected night after flagging.
+\caption{RFI from Digital Video Broadcasting (DVB), which is visible in 1 of the 9 observations (GLEAM~2014-03-17) that cover the DVB frequencies.
+Left plot: dynamic spectrum for the worst affected snapshot; right plot: visibility RMS in a few of the DVB bands for the affected night after flagging.
 Because of its broadband nature, this kind of RFI is not well flagged in the initial \textsc{aoflagger} step, but it is detectable in the global statistics.
@@
 
-Despite not being detecting by the flagger, the right plot of Fig.~\ref{fig:dvb} shows that it is easy to recognize DVB interference from the data statistics, for example by testing whether the RMS in any of the occupied DVB bands exceeds the RMS in DVB band 4, which is not used for broadcasting.
+Although the flagger does not adequately flag DVB, the right plot of Fig.~\ref{fig:dvb} shows that it is easy to recognize DVB interference from the data statistics.
+An option is to test whether the visibility RMS of a snapshot in any of the occupied DVB bands exceeds the RMS in DVB band 4, which is not used for broadcasting.
 When DVB is detected in a snapshot, in most cases the best action will be to remove the entire snapshot, as it is not worth having special mechanisms in place to salvage the small amount of residual good data within an affected snapshot.
-With 2 out of 7 hours affected in 1 out of 9 observed nights, a rough probability of 3\% can be given for observing DVB, as is detailed in the last column of Table~\ref{tbl:obs-list}.
+Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability of observing DVB is roughly 3\%.
+This is detailed in the last column of Table~\ref{tbl:obs-list}.
 
@@
 }\\\vspace{3mm}%
-\subcaptionbox{Short RFI burst of 2 s centred on 150.17 MHz. This shows that flagging at high time and frequency resolution is necessary to accurately detect RFI.}[\linewidth]{%
+\subcaptionbox{Short RFI burst of 2 s centred on 150.17 MHz. Detection of this kind of RFI requires flagging at high time and frequency resolution.}[\linewidth]{%
 % obs data: 1061583576, 2013-08-26.
@@
 }%
-\caption{Examples of RFI events found in the EoR low band (138.9--197.7 MHz) in a 2-min snapshot with relatively high RFI contamination.
+\caption{Examples of RFI events found in the EoR low band (138.9--169.6 MHz) in a 2-min snapshot with relatively high RFI contamination.
 These panels show the Stokes I amplitudes.
@@
 The horizontally flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
+\end{figure*}
+
+\noindent\begin{figure*}%
+% obs data: 1075641128
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=8cm]{img/EoR-high-band-DVB-burst-example-noflags}\includegraphics[width=8cm]{img/EoR-high-band-DVB-burst-example-withflags}
+\end{center}
+\caption{???. }
+\label{fig:dvb-burst}
 \end{figure*}
@@
 
+\subsection{Computational performance of \textsc{cotter}}
+Because \textsc{cotter} processes the data at the highest resolution, its computational performance is an important consideration.
+Because of the large volume of MWA data, its runtime is strongly dependent on the input-output (IO) performance of the host system.
+Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent.
+In \citet{lofar-radio-environment} it is shown that the performance of the \textsc{aoflagger} is in normal situations dominated by IO, and hence the same holds for \textsc{cotter}: on a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup, processing a single 2~min snapshot with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, approximately 3 min are spend on reading the data, 5 min on RFI detection and 3.5 min on writing the data.
+Real-time processing can therefore be achieved by using 6 of such nodes in parallel.
+
 \begin{itemize}
- \item An example of RFI found (e.g., the EoR DVB, ..)
+ \item Describe RFI examples (EoR low band \& high band)
  \item Performance
@@
  \item SKA -- how to implement a Cotter for SKA?
+ \item PAPER etc needs to use combinatorial/morphological flagging
 \end{itemize}
+\subsection{Comparison with LOFAR}
 

### 2014-10-14 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Work

@@
 \hline
-\multicolumn{4}{|r|}{\textbf{With uniform channel weighting}} & \textbf{1.13\%} & \\
+\multicolumn{4}{|r|}{\textbf{With uniform channel coverage}} & \textbf{1.13\%} & \\
 \hline
@@
 This is mostly caused by the FM bands, which are only observed in GLEAM observations.
-The EoR low-band observation shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%.
+The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%.
 The total RFI occupancy in all observations is 0.96\%.
@@
 Fig.~\ref{fig:rfi-per-band} shows the overal detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17.
-The latter is the only observation affected by interference from digital video broadcasting (DVB), and we will analyse that observation later.
+The latter is the only night affected by interference from digital video broadcasting (DVB), and will be analysed later.
 The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data.
@@
 
-Although the flagger does not adequately flag DVB, the right plot of Fig.~\ref{fig:dvb} shows that it is easy to recognize DVB interference from the data statistics.
+Although the flagger does not adequately flag DVB, the right plot of Fig.~\ref{fig:dvb} shows that it is possible to determine whether a snapshot is affected by DVB interference from the data statistics.
 An option is to test whether the visibility RMS of a snapshot in any of the occupied DVB bands exceeds the RMS in DVB band 4, which is not used for broadcasting.
-When DVB is detected in a snapshot, in most cases the best action will be to remove the entire snapshot, as it is not worth having special mechanisms in place to salvage the small amount of residual good data within an affected snapshot.
+These statistics are collected by \textsc{cotter}.
+When DVB is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used.
+However, it might not be worth it to have special mechanisms, such as having to account for lower SNR and a decreased $uv$ coverage, in place to salvage the residual good data within an affected snapshot.
 Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability of observing DVB is roughly 3\%.
@@
 \subsection{Computational performance of \textsc{cotter}}
-Because \textsc{cotter} processes the data at the highest resolution, its computational performance is an important consideration.
-Because of the large volume of MWA data, its runtime is strongly dependent on the input-output (IO) performance of the host system.
-Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent.
-In \citet{lofar-radio-environment} it is shown that the performance of the \textsc{aoflagger} is in normal situations dominated by IO, and hence the same holds for \textsc{cotter}: on a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup, processing a single 2~min snapshot with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, approximately 3 min are spend on reading the data, 5 min on RFI detection and 3.5 min on writing the data.
+Because \textsc{cotter} processes the data at high resolution, its computational performance is an important consideration.
+A major contribution to the runtime is the reading and writing of the data, and the runtime is thus influenced by the input-output (IO) performance of the host system.
+Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases in total the runtime by approximately a percent.
+On a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup, processing a single 2~min snapshot with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, approximately 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data.
 Real-time processing can therefore be achieved by using 6 of such nodes in parallel.
@@
  \item Describe RFI examples (EoR low band \& high band)
- \item Performance
  \item Log N log S Histograms

### 2014-10-21 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Work on article

@@
 
-\section{Data description}
-
 \begin{table*}
@@
 
+\section{Data description}
 We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall).
@@
 
-\section{Detection results}
 \noindent\begin{figure*}
@@
 
+\noindent\begin{figure*}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
+\caption{Observed noise in the high-resolution channels for each observation.
+This is the residual noise in the visibilities after RFI detection.
+Observation `GLEAM 2014-03-17' shows DVB contamination around 180~MHz and higher, and a few channels in the FM and 2-m amateur bands show a deviating standard deviation.}
+\label{fig:stddev-per-set}
+\end{center}
+\end{figure*}
+
+\section{Detection results}
 The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger}.
@@
 
-\noindent\begin{figure*}
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
-\caption{Observed noise in the high-resolution channels for each observation.
-This is the residual noise in the visibilities after RFI detection.
-Observation `GLEAM 2014-03-17' shows DVB contamination around 180~MHz and higher, and a few channels in the FM and 2-m amateur bands show a deviating standard deviation.}
-\label{fig:stddev-per-set}
-\end{center}
-\end{figure*}
-
 The residual noise levels after flagging can be used to validate whether the flagged data is free of RFI.
@@
 The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI.
-Other than that, the RFI leakage is minimal.
-With exception of the ORBCOMM frequencies and DVB affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts.
+With the exception of the ORBCOMM frequencies and DVB affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts.
 This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.
@@
 Because of its broadband nature, this kind of RFI is not well flagged in the initial \textsc{aoflagger} step, but it is detectable in the global statistics.
-When observing at 182~MHz, this type of RFI makes most of the 30.72~MHz bandwidth unusable when it is visible.
 }
@@
 The DVB transmitters are terrestrial, and the fact that only this night observes the DVB, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity.
-As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth.
+As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can fully contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth.
 Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive for steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive for RFI that is this broad.
@@
 An option is to test whether the visibility RMS of a snapshot in any of the occupied DVB bands exceeds the RMS in DVB band 4, which is not used for broadcasting.
-These statistics are collected by \textsc{cotter}.
+These statistics are calculated by \textsc{cotter}, hence this can be validated without having to read the data again.
 When DVB is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used.
-However, it might not be worth it to have special mechanisms, such as having to account for lower SNR and a decreased $uv$ coverage, in place to salvage the residual good data within an affected snapshot.
+However, it might not be worth it to have special mechanisms in place, such as having to account for lower SNR and a decreased $uv$ coverage, to salvage the residual good data within an affected snapshot.
 Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability of observing DVB is roughly 3\%.
 This is detailed in the last column of Table~\ref{tbl:obs-list}.
+
+\noindent\begin{figure*}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/lognlogs-multiplot}
+\caption{Visibility amplitude distributions with logarithmic axes.
+The dashed line represent fits over a reasonably constant part of the tail (selected by eye) of the RFI distribution: $N=\beta S^{-1.43}$ for GLEAM, $N=\beta S^{-4.47}$ for EoR high and $N=\beta S^{-1.33}$ for EoR low.}
+\label{fig:lognlogs}
+\end{center}
+\end{figure*}
+
+\subsection{Distribution analysis}
+In \citet{offringa-rfi-distributions}, using LOFAR it is shown that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail with an exponent of approximately -1.5 to -1.6.
+For the GLEAM, EoR high-band and EoR low-band observations of this work, the distributions are drawn in Fig.~\ref{fig:lognlogs}.
+The three distributions are different due to the fact that each observation covers different frequencies.
+Separate distributions are shown for visibilities that have been detected as RFI (red lines) and those that have been classified as RFI free (green lines).
+The EoR high-band curves do not include the night that was contaminated by DVB RFI.
+
+In the ideal case, the residual visibility distributions show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}).
+The tails of the residual curves of Fig.~\ref{fig:lognlogs} are however not smooth.
+Especially in the case of the EoR high-band and low-band observations, the residual curve shows RFI that has not been properly detected as such.
+In the case of EoR low-band, samples are not detected as RFI even though they are an order of magnitude brighter than where the GLEAM residual curve ends.
+The reason such samples leak through detection is because the detection strategy includes a bandpass correction per 1.28-MHz sub-band.
+When RFI persistently contaminates approximately a full subband, its effect will be absorbed in the bandpass correction and becomes invisible for the flagger.
+In some cases, this will cause the calibration to fail on those channels, after which it will be flagged nevertheless.
+In other cases, such RFI is included in the imaging step or further processing, which is obviously very undesirable.
+
+The distribution curves do not show a very significant power-law tail.
+The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N=\beta S^{-4.47}$, with $\beta$ being an arbitrary scaling factor.
+The GLEAM and EoR low-band observations do show a small power-law part, but the fitted exponent depends strongly on what part of the tail is selected.
+Fits over by-eye selected regions have been added, with $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and Eor-low curves of Fig.~\ref{fig:lognlogs}, respectively.
 
@@
 % obs data: 1061583576, 2013-08-26.
-\includegraphics[width=9.1cm]{img/150_2_mhz_example}\includegraphics[width=9.1cm]{img/150_2_mhz_flagged}
+\includegraphics[width=9.1cm]{img/150_2_mhz_example_ann}\includegraphics[width=9.1cm]{img/150_2_mhz_flagged_ann}
 \label{fig:150_2}
-}\\\vspace{3mm}%
-\subcaptionbox{A weakly-observed transmitter centred on 156.66 MHz.
-This same frequency is flagged in several correlations, and it can therefore be assumed this is real RFI.
-This is an example in which the \textsc{aoflagger} detects an event that would not be detected by regular thresholding.
-}[\linewidth]{%
-% obs data: 1061583576, 2013-08-26.
-\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}
-\label{fig:156_7}
 }%
-\caption{Examples of RFI events found in the EoR low band (138.9--169.6 MHz) in a 2-min snapshot with relatively high RFI contamination.
+\caption{RFI events found in the EoR low band (138.9--169.6 MHz) in a 2-min snapshot with relatively high RFI contamination.
 These panels show the Stokes I amplitudes.
 In the right figures, the result of RFI detection is shown with purple.
-The horizontally flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
+The horizontal flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
+\label{fig:2m-amateur-band}
 \end{figure*}
@@
 \end{center}
-\caption{???. }
+\caption{DVB RFI briefly visible in an EoR high-band observation. The plots show the Stokes~I amplitudes for a single correlation. The right plot shows the result of RFI detection marked in purple.}
 \label{fig:dvb-burst}
@@
 
-%\noindent\begin{figure*}
-%\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set-eor}\vspace{-1cm}
-%\caption{..}
-%\label{fig:rfi-per-set-eor}
-%\end{center}
-%\end{figure*}
+\noindent\begin{figure*}%
+% obs data: 1075637224
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=8cm]{img/broadband-pulse-unflagged}\includegraphics[width=8cm]{img/broadband-pulse-flagged}
+\end{center}
+\caption{Broadband pulses found in an EoR high-band observation.
+The two panels show Stokes~I visibility amplitudes of a single correlation.
+In the right plot, the RFI detection flags are marked with purple.
+Each event occupies 2 timesteps of 0.5~s.}
+\label{fig:broadband-pulse}
+\end{figure*}
 
-%\noindent\begin{figure*}
-%\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set-gleam}\vspace{-1cm}
-%\caption{..}
-%\label{fig:rfi-per-set-gleam}
-%\end{center}
-%\end{figure*}
-
-\noindent\begin{figure*}
-\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/lognlogs-multiplot}
-\caption{Log N log S}
-\label{fig:lognlogs}
+\noindent\begin{figure*}%
+% obs data: 1061583576, 2013-08-26.
+\begin{center}
+\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}\\
+\includegraphics[width=9.1cm]{img/156_7_mhz_stddev}\includegraphics[width=9.1cm]{img/156_7_mhz_stddev_flagged}
+\caption{A weakly-observed transmitter centred on 156.66 MHz.
+Top-left window: a single correlation, where the transmitter is barely visible; Top-right window: detected RFI marked with purple.
+When performing per-baseline flagging, the transmitter is partly detected; Bottom-left window: when combining data from all baselines, the transmitter becomes clearly visible; Bottom-right window: in yellow, result of executing \textsc{aoflagger} on the standard deviations calculated over all baselines.}\label{fig:156_7}
 \end{center}
 \end{figure*}
+
+\subsection{RFI types}
+A variety of RFI events types are observed in the data sets.
+While there are too many transmitters to show examples for each, it is helpful to understand what kind of events are visible and how they are flagged by the \textsc{aoflagger}.
+Therefore, a few typical examples are shown.
+
+Fig.~\ref{fig:2m-amateur-band} shows two examples of RFI events in the same EoR low-band snapshot: the top panels display the Stokes~I values of a single correlation, in which a transmitter has been observed in the 2-m amateur band (146 MHz).
+This is the worst contamination in our data sets by this transmitter, and there are also snapshots in which the transmitter is not visible at all.
+This variability might be caused by intrinsic variation or movement of the transmitter or varying propagation conditions.
+While a change in pointing can also change the appearance of the transmitter, the beam does not change within a single snapshot, while in Fig.~\ref{fig:2m-amateur-band} the strength and affected bandwidth of the transmitter does change during the snapshot.
+In the bottom panel of Fig.~\ref{fig:2m-amateur-band}, the same snapshot is showed but is zoomed in on a briefly-observed RFI event at 150.17 MHz.
+This event occupies only a single 40~kHz channel for 2 seconds, and thus is an event which requires flagging at approximately the observed time and frequency resolution or higher for accurate detection.
+
+Besides persistent transmissions that occupy a few channels, transient broad-band events are observed as well.
+Occassionally, DVB RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}.
+As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}.
+Another transient broad-band example is displayed in Fig.~\ref{fig:broadband-pulse}, which shows strong broad-band pulses of a second in length.
+Strong pulses such as these are rare and well flagged, but a few weak broad-band pulses are observed in almost every 2-min snapshot.
+These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines are added together.
+We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time.
+The MWA Voltage Capture System (VCS; TODO cite) might help to analyze these phenomena.
+
+Fig.~\ref{fig:156_7} shows a longer RFI contamination at 156.66~MHz that is hardly visible in a single correlation.
+The \textsc{aoflagger} detects this RFI partially (Fig.~\ref{fig:156_7}, top-right plot), but when plotting the standard deviation over all correlations in a dynamic spectrum, it is evident that this RFI transmitter is not completely detected (Fig.~\ref{fig:156_7}, bottom-left plot).
+The detection becomes more complete when the \textsc{aoflagger} is executed on the standard deviations over all baselines (Fig.~\ref{fig:156_7}, bottom-right plot).
+This kind of detection is currently not implemented in \textsc{cotter}.
 
@@
 A major contribution to the runtime is the reading and writing of the data, and the runtime is thus influenced by the input-output (IO) performance of the host system.
-Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases in total the runtime by approximately a percent.
+Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent.
 On a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup, processing a single 2~min snapshot with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, approximately 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data.
 Real-time processing can therefore be achieved by using 6 of such nodes in parallel.
-
-\begin{itemize}
- \item Describe RFI examples (EoR low band \& high band)
- \item Log N log S Histograms
-\end{itemize}
 
@@
  \item PAPER etc needs to use combinatorial/morphological flagging
+ \item The presence of strong interference implies there might be weak interference present that is easily seen or detected. 
 \end{itemize}
@@
  \item FM not completely empty
+ \item Frequency behaviour of instrument makes things harder! Need to fix pass-band correction.
 \end{itemize}

### 2014-10-22 cotter/flagreader.h
Fixing possible issue with reading in flag files ; in misaligned gpuboxes, the flags might have been placed at the wrong positions

@@ -80,13 +80,17 @@ public:
 		for(size_t subband=_sbStart; subband!=_sbEnd; ++subband)
 		{
 			size_t fileIndex = subband - _sbStart;
-			size_t row = (timestep + _hduOffsets[_subbandToGPUBoxFileIndex[subband]]) * _baselineCount + baseline + 1;
-			fits_read_col(_files[fileIndex], TBIT, /*colnum*/ _colNums[fileIndex], /*firstrow*/ row, /*firstelem*/ 1,
-       /*nelements*/ _channelsPerGPUBox, /*(*)nulval*/ 0, &_buffer[0], 0 /*(*)anynul*/, &status);
-			for(size_t ch=0; ch!=_channelsPerGPUBox; ++ch)
+			int offset = _hduOffsets[_subbandToGPUBoxFileIndex[subband]];
+			if(int(timestep) >= offset)
 			{
-				size_t channelIndex = ch + fileIndex*_channelsPerGPUBox;
-				bufferStartPos[channelIndex*bufferStride] = bool(_buffer[ch]);
+				size_t row = (timestep - offset) * _baselineCount + baseline + 1;
+				fits_read_col(_files[fileIndex], TBIT, /*colnum*/ _colNums[fileIndex], /*firstrow*/ row, /*firstelem*/ 1,
+				/*nelements*/ _channelsPerGPUBox, /*(*)nulval*/ 0, &_buffer[0], 0 /*(*)anynul*/, &status);
+				for(size_t ch=0; ch!=_channelsPerGPUBox; ++ch)
+				{
+					size_t channelIndex = ch + fileIndex*_channelsPerGPUBox;
+					bufferStartPos[channelIndex*bufferStride] = bool(_buffer[ch]);
+				}
 			}
 		}
 	}

### 2014-10-22 cotter/version.h
Increasing version of Cotter

@@ -1,9 +1,9 @@
 #ifndef COTTER_VERSION_H
 #define COTTER_VERSION_H
 
-#define COTTER_VERSION_STR  "3.1"
+#define COTTER_VERSION_STR  "3.2"
 #define COTTER_VERSION_MAJOR 3
-#define COTTER_VERSION_MINOR 1
-#define COTTER_VERSION_DATE "2014-10-05"
+#define COTTER_VERSION_MINOR 2
+#define COTTER_VERSION_DATE "2014-10-22"
 
 #endif

### 2014-10-22 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Some more work on the article

@@
 In the ideal case, the residual visibility distributions show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}).
-The tails of the residual curves of Fig.~\ref{fig:lognlogs} are however not smooth.
-Especially in the case of the EoR high-band and low-band observations, the residual curve shows RFI that has not been properly detected as such.
-In the case of EoR low-band, samples are not detected as RFI even though they are an order of magnitude brighter than where the GLEAM residual curve ends.
+However, especially the tails of the residual curves of the EoR high-band and low-band observations are not smooth in Fig.~\ref{fig:lognlogs}, and the residual curve shows RFI that has not been properly detected as such.
+In the case of EoR low-band, samples are not detected as RFI even though they are an order of magnitude brighter than the brightest samples in the GLEAM observations.
 The reason such samples leak through detection is because the detection strategy includes a bandpass correction per 1.28-MHz sub-band.
@@
 The distribution curves do not show a very significant power-law tail.
+Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}.
 The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N=\beta S^{-4.47}$, with $\beta$ being an arbitrary scaling factor.
 The GLEAM and EoR low-band observations do show a small power-law part, but the fitted exponent depends strongly on what part of the tail is selected.
-Fits over by-eye selected regions have been added, with $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and Eor-low curves of Fig.~\ref{fig:lognlogs}, respectively.
+The selected regions result in $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and Eor-low curves, respectively.
 

### 2014-10-24 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Lots of updates

@@
 We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection.
-Additionally, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume.
+RFI detection is only performed on cross-correlations.
+Auto-correlations are normally removed, because they are not used in imaging.
+In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume.
 It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}.
@@
 This is the residual noise in the visibilities after RFI detection.
-Observation `GLEAM 2014-03-17' shows DVB contamination around 180~MHz and higher, and a few channels in the FM and 2-m amateur bands show a deviating standard deviation.}
+A few channels in the FM bands show a deviating standard deviation, and observation `GLEAM 2014-03-17' shows DVB contamination around 180~MHz.}
 \label{fig:stddev-per-set}
@@
 \section{Detection results}
-The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger}.
+The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations.
 This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% detections.
@@
 
-In the ideal case, the residual visibility distributions show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}).
-However, especially the tails of the residual curves of the EoR high-band and low-band observations are not smooth in Fig.~\ref{fig:lognlogs}, and the residual curve shows RFI that has not been properly detected as such.
-In the case of EoR low-band, samples are not detected as RFI even though they are an order of magnitude brighter than the brightest samples in the GLEAM observations.
-The reason such samples leak through detection is because the detection strategy includes a bandpass correction per 1.28-MHz sub-band.
-When RFI persistently contaminates approximately a full subband, its effect will be absorbed in the bandpass correction and becomes invisible for the flagger.
-In some cases, this will cause the calibration to fail on those channels, after which it will be flagged nevertheless.
-In other cases, such RFI is included in the imaging step or further processing, which is obviously very undesirable.
-
-The distribution curves do not show a very significant power-law tail.
+The distribution curves of RFI-detected samples do not show a very significant power-law tail.
 Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}.
-The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N=\beta S^{-4.47}$, with $\beta$ being an arbitrary scaling factor.
+The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N=\beta S^{-4.47}$, with $\beta$ being a scaling factor.
 The GLEAM and EoR low-band observations do show a small power-law part, but the fitted exponent depends strongly on what part of the tail is selected.
 The selected regions result in $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and Eor-low curves, respectively.
+
+In the ideal case, the residual visibility distributions show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}).
+The residual curves in Fig.~\ref{fig:lognlogs} do not go down as quickly as a Rayleigh curve would, and therefore the distributions have a slight excess of samples with higher amplitudes.
+This excess is caused by the multi-variate nature of the data, for example because the noise level increases when the Galaxy goes through the beam.
+The EoR-high distribution has some extra features in its tails that are not smooth.
+Further analyses showed that these are also caused by the Galaxy, causing a few short baselines to observe samples with high amplitudes.
+Visual inspection of snapshots with a non-smooth residual distribution tail did not show residual RFI in such sets.
 
@@
  \item PAPER etc needs to use combinatorial/morphological flagging
- \item The presence of strong interference implies there might be weak interference present that is easily seen or detected. 
+ \item The presence of strong interference implies there might be weak interference present that is not easily seen or detected. 
 \end{itemize}
+
+\noindent\begin{figure}
+\begin{center}\hspace*{-0.2cm}\includegraphics[width=9cm]{img/LOFAR-MWA-occupancy}
+\caption{Comparison between LOFAR and MWA RFI occupancies. The statistics are regridded to the same frequency resolution of 48~kHz. }
+\label{fig:lofar-mwa-occupancy}
+\end{center}
+\end{figure}
+
 \subsection{Comparison with LOFAR}
+The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies.
+The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}.
+The statistics are regridded to the same frequency resolution of 48~kHz.
+The average RFI occupancy in this frequency range is 1.65\% for the MWA, while O2012 reports a 3.18\% occupancy for LOFAR.
+
+It should be noted that the RFI detection for LOFAR was performed at 0.78~kHz while for the MWA it is performed at 40~kHz.
+As shown in O2012, LOFAR observes RFI that is only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions.
+O2012 also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA.
+In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources: Cassiopeia~A is never visible at the MWA, and Cygnus~A is only visible at low declinations.
+An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations.
+Taking into account these differences, it becomes hard to compare the environments fairly based on detected RFI levels.
+Nevertheless, for both telescopes, the automatic detection strategies have been optimized such that as little data as possible is thrown away but the detection is sufficient for further data reduction.
+For the MWA, this goal was reached by throwing away a factor of two less data.
+
+After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging.
+While the difference between a 1.65\% or 3.18\% loss of data is hardly a reason to spend resources to build at a remote site like the Murchison Radio Observatory, there are additional reasons to do so.
+A benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the FM-station bands (88--108~MHz).
+Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}.
+Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR.
+Such a detection requires very deep integrations, and it could be that RFI becomes problematic when reaching lower noise levels.
+However, initial results with LOFAR predict that RFI will not prevent a detection of EoR signals \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.
 

### 2014-10-27 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Started on the conclusions

@@
 Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{post-correlation-rfi-classification}.
-The consequence of the latter approach is that a certain amount of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed.
-Analysing the results of interference detection and improving the understanding of RFI contamination can aid in several ways: understanding the effects on the science; improved observation scheduling; designing hardware and location of future telescopes with a maximal cost/benefit approach; and enforcing spectrum management strategies that are most beneficial for the observatories.
+The consequence of this approach is that a certain amount of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed.
+It is important to analyze the impact of RFI on a specific instrument to consider the right RFI mitigation approach.
+Additionally, an increased understanding of the RFI situation will help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; chosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.
 
@@
 The interferometric arrays LOFAR, PAPER and the Long Wavelength Array (LWA) are all designed to perform 21-cm EoR science, and all observe in approximately the same frequency range as the MWA.
-Comparing the MWA to these arrays therefore allows the comparison of the observatory site and hardware design with respect to the RFI mitigation effectivity.
-In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose any issues on regular observations despite that, unlike the MWA, LOFAR is build in a populated area.
+Comparing the MWA to these arrays therefore allows the comparison of the observatory site and hardware design with respect to the RFI impact.
+TODO say something about PAPER and MWA RFI.
+In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose issues on regular observations even though LOFAR is build in a populated area.
 This is confirmed by \citet{ncp-eor-yatawatta}, where the authors analyse the first imaging of EoR long-integration observations with LOFAR.
@@
 It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}.
-Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern -- in particular for many-element arrays.
+Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays.
 In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate.
@@
 For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used.
-This flagger has shown to have a good accuracy, is fast and has a library interface, which allows it to be integrated in a pipeline.
+It has been shown that this flagger has a good accuracy and is fast.
+It also has a library interface, which allows it to be integrated in a pipeline.
 
@@
 
-The \textsc{aoflagger} software provides a C++ library%TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
+The \textsc{aoflagger} software provides a C++ library %TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
 that can be integrated in a pipeline, thereby minimizing reading/writing of data.
@@
 \textsc{casa} or \textsc{miriad}, the \textsc{cotter} output is set to either the \textsc{casa} or uvfits format.
-The output is then directly readable by the astronomical software.
+The output is then directly readable by the most common astronomical software packages.
 
@@
 When less memory is available, \textsc{cotter} will split the observation in different time chunks and flags those independently.
-The \textsc{ndppp} software used by LOFAR (TODO reference?) uses a similar partitioning of data to overcome the time-ordering problem.
+This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR (TODO reference?) to overcome the time-ordering problem.
 Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy.
-Because not all astronomers might have easily access to high-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations.
+Because not all astronomers have easily access to high-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations.
+For this use-case, \textsc{cotter} is run in a flagging-only mode on the high-memory machine.
+The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to convert the raw files and apply the RFI detection from the first run.
 
@@
 We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall).
-The observation nights are listed in Table~\ref{tbl:obs-list}.
+The nights of observations are listed in Table~\ref{tbl:obs-list}.
 
@@
 However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels.
-Therefore, for the analyses in this paper we increase the number of removed edge channels to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.
+Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.
 
@@
 The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data.
-Excluding the RFI from DVB, the EoR high band is again slightly cleaner than the EoR low band, as its worst subband at 188.2 MHz has 1.03\% occupancy.
+Excluding the RFI from DVB, the EoR high band is slightly cleaner than the EoR low band, as its worst subband at 188.2 MHz has 1.03\% occupancy.
 The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy.
@@
 Therefore, observations have to be cleaned of DVB with another method.
-Additionally, some channels in the FM and 2-m amateur bands show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and because of RFI leakage, but because the effect is small these frequencies can be calibrated and imaged without a problem.
+Additionally, some channels in the FM radio band show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and possibly because of RFI leakage.
+Nevertheless, because the effect is small these frequencies can be calibrated and imaged without a problem.
 The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI.
@@
 As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can fully contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth.
-Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive for steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive for RFI that is this broad.
+Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive to steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive for RFI that is this broad.
 
@@
 The selected regions result in $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and Eor-low curves, respectively.
+Due to the small part over which the power laws hold, as well as the high possible error in the fits because a fitting region needs to be selected, it is hard to infer if a uniform distribution of transmitters is observed.
 
@@
 This is the worst contamination in our data sets by this transmitter, and there are also snapshots in which the transmitter is not visible at all.
-This variability might be caused by intrinsic variation or movement of the transmitter or varying propagation conditions.
+This variability might be caused by intrinsic variation, movement of the transmitter or varying propagation conditions.
 While a change in pointing can also change the appearance of the transmitter, the beam does not change within a single snapshot, while in Fig.~\ref{fig:2m-amateur-band} the strength and affected bandwidth of the transmitter does change during the snapshot.
@@
 We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time.
-The MWA Voltage Capture System (VCS; TODO cite) might help to analyze these phenomena.
+The MWA Voltage Capture System (VCS; TODO cite), which allows high time resolution observations, might help to analyze these phenomena.
 
 Fig.~\ref{fig:156_7} shows a longer RFI contamination at 156.66~MHz that is hardly visible in a single correlation.
-The \textsc{aoflagger} detects this RFI partially (Fig.~\ref{fig:156_7}, top-right plot), but when plotting the standard deviation over all correlations in a dynamic spectrum, it is evident that this RFI transmitter is not completely detected (Fig.~\ref{fig:156_7}, bottom-left plot).
+The \textsc{aoflagger} detects this RFI partially (Fig.~\ref{fig:156_7}, top-right plot), but when plotting the standard deviation over all correlations in a dynamic spectrum, it is evident that this RFI transmitter extends beyond what is detected (Fig.~\ref{fig:156_7}, bottom-left plot).
 The detection becomes more complete when the \textsc{aoflagger} is executed on the standard deviations over all baselines (Fig.~\ref{fig:156_7}, bottom-right plot).
@@
 Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent.
-On a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup, processing a single 2~min snapshot with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, approximately 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data.
+A high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup, is used to time \textsc{cotter}.
+The wall-clock runtime for processing a single 2~min snapshot with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, is split up as follows: 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data.
 Real-time processing can therefore be achieved by using 6 of such nodes in parallel.
-
-\section{Discussion}
-\begin{itemize}
- \item SKA -- how to implement a Cotter for SKA?
- \item PAPER etc needs to use combinatorial/morphological flagging
- \item The presence of strong interference implies there might be weak interference present that is not easily seen or detected. 
-\end{itemize}
 
@@
 \begin{center}\hspace*{-0.2cm}\includegraphics[width=9cm]{img/LOFAR-MWA-occupancy}
-\caption{Comparison between LOFAR and MWA RFI occupancies. The statistics are regridded to the same frequency resolution of 48~kHz. }
+\caption{Comparison between LOFAR and MWA RFI occupancies. The statistics are resampled to the same frequency resolution of 48~kHz. }
 \label{fig:lofar-mwa-occupancy}
@@
 
-\subsection{Comparison with LOFAR}
+\section{Comparison with LOFAR}
 The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies.
@@
 It should be noted that the RFI detection for LOFAR was performed at 0.78~kHz while for the MWA it is performed at 40~kHz.
-As shown in O2012, LOFAR observes RFI that is only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions.
+As shown in O2012, LOFAR observes many RFI events that are only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions.
 O2012 also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA.
@@
 An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations.
-Taking into account these differences, it becomes hard to compare the environments fairly based on detected RFI levels.
-Nevertheless, for both telescopes, the automatic detection strategies have been optimized such that as little data as possible is thrown away but the detection is sufficient for further data reduction.
-For the MWA, this goal was reached by throwing away a factor of two less data.
+Because of these differences, it is hard to compare the environments based on detected RFI levels.
+Nevertheless, for both telescopes, the automatic detection strategies have been optimized such that as little data as possible are thrown away, but to be sufficient for further data reduction.
 
 After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging.
-While the difference between a 1.65\% or 3.18\% loss of data is hardly a reason to spend resources to build at a remote site like the Murchison Radio Observatory, there are additional reasons to do so.
-A benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the FM-station bands (88--108~MHz).
+From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar computational requirements as for LOFAR.
+Both RFI occupancy levels are so small that the RFI contamination in the 115--163~MHz range should not be a consideration in selecting a site.
+However, a benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station bands.
 Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}.
 Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR.
-Such a detection requires very deep integrations, and it could be that RFI becomes problematic when reaching lower noise levels.
-However, initial results with LOFAR predict that RFI will not prevent a detection of EoR signals \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.
+The required integration time for such a detection of approximately a hundred nights has currently not been achieved, and it could be that RFI becomes problematic when reaching lower noise levels.
+Initial results with LOFAR predict that RFI will not prevent such a detection \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.
 
-\section{Conclusions}
+\section{Conclusions \& Discussion}
+We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights.
+After RFI detection, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz.
+Also, DVD signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data.
+Over the full GLEAM range of 72--231~MHz, 1.1\% of the data is flagged by the RFI detection, and these RFI events are attributed to several different transmitters.
+Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection.
+The issue of RFI has become smaller by building at a remote site, but still requires adequate handling.
+
 \begin{itemize}
- \item Much less RFI than LOFAR (but at a cost)
- \item About half the observations don't need flagging at all. 
- \item FM not completely empty
  \item Frequency behaviour of instrument makes things harder! Need to fix pass-band correction.
+ \item High time resolution
+ \item Location of SKA is not free of RFI --> Need good ADCs etc
+ \item SKA -- how to implement a Cotter for SKA?
+ \item PAPER etc needs to use combinatorial/morphological flagging
+ \item The presence of strong interference implies there might be weak interference present that is not easily seen or detected. 
 \end{itemize}

### 2014-10-28 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Filled in all the text

@@
 In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency direction.
-Subsequently, line-shaped features are detected by the \textsc{sumthreshold} method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}.
+Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}.
 After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask.
@@
 The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID.
-For more details about the correlator, see TODO fix reference Ord et al.
-(2014).
+For more details about the correlator, see \citet{ord-2014-mwa-fpga-gpu}.
 Currently, the \textsc{cotter} preprocessing pipeline is ran by the scientist that calibrates and images the data.
@@
 One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline.
-The \textsc{sumthreshold} technique and the scale-invariant rank operator that are used by the flagging strategy, use statistics over large time and frequency ranges.
+The SumThreshold and SIR operator algorithms that are used by the flagging strategy, use statistics over large time and frequency ranges.
 Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible.
@@
 When less memory is available, \textsc{cotter} will split the observation in different time chunks and flags those independently.
-This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR (TODO reference?) to overcome the time-ordering problem.
+This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem.
 Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy.
@@
 We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time.
-The MWA Voltage Capture System (VCS; TODO cite), which allows high time resolution observations, might help to analyze these phenomena.
+The MWA Voltage Capture System (VCS; \citealt{mwa-voltage-capture-2014}), which allows high time resolution observations, might help to analyze these phenomena.
 
@@
 Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent.
-A high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup, is used to time \textsc{cotter}.
-The wall-clock runtime for processing a single 2~min snapshot with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, is split up as follows: 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data.
+To time \textsc{cotter}, we use a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup.
+The wall-clock runtime for processing a single 2~min snapshot of 50~GB with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, is split up as follows: 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data.
 Real-time processing can therefore be achieved by using 6 of such nodes in parallel.
@@
 Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection.
-The issue of RFI has become smaller by building at a remote site, but still requires adequate handling.
+The issue of RFI has become smaller by building at a remote site, but still requires adequate mitigation.
 
-\begin{itemize}
- \item Frequency behaviour of instrument makes things harder! Need to fix pass-band correction.
- \item High time resolution
- \item Location of SKA is not free of RFI --> Need good ADCs etc
- \item SKA -- how to implement a Cotter for SKA?
- \item PAPER etc needs to use combinatorial/morphological flagging
- \item The presence of strong interference implies there might be weak interference present that is not easily seen or detected. 
-\end{itemize}
+SKA-low will be build at the same location as the MWA, and hence several lessons can be learned from the MWA.
+First of all, it is clear that the SKA will need to be able to handle some amount of RFI.
+This includes the availability of computational power to perform the RFI detection, analogue-ditigal converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI TODO say something about base design?.
+\textsc{cotter} relies on the fact that a standard desktop can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA.
+Flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains.
+For accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth.
+
+The \textsc{aoflagger} RFI detection strategy, originally developed for LOFAR, works well for the MWA.
+Faint RFI events such as the ones in Figs.~\ref{fig:broadband-pulse} and \ref{fig:156_7}, or complex events such as the one in Fig.~\ref{fig:dvb-burst}, are not adequately detected by a single-sample thresholding algorithm, but \textsc{aoflagger}'s SumThreshold and SIR-operator algorithms are able to flag such events.
+These algorithms have gained some popularity (e.g.
+TODO Miriad, eMERLIN), but other projects still use single-sample thresholding, e.g.
+PAPER.
+Since strong RFI is practically seen everywhere, and because the apparent strength of RFI events will follow a power-law distribution, faint transmitters will interfere with observations of any (terrestrial) radio observatory.
+Using algorithms with low sensitivity will detect fewer of these, and thus result in RFI becoming more quickly a problem in long-integration projects.
+
+The strategy does not perform well on continuous broadband RFI such as DVB, which is occassionally present due to tropospheric ducting or ionospheric activity.
+To remove DVB, a second detection round is required, and the current methodology to handle DVB RFI is to delete all affected snapshots, or even the entire night.
+This requires hardly any computational power, because the visibility statistics are collected in \textsc{cotter}.
+The current sensitivity of the flagger is enough for calibration and imaging of MWA data.
+For deep-integration projects, such as the EoR projects, it might be that low-level RFI will show up at lower noise levels.
+One way to increase the sensitivity of the flagger, is by executing the flagger on the summed power of multiple baselines.
+Fig.~\ref{fig:156_7} shows that this increases the detectability of an RFI event significantly.
 

### 2014-10-28 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Finsihed the first draft of the MWA RFI environment article :-)

@@
 In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of radio observatories.
-Spectrum allocation management and radio-quiet zones help to limit the interference, but do not solve interference from air-based transmitters or accidential electromagnetic radiation, for example from cars or wind turbines.
+Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-based transmitters or accidential electromagnetic radiation, for example from cars or wind turbines.
 
 While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI.
-Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{post-correlation-rfi-classification}.
-The consequence of this approach is that a certain amount of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed.
+Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}.
+The consequence of this approach is that certain ratio of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed.
 It is important to analyze the impact of RFI on a specific instrument to consider the right RFI mitigation approach.
-Additionally, an increased understanding of the RFI situation will help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; chosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.
+An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; chosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.
 
@@
 
-The interferometric arrays LOFAR, PAPER and the Long Wavelength Array (LWA) are all designed to perform 21-cm EoR science, and all observe in approximately the same frequency range as the MWA.
-Comparing the MWA to these arrays therefore allows the comparison of the observatory site and hardware design with respect to the RFI impact.
-TODO say something about PAPER and MWA RFI.
-In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose issues on regular observations even though LOFAR is build in a populated area.
-This is confirmed by \citet{ncp-eor-yatawatta}, where the authors analyse the first imaging of EoR long-integration observations with LOFAR.
-They reach near-thermal noise sensitivity, and argue that RFI is not a limitation for further increasing the sensitivity.
+% TODO mention GMRT ? and the Long Wavelength Array (LWA; \citealt{lwa1-ellingson-2013})  ?
+The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA.
+For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization.
+Comparing the MWA to these arrays allows the comparison of the observatory site and hardware design with respect to the RFI impact.
+So far, initial observations with the MWA have produced the first scientific results without RFI issues (e.g.
+\citealt{mwacs-2014}), which of course is not surprising given its remote location.
+PAPER has shown that their radio environment in the Karoo desert of South Africa has been sufficiently clean to integrate sufficiently long to reach a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, which is one magnitude away from expected EoR signals.
+However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI.
+In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose issues on regular observations, even though LOFAR is build in a populated area.
+This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR.
+They argue that RFI is not a limitation for further increasing the sensitivity.
 Further analysis showed that RFI models for LOFAR can estimate the leakage of the RFI detection \citep{offringa-rfi-distributions}.
-Those analyses concluded that with sufficient precautions, such as good receiver design and accurate detection methods, the leaked RFI is weak and averages down similar to Gaussian noise.
+Those analyses conclude that with sufficient precautions, such as good receiver design, accurate detection methods and high time and frequency resolutions, the leaked RFI is weak and averages down similar to Gaussian noise.
+However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.
 
-TODO paper structure.
+In this paper, we will describe the mitigation efforts implemented for the MWA, and show examples of RFI found and results of the RFI detection.
+Sect.~\ref{ch:method} describes the approach taken for the MWA, including software, algorithms and computational challenges involved.
+This strategy is applied to 63~h of MWA data, and these data are described in Sect.~\ref{ch:data-description}.
+Sect.~\ref{ch:detection-results} presents the RFI that was found in these data, as well as the effectivity of the RFI detection.
+In Sect.~\ref{ch:lofar-comparison}, these results will be compared to a previously performed LOFAR RFI survey.
+Finally, conclusions are drawn and discussed in Sect.~\ref{ch:conclusions-and-discussion}.
 
-\section{Method}
+\section{Method} \label{ch:method}
 RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing.
@@
 
-\section{Data description}
+\section{Data description} \label{ch:data-description}
 We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall).
@@
 
-\section{Detection results}
+\section{Detection results} \label{ch:detection-results}
 The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations.
@@
 
-\section{Comparison with LOFAR}
+\section{Comparison with LOFAR} \label{ch:lofar-comparison}
 The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies.
@@
 An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations.
-Because of these differences, it is hard to compare the environments based on detected RFI levels.
+
+Because of the differences in resolution and the beam forming, it is hard to compare the LOFAR and MWA environments based on detected RFI levels.
 Nevertheless, for both telescopes, the automatic detection strategies have been optimized such that as little data as possible are thrown away, but to be sufficient for further data reduction.
+These values can therefore be interpreted and compared as being the minimum loss of data due to RFI.
 
@@
 
-\section{Conclusions \& Discussion}
+\section{Conclusions \& discussion} \label{ch:conclusions-and-discussion}
 We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights.
@@
 
+\begin{itemize}
+ \item TODO Say something about LEDA (LWA project) and HERA (Parsons' project)
+\end{itemize}
+
 \section*{Acknowledgments}

### 2014-10-29 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Small textualities

@@
 Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}.
-The consequence of this approach is that certain ratio of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed.
-It is important to analyze the impact of RFI on a specific instrument to consider the right RFI mitigation approach.
+The consequence of this approach is that a certain ratio of data is lost due to interference, and that frequencies that are continuously occupied by transmitters cannot be observed.
+It is important to analyze the impact of RFI on a specific instrument to optimize the RFI mitigation approach.
 An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; chosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.
@@
 \citealt{mwacs-2014}), which of course is not surprising given its remote location.
-PAPER has shown that their radio environment in the Karoo desert of South Africa has been sufficiently clean to integrate sufficiently long to reach a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, which is one magnitude away from expected EoR signals.
+PAPER has shown that their radio environment in the Karoo desert of South Africa has been sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength.
 However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI.
@@
 This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR.
-They argue that RFI is not a limitation for further increasing the sensitivity.
-Further analysis showed that RFI models for LOFAR can estimate the leakage of the RFI detection \citep{offringa-rfi-distributions}.
-Those analyses conclude that with sufficient precautions, such as good receiver design, accurate detection methods and high time and frequency resolutions, the leaked RFI is weak and averages down similar to Gaussian noise.
+They argue that RFI is not a limitation for further increasing the sensitivity, and conclude in \citet{offringa-rfi-distributions} that with sufficient precautions, such as good receiver design, accurate detection methods and high time and frequency resolutions, residual RFI is weak and averages down similar to Gaussian noise.
 However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.
 
-In this paper, we will describe the mitigation efforts implemented for the MWA, and show examples of RFI found and results of the RFI detection.
+In this paper, we will describe the mitigation strategy implemented for the MWA, and show examples of RFI found and results of the RFI detection.
 Sect.~\ref{ch:method} describes the approach taken for the MWA, including software, algorithms and computational challenges involved.
-This strategy is applied to 63~h of MWA data, and these data are described in Sect.~\ref{ch:data-description}.
-Sect.~\ref{ch:detection-results} presents the RFI that was found in these data, as well as the effectivity of the RFI detection.
-In Sect.~\ref{ch:lofar-comparison}, these results will be compared to a previously performed LOFAR RFI survey.
-Finally, conclusions are drawn and discussed in Sect.~\ref{ch:conclusions-and-discussion}.
+This strategy is applied to 63~h of MWA data.
+In Sect.~\ref{ch:data-description}, these data are described.
+Sect.~\ref{ch:detection-results} presents examples of RFI that were found in these data, as well as the effectivity of the RFI detection.
+In Sect.~\ref{ch:lofar-comparison}, the results are compared to a previously performed LOFAR RFI survey.
+Finally, in Sect.~\ref{ch:conclusions-and-discussion} conclusions are drawn and discussed.
 
@@
 For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used.
-It has been shown that this flagger has a good accuracy and is fast.
-It also has a library interface, which allows it to be integrated in a pipeline.
+It has been shown that this flagger has a good accuracy and is fast \citep{lofar-radio-environment}.
+It also has a library interface\footnote{The documentation for the \textsc{aoflagger} library interface can be found at \url{http://aoflagger.sourceforge.net/doc/api/}}, which allows it to be integrated in a pipeline.
 
@@
 Because not all astronomers have easily access to high-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations.
-For this use-case, \textsc{cotter} is run in a flagging-only mode on the high-memory machine.
+In this use-case, \textsc{cotter} is run in a flagging-only mode on the high-memory machine.
 The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to convert the raw files and apply the RFI detection from the first run.
@@
 \begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
-\caption{RFI occupancy per subband, calculated over all observations except GLEAM 2014-03-17. The latter has been left out because it is affected by DVB.}
+\caption{RFI occupancy per subband, calculated over all observation nights except GLEAM 2014-03-17. The latter has been left out because it is affected by DVB.}
 \label{fig:rfi-per-band}
@@
 The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data.
-Excluding the RFI from DVB, the EoR high band is slightly cleaner than the EoR low band, as its worst subband at 188.2 MHz has 1.03\% occupancy.
+Excluding the RFI from DVB, the EoR high band is slightly cleaner than the EoR low band, and its worst subband at 188.2 MHz has 1.03\% occupancy.
 The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy.
@@
 
-Because significant DVB interference is left in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively.
-The DVB transmitters are terrestrial, and the fact that only this night observes the DVB, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity.
+Because significant DVB interference residuals are visible in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively.
+The DVB transmitters are terrestrial, and the fact that only this night observes the DVB, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity or tropospheric ducting.
 As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can fully contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth.
@@
 Strong pulses such as these are rare and well flagged, but a few weak broad-band pulses are observed in almost every 2-min snapshot.
-These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines are added together.
+These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines is added together.
 We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time.
@@
 This includes the availability of computational power to perform the RFI detection, analogue-ditigal converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI TODO say something about base design?.
-\textsc{cotter} relies on the fact that a standard desktop can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA.
-Flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains.
+Second, \textsc{cotter} relies on the fact that a standard desktop can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA.
+Finally, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains.
 For accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth.
@@
 PAPER.
-Since strong RFI is practically seen everywhere, and because the apparent strength of RFI events will follow a power-law distribution, faint transmitters will interfere with observations of any (terrestrial) radio observatory.
+Since strong RFI is practically seen everywhere, and because the apparent strength of RFI events will follow a power-law distribution, many faint transmitters will interfere with observations of any (terrestrial) radio observatory.
 Using algorithms with low sensitivity will detect fewer of these, and thus result in RFI becoming more quickly a problem in long-integration projects.

### 2014-10-31 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Randall's comments

@@
 In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of radio observatories.
-Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-based transmitters or accidential electromagnetic radiation, for example from cars or wind turbines.
+Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-based or satellite transmitters or accidential electromagnetic radiation, for example from cars or wind turbines.
 
@@
 Its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
-To avoid as much RFI as possible, the MWA is located in the Murchison region in Western Australia.
-Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be build near the location of the MWA.
+To avoid as much RFI as possible, the MWA is located in the Murchison Radio-astronomy Obervatory in the Murchison region of Western Australia.
+Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.
+TODO mention something about the extent to which protections have been put in place for radio quietness at the MRO.
+Randall: ``I'm sure there's a document somewhere that describes it, otherwise you could contact Carol Wilson at CSIRO who's responsible for such things.
+(carol.wilson@csiro.au)''
 
@@
 However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.
+
+Recently, experiments to use the moon as a calibrator for EoR experiments have shown that the reflection of the moon of terrestrial transmitters complicates such an experiment \citep{mckinley-moon-2013}.
+However, reflections are spatially isolated to the centre of the observed disk caused by the moon, and the high resolution of LOFAR allows separating the reflected and intrinsic power from the moon \citep{vedantham-2014-todo}.
+In another study it is described that objects in space of about a meter in size, such as satellites and space debris, also reflect terrestrial transmissions with enough strength to be observable by the MWA \citep{tingay-space-debris-2013}.
+While tracking space debris is a useful asset, such reflected RFI can be a problem for EoR experiments, especially all-sky experiments that try to measure the global EoR signal \citep{vedantham-2014-todo}.
 
@@
 The \textsc{aoflagger} allows customizing its flagging strategy, such as changing the threshold levels and expected smoothness of good data.
-Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions and varying field of views.
+Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions and varying fields of view.
 Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA specific strategy which is used in this work.
@@
 Not all observations cover the DVB frequencies.
-}\label{tbl:obs-list}
+(TODO this table excludes RF11 and RF12 DVB interference; needs fix)}\label{tbl:obs-list}
 \begin{center}%
@@
 visible in Fig.~\ref{fig:bandpass}.
-The poly-phase filter band-pass shape is hard to take out completely, because the filters are imperfect and each sub-band is effected by leakage from signals in its adjacent sub-bands.
+The poly-phase filter band-pass shape is hard to take out completely, because the filters are imperfect and each sub-band is affected by leakage from signals in its adjacent sub-bands.
 The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent.
@@
 In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations.
-Observation `GLEAM 2014-03-17' shows residual DVB interference, and it is clear that this RFI has not been adequately flagged.
+Observation `GLEAM 2014-03-17' shows residual DVB interference, both in the frequency range 174--195~MHz (RF 6, 7 and 8) and 216--230~MHz (RF 11 and 12), and it is clear that this RFI has not been adequately flagged.
 Therefore, observations have to be cleaned of DVB with another method.
@@
 The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies.
-The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}.
+The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency subrange of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}.
 The statistics are regridded to the same frequency resolution of 48~kHz.
@@
 O2012 also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA.
-In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources: Cassiopeia~A is never visible at the MWA, and Cygnus~A is only visible at low declinations.
+In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources: Cassiopeia~A is never visible at the MWA, and Cygnus~A is only visible at low elevations.
 An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations.
@@
 Both RFI occupancy levels are so small that the RFI contamination in the 115--163~MHz range should not be a consideration in selecting a site.
-However, a benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station bands.
+However, a benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DVB bands.
 Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}.
@@
 After RFI detection, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz.
-Also, DVD signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data.
+Also, DVB signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data.
 Over the full GLEAM range of 72--231~MHz, 1.1\% of the data is flagged by the RFI detection, and these RFI events are attributed to several different transmitters.
@@
 This includes the availability of computational power to perform the RFI detection, analogue-ditigal converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI TODO say something about base design?.
-Second, \textsc{cotter} relies on the fact that a standard desktop can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA.
+Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA.
 Finally, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains.

### 2014-11-01 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Adding abstract and some more things

@@
 
-\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\editmark{Corresponding author.} E-mail: \url{andre.offringa@anu.edu.au}},
+\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\editmark{Corresponding author.} E-mail: \url{offringa@gmail.com}},
 R.~B.~Wayth\Curtin$^,$\CAASTRO,
@@
 \begin{document}
+\begin{abstract}
+The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at the location of the Square Kilometre Array (SKA).
+We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and RFI statistics are shown in the range 72--231~MHz and over various nights.
+RFI detection removes 1.1\% of the data, and another $\sim3\%$ is lost because of occasional ionospheric or atmospheric propagation of RFI from digital TV.
+After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DVB bands.
+The results are compared to a previously published LOFAR RFI survey.
+The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before reduction of observations.
+Advises are given to make the SKA robust to RFI, which include: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient bits; a smooth band-pass response and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
+\end{abstract}
+
 \maketitle
@@
 \label{firstpage}
-
-\begin{abstract}
-...
-\end{abstract}
 
@@
 
-In this article, we will look specifically at the RFI situation of the MWA \citep{mwa}.
+In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa}).
 The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 80--300~MHz with a 31-MHz instantaneous bandwidth.
@@
 % TODO mention GMRT ? and the Long Wavelength Array (LWA; \citealt{lwa1-ellingson-2013})  ?
+\defcitealias{lofar-radio-environment}{O2013a}
 The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA.
@@
 \citealt{mwacs-2014}), which of course is not surprising given its remote location.
-PAPER has shown that their radio environment in the Karoo desert of South Africa has been sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength.
+PAPER has shown that their radio environment in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength.
 However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI.
-In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose issues on regular observations, even though LOFAR is build in a populated area.
+In \citet{lofar-radio-environment} (hereafter referred to as \citetalias{lofar-radio-environment}) it is shown that the radio environment of LOFAR does not pose issues on regular observations, even though LOFAR is build in a populated area.
 This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR.
@@
 RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing.
-It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}.
+It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citepalias{lofar-radio-environment}.
 Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays.
@@
 For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used.
-It has been shown that this flagger has a good accuracy and is fast \citep{lofar-radio-environment}.
+It has been shown that this flagger has a good accuracy and is fast \citepalias{lofar-radio-environment}.
 It also has a library interface\footnote{The documentation for the \textsc{aoflagger} library interface can be found at \url{http://aoflagger.sourceforge.net/doc/api/}}, which allows it to be integrated in a pipeline.
@@
 Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA specific strategy which is used in this work.
-The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citet{lofar-radio-environment}.
+The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citetalias{lofar-radio-environment}.
 In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency direction.
@@
 The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations.
-This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% detections.
+This includes false detections, which are estimated in \citetalias{lofar-radio-environment} to account for approximately 0.4\% detections.
 Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI.
@@
 The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy.
-TODO what are these transmitters?
 
@@
 The GLEAM and EoR low-band observations do show a small power-law part, but the fitted exponent depends strongly on what part of the tail is selected.
-The selected regions result in $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and Eor-low curves, respectively.
+The selected regions result in $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and EoR-low curves, respectively.
 Due to the small part over which the power laws hold, as well as the high possible error in the fits because a fitting region needs to be selected, it is hard to infer if a uniform distribution of transmitters is observed.
@@
 Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent.
-To time \textsc{cotter}, we use a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup.
+To time \textsc{cotter}, we use a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5-disc RAID setup.
 The wall-clock runtime for processing a single 2~min snapshot of 50~GB with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, is split up as follows: 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data.
 Real-time processing can therefore be achieved by using 6 of such nodes in parallel.
+TODO mention samples/minute and that this is constant.
 
@@
 The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies.
-The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency subrange of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}.
+The RFI occupancies for LOFAR from \citetalias{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency subrange of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}.
 The statistics are regridded to the same frequency resolution of 48~kHz.
-The average RFI occupancy in this frequency range is 1.65\% for the MWA, while O2012 reports a 3.18\% occupancy for LOFAR.
+The average RFI occupancy in this frequency range is 1.65\% for the MWA, while \citetalias{lofar-radio-environment} reports a 3.18\% occupancy for LOFAR.
 
 It should be noted that the RFI detection for LOFAR was performed at 0.78~kHz while for the MWA it is performed at 40~kHz.
-As shown in O2012, LOFAR observes many RFI events that are only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions.
-O2012 also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA.
+As shown by \citetalias{lofar-radio-environment}, LOFAR observes many RFI events that are only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions.
+\citetalias{lofar-radio-environment} also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA.
 In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources: Cassiopeia~A is never visible at the MWA, and Cygnus~A is only visible at low elevations.
@@
 The issue of RFI has become smaller by building at a remote site, but still requires adequate mitigation.
+When observing continuously with the MWA, a few fast computing nodes are permanently required for real-time RFI detection.
 
@@
 Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA.
-Finally, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains.
+Thirdly, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains.
 For accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth.
+Finally, the presence of short and spectrally-narrow RFI events confirms that detection at high time and frequency resolution improves the accuracy.
 

### 2014-11-03 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Finished TODO's, article ready for reviews

@@
 The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at the location of the Square Kilometre Array (SKA).
-We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and RFI statistics are shown in the range 72--231~MHz and over various nights.
-RFI detection removes 1.1\% of the data, and another $\sim3\%$ is lost because of occasional ionospheric or atmospheric propagation of RFI from digital TV.
-After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DVB bands.
+We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights.
+RFI detection removes 1.1\% of the data, and another $\sim3\%$ is lost because of occasional ionospheric or atmospheric propagation of RFI from digital TV (DTV).
+After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands.
 The results are compared to a previously published LOFAR RFI survey.
 The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before reduction of observations.
-Advises are given to make the SKA robust to RFI, which include: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient bits; a smooth band-pass response and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
+Some advise is deducted to make the SKA robust to RFI, which include: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient bits; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
 \end{abstract}
@@
 Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.
-TODO mention something about the extent to which protections have been put in place for radio quietness at the MRO.
-Randall: ``I'm sure there's a document somewhere that describes it, otherwise you could contact Carol Wilson at CSIRO who's responsible for such things.
-(carol.wilson@csiro.au)''
 
-% TODO mention GMRT ? and the Long Wavelength Array (LWA; \citealt{lwa1-ellingson-2013})  ?
+Several measures are in place to protect the radio quietness of the MRO: the frequency plan of the Australian Communications and Media Authority (ACMA) states that spectrum licenses cannot interfere within a radius of 70~km around the centre of the MRO.
+Within a radius of 150~km, consultation is required.
+Licenses for individual apparatus are controlled by the Radiocommunications Assignment and Licensing Instruction (RALI, document MS 32), which prescribes a 150~km restricted and 230~km consultation zone for frequencies 100--230~MHz, and therefore covers most of the bandwidth of the MWA.
+Cases that are not covered by the radio-quiet zone are: i) transmitters with frequencies below 70~MHz; ii) aircraft transmissions; iii) satellite transmissions; and iv) transmitters beyond 260~km from the centre of the MRO.
+
 \defcitealias{lofar-radio-environment}{O2013a}
@@
 However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI.
-In \citet{lofar-radio-environment} (hereafter referred to as \citetalias{lofar-radio-environment}) it is shown that the radio environment of LOFAR does not pose issues on regular observations, even though LOFAR is build in a populated area.
+In \citeauthor{lofar-radio-environment} (\citeyear{lofar-radio-environment}; hereafter referred to as \citetalias{lofar-radio-environment}) it is shown that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is build in a populated area.
 This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR.
@@
 RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing.
-It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citepalias{lofar-radio-environment}.
+It is important to perform initial RFI detection at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citepalias{lofar-radio-environment}.
 Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays.
@@
 The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (which includes false positives).
-The DVB column displays the fraction of visibilities that were unusable because of the presence of DVB signals, and which the flagger does not flag.
-Not all observations cover the DVB frequencies.
-(TODO this table excludes RF11 and RF12 DVB interference; needs fix)}\label{tbl:obs-list}
+Occasionally, interference from digital TV (DTV) is visible.
+The DTV column displays the fraction of visibilities that were unusable because of the presence of DTV signals, and which the flagger does not flag.
+Not all observations cover the DTV frequencies.}\label{tbl:obs-list}
 \begin{center}%
@@
 \hline
-\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DVB \\
+\textbf{Project} &\textbf{Date} & \textbf{Frequencies (MHz)} & \textbf{Duration} & RFI & DTV \\
 \hline
@@
 GLEAM & 2014-03-16 & 72.3--133.8, 138.9--230.8 & 7h & 0.79\% & 0\% \\
-GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 30\% \\
+GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 1.29\% \\
 EoR high& 2014-04-10 & 167.0--197.7              & 6h & 0.68\% & 0\%\\
@@
 \hline
-\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.96\%} & \textbf{3\%}\\
+\multicolumn{3}{|r|}{\textbf{Total}} & \textbf{63h} & \textbf{0.96\%} & \textbf{0.14\%}\\
 \hline
@@
 \begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
-\caption{RFI occupancy per subband, calculated over all observation nights except GLEAM 2014-03-17. The latter has been left out because it is affected by DVB.}
+\caption{RFI occupancy per subband, calculated over all observation nights except GLEAM 2014-03-17. The latter has been left out because it is affected by DTV.}
 \label{fig:rfi-per-band}
@@
 This is the residual noise in the visibilities after RFI detection.
-A few channels in the FM bands show a deviating standard deviation, and observation `GLEAM 2014-03-17' shows DVB contamination around 180~MHz.}
+A few channels in the FM bands show a deviating standard deviation, and observation `GLEAM 2014-03-17' shows DTV contamination around 180~MHz.}
 \label{fig:stddev-per-set}
@@
 Fig.~\ref{fig:rfi-per-band} shows the overal detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17.
-The latter is the only night affected by interference from digital video broadcasting (DVB), and will be analysed later.
+The latter is the only night affected by interference from digital TV (DTV), and will be analysed later.
 The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data.
-Excluding the RFI from DVB, the EoR high band is slightly cleaner than the EoR low band, and its worst subband at 188.2 MHz has 1.03\% occupancy.
+Excluding the RFI from DTV, the EoR high band is slightly cleaner than the EoR low band, and its worst subband at 188.2 MHz has 1.03\% occupancy.
 The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy.
@@
 In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations.
-Observation `GLEAM 2014-03-17' shows residual DVB interference, both in the frequency range 174--195~MHz (RF 6, 7 and 8) and 216--230~MHz (RF 11 and 12), and it is clear that this RFI has not been adequately flagged.
-Therefore, observations have to be cleaned of DVB with another method.
+Observation `GLEAM 2014-03-17' shows residual DTV interference, both in the frequency range 174--195~MHz (RF 6, 7 and 8) and 216--230~MHz (RF 11 and 12), and it is clear that this RFI has not been adequately flagged.
+Therefore, DTV interference has to be detected with another method.
 Additionally, some channels in the FM radio band show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and possibly because of RFI leakage.
@@
 The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI.
-With the exception of the ORBCOMM frequencies and DVB affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts.
+With the exception of the ORBCOMM frequencies and DTV affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts.
 This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.
@@
 \end{center}
-\caption{RFI from Digital Video Broadcasting (DVB), which is visible in 1 of the 9 observations (GLEAM~2014-03-17) that cover the DVB frequencies.
-Left plot: dynamic spectrum for the worst affected snapshot; right plot: visibility RMS in a few of the DVB bands for the affected night after flagging.
+\caption{RFI from digital TV (DTV), which is visible in 1 of the 9 observations (GLEAM~2014-03-17) that cover the DTV frequencies.
+Left plot: dynamic spectrum for the worst affected snapshot; right plot: visibility RMS in a few of the DTV bands for the affected night after flagging.
 Because of its broadband nature, this kind of RFI is not well flagged in the initial \textsc{aoflagger} step, but it is detectable in the global statistics.
@@
 
-Because significant DVB interference residuals are visible in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively.
-The DVB transmitters are terrestrial, and the fact that only this night observes the DVB, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity or tropospheric ducting.
+Because significant DTV interference residuals are visible in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively.
+The DTV transmitters are terrestrial, and the fact that only this night observes the DTV, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity or tropospheric ducting.
 As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can fully contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth.
@@
 
-Although the flagger does not adequately flag DVB, the right plot of Fig.~\ref{fig:dvb} shows that it is possible to determine whether a snapshot is affected by DVB interference from the data statistics.
-An option is to test whether the visibility RMS of a snapshot in any of the occupied DVB bands exceeds the RMS in DVB band 4, which is not used for broadcasting.
+Although the flagger does not adequately flag DTV, the right plot of Fig.~\ref{fig:dvb} shows that it is possible to determine whether a snapshot is affected by DTV interference from the data statistics.
+An option is to test whether the visibility RMS of a snapshot in any of the occupied DTV bands exceeds the RMS in DTV band 4, which is not used for broadcasting.
 These statistics are calculated by \textsc{cotter}, hence this can be validated without having to read the data again.
-When DVB is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used.
+When DTV is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used.
 However, it might not be worth it to have special mechanisms in place, such as having to account for lower SNR and a decreased $uv$ coverage, to salvage the residual good data within an affected snapshot.
-Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability of observing DVB is roughly 3\%.
-This is detailed in the last column of Table~\ref{tbl:obs-list}.
+Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability that DTV interference occurs is roughly 3\%.
+The bands involved are RF 5, 6, 7, 11 and 12, each of 7 MHz.
+Therefore, when DTV interference occurs, it affects 35 of the 159~MHz bandwidth.
+The visibility ratios that are lost because of DTV RFI are detailed in the last column of Table~\ref{tbl:obs-list}.
+Because these frequencies correspond to the EoR redshift, in reality these frequencies are observed more than average, so the actual loss is somewhat higher.
 
@@
 Separate distributions are shown for visibilities that have been detected as RFI (red lines) and those that have been classified as RFI free (green lines).
-The EoR high-band curves do not include the night that was contaminated by DVB RFI.
+The EoR high-band curves do not include the night that was contaminated by DTV RFI.
 
@@
 \end{center}
-\caption{DVB RFI briefly visible in an EoR high-band observation. The plots show the Stokes~I amplitudes for a single correlation. The right plot shows the result of RFI detection marked in purple.}
+\caption{DTV RFI briefly visible in an EoR high-band observation. The plots show the Stokes~I amplitudes for a single correlation. The right plot shows the result of RFI detection marked in purple.}
 \label{fig:dvb-burst}
@@
 Besides persistent transmissions that occupy a few channels, transient broad-band events are observed as well.
-Occassionally, DVB RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}.
+Occassionally, DTV RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}.
 As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}.
@@
 Real-time processing can therefore be achieved by using 6 of such nodes in parallel.
-TODO mention samples/minute and that this is constant.
+Assuming a 138~GFLOPS (Giga-floating-point operations per second) performance of the host computer, the RFI detection requires 25 FLOP/sample.
+When expressed as visibilities per time unit, the computational performance of the flagger is independent of the frequency resolution, time resolution and number of antennas.
+This number can therefore be extrapolated to other telescopes, although the performance of a pipeline which incorporates RFI detection will be strongly dependent on available IO performance and memory bandwidth.
 
@@
 Both RFI occupancy levels are so small that the RFI contamination in the 115--163~MHz range should not be a consideration in selecting a site.
-However, a benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DVB bands.
+However, a benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DTV bands.
 Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}.
@@
 After RFI detection, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz.
-Also, DVB signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data.
+Also, DTV signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data.
 Over the full GLEAM range of 72--231~MHz, 1.1\% of the data is flagged by the RFI detection, and these RFI events are attributed to several different transmitters.
@@
 First of all, it is clear that the SKA will need to be able to handle some amount of RFI.
-This includes the availability of computational power to perform the RFI detection, analogue-ditigal converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI TODO say something about base design?.
+This includes the availability of computational power to perform the RFI detection, analogue-ditigal converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI.
 Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA.
@@
 Faint RFI events such as the ones in Figs.~\ref{fig:broadband-pulse} and \ref{fig:156_7}, or complex events such as the one in Fig.~\ref{fig:dvb-burst}, are not adequately detected by a single-sample thresholding algorithm, but \textsc{aoflagger}'s SumThreshold and SIR-operator algorithms are able to flag such events.
-These algorithms have gained some popularity (e.g.
-TODO Miriad, eMERLIN), but other projects still use single-sample thresholding, e.g.
-PAPER.
+These algorithms have gained some popularity (e.g., besides the use of \textsc{aoflagger} by individual astronomers, the SumThreshold algorithm is implemented in the pipeline for eMERLIN \citep{serpent-peck-2013} and in Miriad), but other projects still use single-sample thresholding, e.g.
+PAPER \citep{parsons-paper-eorlimit-2014}.
 Since strong RFI is practically seen everywhere, and because the apparent strength of RFI events will follow a power-law distribution, many faint transmitters will interfere with observations of any (terrestrial) radio observatory.
@@
 
-The strategy does not perform well on continuous broadband RFI such as DVB, which is occassionally present due to tropospheric ducting or ionospheric activity.
-To remove DVB, a second detection round is required, and the current methodology to handle DVB RFI is to delete all affected snapshots, or even the entire night.
+The strategy does not perform well on continuous broadband RFI such as DTV, which is occassionally present due to tropospheric ducting or ionospheric activity.
+To remove DTV, a second detection round is required, and the current methodology to handle DTV RFI is to delete all affected snapshots, or even the entire night.
 This requires hardly any computational power, because the visibility statistics are collected in \textsc{cotter}.
@@
 Fig.~\ref{fig:156_7} shows that this increases the detectability of an RFI event significantly.
-
-\begin{itemize}
- \item TODO Say something about LEDA (LWA project) and HERA (Parsons' project)
-\end{itemize}
 

### 2014-11-03 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Spell check

@@
 In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of radio observatories.
-Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-based or satellite transmitters or accidential electromagnetic radiation, for example from cars or wind turbines.
+Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-based or satellite transmitters or accidental electromagnetic radiation, for example from cars or wind turbines.
 
@@
 The consequence of this approach is that a certain ratio of data is lost due to interference, and that frequencies that are continuously occupied by transmitters cannot be observed.
-It is important to analyze the impact of RFI on a specific instrument to optimize the RFI mitigation approach.
-An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; chosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.
+It is important to analyse the impact of RFI on a specific instrument to optimise the RFI mitigation approach.
+An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; choosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.
 
 In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa}).
-The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 80--300~MHz with a 31-MHz instantaneous bandwidth.
+The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 80--300~MHz with a 31-MHz instantaneous bandwidth.
 Its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
-To avoid as much RFI as possible, the MWA is located in the Murchison Radio-astronomy Obervatory in the Murchison region of Western Australia.
+To avoid as much RFI as possible, the MWA is located in the Murchison Radio-astronomy Observatory in the Murchison region of Western Australia.
 Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.
@@
 This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR.
-They argue that RFI is not a limitation for further increasing the sensitivity, and conclude in \citet{offringa-rfi-distributions} that with sufficient precautions, such as good receiver design, accurate detection methods and high time and frequency resolutions, residual RFI is weak and averages down similar to Gaussian noise.
+They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods and high time and frequency resolutions, residual RFI is weak and averages down similar to Gaussian noise.
 However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.
@@
 \caption{Correlator output RMS over frequency in an MWA high-frequency observation, calculated over all cross-correlated baselines and 112 seconds of data.
-In this band, the band-pass shows two large discontinuities over frequency, because the MWA receivers apply different digital gains at different frequencies to minimize quantization noise.
+In this band, the band-pass shows two large discontinuities over frequency, because the MWA receivers apply different digital gains at different frequencies to minimise quantisation noise.
 The 1.28~MHz pattern is due to the imperfect poly-phase filter.
@@
 To detect RFI in MWA observations, we have used the \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data.
-The \textsc{aoflagger} allows customizing its flagging strategy, such as changing the threshold levels and expected smoothness of good data.
+The \textsc{aoflagger} allows customising its flagging strategy, such as changing the threshold levels and expected smoothness of good data.
 Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions and varying fields of view.
@@
 For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation.
-This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimize quantization noise.
+This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimise quantisation noise.
 An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}.
@@
 The \textsc{aoflagger} software provides a C++ library %TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
-that can be integrated in a pipeline, thereby minimizing reading/writing of data.
+that can be integrated in a pipeline, thereby minimising reading/writing of data.
 We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection.
@@
 When time or frequency averaging is requested, \textsc{cotter} averages samples together that have not been flagged.
-When all input samples are flagged, the average of all input visibilities is stored into the the output sample, and the output sample is flagged.
+When all input samples are flagged, the average of all input visibilities is stored into the output sample, and the output sample is flagged.
 This method makes it possible to superficially analyse flagged samples in the output, even though information is lost in the averaging.
@@
 In practice, most observations are recorded at a resolution of 0.5 s and 40 kHz, and averaged to a resolution of 2--4 s and 40--80 kHz to reduce the data volume.
-This does not significantly cause any time or bandwidth decorrelation up to the first beam nul.
+This does not significantly cause any time or bandwidth decorrelation up to the first beam null.
 \textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $uvw$ values for the central time and frequency of the output sample.
@@
 \section{Data description} \label{ch:data-description}
-We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall).
+We analyse the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall).
 The nights of observations are listed in Table~\ref{tbl:obs-list}.
@@
 
-Both projects avoid the subbands that cover the frequency range 133.8--138.9~MHz, because the ORBCOMM low Earth orbitting (LEO) satellites transmit in these subbands.
-When these frequencies are observed, the subbands are often so strongly contaminated that it affects imaging sensitivity.
+Both projects avoid the sub-bands that cover the frequency range 133.8--138.9~MHz, because the ORBCOMM low Earth orbiting (LEO) satellites transmit in these sub-bands.
+When these frequencies are observed, the sub-bands are often so strongly contaminated that it affects imaging sensitivity.
 For completeness, we include a 16-min observation that covers this frequency range.
 
-The poly-phase filter bank that performs the first separation into subbands, introduces a 1.28~MHz periodic spectral signature.
-This shape is e.g.
-visible in Fig.~\ref{fig:bandpass}.
+The poly-phase filter bank that performs the first separation into sub-bands, introduces a 1.28~MHz periodic spectral signature, which is shown in Fig.~\ref{fig:bandpass}.
 The poly-phase filter band-pass shape is hard to take out completely, because the filters are imperfect and each sub-band is affected by leakage from signals in its adjacent sub-bands.
@@
 Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.
-
 \noindent\begin{figure*}
@@
 
-Fig.~\ref{fig:rfi-per-band} shows the overal detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17.
+Fig.~\ref{fig:rfi-per-band} shows the overall detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17.
 The latter is the only night affected by interference from digital TV (DTV), and will be analysed later.
@@
 
-Although the flagger does not adequately flag DTV, the right plot of Fig.~\ref{fig:dvb} shows that it is possible to determine whether a snapshot is affected by DTV interference from the data statistics.
+Although the flagger does not adequately flag DTV interference, the right plot of Fig.~\ref{fig:dvb} shows that from the data statistics it is possible to determine whether a snapshot is affected by DTV interference.
 An option is to test whether the visibility RMS of a snapshot in any of the occupied DTV bands exceeds the RMS in DTV band 4, which is not used for broadcasting.
@@
 The visibility ratios that are lost because of DTV RFI are detailed in the last column of Table~\ref{tbl:obs-list}.
-Because these frequencies correspond to the EoR redshift, in reality these frequencies are observed more than average, so the actual loss is somewhat higher.
+Because these frequencies overlap with the 21-cm HI frequencies redshifted to the EoR, in reality these frequencies are observed more than average, so the actual loss is somewhat higher.
 
@@
 In the right figures, the result of RFI detection is shown with purple.
-The horizontal flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
+The horizontal flagged lines are flagged because they are W-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
 \label{fig:2m-amateur-band}
@@
 Besides persistent transmissions that occupy a few channels, transient broad-band events are observed as well.
-Occassionally, DTV RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}.
+Occasionally, DTV RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}.
 As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}.
@@
 We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time.
-The MWA Voltage Capture System (VCS; \citealt{mwa-voltage-capture-2014}), which allows high time resolution observations, might help to analyze these phenomena.
+The MWA Voltage Capture System (VCS; \citealt{mwa-voltage-capture-2014}), which allows high time resolution observations, might help to analyse these phenomena.
 
@@
 The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies.
-The RFI occupancies for LOFAR from \citetalias{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency subrange of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}.
+The RFI occupancies for LOFAR from \citetalias{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency sub-range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}.
 The statistics are regridded to the same frequency resolution of 48~kHz.
@@
 Because of the differences in resolution and the beam forming, it is hard to compare the LOFAR and MWA environments based on detected RFI levels.
-Nevertheless, for both telescopes, the automatic detection strategies have been optimized such that as little data as possible are thrown away, but to be sufficient for further data reduction.
+Nevertheless, for both telescopes, the automatic detection strategies have been optimised such that as little data as possible are thrown away, but to be sufficient for further data reduction.
 These values can therefore be interpreted and compared as being the minimum loss of data due to RFI.
@@
 After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging.
-From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar computational requirements as for LOFAR.
+From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar per-sample computational requirements as for LOFAR.
 Both RFI occupancy levels are so small that the RFI contamination in the 115--163~MHz range should not be a consideration in selecting a site.
@@
 First of all, it is clear that the SKA will need to be able to handle some amount of RFI.
-This includes the availability of computational power to perform the RFI detection, analogue-ditigal converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI.
+This includes the availability of computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI.
 Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA.
@@
 Faint RFI events such as the ones in Figs.~\ref{fig:broadband-pulse} and \ref{fig:156_7}, or complex events such as the one in Fig.~\ref{fig:dvb-burst}, are not adequately detected by a single-sample thresholding algorithm, but \textsc{aoflagger}'s SumThreshold and SIR-operator algorithms are able to flag such events.
-These algorithms have gained some popularity (e.g., besides the use of \textsc{aoflagger} by individual astronomers, the SumThreshold algorithm is implemented in the pipeline for eMERLIN \citep{serpent-peck-2013} and in Miriad), but other projects still use single-sample thresholding, e.g.
+These algorithms have gained some popularity (e.g., besides the use of \textsc{aoflagger} by individual astronomers, the SumThreshold algorithm is implemented in the pipeline for eMERLIN \citep{serpent-peck-2013} and in \textsc{miriad}), but other projects still use single-sample thresholding, e.g.
 PAPER \citep{parsons-paper-eorlimit-2014}.
@@
 
-The strategy does not perform well on continuous broadband RFI such as DTV, which is occassionally present due to tropospheric ducting or ionospheric activity.
+The strategy does not perform well on continuous broadband RFI such as DTV, which is occasionally present due to tropospheric ducting or ionospheric activity.
 To remove DTV, a second detection round is required, and the current methodology to handle DTV RFI is to delete all affected snapshots, or even the entire night.
@@
 
-\section*{Acknowledgments}
+\section*{Acknowledgements}
 This scientific work makes use of the Murchison Radio-astronomy Observatory, operated by CSIRO.

### 2014-11-04 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Implemented Steven's comments (and Carol's comment)

@@
 
-\title[The Murchison radio environment]{The low-frequency Murchison radio environment:\\interference analysis and mitigation}
+\title[The MWA radio environment]{The low-frequency environment of the Murchison Widefield Array: radio frequency interference analysis and mitigation}
 
@@
 \begin{abstract}
-The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at the location of the Square Kilometre Array (SKA).
+The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at the location of the future Square Kilometre Array (SKA).
 We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights.
@@
 The results are compared to a previously published LOFAR RFI survey.
-The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before reduction of observations.
-Some advise is deducted to make the SKA robust to RFI, which include: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient bits; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
+The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before analysis of data.
+Some advise is suggested to make the SKA robust to RFI, which include: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient bits; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
 \end{abstract}
@@
 In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of radio observatories.
-Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-based or satellite transmitters or accidental electromagnetic radiation, for example from cars or wind turbines.
+Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-born or satellite transmitters or accidental electromagnetic radiation, for example from cars or wind turbines.
 
@@
 
-In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa}).
-The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 80--300~MHz with a 31-MHz instantaneous bandwidth.
-Its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
-To avoid as much RFI as possible, the MWA is located in the Murchison Radio-astronomy Observatory in the Murchison region of Western Australia.
+In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}).
+The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 80--300~MHz with a 30.72-MHz instantaneous bandwidth; its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
+To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory in the Murchison region of Western Australia.
 Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.
 
-Several measures are in place to protect the radio quietness of the MRO: the frequency plan of the Australian Communications and Media Authority (ACMA) states that spectrum licenses cannot interfere within a radius of 70~km around the centre of the MRO.
-Within a radius of 150~km, consultation is required.
-Licenses for individual apparatus are controlled by the Radiocommunications Assignment and Licensing Instruction (RALI, document MS 32), which prescribes a 150~km restricted and 230~km consultation zone for frequencies 100--230~MHz, and therefore covers most of the bandwidth of the MWA.
-Cases that are not covered by the radio-quiet zone are: i) transmitters with frequencies below 70~MHz; ii) aircraft transmissions; iii) satellite transmissions; and iv) transmitters beyond 260~km from the centre of the MRO.
+Several levels of protection are in place to protect the radio quietness of the MRO.
+Within a 70~km radius, the Australian Communications and Media Authority\footnote{\url{http://www.acma.gov.au/}} and WA government provide the strongest level of protection from other radio equipment across the frequency range 70~MHz to 25.25~GHz.
+Radio devices in this zone must not cause interference to radio astronomy.
+Beyond 70~km, coordination zones extend out to~260 km radius at the lowest frequencies and reduce in size with increasing frequency.
+Individually licensed or spectrum-licensed radio devices in these zones must be coordinated with radio-astronomy requirements to eliminate or minimize interference.
+Situations not covered by the radio-quiet zone are: i) transmitters below 70~MHz; ii) most aircraft transmissions; iii) satellite transmissions; and iv) transmitters beyond 260~km from the centre of the MRO.
 
-\defcitealias{lofar-radio-environment}{O2013a}
+%\defcitealias{lofar-radio-environment}{O2013a}
 The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA.
@@
 However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI.
-In \citeauthor{lofar-radio-environment} (\citeyear{lofar-radio-environment}; hereafter referred to as \citetalias{lofar-radio-environment}) it is shown that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is build in a populated area.
+In \citet{lofar-radio-environment} it is shown that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is built in a populated area.
 This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR.
-They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods and high time and frequency resolutions, residual RFI is weak and averages down similar to Gaussian noise.
+They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down similar to Gaussian noise.
 However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.
@@
 \section{Method} \label{ch:method}
-RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing.
-It is important to perform initial RFI detection at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citepalias{lofar-radio-environment}.
+RFI detection, often referred to as ``data flagging'', is one of the first steps in data processing.
+It is important to perform initial RFI detection at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}.
 Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays.
 In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate.
-For example, a $3\sigma$ threshold is used for reducing PAPER data in \citet{parsons-paper-eorlimit-2014}.
+For example, a $3\sigma$ threshold is used for analysing PAPER data in \citet{parsons-paper-eorlimit-2014}.
 Several observatories or projects have designed pipelines that include more advanced RFI mitigation.
@@
 For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used.
-It has been shown that this flagger has a good accuracy and is fast \citepalias{lofar-radio-environment}.
+It has been shown that this flagger has a good accuracy and is fast \citep{lofar-radio-environment}.
 It also has a library interface\footnote{The documentation for the \textsc{aoflagger} library interface can be found at \url{http://aoflagger.sourceforge.net/doc/api/}}, which allows it to be integrated in a pipeline.
@@
 
-To detect RFI in MWA observations, we have used the \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data.
+To detect RFI in MWA observations, we have used \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data.
 The \textsc{aoflagger} allows customising its flagging strategy, such as changing the threshold levels and expected smoothness of good data.
-Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions and varying fields of view.
+Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions, and varying fields of view.
 Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA specific strategy which is used in this work.
-The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citetalias{lofar-radio-environment}.
-In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency direction.
+The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citet{lofar-radio-environment}.
+In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency directions.
 Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}.
@@
 The bandpass corrections are not permanently applied to the data, but only applied during flagging.
-Normally, a per-channel calibration is performed after flagging for getting accurate flux calibration, which corrects the gain discontinuities permanently.
+Normally, a per-channel calibration is performed after flagging for getting accurate flux density calibration, which corrects the gain discontinuities permanently.
 Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but this step was still included in this work.
@@
 Auto-correlations are normally removed, because they are not used in imaging.
-In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume.
+In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency and/or time directions to reduce the data volume.
 It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}.
@@
 For more details about the correlator, see \citet{ord-2014-mwa-fpga-gpu}.
-Currently, the \textsc{cotter} preprocessing pipeline is ran by the scientist that calibrates and images the data.
+Currently, the \textsc{cotter} preprocessing pipeline is run by the scientist that calibrates and images the data.
 Once the scientist has downloaded the raw files for a given observation ID, there are various ways of processing MWA data.
-For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is ran in a special mode such that it only flags the data, and does not convert the raw correlator files.
+For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is run in a special mode such that it only flags the data, and does not convert the raw correlator files.
 After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS.
@@
 One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline.
-The SumThreshold and SIR operator algorithms that are used by the flagging strategy, use statistics over large time and frequency ranges.
+The SumThreshold and SIR operator algorithms that are used by the flagging strategy use statistics over large time and frequency ranges.
 Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible.
-However, a typical snapshot is about 50 GB in size, and without increasing reading and writing, it requires thus 50 GB of memory to perform flagging on the full data.
-When less memory is available, \textsc{cotter} will split the observation in different time chunks and flags those independently.
+However, a typical snapshot is about 50 GB in size and without increasing reading and writing it requires 50 GB of memory to perform flagging on the full data.
+When less memory is available, \textsc{cotter} will split the observation in different time segments and flags those independently.
 This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem.
 Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy.
-Because not all astronomers have easily access to high-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations.
-In this use-case, \textsc{cotter} is run in a flagging-only mode on the high-memory machine.
+Because not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations.
+In this use-case, \textsc{cotter} is run in a flagging-only mode on the large-memory machine.
 The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to convert the raw files and apply the RFI detection from the first run.
@@
 The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent.
-To solve this, the bordering 80~kHz on both sides of a 1.28~Mhz subband are normally flagged by \textsc{cotter}.
-This implies that in normal observations, 13\% of the data is lost due to the poly-phase filter.
+To solve this, the bordering 80~kHz on both sides of a 1.28~MHz subband are normally flagged by \textsc{cotter}.
+This implies that in normal observations, 13\% of the data are lost due to the poly-phase filter.
 However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels.
@@
 \begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
-\caption{RFI occupancy per subband, calculated over all observation nights except GLEAM 2014-03-17. The latter has been left out because it is affected by DTV.}
+\caption{RFI occupancy per subband, calculated over all observation nights except GLEAM 2014-03-17.
+The latter has been left out because it is affected by DTV.
+The horizontal gray line represents the false-positives rate of the RFI detection.}
 \label{fig:rfi-per-band}
@@
 The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations.
-This includes false detections, which are estimated in \citetalias{lofar-radio-environment} to account for approximately 0.4\% detections.
+This includes false detections, which are estimated in \citep{lofar-radio-environment} to account for approximately 0.4\% detections.
 Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI.
@@
 
-The residual noise levels after flagging can be used to validate whether the flagged data is free of RFI.
+The residual noise levels after flagging can be used to validate whether the flagged data are free of RFI.
 In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations.
@@
 Therefore, DTV interference has to be detected with another method.
-Additionally, some channels in the FM radio band show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and possibly because of RFI leakage.
+Additionally, some channels in the FM radio band show standard deviations that deviate slightly as a result of the smaller amount of available data after flagging and possibly because of RFI leakage.
 Nevertheless, because the effect is small these frequencies can be calibrated and imaged without a problem.
@@
 For the GLEAM, EoR high-band and EoR low-band observations of this work, the distributions are drawn in Fig.~\ref{fig:lognlogs}.
-The three distributions are different due to the fact that each observation covers different frequencies.
+The three distributions are different due to the fact that observations cover different frequencies.
 Separate distributions are shown for visibilities that have been detected as RFI (red lines) and those that have been classified as RFI free (green lines).
@@
 The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N=\beta S^{-4.47}$, with $\beta$ being a scaling factor.
-The GLEAM and EoR low-band observations do show a small power-law part, but the fitted exponent depends strongly on what part of the tail is selected.
+The GLEAM and EoR low-band observations do show a small power-law component but the fitted exponent depends strongly on what part of the tail is selected.
 The selected regions result in $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and EoR-low curves, respectively.
@@
 Strong pulses such as these are rare and well flagged, but a few weak broad-band pulses are observed in almost every 2-min snapshot.
-These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines is added together.
-We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time.
-The MWA Voltage Capture System (VCS; \citealt{mwa-voltage-capture-2014}), which allows high time resolution observations, might help to analyse these phenomena.
+These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power on all baselines is added together.
+We currently do not know what their origin is.
+The MWA Voltage Capture System (VCS; \citealt{mwa-voltage-capture-2014}), which allows high time resolution observations, might help to analyse these signals.
 
@@
 To time \textsc{cotter}, we use a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5-disc RAID setup.
-The wall-clock runtime for processing a single 2~min snapshot of 50~GB with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, is split up as follows: 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data.
+The wall-clock runtime for processing a single 2~min snapshot of 50~GB with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, is split up as follows: 3~min are spent on reading the data, 5~min on RFI detection, and 3.5~min on writing the data.
 Real-time processing can therefore be achieved by using 6 of such nodes in parallel.
@@
 \section{Comparison with LOFAR} \label{ch:lofar-comparison}
-The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies.
-The RFI occupancies for LOFAR from \citetalias{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency sub-range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}.
+The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows a comparison of RFI at the same frequencies between the telescopes.
+The RFI occupancies for LOFAR from \citep{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency sub-range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}.
 The statistics are regridded to the same frequency resolution of 48~kHz.
-The average RFI occupancy in this frequency range is 1.65\% for the MWA, while \citetalias{lofar-radio-environment} reports a 3.18\% occupancy for LOFAR.
+The average RFI occupancy in this frequency range is 1.65\% for the MWA, while \citep{lofar-radio-environment} reports a 3.18\% occupancy for LOFAR.
 
 It should be noted that the RFI detection for LOFAR was performed at 0.78~kHz while for the MWA it is performed at 40~kHz.
-As shown by \citetalias{lofar-radio-environment}, LOFAR observes many RFI events that are only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions.
-\citetalias{lofar-radio-environment} also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA.
-In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources: Cassiopeia~A is never visible at the MWA, and Cygnus~A is only visible at low elevations.
+As shown by \citep{lofar-radio-environment}, LOFAR observes many RFI events that are only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions.
+\citep{lofar-radio-environment} also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA.
+In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources; Cassiopeia~A is never visible at the MWA, and Cygnus~A is only visible at low elevations.
 An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations.
@@
 Also, DTV signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data.
-Over the full GLEAM range of 72--231~MHz, 1.1\% of the data is flagged by the RFI detection, and these RFI events are attributed to several different transmitters.
+Over the full GLEAM range of 72--231~MHz, 1.1\% of the data are flagged by the RFI detection, and these RFI events are attributed to several different transmitters.
 Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection.
@@
 
-SKA-low will be build at the same location as the MWA, and hence several lessons can be learned from the MWA.
+SKA-low will be built at the same location as the MWA, and hence several lessons can be learned from the MWA.
 First of all, it is clear that the SKA will need to be able to handle some amount of RFI.
-This includes the availability of computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI.
-Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA.
+This includes the availability of computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient bits, and a signal path with headroom in order not to be saturated by RFI.
+Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonable amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem.
 Thirdly, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains.

### 2014-11-04 cotter/metafitsfile.cpp
Warning message rewording

@@ -184,7 +184,7 @@ void MetaFitsFile::ReadTiles(std::vector<MWAInput>& inputs, std::vector<MWAAnten
 		if(input >= nrow)
 		{
 			std::ostringstream errstr;
-			errstr << "Antenna " << antennae[antenna].name << " has an input with index " << input << ", which is beyond the maximum index of " << (nrow-1);
+			errstr << "Tile " << tile << " has an input with index " << input << ", which is beyond the maximum index of " << (nrow-1);
 			throw std::runtime_error(errstr.str());
 		}
 		inputs[input].inputIndex = input;

### 2014-11-04 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Natashas comments

@@
 
-\title[The MWA radio environment]{The low-frequency environment of the Murchison Widefield Array: radio frequency interference analysis and mitigation}
+\title[The MWA radio environment]{The low-frequency environment of the Murchison Widefield Array: radio-frequency interference analysis and mitigation}
 
@@
 We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights.
-RFI detection removes 1.1\% of the data, and another $\sim3\%$ is lost because of occasional ionospheric or atmospheric propagation of RFI from digital TV (DTV).
+RFI detection removes 1.1\% of the data, and another 0.14\% is lost because of occasional ionospheric or atmospheric propagation of RFI from digital TV (DTV).
 After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands.
@@
 The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before analysis of data.
-Some advise is suggested to make the SKA robust to RFI, which include: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient bits; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
+Some advise is given to ensure the SKA is robust to RFI, which includes: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient bits; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
 \end{abstract}
@@
 \section{Introduction}
-In recent years, the issue of radio-frequency interference (RFI) in radio astronomy caused by man-made transmitters has become larger due to an increased number of transmitters and wider frequency bandwidths of radio observatories.
+Recent years have seen growth in the impact of radio-frequency interference (RFI) on radio astronomy, due to increased number of transmitters and wider frequency bandwidths of radio observatories.
 Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-born or satellite transmitters or accidental electromagnetic radiation, for example from cars or wind turbines.
@@
 In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}).
-The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 80--300~MHz with a 30.72-MHz instantaneous bandwidth; its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
+The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
 To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory in the Murchison region of Western Australia.
@@
 However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI.
-In \citet{lofar-radio-environment} it is shown that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is built in a populated area.
+In \citet{lofar-radio-environment} it is shown that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR exists in a populated area.
 This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR.
-They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down similar to Gaussian noise.
+They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise.
 However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.
 
-Recently, experiments to use the moon as a calibrator for EoR experiments have shown that the reflection of the moon of terrestrial transmitters complicates such an experiment \citep{mckinley-moon-2013}.
-However, reflections are spatially isolated to the centre of the observed disk caused by the moon, and the high resolution of LOFAR allows separating the reflected and intrinsic power from the moon \citep{vedantham-2014-todo}.
+Recently, experiments to use the Moon as a calibrator for EoR experiments have shown that the reflection of the Moon of terrestrial transmitters complicates such an experiment \citep{mckinley-moon-2013}.
+However, reflections are spatially isolated to the centre of the observed disk caused by the Moon, and the high resolution of LOFAR allows the separation of the reflected and intrinsic power from the Moon \citep{vedantham-2014-todo}.
 In another study it is described that objects in space of about a meter in size, such as satellites and space debris, also reflect terrestrial transmissions with enough strength to be observable by the MWA \citep{tingay-space-debris-2013}.
@@
 In Sect.~\ref{ch:data-description}, these data are described.
-Sect.~\ref{ch:detection-results} presents examples of RFI that were found in these data, as well as the effectivity of the RFI detection.
+Sect.~\ref{ch:detection-results} presents examples of RFI that were found in these data, as well as the efficacy of the RFI detection.
 In Sect.~\ref{ch:lofar-comparison}, the results are compared to a previously performed LOFAR RFI survey.
-Finally, in Sect.~\ref{ch:conclusions-and-discussion} conclusions are drawn and discussed.
+Finally, in Sect.~\ref{ch:conclusions-and-discussion}, conclusions are drawn and discussed.
 
@@
 RFI detection, often referred to as ``data flagging'', is one of the first steps in data processing.
-It is important to perform initial RFI detection at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}.
+It is important to perform initial RFI detection and excision at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}.
 Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays.
@@
 \begin{center}\hspace*{-0.8cm}\includegraphics[width=10.5cm]{img/bandpass}
-\caption{Correlator output RMS over frequency in an MWA high-frequency observation, calculated over all cross-correlated baselines and 112 seconds of data.
+\caption{Correlator output RMS with respect to frequency in an MWA high-frequency observation, calculated over all cross-correlated baselines and 112 seconds of data.
 In this band, the band-pass shows two large discontinuities over frequency, because the MWA receivers apply different digital gains at different frequencies to minimise quantisation noise.
-The 1.28~MHz pattern is due to the imperfect poly-phase filter.
+The 1.28-MHz pattern is due to the imperfect poly-phase filter.
 }
@@
 To detect RFI in MWA observations, we have used \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data.
-The \textsc{aoflagger} allows customising its flagging strategy, such as changing the threshold levels and expected smoothness of good data.
+The \textsc{aoflagger} allows the customisation of its flagging strategy, such as changing the threshold levels and expected smoothness of good data.
 Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions, and varying fields of view.
-Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA specific strategy which is used in this work.
+Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA-specific strategy which is used in this work.
 The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citet{lofar-radio-environment}.
@@
 The bandpass corrections are not permanently applied to the data, but only applied during flagging.
-Normally, a per-channel calibration is performed after flagging for getting accurate flux density calibration, which corrects the gain discontinuities permanently.
+Normally, a per-channel calibration is performed after flagging in order to obtain accurate flux density calibration, which corrects the gain discontinuities permanently.
 Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but this step was still included in this work.
@@
 Auto-correlations are normally removed, because they are not used in imaging.
-In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency and/or time directions to reduce the data volume.
+In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or \textsc{uv-fits} files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $u,v,w$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency and/or time directions to reduce the data volume.
 It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}.
@@
 For more traditional data processing with e.g.
-\textsc{casa} or \textsc{miriad}, the \textsc{cotter} output is set to either the \textsc{casa} or uvfits format.
+the Common Astronomy Software Applications (\textsc{casa}; \citealt{casa-2008}), \textsc{miriad} \citep{miriad-sault-1995} and/or \textsc{wsclean} \citep{offringa-wsclean-2014}, the \textsc{cotter} output is set to either the \textsc{casa} or \textsc{uv-fits} format.
 The output is then directly readable by the most common astronomical software packages.
@@
 Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy.
-Because not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations.
+As not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been set up that pre-runs \textsc{cotter} on all observations.
 In this use-case, \textsc{cotter} is run in a flagging-only mode on the large-memory machine.
@@
 In practice, most observations are recorded at a resolution of 0.5 s and 40 kHz, and averaged to a resolution of 2--4 s and 40--80 kHz to reduce the data volume.
-This does not significantly cause any time or bandwidth decorrelation up to the first beam null.
-\textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $uvw$ values for the central time and frequency of the output sample.
+This does not cause any significant time or bandwidth decorrelation up to the first beam null.
+\textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $u,v,w$ values for the central time and frequency of the output sample.
 This helps to prevent time/frequency decorrelation.
@@
 The poly-phase filter bank that performs the first separation into sub-bands, introduces a 1.28~MHz periodic spectral signature, which is shown in Fig.~\ref{fig:bandpass}.
-The poly-phase filter band-pass shape is hard to take out completely, because the filters are imperfect and each sub-band is affected by leakage from signals in its adjacent sub-bands.
+The poly-phase filter band-pass shape is hard to remove completely, because the filters are imperfect and each sub-band is affected by leakage from signals in its adjacent sub-bands.
 The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent.
@@
 \section{Detection results} \label{ch:detection-results}
-The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations.
+The ``RFI'' column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations.
 This includes false detections, which are estimated in \citep{lofar-radio-environment} to account for approximately 0.4\% detections.
@@
 With the exception of the ORBCOMM frequencies and DTV affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts.
-This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.
+This has been verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.
 
@@
 Left plot: dynamic spectrum for the worst affected snapshot; right plot: visibility RMS in a few of the DTV bands for the affected night after flagging.
-Because of its broadband nature, this kind of RFI is not well flagged in the initial \textsc{aoflagger} step, but it is detectable in the global statistics.
+Because of its broadband nature, this kind of RFI is not well-flagged in the initial \textsc{aoflagger} step, but it is detectable in the global statistics.
 }
@@
 As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can fully contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth.
-Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive to steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive for RFI that is this broad.
+Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive to steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive to RFI covering such a broad spectrum.
 
@@
 Therefore, when DTV interference occurs, it affects 35 of the 159~MHz bandwidth.
-The visibility ratios that are lost because of DTV RFI are detailed in the last column of Table~\ref{tbl:obs-list}.
+The visibility ratios that are lost because of DTV RFI are detailed in ``DTV'' column of Table~\ref{tbl:obs-list}.
 Because these frequencies overlap with the 21-cm HI frequencies redshifted to the EoR, in reality these frequencies are observed more than average, so the actual loss is somewhat higher.
@@
 \subsection{Distribution analysis}
-In \citet{offringa-rfi-distributions}, using LOFAR it is shown that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail with an exponent of approximately -1.5 to -1.6.
-For the GLEAM, EoR high-band and EoR low-band observations of this work, the distributions are drawn in Fig.~\ref{fig:lognlogs}.
+Using LOFAR, \citet{offringa-rfi-distributions} show that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail with an exponent of approximately -1.5 to -1.6.
+For the GLEAM, EoR high-band and EoR low-band observations of this work, the distributions are shown in Fig.~\ref{fig:lognlogs}.
 The three distributions are different due to the fact that observations cover different frequencies.
-Separate distributions are shown for visibilities that have been detected as RFI (red lines) and those that have been classified as RFI free (green lines).
+Separate distributions are shown for visibilities that have been detected as RFI (red lines) and those that have been classified as RFI-free (green lines).
 The EoR high-band curves do not include the night that was contaminated by DTV RFI.
@@
 The selected regions result in $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and EoR-low curves, respectively.
-Due to the small part over which the power laws hold, as well as the high possible error in the fits because a fitting region needs to be selected, it is hard to infer if a uniform distribution of transmitters is observed.
+Due to the small region over which the power laws hold, as well as the high possible error in the fits because a fitting region needs to be selected, it is hard to infer if a uniform distribution of transmitters is observed.
 
@@
 In the right figures, the result of RFI detection is shown with purple.
-The horizontal flagged lines are flagged because they are W-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
+The horizontal flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
 \label{fig:2m-amateur-band}
@@
 \end{center}
-\caption{DTV RFI briefly visible in an EoR high-band observation. The plots show the Stokes~I amplitudes for a single correlation. The right plot shows the result of RFI detection marked in purple.}
+\caption{DTV RFI briefly visible in an EoR high-band observation.
+The plots show the Stokes~I amplitudes for a single correlation.
+The right plot shows the result of RFI detection and invalid channels marked in purple.}
 \label{fig:dvb-burst}
@@
 The two panels show Stokes~I visibility amplitudes of a single correlation.
-In the right plot, the RFI detection flags are marked with purple.
+In the right plot, the RFI detection flags and invalid channels are marked with purple.
 Each event occupies 2 timesteps of 0.5~s.}
@@
 \caption{A weakly-observed transmitter centred on 156.66 MHz.
-Top-left window: a single correlation, where the transmitter is barely visible; Top-right window: detected RFI marked with purple.
+Top-left window: a single correlation, where the transmitter is barely visible; Top-right window: detected RFI and invalid channels with purple.
 When performing per-baseline flagging, the transmitter is partly detected; Bottom-left window: when combining data from all baselines, the transmitter becomes clearly visible; Bottom-right window: in yellow, result of executing \textsc{aoflagger} on the standard deviations calculated over all baselines.}\label{fig:156_7}
@@
 While a change in pointing can also change the appearance of the transmitter, the beam does not change within a single snapshot, while in Fig.~\ref{fig:2m-amateur-band} the strength and affected bandwidth of the transmitter does change during the snapshot.
-In the bottom panel of Fig.~\ref{fig:2m-amateur-band}, the same snapshot is showed but is zoomed in on a briefly-observed RFI event at 150.17 MHz.
+In the bottom panel of Fig.~\ref{fig:2m-amateur-band}, the same snapshot is shown but is zoomed in on a briefly-observed RFI event at 150.17 MHz.
 This event occupies only a single 40~kHz channel for 2 seconds, and thus is an event which requires flagging at approximately the observed time and frequency resolution or higher for accurate detection.
@@
 Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent.
-To time \textsc{cotter}, we use a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5-disc RAID setup.
+To time \textsc{cotter}, we use a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5-disc RAID5 setup.
 The wall-clock runtime for processing a single 2~min snapshot of 50~GB with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, is split up as follows: 3~min are spent on reading the data, 5~min on RFI detection, and 3.5~min on writing the data.
@@
 \citep{lofar-radio-environment} also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA.
-In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources; Cassiopeia~A is never visible at the MWA, and Cygnus~A is only visible at low elevations.
+In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources; Cassiopeia~A and Cygnus~A are only visible at low elevations.
 An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations.
@@
 Faint RFI events such as the ones in Figs.~\ref{fig:broadband-pulse} and \ref{fig:156_7}, or complex events such as the one in Fig.~\ref{fig:dvb-burst}, are not adequately detected by a single-sample thresholding algorithm, but \textsc{aoflagger}'s SumThreshold and SIR-operator algorithms are able to flag such events.
-These algorithms have gained some popularity (e.g., besides the use of \textsc{aoflagger} by individual astronomers, the SumThreshold algorithm is implemented in the pipeline for eMERLIN \citep{serpent-peck-2013} and in \textsc{miriad}), but other projects still use single-sample thresholding, e.g.
+These algorithms have gained some popularity; besides the use of \textsc{aoflagger} by individual astronomers, the SumThreshold algorithm is implemented in the pipeline for eMERLIN \citep{serpent-peck-2013} and in \textsc{miriad}).
+Nevertheless, other projects still use single-sample thresholding, e.g.
 PAPER \citep{parsons-paper-eorlimit-2014}.
-Since strong RFI is practically seen everywhere, and because the apparent strength of RFI events will follow a power-law distribution, many faint transmitters will interfere with observations of any (terrestrial) radio observatory.
+Since strong RFI is seen practically everywhere, and because the apparent strength of RFI events will follow a power-law distribution, many faint transmitters will interfere with observations of any (terrestrial) radio observatory.
 Using algorithms with low sensitivity will detect fewer of these, and thus result in RFI becoming more quickly a problem in long-integration projects.

### 2014-11-07 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
David Kaplan's comments and finishing the author lists

@@
 \def\CAASTRO{$^{2}$}
-\def\Curtin{$^{3}$}
+\def\ASTRON{$^{3}$}
+\def\Curtin{$^{4}$}
+\def\UWisc{$^{5}$}
+\def\UW{$^{6}$}
+\def\CASS{$^{7}$}
+\def\SKASA{$^{8}$}
+\def\ASU{$^{9}$}
+\def\USydney{$^{10}$}
+\def\Haystack{$^{11}$}
+\def\RRI{$^{12}$}
+\def\MIT{$^{13}$}
+\def\UWA{$^{14}$}
+\def\CfA{$^{15}$}
+\def\UMelbourne{$^{16}$}
+\def\Victoria{$^{17}$}
+\def\Tata{$^{18}$}
+%\def\UMichigan{$^{19}$}
+%\def\NRAO{$^{20}$}
 
-\author[A.~R.~Offringa et al.]{A.~R.~Offringa$^{1,2}$\thanks{\editmark{Corresponding author.} E-mail: \url{offringa@gmail.com}},
+\author[A.~R.~Offringa et al.]{A.~R.~Offringa\ANU$^,$\CAASTRO$^,$\ASTRON\thanks{Corresponding author. E-mail: \url{offringa@gmail.com}},
 R.~B.~Wayth\Curtin$^,$\CAASTRO,
 N.~Hurley-Walker\Curtin,
-D.~Kaplan...,
-... , %\newauthor
-builders list
+D.~L.~Kaplan\UWisc,
+N.~Barry\UW,
+A.~P.~Beardsley\UW,
+M.~E.~Bell\CASS,
+G.~Bernardi\SKASA,
+J.~D.~Bowman\ASU, 
+F.~Briggs\ANU,
+J.~Callingham\USydney,
+R.~J.~Cappallo\Haystack, 
+P.~Carroll\UW,
+%A.~de~Oliveira-Costa
+A.~A.~Deshpande\RRI, 
+J.~S.~Dillon\MIT,
+K.~S.~Dwarakanath\RRI,
+A.~Ewall-Wice\MIT,
+L.~Feng\MIT,
+B.-Q.~For\UWA,
+B.~M.~Gaensler\USydney$^,$\CAASTRO, 
+L.~J.~Greenhill\CfA,
+P.~Hancock\Curtin,
+K.~Han-Seek\UMelbourne,
+B.~J.~Hazelton\UW,
+J.~N.~Hewitt\MIT,
+L.~Hindson\Victoria,
+D.~C.~Jacobs\ASU, 
+M.~Johnston-Hollitt\Victoria,
+A.~Kapinska\UWA$^,$\CAASTRO,
+D.~L.~Kaplan\UWisc,
+P.~Kittiwisit\ASU,
+E.~Lenc\USydney$^,$\CAASTRO,
+J.~Line\UMelbourne,
+A.~Loeb\CfA,
+C.~J.~Lonsdale\Haystack, 
+B.~McKinley\ANU$^,$\CAASTRO,
+S.~R.~McWhirter\Haystack,
+D.~A.~Mitchell\CASS$^,$\CAASTRO, 
+M.~F.~Morales\UW, 
+E.~Morgan\MIT, 
+J.~Morgan\Curtin,
+A.~R.~Neben\MIT,
+D.~Oberoi\Tata, 
+S.~M.~Ord\Curtin$^,$\CAASTRO,
+B.~Pindor\UMelbourne,
+J.~C.~Pober\UW,
+T.~Prabu\RRI, 
+P.~Procopio\UMelbourne,
+J.~Riding\UMelbourne,  
+N.~Udaya~Shankar\RRI, 
+P.~Sourabh\RRI,
+K.~S.~Srivani\RRI,
+L.~Staveley-Smith\UWA,
+R.~Subrahmanyan\RRI$^,$\CAASTRO, 
+I.~S.~Sullivan\UW,
+M.~Tegmark\MIT,
+N.~Thyagarajan\ASU,
+S.~J.~Tingay\Curtin$^,$\CAASTRO, 
+C.~M.~Trott\Curtin$^,$\CAASTRO,
+R.~B.~Wayth\Curtin$^,$\CAASTRO, 
+R.~L.~Webster\UMelbourne$^,$\CAASTRO, 
+A.~Williams\Curtin, 
+C.~L.~Williams\MIT,
+C.~Wu\UWA,
+J.~S.~Wyithe\UMelbourne
 \\
 \ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
-\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO), Australian National University, Canberra, ACT 2611, Australia \\
+\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
+\ASTRON{}Netherlands Institute for Radio Astronomy (ASTRON), PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
 \Curtin{}International Centre for Radio Astronomy Research, Curtin University, Bentley, WA 6102, Australia\\
+\UWisc{}Department of Physics, University of Wisconsin--Milwaukee, Milwaukee, WI 53201, USA\\
+\UW{}Department of Physics, University of Washington, Seattle, WA 98195, USA\\
+\CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
+\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town, 7405, South Africa\\
+\ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
+\USydney{}Sydney Institute for Astronomy, School of Physics, University of Sydney, NSW 2006, Australia\\
+\Haystack{}MIT Haystack Observatory, Westford, MA 01886, USA\\
+\RRI{}Raman Research Institute, Bangalore 560080, India\\
+\MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
+\UWA{}University of Western Australia, Crawley, WA 6009, Perth, Australia\\
+\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
+\UMelbourne{}School of Physics, University of Melbourne, Parkville, VIC 3010, Australia\\
+\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
+\Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India
+%\UMichigan{}University of Michigan, Ann Harbor, MI 48109, USA
 }
@@
 \begin{abstract}
-The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at the location of the future Square Kilometre Array (SKA).
+The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at one of the locations of the future Square Kilometre Array (SKA).
 We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights.
-RFI detection removes 1.1\% of the data, and another 0.14\% is lost because of occasional ionospheric or atmospheric propagation of RFI from digital TV (DTV).
+RFI detection removes 1.1\% of the data.
+RFI from digital TV (DTV) is observed 3\% of the time due to occasional ionospheric or atmospheric propagation.
 After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands.
@@
 The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before analysis of data.
-Some advise is given to ensure the SKA is robust to RFI, which includes: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient bits; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
+We include specific recommendations designed to make the SKA more robust to RFI, including: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient dynamic range; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
 \end{abstract}
@@
 \section{Introduction}
-Recent years have seen growth in the impact of radio-frequency interference (RFI) on radio astronomy, due to increased number of transmitters and wider frequency bandwidths of radio observatories.
+Recent years have seen growth in the impact of radio-frequency interference (RFI) on radio astronomy, due to increased number of transmitters and wider bandwidths of radio observatories.
 Spectrum allocation management and radio-quiet zones help to limit the interference, but do not include all terrestrial transmissions, nor solve interference from air-born or satellite transmitters or accidental electromagnetic radiation, for example from cars or wind turbines.
@@
 Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}.
-The consequence of this approach is that a certain ratio of data is lost due to interference, and that frequencies that are continuously occupied by transmitters cannot be observed.
+The consequence of this approach is that a certain fraction of data is lost due to interference, and that frequencies that are continuously occupied by transmitters cannot be observed.
 It is important to analyse the impact of RFI on a specific instrument to optimise the RFI mitigation approach.
@@
 The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
-To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory in the Murchison region of Western Australia.
+To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory (MRO) in the Murchison region of Western Australia.
 Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.
@@
 Comparing the MWA to these arrays allows the comparison of the observatory site and hardware design with respect to the RFI impact.
-So far, initial observations with the MWA have produced the first scientific results without RFI issues (e.g.
+So far, initial observations with the MWA have produced scientific results without RFI issues (e.g.
 \citealt{mwacs-2014}), which of course is not surprising given its remote location.
@@
 \section{Method} \label{ch:method}
-RFI detection, often referred to as ``data flagging'', is one of the first steps in data processing.
+RFI detection, often referred to as ``data flagging'', is one of the first steps in processing the data from any interferometer.
 It is important to perform initial RFI detection and excision at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}.
@@
 To detect RFI in MWA observations, we have used \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data.
-The \textsc{aoflagger} allows the customisation of its flagging strategy, such as changing the threshold levels and expected smoothness of good data.
-Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions, and varying fields of view.
-Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA-specific strategy which is used in this work.
-The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citet{lofar-radio-environment}.
-In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency directions.
+\textsc{aoflagger} is a general-purpose RFI flagging tool developed originally for LOFAR \citep{lofar-radio-environment}.
+Specific customisations, such as changing the threshold levels and expected smoothness of good data, can be made for different telescopes to optimize the detection for different band-pass shapes, time and frequency resolutions, and fields of view.
+Strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA-specific strategy which is used in this work.
+In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency axes.
 Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}.
@@
 
-The MWA and LOFAR strategies differ on one aspect.
+The MWA and LOFAR strategies differ qualitatively in one aspect.
 For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation.
@@
 Auto-correlations are normally removed, because they are not used in imaging.
-In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or \textsc{uv-fits} files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $u,v,w$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency and/or time directions to reduce the data volume.
+In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or \textsc{uv-fits} files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $u,v,w$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency and/or time to reduce the data volume.
 It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}.
-Various tools are available to analyse these statistics, e.g.
+Tools are available to analyse these statistics, e.g.
 the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.
@@
 \section{Detection results} \label{ch:detection-results}
-The ``RFI'' column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations.
-This includes false detections, which are estimated in \citep{lofar-radio-environment} to account for approximately 0.4\% detections.
+The ``RFI'' column of Table~\ref{tbl:obs-list} lists for each of the observation nights the fraction of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations.
+This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% of detections.
 Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI.
@@
 The total RFI occupancy in all observations is 0.96\%.
-Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global expected RFI ratio in the 72.3--230.8 MHz range.
+Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global RFI fraction in the 72.3--230.8 MHz range.
 
@@
 \caption{Visibility amplitude distributions with logarithmic axes.
-The dashed line represent fits over a reasonably constant part of the tail (selected by eye) of the RFI distribution: $N=\beta S^{-1.43}$ for GLEAM, $N=\beta S^{-4.47}$ for EoR high and $N=\beta S^{-1.33}$ for EoR low.}
+The dashed line represent fits to the function $N=\beta S^\alpha$ over a reasonably constant part of the tail (selected by eye) of the RFI distribution, resulting in $\alpha=-1.43$ for GLEAM, $\alpha=-4.47$ for EoR high, and $\alpha=-1.33$ for EoR low.}
 \label{fig:lognlogs}
@@
 \subsection{Distribution analysis}
-Using LOFAR, \citet{offringa-rfi-distributions} show that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail with an exponent of approximately -1.5 to -1.6.
+Using LOFAR, \citet{offringa-rfi-distributions} show that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail described by $N\propto S^\alpha$, where $N$ is the rate of visibilities, $S$ is the visibility brightness and the exponent $\alpha$ is typically -1.5 to -1.6.
 For the GLEAM, EoR high-band and EoR low-band observations of this work, the distributions are shown in Fig.~\ref{fig:lognlogs}.
@@
 Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}.
-The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N=\beta S^{-4.47}$, with $\beta$ being a scaling factor.
+The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N\propto S^{-4.47}$.
 The GLEAM and EoR low-band observations do show a small power-law component but the fitted exponent depends strongly on what part of the tail is selected.
-The selected regions result in $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and EoR-low curves, respectively.
+The selected regions result in $N\propto S^{-1.43}$ and $N\propto S^{-1.33}$ for the GLEAM and EoR-low curves, respectively.
 Due to the small region over which the power laws hold, as well as the high possible error in the fits because a fitting region needs to be selected, it is hard to infer if a uniform distribution of transmitters is observed.
@@
 The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows a comparison of RFI at the same frequencies between the telescopes.
-The RFI occupancies for LOFAR from \citep{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency sub-range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}.
+The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency sub-range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}.
 The statistics are regridded to the same frequency resolution of 48~kHz.
-The average RFI occupancy in this frequency range is 1.65\% for the MWA, while \citep{lofar-radio-environment} reports a 3.18\% occupancy for LOFAR.
+The average RFI occupancy in this frequency range is 1.65\% for the MWA, while \citet{lofar-radio-environment} reports a 3.18\% occupancy for LOFAR.
 
 It should be noted that the RFI detection for LOFAR was performed at 0.78~kHz while for the MWA it is performed at 40~kHz.
-As shown by \citep{lofar-radio-environment}, LOFAR observes many RFI events that are only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions.
-\citep{lofar-radio-environment} also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA.
+As shown by \citet{lofar-radio-environment}, LOFAR observes many RFI events that are only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions.
+\citet{lofar-radio-environment} also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA.
 In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources; Cassiopeia~A and Cygnus~A are only visible at low elevations.
@@
 First of all, it is clear that the SKA will need to be able to handle some amount of RFI.
-This includes the availability of computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient bits, and a signal path with headroom in order not to be saturated by RFI.
+This includes the availability of computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient dynamic range, and a signal path with headroom in order not to be saturated by RFI.
 Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonable amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem.
@@
 Faint RFI events such as the ones in Figs.~\ref{fig:broadband-pulse} and \ref{fig:156_7}, or complex events such as the one in Fig.~\ref{fig:dvb-burst}, are not adequately detected by a single-sample thresholding algorithm, but \textsc{aoflagger}'s SumThreshold and SIR-operator algorithms are able to flag such events.
-These algorithms have gained some popularity; besides the use of \textsc{aoflagger} by individual astronomers, the SumThreshold algorithm is implemented in the pipeline for eMERLIN \citep{serpent-peck-2013} and in \textsc{miriad}).
+These algorithms have gained some popularity; besides the use of \textsc{aoflagger} by individual astronomers, the SumThreshold algorithm is implemented in the pipeline for eMERLIN \citep{serpent-peck-2013} and in \textsc{miriad}.
 Nevertheless, other projects still use single-sample thresholding, e.g.

### 2014-11-07 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Correcting author list: removing double Kaplan and Wayth, and correcting Sourabh Paul's name

@@
 A.~Kapinska\UWA$^,$\CAASTRO,
-D.~L.~Kaplan\UWisc,
 P.~Kittiwisit\ASU,
@@
 S.~M.~Ord\Curtin$^,$\CAASTRO,
+S.~Paul\RRI,
 B.~Pindor\UMelbourne,
@@
 N.~Udaya~Shankar\RRI, 
-P.~Sourabh\RRI,
 K.~S.~Srivani\RRI,
@@
 C.~M.~Trott\Curtin$^,$\CAASTRO,
-R.~B.~Wayth\Curtin$^,$\CAASTRO, 
 R.~L.~Webster\UMelbourne$^,$\CAASTRO, 

### 2014-11-07 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Adding Zheng to author list, fixing spaces in table

@@
 C.~Wu\UWA,
-J.~S.~Wyithe\UMelbourne
+J.~S.~Wyithe\UMelbourne,
+Q. Zheng\Victoria
 \\
@@
 \hline
-EoR high& 2013-08-23 & 167.0--197.7              & 3h & 0.53\% & 0\% \\
-GLEAM & 2013-08-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.94\% & 0\% \\
-EoR low& 2013-08-26 & 138.9--169.6              & 6h & 0.81\% & ---\\
-GLEAM & 2013-11-05 & 72.3--133.8, 138.9--230.8 & 7h & 1.27\% & 0\% \\
-GLEAM & 2013-11-25 & 72.3--133.8, 138.9--230.8 & 7h & 0.69\% & 0\% \\
-EoR high& 2014-02-05 & 167.0--197.7              & 6h & 0.54\% & 0\%\\
-GLEAM & 2014-03-16 & 72.3--133.8, 138.9--230.8 & 7h & 0.79\% & 0\% \\
-GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7h & 1.64\% & 1.29\% \\
-EoR high& 2014-04-10 & 167.0--197.7              & 6h & 0.68\% & 0\%\\
-GLEAM & 2014-06-18 & 72.3--133.8, 138.9--230.8 & 7h & 0.98\% & 0\% \\
-Orbcomm RFI test&2014-08-27& 131.2--161.9     & 2$\times$8 m& 1.85\% & --- \\
+EoR high& 2013-08-23 & 167.0--197.7              & 3~h & 0.53\% & 0\% \\
+GLEAM & 2013-08-25 & 72.3--133.8, 138.9--230.8 & 7~h & 0.94\% & 0\% \\
+EoR low& 2013-08-26 & 138.9--169.6              & 6~h & 0.81\% & ---\\
+GLEAM & 2013-11-05 & 72.3--133.8, 138.9--230.8 & 7~h & 1.27\% & 0\% \\
+GLEAM & 2013-11-25 & 72.3--133.8, 138.9--230.8 & 7~h & 0.69\% & 0\% \\
+EoR high& 2014-02-05 & 167.0--197.7              & 6~h & 0.54\% & 0\%\\
+GLEAM & 2014-03-16 & 72.3--133.8, 138.9--230.8 & 7~h & 0.79\% & 0\% \\
+GLEAM & 2014-03-17 & 72.3--133.8, 138.9--230.8 & 7~h & 1.64\% & 1.29\% \\
+EoR high& 2014-04-10 & 167.0--197.7              & 6~h & 0.68\% & 0\%\\
+GLEAM & 2014-06-18 & 72.3--133.8, 138.9--230.8 & 7~h & 0.98\% & 0\% \\
+Orbcomm RFI test&2014-08-27& 131.2--161.9     & 2$\times$8~m& 1.85\% & --- \\
 \hline

### 2014-11-09 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Adding S. Sethi (RRI) to author list -- had apparently overlooked him

@@
 N.~Udaya~Shankar\RRI, 
+S.~Sethi\RRI,
 K.~S.~Srivani\RRI,

### 2014-11-10 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Adding Melanies comments, except her last point

@@
 F.~Briggs\ANU,
-J.~Callingham\USydney,
+J.~R.~Callingham\USydney$^,$\CAASTRO,
 R.~J.~Cappallo\Haystack, 
@@
 The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at one of the locations of the future Square Kilometre Array (SKA).
-We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights.
+We describe the automated RFI detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights.
 RFI detection removes 1.1\% of the data.
@@
 In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}).
-The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
+The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; one of its main science drivers is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
 To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory (MRO) in the Murchison region of Western Australia.
-Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.
+Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.
 
@@
 %\defcitealias{lofar-radio-environment}{O2013a}
-The interferometric arrays LOFAR \citep{lofar-2013}, PAPER \citep{parsons-paper-eorlimit-2014} and the GMRT \citep{the-gmrt-swarup-chapter-2013} observe in approximately the same frequency range as the MWA.
+The interferometric arrays Low-Frequency Array (LOFAR; \citealt{lofar-2013}), Precision Array to Probe the Epoch of Reionization (PAPER; \citealt{parsons-paper-eorlimit-2014}) and the Giant Metrewave Radio Telescope (GMRT; \citealt{the-gmrt-swarup-chapter-2013}) observe in approximately the same frequency range as the MWA.
 For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization.
-Comparing the MWA to these arrays allows the comparison of the observatory site and hardware design with respect to the RFI impact.
-So far, initial observations with the MWA have produced scientific results without RFI issues (e.g.
-\citealt{mwacs-2014}), which of course is not surprising given its remote location.
+The overlap in frequency gives an opportunity to compare the observatory sites and hardware designs with respect to the RFI impact.
+So far, initial observations with the MWA have produced scientific results without RFI issues \citep{mwacs-2014, mckinley-fornaxa-2014, hindson-cluster-emission-2014}, which of course is not surprising given its remote location.
 PAPER has shown that their radio environment in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength.
@@
 Because the MWA has a 30.72~MHz instantaneous bandwidth, this range has been split up so that each night observes with a central frequency of either 154.2 or 182.4~MHz, with an overlap of 2.6 MHz.
+These will be referred to as the EoR low and high bands, respectively.
 In these observations, the selected EoR field is tracked by the primary beam, such that the sensitivity towards the field is maximal.
@@
 \begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
-\caption{Observed noise in the high-resolution channels for each observation.
-This is the residual noise in the visibilities after RFI detection.
-A few channels in the FM bands show a deviating standard deviation, and observation `GLEAM 2014-03-17' shows DTV contamination around 180~MHz.}
+\caption{Observed visibility RMS in the high-resolution (40~kHz) channels for each observation.
+This is the residual noise in the visibilities after RFI detection and flagging.
+A few channels in the FM bands show a deviating standard deviation, and observation `GLEAM 2014-03-17' shows DTV contamination around 180~MHz.
+The variability over nights is caused by the different celestial observing times and pointing directions, and therefore different apparent brightness of e.g.
+the Galaxy.}
 \label{fig:stddev-per-set}
@@
 With the exception of the ORBCOMM frequencies and DTV affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts.
-This has been verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.
+This has been verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{hindson-cluster-emission-2014, mwacs-2014, murphy-mwa-exoplanets-2014}.
 

### 2014-11-11 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Paul Hancock's comments

@@
 \ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
-\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
+%\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
+\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO) \\
 \ASTRON{}Netherlands Institute for Radio Astronomy (ASTRON), PO Box 2, 7990 AA Dwingeloo, The Netherlands\\
@@
 The results are compared to a previously published LOFAR RFI survey.
-The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before analysis of data.
+The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before data can be analysed.
 We include specific recommendations designed to make the SKA more robust to RFI, including: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient dynamic range; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
@@
 Because of this, when no averaging is requested, the output is 50\% larger than the input (one extra float per complex float visibility).
-In practice, most observations are recorded at a resolution of 0.5 s and 40 kHz, and averaged to a resolution of 2--4 s and 40--80 kHz to reduce the data volume.
-This does not cause any significant time or bandwidth decorrelation up to the first beam null.
+In practice, most MWA observations are recorded at a resolution of 0.5~s and 40~kHz, and averaged to a resolution of 2--4 s and 40--80 kHz to reduce the data volume.
+This does not cause any significant time or bandwidth decorrelation up to the first beam null for MWA data.
 \textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $u,v,w$ values for the central time and frequency of the output sample.
@@
 \caption{Observed visibility RMS in the high-resolution (40~kHz) channels for each observation.
-This is the residual noise in the visibilities after RFI detection and flagging.
+This is the residual noise in the visibilities after RFI excision.
 A few channels in the FM bands show a deviating standard deviation, and observation `GLEAM 2014-03-17' shows DTV contamination around 180~MHz.
-The variability over nights is caused by the different celestial observing times and pointing directions, and therefore different apparent brightness of e.g.
+The variability over nights is caused by the different celestial observing times and pointing directions, and therefore the difference in apparent brightness of e.g.
 the Galaxy.}
@@
 From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar per-sample computational requirements as for LOFAR.
-Both RFI occupancy levels are so small that the RFI contamination in the 115--163~MHz range should not be a consideration in selecting a site.
-However, a benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DTV bands.
+When comparing the 115--163~MHz range, the RFI occupancy levels are small and the RFI flagging is effective for both sites.
+Therefore, the MWA's larger radio-quiet zone does not provide a significant benefit at these frequencies.
+However, a benefit of the remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DTV bands.
 Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}.

### 2014-11-17 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Implemented Emil's comments

@@
 To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory (MRO) in the Murchison region of Western Australia.
-Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.
+Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture array are planned to be built near the location of the MWA.
 
 Several levels of protection are in place to protect the radio quietness of the MRO.
-Within a 70~km radius, the Australian Communications and Media Authority\footnote{\url{http://www.acma.gov.au/}} and WA government provide the strongest level of protection from other radio equipment across the frequency range 70~MHz to 25.25~GHz.
+Within a 70~km radius, the Australian Communications and Media Authority\footnote{\url{http://www.acma.gov.au/}} and Western Australia government provide the strongest level of protection from other radio equipment across the frequency range 70~MHz to 25.25~GHz.
 Radio devices in this zone must not cause interference to radio astronomy.
@@
 Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy.
-As not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been set up that pre-runs \textsc{cotter} on all observations.
+As not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been set up that runs a partial \textsc{cotter} preprocessing on all observations.
 In this use-case, \textsc{cotter} is run in a flagging-only mode on the large-memory machine.
@@
 Fig.~\ref{fig:rfi-per-band} shows the overall detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17.
-The latter is the only night affected by interference from digital TV (DTV), and will be analysed later.
+The latter is the only night affected by interference from digital TV (DTV), and will be analysed later in this section.
 The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data.
@@
 Faint RFI events such as the ones in Figs.~\ref{fig:broadband-pulse} and \ref{fig:156_7}, or complex events such as the one in Fig.~\ref{fig:dvb-burst}, are not adequately detected by a single-sample thresholding algorithm, but \textsc{aoflagger}'s SumThreshold and SIR-operator algorithms are able to flag such events.
-These algorithms have gained some popularity; besides the use of \textsc{aoflagger} by individual astronomers, the SumThreshold algorithm is implemented in the pipeline for eMERLIN \citep{serpent-peck-2013} and in \textsc{miriad}.
+These algorithms have gained some popularity; besides the use of \textsc{aoflagger} by individual astronomers, the SumThreshold algorithm is implemented in the pipeline for eMERLIN \citep{serpent-peck-2013} and in \textsc{miriad}'s PGFLAG task.
 Nevertheless, other projects still use single-sample thresholding, e.g.

### 2014-11-17 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Changing text of Miriad, as it uses the SIR-operator

@@
 Faint RFI events such as the ones in Figs.~\ref{fig:broadband-pulse} and \ref{fig:156_7}, or complex events such as the one in Fig.~\ref{fig:dvb-burst}, are not adequately detected by a single-sample thresholding algorithm, but \textsc{aoflagger}'s SumThreshold and SIR-operator algorithms are able to flag such events.
-These algorithms have gained some popularity; besides the use of \textsc{aoflagger} by individual astronomers, the SumThreshold algorithm is implemented in the pipeline for eMERLIN \citep{serpent-peck-2013} and in \textsc{miriad}'s PGFLAG task.
+These algorithms have gained some popularity; besides the use of \textsc{aoflagger} by individual astronomers, \textsc{miriad}'s PGFLAG task implements both the SumThreshold and SIR-operator algorithms, and the pipeline for eMERLIN \citep{serpent-peck-2013} implements the SumThreshold algorithm.
 Nevertheless, other projects still use single-sample thresholding, e.g.

### 2014-11-17 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
John's larger comments

@@
 In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations.
-Observation `GLEAM 2014-03-17' shows residual DTV interference, both in the frequency range 174--195~MHz (RF 6, 7 and 8) and 216--230~MHz (RF 11 and 12), and it is clear that this RFI has not been adequately flagged.
+Observation `GLEAM 2014-03-17' shows residual DTV interference, both in the frequency range 174--195~MHz (radio frequencies (RF) 6, 7 and 8) and 216--230~MHz (RF 11 and 12), and it is clear that this RFI has not been adequately flagged.
 Therefore, DTV interference has to be detected with another method.
@@
 Nevertheless, because the effect is small these frequencies can be calibrated and imaged without a problem.
+FM-band RFI is noticeably worse when pointing at the southern horizon, however beyond this we do not have sufficient data to explore any correlation between pointing direction and RFI.
 The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI.
@@
 Although the flagger does not adequately flag DTV interference, the right plot of Fig.~\ref{fig:dvb} shows that from the data statistics it is possible to determine whether a snapshot is affected by DTV interference.
-An option is to test whether the visibility RMS of a snapshot in any of the occupied DTV bands exceeds the RMS in DTV band 4, which is not used for broadcasting.
+An option is to test whether the visibility RMS of a snapshot in any of the occupied DTV bands exceeds the RMS in the Digital Video Broadcasting (DVB) band 4 (167--174~MHz), which is not used for broadcasting.
 These statistics are calculated by \textsc{cotter}, hence this can be validated without having to read the data again.
@@
 
-After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging.
+After RFI detection and excision, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging.
 From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar per-sample computational requirements as for LOFAR.
@@
 Thirdly, flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains.
-For accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth.
+The extra per-subband gain correction that is required for the MWA makes the detection less stable, and the sub-band bandpass makes it harder to recognize RFI patterns in frequency direction.
+Therefore, for accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth.
 Finally, the presence of short and spectrally-narrow RFI events confirms that detection at high time and frequency resolution improves the accuracy.

### 2014-11-17 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
John's small comments

@@
 RFI from digital TV (DTV) is observed 3\% of the time due to occasional ionospheric or atmospheric propagation.
-After RFI detection, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands.
+After RFI detection and excision, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands.
 The results are compared to a previously published LOFAR RFI survey.
-The remote location of the MWA has resulted in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before data can be analysed.
-We include specific recommendations designed to make the SKA more robust to RFI, including: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient dynamic range; a smooth band-pass response; and the possibility for RFI detection at high time and frequency resolution (second and kHz-scale respectively).
+The remote location of the MWA results in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before data can be analysed.
+We include specific recommendations designed to make the SKA more robust to RFI, including: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient dynamic range; a smooth band-pass response; and the capability of RFI detection at high time and frequency resolution (second and kHz-scale respectively).
 \end{abstract}
@@
 Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}.
-The consequence of this approach is that a certain fraction of data is lost due to interference, and that frequencies that are continuously occupied by transmitters cannot be observed.
+The consequences of this approach are that a certain fraction of data is lost due to interference, and frequencies that are continuously occupied by transmitters cannot be observed.
 It is important to analyse the impact of RFI on a specific instrument to optimise the RFI mitigation approach.
@@
 To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory (MRO) in the Murchison region of Western Australia.
-Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture array are planned to be built near the location of the MWA.
+Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture array are planned to be built in this vicinity.
 
@@
 However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI.
-In \citet{lofar-radio-environment} it is shown that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR exists in a populated area.
+\citet{lofar-radio-environment} show that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is located in a populated area.
 This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR.
@@
 Recently, experiments to use the Moon as a calibrator for EoR experiments have shown that the reflection of the Moon of terrestrial transmitters complicates such an experiment \citep{mckinley-moon-2013}.
-However, reflections are spatially isolated to the centre of the observed disk caused by the Moon, and the high resolution of LOFAR allows the separation of the reflected and intrinsic power from the Moon \citep{vedantham-2014-todo}.
-In another study it is described that objects in space of about a meter in size, such as satellites and space debris, also reflect terrestrial transmissions with enough strength to be observable by the MWA \citep{tingay-space-debris-2013}.
+However, reflections are spatially restricted to the centre of the Lunar disk, and the high resolution of LOFAR allows the separation of the reflected and intrinsic power from the Moon \citep{vedantham-2014-todo}.
+Another study shows that objects in space of about a meter in size, such as satellites and space debris, may also reflect terrestrial transmissions with enough strength to be observable by the MWA \citep{tingay-space-debris-2013}.
 While tracking space debris is a useful asset, such reflected RFI can be a problem for EoR experiments, especially all-sky experiments that try to measure the global EoR signal \citep{vedantham-2014-todo}.
@@
 Normally, a per-channel calibration is performed after flagging in order to obtain accurate flux density calibration, which corrects the gain discontinuities permanently.
-Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but this step was still included in this work.
+Recently, the applied digital gains have been smoothed to prevent these discontinuities, which makes it possible to skip the band-pass correction step, but it was necessary to apply this correction for the data used here.
 
@@
 In these observations, the selected EoR field is tracked by the primary beam, such that the sensitivity towards the field is maximal.
-The GLEAM survey covers 72.3--230.8 MHz split into 5 bands of 30.72 MHz.
-The GLEAM observations are made in drift-scan mode, and the 5 bands are observed one by one in 5 subsequent snapshots.
+The GLEAM survey covers 72.3--230.8~MHz split into 5 bands of 30.72~MHz.
+The GLEAM observations are made in drift-scan mode, with the 5 bands rotated through in sequence on a 2-minute cadence.
 
@@
 This is the residual noise in the visibilities after RFI excision.
-A few channels in the FM bands show a deviating standard deviation, and observation `GLEAM 2014-03-17' shows DTV contamination around 180~MHz.
+A few channels in the FM bands show an abnormal standard deviation, and observation `GLEAM 2014-03-17' shows DTV contamination around 180~MHz.
 The variability over nights is caused by the different celestial observing times and pointing directions, and therefore the difference in apparent brightness of e.g.
@@
 \section{Detection results} \label{ch:detection-results}
-The ``RFI'' column of Table~\ref{tbl:obs-list} lists for each of the observation nights the fraction of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations.
+The ``RFI'' column of Table~\ref{tbl:obs-list} lists the fraction of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations for each observing night.
 This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% of detections.
@@
 When DTV is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used.
-However, it might not be worth it to have special mechanisms in place, such as having to account for lower SNR and a decreased $uv$ coverage, to salvage the residual good data within an affected snapshot.
+This could be dealt with by accounting for lower SNR and a decreased $uv$ coverage, however putting such a mechanism in place (rather than just deleting affected snapshots entirely) may not be worth the effort.
 Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability that DTV interference occurs is roughly 3\%.
@@
 Using LOFAR, \citet{offringa-rfi-distributions} show that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail described by $N\propto S^\alpha$, where $N$ is the rate of visibilities, $S$ is the visibility brightness and the exponent $\alpha$ is typically -1.5 to -1.6.
-For the GLEAM, EoR high-band and EoR low-band observations of this work, the distributions are shown in Fig.~\ref{fig:lognlogs}.
+For the GLEAM, EoR high-band and EoR low-band observations studied here, the distributions are shown in Fig.~\ref{fig:lognlogs}.
 The three distributions are different due to the fact that observations cover different frequencies.
@@
 The selected regions result in $N\propto S^{-1.43}$ and $N\propto S^{-1.33}$ for the GLEAM and EoR-low curves, respectively.
-Due to the small region over which the power laws hold, as well as the high possible error in the fits because a fitting region needs to be selected, it is hard to infer if a uniform distribution of transmitters is observed.
+Due to the small region over which the power laws hold, as well as the possible high error in the fits because of the subjective data selection, it is hard to infer if a uniform distribution of transmitters is observed.
 
-In the ideal case, the residual visibility distributions show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}).
+In the ideal case, the residual visibility distribution would show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}).
 The residual curves in Fig.~\ref{fig:lognlogs} do not go down as quickly as a Rayleigh curve would, and therefore the distributions have a slight excess of samples with higher amplitudes.
@@
 \subsection{RFI types}
-A variety of RFI events types are observed in the data sets.
+A variety of RFI events are observed in the data sets.
 While there are too many transmitters to show examples for each, it is helpful to understand what kind of events are visible and how they are flagged by the \textsc{aoflagger}.
@@
 We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights.
-After RFI detection, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz.
-Also, DTV signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data.
-Over the full GLEAM range of 72--231~MHz, 1.1\% of the data are flagged by the RFI detection, and these RFI events are attributed to several different transmitters.
+After RFI detection and excision, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz.
+Also, DTV signals are seen $\sim3\%$ of the time, and when present make observing in the 174--195~MHz range impossible.
+Over the full GLEAM range of 72--231~MHz, 1.1\% of the data are detected and flagged by the \textsc{aoflagger}, and these RFI events are attributed to several different transmitters.
 Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection.
@@
 
-SKA-low will be built at the same location as the MWA, and hence several lessons can be learned from the MWA.
+SKA-low will be built at the same location as the MWA, and hence several lessons can be learned.
 First of all, it is clear that the SKA will need to be able to handle some amount of RFI.
-This includes the availability of computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient dynamic range, and a signal path with headroom in order not to be saturated by RFI.
+This requires computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient dynamic range, and a signal path with headroom sufficient to avoid saturation by RFI.
 Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonable amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem.
@@
 Therefore, for accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth.
-Finally, the presence of short and spectrally-narrow RFI events confirms that detection at high time and frequency resolution improves the accuracy.
+Finally, the presence of short and spectrally-narrow RFI events confirms that detection at high time and frequency resolution improves accuracy.
 
@@
 
-The strategy does not perform well on continuous broadband RFI such as DTV, which is occasionally present due to tropospheric ducting or ionospheric activity.
+The current sensitivity of \textsc{aoflagger} is enough for calibration and imaging of MWA data.
+However, \textsc{aoflagger} does not perform well on continuous broadband RFI such as DTV, which is occasionally present due to tropospheric ducting or ionospheric activity.
 To remove DTV, a second detection round is required, and the current methodology to handle DTV RFI is to delete all affected snapshots, or even the entire night.
 This requires hardly any computational power, because the visibility statistics are collected in \textsc{cotter}.
-The current sensitivity of the flagger is enough for calibration and imaging of MWA data.
 For deep-integration projects, such as the EoR projects, it might be that low-level RFI will show up at lower noise levels.

### 2014-11-17 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Slight wording changes

@@
 In this use-case, \textsc{cotter} is run in a flagging-only mode on the large-memory machine.
-The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to convert the raw files and apply the RFI detection from the first run.
+The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to apply the RFI detection from the first run and convert the raw files to his/her preferred output.
 
@@
 This does not cause any significant time or bandwidth decorrelation up to the first beam null for MWA data.
-\textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $u,v,w$ values for the central time and frequency of the output sample.
+\textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $u,v,w$-values for the central time and frequency of the output sample.
 This helps to prevent time/frequency decorrelation.
@@
 The total RFI occupancy in all observations is 0.96\%.
-Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global RFI fraction in the 72.3--230.8 MHz range.
+Weighting the occupancy in each channel by the inverse time that it is observed results in a global RFI occupancy of 1.13\% in the 72.3--230.8 MHz range.
 

### 2014-11-18 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Abraham's comments

@@
 The poly-phase filter bank that performs the first separation into sub-bands, introduces a 1.28~MHz periodic spectral signature, which is shown in Fig.~\ref{fig:bandpass}.
-The poly-phase filter band-pass shape is hard to remove completely, because the filters are imperfect and each sub-band is affected by leakage from signals in its adjacent sub-bands.
+The poly-phase filter band-pass shape is hard to remove completely, because of aliasing each sub-band is affected by leakage from signals in its adjacent sub-bands.
 The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent.
@@
 However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels.
+In particular, the detected fraction of RFI over frequency is about 0.5\% higher in the edge channels.
 Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.

### 2014-11-18 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Mark Waterson's comments

@@
 
-Both projects avoid the sub-bands that cover the frequency range 133.8--138.9~MHz, because the ORBCOMM low Earth orbiting (LEO) satellites transmit in these sub-bands.
+Both projects avoid the sub-bands that cover the frequency range 133.8--138.9~MHz, because the ORBCOMM low-Earth-orbiting (LEO) satellites transmit in these sub-bands.
 When these frequencies are observed, the sub-bands are often so strongly contaminated that it affects imaging sensitivity.
 For completeness, we include a 16-min observation that covers this frequency range.
+Our analyses do not include frequencies above 231~MHz, although observations above 231~MHz are possible with the MWA.
+Data above 230~MHz have shown some contamination from the constellation of Milstar communication satellites.
 
-The poly-phase filter bank that performs the first separation into sub-bands, introduces a 1.28~MHz periodic spectral signature, which is shown in Fig.~\ref{fig:bandpass}.
-The poly-phase filter band-pass shape is hard to remove completely, because of aliasing each sub-band is affected by leakage from signals in its adjacent sub-bands.
+The poly-phase filter bank that performs the first separation into sub-bands introduces a 1.28~MHz periodic spectral signature.
+As shown in Fig.~\ref{fig:bandpass}, the poly-phase filter band-pass shape is hard to remove completely, because of aliasing each sub-band is affected by leakage from signals in its adjacent sub-bands.
 The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent.
@@
 Therefore, DTV interference has to be detected with another method.
-Additionally, some channels in the FM radio band show standard deviations that deviate slightly as a result of the smaller amount of available data after flagging and possibly because of RFI leakage.
+Additionally, some channels in the FM radio band show higher standard deviations as a result of the smaller amount of available data after flagging and possibly because of RFI leakage.
 Nevertheless, because the effect is small these frequencies can be calibrated and imaged without a problem.
@@
 For deep-integration projects, such as the EoR projects, it might be that low-level RFI will show up at lower noise levels.
-One way to increase the sensitivity of the flagger, is by executing the flagger on the summed power of multiple baselines.
-Fig.~\ref{fig:156_7} shows that this increases the detectability of an RFI event significantly.
+One way to increase the sensitivity of the flagger would be to operate on the summed power of multiple baselines.
+Fig.~\ref{fig:156_7} shows that this increases the detectability of certain RFI events significantly.
 

### 2014-11-18 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Implemented Bryan Gaensler's comments

@@
 \def\CfA{$^{15}$}
-\def\UMelbourne{$^{16}$}
-\def\Victoria{$^{17}$}
+\def\Victoria{$^{16}$}
+\def\UMelbourne{$^{17}$}
 \def\Tata{$^{18}$}
@@
 A.~P.~Beardsley\UW,
-M.~E.~Bell\CASS,
+M.~E.~Bell\CASS$^,$\CAASTRO,
 G.~Bernardi\SKASA,
 J.~D.~Bowman\ASU, 
-F.~Briggs\ANU,
+F.~Briggs\ANU$^,$\CAASTRO,
 J.~R.~Callingham\USydney$^,$\CAASTRO,
@@
 L.~J.~Greenhill\CfA,
-P.~Hancock\Curtin,
-K.~Han-Seek\UMelbourne,
+P.~Hancock\Curtin$^,$\CAASTRO,
 B.~J.~Hazelton\UW,
@@
 A.~Kapinska\UWA$^,$\CAASTRO,
+H.-S.~Kim\UMelbourne$^,$\CAASTRO,
 P.~Kittiwisit\ASU,
 E.~Lenc\USydney$^,$\CAASTRO,
-J.~Line\UMelbourne,
+J.~Line\UMelbourne$^,$\CAASTRO,
 A.~Loeb\CfA,
 C.~J.~Lonsdale\Haystack, 
-B.~McKinley\ANU$^,$\CAASTRO,
+B.~McKinley\ANU$^,$\CAASTRO$^,$\UMelbourne,
 S.~R.~McWhirter\Haystack,
@@
 T.~Prabu\RRI, 
-P.~Procopio\UMelbourne,
-J.~Riding\UMelbourne,  
+P.~Procopio\UMelbourne$^,$\CAASTRO,
+J.~Riding\UMelbourne$^,$\CAASTRO,
 N.~Udaya~Shankar\RRI, 
@@
 K.~S.~Srivani\RRI,
-L.~Staveley-Smith\UWA,
+L.~Staveley-Smith\UWA$^,$\CAASTRO,
 R.~Subrahmanyan\RRI$^,$\CAASTRO, 
@@
 C.~Wu\UWA,
-J.~S.~Wyithe\UMelbourne,
+J.~S.~Wyithe\UMelbourne$^,$\CAASTRO,
 Q. Zheng\Victoria
@@
 \CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
+\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
 \UMelbourne{}School of Physics, University of Melbourne, Parkville, VIC 3010, Australia\\
-\Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
 \Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India
@@
 The Murchison Widefield Array (MWA) is a new low-frequency interferometric radio telescope built in Western Australia at one of the locations of the future Square Kilometre Array (SKA).
-We describe the automated RFI detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights.
+We describe the automated radio-frequency interference (RFI) detection strategy implemented for the MWA, which is based on the \textsc{aoflagger} platform, and present 72--231-MHz RFI statistics from 10 observing nights.
 RFI detection removes 1.1\% of the data.
@@
 After RFI detection and excision, almost all data can be calibrated and imaged without further RFI mitigation efforts, including observations within the FM and DTV bands.
-The results are compared to a previously published LOFAR RFI survey.
+The results are compared to a previously published Low-Frequency Array (LOFAR) RFI survey.
 The remote location of the MWA results in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before data can be analysed.
@@
 
-While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI.
-Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}.
+While there has been some success in RFI mitigation by actual removal of the interference while retaining the underlying data \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI.
+Therefore, a common approach is detection and flagging of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg-2006,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}.
 The consequences of this approach are that a certain fraction of data is lost due to interference, and frequencies that are continuously occupied by transmitters cannot be observed.
-It is important to analyse the impact of RFI on a specific instrument to optimise the RFI mitigation approach.
+It is important to analyse the impact of RFI on a specific instrument to optimize the RFI mitigation approach.
 An increased understanding of the RFI situation will also help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; choosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.
@@
 In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}).
-The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; one of its main science drivers is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
+The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; one of its main science drivers is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
 To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory (MRO) in the Murchison region of Western Australia.
@@
 The overlap in frequency gives an opportunity to compare the observatory sites and hardware designs with respect to the RFI impact.
-So far, initial observations with the MWA have produced scientific results without RFI issues \citep{mwacs-2014, mckinley-fornaxa-2014, hindson-cluster-emission-2014}, which of course is not surprising given its remote location.
-PAPER has shown that their radio environment in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, one magnitude away from the expected EoR signal strength.
-However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER \citep{parsons-paper-eorlimit-2014}, which makes it hard to identify low-level RFI.
+So far, initial observations with the MWA have produced scientific results without RFI issues \citep{hurley-walker-mwacs-2014, mckinley-fornaxa-2014, hindson-cluster-emission-2014}, which of course is not surprising given its remote location.
+\citet{parsons-paper-eorlimit-2014} have shown that the radio environment for PAPER in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, two orders of magnitude (in $\textrm{mK}^2$) away from the expected EoR signal strength.
+However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER, which makes it hard to identify low-level RFI.
 \citet{lofar-radio-environment} show that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is located in a populated area.
@@
 Several observatories or projects have designed pipelines that include more advanced RFI mitigation.
-Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}).
+Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); \textsc{pieflag} \citep{pieflag-middelberg-2006} and \textsc{mirflag} \citep{lenc-mirflag-2010} mostly used for the Australia Telescope Compact Array (ATCA); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}).
 For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used.
@@
 \caption{Correlator output RMS with respect to frequency in an MWA high-frequency observation, calculated over all cross-correlated baselines and 112 seconds of data.
-In this band, the band-pass shows two large discontinuities over frequency, because the MWA receivers apply different digital gains at different frequencies to minimise quantisation noise.
-The 1.28-MHz pattern is due to the imperfect poly-phase filter.
+In this band, the band-pass shows two large discontinuities over frequency, because the MWA receivers apply different digital gains at different frequencies to minimize quantization noise.
+The 1.28~MHz sub-bands have already been corrected for the band-pass shape of the poly-phase filter, but a residual 1.28-MHz pattern is visible due to aliasing.
 }
@@
 \textsc{aoflagger} is a general-purpose RFI flagging tool developed originally for LOFAR \citep{lofar-radio-environment}.
-Specific customisations, such as changing the threshold levels and expected smoothness of good data, can be made for different telescopes to optimize the detection for different band-pass shapes, time and frequency resolutions, and fields of view.
+Specific customizations, such as changing the threshold levels and expected smoothness of good data, can be made for different telescopes to optimize the detection for different band-pass shapes, time and frequency resolutions, and fields of view.
 Strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA-specific strategy which is used in this work.
@@
 For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation.
-This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimise quantisation noise.
+This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimize quantization noise.
 An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}.
@@
 \section{Data description} \label{ch:data-description}
-We analyse the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall).
+We analyse the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (Wayth et al., in prep.).
 The nights of observations are listed in Table~\ref{tbl:obs-list}.
@@
 With the exception of the ORBCOMM frequencies and DTV affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts.
-This has been verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{hindson-cluster-emission-2014, mwacs-2014, murphy-mwa-exoplanets-2014}.
+This has been verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{hindson-cluster-emission-2014, hurley-walker-mwacs-2014, murphy-mwa-exoplanets-2014}.
 
@@
 Because of the differences in resolution and the beam forming, it is hard to compare the LOFAR and MWA environments based on detected RFI levels.
-Nevertheless, for both telescopes, the automatic detection strategies have been optimised such that as little data as possible are thrown away, but to be sufficient for further data reduction.
+Nevertheless, for both telescopes, the automatic detection strategies have been optimized such that as little data as possible are thrown away, but to be sufficient for further data reduction.
 These values can therefore be interpreted and compared as being the minimum loss of data due to RFI.
@@
 However, a benefit of the remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DTV bands.
-Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}.
+Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,hurley-walker-mwacs-2014}.
 Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR.

### 2014-11-21 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Comments by Jonnie Pober

@@
 So far, initial observations with the MWA have produced scientific results without RFI issues \citep{hurley-walker-mwacs-2014, mckinley-fornaxa-2014, hindson-cluster-emission-2014}, which of course is not surprising given its remote location.
-\citet{parsons-paper-eorlimit-2014} have shown that the radio environment for PAPER in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, two orders of magnitude (in $\textrm{mK}^2$) away from the expected EoR signal strength.
-However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER, which makes it hard to identify low-level RFI.
+\citet{parsons-paper-eorlimit-2014} have shown that the radio environment for PAPER in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit for the EoR brightness temperature at one scale and redshift, two orders of magnitude (in $\textrm{mK}^2$) away from the expected EoR signal strength.
 \citet{lofar-radio-environment} show that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is located in a populated area.
@@
 
-Recently, experiments to use the Moon as a calibrator for EoR experiments have shown that the reflection of the Moon of terrestrial transmitters complicates such an experiment \citep{mckinley-moon-2013}.
+Recently, experiments to use the Moon as a calibrator for EoR experiments have shown that the reflection of terrestrial transmitters by the Moon complicates such an experiment \citep{mckinley-moon-2013}.
 However, reflections are spatially restricted to the centre of the Lunar disk, and the high resolution of LOFAR allows the separation of the reflected and intrinsic power from the Moon \citep{vedantham-2014-todo}.
@@
 After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS.
-For more traditional data processing with e.g.
-the Common Astronomy Software Applications (\textsc{casa}; \citealt{casa-2008}), \textsc{miriad} \citep{miriad-sault-1995} and/or \textsc{wsclean} \citep{offringa-wsclean-2014}, the \textsc{cotter} output is set to either the \textsc{casa} or \textsc{uv-fits} format.
+For data processing with the Common Astronomy Software Applications (\textsc{casa}; \citealt{casa-2008}), \textsc{miriad} \citep{miriad-sault-1995}, \textsc{wsclean} \citep{offringa-wsclean-2014} and/or the Fast Holographic Deconvolution software (\textsc{fhd}, \citealt{fhd-sullivan-2012}), the \textsc{cotter} output is set to either the \textsc{casa} or \textsc{uv-fits} format.
 The output is then directly readable by the most common astronomical software packages.
@@
 These will be referred to as the EoR low and high bands, respectively.
-In these observations, the selected EoR field is tracked by the primary beam, such that the sensitivity towards the field is maximal.
+For the EoR observations, the beam formers are changed every 30~minutes to track the selected field with the primary beam, such that the sensitivity towards the field is maximal.
 The GLEAM survey covers 72.3--230.8~MHz split into 5 bands of 30.72~MHz.
@@
 However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels.
-In particular, the detected fraction of RFI over frequency is about 0.5\% higher in the edge channels.
+In particular, the detected fraction of RFI over frequency is about 0.5\% higher in the edge channels after flagging 80~kHz of the edges.
 Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.
@@
 The latter has been left out because it is affected by DTV.
-The horizontal gray line represents the false-positives rate of the RFI detection.}
+The horizontal gray line represents the false-positives rate of the RFI detection.
+The RFI fractions are consistently higher than the false-positives rate because of transient broad-band RFI.}
 \label{fig:rfi-per-band}
@@
 \caption{Visibility amplitude distributions with logarithmic axes.
-The dashed line represent fits to the function $N=\beta S^\alpha$ over a reasonably constant part of the tail (selected by eye) of the RFI distribution, resulting in $\alpha=-1.43$ for GLEAM, $\alpha=-4.47$ for EoR high, and $\alpha=-1.33$ for EoR low.}
+The dashed line represent fits to the function $N=\beta S^\alpha$ over a reasonably constant part of the tail (selected by eye) of the RFI distribution, resulting in $\alpha=-1.37$ for GLEAM, $\alpha=-4.47$ for EoR high, and $\alpha=-1.33$ for EoR low.}
 \label{fig:lognlogs}
@@
 The GLEAM and EoR low-band observations do show a small power-law component but the fitted exponent depends strongly on what part of the tail is selected.
-The selected regions result in $N\propto S^{-1.43}$ and $N\propto S^{-1.33}$ for the GLEAM and EoR-low curves, respectively.
+The selected regions result in $N\propto S^{-1.37}$ and $N\propto S^{-1.33}$ for the GLEAM and EoR-low curves, respectively.
 Due to the small region over which the power laws hold, as well as the possible high error in the fits because of the subjective data selection, it is hard to infer if a uniform distribution of transmitters is observed.
@@
 
+Projects that try to detect very weak signals, such as the EoR projects, have to be careful that a (non)detection has not been affected by RFI leakage.
+One consideration is the storage of data: the symptoms of leaked RFI are harder to identify once visibilities have been averaged or compression techniques such as Delay/Delay-Rate filtering \citep{parsons-paper-eorlimit-2014} have been applied.
+Storing more information allows better verification of a detection and, if necessary, increases the chance of successfully performing further RFI excision.
+
 \section*{Acknowledgements}

### 2014-11-24 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Implemented Anna's changes

@@
 M.~Johnston-Hollitt\Victoria,
-A.~Kapinska\UWA$^,$\CAASTRO,
+A.~D.~Kapi\'{n}ska\UWA$^,$\CAASTRO,
 H.-S.~Kim\UMelbourne$^,$\CAASTRO,
@@
 \MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
-\UWA{}University of Western Australia, Crawley, WA 6009, Perth, Australia\\
+\UWA{}ICRAR, University of Western Australia, Crawley, WA 6009, Perth, Australia\\
 \CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\

### 2014-11-24 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Implemented some final comments of Melanie and some rewording of ADC discussions

@@
 The remote location of the MWA results in a substantially cleaner RFI environment compared to LOFAR's radio environment, but adequate detection of RFI is still required before data can be analysed.
-We include specific recommendations designed to make the SKA more robust to RFI, including: the availability of sufficient computing power for RFI detection; analogue-digital converters with sufficient dynamic range; a smooth band-pass response; and the capability of RFI detection at high time and frequency resolution (second and kHz-scale respectively).
+We include specific recommendations designed to make the SKA more robust to RFI, including: the availability of sufficient computing power for RFI detection; accounting for RFI in the receiver design; a smooth band-pass response; and the capability of RFI detection at high time and frequency resolution (second and kHz-scale respectively).
 \end{abstract}
@@
 From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar per-sample computational requirements as for LOFAR.
-When comparing the 115--163~MHz range, the RFI occupancy levels are small and the RFI flagging is effective for both sites.
-Therefore, the MWA's larger radio-quiet zone does not provide a significant benefit at these frequencies.
-However, a benefit of the remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station and 174-230 MHz DTV bands.
+In both cases, RFI occupancy levels are small and RFI flagging is effective.
+Nevertheless, a benefit of MWA's remote radio-quiet site is that it allows observations in the 88--108~MHz FM-station and 174-230 MHz DTV bands.
 Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,hurley-walker-mwacs-2014}.
@@
 First of all, it is clear that the SKA will need to be able to handle some amount of RFI.
-This requires computational power to perform the RFI detection, analogue-digital converters (ADCs) with sufficient dynamic range, and a signal path with headroom sufficient to avoid saturation by RFI.
+This requires computational power to perform the RFI detection, and a receiver signal path with headroom sufficient to avoid saturation by RFI.
 Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonable amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem.

### 2014-11-25 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Small details

@@
 
-\ifpdf\pdfinfo{/Title      (Analysis of the Murchison radio environment for low-frequency astronomy)
+\ifpdf\pdfinfo{/Title      (The low-frequency environment of the Murchison Widefield Array: radio-frequency interference analysis and mitigation)
                /Author     (A. R. Offringa et al.)
@@
 \caption{Visibility amplitude distributions with logarithmic axes.
-The dashed line represent fits to the function $N=\beta S^\alpha$ over a reasonably constant part of the tail (selected by eye) of the RFI distribution, resulting in $\alpha=-1.37$ for GLEAM, $\alpha=-4.47$ for EoR high, and $\alpha=-1.33$ for EoR low.}
+The dashed line represent fits to the function $N=\beta S^\alpha$ over a reasonably constant part of the tail (gray area, selected by eye) of the RFI distribution, resulting in $\alpha=-1.37$ for GLEAM, $\alpha=-4.47$ for EoR high, and $\alpha=-1.33$ for EoR low.}
 \label{fig:lognlogs}

### 2014-11-26 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Fixed Bernardi's affiliation

@@
 \def\SKASA{$^{8}$}
-\def\ASU{$^{9}$}
-\def\USydney{$^{10}$}
-\def\Haystack{$^{11}$}
-\def\RRI{$^{12}$}
-\def\MIT{$^{13}$}
-\def\UWA{$^{14}$}
-\def\CfA{$^{15}$}
-\def\Victoria{$^{16}$}
-\def\UMelbourne{$^{17}$}
-\def\Tata{$^{18}$}
+\def\Rhodes{$^{9}$}
+\def\CfA{$^{10}$}
+\def\ASU{$^{11}$}
+\def\USydney{$^{12}$}
+\def\Haystack{$^{13}$}
+\def\RRI{$^{14}$}
+\def\MIT{$^{15}$}
+\def\UWA{$^{16}$}
+\def\Victoria{$^{17}$}
+\def\UMelbourne{$^{18}$}
+\def\Tata{$^{19}$}
 %\def\UMichigan{$^{19}$}
@@
 M.~E.~Bell\CASS$^,$\CAASTRO,
-G.~Bernardi\SKASA,
+G.~Bernardi\SKASA$^,$\Rhodes$^,$\CfA,
 J.~D.~Bowman\ASU, 
@@
 \CASS{}CSIRO Astronomy and Space Science, Marsfield, NSW 2122, Australia\\
-\SKASA{}Square Kilometre Array South Africa (SKA SA), Cape Town, 7405, South Africa\\
+\SKASA{}SKA SA, 3rd Floor, The Park, Park Road, Pinelands, 7405, South Africa\\
+\Rhodes{}Department of Physics and Electronics, Rhodes University, PO Box 94, Grahamstown, 6140, South Africa\\
+\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
 \ASU{}School of Earth and Space Exploration, Arizona State University, Tempe, AZ 85287, USA\\
@@
 \UWA{}ICRAR, University of Western Australia, Crawley, WA 6009, Perth, Australia\\
-\CfA{}Harvard-Smithsonian Center for Astrophysics, Cambridge, MA 02138, USA\\
 \Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\

### 2015-01-08 articles/2014/MWA-Radio-Environment/response-to-referee.txt
Implementing changes requested by referee

In the sentece: "alpha is typically -1.5 to -1.6"

Referee: "Presumably these nominal values of alpha are determined empirically?  This should be stated explicitly."

Answer: This is both empirically found as well as mathematically derived for uniform distributions of sources with a given average transmitter height. We think explaining this is a bit too much for this paper, so we have changed this sentence to "the exponent alpha is found by Offringa et al (2013b) to be typically -1.5 to -1.6."

"This excess is caused by the multi-variate nature of the data, for example because the noise level increases when the Galaxy goes..."

Referee: "Not sure "multi-variate" is the correct term here.  Perhaps non-stationarity and non-gaussanity of RFI is also a factor."

Answer:

Referee: "Such seemingly random impulsive RFI is ubiquitous - probably caused by ESD "sparking" and relay contacts."

Answer:

Figure 6.

"Are the centre channels of the PFB bands also flagged?  Otherwise, what are narrow horizontal lines?"

Answer: This is explained below the image: "[..] The horizontal flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively."
These images are not correctly placed in the PASA-generated PDF, so this is a bit confusing.

### 2015-01-08 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Fixing ICRAR affil
Implementing changes requested by referee

@@
 \MIT{}Kavli Institute for Astrophysics and Space Research, Massachusetts Institute of Technology, Cambridge, MA 02139, USA\\
-\UWA{}ICRAR, University of Western Australia, Crawley, WA 6009, Perth, Australia\\
+\UWA{}International Centre for Radio Astronomy Research, University of Western Australia, Crawley, WA 6009, Perth, Australia\\
 \Victoria{}School of Chemical \& Physical Sciences, Victoria University of Wellington, Wellington 6140, New Zealand\\
@@
 Therefore, a common approach is detection and flagging of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg-2006,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}.
-The consequences of this approach are that a certain fraction of data is lost due to interference, and frequencies that are continuously occupied by transmitters cannot be observed.
+The consequences of this approach are that a certain fraction of data is lost due to interference, and frequency \editmark{channels} that are continuously occupied by transmitters cannot be observed.
 It is important to analyse the impact of RFI on a specific instrument to optimize the RFI mitigation approach.
@@
 In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}).
-The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; one of its main science drivers is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
+The MWA is a low-frequency array consisting of 128 tiles, each \editmark{tile comprising of a 4x4 array of dual-polarization dipoles}, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; one of its main science drivers is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
 To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory (MRO) in the Murchison region of Western Australia.
-Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture array are planned to be built in this vicinity.
+Analysing the interference \editmark{environment} of the MWA \editmark{may} improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture array are planned to be built in this vicinity.
 
@@
 \citet{parsons-paper-eorlimit-2014} have shown that the radio environment for PAPER in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit for the EoR brightness temperature at one scale and redshift, two orders of magnitude (in $\textrm{mK}^2$) away from the expected EoR signal strength.
-\citet{lofar-radio-environment} show that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is located in a populated area.
+\citet{lofar-radio-environment} show that for regular observations the radio environment of LOFAR does not pose \editmark{unsurmountable} issues, even though LOFAR is located in a populated area.
 This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR.
-They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise.
+They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} conclude that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise.
 However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.
@@
 RFI detection, often referred to as ``data flagging'', is one of the first steps in processing the data from any interferometer.
-It is important to perform initial RFI detection and excision at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}.
+\editmark{One measure of the performance of an RFI detection method is its accuracy, which is often quantified by the average number of false-positive and true-positive detections resulting from the method.} It is important to perform initial RFI detection and excision at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}.
 Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays.
-In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate.
+In some projects, simple \editmark{amplitude} thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate.
 For example, a $3\sigma$ threshold is used for analysing PAPER data in \citet{parsons-paper-eorlimit-2014}.
@@
 \textsc{aoflagger} is a general-purpose RFI flagging tool developed originally for LOFAR \citep{lofar-radio-environment}.
-Specific customizations, such as changing the threshold levels and expected smoothness of good data, can be made for different telescopes to optimize the detection for different band-pass shapes, time and frequency resolutions, and fields of view.
+Specific customizations, such as changing the threshold levels and expected smoothness of good data, can be made for different telescopes to optimize the detection \editmark{accuracy} for different band-pass shapes, time and frequency resolutions, and fields of view.
 Strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA-specific strategy which is used in this work.
-In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency axes.
+In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency \editmark{domains}.
 Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}.
@@
 The MWA and LOFAR strategies differ qualitatively in one aspect.
-For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation.
+For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth \editmark{into 48 equal sub-bands and dividing each sub-band} by its Winsorized standard deviation \editmark{(see \citet{variance-estimates} for an explanation of Winsorized statistics}).
 This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimize quantization noise.
 An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}.
-The bandpass corrections are not permanently applied to the data, but only applied during flagging.
+The bandpass corrections are not permanently applied to the data, but only used during flagging.
 Normally, a per-channel calibration is performed after flagging in order to obtain accurate flux density calibration, which corrects the gain discontinuities permanently.
@@
 The \textsc{aoflagger} software provides a C++ library %TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
-that can be integrated in a pipeline, thereby minimising reading/writing of data.
-We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection.
+that can be integrated in a pipeline \editmark{such that intermediate data can be kept in memory.
+This minimizes the reading and writing} of data.
+We have written an MWA-specific preprocessing pipeline named \textsc{cotter}\footnote{\editmark{Etymology: Cotter is a geographical area around Cotter River, near Mount Stromlo Observatory in Canberra.}} that uses the \textsc{aoflagger} library for RFI detection.
 RFI detection is only performed on cross-correlations.
-Auto-correlations are normally removed, because they are not used in imaging.
-In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or \textsc{uv-fits} files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $u,v,w$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency and/or time to reduce the data volume.
+Auto-correlations are normally \editmark{ignored}, because they are not used in imaging.
+In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or \textsc{uv-fits} files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $u,v,w$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the \editmark{visibilities} in frequency and/or time to reduce the data volume.
 It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}.
@@
 MWA observations are split into snapshots of a few minutes by the correlator.
-The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID.
+The MWA archive stores the raw correlator outputs for \editmark{each} snapshot as an observation that can be referenced by its observation ID.
 For more details about the correlator, see \citet{ord-2014-mwa-fpga-gpu}.
@@
 One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline.
-The SumThreshold and SIR operator algorithms that are used by the flagging strategy use statistics over large time and frequency ranges.
-Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible.
-However, a typical snapshot is about 50 GB in size and without increasing reading and writing it requires 50 GB of memory to perform flagging on the full data.
-When less memory is available, \textsc{cotter} will split the observation in different time segments and flags those independently.
+The SumThreshold and SIR operator algorithms that are used by the flagging strategy use statistics \editmark{calculated} over large time and frequency ranges.
+Therefore, \editmark{detection} accuracy is improved when the time-frequency flagging intervals are as large as possible.
+However, a typical snapshot is about 50 GB in size and without \editmark{increasing disk I/O overhead} it requires 50 GB of memory to perform flagging on the full data.
+When less memory is available, \textsc{cotter} will split the observation \editmark{into a number of shorter time segments and flag these independently}.
 This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem.
-Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy.
+Partitioning the data has the undesirable \editmark{consequence} that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy.
 As not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been set up that runs a partial \textsc{cotter} preprocessing on all observations.
 In this use-case, \textsc{cotter} is run in a flagging-only mode on the large-memory machine.
-The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to apply the RFI detection from the first run and convert the raw files to his/her preferred output.
+The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to apply the RFI detection from the first run and convert the raw files to his/her preferred output \editmark{format}.
 
@@
 This helps to prevent time/frequency decorrelation.
+\editmark{The central time/frequency of an output sample is set to the time/frequency mean of the corresponding input samples, independently of what input samples are flagged.}
 
@@
 \caption{List of observations used in the analyses.
-The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (which includes false positives).
+The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (including false positives, \editmark{excluding data loss caused by band edges}).
 Occasionally, interference from digital TV (DTV) is visible.
@@
 We analyse the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (Wayth et al., in prep.).
-The nights of observations are listed in Table~\ref{tbl:obs-list}.
+The  \editmark{schedule of night-time observations is listed} in Table~\ref{tbl:obs-list}.
 
 The MWA EoR project observes primarily in the 138.9--197.7~MHz range, covering the HI 21-cm line with redshift 6.1--9.2.
-Because the MWA has a 30.72~MHz instantaneous bandwidth, this range has been split up so that each night observes with a central frequency of either 154.2 or 182.4~MHz, with an overlap of 2.6 MHz.
-These will be referred to as the EoR low and high bands, respectively.
-For the EoR observations, the beam formers are changed every 30~minutes to track the selected field with the primary beam, such that the sensitivity towards the field is maximal.
+The MWA has a 30.72~MHz instantaneous bandwidth, and \editmark{observing at two different central frequencies is therefore required to cover the desired EoR range.
+Observations are centred on either 154.2 or 182.4~MHz, with an overlap of 2.6~MHz.
+The bands covered by these central frequencies} will be referred to as the EoR low and high bands, respectively.
+For the EoR observations, the beam formers are changed every 30~minutes to track the selected field with the primary \editmark{tied-array} beam, such that the sensitivity towards the field is maximal.
 The GLEAM survey covers 72.3--230.8~MHz split into 5 bands of 30.72~MHz.
@@
 Our analyses do not include frequencies above 231~MHz, although observations above 231~MHz are possible with the MWA.
-Data above 230~MHz have shown some contamination from the constellation of Milstar communication satellites.
+Data above 230~MHz have shown \editmark{contamination} from the constellation of Milstar communication satellites.
 
@@
 However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels.
-In particular, the detected fraction of RFI over frequency is about 0.5\% higher in the edge channels after flagging 80~kHz of the edges.
+In particular, the detected fraction of RFI over frequency is about 0.5\% higher \editmark{(i.e., $\sim1.5\%$ instead of $\sim1\%$)} in the edge channels after flagging 80~kHz of the edges.
 Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.
@@
 The latter is the only night affected by interference from digital TV (DTV), and will be analysed later in this section.
-The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data.
+\editmark{RFI occupancy is calculated as the percentage of discrete visibilities within the frequency range that are detected as RFI by the flagger.} The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data.
 Excluding the RFI from DTV, the EoR high band is slightly cleaner than the EoR low band, and its worst subband at 188.2 MHz has 1.03\% occupancy.
@@
 Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability that DTV interference occurs is roughly 3\%.
-The bands involved are RF 5, 6, 7, 11 and 12, each of 7 MHz.
+The bands involved are RF 5, 6, 7, 11 and 12, each of 7~MHz \editmark{bandwidth}.
 Therefore, when DTV interference occurs, it affects 35 of the 159~MHz bandwidth.
-The visibility ratios that are lost because of DTV RFI are detailed in ``DTV'' column of Table~\ref{tbl:obs-list}.
-Because these frequencies overlap with the 21-cm HI frequencies redshifted to the EoR, in reality these frequencies are observed more than average, so the actual loss is somewhat higher.
+The visibility ratios that are lost because of DTV RFI are detailed in \editmark{the} ``DTV'' column of Table~\ref{tbl:obs-list}.
+Because these frequencies overlap with the 21-cm HI frequencies redshifted to the EoR, in reality these frequencies are \editmark{observed more often}, so the actual loss is somewhat higher.
 
@@
 \subsection{Distribution analysis}
-Using LOFAR, \citet{offringa-rfi-distributions} show that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail described by $N\propto S^\alpha$, where $N$ is the rate of visibilities, $S$ is the visibility brightness and the exponent $\alpha$ is typically -1.5 to -1.6.
+Using LOFAR, \citet{offringa-rfi-distributions} show that when \editmark{a uniform spatial distribution with sufficient} interfering transmitters is observed, the distribution of the visibility brightness will have a power-law tail described by $N\propto S^\alpha$, where $N$ is the \editmark{differential number of visibilities}, $S$ is the visibility brightness and the exponent $\alpha$ \editmark{is found by \citet{offringa-rfi-distributions} to be} typically -1.5 to -1.6.
 For the GLEAM, EoR high-band and EoR low-band observations studied here, the distributions are shown in Fig.~\ref{fig:lognlogs}.
@@
 The distribution curves of RFI-detected samples do not show a very significant power-law tail.
-Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}.
-The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N\propto S^{-4.47}$.
+Fits over regions selected by eye have been \editmark{overlaid in} Fig.~\ref{fig:lognlogs}.
+The EoR high-band observes too little RFI to generate such a \editmark{power-law tail}, which is reflected by its fit of $N\propto S^{-4.47}$.
 The GLEAM and EoR low-band observations do show a small power-law component but the fitted exponent depends strongly on what part of the tail is selected.
 The selected regions result in $N\propto S^{-1.37}$ and $N\propto S^{-1.33}$ for the GLEAM and EoR-low curves, respectively.
-Due to the small region over which the power laws hold, as well as the possible high error in the fits because of the subjective data selection, it is hard to infer if a uniform distribution of transmitters is observed.
+Due to the small region over which the power laws hold, as well as the possible high error in the fits because of the subjective data selection, it is hard to infer if \editmark{the transmitters have a uniform spatial distribution.}
 
 In the ideal case, the residual visibility distribution would show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}).
-The residual curves in Fig.~\ref{fig:lognlogs} do not go down as quickly as a Rayleigh curve would, and therefore the distributions have a slight excess of samples with higher amplitudes.
+The residual curves in Fig.~\ref{fig:lognlogs} \editmark{do not fall off with RFI power as quickly} as a Rayleigh curve would, and therefore the distributions have a slight excess of samples with higher amplitudes.
 This excess is caused by the multi-variate nature of the data, for example because the noise level increases when the Galaxy goes through the beam.
@@
 Fig.~\ref{fig:2m-amateur-band} shows two examples of RFI events in the same EoR low-band snapshot: the top panels display the Stokes~I values of a single correlation, in which a transmitter has been observed in the 2-m amateur band (146 MHz).
-This is the worst contamination in our data sets by this transmitter, and there are also snapshots in which the transmitter is not visible at all.
+This is the worst \editmark{example of} contamination in our data sets by this transmitter, and there are snapshots in which the transmitter is not visible at all.
 This variability might be caused by intrinsic variation, movement of the transmitter or varying propagation conditions.
@@
 Occasionally, DTV RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}.
-As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}.
+As can be seen in the \editmark{right-hand} plot of Fig.~\ref{fig:dvb-burst}, such a \editmark{brief interference event} is well flagged by the \textsc{aoflagger}.
 Another transient broad-band example is displayed in Fig.~\ref{fig:broadband-pulse}, which shows strong broad-band pulses of a second in length.
@@
 Fig.~\ref{fig:156_7} shows a longer RFI contamination at 156.66~MHz that is hardly visible in a single correlation.
-The \textsc{aoflagger} detects this RFI partially (Fig.~\ref{fig:156_7}, top-right plot), but when plotting the standard deviation over all correlations in a dynamic spectrum, it is evident that this RFI transmitter extends beyond what is detected (Fig.~\ref{fig:156_7}, bottom-left plot).
+The \textsc{aoflagger} detects this RFI partially (Fig.~\ref{fig:156_7}, top-right plot), but when plotting the standard deviation over all correlations in a dynamic spectrum, it is evident that \editmark{this RFI event extends} in time beyond what is detected (Fig.~\ref{fig:156_7}, bottom-left plot).
 The detection becomes more complete when the \textsc{aoflagger} is executed on the standard deviations over all baselines (Fig.~\ref{fig:156_7}, bottom-right plot).
@@
 \subsection{Computational performance of \textsc{cotter}}
-Because \textsc{cotter} processes the data at high resolution, its computational performance is an important consideration.
-A major contribution to the runtime is the reading and writing of the data, and the runtime is thus influenced by the input-output (IO) performance of the host system.
-Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent.
-To time \textsc{cotter}, we use a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5-disc RAID5 setup.
+Because \textsc{cotter} processes the data at \editmark{high time and frequency} resolution, its computational performance is an important consideration.
+A major contribution to the runtime is the reading and writing of the data, and the runtime is thus influenced by the input-output (IO) \editmark{disk} performance of the host system.
+Excluding IO, the main computational \editmark{burden} of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately \editmark{$1\%$}.
+To time \textsc{cotter}, we use a high-end desktop \editmark{computer} with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5-disc RAID5 setup.
 The wall-clock runtime for processing a single 2~min snapshot of 50~GB with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, is split up as follows: 3~min are spent on reading the data, 5~min on RFI detection, and 3.5~min on writing the data.
 Real-time processing can therefore be achieved by using 6 of such nodes in parallel.
-Assuming a 138~GFLOPS (Giga-floating-point operations per second) performance of the host computer, the RFI detection requires 25 FLOP/sample.
+Assuming a 138~GFLOPS (Giga-floating-point operations per second) performance of the host computer, the RFI detection requires 25 FLOP/\editmark{visibility}.
 When expressed as visibilities per time unit, the computational performance of the flagger is independent of the frequency resolution, time resolution and number of antennas.
-This number can therefore be extrapolated to other telescopes, although the performance of a pipeline which incorporates RFI detection will be strongly dependent on available IO performance and memory bandwidth.
+This number can therefore be extrapolated to other telescopes, although the performance of a pipeline which incorporates RFI detection will be strongly dependent on available IO performance, memory bandwidth \editmark{and other system properties}.
 
@@
 The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency sub-range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}.
-The statistics are regridded to the same frequency resolution of 48~kHz.
+The statistics are regridded to \editmark{a common frequency resolution} of 48~kHz.
 The average RFI occupancy in this frequency range is 1.65\% for the MWA, while \citet{lofar-radio-environment} reports a 3.18\% occupancy for LOFAR.
@@
 In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources; Cassiopeia~A and Cygnus~A are only visible at low elevations.
-An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations.
+An additional explanation could also be that the sidelobe behaviour \editmark{of the tiled-array beam} is also different between MWA and LOFAR due to the difference between MWA tiles and LOFAR stations.
 
-Because of the differences in resolution and the beam forming, it is hard to compare the LOFAR and MWA environments based on detected RFI levels.
+Because of the differences in resolution and the beam forming, it is hard to compare the LOFAR and MWA environments based on detected RFI \editmark{statistics}.
 Nevertheless, for both telescopes, the automatic detection strategies have been optimized such that as little data as possible are thrown away, but to be sufficient for further data reduction.
@@
 
-After RFI detection and excision, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging.
+After RFI detection and excision, power spectra from both MWA and LOFAR show smooth curves without \editmark{artefacts that can be attributed to leakage}, and \editmark{RFI does not lead to detectable artifacts in the resulting image at the thermal noise limit}.
 From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar per-sample computational requirements as for LOFAR.
@@
 We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights.
-After RFI detection and excision, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz.
+After RFI detection and excision, data can be calibrated and imaged without \editmark{artefacts visible at the thermal or confusion noise level}, with the exception of the ORBCOMM bands at 137~MHz.
 Also, DTV signals are seen $\sim3\%$ of the time, and when present make observing in the 174--195~MHz range impossible.
@@
 Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection.
-The issue of RFI has become smaller by building at a remote site, but still requires adequate mitigation.
+The issue of RFI has become smaller by building at a \editmark{radio quiet} site, but still requires adequate mitigation.
 When observing continuously with the MWA, a few fast computing nodes are permanently required for real-time RFI detection.
@@
 First of all, it is clear that the SKA will need to be able to handle some amount of RFI.
-This requires computational power to perform the RFI detection, and a receiver signal path with headroom sufficient to avoid saturation by RFI.
+This requires computational power to perform the RFI detection, and a receiver signal path with headroom sufficient to avoid \editmark{gain compression} by RFI.
 Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonable amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem.
@@
 Since strong RFI is seen practically everywhere, and because the apparent strength of RFI events will follow a power-law distribution, many faint transmitters will interfere with observations of any (terrestrial) radio observatory.
-Using algorithms with low sensitivity will detect fewer of these, and thus result in RFI becoming more quickly a problem in long-integration projects.
+Using algorithms with low sensitivity will detect fewer of these, and thus result in RFI becoming more quickly a problem in \editmark{deep integration projects}.
 
@@
 To remove DTV, a second detection round is required, and the current methodology to handle DTV RFI is to delete all affected snapshots, or even the entire night.
-This requires hardly any computational power, because the visibility statistics are collected in \textsc{cotter}.
+This requires hardly any computational power, because the \editmark{required} visibility statistics are collected in \textsc{cotter}.
 For deep-integration projects, such as the EoR projects, it might be that low-level RFI will show up at lower noise levels.

### 2015-01-14 articles/2014/MWA-Radio-Environment/response-to-referee.txt
Referee report implemented

@@
-In the sentece: "alpha is typically -1.5 to -1.6"
+Dear referee and editor,
+
+We would like to thank the referee for his comments on our article.
+We agree with most suggestions by the referee.
+For a few comments by the referee, we think no changes were required or we have made a change slightly different as suggested.
+These cases are detailed below and we think are minor differences.
+Other than those, all suggestions were implemented, and we think the paper is ready for publication.
+All changes have been marked in the article with a red/bold font.
+
+Page 7
+"Because the AOFLAGGER determines its thresholding levels from the data, and because it needs to be insensitive to steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.
+1), it is insensitive to RFI covering such a broad spectrum."
+
+Referee: "Because of the increase in spread-spectrum transmissions, this is a quite serious weakness in AOFLAGGER.
+The process described in the next paragraph might be included into the pipeline in a generic way to detect any spread-spectrum RFI."
+
+and from the decision letter:
+"AOFLAGGER fails to detect what is becoming an increasingly troublesome source of RFI: persistent spread-spectrum signals.
+A methodology for supplementing AOFLAGGER using COTTER statistics is suggested, but some discussion of how AOFLAGGER may be modified to handle this class of RFI would be useful.
+"
+
+Answer: The extra flagging step runs fast compared to AOFlagger (which is described in the paper).
+It is therefore not necessary to include this step inside Cotter or AOFlagger -- a reason for inclusion in the initial pre-processing pipeline is when it needs the data at high resolution, but that's not the case.
+While this RFI is serious, it's not a serious weakness in the flagger that such RFI is not immediately flagged; the initial flagger is aimed at flagging out 'high-resolution' RFI that occupies a few channels or timesteps, and a second flagging step can recognize broadband RFI at a lower data resolution.
+We think this is described well enough in the paper and no further changes are required.
+
+Page 8
+Sentence: "[..] and the exponent alpha is typically -1.5 to -1.6"
 
@@
 
-Answer: This is both empirically found as well as mathematically derived for uniform distributions of sources with a given average transmitter height.
-We think explaining this is a bit too much for this paper, so we have changed this sentence to "the exponent alpha is found by Offringa et al (2013b) to be typically -1.5 to -1.6."
+Answer: In Offringa et al (2013b), this is both empirically found as well as mathematically derived to be true for uniform distributions of sources with a given average transmitter height.
+We think explaining this completely is not necessary, so instead we now refer to the paper: "the exponent alpha is found by Offringa et al (2013b) to be typically -1.5 to -1.6."
+and explain the uniformness property: "Using LOFAR, Offringa et al.
+(2013b) show that when a uniform spatial distribution with sufficient interfering transmitters is observed, the distribution of the visibility brightness will have a power-law tail [..]"
 
-"This excess is caused by the multi-variate nature of the data, for example because the noise level increases when the Galaxy goes..."
-
-Referee: "Not sure "multi-variate" is the correct term here.  Perhaps non-stationarity and non-gaussanity of RFI is also a factor."
-
-Answer:
+Page 9
+"[..] a few weak broad-band pulses are observed in almost every 2-min snapshot.
+These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power on all baselines is added together.
+We currently do not know what their origin is"
 
@@
 
-Answer:
+Answer: We agree with the referee, but rather don't speculate to what is the cause.
 
-Figure 6.
+Page 10, Figure 6.
+Referee: "Are the centre channels of the PFB bands also flagged?  Otherwise, what are narrow horizontal lines?"
 
-"Are the centre channels of the PFB bands also flagged?  Otherwise, what are narrow horizontal lines?"
+Answer: This is explained in the caption below the image (which is placed at a somewhat incorrect position by the PASA-generated PDF, so that might have been a bit confusing) : "[..] The horizontal flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively."
 
-Answer: This is explained below the image: "[..] The horizontal flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively."
-These images are not correctly placed in the PASA-generated PDF, so this is a bit confusing.
+Page 12
+"In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources; Cassiopeia A and Cygnus A are only visible at low elevations."
+
+Referee: "The southern hemisphere has its own strong discrete sources - are these not evident?"
+
+Answer: At these frequencies, Cyg A and Cas A are almost an order of magnitude larger (~5-10 KJy) than the strongest Southern sources such as M1 (~1 KJy).

### 2015-01-14 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Referee report implemented

@@
 This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR.
-They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} conclude that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise.
+They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} \editmark{conclude} that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise.
 However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.
@@
 An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}.
-The bandpass corrections are not permanently applied to the data, but only used during flagging.
+The bandpass corrections are not permanently applied to the data, but only \editmark{used} during flagging.
 Normally, a per-channel calibration is performed after flagging in order to obtain accurate flux density calibration, which corrects the gain discontinuities permanently.
@@
 The MWA EoR project observes primarily in the 138.9--197.7~MHz range, covering the HI 21-cm line with redshift 6.1--9.2.
-The MWA has a 30.72~MHz instantaneous bandwidth, and \editmark{observing at two different central frequencies is therefore required to cover the desired EoR range.
-Observations are centred on either 154.2 or 182.4~MHz, with an overlap of 2.6~MHz.
+The MWA has a 30.72-MHz instantaneous bandwidth, \editmark{which makes it necessary to observe at two different frequencies to cover the desired EoR bandwidth.
+An observing night is centred on either 154.2 or 182.4~MHz, which results in an overlap of 2.6~MHz.
 The bands covered by these central frequencies} will be referred to as the EoR low and high bands, respectively.
@@
 The latter is the only night affected by interference from digital TV (DTV), and will be analysed later in this section.
-\editmark{RFI occupancy is calculated as the percentage of discrete visibilities within the frequency range that are detected as RFI by the flagger.} The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data.
+\editmark{RFI occupancy is calculated as the percentage of discrete visibilities that are detected as RFI by the flagger at the resolution of the correlator output.} The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data.
 Excluding the RFI from DTV, the EoR high band is slightly cleaner than the EoR low band, and its worst subband at 188.2 MHz has 1.03\% occupancy.
@@
 The residual curves in Fig.~\ref{fig:lognlogs} \editmark{do not fall off with RFI power as quickly} as a Rayleigh curve would, and therefore the distributions have a slight excess of samples with higher amplitudes.
-This excess is caused by the multi-variate nature of the data, for example because the noise level increases when the Galaxy goes through the beam.
+This excess is caused by the \editmark{variable} nature of the data, for example because the noise level increases when the Galaxy goes through the beam.
 The EoR-high distribution has some extra features in its tails that are not smooth.
@@
 Fig.~\ref{fig:2m-amateur-band} shows two examples of RFI events in the same EoR low-band snapshot: the top panels display the Stokes~I values of a single correlation, in which a transmitter has been observed in the 2-m amateur band (146 MHz).
-This is the worst \editmark{example of} contamination in our data sets by this transmitter, and there are snapshots in which the transmitter is not visible at all.
+This is the worst \editmark{example of} contamination in our data sets by this transmitter, and there are \editmark{snapshots} in which the transmitter is not visible at all.
 This variability might be caused by intrinsic variation, movement of the transmitter or varying propagation conditions.
@@
 \section*{Acknowledgements}
+\editmark{This scientific work makes use of the NCI National Facility in Canberra, Australia, which is supported by the Australian Commonwealth Government. }
+
 This scientific work makes use of the Murchison Radio-astronomy Observatory, operated by CSIRO.

### 2015-01-16 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Removing edit marks and other comments to prepare for Arxiv submission

--- 2014-11-26 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
@@
 \bibliographystyle{apj}
-%\usepackage[utf8]{inputenc}
 \usepackage{graphicx}
@@
 \usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
-%\usepackage{xspace}
 \usepackage{subcaption}
-
-%\usepackage{fixltx2e} % fixes placing of image positions
 
@@
 
-\newcommand{\editmark}[1]{#1}
+%\newcommand{\editmark}[1]{#1}
 %\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
@@
 \def\Tata{$^{19}$}
-%\def\UMichigan{$^{19}$}
-%\def\NRAO{$^{20}$}
 
@@
 P.~Carroll\UW,
-%A.~de~Oliveira-Costa
 A.~A.~Deshpande\RRI, 
@@
 \ANU{}Research School of Astronomy and Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
-%\CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics, Australian National University, Canberra, ACT 2611, Australia \\
 \CAASTRO{}ARC Centre of Excellence for All-sky Astrophysics (CAASTRO) \\
@@
 \Tata{}National Centre for Radio Astrophysics, Tata Institute for Fundamental Research, Pune 411007, India
-%\UMichigan{}University of Michigan, Ann Harbor, MI 48109, USA
 }
@@
 
-%\date{Accepted TODO. Received TODO; in original form TODO}
-%\pagerange{\pageref{firstpage}--\pageref{lastpage}}
-%\pubyear{2014}
-
 \label{firstpage}
@@
 Therefore, a common approach is detection and flagging of contaminated data and ignoring these samples in further data analysis \citep{statistical-rfi-removal,pieflag-middelberg-2006,post-correlation-rfi-classification,prasad-flagcal-2012,serpent-peck-2013}.
-The consequences of this approach are that a certain fraction of data is lost due to interference, and frequencies that are continuously occupied by transmitters cannot be observed.
+The consequences of this approach are that a certain fraction of data is lost due to interference, and frequency channels that are continuously occupied by transmitters cannot be observed.
 It is important to analyse the impact of RFI on a specific instrument to optimize the RFI mitigation approach.
@@
 In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}).
-The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; one of its main science drivers is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
+The MWA is a low-frequency array consisting of 128 tiles, each tile comprising of a 4x4 array of dual-polarization dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; one of its main science drivers is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}).
 To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory (MRO) in the Murchison region of Western Australia.
-Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture array are planned to be built in this vicinity.
+Analysing the interference environment of the MWA may improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture array are planned to be built in this vicinity.
 
@@
 
-%\defcitealias{lofar-radio-environment}{O2013a}
 The interferometric arrays Low-Frequency Array (LOFAR; \citealt{lofar-2013}), Precision Array to Probe the Epoch of Reionization (PAPER; \citealt{parsons-paper-eorlimit-2014}) and the Giant Metrewave Radio Telescope (GMRT; \citealt{the-gmrt-swarup-chapter-2013}) observe in approximately the same frequency range as the MWA.
@@
 \citet{parsons-paper-eorlimit-2014} have shown that the radio environment for PAPER in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit for the EoR brightness temperature at one scale and redshift, two orders of magnitude (in $\textrm{mK}^2$) away from the expected EoR signal strength.
-\citet{lofar-radio-environment} show that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is located in a populated area.
+\citet{lofar-radio-environment} show that for regular observations the radio environment of LOFAR does not pose unsurmountable issues, even though LOFAR is located in a populated area.
 This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR.
-They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise.
+They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} conclude that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise.
 However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.
@@
 RFI detection, often referred to as ``data flagging'', is one of the first steps in processing the data from any interferometer.
+One measure of the performance of an RFI detection method is its accuracy, which is often quantified by the average number of false-positive and true-positive detections resulting from the method.
 It is important to perform initial RFI detection and excision at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}.
 Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays.
-In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate.
+In some projects, simple amplitude thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate.
 For example, a $3\sigma$ threshold is used for analysing PAPER data in \citet{parsons-paper-eorlimit-2014}.
@@
 \textsc{aoflagger} is a general-purpose RFI flagging tool developed originally for LOFAR \citep{lofar-radio-environment}.
-Specific customizations, such as changing the threshold levels and expected smoothness of good data, can be made for different telescopes to optimize the detection for different band-pass shapes, time and frequency resolutions, and fields of view.
+Specific customizations, such as changing the threshold levels and expected smoothness of good data, can be made for different telescopes to optimize the detection accuracy for different band-pass shapes, time and frequency resolutions, and fields of view.
 Strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA-specific strategy which is used in this work.
-In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency axes.
+In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency domains.
 Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}.
@@
 The MWA and LOFAR strategies differ qualitatively in one aspect.
-For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth in 48 parts and dividing each part by its Winsorized standard deviation.
+For the MWA, an extra bandpass correction is added, which is performed by dividing the bandwidth into 48 equal sub-bands and dividing each sub-band by its Winsorized standard deviation (see \citet{variance-estimates} for an explanation of Winsorized statistics).
 This step is required to smooth out discontinuities due to varying digital gains over frequency that are applied by the receivers to minimize quantization noise.
 An example of these discontinuities is shown in Fig.~\ref{fig:bandpass}.
-The bandpass corrections are not permanently applied to the data, but only applied during flagging.
+The bandpass corrections are not permanently applied to the data, but only used during flagging.
 Normally, a per-channel calibration is performed after flagging in order to obtain accurate flux density calibration, which corrects the gain discontinuities permanently.
@@
 
-The \textsc{aoflagger} software provides a C++ library %TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
-that can be integrated in a pipeline, thereby minimising reading/writing of data.
-We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection.
+The \textsc{aoflagger} software provides a C++ library that can be integrated in a pipeline such that intermediate data can be kept in memory.
+This minimizes the reading and writing of data.
+We have written an MWA-specific preprocessing pipeline named \textsc{cotter}\footnote{Etymology: Cotter is a geographical area around Cotter River, near Mount Stromlo Observatory in Canberra.} that uses the \textsc{aoflagger} library for RFI detection.
 RFI detection is only performed on cross-correlations.
-Auto-correlations are normally removed, because they are not used in imaging.
-In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or \textsc{uv-fits} files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $u,v,w$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency and/or time to reduce the data volume.
+Auto-correlations are normally ignored, because they are not used in imaging.
+In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or \textsc{uv-fits} files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $u,v,w$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the visibilities in frequency and/or time to reduce the data volume.
 It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}.
@@
 MWA observations are split into snapshots of a few minutes by the correlator.
-The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID.
+The MWA archive stores the raw correlator outputs for each snapshot as an observation that can be referenced by its observation ID.
 For more details about the correlator, see \citet{ord-2014-mwa-fpga-gpu}.
@@
 One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline.
-The SumThreshold and SIR operator algorithms that are used by the flagging strategy use statistics over large time and frequency ranges.
-Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible.
-However, a typical snapshot is about 50 GB in size and without increasing reading and writing it requires 50 GB of memory to perform flagging on the full data.
-When less memory is available, \textsc{cotter} will split the observation in different time segments and flags those independently.
+The SumThreshold and SIR operator algorithms that are used by the flagging strategy use statistics calculated over large time and frequency ranges.
+Therefore, detection accuracy is improved when the time-frequency flagging intervals are as large as possible.
+However, a typical snapshot is about 50 GB in size and without increasing disk I/O overhead it requires 50 GB of memory to perform flagging on the full data.
+When less memory is available, \textsc{cotter} will split the observation into a number of shorter time segments and flag these independently.
 This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem.
-Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy.
+Partitioning the data has the undesirable consequence that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy.
 As not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been set up that runs a partial \textsc{cotter} preprocessing on all observations.
 In this use-case, \textsc{cotter} is run in a flagging-only mode on the large-memory machine.
-The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to apply the RFI detection from the first run and convert the raw files to his/her preferred output.
+The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to apply the RFI detection from the first run and convert the raw files to his/her preferred output format.
 
@@
 This helps to prevent time/frequency decorrelation.
+The central time/frequency of an output sample is set to the time/frequency mean of the corresponding input samples, independently of what input samples are flagged.
 
@@
 \caption{List of observations used in the analyses.
-The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (which includes false positives).
+The RFI column contains the fraction of visibilities that the initial \textsc{aoflagger} run has classified as RFI (including false positives, excluding data loss caused by band edges).
 Occasionally, interference from digital TV (DTV) is visible.
@@
 We analyse the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (Wayth et al., in prep.).
-The nights of observations are listed in Table~\ref{tbl:obs-list}.
+The schedule of night-time observations is listed in Table~\ref{tbl:obs-list}.
 
 The MWA EoR project observes primarily in the 138.9--197.7~MHz range, covering the HI 21-cm line with redshift 6.1--9.2.
-Because the MWA has a 30.72~MHz instantaneous bandwidth, this range has been split up so that each night observes with a central frequency of either 154.2 or 182.4~MHz, with an overlap of 2.6 MHz.
-These will be referred to as the EoR low and high bands, respectively.
-For the EoR observations, the beam formers are changed every 30~minutes to track the selected field with the primary beam, such that the sensitivity towards the field is maximal.
+The MWA has a 30.72-MHz instantaneous bandwidth, which makes it necessary to observe at two different frequencies to cover the desired EoR bandwidth.
+An observing night is centred on either 154.2 or 182.4~MHz, which results in an overlap of 2.6~MHz.
+The bands covered by these central frequencies will be referred to as the EoR low and high bands, respectively.
+For the EoR observations, the beam formers are changed every 30~minutes to track the selected field with the primary tied-array beam, such that the sensitivity towards the field is maximal.
 The GLEAM survey covers 72.3--230.8~MHz split into 5 bands of 30.72~MHz.
@@
 Our analyses do not include frequencies above 231~MHz, although observations above 231~MHz are possible with the MWA.
-Data above 230~MHz have shown some contamination from the constellation of Milstar communication satellites.
+Data above 230~MHz have shown contamination from the constellation of Milstar communication satellites.
 
@@
 However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels.
-In particular, the detected fraction of RFI over frequency is about 0.5\% higher in the edge channels after flagging 80~kHz of the edges.
+In particular, the detected fraction of RFI over frequency is about 0.5\% higher (i.e., $\sim1.5\%$ instead of $\sim1\%$) in the edge channels after flagging 80~kHz of the edges.
 Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.
@@
 The latter is the only night affected by interference from digital TV (DTV), and will be analysed later in this section.
+RFI occupancy is calculated as the percentage of discrete visibilities that are detected as RFI by the flagger at the resolution of the correlator output.
 The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data.
@@
 Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability that DTV interference occurs is roughly 3\%.
-The bands involved are RF 5, 6, 7, 11 and 12, each of 7 MHz.
+The bands involved are RF 5, 6, 7, 11 and 12, each of 7~MHz bandwidth.
 Therefore, when DTV interference occurs, it affects 35 of the 159~MHz bandwidth.
-The visibility ratios that are lost because of DTV RFI are detailed in ``DTV'' column of Table~\ref{tbl:obs-list}.
-Because these frequencies overlap with the 21-cm HI frequencies redshifted to the EoR, in reality these frequencies are observed more than average, so the actual loss is somewhat higher.
+The visibility ratios that are lost because of DTV RFI are detailed in the ``DTV'' column of Table~\ref{tbl:obs-list}.
+Because these frequencies overlap with the 21-cm HI frequencies redshifted to the EoR, in reality these frequencies are observed more often, so the actual loss is somewhat higher.
 
@@
 \subsection{Distribution analysis}
-Using LOFAR, \citet{offringa-rfi-distributions} show that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail described by $N\propto S^\alpha$, where $N$ is the rate of visibilities, $S$ is the visibility brightness and the exponent $\alpha$ is typically -1.5 to -1.6.
+Using LOFAR, \citet{offringa-rfi-distributions} show that when a uniform spatial distribution with sufficient interfering transmitters is observed, the distribution of the visibility brightness will have a power-law tail described by $N\propto S^\alpha$, where $N$ is the differential number of visibilities, $S$ is the visibility brightness and the exponent $\alpha$ is found by \citet{offringa-rfi-distributions} to be typically -1.5 to -1.6.
 For the GLEAM, EoR high-band and EoR low-band observations studied here, the distributions are shown in Fig.~\ref{fig:lognlogs}.
@@
 The distribution curves of RFI-detected samples do not show a very significant power-law tail.
-Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}.
-The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N\propto S^{-4.47}$.
+Fits over regions selected by eye have been overlaid in Fig.~\ref{fig:lognlogs}.
+The EoR high-band observes too little RFI to generate such a power-law tail, which is reflected by its fit of $N\propto S^{-4.47}$.
 The GLEAM and EoR low-band observations do show a small power-law component but the fitted exponent depends strongly on what part of the tail is selected.
 The selected regions result in $N\propto S^{-1.37}$ and $N\propto S^{-1.33}$ for the GLEAM and EoR-low curves, respectively.
-Due to the small region over which the power laws hold, as well as the possible high error in the fits because of the subjective data selection, it is hard to infer if a uniform distribution of transmitters is observed.
+Due to the small region over which the power laws hold, as well as the possible high error in the fits because of the subjective data selection, it is hard to infer if the transmitters have a uniform spatial distribution.
 
 In the ideal case, the residual visibility distribution would show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}).
-The residual curves in Fig.~\ref{fig:lognlogs} do not go down as quickly as a Rayleigh curve would, and therefore the distributions have a slight excess of samples with higher amplitudes.
-This excess is caused by the multi-variate nature of the data, for example because the noise level increases when the Galaxy goes through the beam.
+The residual curves in Fig.~\ref{fig:lognlogs} do not fall off with RFI power as quickly as a Rayleigh curve would, and therefore the distributions have a slight excess of samples with higher amplitudes.
+This excess is caused by the variable nature of the data, for example because the noise level increases when the Galaxy goes through the beam.
 The EoR-high distribution has some extra features in its tails that are not smooth.
@@
 Fig.~\ref{fig:2m-amateur-band} shows two examples of RFI events in the same EoR low-band snapshot: the top panels display the Stokes~I values of a single correlation, in which a transmitter has been observed in the 2-m amateur band (146 MHz).
-This is the worst contamination in our data sets by this transmitter, and there are also snapshots in which the transmitter is not visible at all.
+This is the worst example of contamination in our data sets by this transmitter, and there are snapshots in which the transmitter is not visible at all.
 This variability might be caused by intrinsic variation, movement of the transmitter or varying propagation conditions.
@@
 Occasionally, DTV RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}.
-As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}.
+As can be seen in the right-hand plot of Fig.~\ref{fig:dvb-burst}, such a brief interference event is well flagged by the \textsc{aoflagger}.
 Another transient broad-band example is displayed in Fig.~\ref{fig:broadband-pulse}, which shows strong broad-band pulses of a second in length.
@@
 Fig.~\ref{fig:156_7} shows a longer RFI contamination at 156.66~MHz that is hardly visible in a single correlation.
-The \textsc{aoflagger} detects this RFI partially (Fig.~\ref{fig:156_7}, top-right plot), but when plotting the standard deviation over all correlations in a dynamic spectrum, it is evident that this RFI transmitter extends beyond what is detected (Fig.~\ref{fig:156_7}, bottom-left plot).
+The \textsc{aoflagger} detects this RFI partially (Fig.~\ref{fig:156_7}, top-right plot), but when plotting the standard deviation over all correlations in a dynamic spectrum, it is evident that this RFI event extends in time beyond what is detected (Fig.~\ref{fig:156_7}, bottom-left plot).
 The detection becomes more complete when the \textsc{aoflagger} is executed on the standard deviations over all baselines (Fig.~\ref{fig:156_7}, bottom-right plot).
@@
 \subsection{Computational performance of \textsc{cotter}}
-Because \textsc{cotter} processes the data at high resolution, its computational performance is an important consideration.
-A major contribution to the runtime is the reading and writing of the data, and the runtime is thus influenced by the input-output (IO) performance of the host system.
-Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent.
-To time \textsc{cotter}, we use a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5-disc RAID5 setup.
+Because \textsc{cotter} processes the data at high time and frequency resolution, its computational performance is an important consideration.
+A major contribution to the runtime is the reading and writing of the data, and the runtime is thus influenced by the input-output (IO) disk performance of the host system.
+Excluding IO, the main computational burden of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately $1\%$.
+To time \textsc{cotter}, we use a high-end desktop computer with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5-disc RAID5 setup.
 The wall-clock runtime for processing a single 2~min snapshot of 50~GB with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, is split up as follows: 3~min are spent on reading the data, 5~min on RFI detection, and 3.5~min on writing the data.
 Real-time processing can therefore be achieved by using 6 of such nodes in parallel.
-Assuming a 138~GFLOPS (Giga-floating-point operations per second) performance of the host computer, the RFI detection requires 25 FLOP/sample.
+Assuming a 138~GFLOPS (Giga-floating-point operations per second) performance of the host computer, the RFI detection requires 25 FLOP/visibility.
 When expressed as visibilities per time unit, the computational performance of the flagger is independent of the frequency resolution, time resolution and number of antennas.
-This number can therefore be extrapolated to other telescopes, although the performance of a pipeline which incorporates RFI detection will be strongly dependent on available IO performance and memory bandwidth.
+This number can therefore be extrapolated to other telescopes, although the performance of a pipeline which incorporates RFI detection will be strongly dependent on available IO performance, memory bandwidth and other system properties.
 
@@
 The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency sub-range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}.
-The statistics are regridded to the same frequency resolution of 48~kHz.
+The statistics are regridded to a common frequency resolution of 48~kHz.
 The average RFI occupancy in this frequency range is 1.65\% for the MWA, while \citet{lofar-radio-environment} reports a 3.18\% occupancy for LOFAR.
@@
 In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources; Cassiopeia~A and Cygnus~A are only visible at low elevations.
-An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations.
+An additional explanation could also be that the sidelobe behaviour of the tiled-array beam is also different between MWA and LOFAR due to the difference between MWA tiles and LOFAR stations.
 
-Because of the differences in resolution and the beam forming, it is hard to compare the LOFAR and MWA environments based on detected RFI levels.
+Because of the differences in resolution and the beam forming, it is hard to compare the LOFAR and MWA environments based on detected RFI statistics.
 Nevertheless, for both telescopes, the automatic detection strategies have been optimized such that as little data as possible are thrown away, but to be sufficient for further data reduction.
@@
 
-After RFI detection and excision, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging.
+After RFI detection and excision, power spectra from both MWA and LOFAR show smooth curves without artefacts that can be attributed to leakage, and RFI does not lead to detectable artifacts in the resulting image at the thermal noise limit.
 From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar per-sample computational requirements as for LOFAR.
@@
 We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights.
-After RFI detection and excision, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz.
+After RFI detection and excision, data can be calibrated and imaged without artefacts visible at the thermal or confusion noise level, with the exception of the ORBCOMM bands at 137~MHz.
 Also, DTV signals are seen $\sim3\%$ of the time, and when present make observing in the 174--195~MHz range impossible.
@@
 Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection.
-The issue of RFI has become smaller by building at a remote site, but still requires adequate mitigation.
+The issue of RFI has become smaller by building at a radio-quiet site, but still requires adequate mitigation.
 When observing continuously with the MWA, a few fast computing nodes are permanently required for real-time RFI detection.
@@
 First of all, it is clear that the SKA will need to be able to handle some amount of RFI.
-This requires computational power to perform the RFI detection, and a receiver signal path with headroom sufficient to avoid saturation by RFI.
+This requires computational power to perform the RFI detection, and a receiver signal path with headroom sufficient to avoid gain compression by RFI.
 Second, \textsc{cotter} relies on the fact that a single computing node can still hold a reasonable amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem.
@@
 Since strong RFI is seen practically everywhere, and because the apparent strength of RFI events will follow a power-law distribution, many faint transmitters will interfere with observations of any (terrestrial) radio observatory.
-Using algorithms with low sensitivity will detect fewer of these, and thus result in RFI becoming more quickly a problem in long-integration projects.
+Using algorithms with low sensitivity will detect fewer of these, and thus result in RFI becoming more quickly a problem in deep integration projects.
 
@@
 To remove DTV, a second detection round is required, and the current methodology to handle DTV RFI is to delete all affected snapshots, or even the entire night.
-This requires hardly any computational power, because the visibility statistics are collected in \textsc{cotter}.
+This requires hardly any computational power, because the required visibility statistics are collected in \textsc{cotter}.
 For deep-integration projects, such as the EoR projects, it might be that low-level RFI will show up at lower noise levels.
@@
 \section*{Acknowledgements}
+This scientific work makes use of the NCI National Facility in Canberra, Australia, which is supported by the Australian Commonwealth Government.
+
 This scientific work makes use of the Murchison Radio-astronomy Observatory, operated by CSIRO.
